### 2017-04-28 ms.tex
big updates and revisions

\documentclass[DM,lsstdraft,toc]{lsstdoc}
\usepackage{graphicx}
\usepackage{url}
\usepackage{latexsym}
\usepackage{color}
% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.

\title[LSST Special Programs]{Data Management \\ and LSST Special Programs}

\author{M.~L.~Graham}

\setDocRef{DMTN-nnn}
\date{\today}
\setDocRevision{TBD}
\setDocStatus{draft}

\setDocAbstract{\textbf{WORKING DRAFT } but intended to be an overview of any and all potential requirements on data management (DM) from special programs (SP) such as the deep drilling fields (DDF) or mini-surveys (MS). We also summarize the plans for how the special programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future special program white papers and preparations for Level 3 pipelines. \textbf{Preliminary action items are listed in Section \ref{ssec:intro_action}.}

\medskip MLG Summary: The very high level requirements for Level 3 and Special Programs are written down in the DMSR, but they're broad. The next document down-flow, the LDM-151, is focused on the Level 1/2 algorithms with no accountability to the DMSR requirements related to SP and Level 3 (and that's fine). LDM-151, Section 7, mentions the DAX and SUIT, and then also mentions that it will be detailed in a different document. At this stage it seems it is maybe a mite too early to conclude anything, but we can go through the "Action Items" and try to resolve the questions that exist, for now.
}

\setDocChangeRecord{%
\addtohist{1}{2017-04-??}{Internal working document.}{Melissa Graham}
%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
}

\begin{document}

\maketitle

% CITATION EXAMPLES
% \verb|\citellp|: \citellp{LPM-17, LSE-30} \\
% \verb|\citell|: (SRD; \citell{LPM-17,LSE-29}) \\
% \verb|\citep[][]|: \citep[e.g.,][are interesting]{LPM-17,LSE-29} \\
% \verb|\cite|: \cite{LPM-17,LSE-29}

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{Introduction} \label{sec:intro}

\noindent The current status of this work is that it is an internal, ongoing study with several interim preliminary action items or outstanding questions so far (Section \ref{ssec:intro_action}).

\noindent The purpose of this study is to: \\
(1) summarize the current plans of the DM pipelines with respect to special programs, \\
(2) summarize the processing that is required to enable science for special programs, \\
(3) assess whether the DM pipelines meet the expected needs of special programs, and \\
(4) ensure any necessary changes get written into the requirements, designs, and plans.

\noindent This will be accomplished by: \\
(1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
(2) discussing initial issues internally with DM (Mario, K-T, Tim), and then \\
(3) ensure the identified needs become specified requirements (i.e., action items Section \ref{ssec:intro_action}).

{\it Mario: "Examples include making sure there's a requirement that the system is able to use an arbitrary template for some (deep drilling) field and a plan to deliver that functionality; or that the requirements define how a user-supplied code will be able to run in the batch system to process the last-night's data with shift-and-stack code for KBOs."}

\subsection{Current Action Items}\label{ssec:intro_action}

Action items for Data Management can be done internally as needed -- some are just questions of things to investigate. Items that involve gather feedback from the Science Collaborations should probably wait and happen at the same time as the next calls for community input (see Section \ref{ssec:intro_forum}).

\noindent \textbf{Data Management: } \\
\textbf{(1)} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time than the WFD images, by user request. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(2)} Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(3)} Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it. (Section \ref{ssec:dmdocs_review}) \\
\textbf{(4)} Find out what the expected filter change time will be (maximum is 120s). (Section \ref{ssec:dmdocs_review}) \\
\textbf{(5)} Confirm whether alternative products can contribute to the Alert Stream. (Section \ref{ssec:issues_alerts}) \\
\textbf{(6)} Update the database schema for {\tt DIASource} to add an element that contains information about which template was used to create the difference image (e.g., {\tt tractId} or {\tt patchId}). (Section \ref{ssec:issues_databases}) \\
\textbf{(7)} Clarify whether database schema key {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. (Section \ref{ssec:issues_databases}) \\
\textbf{(8)} Find out whether a SP visit with a single exposure (no snaps) disqualifies it from incorporation into Level 1. (Section \ref{ssec:issues_snaps}) \\
\textbf{(9)} Confirm that the ISR algorithms will cover a range of exposure times and dither pattersn. (Section \ref{ssec:issues_calibrations}) \\
\textbf{(10)} Implanting fake sources remains TBD. (Section \ref{ssec:issues_calibrations}) \\
\textbf{(11)} Confirm that science users could write a Level 3 SAS with the existing planned algorithms? (Section \ref{ssec:issues_mops})

\noindent \textbf{Science Collaborations: } \\
\textbf{(1)} SC should be solicited to provide input about all special program science goals that require 60 second alert latency from non-standard exposure times. (Section \ref{ssec:issues_alerts})\\
\textbf{(2)} SC should be solicited to provide input on whether the space allotted for object characterization parameters in the database schema such as variability and morphology are adequate to describe objects with observations over a wider range of cadences and depths than the WFD observations. E.g., can the variability parameter characterize fast cadence data, does the morphology parameter appropriately identify low-surface brightness features. DM will first compile and provide a list of planned parameters from the schema. (Section \ref{ssec:issues_databases}) \\
\textbf{(3)} SC should be solicited to provide input on whether there are science use-cases for specialized, auto-generated template images e.g., with constant $\Delta t$, or image-image differencing (instead of image-template). DM should then assess whether this will require additional database elements. (Section \ref{ssec:issues_templates}) \\

\subsection{Interacting with the Science Community} \label{ssec:intro_forum}

Several of the Action Items suggest soliciting input from the science collaborations (SC). However, it may be wiser/easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.

\noindent Recall that there is already a community forum on this topic, so far minimally used but ready to be revised in the future when the next calls soliciting white papers are released. \\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.

\noindent Remember also that the SC Coordinator for DDF: Neil Brandt.

\noindent The questions opened up on that forum are: \\
(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided? ``Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely). \\
(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Current DM Plans for Special Programs} \label{sec:dmdocs}

In this section we review what is written down in LSST documents with regards to the processing of Special Programs data. Below, we review the relevant contents of the three Data Management documents, and then the non-DM LSST documents.

\textbf{Overview:} Although the data from some special programs might be incorporated into the Level 1, 2, and/or MOPS pipelines (see Section \ref{ssec:dmdocs_SPinWFD} below), every special program will also have one or more associated Level 3 processing pipelines. These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas (so that e.g., the Level 3 databases that can be joined to the Level 1/2 databases), and also with custom codes. It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that an effort will be made to ensure the DM codes are extendable to Level 3. DM will also ensure that the relevant software infrastructure exists for science users, and for LSST to automatically run user-provided Level 3 pipelines (e.g., difference imaging on deep drilling nightly stacks). To enable science goals, some Level 3 pipelines can be run by LSST with a projected latency of (at most) tens of minutes (as opposed to the 60 second guaranteed for WFD). Level 3 pipelines for stacks and co-adds that do not have the same real-time urgency can simply be run by users through the Science Platform.

\subsection{Documentation Review}\label{ssec:dmdocs_review}

\noindent \textbf{DMSR, LSE-61} \\
Data Management Subsystems Requirements \citep{LSE-61} contains the high-level requirements for all processing and products. These have flowed down from the SRD (LPM-17), the SR (LSE-29), and the OSS (LSE-30), which are discussed in the next section. \\
$\bullet$ The only part of the DMSR that directly references Special Programs (or DDF or mini-surveys) is Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069), which specifies that "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing. They will be re-generated for users on-demand using the latest processing software and calibrations", but also says that "This aspect of the processing for Special Programs data is specific to each program". \\
$\rightarrow$ \textcolor{red}{It is unclear whether processed SP data will remain on disk and accessible when required by the science users, or if it will only live a short while}. \\
$\rightarrow$ \textbf{Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.} \\
$\bullet$ Section 2.9, "Level 3 Production", specifies high-level services of the DMS for Level 3 processing, which are relevant to special programs. \\
2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. $\rightarrow$ \textcolor{red}{(?).} \\
2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
2.9.5/6. Software Frameworks (ID: DMS-REQ-0125,DMS-REQ-0128): to provide a means for applying user-provided processing to catalog and image data, and for assessing the completeness of that application. \\
3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. $\rightarrow$ \textcolor{red}{I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?} \\

\noindent \textbf{DMAD, LDM-151}\\
LSST Data Management Applications Design \citep{LDM-151} outlines the code and algorithms that will be used in the processing of images in the AP (Level 1), DRP (Level 2), and MOPS. As such, it does not say much about Special Programs. It does say that "LSST DM is required to facilitate the creation of Level 3 data products by providing suitable APIs, software components, and computing infrastructure, but will not by itself create any Level 3 data products." The latter meaning that DM will not create any processing algorithms that are specific to the needs of any Special Programs data. Section 7 briefly describes the Data Access and the Science User Interface and Tool (SUIT), also known as the Science Platform, with a mention that they will be described in full in later documents. The Data Access layer (DAX) will be for sharing, accessing, searching (etc.) products, and the SUIT/Platform will provide an interactive environment for users, and also serve as portal for Level 3 pipeline development (including installed version of the LSST Software Stack), execution, retrieval, and analysis.

\noindent \textbf{DPDD, LSE-163}\\
Data Products Definitions Document \citep{LSE-163} specifies that the data products for Special Programs will ``be created using the same software and hardware as Levels 1 and 2". Section 7 is devoted to the data products of Special Programs, explaining how the current plan is to have both the Level 1 and 2 processing pipelines run on SP data in the same way that they run on the WFD, and produce all of the same data products and alerts, when possible/desirable. The Level 3 products for SP data can/will be kept in separate (joinable) databases that have identical schema as the WFD databases. \\
$\bullet$ It is specified that SP data processing will be limited to not use more than 10\% of the DM computational and storage facilities. During the telecon on Wed Mar 29, KT Lam reviewed the computational budget but it seemed mainly for WFD with only passing mention of DDF or special programs. \\
$\rightarrow$ \textcolor{red}{10\% of the computational budget seems like an underestimate. Although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack). On top of this, add the Level 3 processing that will be applied to WFD data, and we could easily find that the Special Programs data are processed $\sim4\times$ as much as WFD.} \\
$\rightarrow$ \textbf{Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing.}

\noindent \textbf{SRD, LPM-17}\\
The Science Requirements Document \citep{LPM-17} makes no reference to special programs, deep drilling fields, or mini-surveys; it's scope is the science of the WFD only.

\noindent \textbf{LSST SR, LSE-29}\\
The System Requirements document \citep{LSE-29} is derived from the LSST Science Requirements Document (SRD, \citep{LPM-17}). It contains the high-level specifications about fraction of total observing time delegated for special programs, minimum exposure times, maximum filter change time, etc., that flow down to the DMSR and OSS (detailed and discussed above and below).

\noindent \textbf{OSS, LSE-30}\\
The Observatory System Specifications (OSS) document \citep{LSE-30} contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from LPM-17 and LSE-29). There are two points worth mentioning: \\
$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
$\rightarrow$ \textcolor{red}{There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that thermal stability might also be affected with longer exposure times.}\\
$\rightarrow$ \textbf{Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it.} \\
$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
$\rightarrow$ \textcolor{red}{This maximum time for operational filter change would be a debilitating overhead for a deep drilling cadence of consecutive 30 second images that rotates through filters.}\\
$\rightarrow$ \textbf{Find out what the expected filter change time will be.} \\
$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
$\rightarrow$ \textcolor{red}{There is no related specification limiting the maximum number of total filter changes to avoid wear-and-tear on the mechanism?}

\noindent \textbf{LSE-180}\\
Level 2 Photometric Calibration for the LSST Survey \cite{LSE-180} is built on OpSim runs that do include some nominal deep drilling fields, but the photometric calibration investigated in this work does not deal with potential issues induced by non-standard visit patterns or exposure times of special programs -- it's scope is that of the WFD Main Survey.

% % % % % % % % % % % % % % % % % % 
\subsection{Incorporating SP Data into the WFD Main Survey}\label{ssec:dmdocs_SPinWFD}

I haven't found much written down in terms of guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines -- especially when these data deviate from the standard 30 second visit -- but it seems feasible. Here we make these considerations.

\noindent \textbf{Incorporating SP data into Level 1 and the Alert Stream.} \\
It is reasonable to attempt to include as many LSST images into Level 1 and the Alert Stream as possible. When the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, this is obviously fine. When the exposure time is shorter than a WFD visit image, this should be fine as well. The minimum supported exposure time is currently specified as 1 second, which spans a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (all other filters have a similarly large overlap). The exposure times of a special program image might be longer than the WFD visit image, but given that the template will typically be a stack of a year or two of data (at least), and that it's hard to imagine a reason for even doing $150$ second exposures (given that the saturation would be $r\approx 18.3$ and the cosmic rays would intensify), it seems unlikely that any images would be rejected from Level 1 for having too long an exposure time. This means that the limiting factor of including any/all special programs data into the Level 1 and Alert Stream will be a processing back-up caused by the higher cadence of short exposures, or a lack of template images for special programs that go outside the WFD search area.

\noindent \textbf{Incorporating SP data into MOPS.} \\
Needs and capabilities are currently undefined.

\noindent \textbf{Incorporating SP data into the Level 2 DRP.} \\
The core science needs for uniform-depth images and catalogs for the WFD main survey probably means this will not happen, but identical Level 3 data pipelines and products may be used for regions with deeper coverage and the two databases can be joined.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{Questions and Potential Issues}\label{sec:issues}

In this section we list questions or potential problems identified only from reading LSST documents \citep{LSE-163,LDM-151}, or issues raised also in Gregory's AHM 2016 slides\footnote{``DM Considerations for Deep Drilling", LSST AHM Aug 2016 presentation by Gregory Dubois-Felsmann \url{https://zenodo.org/record/61402\#.WNVk6hIrIUF}}. Another useful resource is the LSST Database Schema Browser: \url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}.

For each of these questions or potential issues, a response will be: \\
\textcolor{blue}{blue if the answer is known} (even if the answer is a negative), \\
\textcolor{red}{red if the answer is still to be determined}, and \\
\textbf{bold if an action item is required}.

% % % % % % % % % % % % % % % % % % 
\subsection{Alert Stream}\label{ssec:issues_alerts}

$\bullet$ Will the near-real-time (tens of minutes) difference-image processing of Level 3 pipelines be set up to contribute to the same Alert Stream as Level 1? (E.g., transient source detections in nightly stacks of a deep drilling field.) \\
$\rightarrow$  \textcolor{red}{As it is described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.} \\
$\rightarrow$ \textbf{Confirm whether alternative products can contribute to the Alert Stream.}

$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1 then this higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA. Is there a science need for 60 second latency on a consecutive time-series of short exposures? Or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred? \\
$\rightarrow$ \textbf{Query the community about all special program science goals that require 60 second alert latency from non-standard exposure times.}

$\bullet$ Alerts on a {\tt DIASource} will link to the associated {\tt DIAObject}, but in cases where a special program obtains a sequence of images without switching fields, will the {\tt DIAObject} catalog have had enough time to be updated to include the {\tt DIASource} from the previous image? \\
$\rightarrow$  \textcolor{blue}{Immediate updates of the {\tt DIAObject} catalog is an expected capability, although it is currently uncertain how it will be accomplished, technically.}

$\bullet$ LDM-151, Section 3.3.4 `Alert queuing and persistance' mentions that the {\it ``event message stream and the AlertDB will be synchronized at least once every 24 hours"} and that {\it ``Prior to the start of the subsequent night's observations, the message queue will be flushed and synchronized with the AlertDB. It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field. \\
$\rightarrow$  \textcolor{blue}{Upon review, this requirement seems unrelated to visit locations, and thus not an issue for special programs.} \\

% % % % % % % % % % % % % % % % % % 
\subsection{Databases and Schema}\label{ssec:issues_databases}

$\bullet$ The database schema for {\tt DIASource} needs to have an element added to identify the template image that was used; this is relevant for the Main Survey WFD Level 1 and also all Level 3 differencing products. \\
$\rightarrow$  \textbf{Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.}\\

$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to special programs, but it is unclear how such targets will be identified or flagged as such in the databases. \\
$\rightarrow$ \textcolor{red}{This is definitely a promised Level 1 product, but it is unclear whether we need to add an element to the database schema. Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType. Indicates which input was used to produce a given object."} Is that all we need?} \\
$\rightarrow$  \textbf{Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.}

$\bullet$ Will the following element sizes be enough the characterize the different kinds of variability that could be measured with cadences that are quite different from the main survey (i.e., longer or shorter cadenced special programs)? \\
{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
$\rightarrow$ \textcolor{red}{Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given special program, we need to ensure that they are adequate for the Level 1 data which might incorporate many timescales.}\\
$\rightarrow$  \textbf{Contact the Science Collaborations for input on the characterization parameters for variability.}

$\bullet$ Low surface-brightness features and/or objects are required for some galaxy science, and will show up more in e.g., a DDF than the WFD stack. Will the related shape parameters such as extendedness that will be included in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features afforded by e.g., DDF stacks? \\
$\rightarrow$ \textcolor{red}{As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given special program. It's unlikely that extremely deep patches will be included in the Level 2 products, but this needs assessment.}\\
$\rightarrow$  \textbf{Contact the Science Collaborations for input on the characterization parameters for extended objects.}

$\bullet$ Images obtained during special programs might be used with one or more different types of templates -- the same images could be used as multiple surveys. Is it feasible to have separate databases for each? \\ 
$\rightarrow$ \textcolor{blue}{Yes, Level 3 pipelines can create as many separate databases as necessary. There should at least be a separate database for each kind of template image. These databases can have the same base schema, and elements of the schema can be ``turned off" if the survey will not use them, to save processing time and disk space.} \\

$\bullet$ Data from special programs might be incorporated into the Level 1 and/or 2 databases; if so, how will it be flagged in these databases? \\
$\rightarrow$  \textcolor{blue}{In the schema for {\tt DIASource} and {\tt Source} there is an element for {\tt ccdVisitId}, and the database {\tt CcdVisit} has entry for {\tt visitId}, and the database {\tt Visit} has an entry for {\tt programId}, which is currently an {\tt integer} containing the {\it ``Observing program id (e.g., universal cadence, or one of the deep drilling programs, etc.)."}.} \\

$\bullet$ Will a Level 3 database for a Special Program be able to generate and incorporate forced photometry from WFD images? \\
$\rightarrow$ \textcolor{blue}{Yes. All Level 3 pipelines have access to all LSST data.}

$\bullet$ Will the Level 1 requirement to generate forced photometry for all new {\tt DIAObjects} within 24 hours be able to run on Level 3 image products of Special Programs data, such as deep nightly stacks of DDF images? \\
$\rightarrow$ \textcolor{blue}{This seems unlikely to be included as Level 1 (and unnecessary). Performing forced photometry for new {\tt DIAObjects} on a Level 3 image product would itself be a Level 3 pipeline.}

% % % % % % % % % % % % % % % % % % 
\subsection{Templates}\label{ssec:issues_templates}

$\bullet$ Typically we think of a template as a single deep coadd of image obtained e.g., $\Delta t \geq 1$ year ago. However, the use of a template with constant $\Delta t$ to the new images (e.g., $\Delta t = 7$ days to mitigate proper motion induced issues in a difference image for a Galactic plane mini-survey), or the use of a template that is simply a specific other image, might be scientifically necessary. Will this be supported by DM? \\
$\rightarrow$ \textcolor{blue}{Difference imaging with a custom template is Level 3, and the requirement for the generation of custom templates already exists as} \textbf{DMS-REQ-0032:} The DMS shall provide software to perform image differencing, generating Difference Exposures from the comparison of single exposures and/or coadded images \citep{LSE-61}.

% % % % % % % % % % % % % % % % % % 
\subsection{Snap Handling}\label{ssec:issues_snaps}

$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}, but is there a lower limit to exposure time for this? E.g., would not apply to a 1s twilight survey? \\
$\rightarrow$ \textcolor{red}{Have not yet tried to answer this question, but I wonder if for Special Programs that have single-image visits instead of two snaps, whether the lack of snaps will disqualify them from Level 1 inclusion?}\\
$\rightarrow$ \textbf{Does a SP visit with a single exposure (no snaps) disqualify it from incorporation into Level 1?}

% % % % % % % % % % % % % % % % % % 
\subsection{Calibrations}\label{ssec:issues_calibrations}

$\bullet$ The Special Programs data may have a different exposure time or dither pattern than the WFD survey. The concern here is that if the instrument signature removal (ISR) calibration pipeline, and/or the photometric/astrometric calibration pipeline, have been designed to work for WFD data that problems might be encountered for Special Programs data (e.g., dark frames; brighter-fatter correction; artifacts with an exposure time dependence; charge build-up from no-dither with bright stars). \\
$\rightarrow$ \textcolor{red}{Since there already exists a requirement that DM deliver processed visit images (ID: DMS-REQ-0069) to users (and to internal pipelines), it seems that at least the ISR will be characterized for a variety of exposure times -- but will it for sure?}. \\
$\rightarrow$ Note that LSE-180 mentions that it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
$\rightarrow$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD. The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD the corrections can be applied to the much smaller amount of data from the special programs (unless it is a function of exposure time and the special program uses a different exposure time). \\
$\rightarrow$ \textbf{Confirm whether it is implicit in the LDM-151 that all the ISR algorithms will automatically investigate and derive corrections for a range of exposure times, dither patterns, etc.}

$\bullet$ Planting fake sources may be both more manageable and more important to some special programs. How will DM support this? (Answer needs to distinguish between planting fakes in the single images for detection efficiencies vs. planting fakes in the coadds for point-source limiting magnitudes). \\
$\rightarrow$ The word 'fake' appears just once across LDM-151, LSE-61, and LSE-180 combined, and it is in Section 5.6.3 'MakeSelectionMaps' of LDM-151, which comments on how the depth and detection efficiency maps will be created, suggesting that fake implanting is one way. \\
$\rightarrow$ \textcolor{red}{Everything related to fake sources remains very unclear, even at Level 1 and 2.} \\
$\rightarrow$ \textbf{Figure out fake implantation?}

$\bullet$ Multiple special programs request short-exposure survey, which would provide less/different stars for photometric/astrometric calibration. Would a 1s exposure contain enough stars for calibration? \\
$\rightarrow$ \textcolor{blue}{A 1 second exposure will contain stars from $12.9<r<21.0$, which should be more than enough for a full photometric calibration.} \\
$\rightarrow$ Note that LSE-180 only considers calibration for the standard WFD data.

% % % % % % % % % % % % % % % % % % 
\subsection{Custom Co-Adds}\label{ssec:issues_coadds}

$\bullet$ Building deep co-adds with no transient contamination (e.g., SN hosts, cosmology uses host parameters correlated with intrinsic SN brightness), or creating deep stacks across filters for faint-object detection at $<5\sigma$ in a single filter. \\
$\rightarrow$ \textcolor{blue}{Creating custom co-adds will be a Level 3 capability that will be possible with the Level 2 DRP pipeline algorithms described in LDM-151. This includes multi-filter stacks (DMSR Section 1.3.4. 'Multi-band Coadds' ID:DMS-REQ-0281).}

% % % % % % % % % % % % % % % % % % 
\subsection{Moving Objects}\label{ssec:issues_mops}

$\bullet$ For faint moving objects, users will develop shift-and-stack (SAS) pipelines in Level 3 (MOPS does not SAS). Will they have all the necessary infrastructure for this intensive process? Will they be able to use the LSST real/bogus on their shift-and-stacked frames? \\
$\rightarrow$ \textcolor{red}{In particular, people may want to SAS on difference images, which would require DM making these available for some time.} \\
$\rightarrow$ \textbf{Confirm that science users could write a Level 3 SAS with the existing planned algorithms?}

$\bullet$ Can MOPS link newly discovered {\tt sources} to those discovered earlier on the same night (e.g., in a 4-visits-per-night scenario), or does it have a similar potential problem to {\tt DIASource} and {\tt DIAObject}? \\
$\rightarrow$  \textcolor{blue}{MOPS does not run during the night in real time. This is not a problem.}.

\clearpage
% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{Science Cases and their DM Requirements} \label{sec:science}

In this section we review science documents from LSST and the community. We compile information about the nominal observing plans for special programs that are already written down (Section \ref{ssec:science_plans}), and summarize the science motivations for special programs that have so far been proposed (Section \ref{ssec:science_descriptions}). Potential issues for DM are inferred from these science motivations and have been incorporated into our list of issues in Section \ref{sec:issues}.

\noindent Resources: \\
$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" \cite{2008arXiv0805.2366I} \\
$\bullet$ The LSST Science Requirements Document \cite{LPM-17} \\
$\bullet$ ``General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}\\
$\bullet$ ``LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{OSWP} \\
$\bullet$ List of LSST Deep Drilling white papers: \url{https://project.lsst.org/content/whitepapers32012} \\

\subsection{Nominal Observing Plans for DDF and MS}\label{ssec:science_plans}

\cite{2008arXiv0805.2366I}, Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months. Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$. The LSST Science Requirements Document \cite{LPM-17} doesn't contain the terms ``deep drilling field" or ``mini-survey"; I assume that the SRD is only for WFD survey.

Four extragalactic deep drilling fields have already been specified (Table \ref{tab:ddfms}). From a scouring of mainly the presentations of Brandt and Ridgway at the 2016 AHM, an {\it incomplete} list of potential mini-surveys that people are thinking about in Table \ref{tab:ddfms} also.

\begin{table}[h]
\begin{center}
\begin{footnotesize}
\caption{Approved DDF and Incomplete List of Potential MS.}
\label{tab:ddfms}
\begin{tabular}{lll}
\hline \hline
Name & Coordinates & Description  \\
\hline
DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
DDF TBD  & & TBD \\
North Ecliptic Spur      & & solar system objects (find and characterize) \\
Galactic Plane             & & more intensive stellar surveying \\
South Equatorial Cap  & & S/LMC and more Galactic science \\
Twilight                        & & short exposures (0.1s) for bright stars \\
Mini-Moons                     &  & finding mini-moons \\
Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
Meter-Sized Impactors     & & detection a week before impact \\
GW Optical Counterparts & & search and recovery \\
Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
\hline
\end{tabular}
\end{footnotesize}
\end{center}
\end{table}

\subsection{Science Descriptions (Not Comprehensive)}\label{ssec:science_descriptions}

The following is a collection of rough notes on the science goals and potential data processing/product issues for some proposed special programs, classified by field. It is not comprehensive. The questions for DM that were inspired by considering the proposed special programs are still mingled in amongst these notes, but have also been incorporated into Section \ref{sec:issues}, and so the answers and solutions to any questions or problems are all in Section \ref{sec:issues} also.

\medskip
\noindent \textbf{Solar System Objects}\\
$\bullet$ Strategy would be to observe 3x3 grid of 9 fields centered on coordinate of an ecliptic latitude equal to 0 and a longitude at a conjunction of Trojan clouds of Jupiter and Neptune, 4 groups of 2 visits on a 2--3 day revisit schedule, with the 4 groups targeted at e.g., opposition, all done in $r$-band ($g$ suitable too). (Is this within the North Ecliptic Spur?) \citep{BeckerWP}. \\
$\bullet$ North Ecliptic Spur region will yield more $\geq140$ m NEOs (Brandt talk).  \\
$\bullet$ Are the three SSO special programs described in Chapter 10.2 of \cite{OSWP} mandated by NASA? \\
$\bullet$ SSO science goals require enabling the discovery of faint SSO via shift-and-stack of images in specific fields where SSO discovery is maximized \citep{BeckerWP}. \\
$\bullet$ Does MOPS processing do shift-and-stack, and/or is the MOPS database schema ready to incorporate a shift-and-stack output? \\
$\bullet$ Regarding shift-and-stack, \cite{BeckerWP} says that ``the multi-fit algorithm ... naturally provides a base infrastructure for this process. In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required. However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." True? \\
$\bullet$ To detect Earth impactors {\it ``hours before impact"} using the four visits per night will require real-time linking of new {\tt DIASource}s -- or maybe the MOPS handles this differently? \\

\noindent \textbf{Stars in the Milky Way and Magellanics} \\
$\bullet$ Special programs will provide images deeper than the main survey in order to identify useful stellar populations; getting the images over a short time span mitigates proper motion losses and increase flare detection rates \citep{DhitalWP}. \\
$\bullet$ Target Galactic plane regions and/or open clusters, typically 1 to 3 filters needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}. \\
$\bullet$ Characterize stellar variability over the full L/SMC galaxies to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. \\
$\bullet$ Special co-adds may be required, e.g., ``To reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables" \citep{SzkodyWP}. \\
$\bullet$ Light curve characterization parameters in the databases should include one for very short events like flares; ensure there's adequate space in the database for multiple variability parameterizations. \\
$\bullet$ How might LC-characterization parameters fail on e.g., an LBV-turned-SN? \\
$\bullet$ A Twilight Short Exposure survey would need good on-the-fly sky subtraction of a rapidly changing background; ensure plans are adeqaute? \\
$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{OSWP}, suggests using {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}; feasible? \\

\noindent \textbf{Exoplanets} \\ 
$\bullet$ Transits. The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits. A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence. Microlensing events can also be detected with this data set. In both cases, follow-up is required. The Galactic Plane MS is proposed for this. \cite{2013arXiv1304.3455G} describes how transits can be extract from the same data set.\\
$\bullet$ Microlensing. Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates (for follow-up with e.g., LCO). However, more Galactic regions must be included, and this will require image differencing in crowded fields. \cite{2013arXiv1304.3455G} claims that imaging in the necessary regions of the galaxy has been disfavored by the project on the basis that the eventual deep co-adds would be uselessly confusion limited. I don't think DM could fix this fact, and deep co-adds (and all DRP) could possibly just be skipped for these regions. Otherwise, the already-planned AP should work fine for Galactic plane survey.\\

\noindent \textbf{Supernovae:} \\
$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for nightly stacks with a limit of $r\sim26.5$, extending the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses. \\
$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
$\bullet$ \cite{KesslerWP} describes their optimal exposure time distribution as 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures. \\
$\bullet$ Template generation: e.g., supernovae will want a DDF template from $>30$ days ago (minimum for SNeIa); $>100$ days is better (SNII), and $>200$ days safest (SLSNe, SNIIn). So, last DRP will be fine. But some last longer, thousands of days, and so users will want to know the date of the template. \\
$\bullet$ Building special, deep-as-possible, SN-free host galaxy images in order to measure correlated properties and make corrections to the LC, and do host science. The regular DRP will be inadequate for this (this could be a Level 3 tool to stack requested epochs only?). With short runs on a DDF this might be impossible, would have to revisit.  \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but can we calibrate as well with the limited set of stars? \\

\noindent \textbf{Galaxies:} \\
$\bullet$ \cite{FergusonWP} mentions {\it ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations"} and {\it ``characterization of low-surface-brightness extended features around both nearby and distant galaxies"} -- building a large collection of low-$\mu$ objects is a science driver but do these even get in the {\tt Object} database (\textbf{check on the `extendedness'-related parameters, e.g. {\sc Galfit} modeling? to what $\mu$?}), or is this assumed entirely Level 3? \\
$\bullet$ It seems that e.g., characterization of high-$z$ clusters will depend on ability to deblend extended objects. \\
$\bullet$ AGN monitoring in well-characterized galaxies \citep{FergusonWP}, and for a long-term DDF on many timescales \citep{GawiserWP} \\
$\bullet$ Effect of AGN on e.g., bulge-disk decomposition parameters? \\

\noindent \textbf{Weak lensing:} \\
$\bullet$ DDF can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
$\bullet$ ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey" (community forum, Jim Bosch) \\
$\bullet$ Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well. (community forum, Jim Bosch)  \\
$\bullet$ photo-$z$ are very important \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized \\

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{Conclusions}\label{sec:conc}

Conclusions section.

%\acknowledgments
%Acknowledgments.
%
%\appendix
%Appendix.

\bibliography{ms,lsst,refs,books,refs_ads}

\end{document}

%\begin{center}
%\includegraphics[width=8cm]{figures/}
%\includegraphics[width=8cm]{figures/}
%\caption{ \label{fig:}}
%\end{center}
%\end{figure}

### 2017-05-02 ms.tex
small update

@@
 \author{M.~L.~Graham}
+\author{M.~Juri\'{c}}
 
@@
 \medskip MLG Summary: The very high level requirements for Level 3 and Special Programs are written down in the DMSR, but they're broad.
-The next document down-flow, the LDM-151, is focused on the Level 1/2 algorithms with no accountability to the DMSR requirements related to SP and Level 3 (and that's fine).
+The next document down-flow, the LDM-151, is focused on the Level 1 and 2 algorithms with no accountability to the DMSR requirements related to SP and Level 3 (and that's fine).
 LDM-151, Section 7, mentions the DAX and SUIT, and then also mentions that it will be detailed in a different document.
@@
 (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(2)} Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing. (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(3)} Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it. (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(4)} Find out what the expected filter change time will be (maximum is 120s). (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(5)} Confirm whether alternative products can contribute to the Alert Stream. (Section \ref{ssec:issues_alerts}) \\
+\textbf{(2)} Confirm whether the computational/storage budget requires more than 10\% of the total resources for Special Programs and Level 3 processing. (Section \ref{ssec:dmdocs_review}) \\
+\textbf{(3)} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
+(Section \ref{ssec:dmdocs_review}) \\
+\textbf{(4)} Find out whether we can expect a filter change time shorter than 120 seconds (the specified maximum), which would enable a DDF to sequence filters. (Section \ref{ssec:dmdocs_review}) \\
+\textbf{(5)} Confirm whether alternative data products can contribute to the Alert Stream, or if only the Level 1 processing system will create alerts. (Section \ref{ssec:issues_alerts}) \\
 \textbf{(6)} Update the database schema for {\tt DIASource} to add an element that contains information about which template was used to create the difference image (e.g., {\tt tractId} or {\tt patchId}).
@@
 \textbf{(8)} Find out whether a SP visit with a single exposure (no snaps) disqualifies it from incorporation into Level 1. (Section \ref{ssec:issues_snaps}) \\
-\textbf{(9)} Confirm that the ISR algorithms will cover a range of exposure times and dither pattersn. (Section \ref{ssec:issues_calibrations}) \\
-\textbf{(10)} Implanting fake sources remains TBD. (Section \ref{ssec:issues_calibrations}) \\
-\textbf{(11)} Confirm that science users could write a Level 3 SAS with the existing planned algorithms? (Section \ref{ssec:issues_mops})
+\textbf{(9)} Confirm that the ISR algorithms will cover a range of exposure times and dither patterns. (Section \ref{ssec:issues_calibrations}) \\
+\textbf{(10)} Implanting fake sources remains TBD; will DM provide a code base to do this in Level 3? (Section \ref{ssec:issues_calibrations}) \\
 
@@
 \textbf{(1)} SC should be solicited to provide input about all special program science goals that require 60 second alert latency from non-standard exposure times. (Section \ref{ssec:issues_alerts})\\
-\textbf{(2)} SC should be solicited to provide input on whether the space allotted for object characterization parameters in the database schema such as variability and morphology are adequate to describe objects with observations over a wider range of cadences and depths than the WFD observations.
-E.g., can the variability parameter characterize fast cadence data, does the morphology parameter appropriately identify low-surface brightness features.
+\textbf{(2)} SC should be solicited to provide input on whether the space allotted for object characterization parameters in the database schema such as variability (e.g., {\tt lcPeriodic = float[6 x 32]}) or morphology (e.g., {\tt extendedness = float[1]}) are adequate to describe objects with Special Programs observations, which may include a wider range of cadences and depths than the WFD observations.
+Two examples include: (1) Does the variability parameter {\tt lcPeriodic} have enough space to adequately characterize high and low cadence observations with a variable dynamic range from a Twilight Survey?
+(2) Do the morphology parameters have enough elements to adequately represent both bright spiral arms and ultra-low-surface brightness features like tidal tails?
 DM will first compile and provide a list of planned parameters from the schema.
@@
 \textbf{Overview:} Although the data from some special programs might be incorporated into the Level 1, 2, and/or MOPS pipelines (see Section \ref{ssec:dmdocs_SPinWFD} below), every special program will also have one or more associated Level 3 processing pipelines.
-These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas (so that e.g., the Level 3 databases that can be joined to the Level 1/2 databases), and also with custom codes.
+These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas (so that e.g., the Level 3 databases that can be joined to the Level 1 and 2 databases), and also with custom codes.
 It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that an effort will be made to ensure the DM codes are extendable to Level 3.
@@
 On top of this, add the Level 3 processing that will be applied to WFD data, and we could easily find that the Special Programs data are processed $\sim4\times$ as much as WFD.} \\
-$\rightarrow$ \textbf{Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing.}
+$\rightarrow$ \textbf{Confirm whether the computational/storage budget requires more than 10\% of the total resources for Special Programs and Level 3 processing.}
 
@@
 $\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
-$\rightarrow$ \textcolor{red}{There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that thermal stability might also be affected with longer exposure times.}\\
-$\rightarrow$ \textbf{Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it.} \\
+$\rightarrow$ \textcolor{red}{There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that thermal stability might also be affected with longer exposure times -- but there doesn't seem to be an official limit on 'thermal stability'.}\\
+$\rightarrow$ \textbf{Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.} \\
 $\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
-$\rightarrow$ \textcolor{red}{This maximum time for operational filter change would be a debilitating overhead for a deep drilling cadence of consecutive 30 second images that rotates through filters.}\\
-$\rightarrow$ \textbf{Find out what the expected filter change time will be.} \\
+$\rightarrow$ \textcolor{red}{The specified maximum filter change time is 120 seconds, but this could be a debilitating overhead for, e.g., a deep drilling cadence of consecutive 30 second images that rotates through filters.}\\
+$\rightarrow$ \textbf{Find out whether we can expect a filter change time shorter than 120 seconds (the specified maximum), which would enable a DDF to sequence filters.} \\
 $\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
@@
 This means that the limiting factor of including any/all special programs data into the Level 1 and Alert Stream will be a processing back-up caused by the higher cadence of short exposures, or a lack of template images for special programs that go outside the WFD search area.
+(Also, I checked with David Reiss and he is not aware of anyone considering problems arising from differencing shorter/longer exposure times.)
 
@@
 $\rightarrow$ \textcolor{red}{Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given special program, we need to ensure that they are adequate for the Level 1 data which might incorporate many timescales.}\\
-$\rightarrow$  \textbf{Contact the Science Collaborations for input on the characterization parameters for variability.}
+$\rightarrow$  \textbf{Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths afforded by a special programs observing strategy.}
 
@@
 $\rightarrow$ \textcolor{red}{Everything related to fake sources remains very unclear, even at Level 1 and 2.} \\
-$\rightarrow$ \textbf{Figure out fake implantation?}
+$\rightarrow$ \textbf{Figure out fake planting for detection efficiencies etc.?}
 
@@
 $\rightarrow$ \textcolor{red}{In particular, people may want to SAS on difference images, which would require DM making these available for some time.} \\
-$\rightarrow$ \textbf{Confirm that science users could write a Level 3 SAS with the existing planned algorithms?}
+$\rightarrow$ \textbf{Confirm that science users could write a Level 3 SAS with the existing planned algorithms? E.g., see Section \ref{ssec:science_dmsumex}.}
 
@@
 
+\subsection{An Example of a ``Special Programs Processing Summary"}\label{ssec:science_dmsumex}
+
+For further insight to the DM-related needs of potential Special Programs, we can write out all of the acquisition and processing steps, in order.
+This kind of section describing the reductions and analysis would be typical in some kinds of observing proposal, and would be a reasonable expectation for a whitepaper.
+
+\noindent By Mario
+
+It would be extremely useful to describe a few concrete examples of how special programs processing would unfold, to make sure everyone is on the same page (first DM internally, then the science collaboration).
+Here's a *very rough* example I tried to put together while thinking about this.
+
+\noindent \textbf{DDF searches for TNOs}
+
+(1) The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec).
+[and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].
+{\it MLG: For example, \citep{BeckerWP} proposes four ``re-visit" sequences per field, each composed of two sets of $2$ $\times$ $15$ second exposures obtained 2-3 nights apart.
+The four sequences are obtained a (week?)/month/year apart.}
+
+(2) The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database.
+This will happen automatically for all images (perhaps within some range of exposure times?).
+{\it MLG: Yes, images from 1 to 120 seconds all seem OK to include in Level 1; doing shorter/longer exposures for any reason seems doubtful.
+However, this has not yet been tested (let alone confirmed) in LSST simulations of difference imaging (that I know of).}
+
+(3) The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes.
+This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it?
+or is is the same one that's used to process calibrations?
+have these systems been sized?].
+
+(3a) The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team.
+That is, we don't expect someone external to the ops team to babysit the code on a nightly basis.
+In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
+{\it MLG: I'm unsure of even which document would contain details this specific at this point -- not sure this can be written down anywhere yet?}
+
+(4) There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution.
+We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
+{\it MLG: I think this is covered by the DMSR Section 3.3.5 ``Provide Pipeline Execution Services" (ID: DMS-REQ-0156) and associated sub-requirements to provide production orchestration, monitoring, and fault tolerance software.
+The wording does not reference Level 1, 2, or Calibrations pipelines and so must apply generically to Level 3 as well.}
+
+(5) Once the processing finishes, the results of will be stored to a program-specific database.
+No alerts (in VOEvent sense) will be issued.
+We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store.
+[This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
+{\it MLG: This kind of notification does not appear to be covered by any item in the DMSR, but it seems more appropriate to be described in the Science Platform deliverables?}
+
+(6) The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables).
+The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files).
+The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).
+
+(7) When the images are made available to the batch system (step 3), they also become available to *everyone*.
+I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
+[this isn't in the requirements right now -- right now we say that images will become available in 24hrs.
+But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have 3, you get 6 for free as well.]
+
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 

### 2017-05-10 ms.tex
major revision

--- 2017-04-28 ms.tex
@@
 % black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
+\usepackage{enumitem}
 
@@
 
-\author{M.~L.~Graham}
+\author{M.~L.~Graham and M.~Juri\'{c}}
 
@@
 
-\setDocAbstract{\textbf{WORKING DRAFT } but intended to be an overview of any and all potential requirements on data management (DM) from special programs (SP) such as the deep drilling fields (DDF) or mini-surveys (MS).
-We also summarize the plans for how the special programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2).
-It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future special program white papers and preparations for Level 3 pipelines.
-\textbf{Preliminary action items are listed in Section \ref{ssec:intro_action}.}
-
-\medskip MLG Summary: The very high level requirements for Level 3 and Special Programs are written down in the DMSR, but they're broad.
-The next document down-flow, the LDM-151, is focused on the Level 1/2 algorithms with no accountability to the DMSR requirements related to SP and Level 3 (and that's fine).
-LDM-151, Section 7, mentions the DAX and SUIT, and then also mentions that it will be detailed in a different document.
-At this stage it seems it is maybe a mite too early to conclude anything, but we can go through the "Action Items" and try to resolve the questions that exist, for now.
+\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys.
+We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2).
+It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.
 }
@@
 
-\noindent The current status of this work is that it is an internal, ongoing study with several interim preliminary action items or outstanding questions so far (Section \ref{ssec:intro_action}).
-
 \noindent The purpose of this study is to: \\
-(1) summarize the current plans of the DM pipelines with respect to special programs, \\
-(2) summarize the processing that is required to enable science for special programs, \\
-(3) assess whether the DM pipelines meet the expected needs of special programs, and \\
+(1) summarize the current plans of the DM pipelines with respect to Special Programs, \\
+(2) summarize the processing that is required to enable science for Special Programs, \\
+(3) assess whether the DM pipelines meet the expected needs of Special Programs, and \\
 (4) ensure any necessary changes get written into the requirements, designs, and plans.
@@
 (1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
-(2) discussing initial issues internally with DM (Mario, K-T, Tim), and then \\
-(3) ensure the identified needs become specified requirements (i.e., action items Section \ref{ssec:intro_action}).
+(2) discussing initial issues internally with DM (Mario, the SST), engineering (K-T, Tim), and then \\
+(3) ensuring the identified needs become specified requirements (i.e., spawn change requests).
 
@@
 
-\subsection{Current Action Items}\label{ssec:intro_action}
+% % % % % % % % % % % % % % % % % % 
+\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}
 
-Action items for Data Management can be done internally as needed -- some are just questions of things to investigate.
-Items that involve gather feedback from the Science Collaborations should probably wait and happen at the same time as the next calls for community input (see Section \ref{ssec:intro_forum}).
+In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community.
+The LSST project will need to define boundaries for these observations.
+Although not all of these boundaries will originate with DM constraints, we start that process here (\textcolor{red}{MLG: Are any initially derived boundaries for Special Programs observations specified elsewhere?}).
+The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}
 
-\noindent \textbf{Data Management: } \\
-\textbf{(1)} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time than the WFD images, by user request.
-(Section \ref{ssec:dmdocs_review}) \\
-\textbf{(2)} Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing. (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(3)} Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it. (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(4)} Find out what the expected filter change time will be (maximum is 120s). (Section \ref{ssec:dmdocs_review}) \\
-\textbf{(5)} Confirm whether alternative products can contribute to the Alert Stream. (Section \ref{ssec:issues_alerts}) \\
-\textbf{(6)} Update the database schema for {\tt DIASource} to add an element that contains information about which template was used to create the difference image (e.g., {\tt tractId} or {\tt patchId}).
-(Section \ref{ssec:issues_databases}) \\
-\textbf{(7)} Clarify whether database schema key {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
-(Section \ref{ssec:issues_databases}) \\
-\textbf{(8)} Find out whether a SP visit with a single exposure (no snaps) disqualifies it from incorporation into Level 1. (Section \ref{ssec:issues_snaps}) \\
-\textbf{(9)} Confirm that the ISR algorithms will cover a range of exposure times and dither pattersn. (Section \ref{ssec:issues_calibrations}) \\
-\textbf{(10)} Implanting fake sources remains TBD. (Section \ref{ssec:issues_calibrations}) \\
-\textbf{(11)} Confirm that science users could write a Level 3 SAS with the existing planned algorithms? (Section \ref{ssec:issues_mops})
+Start of a list of observing restrictions. \\
+\textbf{Sky coordinates --} (The RA and Dec that will be accessible from the site.) \\
+\textbf{Pointing alt/az, airmass --} (Telescope pointing constraints.) \\
+\textbf{Sky background level --} (e.g., twilight observations cannot start before X) \\
+\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
+\textbf{Filters --} (e.g., list filter combinations; $gri$ always be on?; change times, etc.) \\
+\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
+\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD Main Survey). \\
 
-\noindent \textbf{Science Collaborations: } \\
-\textbf{(1)} SC should be solicited to provide input about all special program science goals that require 60 second alert latency from non-standard exposure times. (Section \ref{ssec:issues_alerts})\\
-\textbf{(2)} SC should be solicited to provide input on whether the space allotted for object characterization parameters in the database schema such as variability and morphology are adequate to describe objects with observations over a wider range of cadences and depths than the WFD observations.
-E.g., can the variability parameter characterize fast cadence data, does the morphology parameter appropriately identify low-surface brightness features.
-DM will first compile and provide a list of planned parameters from the schema.
-(Section \ref{ssec:issues_databases}) \\
-\textbf{(3)} SC should be solicited to provide input on whether there are science use-cases for specialized, auto-generated template images e.g., with constant $\Delta t$, or image-image differencing (instead of image-template).
-DM should then assess whether this will require additional database elements.
-(Section \ref{ssec:issues_templates}) \\
-
+% % % % % % % % % % % % % % % % % % 
 \subsection{Interacting with the Science Community} \label{ssec:intro_forum}
 
-Several of the Action Items suggest soliciting input from the science collaborations (SC).
-However, it may be wiser/easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers.
+To resolve some open issues we will need to solicit input from the science collaborations (SC).
+However, it may be better and easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers.
 The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
@@
 
+% % % % % % % % % % % % % % % % % % 
+\subsection{A Summary of Open Action Items} \label{ssec:intro_AI}
+
+\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
+\end{enumerate}
+\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
+\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time.
+\end{enumerate}
+
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
 \clearpage
-\section{The Current DM Plans for Special Programs} \label{sec:dmdocs}
+\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmdocs}
 
-In this section we review what is written down in LSST documents with regards to the processing of Special Programs data.
-Below, we review the relevant contents of the three Data Management documents, and then the non-DM LSST documents.
+In this section we summarize how DM intends to process the data obtained during Special Programs observations.
+This high-level "intent" is a combination of specifications that are actually written down in DM-related documents, and unwritten plans that exist in e.g., Mario's Brain.
+{\it MLG: Note that Sections \ref{ssec:dmdocs_SPinWFD} and \ref{ssec:dmdocs_L3} currently contain a mixture of what is officially written down and what is unofficially planned or expected, usually my interpretation of a conversation with Mario, while Section \ref{ssec:dmdocs_review} is a direct review of the DM documentation.
+Once the SST and engineering have converged on a treatment plan for Special Programs data via this work, my plan is to return to the official DM documentation and identify where change requests need to be initiated.}
 
-\textbf{Overview:} Although the data from some special programs might be incorporated into the Level 1, 2, and/or MOPS pipelines (see Section \ref{ssec:dmdocs_SPinWFD} below), every special program will also have one or more associated Level 3 processing pipelines.
-These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas (so that e.g., the Level 3 databases that can be joined to the Level 1/2 databases), and also with custom codes.
-It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that an effort will be made to ensure the DM codes are extendable to Level 3.
-DM will also ensure that the relevant software infrastructure exists for science users, and for LSST to automatically run user-provided Level 3 pipelines (e.g., difference imaging on deep drilling nightly stacks).
-To enable science goals, some Level 3 pipelines can be run by LSST with a projected latency of (at most) tens of minutes (as opposed to the 60 second guaranteed for WFD).
-Level 3 pipelines for stacks and co-adds that do not have the same real-time urgency can simply be run by users through the Science Platform.
+\textbf{Overview:} Although the data from some Special Programs might be incorporated into the Level 1, 2, and/or MOPS pipelines, every Special Program will probably also need one or more associated Level 3 processing pipelines.
+These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas so that e.g., the Level 3 databases that can be easily joined to the Level 1 and 2 databases.
+It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that DM will ensure all pipeline code is extendable to Level 3 and that the relevant software infrastructure exists for science users to create and run their Level 3 pipelines.
+In addition, LSST will "federate" some Level 3 pipelines, putting them under change control and taking responsibility for their processing (e.g., shift-and-stack for faint moving objects is expected to become a federated Level 3 processing).
+All user access to processing codes and data products will be through the Science Platform.
 
-\subsection{Documentation Review}\label{ssec:dmdocs_review}
+\textbf{Section contents}: \\
+Section \ref{ssec:dmdocs_SPinWFD} -- an overview of whether and how DM will incorporate Special Programs data into the Level 1, 2, and MOPS pipelines and products. \\
+Section \ref{ssec:dmdocs_L3} -- an overview of how federated and user-run Level 3 pipelines will process Special Programs data, incorporate WFD data, and/or join to Level 1 and 2 products. \\
+Section \ref{ssec:dmdocs_review} -- a detailed break-down of Special Programs (or Level 3) related content in the three DM documents (DPDD, DMSR, and DMAD), and non-DM LSST documents. \\
 
-\noindent \textbf{DMSR, LSE-61} \\
-Data Management Subsystems Requirements \citep{LSE-61} contains the high-level requirements for all processing and products.
-These have flowed down from the SRD (LPM-17), the SR (LSE-29), and the OSS (LSE-30), which are discussed in the next section.
-\\
-$\bullet$ The only part of the DMSR that directly references Special Programs (or DDF or mini-surveys) is Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069), which specifies that "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing.
-They will be re-generated for users on-demand using the latest processing software and calibrations", but also says that "This aspect of the processing for Special Programs data is specific to each program".
-\\
-$\rightarrow$ \textcolor{red}{It is unclear whether processed SP data will remain on disk and accessible when required by the science users, or if it will only live a short while}. \\
-$\rightarrow$ \textbf{Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.} \\
-$\bullet$ Section 2.9, "Level 3 Production", specifies high-level services of the DMS for Level 3 processing, which are relevant to special programs. \\
+% % % % % % % % % % % % % % % % % % 
+\subsection{Incorporating SP Data into the WFD Main Survey}\label{ssec:dmdocs_SPinWFD}
+
+There is not much, if anything, specified regarding guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines that will automatically run on all WFD main survey observations.
+The concern is that drastic changes in the exposure time or total depth of the observations might negatively impact the science goals that depend on the WFD main survey data products.
+Here we summarize these concerns.
+
+\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmdocs_SPinWFD_L1}
+
+It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
+Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream).
+Only images that qualify for Level 1 processing will contribute to the Alert Stream.
+The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template.
+So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine.
+However, the suitability of shorter and longer exposures is unknown.
+Here we consider the suitability -- or what we need to do in order to determine the suitability -- of shorter and longer exposures for the Level 1 and Alert Stream.
+These considerations are made independently of the science motivations of the Special Programs that proposed the different exposure times.
+
+% Shorter exposures: twilight survey, getting bright stars on the same photometric system, variability in brighter stars
+% Longer exposures: the weak lensing community might propose long-exposure fields if they remain concerned that 2x15 seconds is insufficient to obtain a stable PSF with no spatial variation dominated by the atmosphere
+
+(1) Shorter Exposures.
+The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint.
+The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified.
+We consider this issue separately in Section \ref{ssec:issues_calibrations}.
+Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
+A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
+However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
+Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
+
+(2) Longer Exposures.
+There is no maximum exposure time specified for an LSST image.
+Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
+However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
+With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+
+In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
+One possibly useful study is Chang et al.
+(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
+They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
+They present a new software package to do mitigate the effects.
+Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
+2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
+
+\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
+\end{enumerate}
+
+\subsubsection{MOPS}\label{ssec:dmdocs_SPinWFD_MOPS}
+
+The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
+While it is expected that MOPS will be able to identify appropriate images and incorporate as much Special Programs data as possible, the capabilities are currently undefined, and we leave it for the MOPS pipeline builders to decide.
+We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.
+
+\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmdocs_SPinWFD_L2}
+
+Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
+We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
+When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
+For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
+In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{SP Data Processing with Federated and User-Run Level 3}\label{ssec:dmdocs_L3}
+
+Through the Science Platform, member of the science community can assemble Level 3 processing packages using both LSST and custom codes.
+The DM team is required to write their codes to be extendable to Level 3, and to provide the infrastructure for users to run them on LSST servers, monitor their progress, and analyze their output (including a prioritization system to allocate processing resources).
+All Level 3 pipelines will have access to processed single visits from both the WFD main survey and Special Programs, as well as difference images, CoAdds, and catalogs from the Level 1 and 2 pipelines.
+It is expected that some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team, such as those developed for time domain science that need to run automatically and repeatedly.
+For both federated and user-run code, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
+
+Level 3 pipelines for Special Programs data for time-domain astronomy will need access to images on a timescale from minutes to weeks after acquisition, as opposed to static-sky Level 3 pipelines that may run on e.g., a yearly schedule like the Level 2 DRP.
+It is currently unclear what that real-time latency will be, and how processing resources will be allocated; here we break down some of the options.
+
+If the images obtained with a Special Program are incorporated into Level 1 (see Section \ref{ssec:dmdocs_SPinWFD}), then the Level 3 processing can begin as soon as the Level 1 imaging products are available: raw images, processed single visits (a.k.a.
+calexps), difference images, and their source catalogs.
+The current maximum time for this is {\tt \textbf{L1PublicT}}, specified by LSE-29 (\citep{LSE-29}) to be 24 hours (LSR-REQ-0104) -- but this seems far too long, especially for federated Level 3 pipelines, given that they are finished within $60$ seconds.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
+\end{enumerate}
+
+If the images obtained with a Special Program cannot be incorporated into Level 1, then the latency with which they will be made available is unspecified.
+Ultimately, their delivery will depend why they are not incorporated into Level 1, which will probably fall into two categories: (1) there is no template at their location, or (2) their exposure time prohibits difference imaging.
+In the former case, it seems likely that single processed images can be provided on a timescale of minutes, but this needs to be specified.
+In the latter case, it will depend on the exposure time.
+As mentioned in Section\label{ssec:dmdocs_SPinWFD_L1}, the minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified, and discussed separately in Section \ref{ssec:issues_calibrations}.
+It may be that single-visit processing is simply not supported by DM below some exposure time threshold, and so the latency on raw image delivery should be specified.
+Although it is likely that slightly shorter/longer exposure images can be reduced and calibrated with the same pipeline as $15$ second exposures, even if they are unsuitable for difference imaging, this has not yet been confirmed, and needs to be investigated and specified.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
+\end{enumerate}
+
+In the DPDD \citep{LSE-163}, it specifies that Special Programs data processing will be limited to not more than 10\% of the DM computational and storage facilities.
+During the telecon on Wed Mar 29, KT Lam reviewed the computational budget, but it seemed mainly for WFD with only passing mention of DDF or Special Programs.
+This 10\% of the computational budget seems like an underestimate: although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack).
+It seems that we could easily find that Special Programs data are processed $\sim4\times$ as much as WFD.
+Furthermore, it is unclear whether the location for all this processing has been identified; Mario asks, for federated real-time Level 3 processing, {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that?s used to process calibrations.
+Have these systems been sized?"}.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
+\end{enumerate}
+
+\subsection{In-Depth Documentation Review}\label{ssec:dmdocs_review}
+
+In this section we attempt to track the flow-down requirements through the official LSST and DM documentation.
+We find some hardware-related constraints in the OSS that will set the maximum possible divergence of Special Programs raw images from WFD main survey images (i.e., exposure times, rate of filter changes), and these will be relevant to DM processing.
+We also find that some broad, high-level requirements for Level 3 and Special Programs are written down in the DMSR.
+The next document down-flow, the DMAD, is focused on the Level 1 and 2 algorithms, and although its Section 7 mentions the DAX and SUIT, it says that they are to be detailed in a different document.
+The DPDD also does not have much content that directly pertains to Special Programs and Level 3 processing.
+
+\noindent \textbf{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\\
+The SRD makes no direct reference to Special Programs, deep drilling fields, or mini-surveys. It's scope is the science of the WFD only.
+
+\noindent \textbf{LSST System Requirements, SR, LSE-29, \cite{LSE-29}}\\
+The SR is derived from the SRD.
+It contains the high-level specifications about fraction of total observing time delegated for Special Programs, minimum exposure times, maximum filter change time, etc., that flow-down to and are repeated in the DMSR and OSS (discussed below).
+
+\noindent \textbf{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\\
+The OSS contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from the SRD and SR).
+There are two points worth mentioning: \\
+$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
+$\rightarrow$ There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times.
+However, there doesn't seem to be a specified limit on 'thermal stability'.
+Knowing the full range of exposure times and image cadences that DM must consider is important for estimating the predicted latencies providing raw, reduced and calibrated images (Section \ref{ssec:dmdocs_L3}) and testing the incorporation of non-standard images into the calibration and difference imaging pipelines (Section \ref{ssec:dmdocs_SPinWFD_L1}).
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
+\end{enumerate}
+$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
+$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
+$\rightarrow$ The specified maximum filter change time could be a debilitating overhead for, e.g., a deep drilling cadence of consecutive 30 second images wants to rotate through filters.
+Furthermore, we might expect that in reality the filter change time will be shorter than 120 seconds, which would enable a DDF to sequence filters.
+Furthermore, while the minimum filter change count is not an issue, I can't find the nightly (or total lifetime) maximum to limit wear and tear on the mechanism (as I've heard about).
+However, there is already a change request in progress regarding this issue: \url{https://project.lsst.org/groups/ccb/node/938}.
+
+\noindent \textbf{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}} \\
+The DMSR contains the high-level requirements for all processing and products. There are two main areas that pertain to Special Programs. \\
+$\bullet$ Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069)\\
+$\rightarrow$ Specifies that {\it "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing.
+They will be re-generated for users on-demand using the latest processing software and calibrations"}, but also says that {\it "This aspect of the processing for Special Programs data is specific to each program"}.
+It is unclear whether processed SP data will remain on disk and accessible for as long as it is required by the science users, or if it will only live a short while.
+\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
+\end{enumerate}
+$\bullet$ Section 2.9, "Level 3 Production" \\
+Specifies services of the DMS for Level 3 processing which are relevant to Special Programs: \\
 2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
 2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
-2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. $\rightarrow$ \textcolor{red}{(?).} \\
+2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. \\
 2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
@@
 \\
-3.1.
-Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops.
-$\rightarrow$ \textcolor{red}{I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?} \\
+3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. \\
+$\rightarrow$ \textcolor{red}{Questions}: Regarding 2.9.3, I'm not sure what {\it "self-consistent inputs"} means and so cannot assess any potential issues.
+Regarding 3.1, I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?
 
-\noindent \textbf{DMAD, LDM-151}\\
-LSST Data Management Applications Design \citep{LDM-151} outlines the code and algorithms that will be used in the processing of images in the AP (Level 1), DRP (Level 2), and MOPS.
+\noindent \textbf{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\\
+LDM-151 outlines the code and algorithms that will be used in the processing of images in the AP (Level 1), DRP (Level 2), and MOPS.
 As such, it does not say much about Special Programs.
-It does say that "LSST DM is required to facilitate the creation of Level 3 data products by providing suitable APIs, software components, and computing infrastructure, but will not by itself create any Level 3 data products." The latter meaning that DM will not create any processing algorithms that are specific to the needs of any Special Programs data.
-Section 7 briefly describes the Data Access and the Science User Interface and Tool (SUIT), also known as the Science Platform, with a mention that they will be described in full in later documents.
+It does say that {\it "LSST DM is required to facilitate the creation of Level 3 data products by providing suitable APIs, software components, and computing infrastructure, but will not by itself create any Level 3 data products"}.
+The latter means that DM will not create any processing algorithms that are specific to the needs of any Special Programs data.
+Section 7 briefly describes the Data Access and the Science User Interface and Tool (SUIT), also known as the Science Platform, with a mention that they will be described in full in later documents  (probably this one: \url{https://github.com/lsst-dmsst/lsst_science_platform/blob/master/science-platform.pdf}).
 The Data Access layer (DAX) will be for sharing, accessing, searching (etc.) products, and the SUIT/Platform will provide an interactive environment for users, and also serve as portal for Level 3 pipeline development (including installed version of the LSST Software Stack), execution, retrieval, and analysis.
 
-\noindent \textbf{DPDD, LSE-163}\\
-Data Products Definitions Document \citep{LSE-163} specifies that the data products for Special Programs will ``be created using the same software and hardware as Levels 1 and 2".
-Section 7 is devoted to the data products of Special Programs, explaining how the current plan is to have both the Level 1 and 2 processing pipelines run on SP data in the same way that they run on the WFD, and produce all of the same data products and alerts, when possible/desirable.
-The Level 3 products for SP data can/will be kept in separate (joinable) databases that have identical schema as the WFD databases.
-\\
-$\bullet$ It is specified that SP data processing will be limited to not use more than 10\% of the DM computational and storage facilities.
-During the telecon on Wed Mar 29, KT Lam reviewed the computational budget but it seemed mainly for WFD with only passing mention of DDF or special programs.
-\\
-$\rightarrow$ \textcolor{red}{10\% of the computational budget seems like an underestimate.
-Although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack).
-On top of this, add the Level 3 processing that will be applied to WFD data, and we could easily find that the Special Programs data are processed $\sim4\times$ as much as WFD.} \\
-$\rightarrow$ \textbf{Confirm if the budget requires additional computational resources for Special Programs and Level 3 processing.}
+\noindent \textbf{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\\
+The DPDD specifies that the data products for Special Programs will {\it "be created using the same software and hardware as Levels 1 and 2"}.
+Like LDM-151, Section 7 is devoted to the data products of Special Programs, explaining how the current plan is to have both the Level 1 and 2 processing pipelines run on SP data in the same way that they run on the WFD, and produce all of the same data products and alerts, when possible.
+It also states that the Level 3 products for SP data can (should) be kept in separate, but joinable, databases that have identical schema as the WFD databases.
 
-\noindent \textbf{SRD, LPM-17}\\
-The Science Requirements Document \citep{LPM-17} makes no reference to special programs, deep drilling fields, or mini-surveys; it's scope is the science of the WFD only.
-
-\noindent \textbf{LSST SR, LSE-29}\\
-The System Requirements document \citep{LSE-29} is derived from the LSST Science Requirements Document (SRD, \citep{LPM-17}).
-It contains the high-level specifications about fraction of total observing time delegated for special programs, minimum exposure times, maximum filter change time, etc., that flow down to the DMSR and OSS (detailed and discussed above and below).
-
-\noindent \textbf{OSS, LSE-30}\\
-The Observatory System Specifications (OSS) document \citep{LSE-30} contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from LPM-17 and LSE-29).
-There are two points worth mentioning: \\
-$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
-$\rightarrow$ \textcolor{red}{There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that thermal stability might also be affected with longer exposure times.}\\
-$\rightarrow$ \textbf{Find out the thermal tolerances and determine the minimum and maximum exposure times (and/or cadence of short exposures) that will satisfy it.} \\
-$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
-$\rightarrow$ \textcolor{red}{This maximum time for operational filter change would be a debilitating overhead for a deep drilling cadence of consecutive 30 second images that rotates through filters.}\\
-$\rightarrow$ \textbf{Find out what the expected filter change time will be.} \\
-$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
-$\rightarrow$ \textcolor{red}{There is no related specification limiting the maximum number of total filter changes to avoid wear-and-tear on the mechanism?}
-
-\noindent \textbf{LSE-180}\\
-Level 2 Photometric Calibration for the LSST Survey \cite{LSE-180} is built on OpSim runs that do include some nominal deep drilling fields, but the photometric calibration investigated in this work does not deal with potential issues induced by non-standard visit patterns or exposure times of special programs -- it's scope is that of the WFD Main Survey.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Incorporating SP Data into the WFD Main Survey}\label{ssec:dmdocs_SPinWFD}
-
-I haven't found much written down in terms of guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines -- especially when these data deviate from the standard 30 second visit -- but it seems feasible.
-Here we make these considerations.
-
-\noindent \textbf{Incorporating SP data into Level 1 and the Alert Stream.} \\
-It is reasonable to attempt to include as many LSST images into Level 1 and the Alert Stream as possible.
-When the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, this is obviously fine.
-When the exposure time is shorter than a WFD visit image, this should be fine as well.
-The minimum supported exposure time is currently specified as 1 second, which spans a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (all other filters have a similarly large overlap).
-The exposure times of a special program image might be longer than the WFD visit image, but given that the template will typically be a stack of a year or two of data (at least), and that it's hard to imagine a reason for even doing $150$ second exposures (given that the saturation would be $r\approx 18.3$ and the cosmic rays would intensify), it seems unlikely that any images would be rejected from Level 1 for having too long an exposure time.
-This means that the limiting factor of including any/all special programs data into the Level 1 and Alert Stream will be a processing back-up caused by the higher cadence of short exposures, or a lack of template images for special programs that go outside the WFD search area.
-
-\noindent \textbf{Incorporating SP data into MOPS.} \\
-Needs and capabilities are currently undefined.
-
-\noindent \textbf{Incorporating SP data into the Level 2 DRP.} \\
-The core science needs for uniform-depth images and catalogs for the WFD main survey probably means this will not happen, but identical Level 3 data pipelines and products may be used for regions with deeper coverage and the two databases can be joined.
+\noindent \textbf{Level 2 Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}\\
+LSE-180 is built on OpSim runs that do include some nominal deep drilling fields, but the photometric calibration investigated in this work does not deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs -- it's scope the WFD Main Survey.
 
@@
 
-In this section we list questions or potential problems identified only from reading LSST documents \citep{LSE-163,LDM-151}, or issues raised also in Gregory's AHM 2016 slides\footnote{``DM Considerations for Deep Drilling", LSST AHM Aug 2016 presentation by Gregory Dubois-Felsmann \url{https://zenodo.org/record/61402\#.WNVk6hIrIUF}}.
+In this section we list questions or potential problems identified from reading LSST documents \citep{LSE-163,LDM-151}, Gregory's AHM 2016 slides\footnote{``DM Considerations for Deep Drilling", LSST AHM Aug 2016 presentation by Gregory Dubois-Felsmann \url{https://zenodo.org/record/61402\#.WNVk6hIrIUF}}, or using our imaginations.
 Another useful resource is the LSST Database Schema Browser: \url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}.
@@
 \textcolor{blue}{blue if the answer is known} (even if the answer is a negative), \\
-\textcolor{red}{red if the answer is still to be determined}, and \\
-\textbf{bold if an action item is required}.
+\textcolor{red}{red if the answer is still to be determined}. \\
+If it seems particularly urgent or involved to answer, it will spawn an Action Item.
 
@@
 
-$\bullet$ Will the near-real-time (tens of minutes) difference-image processing of Level 3 pipelines be set up to contribute to the same Alert Stream as Level 1?
-(E.g., transient source detections in nightly stacks of a deep drilling field.) \\
-$\rightarrow$  \textcolor{red}{As it is described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.} \\
-$\rightarrow$ \textbf{Confirm whether alternative products can contribute to the Alert Stream.}
-
-$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1 then this higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA.
-Is there a science need for 60 second latency on a consecutive time-series of short exposures?
-Or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred?
+$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA.
+Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred?
+(I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales).
+Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations?
 \\
-$\rightarrow$ \textbf{Query the community about all special program science goals that require 60 second alert latency from non-standard exposure times.}
-
-$\bullet$ Alerts on a {\tt DIASource} will link to the associated {\tt DIAObject}, but in cases where a special program obtains a sequence of images without switching fields, will the {\tt DIAObject} catalog have had enough time to be updated to include the {\tt DIASource} from the previous image?
-\\
-$\rightarrow$  \textcolor{blue}{Immediate updates of the {\tt DIAObject} catalog is an expected capability, although it is currently uncertain how it will be accomplished, technically.}
+$\rightarrow$ \textcolor{red}{Unanswered.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging.
+\end{enumerate}
 
 $\bullet$ LDM-151, Section 3.3.4 `Alert queuing and persistance' mentions that the {\it ``event message stream and the AlertDB will be synchronized at least once every 24 hours"} and that {\it ``Prior to the start of the subsequent night's observations, the message queue will be flushed and synchronized with the AlertDB.
-It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field.
+It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field?
 \\
-$\rightarrow$  \textcolor{blue}{Upon review, this requirement seems unrelated to visit locations, and thus not an issue for special programs.} \\
+$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs?
+
+$\bullet$ Will near-real-time Level 3 difference-image processing be able to contribute to the same Alert Stream as Level 1?
+(E.g., transient source detections in nightly stacks of a deep drilling field.) \\
+$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.
+
+$\bullet$ Alerts on a {\tt DIASource} will link to the associated {\tt DIAObject}, but in cases where a Special Program obtains a sequence of images without switching fields, will the {\tt DIAObject} catalog have had enough time to be updated to include the {\tt DIASource} from the previous image?
+\\
+$\rightarrow$  \textcolor{blue}{Answered.} Immediate updates of the {\tt DIAObject} catalog is an expected capability, although it is currently uncertain how it will be accomplished, technically.
 
@@
 
-$\bullet$ The database schema for {\tt DIASource} needs to have an element added to identify the template image that was used; this is relevant for the Main Survey WFD Level 1 and also all Level 3 differencing products.
+$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products.
 \\
-$\rightarrow$  \textbf{Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.}\\
+$\rightarrow$ \textcolor{red}{Unaddressed.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\end{enumerate}
 
-$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to special programs, but it is unclear how such targets will be identified or flagged as such in the databases.
+$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to Special Programs, but it is unclear how such targets will be identified or flagged as such in the databases, and unclear whether we need to add an element to the database schema for this.
+Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType.
+Indicates which input was used to produce a given object."} Is that all we need?
 \\
-$\rightarrow$ \textcolor{red}{This is definitely a promised Level 1 product, but it is unclear whether we need to add an element to the database schema.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?} \\
-$\rightarrow$  \textbf{Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.}
+$\rightarrow$ \textcolor{red}{Unanswered.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
+\end{enumerate}
 
-$\bullet$ Will the following element sizes be enough the characterize the different kinds of variability that could be measured with cadences that are quite different from the main survey (i.e., longer or shorter cadenced special programs)?
+$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)?
+Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales.
 \\
@@
 {\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-$\rightarrow$ \textcolor{red}{Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given special program, we need to ensure that they are adequate for the Level 1 data which might incorporate many timescales.}\\
-$\rightarrow$  \textbf{Contact the Science Collaborations for input on the characterization parameters for variability.}
+$\rightarrow$ \textcolor{red}{Unanswered.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+\end{enumerate}
 
-$\bullet$ Low surface-brightness features and/or objects are required for some galaxy science, and will show up more in e.g., a DDF than the WFD stack.
-Will the related shape parameters such as extendedness that will be included in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features afforded by e.g., DDF stacks?
-\\
-$\rightarrow$ \textcolor{red}{As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given special program.
-It's unlikely that extremely deep patches will be included in the Level 2 products, but this needs assessment.}\\
-$\rightarrow$  \textbf{Contact the Science Collaborations for input on the characterization parameters for extended objects.}
+$\bullet$ Low surface-brightness features and/or faint extended objects are required for some galaxy science, and will show up more in e.g., a DDF than a WFD CoAdd.
+Will the related shape parameters such as extendedness in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features that might come from deeper Special Programs CoAdds?
+As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program.
+Furthermore, it's unlikely that extremely deep patches will be included in the Level 2 products, but this still should be assessed.\\
+$\rightarrow$ \textcolor{red}{Unassessed.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
 
-$\bullet$ Images obtained during special programs might be used with one or more different types of templates -- the same images could be used as multiple surveys.
+$\bullet$ Images obtained during Special Programs might be used with one or more different types of templates -- the same images could be used as multiple surveys.
 Is it feasible to have separate databases for each?
 \\
-$\rightarrow$ \textcolor{blue}{Yes, Level 3 pipelines can create as many separate databases as necessary.
+$\rightarrow$ \textcolor{blue}{Answered.} Yes, Level 3 pipelines can create as many separate databases as necessary.
 There should at least be a separate database for each kind of template image.
-These databases can have the same base schema, and elements of the schema can be ``turned off" if the survey will not use them, to save processing time and disk space.} \\
+These databases can have the same base schema, and elements of the schema can be ``turned off" if the survey will not use them, to save processing time and disk space.
 
-$\bullet$ Data from special programs might be incorporated into the Level 1 and/or 2 databases; if so, how will it be flagged in these databases? \\
-$\rightarrow$  \textcolor{blue}{In the schema for {\tt DIASource} and {\tt Source} there is an element for {\tt ccdVisitId}, and the database {\tt CcdVisit} has entry for {\tt visitId}, and the database {\tt Visit} has an entry for {\tt programId}, which is currently an {\tt integer} containing the {\it ``Observing program id (e.g., universal cadence, or one of the deep drilling programs, etc.)."}.} \\
+$\bullet$ Data from Special Programs might be incorporated into the Level 1 and/or 2 databases; if so, how will it be flagged in these databases? \\
+$\rightarrow$  \textcolor{blue}{Answered.} In the schema for {\tt DIASource} and {\tt Source} there is an element for {\tt ccdVisitId}, and the database {\tt CcdVisit} has entry for {\tt visitId}, and the database {\tt Visit} has an entry for {\tt programId}, which is currently an {\tt integer} containing the {\it ``Observing program id (e.g., universal cadence, or one of the deep drilling programs, etc.)."}.
 
 $\bullet$ Will a Level 3 database for a Special Program be able to generate and incorporate forced photometry from WFD images? \\
-$\rightarrow$ \textcolor{blue}{Yes. All Level 3 pipelines have access to all LSST data.}
+$\rightarrow$ \textcolor{blue}{Answered.} Yes. All Level 3 pipelines have access to all LSST data.
 
@@
 \\
-$\rightarrow$ \textcolor{blue}{This seems unlikely to be included as Level 1 (and unnecessary).
-Performing forced photometry for new {\tt DIAObjects} on a Level 3 image product would itself be a Level 3 pipeline.}
+$\rightarrow$ \textcolor{blue}{Answered.} No.
+This forced photometry will run only on other images that are incorporated into Level 1, and this will not include the data products of Level 3 pipelines.
+Performing forced photometry for new {\tt DIAObjects} on a Level 3 image product would itself be a Level 3 pipeline.
 
@@
 \\
-$\rightarrow$ \textcolor{blue}{Difference imaging with a custom template is Level 3, and the requirement for the generation of custom templates already exists as} \textbf{DMS-REQ-0032:} The DMS shall provide software to perform image differencing, generating Difference Exposures from the comparison of single exposures and/or coadded images \citep{LSE-61}.
+$\rightarrow$ \textcolor{blue}{Answered.} Yes.
+Difference imaging with a custom template is Level 3, and the requirement for the generation of custom templates already exists as \textbf{DMS-REQ-0032:} The DMS shall provide software to perform image differencing, generating Difference Exposures from the comparison of single exposures and/or coadded images \citep{LSE-61}.
 
@@
 
-$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}, but is there a lower limit to exposure time for this?
-E.g., would not apply to a 1s twilight survey?
+$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}.
+Could Special Programs data that is only a single exposure, not a visit that is the combination of two snaps, be incorporated into the Level 1 AP?
 \\
-$\rightarrow$ \textcolor{red}{Have not yet tried to answer this question, but I wonder if for Special Programs that have single-image visits instead of two snaps, whether the lack of snaps will disqualify them from Level 1 inclusion?}\\
-$\rightarrow$ \textbf{Does a SP visit with a single exposure (no snaps) disqualify it from incorporation into Level 1?}
+$\rightarrow$ \textcolor{red}{Unanswered.}
 
@@
 
-$\bullet$ The Special Programs data may have a different exposure time or dither pattern than the WFD survey.
+$\bullet$ \textbf{Image Reduction} The Special Programs data may have a different exposure time or dither pattern than the WFD survey.
 The concern here is that if the instrument signature removal (ISR) calibration pipeline, and/or the photometric/astrometric calibration pipeline, have been designed to work for WFD data that problems might be encountered for Special Programs data (e.g., dark frames; brighter-fatter correction; artifacts with an exposure time dependence; charge build-up from no-dither with bright stars).
+Since there already exists a requirement that DM deliver processed visit images (ID: DMS-REQ-0069) to users (and to internal pipelines), it seems that at least the ISR must be developed for a variety of exposure times, but is this truly the case?
+Or are there exposure time boundaries beyond which DM will not guarantee delivery of processed images?
 \\
-$\rightarrow$ \textcolor{red}{Since there already exists a requirement that DM deliver processed visit images (ID: DMS-REQ-0069) to users (and to internal pipelines), it seems that at least the ISR will be characterized for a variety of exposure times -- but will it for sure?}.
+$\rightarrow$ \textcolor{red}{Unanswered.}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time.
+\end{enumerate}
+
+$\bullet$ Regarding the above point, LSE-180 offers some relevant information regarding the reduction and calibration of non-WFD images: \\
+$\rightarrow$ LSE-180 mentions that it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
+$\rightarrow$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
+The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD the corrections can be applied to the much smaller amount of data from the Special Programs (unless it is a function of exposure time and the Special Program uses a different exposure time).
+
+$\bullet$ \textbf{Image Detection Efficiency} Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
+Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
+What are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
+(The answer should distinguish between planting fakes in the single images for transient detection efficiencies with difference imaging, $vs.$ planting fakes in the CoAdds for point-source limiting magnitudes).
 \\
-$\rightarrow$ Note that LSE-180 mentions that it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be.
+$\rightarrow$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
+will record information on the pixel variance ...
+as well as the per-pixel masks ...
+These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
 \\
-$\rightarrow$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD the corrections can be applied to the much smaller amount of data from the special programs (unless it is a function of exposure time and the special program uses a different exposure time).
+$\rightarrow$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
+detection efficiency for point sources vs.
+mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
 \\
-$\rightarrow$ \textbf{Confirm whether it is implicit in the LDM-151 that all the ISR algorithms will automatically investigate and derive corrections for a range of exposure times, dither patterns, etc.}
+$\rightarrow$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
+The details of what metrics will be mapped, the format and scale of the maps (e.g.
+hierarchical pixelizations vs.
+polygons), and the way the metrics will be computed are all unknown"}.
+It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
+(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
+Source detection thresholds can be estimated through the use of sky sources"}.)\\
+$\rightarrow$ Suchyta et al.
+(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
+\\
+$\rightarrow$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
+\\
+$\rightarrow$ \textcolor{red}{Unassessed.} But I would like to take this on in the near future with an assessment of our options, their costs, and the scientific payoff.
 
-$\bullet$ Planting fake sources may be both more manageable and more important to some special programs.
-How will DM support this?
-(Answer needs to distinguish between planting fakes in the single images for detection efficiencies vs.
-planting fakes in the coadds for point-source limiting magnitudes).
-\\
-$\rightarrow$ The word 'fake' appears just once across LDM-151, LSE-61, and LSE-180 combined, and it is in Section 5.6.3 'MakeSelectionMaps' of LDM-151, which comments on how the depth and detection efficiency maps will be created, suggesting that fake implanting is one way.
-\\
-$\rightarrow$ \textcolor{red}{Everything related to fake sources remains very unclear, even at Level 1 and 2.} \\
-$\rightarrow$ \textbf{Figure out fake implantation?}
-
-$\bullet$ Multiple special programs request short-exposure survey, which would provide less/different stars for photometric/astrometric calibration.
+$\bullet$ Multiple Special Programs request short-exposure survey, which would provide less/different stars for photometric/astrometric calibration.
 Would a 1s exposure contain enough stars for calibration?
 \\
-$\rightarrow$ \textcolor{blue}{A 1 second exposure will contain stars from $12.9<r<21.0$, which should be more than enough for a full photometric calibration.} \\
-$\rightarrow$ Note that LSE-180 only considers calibration for the standard WFD data.
+$\rightarrow$ \textcolor{blue}{Answered.} 1 second exposure will contain stars from $12.9<r<21.0$, which should be more than enough for a full photometric calibration.
+However, LSE-180 only considers calibration for the standard WFD data, and so this has not been tested.
 
@@
 
-$\bullet$ Building deep co-adds with no transient contamination (e.g., SN hosts, cosmology uses host parameters correlated with intrinsic SN brightness), or creating deep stacks across filters for faint-object detection at $<5\sigma$ in a single filter.
+$\bullet$ Building deep co-adds with no transient contamination (e.g., SN hosts, cosmology uses host parameters correlated with intrinsic SN brightness), or creating deep stacks across filters for faint-object detection at $<5\sigma$ in a single filter, may be a desired capability for Special Programs data.
+Will this be possible?
 \\
-$\rightarrow$ \textcolor{blue}{Creating custom co-adds will be a Level 3 capability that will be possible with the Level 2 DRP pipeline algorithms described in LDM-151.
+$\rightarrow$ \textcolor{blue}{Answered.} Yes.
+Creating custom co-adds will be a Level 3 capability that will be possible with the Level 2 DRP pipeline algorithms described in LDM-151.
 This includes multi-filter stacks (DMSR Section 1.3.4.
-'Multi-band Coadds' ID:DMS-REQ-0281).}
+'Multi-band Coadds' ID:DMS-REQ-0281).
 
@@
 Will they be able to use the LSST real/bogus on their shift-and-stacked frames?
+Will they be able to access the difference images over long time scales?
 \\
-$\rightarrow$ \textcolor{red}{In particular, people may want to SAS on difference images, which would require DM making these available for some time.} \\
-$\rightarrow$ \textbf{Confirm that science users could write a Level 3 SAS with the existing planned algorithms?}
+$\rightarrow$ \textcolor{red}{Unanswered.} All of this seems likely, as each of these questions is covered by the DM requirement to {\it "make all codes extendable to Level 3"}, but ensuring that the answer is yes to each of them will be more involved in reality, and probably happen only with time as the SAS Level 3 pipeline is developed and integrated?
 
@@
 \\
-$\rightarrow$  \textcolor{blue}{MOPS does not run during the night in real time. This is not a problem.}.
+$\rightarrow$  \textcolor{blue}{Answered.} MOPS does not run during the night in real time. This is not a problem.
 
@@
 In this section we review science documents from LSST and the community.
-We compile information about the nominal observing plans for special programs that are already written down (Section \ref{ssec:science_plans}), and summarize the science motivations for special programs that have so far been proposed (Section \ref{ssec:science_descriptions}).
+We compile information about the nominal observing plans for Special Programs that are already written down (Section \ref{ssec:science_plans}), and summarize the science motivations for Special Programs that have so far been proposed (Section \ref{ssec:science_descriptions}).
 Potential issues for DM are inferred from these science motivations and have been incorporated into our list of issues in Section \ref{sec:issues}.
+In Section \ref{ssec:science_dmsums}, we explore some Special Programs processing case studies as an additional avenue to identify potential issues.
 
@@
 
+% % % % % % % % % % % % % % % % % % 
 \subsection{Nominal Observing Plans for DDF and MS}\label{ssec:science_plans}
 
-\cite{2008arXiv0805.2366I}, Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
+Ivezi\'{c} et al.
+(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
 Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
-The LSST Science Requirements Document \cite{LPM-17} doesn't contain the terms ``deep drilling field" or ``mini-survey"; I assume that the SRD is only for WFD survey.
-
+The LSST Science Requirements Document \cite{LPM-17} doesn't contain the terms ``deep drilling field" or ``mini-survey".
 Four extragalactic deep drilling fields have already been specified (Table \ref{tab:ddfms}).
@@
 
+% % % % % % % % % % % % % % % % % % 
 \subsection{Science Descriptions (Not Comprehensive)}\label{ssec:science_descriptions}
 
-The following is a collection of rough notes on the science goals and potential data processing/product issues for some proposed special programs, classified by field.
+The following is a collection of rough notes on the science goals for some proposed Special Programs, classified by area of astronomy.
 It is not comprehensive.
-The questions for DM that were inspired by considering the proposed special programs are still mingled in amongst these notes, but have also been incorporated into Section \ref{sec:issues}, and so the answers and solutions to any questions or problems are all in Section \ref{sec:issues} also.
+Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no Action Items are spawned in this section (although there are some low-level unanswered questions).
 
@@
 \noindent \textbf{Solar System Objects}\\
-$\bullet$ Strategy would be to observe 3x3 grid of 9 fields centered on coordinate of an ecliptic latitude equal to 0 and a longitude at a conjunction of Trojan clouds of Jupiter and Neptune, 4 groups of 2 visits on a 2--3 day revisit schedule, with the 4 groups targeted at e.g., opposition, all done in $r$-band ($g$ suitable too).
-(Is this within the North Ecliptic Spur?) \citep{BeckerWP}.
-\\
-$\bullet$ North Ecliptic Spur region will yield more $\geq140$ m NEOs (Brandt talk).  \\
-$\bullet$ Are the three SSO special programs described in Chapter 10.2 of \cite{OSWP} mandated by NASA? \\
-$\bullet$ SSO science goals require enabling the discovery of faint SSO via shift-and-stack of images in specific fields where SSO discovery is maximized \citep{BeckerWP}. \\
-$\bullet$ Does MOPS processing do shift-and-stack, and/or is the MOPS database schema ready to incorporate a shift-and-stack output? \\
-$\bullet$ Regarding shift-and-stack, \cite{BeckerWP} says that ``the multi-fit algorithm ...
+$\bullet$ Observations in the North Ecliptic Spur region will yield more $\geq140$ m NEOs (Brandt talk). \\
+$\bullet$ Detecting faint SSO down to $r\sim27$ by applying shift-and-stack processing to a mini-survey (Trans-Neptunian Objectss, Trojans, asteroids, long-period comets, dwarf planets).
+\citep{BeckerWP} \\
+$\rightarrow$ Regarding shift-and-stack, \cite{BeckerWP} says that ``the multi-fit algorithm ...
 naturally provides a base infrastructure for this process.
 In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required.
-However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." True?
+However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." Is this true?
+\textcolor{red}{Unanswered}.
 \\
-$\bullet$ To detect Earth impactors {\it ``hours before impact"} using the four visits per night will require real-time linking of new {\tt DIASource}s -- or maybe the MOPS handles this differently?
-\\
+$\bullet$ Mini-moons, temporarily captured satellites of the Earth. (Section 10.2, \cite{OSWP}) \\
+$\bullet$ Find meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{OSWP}) \\
+$\bullet$ Sweetspot: twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{OSWP})
 
 \noindent \textbf{Stars in the Milky Way and Magellanics} \\
-$\bullet$ Special programs will provide images deeper than the main survey in order to identify useful stellar populations; getting the images over a short time span mitigates proper motion losses and increase flare detection rates \citep{DhitalWP}.
+$\bullet$ Deep imaging acquired over a short time span to mitigate proper motion loss (and increase flare detection rates) will identify useful stellar populations. \citep{DhitalWP}. \\
+$\rightarrow$ When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
 \\
-$\bullet$ Target Galactic plane regions and/or open clusters, typically 1 to 3 filters needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
+$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. \\
+$\rightarrow$ Special co-adds may be required, e.g., {\it "To reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
+\citep{SzkodyWP}.
 \\
-$\bullet$ Characterize stellar variability over the full L/SMC galaxies to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. \\
-$\bullet$ Special co-adds may be required, e.g., ``To reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables" \citep{SzkodyWP}. \\
-$\bullet$ Light curve characterization parameters in the databases should include one for very short events like flares; ensure there's adequate space in the database for multiple variability parameterizations.
-\\
-$\bullet$ How might LC-characterization parameters fail on e.g., an LBV-turned-SN? \\
-$\bullet$ A Twilight Short Exposure survey would need good on-the-fly sky subtraction of a rapidly changing background; ensure plans are adeqaute? \\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{OSWP}, suggests using {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}; feasible?
-\\
+$\bullet$ A Twilight Short Exposure survey would enable bright stars (with many historical observations) to be put solidly on the same photometric system as the deeper LSST WFD main survey catalog. \\
+$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{OSWP}, suggests using {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}, but is this feasible?
+\textcolor{red}{Unanswered.}
 
@@
 \cite{2013arXiv1304.3455G} claims that imaging in the necessary regions of the galaxy has been disfavored by the project on the basis that the eventual deep co-adds would be uselessly confusion limited.
-I don't think DM could fix this fact, and deep co-adds (and all DRP) could possibly just be skipped for these regions.
-Otherwise, the already-planned AP should work fine for Galactic plane survey.\\
+(But was this before the Galactic Plane region was defined?)
 
-\noindent \textbf{Supernovae:} \\
-$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for nightly stacks with a limit of $r\sim26.5$, extending the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses.
+\noindent \textbf{Supernovae} \\
+$\bullet$ Extend the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses with the nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$.
 \\
-$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
-$\bullet$ \cite{KesslerWP} describes their optimal exposure time distribution as 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures.
+%$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
+$\rightarrow$ The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures.
+\cite{KesslerWP} \\
+$\bullet$ Generating the best-possible individual SN light curves by building special, deep-as-possible, SN-free host galaxy images and using them as a template.
+This will be necessary for SNe that appear in the template image (e.g., that last $>1000$ days).
+SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
 \\
-$\bullet$ Template generation: e.g., supernovae will want a DDF template from $>30$ days ago (minimum for SNeIa); $>100$ days is better (SNII), and $>200$ days safest (SLSNe, SNIIn).
-So, last DRP will be fine.
-But some last longer, thousands of days, and so users will want to know the date of the template.
+$\rightarrow$ The Level 2 DRP codes to create CoAdds should be adequate for this. \\
+$\rightarrow$ \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
+$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.
+
+\noindent \textbf{Galaxies} \\
+$\bullet$ To build  a large collection of low-$\mu$ objects: \cite{FergusonWP} mentions {\it ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations"} and {\it ``characterization of low-surface-brightness extended features around both nearby and distant galaxies"}.
 \\
-$\bullet$ Building special, deep-as-possible, SN-free host galaxy images in order to measure correlated properties and make corrections to the LC, and do host science.
-The regular DRP will be inadequate for this (this could be a Level 3 tool to stack requested epochs only?).
-With short runs on a DDF this might be impossible, would have to revisit.
-\cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP.
+$\bullet$ Characterization of high-$z$ clusters (will will depend on ability to deblend extended objects). \\
+$\bullet$ AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP}  \citep{GawiserWP} \\
+$\rightarrow$ Study the effects of AGN on e.g., bulge-disk decomposition parameters?
+
+\noindent \textbf{Weak Lensing} \\
+$\bullet$ DDF can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
+$\bullet$ {\it "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (community forum, Jim Bosch) \\
+$\bullet$ {\it "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
+We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (community forum, Jim Bosch)  \\
+$\bullet$ Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Special Programs Processing Case Studies}\label{ssec:science_dmsums}
+
+For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
+This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
+Note that we are not including any analysis in these descriptions.
+Also, they aren't that fully realized yet, and could use some more thought and input, which is why none of the questions associated with these case studies have spawned any "Action Items".
+
+Generic steps currently include: \\
+Step 1. Data Acquisition. \\
+Step 2. Inclusion in Level 1 AP. \\
+Step 3. Delivery of LSST Processed Images. \\
+Step 4. Level 3 Processing. \\
+Step 5. Level 3 Products. \\
+(Step 6. Inclusion in Level 2.) \\
+
+\subsubsection{A Mini-Survey Searching for TNOs with Shift-and-Stack}
+
+This Special Programs processing summary is based on Becker et al.
+(2011) white paper to find TNOs with shift-and stack \citep{BeckerWP}.
+{\it MLG: See Section \ref{sssec:science_dmsums_generic} for a more general processing case-study written by Mario, for a shift-and-stack Level 3 pipeline running regularly on a large amount of data.}
+
+Step 1.
+Data Acquisition.
+The observational sequence is triggered.
+In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
+This sequence is always repeated 2-3 nights later.
+This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
+Data obtained in the $g$-band filter is also acceptable.
+\citep{BeckerWP}
+
+Step 2. Inclusion in Level 1 AP. Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+
+Step 3.
+Delivery of LSST Processed Images.
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+However, this Special Program appears to require a year of dispersed observations before the Level 3 processing would run.
 \\
-$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but can we calibrate as well with the limited set of stars? \\
+\textcolor{red}{Question:} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
 
-\noindent \textbf{Galaxies:} \\
-$\bullet$ \cite{FergusonWP} mentions {\it ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations"} and {\it ``characterization of low-surface-brightness extended features around both nearby and distant galaxies"} -- building a large collection of low-$\mu$ objects is a science driver but do these even get in the {\tt Object} database (\textbf{check on the `extendedness'-related parameters, e.g.
-{\sc Galfit} modeling?
-to what $\mu$?}), or is this assumed entirely Level 3?
+Step 4.
+Level 3 Processing.
+The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
+Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
+The Level 2 difference imaging routine will be used with the same template tract/patch for all.
+Custom code will shift the difference images, then Level 2 routines will stack and do source detection and characterization (including real/bogus?) and generate an object database.
+Custom code will derive orbital parameters for the detections and add them to the database.
+The user is notified with processing status updates.
 \\
-$\bullet$ It seems that e.g., characterization of high-$z$ clusters will depend on ability to deblend extended objects. \\
-$\bullet$ AGN monitoring in well-characterized galaxies \citep{FergusonWP}, and for a long-term DDF on many timescales \citep{GawiserWP} \\
-$\bullet$ Effect of AGN on e.g., bulge-disk decomposition parameters? \\
+\textcolor{red}{Question:} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
+\\
+\textcolor{red}{Question:} Will the internal real/bogus routine be able to run on shift-and-stack difference images? \\
+\textcolor{red}{Question:} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
+I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables?
 
-\noindent \textbf{Weak lensing:} \\
-$\bullet$ DDF can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey" (community forum, Jim Bosch) \\
-$\bullet$ Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well.
-(community forum, Jim Bosch)  \\
-$\bullet$ photo-$z$ are very important \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized \\
+Step 5.
+Level 3 Products.
+The shifted-and-stacked images.
+Databases of sources detected by the shift-and-stack process with the desired schema, some of which will overlap with Level 1, 2, and/or MOPS to facilitate table joins, some of which will be generated by custom codes.
+\\
+\textcolor{red}{Question:} Will it be possible for Level 3 databases to add arbitrary schema elements, or can they only subselect from pre-defined options? \\
+
+(Step 6.
+Inclusion in Level 2.
+Given the extreme difference in final depth of these fields, it is unlikely that they will be incorporated into the Level 2.
+However, external Level 3 processing could certainly make deep CoAdds for these fields, join them to the Level 2 DRP catalog, and use them for other science.)
+
+\subsubsection{Searching for Supernovae in Deep Drilling Fields}
+
+Step 1.
+Data Acquisition.
+On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
+
+Step 2. Inclusion in Level 1 AP. Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+
+Step 3.
+Delivery of LSST Processed Images.
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+
+Step 4.
+Level 3 Processing.
+The federated pipeline runs automatically upon delivery of the Level 1 processed images.
+If a deep template image does not already exist (this would be a Level 3 product), a Level 2 stacking routine will be used to make them.
+This Level 2 stacking routine is also used to make the nightly deep CoAdd in each filter, and then a Level 1 or 2 routine will generate the difference images and run source detection (and real/bogus?).
+These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases.
+\textcolor{red}{Question:} Will the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
+
+Step 5.
+Level 3 Products.
+A deep template for this field, the nightly stacked CoAdds and difference images, and analogs of the {\tt DIAObject} and {\tt DIASource} databases but built only for this specific Deep Drilling Field.
+\\
+\textcolor{red}{Question:} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
+
+(Step 6. Inclusion in Level 2. Unlikely.)
+
+\subsubsection{A Twilight Survey of Bright Stars}
+
+Future work.
+
+\subsubsection{The Galactic Plane Survey for Variable Stars}
+
+Future work.
+
+\subsubsection{GW Optical Counterpart Search}
+
+Future work.
+
+\subsubsection{A More Generic Level 3 Shift-and-Stack Case Study}\label{sssec:science_dmsums_generic}
+
+A Level 3 processing case study for shift-and-stack on a large number images. By Mario.
+
+\#1.
+The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec).
+[ and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].
+
+\#2.
+The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database.
+This will happen automatically for all images (perhaps within some range of exposure times?).
+
+\#3.
+The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes.
+This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it?
+or is is the same one that's used to process calibrations?
+have these systems been sized?].
+
+\#3 a).
+The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team.
+That is, we don't expect someone external to the ops team to babysit the code on a nightly basis.
+In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
+
+\#4.
+There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution.
+We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
+
+\#5.
+Once the processing finishes, the results of will be stored to a program-specific database.
+No alerts (in VOEvent sense) will be issued.
+We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store.
+[This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
+
+\#6.
+The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables).
+The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files).
+The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).
+
+\#7.
+When the images are made available to the batch system (step \#3), they also become available to *everyone*.
+I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
+[this isn't in the requirements right now -- right now we say that images will become available in 24hrs.
+But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have \#3, you get \#6 for free as well.]
 
@@
 %
-%\appendix
-%Appendix.
 

### 2017-06-09 ms.tex
organizational edits

@@
 
-{\it Mario: "Examples include making sure there's a requirement that the system is able to use an arbitrary template for some (deep drilling) field and a plan to deliver that functionality; or that the requirements define how a user-supplied code will be able to run in the batch system to process the last-night's data with shift-and-stack code for KBOs."}
+{\it Current Status: This document is a rough draft, with the goal of being exhaustively complete. Many of the identified \textbf{``Concerns"} may just be misunderstandings or a lack of knowledge.}
 
+The rest of this introduction contains: \\
+\S~\ref{ssec:intro_bounds} -- defining DM-related boundaries for Special Programs. \\
+\S~\ref{ssec:intro_forum} -- existing channels for interaction with Science communities. \\
+\S~\ref{ssec:intro_UC} -- a summary of unresolved \textbf{Concerns} that arose in this study.
+
+Later sections attempt to identify potential \textbf{Concerns} regarding DM's plans for processing the data from Special Programs with a variety of approaches: \\
+\S~\ref{sec:dmdocs} -- a review of the current DM plans for SP data that are either already explicitly written down in DM documentation, or exist in Mario's thoughts. \\
+\S~\ref{sec:issues} -- a brainstorm of potential issues that might arise, organized by pipeline or product, inspired by internal and external documentation. \\
+\S~\ref{sec:science} -- a review of science cases for SP and their potential strain on the DM system, including detailed reduction/analysis case studies in \S~\ref{ssec:science_dmsums}.
+
+\clearpage
 % % % % % % % % % % % % % % % % % % 
@@
 In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community.
-The LSST project will need to define boundaries for these observations.
-Although not all of these boundaries will originate with DM constraints, we start that process here (\textcolor{red}{MLG: Are any initially derived boundaries for Special Programs observations specified elsewhere?}).
-The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}
+The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
+We will need to define boundaries for these observations from a DM prospective.
+The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}.
 
-Start of a list of observing restrictions. \\
-\textbf{Sky coordinates --} (The RA and Dec that will be accessible from the site.) \\
-\textbf{Pointing alt/az, airmass --} (Telescope pointing constraints.) \\
-\textbf{Sky background level --} (e.g., twilight observations cannot start before X) \\
+Start of a list of DM-related considerations for Special Programs data. \\
 \textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
-\textbf{Filters --} (e.g., list filter combinations; $gri$ always be on?; change times, etc.) \\
 \textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
-\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD Main Survey). \\
+\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD).
 
@@
 
-To resolve some open issues we will need to solicit input from the science collaborations (SC).
-However, it may be better and easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers.
-The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
-
-\noindent Recall that there is already a community forum on this topic, so far minimally used but ready to be revised in the future when the next calls soliciting white papers are released.
-\\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
-
-\noindent Remember also that the SC Coordinator for DDF: Neil Brandt.
+To resolve some open issues we will need to solicit input from the science collaborations (SC; also, remember that the SC Coordinator for DDF is Neil Brandt).
+It may be better and easier to wait until the next call for community input, and request that a "DM Requirements" section be included in all white papers (as in \S~\ref{ssec:science_dmsums}).
+Also, there is already a community forum on DDF, so far minimally used but ready to be revived when the next calls soliciting white papers are released.
+\url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
 
@@
 
+\clearpage
 % % % % % % % % % % % % % % % % % % 
-\subsection{A Summary of Open Action Items} \label{ssec:intro_AI}
+\subsection{A Summary of Unresolved Concerns} \label{ssec:intro_UC}
 
-\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
+\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
 \end{enumerate}
-\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
+\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
-\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR.
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
+I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope).
 \end{enumerate}
@@
 
-\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
+\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
 \end{enumerate}
@@
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
 \end{enumerate}
@@
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
 \end{enumerate}
@@
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
 \end{enumerate}
@@
 Knowing the full range of exposure times and image cadences that DM must consider is important for estimating the predicted latencies providing raw, reduced and calibrated images (Section \ref{ssec:dmdocs_L3}) and testing the incorporation of non-standard images into the calibration and difference imaging pipelines (Section \ref{ssec:dmdocs_SPinWFD_L1}).
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
+The science community will also need this information when they propose their DDF cadences and write their Special Programs white papers.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
 \end{enumerate}
@@
 It is unclear whether processed SP data will remain on disk and accessible for as long as it is required by the science users, or if it will only live a short while.
-\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
+\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
 \end{enumerate}
@@
 \textcolor{red}{red if the answer is still to be determined}. \\
-If it seems particularly urgent or involved to answer, it will spawn an Action Item.
+If it seems particularly urgent or involved to answer, it will spawn a \textbf{Concern}.
 
@@
 $\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
 \end{enumerate}
@@
 \\
-$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs?
+$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs.
 
@@
 $\rightarrow$ \textcolor{red}{Unaddressed.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
 \end{enumerate}
@@
 $\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
 \end{enumerate}
@@
 $\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
 \end{enumerate}
@@
 $\rightarrow$ \textcolor{red}{Unassessed.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects.
+\end{enumerate}
 
@@
 \\
-$\rightarrow$ \textcolor{red}{Unanswered.}
+$\rightarrow$ \textcolor{red}{Unanswered.} But since there is talk anyway of using 1x30 instead of 2x15 snaps, I am not spawning a \textbf{Concern} for this question.
 
@@
 $\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR.
 \end{enumerate}
@@
 \\
-$\rightarrow$ \textcolor{red}{Unassessed.} But I would like to take this on in the near future with an assessment of our options, their costs, and the scientific payoff.
+$\rightarrow$ \textcolor{red}{Unassessed.} But I plan to take this on in the near future with an assessment of our options, their costs, and the scientific payoff, and also this is not a question limited to SP so I have not spawned a \textbf{Concern} for this.
 
@@
 It is not comprehensive.
-Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no Action Items are spawned in this section (although there are some low-level unanswered questions).
+Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no \textbf{Concerns} are spawned in this section (although there are some low-level unanswered questions).
 
@@
 However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." Is this true?
-\textcolor{red}{Unanswered}.
+\textcolor{red}{Unanswered}, but this does not seem like an urgent concern for DM, so no \textbf{Concern} is spawned.
 \\
@@
 $\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{OSWP}, suggests using {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}, but is this feasible?
-\textcolor{red}{Unanswered.}
+\textcolor{red}{Unanswered,} but this just seems like a description of a potential Level 3 product, so no \textbf{Concern} is spawned.
 
@@
 Note that we are not including any analysis in these descriptions.
-Also, they aren't that fully realized yet, and could use some more thought and input, which is why none of the questions associated with these case studies have spawned any "Action Items".
+Also, they aren't that fully realized yet, and could use some more thought and input.
 
@@
 Data Acquisition.
-The observational sequence is triggered.
+\\ The observational sequence is triggered.
 In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
@@
 
-Step 2. Inclusion in Level 1 AP. Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+Step 2. Inclusion in Level 1 AP. \\ Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
 
@@
 Delivery of LSST Processed Images.
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
 However, this Special Program appears to require a year of dispersed observations before the Level 3 processing would run.
-\\
-\textcolor{red}{Question:} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
+\end{enumerate}
 
@@
 Level 3 Processing.
-The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
+\\ The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
 Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
@@
 The user is notified with processing status updates.
-\\
-\textcolor{red}{Question:} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
-\\
-\textcolor{red}{Question:} Will the internal real/bogus routine be able to run on shift-and-stack difference images? \\
-\textcolor{red}{Question:} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
 I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables?
+\end{enumerate}
 
@@
 Level 3 Products.
-The shifted-and-stacked images.
+\\ The shifted-and-stacked images.
 Databases of sources detected by the shift-and-stack process with the desired schema, some of which will overlap with Level 1, 2, and/or MOPS to facilitate table joins, some of which will be generated by custom codes.
-\\
-\textcolor{red}{Question:} Will it be possible for Level 3 databases to add arbitrary schema elements, or can they only subselect from pre-defined options? \\
 
-(Step 6.
+Step 6.
 Inclusion in Level 2.
-Given the extreme difference in final depth of these fields, it is unlikely that they will be incorporated into the Level 2.
-However, external Level 3 processing could certainly make deep CoAdds for these fields, join them to the Level 2 DRP catalog, and use them for other science.)
+\\ Given the extreme difference in final depth of these fields, it is unlikely that they will be incorporated into the Level 2.
+However, external Level 3 processing could certainly make deep CoAdds for these fields, join them to the Level 2 DRP catalog, and use them for other science.
 
@@
 Data Acquisition.
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
+\\ On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
 
-Step 2. Inclusion in Level 1 AP. Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+Step 2. Inclusion in Level 1 AP. \\ Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
 
@@
 Delivery of LSST Processed Images.
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
 
@@
 Level 3 Processing.
-The federated pipeline runs automatically upon delivery of the Level 1 processed images.
+\\ The federated pipeline runs automatically upon delivery of the Level 1 processed images.
 If a deep template image does not already exist (this would be a Level 3 product), a Level 2 stacking routine will be used to make them.
@@
 These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases.
-\textcolor{red}{Question:} Will the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
 
@@
 Level 3 Products.
-A deep template for this field, the nightly stacked CoAdds and difference images, and analogs of the {\tt DIAObject} and {\tt DIASource} databases but built only for this specific Deep Drilling Field.
-\\
-\textcolor{red}{Question:} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
+\\ A deep template for this field, the nightly stacked CoAdds and difference images, and analogs of the {\tt DIAObject} and {\tt DIASource} databases but built only for this specific Deep Drilling Field.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
+\end{enumerate}
 
-(Step 6. Inclusion in Level 2. Unlikely.)
+Step 6. Inclusion in Level 2. \\ At the discretion of DM.
 
@@
 
-Future work.
+Step 1.
+Data Acquisition.
+\\ At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (1-5 seconds?).
+Location and exposure times are set by the sky brightness and desired saturation limits.
+Consider: this survey might obtain a larger fraction of high-airmass (i.e., anti-solar) images.
+
+Step 2.
+Inclusion in Level 1 AP.
+\\ Pending a study of short-exposure suitability for psf-matching and differencing with a deeper template, and pending a decision on whether the AP processing capabilities can be extended to accommodate a faster image-input rate, these data could be incorporated.
+
+Step 3.
+Delivery of LSST Processed Images.
+\\ Pending a study of the ISR pipeline's ability to reduce exposures with a short integration time and a high sky background, the raw, reduced and calibrated exposures (and difference images, if created) are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+
+Step 4. Level 3 Processing. \\ User-generated, user-run scripts might include the following, assuming that the ISR pipeline will produce reduced exposures: \\
+(A.) If difference images are required for science and the Level 1 AP is not run, then custom template images and difference images, and catalogs of difference-image sources, could be created using Level 1 or 2 codes.
+If the PSF is not well-enough formed, or there is a problem with DCR corrections (e.g., if obtained at high airmass), them some more sophisticated user-written codes will be necessary.
+\\
+(A2.) If the field includes a nearby supernova, special SN-free template images will be created using Level 1 or 2 image-stacking codes, difference images created with Level 1 DIA codes, and then photometric calibration done using the results of Level 3 processing.\\
+(B.) A script of Level 2 codes to extract and characterize sources in individual exposures, stack exposures into CoAdds (probably not optimized for depth, but maybe for precision), and extract and characterize sources in the CoAdds will be built and run by users.
+A specialized deblender for sources that span a large dynamic range might be necessary?
+\\
+(C.) User-written analysis codes for photometric calibration.
+
+Step 5. Level 3 Products. \\ 
+(A.) Special template images, difference images, and difference-source catalogs. \\
+(B.) Special CoAdds and source catalogs. \\
+(C.) Catalogs of bright objects in the LSST filter set.
+
+Step 6. Inclusion in Level 2. \\ Probably not, given the same regions would be surveyed with WFD exposures, these short exposures probably wouldn't help the DRP's objectives.
 
@@
 
-Future work.
+Step 1. Data Acquisition. \\ The schedule incorporates fields in the Galactic Plane, and executes e.g., 2x15 second exposures.
 
-\subsubsection{GW Optical Counterpart Search}
+Step 2. Inclusion in Level 1 AP. \\ These images are incorporated into the Level 1 pipeline. Alerts are produced and DIAObjects is updated.
 
-Future work.
+Step 3.
+Delivery of LSST Processed Images.
+\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
 
-\subsubsection{A More Generic Level 3 Shift-and-Stack Case Study}\label{sssec:science_dmsums_generic}
+Step 4.
+Level 3 Processing.
+\\ User-created, user-run pipelines that incorporate a mix of Level 1 and 2 DM-provided codes and e.g., specialized deblenders, CoAdds, and photometric calibration routines targeted towards specific stellar populations and/or Galactic latitudes, depending on science goals.
+
+Step 5. Level 3 Products. \\ Possibly a wide variety of difference and CoAdd images and their associated source catalogs.
+
+Step 6.
+Inclusion in Level 2.
+\\ These images are incorporated into the Level 2 DRP to the extent that they assist with the all-sky completeness and photometric calibration; TBD, at the discretion of DM.
+
+\subsubsection{A More Generic Level 3 Shift-and-Stack Case Study, by Mario}\label{sssec:science_dmsums_generic}
 
@@
 have these systems been sized?].
+(\textbf{See \ref{C4}})
 
@@
 In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope).
+\end{enumerate}
 
@@
 We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
+(\textbf{See \ref{C18}, above.})
 
@@
 [This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
+(\textbf{See \ref{C15}}).
 
@@
 But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have \#3, you get \#6 for free as well.]
+ (\textbf{See \ref{C2}}).
 

### 2017-07-31 ms.tex
updates after comments from ECB

@@
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, NCSA about their processing capabilities, and DM about incorporating a functionality for the Alert Pipeline to go into a "skip intervening images" mode (or something similar), for $<30$ second cadence imaging.
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used).
 \end{enumerate}
@@
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
+If so, its outputs might not be reliable because of train/test set mismatch; will the internal LSST routines for real/bogus training be available and adaptable to Level 3 users?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16a} Confirm that resources (packet distribution) for a Level 3 alert stream would be supported and with what latency, whether and how such packets would be identified as non-AP, and that they could be ingested by brokers in real time.
+\end{enumerate}
 \begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
@@
 It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream).
+Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST.).
 Only images that qualify for Level 1 processing will contribute to the Alert Stream.
@@
 The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
-While it is expected that MOPS will be able to identify appropriate images and incorporate as much Special Programs data as possible, the capabilities are currently undefined, and we leave it for the MOPS pipeline builders to decide.
+Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline could be ingested by MOPS.
 We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.
@@
 
-$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA.
-Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred?
-(I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales).
-Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations?
+$\bullet$ Special Programs that use short exposures (e.g., 5 seconds) pose a problem for the Level 1 Alert Pipeline because the higher data acquisition rate (e.g., $\sim$6$\times$ higher) would cause a processing back-up: the allocated processing resource is two parallel chains that take 60 seconds each, and so real-time processing only works for 30 second integrations.
+If there is a demonstrable science need for 60 second latency on each and every one of a consecutive series of short exposures, options for short-term increases in parallel processing power at NCSA will need to be considered (note that this may also be needed for crowded fields with >10k {\tt DIASources}).
+However, the main science goal of a short exposure Special Program may simply be to include brighter stars in the photometric catalogs, and not necessarily to monitor variability on short timescales.
+In this case, there is perhaps no science need to include each and every one of the exposures into the Alert Pipeline, and it might instead be useful enough to just include e.g., every 6$^{\rm th}$ image (which would need only the two parallel processing streams).
+There is currently no written functionality for this option: i.e., there is no way to send only every N$^{\rm th}$ image to the pipeline or, more flexibly, to skip over intervening images and only send an image once the processing pipeline is done with the last.
 \\
-$\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, NCSA about their processing capabilities, and DM about incorporating a functionality for the Alert Pipeline to go into a "skip intervening images" mode (or something similar), for $<30$ second cadence imaging.
 \end{enumerate}
@@
 (E.g., transient source detections in nightly stacks of a deep drilling field.) \\
-$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.
+$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream (but a separate Level 3 alert stream would be possible).
 
@@
 $\rightarrow$ \textcolor{red}{Unaddressed.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used).
 \end{enumerate}
@@
 
-$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)?
-Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales.
-\\
+$\bullet$ These are the database elements reserved for variability characterization: \\
 {\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
 {\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+Will their sizes be adequate to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)?
+Keep in mind that we just need sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and that these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program.
+The concern here is just to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales.
+\\
 $\rightarrow$ \textcolor{red}{Unanswered.}
@@
 $\rightarrow$ \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
-$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.
+$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.\\
+$\bullet$ High-cadence observations to detect fast events, particularly young SNe and Gamma-Ray Bursts. \\
 
@@
 These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
+The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
+If so, its outputs might not be reliable because of train/test set mismatch; will the internal LSST routines for real/bogus training be available and adaptable to Level 3 users?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16a} Confirm that resources (packet distribution) for a Level 3 alert stream would be supported and with what latency, whether and how such packets would be identified as non-AP, and that they could be ingested by brokers in real time.
+\end{enumerate}
 

### 2017-09-01 ms.tex
major revision

--- 2017-06-09 ms.tex
@@
 
-\author{M.~L.~Graham and M.~Juri\'{c}}
+\author{M.~L.~Graham, M.~Juri\'{c}, and E.~Bellm}
 
@@
 \setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys.
-We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey (Levels 1 and 2).
-It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.
-}
+We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey data products(Levels 1 and 2).
+It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.}
 
@@
 
-\noindent The purpose of this study is to: \\
-(1) summarize the current plans of the DM pipelines with respect to Special Programs, \\
-(2) summarize the processing that is required to enable science for Special Programs, \\
-(3) assess whether the DM pipelines meet the expected needs of Special Programs, and \\
-(4) ensure any necessary changes get written into the requirements, designs, and plans.
+The intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
 
-\noindent This will be accomplished by: \\
-(1) starting with the DPDD, whitepapers, etc., and compiling relevant information, \\
-(2) discussing initial issues internally with DM (Mario, the SST), engineering (K-T, Tim), and then \\
-(3) ensuring the identified needs become specified requirements (i.e., spawn change requests).
+\vspace{-20pt}
+\begin{quote}
+LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
+\end{quote}
 
-{\it Current Status: This document is a rough draft, with the goal of being exhaustively complete. Many of the identified \textbf{``Concerns"} may just be misunderstandings or a lack of knowledge.}
+\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent.
+In particular, this study is designed to review: \\
+(1) the processing and products that are required to enable science for Special Programs, \\
+(2) the current plans for DM processing and products with respect to Special Programs, \\
+(3) whether the DM plans meet the expected needs of Special Programs, and \\
+(4) any necessary changes to be written into the requirements, designs, and plans.
 
-The rest of this introduction contains: \\
-\S~\ref{ssec:intro_bounds} -- defining DM-related boundaries for Special Programs. \\
-\S~\ref{ssec:intro_forum} -- existing channels for interaction with Science communities. \\
-\S~\ref{ssec:intro_UC} -- a summary of unresolved \textbf{Concerns} that arose in this study.
+\noindent The concerns raised by this study fall into three main categories: \\
+(1) Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
+(2) Ensuring that DM's intent to process Special Programs data with reconfigured pipelines and provide separate data products that meet the science goals of each program is accurately reflected in all documents, as it is in the DPDD.
+\\
+(3) Clarifying the computational sizing plans for processing and storage of Special Programs data and whether it is this separate from the allocation for end-user Level 3 analysis.
 
-Later sections attempt to identify potential \textbf{Concerns} regarding DM's plans for processing the data from Special Programs with a variety of approaches: \\
-\S~\ref{sec:dmdocs} -- a review of the current DM plans for SP data that are either already explicitly written down in DM documentation, or exist in Mario's thoughts. \\
-\S~\ref{sec:issues} -- a brainstorm of potential issues that might arise, organized by pipeline or product, inspired by internal and external documentation. \\
-\S~\ref{sec:science} -- a review of science cases for SP and their potential strain on the DM system, including detailed reduction/analysis case studies in \S~\ref{ssec:science_dmsums}.
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\section{The Potential Diversity of Special Programs Data} \label{ssec:data}
 
-\clearpage
-% % % % % % % % % % % % % % % % % % 
-\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}
-
-In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community.
-The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
-We will need to define boundaries for these observations from a DM prospective.
-The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}.
-
-Start of a list of DM-related considerations for Special Programs data. \\
-\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
-\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
-\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD).
+To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data.
+The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (standard visit images), then it might not be adequately handled by the DM system.
+To get a handle on the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes (Section \ref{ssec:data_bounds}).
+Next, we explore the current set of proposed science programs for LSST (Section \ref{ssec:data_science}) and the types of observations they would want to do, and then describe when and how we can incorporate revisions to these proposals from the user community in the future (Section \ref{ssec:data_comm}).
+Finally, we explore whether DM can or should set down any technical constraints on how far an observing program can deviate from a sequence of standard visit images and still be processed by DM pipelines, be they the reconfigured and/or the standard Level 1 and 2 pipelines (Section \ref{ssec:data_DMconstraints}).
 
 % % % % % % % % % % % % % % % % % % 
-\subsection{Interacting with the Science Community} \label{ssec:intro_forum}
+\subsection{Observing Boundaries for Special Programs}\label{ssec:data_bounds}
 
-To resolve some open issues we will need to solicit input from the science collaborations (SC; also, remember that the SC Coordinator for DDF is Neil Brandt).
-It may be better and easier to wait until the next call for community input, and request that a "DM Requirements" section be included in all white papers (as in \S~\ref{ssec:science_dmsums}).
-Also, there is already a community forum on DDF, so far minimally used but ready to be revived when the next calls soliciting white papers are released.
-\url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
+We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals.
+So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
+\cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
+Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}).
+We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.
 
-\noindent The questions opened up on that forum are: \\
-(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
-(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
-``Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
+All other aspects of LSST Special Programs remain open to proposals, including: \\
+$\bullet$ additional deep drilling fields \\
+$\bullet$ refined observing strategies for deep drilling fields \\
+$\bullet$ optimized survey areas for the NES, South Pole, and Galactic Plane \\
+$\bullet$ refined observing strategies for the NES, South Pole, and Galactic Plane \\
+$\bullet$ additional mini-surveys \\
+$\bullet$ the total fraction of time spent on Special Programs (nominally 10\%)
+
+There will be some technical limitations on the mini-surveys that will be possible.
+Here's a rough draft list of such potential issues.
+It is not the job of this document to define the camera/observatory technical boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
+This list will be updated in the future (or revised to point to where these boundaries are defined).
+In Section \ref{sssec:dmplans_review_oss}, as we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document we highlight a couple more questions regarding potential boundaries on Special Programs observations.
+
+Filter Changes (and see \ref{OSS-6}): \\
+$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
 \\
-(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
-(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
+$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
+$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
+However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
+\\
+$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
+$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights.
 
-\clearpage
-% % % % % % % % % % % % % % % % % % 
-\subsection{A Summary of Unresolved Concerns} \label{ssec:intro_UC}
+Exposure Times (and see \ref{OSS-5}): \\
+$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
+$\bullet$ The maximum exposure time is not restricted. \\
+$\bullet$ However, for exposure times there are other considerations.
+The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
+Changing the exposure time also affects the photometric and astrometric calibrations.
+Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
+A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
+Additionally, the impact on CR rejection routines is untested for long exposures.
 
-\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
-\end{enumerate}
-\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR.
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
-I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope).
-\end{enumerate}
+Camera Rotation (and see \ref{OSS-7}): \\
+$\bullet$ \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
+Can't remember where I heard it.}
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmdocs}
+Telescope Slews: \\
+$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
 
-In this section we summarize how DM intends to process the data obtained during Special Programs observations.
-This high-level "intent" is a combination of specifications that are actually written down in DM-related documents, and unwritten plans that exist in e.g., Mario's Brain.
-{\it MLG: Note that Sections \ref{ssec:dmdocs_SPinWFD} and \ref{ssec:dmdocs_L3} currently contain a mixture of what is officially written down and what is unofficially planned or expected, usually my interpretation of a conversation with Mario, while Section \ref{ssec:dmdocs_review} is a direct review of the DM documentation.
-Once the SST and engineering have converged on a treatment plan for Special Programs data via this work, my plan is to return to the official DM documentation and identify where change requests need to be initiated.}
+Calibration Images (and see \ref{OSS-2}, \ref{OSS-4}): \\
+$\bullet$ The necessary afternoon calibrations for a Special Program must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained.
 
-\textbf{Overview:} Although the data from some Special Programs might be incorporated into the Level 1, 2, and/or MOPS pipelines, every Special Program will probably also need one or more associated Level 3 processing pipelines.
-These Level 3 processing pipelines will be assembled by the science users, hopefully using existing LSST codes and schemas so that e.g., the Level 3 databases that can be easily joined to the Level 1 and 2 databases.
-It is clearly specified that DM will not write any specialized algorithms for Special Programs data, but that DM will ensure all pipeline code is extendable to Level 3 and that the relevant software infrastructure exists for science users to create and run their Level 3 pipelines.
-In addition, LSST will "federate" some Level 3 pipelines, putting them under change control and taking responsibility for their processing (e.g., shift-and-stack for faint moving objects is expected to become a federated Level 3 processing).
-All user access to processing codes and data products will be through the Science Platform.
+Operations Simulations: \\
+$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
 
-\textbf{Section contents}: \\
-Section \ref{ssec:dmdocs_SPinWFD} -- an overview of whether and how DM will incorporate Special Programs data into the Level 1, 2, and MOPS pipelines and products. \\
-Section \ref{ssec:dmdocs_L3} -- an overview of how federated and user-run Level 3 pipelines will process Special Programs data, incorporate WFD data, and/or join to Level 1 and 2 products. \\
-Section \ref{ssec:dmdocs_review} -- a detailed break-down of Special Programs (or Level 3) related content in the three DM documents (DPDD, DMSR, and DMAD), and non-DM LSST documents. \\
+Total Time: \\
+$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey), but this could be changed if significant science is enabled.
+\\
 
 % % % % % % % % % % % % % % % % % % 
-\subsection{Incorporating SP Data into the WFD Main Survey}\label{ssec:dmdocs_SPinWFD}
+\subsection{Currently Proposed Special Programs} \label{ssec:data_science}
 
-There is not much, if anything, specified regarding guidelines or expectations for incorporating Special Programs data into the Level 1 and Alert Stream, Level 2 DRP, and/or MOPS pipelines that will automatically run on all WFD main survey observations.
-The concern is that drastic changes in the exposure time or total depth of the observations might negatively impact the science goals that depend on the WFD main survey data products.
-Here we summarize these concerns.
+In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed.
+We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science; this Section is mainly descriptive and we do not spawn any {\bf Concerns} here.
+In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs; any {\bf Concerns} that arise are detailed in that section.
+Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
 
-\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmdocs_SPinWFD_L1}
-
-It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream (e.g., no Level 3 difference imaging pipeline can send its detections to the Alert Stream).
-Only images that qualify for Level 1 processing will contribute to the Alert Stream.
-The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template.
-So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine.
-However, the suitability of shorter and longer exposures is unknown.
-Here we consider the suitability -- or what we need to do in order to determine the suitability -- of shorter and longer exposures for the Level 1 and Alert Stream.
-These considerations are made independently of the science motivations of the Special Programs that proposed the different exposure times.
-
-% Shorter exposures: twilight survey, getting bright stars on the same photometric system, variability in brighter stars
-% Longer exposures: the weak lensing community might propose long-exposure fields if they remain concerned that 2x15 seconds is insufficient to obtain a stable PSF with no spatial variation dominated by the atmosphere
-
-(1) Shorter Exposures.
-The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint.
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified.
-We consider this issue separately in Section \ref{ssec:issues_calibrations}.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
-
-(2) Longer Exposures.
-There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
-One possibly useful study is Chang et al.
-(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
-
-\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream.
-\end{enumerate}
-
-\subsubsection{MOPS}\label{ssec:dmdocs_SPinWFD_MOPS}
-
-The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
-While it is expected that MOPS will be able to identify appropriate images and incorporate as much Special Programs data as possible, the capabilities are currently undefined, and we leave it for the MOPS pipeline builders to decide.
-We note that since MOPS does not run during the night, it does not have the same latency requirements, or the associated issues.
-
-\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmdocs_SPinWFD_L2}
-
-Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
-When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
-For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
-In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{SP Data Processing with Federated and User-Run Level 3}\label{ssec:dmdocs_L3}
-
-Through the Science Platform, member of the science community can assemble Level 3 processing packages using both LSST and custom codes.
-The DM team is required to write their codes to be extendable to Level 3, and to provide the infrastructure for users to run them on LSST servers, monitor their progress, and analyze their output (including a prioritization system to allocate processing resources).
-All Level 3 pipelines will have access to processed single visits from both the WFD main survey and Special Programs, as well as difference images, CoAdds, and catalogs from the Level 1 and 2 pipelines.
-It is expected that some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team, such as those developed for time domain science that need to run automatically and repeatedly.
-For both federated and user-run code, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
-
-Level 3 pipelines for Special Programs data for time-domain astronomy will need access to images on a timescale from minutes to weeks after acquisition, as opposed to static-sky Level 3 pipelines that may run on e.g., a yearly schedule like the Level 2 DRP.
-It is currently unclear what that real-time latency will be, and how processing resources will be allocated; here we break down some of the options.
-
-If the images obtained with a Special Program are incorporated into Level 1 (see Section \ref{ssec:dmdocs_SPinWFD}), then the Level 3 processing can begin as soon as the Level 1 imaging products are available: raw images, processed single visits (a.k.a.
-calexps), difference images, and their source catalogs.
-The current maximum time for this is {\tt \textbf{L1PublicT}}, specified by LSE-29 (\citep{LSE-29}) to be 24 hours (LSR-REQ-0104) -- but this seems far too long, especially for federated Level 3 pipelines, given that they are finished within $60$ seconds.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs.
-\end{enumerate}
-
-If the images obtained with a Special Program cannot be incorporated into Level 1, then the latency with which they will be made available is unspecified.
-Ultimately, their delivery will depend why they are not incorporated into Level 1, which will probably fall into two categories: (1) there is no template at their location, or (2) their exposure time prohibits difference imaging.
-In the former case, it seems likely that single processed images can be provided on a timescale of minutes, but this needs to be specified.
-In the latter case, it will depend on the exposure time.
-As mentioned in Section\label{ssec:dmdocs_SPinWFD_L1}, the minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint, but is currently unspecified, and discussed separately in Section \ref{ssec:issues_calibrations}.
-It may be that single-visit processing is simply not supported by DM below some exposure time threshold, and so the latency on raw image delivery should be specified.
-Although it is likely that slightly shorter/longer exposure images can be reduced and calibrated with the same pipeline as $15$ second exposures, even if they are unsuitable for difference imaging, this has not yet been confirmed, and needs to be investigated and specified.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline.
-\end{enumerate}
-
-In the DPDD \citep{LSE-163}, it specifies that Special Programs data processing will be limited to not more than 10\% of the DM computational and storage facilities.
-During the telecon on Wed Mar 29, KT Lam reviewed the computational budget, but it seemed mainly for WFD with only passing mention of DDF or Special Programs.
-This 10\% of the computational budget seems like an underestimate: although SP data will not be more than 10\% of the total data volume, it will be multiply processed: once as WFD Level 1, again perhaps as Level 2, and then possibly in multiple pipelines as Level 3, some of which are extremely computationally intensive (e.g., shift-and-stack).
-It seems that we could easily find that Special Programs data are processed $\sim4\times$ as much as WFD.
-Furthermore, it is unclear whether the location for all this processing has been identified; Mario asks, for federated real-time Level 3 processing, {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that?s used to process calibrations.
-Have these systems been sized?"}.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place.
-\end{enumerate}
-
-\subsection{In-Depth Documentation Review}\label{ssec:dmdocs_review}
-
-In this section we attempt to track the flow-down requirements through the official LSST and DM documentation.
-We find some hardware-related constraints in the OSS that will set the maximum possible divergence of Special Programs raw images from WFD main survey images (i.e., exposure times, rate of filter changes), and these will be relevant to DM processing.
-We also find that some broad, high-level requirements for Level 3 and Special Programs are written down in the DMSR.
-The next document down-flow, the DMAD, is focused on the Level 1 and 2 algorithms, and although its Section 7 mentions the DAX and SUIT, it says that they are to be detailed in a different document.
-The DPDD also does not have much content that directly pertains to Special Programs and Level 3 processing.
-
-\noindent \textbf{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\\
-The SRD makes no direct reference to Special Programs, deep drilling fields, or mini-surveys. It's scope is the science of the WFD only.
-
-\noindent \textbf{LSST System Requirements, SR, LSE-29, \cite{LSE-29}}\\
-The SR is derived from the SRD.
-It contains the high-level specifications about fraction of total observing time delegated for Special Programs, minimum exposure times, maximum filter change time, etc., that flow-down to and are repeated in the DMSR and OSS (discussed below).
-
-\noindent \textbf{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\\
-The OSS contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from the SRD and SR).
-There are two points worth mentioning: \\
-$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
-$\rightarrow$ There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times.
-However, there doesn't seem to be a specified limit on 'thermal stability'.
-Knowing the full range of exposure times and image cadences that DM must consider is important for estimating the predicted latencies providing raw, reduced and calibrated images (Section \ref{ssec:dmdocs_L3}) and testing the incorporation of non-standard images into the calibration and difference imaging pipelines (Section \ref{ssec:dmdocs_SPinWFD_L1}).
-The science community will also need this information when they propose their DDF cadences and write their Special Programs white papers.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications.
-\end{enumerate}
-$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
-$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
-$\rightarrow$ The specified maximum filter change time could be a debilitating overhead for, e.g., a deep drilling cadence of consecutive 30 second images wants to rotate through filters.
-Furthermore, we might expect that in reality the filter change time will be shorter than 120 seconds, which would enable a DDF to sequence filters.
-Furthermore, while the minimum filter change count is not an issue, I can't find the nightly (or total lifetime) maximum to limit wear and tear on the mechanism (as I've heard about).
-However, there is already a change request in progress regarding this issue: \url{https://project.lsst.org/groups/ccb/node/938}.
-
-\noindent \textbf{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}} \\
-The DMSR contains the high-level requirements for all processing and products. There are two main areas that pertain to Special Programs. \\
-$\bullet$ Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069)\\
-$\rightarrow$ Specifies that {\it "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing.
-They will be re-generated for users on-demand using the latest processing software and calibrations"}, but also says that {\it "This aspect of the processing for Special Programs data is specific to each program"}.
-It is unclear whether processed SP data will remain on disk and accessible for as long as it is required by the science users, or if it will only live a short while.
-\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request.
-\end{enumerate}
-$\bullet$ Section 2.9, "Level 3 Production" \\
-Specifies services of the DMS for Level 3 processing which are relevant to Special Programs: \\
-2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
-2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
-2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. \\
-2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
-2.9.5/6.
-Software Frameworks (ID: DMS-REQ-0125,DMS-REQ-0128): to provide a means for applying user-provided processing to catalog and image data, and for assessing the completeness of that application.
-\\
-3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. \\
-$\rightarrow$ \textcolor{red}{Questions}: Regarding 2.9.3, I'm not sure what {\it "self-consistent inputs"} means and so cannot assess any potential issues.
-Regarding 3.1, I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?
-
-\noindent \textbf{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\\
-LDM-151 outlines the code and algorithms that will be used in the processing of images in the AP (Level 1), DRP (Level 2), and MOPS.
-As such, it does not say much about Special Programs.
-It does say that {\it "LSST DM is required to facilitate the creation of Level 3 data products by providing suitable APIs, software components, and computing infrastructure, but will not by itself create any Level 3 data products"}.
-The latter means that DM will not create any processing algorithms that are specific to the needs of any Special Programs data.
-Section 7 briefly describes the Data Access and the Science User Interface and Tool (SUIT), also known as the Science Platform, with a mention that they will be described in full in later documents  (probably this one: \url{https://github.com/lsst-dmsst/lsst_science_platform/blob/master/science-platform.pdf}).
-The Data Access layer (DAX) will be for sharing, accessing, searching (etc.) products, and the SUIT/Platform will provide an interactive environment for users, and also serve as portal for Level 3 pipeline development (including installed version of the LSST Software Stack), execution, retrieval, and analysis.
-
-\noindent \textbf{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\\
-The DPDD specifies that the data products for Special Programs will {\it "be created using the same software and hardware as Levels 1 and 2"}.
-Like LDM-151, Section 7 is devoted to the data products of Special Programs, explaining how the current plan is to have both the Level 1 and 2 processing pipelines run on SP data in the same way that they run on the WFD, and produce all of the same data products and alerts, when possible.
-It also states that the Level 3 products for SP data can (should) be kept in separate, but joinable, databases that have identical schema as the WFD databases.
-
-\noindent \textbf{Level 2 Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}\\
-LSE-180 is built on OpSim runs that do include some nominal deep drilling fields, but the photometric calibration investigated in this work does not deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs -- it's scope the WFD Main Survey.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{Questions and Potential Issues}\label{sec:issues}
-
-In this section we list questions or potential problems identified from reading LSST documents \citep{LSE-163,LDM-151}, Gregory's AHM 2016 slides\footnote{``DM Considerations for Deep Drilling", LSST AHM Aug 2016 presentation by Gregory Dubois-Felsmann \url{https://zenodo.org/record/61402\#.WNVk6hIrIUF}}, or using our imaginations.
-Another useful resource is the LSST Database Schema Browser: \url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}.
-
-For each of these questions or potential issues, a response will be: \\
-\textcolor{blue}{blue if the answer is known} (even if the answer is a negative), \\
-\textcolor{red}{red if the answer is still to be determined}. \\
-If it seems particularly urgent or involved to answer, it will spawn a \textbf{Concern}.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Alert Stream}\label{ssec:issues_alerts}
-
-$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA.
-Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred?
-(I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales).
-Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations?
-\\
-$\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging.
-\end{enumerate}
-
-$\bullet$ LDM-151, Section 3.3.4 `Alert queuing and persistance' mentions that the {\it ``event message stream and the AlertDB will be synchronized at least once every 24 hours"} and that {\it ``Prior to the start of the subsequent night's observations, the message queue will be flushed and synchronized with the AlertDB.
-It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field?
-\\
-$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs.
-
-$\bullet$ Will near-real-time Level 3 difference-image processing be able to contribute to the same Alert Stream as Level 1?
-(E.g., transient source detections in nightly stacks of a deep drilling field.) \\
-$\rightarrow$ \textcolor{blue}{Answered.} As described in LDM-151, the Alert Stream takes as input only {\tt DIAObjects} and specifies that {\it "all alerts should be transmitted within 60s of the closure of the shutter of the final snap within a visit."} Based on this, any Level 3 difference imaging sources would not contribute to the Alert Stream; only imaging that qualifies as Level 1 will contribute to the Alert Stream.
-
-$\bullet$ Alerts on a {\tt DIASource} will link to the associated {\tt DIAObject}, but in cases where a Special Program obtains a sequence of images without switching fields, will the {\tt DIAObject} catalog have had enough time to be updated to include the {\tt DIASource} from the previous image?
-\\
-$\rightarrow$  \textcolor{blue}{Answered.} Immediate updates of the {\tt DIAObject} catalog is an expected capability, although it is currently uncertain how it will be accomplished, technically.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Databases and Schema}\label{ssec:issues_databases}
-
-$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products.
-\\
-$\rightarrow$ \textcolor{red}{Unaddressed.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image.
-\end{enumerate}
-
-$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to Special Programs, but it is unclear how such targets will be identified or flagged as such in the databases, and unclear whether we need to add an element to the database schema for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?
-\\
-$\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry.
-\end{enumerate}
-
-$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)?
-Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales.
-\\
-{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-$\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-\end{enumerate}
-
-$\bullet$ Low surface-brightness features and/or faint extended objects are required for some galaxy science, and will show up more in e.g., a DDF than a WFD CoAdd.
-Will the related shape parameters such as extendedness in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features that might come from deeper Special Programs CoAdds?
-As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program.
-Furthermore, it's unlikely that extremely deep patches will be included in the Level 2 products, but this still should be assessed.\\
-$\rightarrow$ \textcolor{red}{Unassessed.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects.
-\end{enumerate}
-
-$\bullet$ Images obtained during Special Programs might be used with one or more different types of templates -- the same images could be used as multiple surveys.
-Is it feasible to have separate databases for each?
-\\
-$\rightarrow$ \textcolor{blue}{Answered.} Yes, Level 3 pipelines can create as many separate databases as necessary.
-There should at least be a separate database for each kind of template image.
-These databases can have the same base schema, and elements of the schema can be ``turned off" if the survey will not use them, to save processing time and disk space.
-
-$\bullet$ Data from Special Programs might be incorporated into the Level 1 and/or 2 databases; if so, how will it be flagged in these databases? \\
-$\rightarrow$  \textcolor{blue}{Answered.} In the schema for {\tt DIASource} and {\tt Source} there is an element for {\tt ccdVisitId}, and the database {\tt CcdVisit} has entry for {\tt visitId}, and the database {\tt Visit} has an entry for {\tt programId}, which is currently an {\tt integer} containing the {\it ``Observing program id (e.g., universal cadence, or one of the deep drilling programs, etc.)."}.
-
-$\bullet$ Will a Level 3 database for a Special Program be able to generate and incorporate forced photometry from WFD images? \\
-$\rightarrow$ \textcolor{blue}{Answered.} Yes. All Level 3 pipelines have access to all LSST data.
-
-$\bullet$ Will the Level 1 requirement to generate forced photometry for all new {\tt DIAObjects} within 24 hours be able to run on Level 3 image products of Special Programs data, such as deep nightly stacks of DDF images?
-\\
-$\rightarrow$ \textcolor{blue}{Answered.} No.
-This forced photometry will run only on other images that are incorporated into Level 1, and this will not include the data products of Level 3 pipelines.
-Performing forced photometry for new {\tt DIAObjects} on a Level 3 image product would itself be a Level 3 pipeline.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Templates}\label{ssec:issues_templates}
-
-$\bullet$ Typically we think of a template as a single deep coadd of image obtained e.g., $\Delta t \geq 1$ year ago.
-However, the use of a template with constant $\Delta t$ to the new images (e.g., $\Delta t = 7$ days to mitigate proper motion induced issues in a difference image for a Galactic plane mini-survey), or the use of a template that is simply a specific other image, might be scientifically necessary.
-Will this be supported by DM?
-\\
-$\rightarrow$ \textcolor{blue}{Answered.} Yes.
-Difference imaging with a custom template is Level 3, and the requirement for the generation of custom templates already exists as \textbf{DMS-REQ-0032:} The DMS shall provide software to perform image differencing, generating Difference Exposures from the comparison of single exposures and/or coadded images \citep{LSE-61}.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Snap Handling}\label{ssec:issues_snaps}
-
-$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}.
-Could Special Programs data that is only a single exposure, not a visit that is the combination of two snaps, be incorporated into the Level 1 AP?
-\\
-$\rightarrow$ \textcolor{red}{Unanswered.} But since there is talk anyway of using 1x30 instead of 2x15 snaps, I am not spawning a \textbf{Concern} for this question.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Calibrations}\label{ssec:issues_calibrations}
-
-$\bullet$ \textbf{Image Reduction} The Special Programs data may have a different exposure time or dither pattern than the WFD survey.
-The concern here is that if the instrument signature removal (ISR) calibration pipeline, and/or the photometric/astrometric calibration pipeline, have been designed to work for WFD data that problems might be encountered for Special Programs data (e.g., dark frames; brighter-fatter correction; artifacts with an exposure time dependence; charge build-up from no-dither with bright stars).
-Since there already exists a requirement that DM deliver processed visit images (ID: DMS-REQ-0069) to users (and to internal pipelines), it seems that at least the ISR must be developed for a variety of exposure times, but is this truly the case?
-Or are there exposure time boundaries beyond which DM will not guarantee delivery of processed images?
-\\
-$\rightarrow$ \textcolor{red}{Unanswered.}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR.
-\end{enumerate}
-
-$\bullet$ Regarding the above point, LSE-180 offers some relevant information regarding the reduction and calibration of non-WFD images: \\
-$\rightarrow$ LSE-180 mentions that it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
-$\rightarrow$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD the corrections can be applied to the much smaller amount of data from the Special Programs (unless it is a function of exposure time and the Special Program uses a different exposure time).
-
-$\bullet$ \textbf{Image Detection Efficiency} Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
-Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
-What are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
-(The answer should distinguish between planting fakes in the single images for transient detection efficiencies with difference imaging, $vs.$ planting fakes in the CoAdds for point-source limiting magnitudes).
-\\
-$\rightarrow$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
-will record information on the pixel variance ...
-as well as the per-pixel masks ...
-These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
-\\
-$\rightarrow$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
-detection efficiency for point sources vs.
-mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
-\\
-$\rightarrow$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
-The details of what metrics will be mapped, the format and scale of the maps (e.g.
-hierarchical pixelizations vs.
-polygons), and the way the metrics will be computed are all unknown"}.
-It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
-(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
-Source detection thresholds can be estimated through the use of sky sources"}.)\\
-$\rightarrow$ Suchyta et al.
-(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
-\\
-$\rightarrow$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
-\\
-$\rightarrow$ \textcolor{red}{Unassessed.} But I plan to take this on in the near future with an assessment of our options, their costs, and the scientific payoff, and also this is not a question limited to SP so I have not spawned a \textbf{Concern} for this.
-
-$\bullet$ Multiple Special Programs request short-exposure survey, which would provide less/different stars for photometric/astrometric calibration.
-Would a 1s exposure contain enough stars for calibration?
-\\
-$\rightarrow$ \textcolor{blue}{Answered.} 1 second exposure will contain stars from $12.9<r<21.0$, which should be more than enough for a full photometric calibration.
-However, LSE-180 only considers calibration for the standard WFD data, and so this has not been tested.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Custom Co-Adds}\label{ssec:issues_coadds}
-
-$\bullet$ Building deep co-adds with no transient contamination (e.g., SN hosts, cosmology uses host parameters correlated with intrinsic SN brightness), or creating deep stacks across filters for faint-object detection at $<5\sigma$ in a single filter, may be a desired capability for Special Programs data.
-Will this be possible?
-\\
-$\rightarrow$ \textcolor{blue}{Answered.} Yes.
-Creating custom co-adds will be a Level 3 capability that will be possible with the Level 2 DRP pipeline algorithms described in LDM-151.
-This includes multi-filter stacks (DMSR Section 1.3.4.
-'Multi-band Coadds' ID:DMS-REQ-0281).
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Moving Objects}\label{ssec:issues_mops}
-
-$\bullet$ For faint moving objects, users will develop shift-and-stack (SAS) pipelines in Level 3 (MOPS does not SAS).
-Will they have all the necessary infrastructure for this intensive process?
-Will they be able to use the LSST real/bogus on their shift-and-stacked frames?
-Will they be able to access the difference images over long time scales?
-\\
-$\rightarrow$ \textcolor{red}{Unanswered.} All of this seems likely, as each of these questions is covered by the DM requirement to {\it "make all codes extendable to Level 3"}, but ensuring that the answer is yes to each of them will be more involved in reality, and probably happen only with time as the SAS Level 3 pipeline is developed and integrated?
-
-$\bullet$ Can MOPS link newly discovered {\tt sources} to those discovered earlier on the same night (e.g., in a 4-visits-per-night scenario), or does it have a similar potential problem to {\tt DIASource} and {\tt DIAObject}?
-\\
-$\rightarrow$  \textcolor{blue}{Answered.} MOPS does not run during the night in real time. This is not a problem.
-
-\clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\section{Science Cases and their DM Requirements} \label{sec:science}
-
-In this section we review science documents from LSST and the community.
-We compile information about the nominal observing plans for Special Programs that are already written down (Section \ref{ssec:science_plans}), and summarize the science motivations for Special Programs that have so far been proposed (Section \ref{ssec:science_descriptions}).
-Potential issues for DM are inferred from these science motivations and have been incorporated into our list of issues in Section \ref{sec:issues}.
-In Section \ref{ssec:science_dmsums}, we explore some Special Programs processing case studies as an additional avenue to identify potential issues.
-
-\noindent Resources: \\
-$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" \cite{2008arXiv0805.2366I} \\
-$\bullet$ The LSST Science Requirements Document \cite{LPM-17} \\
+\noindent {\bf Resources:} \\
+$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
+$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
+$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
 $\bullet$ ``General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
-$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}\\
+$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
 $\bullet$ ``LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
-$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{OSWP} \\
-$\bullet$ List of LSST Deep Drilling white papers: \url{https://project.lsst.org/content/whitepapers32012} \\
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Nominal Observing Plans for DDF and MS}\label{ssec:science_plans}
-
-Ivezi\'{c} et al.
-(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
-Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
-The LSST Science Requirements Document \cite{LPM-17} doesn't contain the terms ``deep drilling field" or ``mini-survey".
-Four extragalactic deep drilling fields have already been specified (Table \ref{tab:ddfms}).
-From a scouring of mainly the presentations of Brandt and Ridgway at the 2016 AHM, an {\it incomplete} list of potential mini-surveys that people are thinking about in Table \ref{tab:ddfms} also.
+$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
 
@@
 
-% % % % % % % % % % % % % % % % % % 
-\subsection{Science Descriptions (Not Comprehensive)}\label{ssec:science_descriptions}
-
-The following is a collection of rough notes on the science goals for some proposed Special Programs, classified by area of astronomy.
-It is not comprehensive.
-Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no \textbf{Concerns} are spawned in this section (although there are some low-level unanswered questions).
+\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
+(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
+Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
 
 \medskip
-\noindent \textbf{Solar System Objects}\\
-$\bullet$ Observations in the North Ecliptic Spur region will yield more $\geq140$ m NEOs (Brandt talk). \\
-$\bullet$ Detecting faint SSO down to $r\sim27$ by applying shift-and-stack processing to a mini-survey (Trans-Neptunian Objectss, Trojans, asteroids, long-period comets, dwarf planets).
-\citep{BeckerWP} \\
-$\rightarrow$ Regarding shift-and-stack, \cite{BeckerWP} says that ``the multi-fit algorithm ...
+\noindent \textbf{Solar System Objects (SSO)}\\
+$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
+$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
+$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) 
+$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
+$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
+Regarding SAS, \cite{BeckerWP} says that ``the multi-fit algorithm ...
 naturally provides a base infrastructure for this process.
 In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required.
-However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." Is this true?
-\textcolor{red}{Unanswered}, but this does not seem like an urgent concern for DM, so no \textbf{Concern} is spawned.
+However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
+$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
+The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures.
+Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night.
+The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
+See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_MM} for more detailed DM processing case studies.
+
+\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
+$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
+When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
 \\
-$\bullet$ Mini-moons, temporarily captured satellites of the Earth. (Section 10.2, \cite{OSWP}) \\
-$\bullet$ Find meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{OSWP}) \\
-$\bullet$ Sweetspot: twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{OSWP})
-
-\noindent \textbf{Stars in the Milky Way and Magellanics} \\
-$\bullet$ Deep imaging acquired over a short time span to mitigate proper motion loss (and increase flare detection rates) will identify useful stellar populations. \citep{DhitalWP}. \\
-$\rightarrow$ When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. \\
-$\rightarrow$ Special co-adds may be required, e.g., {\it "To reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
+$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}.
+For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
 \citep{SzkodyWP}.
 \\
-$\bullet$ A Twilight Short Exposure survey would enable bright stars (with many historical observations) to be put solidly on the same photometric system as the deeper LSST WFD main survey catalog. \\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{OSWP}, suggests using {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}, but is this feasible?
-\textcolor{red}{Unanswered,} but this just seems like a description of a potential Level 3 product, so no \textbf{Concern} is spawned.
+$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog.
+\\
+$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+\\
+$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
+Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks.
+The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars.
+See Sections \ref{ssec:SPCS_GPVSEx} and \ref{ssec:SPCS_M67} for more detailed DM processing case studies.
 
@@
 The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence.
-Microlensing events can also be detected with this data set.
-In both cases, follow-up is required.
-The Galactic Plane MS is proposed for this.
-\cite{2013arXiv1304.3455G} describes how transits can be extract from the same data set.\\
+A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
+\cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
 $\bullet$ Microlensing.
-Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates (for follow-up with e.g., LCO).
-However, more Galactic regions must be included, and this will require image differencing in crowded fields.
-\cite{2013arXiv1304.3455G} claims that imaging in the necessary regions of the galaxy has been disfavored by the project on the basis that the eventual deep co-adds would be uselessly confusion limited.
-(But was this before the Galactic Plane region was defined?)
+Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
+However, this will require image differencing in crowded fields.
+\cite{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
+Microlensing can also be done with DDF-style observing at low Galactic latitudes.
+\\
+$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images.
+They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
+See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.
 
 \noindent \textbf{Supernovae} \\
-$\bullet$ Extend the SN sample to $z\sim1.2$ and providing more densely sampled light curves for cosmological analyses with the nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$.
+$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
+The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
+\cite{KesslerWP} \\
+$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}.
 \\
-%$\bullet$ \cite{CrottsWP} doesn't say much specific. \\
-$\rightarrow$ The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$, which adds up to $<60$ minutes and could presumably be done with the standard 30s exposures.
-\cite{KesslerWP} \\
-$\bullet$ Generating the best-possible individual SN light curves by building special, deep-as-possible, SN-free host galaxy images and using them as a template.
-This will be necessary for SNe that appear in the template image (e.g., that last $>1000$ days).
+$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
+This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days.
+These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss.
 SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
+The latter, specifically the ``characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
 \\
-$\rightarrow$ The Level 2 DRP codes to create CoAdds should be adequate for this. \\
-$\rightarrow$ \cite{FergusonWP} also mentions {\it ``characterization of ultra-faint SN host galaxies"} in their Galaxies WP. \\
-$\bullet$ A short-exposure survey could include nearby SNeIa on the same photometric system -- but again this comes back to the short-exposure questions of whether we can do photometric calibration as well.
+$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
+$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
+Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
+In particular, the Level 2 DRP codes to create ``transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
+See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
 
 \noindent \textbf{Galaxies} \\
-$\bullet$ To build  a large collection of low-$\mu$ objects: \cite{FergusonWP} mentions {\it ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations"} and {\it ``characterization of low-surface-brightness extended features around both nearby and distant galaxies"}.
+$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
+\cite{FergusonWP} mentions ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and ``characterization of low-surface-brightness extended features around both nearby and distant galaxies".
 \\
-$\bullet$ Characterization of high-$z$ clusters (will will depend on ability to deblend extended objects). \\
-$\bullet$ AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP}  \citep{GawiserWP} \\
-$\rightarrow$ Study the effects of AGN on e.g., bulge-disk decomposition parameters?
+$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
+$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
+$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
+In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.
 
 \noindent \textbf{Weak Lensing} \\
-$\bullet$ DDF can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ {\it "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (community forum, Jim Bosch) \\
-$\bullet$ {\it "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (community forum, Jim Bosch)  \\
-$\bullet$ Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.
+$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
+$\bullet$ {\it Jim Bosch -- ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
+$\bullet$ {\it Jim Bosch -- ``Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
+We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
+$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
+$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
 
 % % % % % % % % % % % % % % % % % % 
-\subsection{Special Programs Processing Case Studies}\label{ssec:science_dmsums}
+\subsection{Future Input from Data Rights Holders} \label{ssec:data_comm}
+
+Community input on LSST Special Programs will be ingested through a formal call for white paper proposals (Section \ref{sssec:data_comm_call}).
+This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}.
+The DM team will also interact informally with the community through, e.g., a {\tt \Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
+
+\subsubsection{Call for White Paper Proposals}\label{sssec:data_comm_call}
+
+The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
+At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call.
+The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
+The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director.
+The criteria to be met by the proposals may include specifications that they: satisfy minimum technical requirements (feasibility, overheads); maximize diverse scientific objectives (serve a wide community); and generate legacy datasets and add value to the products of other facilities.
+After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation.
+While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \cite{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
+
+\textbf{Initial community suggestions regarding the call for white papers.} \\
+$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
+$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations.
+Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers.
+\\
+$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an ``eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
+$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).
+
+\subsubsection{{\tt \Large{Community}} Forum}\label{sssec:data_comm_forum}
+
+As the science community prepares their white paper proposals they have been invited to interact with Data Management through this {\tt \Large{Community}} forum on DDF.
+I am a watcher on this forum.
+So far it has been minimally used.
+\\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
+
+\noindent The questions opened up on that forum are: \\
+(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
+(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
+``Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
+\\
+(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
+(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:data_DMconstraints}
+
+It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines.
+Exposures which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+We consider shorter and longer exposure in turn.
+
+(1) Shorter Exposures.
+The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint.
+The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
+Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
+A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
+However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
+As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for ``well-formed").
+Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
+\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
+At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
+We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
+\label{fig:expt}}
+\end{center}
+\end{figure}
+
+(2) Longer Exposures.
+There is no maximum exposure time specified for an LSST image.
+Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
+However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
+With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+
+In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
+One possibly useful study is Chang et al.
+(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
+They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
+They present a new software package to do mitigate the effects.
+Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
+2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Summary}\label{ssec:data_summary}
+
+Our review of the technical boundaries and previously proposed Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, and exposures that are obtained with a bright sky background during twilight.
+
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
+
+As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
+
+\begin{quote}
+LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
+\end{quote}
+
+In this section we review in greater detail the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
+Later on, in Section \ref{ssec:dmplans_review}, we review each piece of DM documentation in turn and identify requirements and/or specifications that do not adequately represent the intent of LSST DM, as defined in Section 6 of the DPDD, with respect to Special Programs processing.
+Where relevant we suggest clarifications, changes, or additions to these documents.
+
+%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C1} There are currently no requirements on DM to incorporate data from Special Programs; should there be?
+\end{enumerate}
+%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C2} Should DM undertake an investigation to define the exposure boundaries (integration time, sky background) for non-standard visit images that can be processed by DM pipelines, such as ISR and DIA?
+\end{enumerate}
+%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C3} Given the double-processing of Special Programs data, is the system sized appropriately? \end{enumerate}
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
+
+It seems obvious that LSST DM would want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
+In the following three sections we summarize when and how SP data might be incorporated into the pipelines and products of the Level 1 and Alert Stream, the Level 2 DRP, and/or MOPS.
+
+\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
+
+It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
+Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST.
+Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
+The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template.
+So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine.
+However, the suitability of shorter and longer exposures is unknown, as discussed in Section \ref{ssec:data_DMconstraints}.
+
+\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
+
+The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
+Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS.
+As discussed under ``Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
+
+\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
+
+Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
+We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
+When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
+For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
+In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
+
+LSST DM intends to reconfigure its existing pipelines in order to generate imaging and catalog products for Special Programs data, whenever possible.
+In this context, possible means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
+In a possible scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build unique, separate image and catalog products that meet the science needs of that particular program.
+For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
+This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Intent to Support Special Programs Processing with Level 3 Pipelines}\label{ssec:dmplans_L3}
+
+If a Special Program requires processing that cannot be accomplished with existing DM algorithms and codes, and/or requires an intensive amount of additional computational resources, it is considered Level 3 -- and in the latter case the processing might have to be undertaken on a separate computational resource.
+To support such Level 3 processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
+In most cases, data export and processing on external computation resources will not be necessary.
+The majority of users will access the data, build pipelines based on DM codes, and execute them through the Science Platform \cite{LDM-542}.
+Through the Science Platform, DM provides the infrastructure for users to run their Level 3 pipelines on LSST servers, monitor their progress, and analyze their output.
+DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
+All of this applies to both Level 3 processing for Special Programs data, and reprocessing of WFD main survey data.
+It is furthermore expected that, over time, some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team.
+For both federated and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
+
+% % % % % % % % % % % % % % % % % %
+\subsection{Documentation Review}\label{ssec:dmplans_review}
+
+Here we highlight every requirement or issue relevant to Special Programs that we could find in the existing documentation.
+Since the goal of this work is to make any needed changes to existing documentation, we have endeavored to directly tie every possible concern to the relevant sections of existing documentation (or, where possible, to requirement codes) so that we can more easily direct future changes back into the existing documentations.
+
+% % % % % % %
+\subsubsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{sssec:dmplans_review_srd}
+
+This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
+The SRD makes two mentions of Special Programs: that the ``LSST is well suited to conducting Deep Supernova Survey" (Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (Section 3.4).
+The latter we know to be an approximate estimate.
+
+% % % % % % %
+\subsubsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{sssec:dmplans_review_lsr}
+
+The LSR is derived from the SRD, and converts the high-level specifications from the SRD into system requirements, which flow-down to and in some cases repeated in the OSS and DMSR (next sections).
+We have the version of LSE-29 last revised on August 4, 2016.
+
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0} Most notably, there is no written requirement for DM to reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, as it intends to.
+Should there be, and is the LSR right right place for it?
+\end{enumerate}
+
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0b} Similar to \ref{LSR-0}, there is no written requirement for DM to evaluate the benefit of incorporating data from Special Programs into the products of the WFD main survey, and then to do this when it is found to enable/improve science results, as it intends to.
+Should there be, and is the LSR right right place for it?
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-1} There is no requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102; LSE-29).
+Should there be?
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-2} While the requirement that LSST process non-standard visit exposures (LSR-REQ-0111, LSE-29) implies that DM must be able to process the full diversity of Special Programs data, it should probably be verified that this is (a) true and (b) possible.
+For example, will DM impose limitations on the exposure times for which the Instrument Signature Removal (ISR) processing can deliver a science-grade quality image, as considered in Section \ref{ssec:data_DMconstraints}?
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as ``Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-3} It might be helpful to include in this section that Special Programs processing with reconfigured Level 1 and 2 codes, wherever possible, is also an effective requirement similar to LSR-REQ-0032 (LSE-29).
+(This might satisfy \ref{LSR-0}).
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products ``of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-4} The requirements for LSST to provide software, services, and hardware resources to support Level 3 data products (LSR-REQ-0041, LSE-39) all have technical limits (i.e., DM's intent to create and run reconfigured pipelines for Special Programs data ``unless it requires an intensive amount of additional computational resources" as discussed in Section \ref{ssec:dmplans_L3}); should these be mentioned in the LSR?
+(Or are they just in the sizing documents, discussed in Section \ref{sssec:dmplans_review_sizing}?)  \end{enumerate}
+
+$\bullet$ LSR-REQ-0042 requires that LSST produce the data products necessary to support the four primary science missions (dark energy, solar system, transients, milky way; 2.4.5.2, Science Flowdown).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-5} The requirement for LSST to produce the data products necessary for the four primary science missions (LSR-REQ-0042, LSE-29) could be extended with text to include Special Programs data?
+This is might also satisfy \ref{LSR-0}.
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
+This is discussed further in Section \ref{sssec:dmplans_review_sizing}.
+
+$\bullet$ LSR-REQ-0075 defines that the WFD main survey science objectives will be met with 90\% of the observing time, with the remainder left to Special Programs (Section 3.3.1, Survey Time Allocation)
+
+% % % % % % %
+\subsubsection{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\label{sssec:dmplans_review_oss}
+
+The OSS contains the requirements on the site, facility, and camera that are necessary to meet the requirements specified by the LSR (LSE-29) and the SRD (LPM-17).
+We have the version of LSE-30 last revised on February 10, 2017.
+For the most part, the {\bf Concerns} raised in this section are not relevant to Data Management and are probably outside the scope of this study.
+But we keep them because they are relevant to Section \ref{ssec:data_bounds}.
+
+$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 ``science proposals", where these ``proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (2.1.1.2, Multiple Science Programs).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-1} There is a requirement on the minimum number of ``science proposals" that can be optimized by the scheduler ({\tt nSciProp} = 6, OSS-REQ-0027, LSE-30), but is there also a technical constraint on the maximum?
+This might be relevant to some proposed mini-surveys with complicated observing strategies.
+\end{enumerate}
+
+$\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (2.1.1.7, Visit Optimization).
+
+$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively.
+These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
+These estimates appear adequate for Special Programs in general.
+
+$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 per night on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively.
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-2} There is a requirement on the minimum number of calibration frames that will be supported during daytime ({\tt nCalibExpDay} = 450, OSS-REQ-0194, LSE-30); could we imagine a Special Program that would need more than that?
+\end{enumerate}
+
+$\bullet$ OSS-REQ-0125, Section 3.1.5 deals with data products, including requirements for the variety and format of data products (i.e., images and catalogs in Level 1 and 2), the precision of the data products (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (alerts, Level 1, and the Level 2 DRP).
+(Section 3.1.5, Data Products).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-3} There are requirements on the type and quality of data products for Level 1, 2, and 3 (OSS-REQ-0125 and all those following, in LSE-30).
+Should it be specified, for Special Programs data, whether or which of these requirements DM must meet?
+For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be {\tt falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
+\end{enumerate}
+
+$\bullet$ Section 3.5, Photometric Calibration, puts requirements on e.g., the relative contributions to photometric errors from instrumental (OSS-REQ-0282) and atmospheric (OSS-REQ-0276) transmissions, and that a catalog of$10000$ reference stars of $17<r<20$ mag per field be created for use (OSS-REQ-0285).
+It seems that these might not apply to non-standard images from Special Programs (3.5, Photometric Calibration).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-4} Section 3.5 of the OSS (LSE-30) puts requirements on the photometric calibrations for WFD data, but it is unclear how and whether these requirements extend to DM providing calibration of non-standard images from Special Programs (this is related to \ref{LSR-2}).
+\end{enumerate}
+
+$\bullet$ OSS-REQ-0319 sets the requirement that the LSST be capable of continuous operation throughout the night when all visits are separated only by the readout time, thereby enabling a deep drilling field style of observations (3.6.1.3, Continuous Exposures).
+
+$\bullet$ OSS-REQ-0291 defines the minimum exposure time as $1$ second with a stretch goal of $0.1$ seconds, with a note that the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times (3.6.1.4, Minimum Exposure Time).
+\\
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-5} There is a requirement on the minimum exposure time (OSS-REQ-0291, LSE-30), but not on the spacing between short exposures and no specified maximum exposure time.
+It would be great to get this clarified from the camera team and updated in the OSS, so we know if they will they impose boundaries on Special Programs data.
+\end{enumerate}
+
+$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be {\tt tFilterChange} = 120 seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
+There is no officially required minimum time between filter changes.
+Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night\footnote{From Zeljko.}.
+Additionally, changes regarding the requirements on filters in support of ``deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
+(3.6.2 Filter Swaps \& Changes).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-6}   There are currently are no official requirements on the minimum time between filter changes, only that the time for a filter change will be 120 seconds (OSS-REQ-0293).
+Should the OSS include a clarification that there is no technical constraint on the frequency of filter changes?
+({\it MLG -- maybe unnecessary.}) \end{enumerate}
+
+$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking ({\tt rotTrackTime} = 1 hour) and the half-range of the rotator motion ({\tt rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-7} The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
+\end{enumerate}
+
+$\bullet$ OSS-REQ-0380 sets the rate and error limits for nonsidereal tracking, which presumably would only be relevant to a Special Program (3.6.3.7, Non-Sidereal Tracking).
+
+% % % % % % %
+\subsubsection{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}}\label{sssec:dmplans_review_dmsr}
+
+The DMSR contains the top-level requirements for all of Data Management, including processing pipelines and products (revision 2017-08-11).
+
+$\bullet$ DMS-REQ-0068 requires that each raw science image store metadata regarding the date/time, site, telescope, and camera -- missing from this is the scheduler and program information?
+That kind of metadata will be necessary to identify images related to Special Programs (1.2.3, Raw Science Image Metadata).
+Related to this is DMS-REQ-0266, which requires the creation of an Exposure Catalog that also stores this kind of metadata independently, but also does not include scheduler and program information.
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-1} Clarify if and how scheduler and program information (i.e., identification of an image as being associated with a Special Program) is included in the raw science image metadata and/or the Exposure Catalog (DMS-REQ-0068, -0266); if not, should it be added?
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
+It also says that ``this aspect of the processing for Special Programs data is specific to each program".
+It should be specified whether ``this aspect" refers to the reduction or archiving of the processed visit images; if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
+(1.2.2, Processed Visit Images).
+This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-2} Confirm whether processed single visits and/or difference images for Special Programs can stay on disk and accessible to e.g., Level 3 pipelines, for a longer amount of time by user request (DMS-REQ-0069 and -0010, LSE-61).
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0069 and DMS-REQ-0010 require that DM create processed single visit images and a difference image for each of those, but this might not be possible for Special Programs that obtain non-standard exposures (i.e., short exposures in which the PSF is not well formed, or very bright sky backgrounds).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-3} Confirm the exposure boundaries for which DM can produce processed single visits and/or difference images for Special Programs that obtain non-standard images (DMS-REQ-0069 and -0010, LSE-61).
+This is related to \ref{LSR-2} and might require a dedicated additional study with simulated data, as discussed in Section \ref{ssec:data_DMconstraints}.
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0274 sets the content of an Alert, but this does not currently include scheduler or program information; should this be added?
+This would enable users to e.g., use the LSST mini-broker to filter for only Alerts from their desired Special Program, in case they have a dedicated follow-up resource.
+Furthermore, it might facilitate follow-up more generally if users know e.g., that this Alert is from a field that is going to be observed again in $X$ minutes or $Y$ times in a night (1.3.13, Alert Content).
+However, the latter might be adequately available through the predicted visit schedule server specified by DMS-REQ-0353.
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-4} Could and/or should we add the program information and/or predicted revisit schedule to the Alert content (if it isn't included already)?
+(DMS-REQ-0274, -0353) \end{enumerate}
+
+$\bullet$ DMS-REQ-0320 states that ``it shall be possible for special programs to trigger their own data processing recipes", but this could use clarification.
+It could be misinterpreted to mean that the {\it science users} must trigger the processing, thereby classifying it as Level 3.
+More likely, it means that e.g., a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline.
+(1.6.1 Processing of Data From Special Programs)
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-5} Can we clarify the mechanism by which Special Programs trigger their own data processing recipes, as required by DMS-REQ-0320?
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing from Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
+It should probably be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that ``reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
+(1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6} Clarify that only Special Programs data that {\it can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products; see \ref{DMSR-3}), will be incorporated (relevant also to \ref{LSR-0b}, \ref{LSR-2}).
+Perhaps add that in these cases they will contribute directly to the L1 Alert Stream, with a note that generating a separate Level 3 stream of alerts specific to a Special Program should also be possible (DMS-REQ-0321, -0344).
+\end{enumerate}
+
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6b} Related to \ref{DMSR-6} and DMS-REQ-0344, there is also a potential problem with short exposures (e.g., 5 seconds) that can be processed by  the DIA pipeline, but pose a problem for the Alert Stream because the higher data acquisition rate (e.g., $\sim$6$\times$ higher) would cause a processing back-up.
+Should this situation arise, are there options for short-term increases in parallel processing power at NCSA?
+Such options might also apply to e.g., crowded fields with $>$10k {\tt DIASources}.
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0322 specifies that data products from Special Programs processing shall be stored in separate (but joinable) databases.
+
+Section 1.6 ``Special Programs" might benefit from the addition of discussion and/or requirements to address the following: \\
+$\bullet$ Should we add a general specification to the DMSR (\S~1.6) to describe whether and how data from Special Programs could be included in the WFD main survey Level 1 and/or 2 data products, as also raised for the LSR in \ref{LSR-0b}, and as discussed in Section \ref{ssec:dmplans_WFD}?
+\\
+$\bullet$ Should we add a specification to the DMSR (\S~1.6) to describe DM's plans to reconfigure its own pipelines for the processing of Special Programs data, perform and verify this processing, and deliver the unique products (separate but joinable to the WFD L1 and L2 products), as also raised for the LSR in \ref{LSR-0}, and as discussed in Section \ref{ssec:dmplans_reconfig}?
+\\
+$\bullet$ Since DM plans to include Special Programs data to the WFD main survey's L1 and L2 products wherever possible, and also run dedicated processing and generate unique products for all SP data, the associated computational and storage resources for Special Programs data might need to be multiplied by {\it at least} a factor of two.
+This is discussed more in Section \ref{sssec:dmplans_review_sizing}, but should it also be addressed in the DMSR?
+
+$\bullet$ The following items are applicable to users processing data from Special Programs and/or reprocessing data from the WFD main survey.
+Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
+Section 2.9 ``Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
+Sections 3.1 ``Software Architecture to Enable Community Re-Use" and 3.2 ``Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
+In Section 3.3 ``Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
+
+$\bullet$ DMS-REQ-0312 and -0313 require that DM maintain a live L1 and the two most recent DR L1 and L2 catalogs for access by users, but there is no corresponding access requirement for the data products of Special Programs (4.1, Data Archive).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-10} Should there be a requirement for maintenance of user access to data products of Special Programs with the same timescales as the WFD main survey data, similar to DMS-REQ-0312, -0313?
+(This is related to \ref{LSR-1}) \end{enumerate}
+
+% % % % % % %
+\subsubsection{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\label{sssec:dmplans_review_dmad}
+
+The DMAD specifies the design and implementation of the code and algorithms that will be used in the processing of images and creation of catalogs in the Level 1 AP, Level 2 DRP, and MOPS pipelines.
+We have used the version last revised 2017-07-19.
+As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
+In fact, the phrase ``Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
+It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
+This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
+
+The purpose of this study on Special Programs is not to suggest any changes to the DMAD.
+Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3: {\bf Deblending --} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
+It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
+Level 3 deblenders for specific Special Programs fields may require development by the user community.
+{\bf Variability Characterization --} The periodic and aperiodic variability characterizations described in Section 6.21 are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
+From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
+It is conceivable that the goals of a particular Special Program might require different algorithms, and these would be written as Level 3 and either made joinable to the DM reconfigured data products, or perhaps federated and incorporated directly.
+{\bf Photometric Redshifts --} As described in Section 5.6.5 the Level 2 DRP {\tt Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then federated and run at scale by DM.
+It is conceivable that the photo-$z$ algorithm for a Special Programs data set product might be different from that used for the WFD main survey.
+
+% % % % % % %
+\subsubsection{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\label{sssec:dmplans_review_dpdd}
+
+The DPDD describes the data products -- the images and catalogs -- to be delivered from the Level 1 AP, Level 2 DRP, and MOPS pipelines.
+We have used the version last revised 2017-07-01.
+Section 6 describes the data products that DM will provide for Special Programs.
+It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster (see Section \ref{sssec:dmplans_review_sizing} for sizing questions).
+
+The LSST Database Schema Browser\footnote{\url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}} is another way to explore the contents of the planned DM catalogs.
+We identify a couple of potential changes that might be necessary to the schema, which are related to this study on Special Programs.
+These are not high priority.
+
+$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used.
+This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
+DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} might need to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used).
+\end{enumerate}
+
+$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
+It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
+Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType.
+Indicates which input was used to produce a given object."} Is that all we need?
+\\
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an {\tt Object} is an externally provided coordinate.
+\end{enumerate}
+
+$\bullet$ As described in the DMAD and the DPDD, these are the {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization: \\
+{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{sssec:dmplans_review_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
+Are 32 floats in each of the 6 filters always going to be a large enough volume?
+({\it MLG -- I suspect yes and that if not, compression can be used, but am leaving this note here anyway}).
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} DM might want to consider whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+This might be similar to e.g., DMTN-049 for photometric redshifts included in the Level 2 DRP {\tt Object} catalog.
+\end{enumerate}
+
+% % % % % % %
+\subsubsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
+
+LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
+Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:data_DMconstraints} and \ref{LSR-2}.
+Regarding the reduction and calibration of non-standard visit imags, LSE-180 makes two relevant points: \\
+$\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
+$\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
+The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
+
+% % % % % % %
+\subsubsection{Computational Resources Sizing Documentation}\label{sssec:dmplans_review_sizing}
+
+LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
+DM Compute Sizing spreadsheet (LDM-138) and explanation document (LDM-140) estimates the total compute resources needed for all processes, quoting the same 10\% addition for Level 3.
+DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
+As mentioned in Section \ref{sssec:dmplans_review_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
+In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will ``use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
+The following might be specified in one of the sizing documents, but I am still confused:
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-1} Are there two {\it separate} blocks of ``10\% of the computing resources" allocated for DM processing of Special Programs data {\it and} end-user Level 3 analysis, or are these needs coming from the {\it same} 10\%?
+\end{enumerate}
+
+Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
+The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-2} Is ``10\% of the computing resources" allocated for DM processing of Special Programs data adequate, or too small, given that it is most likely to be multiply-reprocessed?
+\end{enumerate}
+
+Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will be done for Special Programs data, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
+Have these systems been sized?"}.
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-3} Confirm on which batch system the bulk of the intensive Level 3 processing -- for Special Programs or WFD main survey -- will take place.
+\end{enumerate}
+
+% % % % % % %
+\subsection{The Limited Documentation Regarding Image Detection Efficiencies}
+
+Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
+Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
+What, exactly, are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
+Note that here we are talking about planting fakes in the single images for transient detection efficiencies with difference imaging analysis, and not planting fakes in the CoAdds for point-source limiting magnitudes).
+
+$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
+will record information on the pixel variance ...
+as well as the per-pixel masks ...
+These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
+
+$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
+detection efficiency for point sources vs.
+mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
+
+$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
+The details of what metrics will be mapped, the format and scale of the maps (e.g.
+hierarchical pixelizations vs.
+polygons), and the way the metrics will be computed are all unknown"}.
+It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
+(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
+Source detection thresholds can be estimated through the use of sky sources"}.)
+
+$\bullet$ Suchyta et al.
+(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
+
+$\bullet$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
+
+$\bullet \bullet \bullet$ {\bf Summary.} {\it MLG - Perhaps I could take this on as a future study, and assess our options, their costs, and the scientific payoff.
+Since this is not a question limited to DM processing of Special Programs data, specifically, I end this side note here and no {\bf Concerns} are spawned.}
+
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\section{Special Programs Processing Case Studies}\label{sec:SPCS}
 
@@
 
-\subsubsection{A Mini-Survey Searching for TNOs with Shift-and-Stack}
+\subsection{A Mini-Survey Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_MM}
 
@@
 Delivery of LSST Processed Images.
-\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
 However, this Special Program appears to require a year of dispersed observations before the Level 3 processing would run.
@@
 
-\subsubsection{Searching for Supernovae in Deep Drilling Fields}
+\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
 
@@
 Delivery of LSST Processed Images.
-\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
 
@@
 
-\subsubsection{A Twilight Survey of Bright Stars}
+\subsection{A Twilight Survey of Bright Stars}\label{ssec:SPCS_Twilight}
 
@@
 Delivery of LSST Processed Images.
-\\ Pending a study of the ISR pipeline's ability to reduce exposures with a short integration time and a high sky background, the raw, reduced and calibrated exposures (and difference images, if created) are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmdocs_L3}).
+\\ Pending a study of the ISR pipeline's ability to reduce exposures with a short integration time and a high sky background, the raw, reduced and calibrated exposures (and difference images, if created) are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
 
@@
 
-\subsubsection{The Galactic Plane Survey for Variable Stars}
+\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
+
+Here we have to add the issue of crowded fields, differencing in crowded fields by the DM pipelines might not hit the needed quality for science?
+This would entail Level 3.
+This should spawn a Concern.
 
@@
 
-\subsubsection{A More Generic Level 3 Shift-and-Stack Case Study, by Mario}\label{sssec:science_dmsums_generic}
+\subsection{Including Saturated Stars in M67}\label{ssec:SPCS_M67}
+
+A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+Expand on this here.
+
+Want to spawn a concern here about using the PSF to measure flux of saturated stars.
+
+\subsection{A More Generic Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
 
@@
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\section{Conclusions}\label{sec:conc}
-
-Conclusions section.
-
-%\acknowledgments
-%Acknowledgments.
-%
-
 \bibliography{ms,lsst,refs,books,refs_ads}

### 2017-09-01 ms.tex
finished with case studies; added K-T feedback

@@
 
-\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys.
-We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey data products(Levels 1 and 2).
-It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.}
+\setDocAbstract{The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community.
+To do this, we begin with a review of the broad science goals of the previously proposed Special Programs and the potential diversity of data they might generate.
+We also consider any technical limitations on the data from the telescope, camera, and/or DM processing capabilities, and the methods by which information could be exchanged with the scientific community on these issues (i.e., the upcoming call for white paper proposals).
+We then move on to summarize the current DM plans for processing Special Programs data, and highlight specific requirements or text in the existing documentation that might need additions or clarifications to reflect these plans.
+We also develop a series of case studies for DM's processing of Special Programs data to better illuminate any complex or hidden issues; these case studies could also serve as examples for the upcoming call for community white papers.}
 
@@
 
-The intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
+The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
 
-\vspace{-20pt}
 \begin{quote}
@@
 \noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent.
-In particular, this study is designed to review: \\
-(1) the processing and products that are required to enable science for Special Programs, \\
-(2) the current plans for DM processing and products with respect to Special Programs, \\
-(3) whether the DM plans meet the expected needs of Special Programs, and \\
-(4) any necessary changes to be written into the requirements, designs, and plans.
+In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans.
 
-\noindent The concerns raised by this study fall into three main categories: \\
-(1) Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
-(2) Ensuring that DM's intent to process Special Programs data with reconfigured pipelines and provide separate data products that meet the science goals of each program is accurately reflected in all documents, as it is in the DPDD.
-\\
-(3) Clarifying the computational sizing plans for processing and storage of Special Programs data and whether it is this separate from the allocation for end-user Level 3 analysis.
+\subsection{Study Overview}\label{ssec:intro_overview}
+
+Section \ref{sec:data} compiles a comprehensive overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
+We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera, the scientific needs of as many currently proposed Special Programs as we can find, and the future mechanisms for LSST to gather more community input.
+We also consider whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
+
+Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data: reconfiguring pipelines to create unique products, incorporating data into the WFD main survey products, and enabling end-user Level 3 processing.
+Having established the {\it intent} of DM, we step through each of the relevant LSST documents and raise {\bf "Concerns"}, which highlight specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
+These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
+{\bf The Concerns raised by this study fall into two main categories:}
+\begin{enumerate}
+\item Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be.
+\item Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD.
+\end{enumerate}
+
+Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys.
+These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional {\bf Concerns} for the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
 
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\section{The Potential Diversity of Special Programs Data} \label{ssec:data}
+\clearpage
+\section{The Potential Diversity of Special Programs Data} \label{sec:data}
 
@@
 \noindent {\bf Resources:} \\
-$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
+$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
 $\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
 $\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
-$\bullet$ ``General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
-$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ ``LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
-$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
+$\bullet$ "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
+$\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
+$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
+$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
 
@@
 \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
+(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
 Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
@@
 $\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) 
+$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) \\ 
 $\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
 $\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
-Regarding SAS, \cite{BeckerWP} says that ``the multi-fit algorithm ...
+Regarding SAS, \cite{BeckerWP} says that "the multi-fit algorithm ...
 naturally provides a base infrastructure for this process.
@@
 The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
-See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_MM} for more detailed DM processing case studies.
+See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_TNO} for more detailed DM processing case studies.
 
@@
 \\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
 \\
@@
 The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars.
-See Sections \ref{ssec:SPCS_GPVSEx} and \ref{ssec:SPCS_M67} for more detailed DM processing case studies.
+See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.
 
@@
 SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the ``characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
+The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
 \\
@@
 Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
-In particular, the Level 2 DRP codes to create ``transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
+In particular, the Level 2 DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
 See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
@@
 $\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
-\cite{FergusonWP} mentions ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and ``characterization of low-surface-brightness extended features around both nearby and distant galaxies".
+\cite{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
 \\
@@
 $\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ {\it Jim Bosch -- ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
-$\bullet$ {\it Jim Bosch -- ``Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
+$\bullet$ {\it Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
+$\bullet$ {\it Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
 We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
@@
 \\
-$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an ``eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
+$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an "eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
 $\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).
@@
 (2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
-``Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
+"Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
 \\
@@
 
-It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines.
-Exposures which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines, such as those with different exposure time, dither pattern, or sky background.
+Alternatively, it might make more sense that DM's processing boundaries apply not to the exposure acquisition parameters, but the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding).
+The boundaries of this parameter space might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{sssec:dmplans_review_dmad}).
+
+Regarding exposure time, images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
 We consider shorter and longer exposure in turn.
@@
 However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for ``well-formed").
+As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
 Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
@@
 
-Our review of the technical boundaries and previously proposed Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, and exposures that are obtained with a bright sky background during twilight.
+Our review of the technical boundaries and currently tabled Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, and images of crowded fields.
 
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\clearpage
 \section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
@@
 Where relevant we suggest clarifications, changes, or additions to these documents.
-
-%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C1} There are currently no requirements on DM to incorporate data from Special Programs; should there be?
-\end{enumerate}
-%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C2} Should DM undertake an investigation to define the exposure boundaries (integration time, sky background) for non-standard visit images that can be processed by DM pipelines, such as ISR and DIA?
-\end{enumerate}
-%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C3} Given the double-processing of Special Programs data, is the system sized appropriately? \end{enumerate}
 
@@
 Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under ``Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
+As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
 
@@
 In most cases, data export and processing on external computation resources will not be necessary.
-The majority of users will access the data, build pipelines based on DM codes, and execute them through the Science Platform \cite{LDM-542}.
-Through the Science Platform, DM provides the infrastructure for users to run their Level 3 pipelines on LSST servers, monitor their progress, and analyze their output.
+The majority of users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their Level 3 pipelines on LSST servers, monitor their progress, and analyze their output.
 DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
@@
 This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
-The SRD makes two mentions of Special Programs: that the ``LSST is well suited to conducting Deep Supernova Survey" (Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (Section 3.4).
+The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4).
 The latter we know to be an approximate estimate.
@@
 
-$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
+$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
 \begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-2} While the requirement that LSST process non-standard visit exposures (LSR-REQ-0111, LSE-29) implies that DM must be able to process the full diversity of Special Programs data, it should probably be verified that this is (a) true and (b) possible.
@@
 
-$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as ``Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
+$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as "Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
 \begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-3} It might be helpful to include in this section that Special Programs processing with reconfigured Level 1 and 2 codes, wherever possible, is also an effective requirement similar to LSR-REQ-0032 (LSE-29).
@@
 
-$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products ``of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-4} The requirements for LSST to provide software, services, and hardware resources to support Level 3 data products (LSR-REQ-0041, LSE-39) all have technical limits (i.e., DM's intent to create and run reconfigured pipelines for Special Programs data ``unless it requires an intensive amount of additional computational resources" as discussed in Section \ref{ssec:dmplans_L3}); should these be mentioned in the LSR?
-(Or are they just in the sizing documents, discussed in Section \ref{sssec:dmplans_review_sizing}?)  \end{enumerate}
+$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products "of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-4} The requirements for LSST to provide software, services, and hardware resources to support Level 3 data products (LSR-REQ-0041, LSE-39) all have technical limits (i.e., DM's intent to create and run reconfigured pipelines for Special Programs data "unless it requires an intensive amount of additional computational resources" as discussed in Section \ref{ssec:dmplans_L3}); should these be mentioned in the LSR?
+\end{enumerate}
 
@@
 
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 ``science proposals", where these ``proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-1} There is a requirement on the minimum number of ``science proposals" that can be optimized by the scheduler ({\tt nSciProp} = 6, OSS-REQ-0027, LSE-30), but is there also a technical constraint on the maximum?
+$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (2.1.1.2, Multiple Science Programs).
+\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-1} There is a requirement on the minimum number of "science proposals" that can be optimized by the scheduler ({\tt nSciProp} = 6, OSS-REQ-0027, LSE-30), but is there also a technical constraint on the maximum?
 This might be relevant to some proposed mini-surveys with complicated observing strategies.
@@
 Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night\footnote{From Zeljko.}.
-Additionally, changes regarding the requirements on filters in support of ``deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
+Additionally, changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
 (3.6.2 Filter Swaps \& Changes).
@@
 $\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
-It also says that ``this aspect of the processing for Special Programs data is specific to each program".
-It should be specified whether ``this aspect" refers to the reduction or archiving of the processed visit images; if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
+It also says that "this aspect of the processing for Special Programs data is specific to each program".
+It should be specified whether "this aspect" refers to the reduction or archiving of the processed visit images; if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
 (1.2.2, Processed Visit Images).
@@
 
-$\bullet$ DMS-REQ-0320 states that ``it shall be possible for special programs to trigger their own data processing recipes", but this could use clarification.
+$\bullet$ DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes", but this could use clarification.
 It could be misinterpreted to mean that the {\it science users} must trigger the processing, thereby classifying it as Level 3.
@@
 $\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing from Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
-It should probably be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that ``reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
+It should probably be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
 (1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
@@
 
-Section 1.6 ``Special Programs" might benefit from the addition of discussion and/or requirements to address the following: \\
+Section 1.6 "Special Programs" might benefit from the addition of discussion and/or requirements to address the following: \\
 $\bullet$ Should we add a general specification to the DMSR (\S~1.6) to describe whether and how data from Special Programs could be included in the WFD main survey Level 1 and/or 2 data products, as also raised for the LSR in \ref{LSR-0b}, and as discussed in Section \ref{ssec:dmplans_WFD}?
@@
 \\
-$\bullet$ Since DM plans to include Special Programs data to the WFD main survey's L1 and L2 products wherever possible, and also run dedicated processing and generate unique products for all SP data, the associated computational and storage resources for Special Programs data might need to be multiplied by {\it at least} a factor of two.
-This is discussed more in Section \ref{sssec:dmplans_review_sizing}, but should it also be addressed in the DMSR?
+$\bullet$ Since DM plans to include Special Programs data to the WFD main survey's L1 and L2 products wherever possible, and also run dedicated processing and generate unique products for all SP data, the associated computational and storage resources for Special Programs data might need to be multiplied by a factor of two.
+This is discussed more in Section \ref{sssec:dmplans_review_sizing}, and for now this unknown seems to be well within the uncertainty of the sizing models, and so probably does not (at this time) need to be addressed in the DMSR.
 
@@
 Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
-Section 2.9 ``Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
-Sections 3.1 ``Software Architecture to Enable Community Re-Use" and 3.2 ``Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
-In Section 3.3 ``Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
+Section 2.9 "Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
+Sections 3.1 "Software Architecture to Enable Community Re-Use" and 3.2 "Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
+In Section 3.3 "Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
 
@@
 As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
-In fact, the phrase ``Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
+In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
 It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
@@
 The purpose of this study on Special Programs is not to suggest any changes to the DMAD.
-Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3: {\bf Deblending --} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
+Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3.
+{\bf Deblending --} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
 It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
@@
 Section 6 describes the data products that DM will provide for Special Programs.
-It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster (see Section \ref{sssec:dmplans_review_sizing} for sizing questions).
+It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster (see Section \ref{sssec:dmplans_review_sizing}).
 
@@
 It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType.
+Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
 Indicates which input was used to produce a given object."} Is that all we need?
@@
 LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
-Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:data_DMconstraints} and \ref{LSR-2}.
+Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}.
 Regarding the reduction and calibration of non-standard visit imags, LSE-180 makes two relevant points: \\
@@
 
+In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will "use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
+This is based on the estimate that Special Programs data comprise $\sim10\%$ of the total observing time.
+Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
+Based on this predicted multiple reprocessing, we might ask whether the "10\% of the computing resources" allocated for DM processing of Special Programs data is too small, but it is still a bit to early to know this.
+It will depend on the exact final mix of Special Programs, which may well change from year to year, and furthermore the current error bars on the sizing are of order $\pm10\%$.
+
+The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
 LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
@@
 As mentioned in Section \ref{sssec:dmplans_review_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will ``use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
-The following might be specified in one of the sizing documents, but I am still confused:
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-1} Are there two {\it separate} blocks of ``10\% of the computing resources" allocated for DM processing of Special Programs data {\it and} end-user Level 3 analysis, or are these needs coming from the {\it same} 10\%?
-\end{enumerate}
+It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
 
-Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
-The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-2} Is ``10\% of the computing resources" allocated for DM processing of Special Programs data adequate, or too small, given that it is most likely to be multiply-reprocessed?
-\end{enumerate}
-
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will be done for Special Programs data, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
+Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
 Have these systems been sized?"}.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{CS-3} Confirm on which batch system the bulk of the intensive Level 3 processing -- for Special Programs or WFD main survey -- will take place.
-\end{enumerate}
+The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing "catch-up" L1, processing
+the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
+There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can "borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
 
@@
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\clearpage
 \section{Special Programs Processing Case Studies}\label{sec:SPCS}
@@
 This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
-Note that we are not including any analysis in these descriptions.
-Also, they aren't that fully realized yet, and could use some more thought and input.
+Note that we are not including any analysis in these descriptions, only processing and products.
+These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
 
-Generic steps currently include: \\
+Basic steps that we use to describe a processing case study: \\
 Step 1. Data Acquisition. \\
-Step 2. Inclusion in Level 1 AP. \\
+Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
 Step 3. Delivery of LSST Processed Images. \\
-Step 4. Level 3 Processing. \\
-Step 5. Level 3 Products. \\
-(Step 6. Inclusion in Level 2.) \\
+Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
+Step 5. Inclusion in Level 2 DRP Products for WFD. \\
+Step 6. Science User Level 3 Pipelines and Products. \\
 
-\subsection{A Mini-Survey Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_MM}
+% % % % % % %
+\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
 
 This Special Programs processing summary is based on Becker et al.
-(2011) white paper to find TNOs with shift-and stack \citep{BeckerWP}.
-{\it MLG: See Section \ref{sssec:science_dmsums_generic} for a more general processing case-study written by Mario, for a shift-and-stack Level 3 pipeline running regularly on a large amount of data.}
+(2011) white paper to find TNOs with shift-and stack (SAS) \citep{BeckerWP}.
+See Section \ref{ssec:SPCS_SAS} for a more general processing case-study of a shift-and-stack Level 3 pipeline running regularly on a large amount of data.
 
-Step 1.
-Data Acquisition.
-\\ The observational sequence is triggered.
+Step 1. Data Acquisition. \\ 
+The observational sequence is triggered.
 In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
@@
 
-Step 2. Inclusion in Level 1 AP. \\ Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
+Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
 
-Step 3.
-Delivery of LSST Processed Images.
-\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
-However, this Special Program appears to require a year of dispersed observations before the Level 3 processing would run.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing?
+Step 3. Delivery of LSST Processed Images. \\ 
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
+
+Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
+Shift-and-stack processing is beyond the scope of DM's algorithms.
+
+Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
+As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
+However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would {\it all} be incorporated.
+
+Step 6. Science User Level 3 Pipelines and Products. \\
+The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
+Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
+The Level 2 difference imaging routine will be used with the same template tract/patch for all.
+Custom, user-supplied algorithms will shift the exposures and create difference images, then Level 2 routines will stack and do source detection and characterization and generate an object database.
+Custom code will derive orbital parameters for the detections and add them to the database.
+
+% % % % % % %
+\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
+
+Step 1. Data Acquisition. \\ 
+On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
+
+Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
+Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+
+Step 3. Delivery of LSST Processed Images. \\ 
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
+
+Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
+The required data products for this science goal can be met by reconfiguring the DM pipelines.
+First, a template image for the field will be made using DM stacking algorithms.
+On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of {\tt DIAObject}, {\tt DIASource}, and {\tt Object} that are unique to this DDF.
+The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-1} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
+Or might there be issues due to train/test set mismatch?
+This is probably a question to be answered with testing, later, and incorporated into the DMAD?
+\end{enumerate}
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-2} Confirm that resources for packet distribution outside of the Alert Stream will be supported (related to \ref{DMSR-4}).
 \end{enumerate}
 
-Step 4.
-Level 3 Processing.
-\\ The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The Level 2 difference imaging routine will be used with the same template tract/patch for all.
-Custom code will shift the difference images, then Level 2 routines will stack and do source detection and characterization (including real/bogus?) and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to the database.
-The user is notified with processing status updates.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image?
+Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
+As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
+However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would {\it all} be incorporated.
+
+Step 6. Science User Level 3 Pipelines and Products. \\
+For the science goal of searching for supernovae in nightly stacked DDF images, no separate Level 3, user-driven software appears necessary.
+
+% % % % % % %
+\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
+
+Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
+The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
+
+Step 1. Data Acquisition. \\
+At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (i.e., 1 second).
+Location and exposure times are set by the sky brightness and desired saturation limits.
+
+Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
+Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds (see \ref{OSS-5} and \ref{DMSR-6b}), these data could be incorporated.
+
+Step 3. Delivery of LSST Processed Images. \\
+Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
+
+Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
+TBD, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
+
+Step 5. Inclusion in Level 2 DRP Products for WFD. \\
+These data would not contribute to the Level 2 data products created for the WFD survey.
+
+Step 6. Science User Level 3 Pipelines and Products. \\
+If short-exposure images cannot be processed with the existing DM algorithms, would the Level 3 user-built pipeline have to start from the raw data and create new, custom algorithms for e.g., ISR? 
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-3} Is it at all conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
 \end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed?
-I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables?
+A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+This would certainly be considered a Level 3 algorithm.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-4} Are any DM algorithms being developed to estimate the flux of saturated stars using e.g., the PSF wings?
+Could the brighter-fatter issue be used for some science advantage?
 \end{enumerate}
 
-Step 5.
-Level 3 Products.
-\\ The shifted-and-stacked images.
-Databases of sources detected by the shift-and-stack process with the desired schema, some of which will overlap with Level 1, 2, and/or MOPS to facilitate table joins, some of which will be generated by custom codes.
-
-Step 6.
-Inclusion in Level 2.
-\\ Given the extreme difference in final depth of these fields, it is unlikely that they will be incorporated into the Level 2.
-However, external Level 3 processing could certainly make deep CoAdds for these fields, join them to the Level 2 DRP catalog, and use them for other science.
-
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
-
-Step 1.
-Data Acquisition.
-\\ On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
-
-Step 2. Inclusion in Level 1 AP. \\ Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3.
-Delivery of LSST Processed Images.
-\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
-
-Step 4.
-Level 3 Processing.
-\\ The federated pipeline runs automatically upon delivery of the Level 1 processed images.
-If a deep template image does not already exist (this would be a Level 3 product), a Level 2 stacking routine will be used to make them.
-This Level 2 stacking routine is also used to make the nightly deep CoAdd in each filter, and then a Level 1 or 2 routine will generate the difference images and run source detection (and real/bogus?).
-These sources will be added to Level 3 equivalents of the {\tt DIAObject} and {\tt DIASource} databases.
-The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
-If so, its outputs might not be reliable because of train/test set mismatch; will the internal LSST routines for real/bogus training be available and adaptable to Level 3 users?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16a} Confirm that resources (packet distribution) for a Level 3 alert stream would be supported and with what latency, whether and how such packets would be identified as non-AP, and that they could be ingested by brokers in real time.
-\end{enumerate}
-
-Step 5.
-Level 3 Products.
-\\ A deep template for this field, the nightly stacked CoAdds and difference images, and analogs of the {\tt DIAObject} and {\tt DIASource} databases but built only for this specific Deep Drilling Field.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction?
-\end{enumerate}
-
-Step 6. Inclusion in Level 2. \\ At the discretion of DM.
-
-\subsection{A Twilight Survey of Bright Stars}\label{ssec:SPCS_Twilight}
-
-Step 1.
-Data Acquisition.
-\\ At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (1-5 seconds?).
-Location and exposure times are set by the sky brightness and desired saturation limits.
-Consider: this survey might obtain a larger fraction of high-airmass (i.e., anti-solar) images.
-
-Step 2.
-Inclusion in Level 1 AP.
-\\ Pending a study of short-exposure suitability for psf-matching and differencing with a deeper template, and pending a decision on whether the AP processing capabilities can be extended to accommodate a faster image-input rate, these data could be incorporated.
-
-Step 3.
-Delivery of LSST Processed Images.
-\\ Pending a study of the ISR pipeline's ability to reduce exposures with a short integration time and a high sky background, the raw, reduced and calibrated exposures (and difference images, if created) are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
-
-Step 4. Level 3 Processing. \\ User-generated, user-run scripts might include the following, assuming that the ISR pipeline will produce reduced exposures: \\
-(A.) If difference images are required for science and the Level 1 AP is not run, then custom template images and difference images, and catalogs of difference-image sources, could be created using Level 1 or 2 codes.
-If the PSF is not well-enough formed, or there is a problem with DCR corrections (e.g., if obtained at high airmass), them some more sophisticated user-written codes will be necessary.
-\\
-(A2.) If the field includes a nearby supernova, special SN-free template images will be created using Level 1 or 2 image-stacking codes, difference images created with Level 1 DIA codes, and then photometric calibration done using the results of Level 3 processing.\\
-(B.) A script of Level 2 codes to extract and characterize sources in individual exposures, stack exposures into CoAdds (probably not optimized for depth, but maybe for precision), and extract and characterize sources in the CoAdds will be built and run by users.
-A specialized deblender for sources that span a large dynamic range might be necessary?
-\\
-(C.) User-written analysis codes for photometric calibration.
-
-Step 5. Level 3 Products. \\ 
-(A.) Special template images, difference images, and difference-source catalogs. \\
-(B.) Special CoAdds and source catalogs. \\
-(C.) Catalogs of bright objects in the LSST filter set.
-
-Step 6. Inclusion in Level 2. \\ Probably not, given the same regions would be surveyed with WFD exposures, these short exposures probably wouldn't help the DRP's objectives.
-
+% % % % % % %
 \subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
 
-Here we have to add the issue of crowded fields, differencing in crowded fields by the DM pipelines might not hit the needed quality for science?
-This would entail Level 3.
-This should spawn a Concern.
+Step 1. Data Acquisition. \\
+The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields, to a shallower depth than the WFD main survey.
 
-Step 1. Data Acquisition. \\ The schedule incorporates fields in the Galactic Plane, and executes e.g., 2x15 second exposures.
+Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
+Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
+Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:data_DMconstraints} and \ref{sssec:dmplans_review_dmad}).
 
-Step 2. Inclusion in Level 1 AP. \\ These images are incorporated into the Level 1 pipeline. Alerts are produced and DIAObjects is updated.
+Step 3. Delivery of LSST Processed Images. \\
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
 
-Step 3.
-Delivery of LSST Processed Images.
-\\ The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt \textbf{L1PublicT}} (currently 24 hours, but see Section \ref{ssec:dmplans_reconfig}).
+Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
+The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Level 1 and 2 pipelines, and not much reconfiguration would be needed.
+The biggest difference might be the federation of a user-supplied deblender algorithm optimized for very crowded fields.
 
-Step 4.
-Level 3 Processing.
-\\ User-created, user-run pipelines that incorporate a mix of Level 1 and 2 DM-provided codes and e.g., specialized deblenders, CoAdds, and photometric calibration routines targeted towards specific stellar populations and/or Galactic latitudes, depending on science goals.
+Step 5. Inclusion in Level 2 DRP Products for WFD. \\
+It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
+({\it MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
 
-Step 5. Level 3 Products. \\ Possibly a wide variety of difference and CoAdd images and their associated source catalogs.
+Step 6. Science User Level 3 Pipelines and Products. \\
+It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
 
-Step 6.
-Inclusion in Level 2.
-\\ These images are incorporated into the Level 2 DRP to the extent that they assist with the all-sky completeness and photometric calibration; TBD, at the discretion of DM.
-
-\subsection{Including Saturated Stars in M67}\label{ssec:SPCS_M67}
-
-A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-Expand on this here.
-
-Want to spawn a concern here about using the PSF to measure flux of saturated stars.
-
-\subsection{A More Generic Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
+% % % % % % %
+\subsection{A More General Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
 
@@
 have these systems been sized?].
-(\textbf{See \ref{C4}})
 
@@
 In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope).
-\end{enumerate}
 
@@
 We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
-(\textbf{See \ref{C18}, above.})
 
@@
 [This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
-(\textbf{See \ref{C15}}).
 
@@
 But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have \#3, you get \#6 for free as well.]
- (\textbf{See \ref{C2}}).
 
+\clearpage
 \bibliography{ms,lsst,refs,books,refs_ads}

### 2017-11-10 ms.tex
major updates to Action Items

@@
 
-\author{M.~L.~Graham, M.~Juri\'{c}, and E.~Bellm}
+\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, and E.~Bellm}
 
@@
 
-\setDocAbstract{The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community.
-To do this, we begin with a review of the broad science goals of the previously proposed Special Programs and the potential diversity of data they might generate.
-We also consider any technical limitations on the data from the telescope, camera, and/or DM processing capabilities, and the methods by which information could be exchanged with the scientific community on these issues (i.e., the upcoming call for white paper proposals).
-We then move on to summarize the current DM plans for processing Special Programs data, and highlight specific requirements or text in the existing documentation that might need additions or clarifications to reflect these plans.
-We also develop a series of case studies for DM's processing of Special Programs data to better illuminate any complex or hidden issues; these case studies could also serve as examples for the upcoming call for community white papers.}
+\setDocAbstract{{\bf This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget.
+Unresolved issues spawn {\bf Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}
 
 \setDocChangeRecord{%
-\addtohist{1}{2017-04-??}{Internal working document.}{Melissa Graham}
+\addtohist{1}{2017-11-??}{Internal working document.}{Melissa Graham}
 %\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
@@
 
-The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
+The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):
 
@@
 
-\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent.
-In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans.
+\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
+This study spawns {\bf Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.
 
-\subsection{Study Overview}\label{ssec:intro_overview}
+%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an {\bf Action} is spawned.
 
-Section \ref{sec:data} compiles a comprehensive overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
-We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera, the scientific needs of as many currently proposed Special Programs as we can find, and the future mechanisms for LSST to gather more community input.
-We also consider whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
+Section \ref{sec:data} compiles an overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
+We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera.
 
-Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data: reconfiguring pipelines to create unique products, incorporating data into the WFD main survey products, and enabling end-user Level 3 processing.
-Having established the {\it intent} of DM, we step through each of the relevant LSST documents and raise {\bf "Concerns"}, which highlight specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
-These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
-{\bf The Concerns raised by this study fall into two main categories:}
-\begin{enumerate}
-\item Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be.
-\item Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD.
-\end{enumerate}
+Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data, including the capabilities for processing non-standard visits and whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
+We also describe DM's intent for incorporating Special Programs data into the WFD main survey products, reconfiguring pipelines to create unique products, and providing support for end-user Level 3 processing.
+
+Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
+%These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
+{\bf The Concerns raised by this study fall into two main categories:} \\
+%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
+%2.
+Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget.
+\\
+
+Section \ref{sec:SP_SC} gathers the existing efforts from the Science Community regarding Special Programs, summarizes the scientific needs of as many currently proposed Special Programs as we can find, and outlines the future mechanisms for LSST to gather more community input -- including some recommendations for the future call for Special Programs white paper proposals.
 
 Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys.
-These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional {\bf Concerns} for the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
+These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional issues regarding the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
 
@@
 To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data.
-The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (standard visit images), then it might not be adequately handled by the DM system.
-To get a handle on the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes (Section \ref{ssec:data_bounds}).
-Next, we explore the current set of proposed science programs for LSST (Section \ref{ssec:data_science}) and the types of observations they would want to do, and then describe when and how we can incorporate revisions to these proposals from the user community in the future (Section \ref{ssec:data_comm}).
-Finally, we explore whether DM can or should set down any technical constraints on how far an observing program can deviate from a sequence of standard visit images and still be processed by DM pipelines, be they the reconfigured and/or the standard Level 1 and 2 pipelines (Section \ref{ssec:data_DMconstraints}).
+The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (i.e., standard visit images), then it might not be adequately handled by the DM system.
+To get an idea of the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes, in Section \ref{ssec:data_bounds}.
+
+We can also get an idea of the expected diversity in exposure type and survey pattern from the current set of proposed Special Programs for LSST, which are reviewed in more detail in Section \ref{sec:SP_SC}.
+We find that most of the proposed Special Programs are likely to use standard visits, two $\sim15$ second exposures, and that the potential diversity of Special Programs data is mainly comprised of exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, long series of exposures obtained of the same field (i.e., deep drilling), and images of crowded fields.
+
+All aspects related to DM's {\it processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.
 
 % % % % % % % % % % % % % % % % % % 
-\subsection{Observing Boundaries for Special Programs}\label{ssec:data_bounds}
+\subsection{Technical Boundaries for Special Programs Data}\label{ssec:data_bounds}
 
 We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals.
+For the purposes of this document we consider the ``observing boundaries" as those imposed by the camera, telescope, or site, and independent of any DM processing capabilities.
+It is not the job of this document to define these observing boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
+
+The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future.
+Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.
+
+Filter Changes: \\
+$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
+\\
+$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
+$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
+$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
+However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
+\\
+$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
+$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights. \\
+
+Exposure Times: \\
+$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
+$\bullet$ The maximum exposure time is not restricted. \\
+$\bullet$ However, for exposure times there are other considerations.
+The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
+Changing the exposure time also affects the photometric and astrometric calibrations.
+Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
+A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
+Additionally, the impact on CR rejection routines is untested for long exposures.
+
+Camera Rotation: \\
+$\bullet$ The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
+\textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
+Can't remember where I heard it.}
+
+Telescope Slews: \\
+$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
+
+Total Time: \\
+$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey).
+This could be increased a bit, but otherwise imposes an upper limit on the total time request of a given Special Program.
+
+Operations Simulations: \\
+$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{Bounds-1} The purpose of this ticket is to request clarification regarding several aspects of the telescope and camera capabilities that might impose boundaries on the diversity of data from Special Programs.
+First, have the following proposed filter change tolerances been confirmed: (1) the minimum time between filter changes has no restrictions from thermal tolerances, but it is recommended to keep the filter change rate lower than once every 10 minutes in order to minimize overheads; (2) the maximum total number of filter changes is 100000 over 15 years (18/night) and the maximum number of filter swaps in and out of the carousel is 3000 in 15 years (once every two nights).
+Second, are there any restrictions on the use of the camera rotator or the total amount of slew time?
+For example, could a Special Program continuously jump between well-separated fields on the sky for an extended amount of time?
+\end{enumerate}
+
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\clearpage
+\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
+
+As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (DPDD, LSE-163, \cite{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).
+
+In this section we review in greater detail the planned ability of DM to process non-standard visit images (Section \ref{ssec:dmplans_NSV}), the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:dmplans_NSV}
+
+A standard visit image is considered to be two snap exposures of $\sim15$ second each, and in the WFD main survey sequential visits will be of different fields.
+As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field.
+In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).
+
+\subsubsection{Exposure Times}
+
+Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+We consider shorter and longer exposure in turn.
+
+{\bf Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
+The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
+Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
+A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
+However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
+As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
+Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
+\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
+At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
+We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
+\label{fig:expt}}
+\end{center}
+\end{figure}
+
+{\bf Longer Exposures.} There is no maximum exposure time specified for an LSST image.
+Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
+However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
+With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+
+In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
+One possibly useful study is Chang et al.
+(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
+They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
+They present a new software package to do mitigate the effects.
+Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
+2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
+
+In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
+
+Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
+This is probably best left ot a policy issue (beyond the scope of this document).
+
+\subsubsection{Image Acquisition Rate}
+
+OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds.
+However, {\it if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated {\tt DIASources}.
+
+{\bf Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds.
+This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later?
+{\bf How long does it take to save an image to disk?}
+
+{\bf Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
+(Such options might be needed anyway to process crowded fields with $>$10k {\tt DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
+However, at NCSA there should be ways to elastically change the amount of processing power available.
+This is also an issue for crowded fields (below).
+
+{\bf Unassociated {\tt DIASources}:} the consistency of {\tt DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field).
+In other words, it is not guaranteed that the next string of processing would have access to the latest {\tt DIAObjects} for association.
+It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations.
+This is also an issue for crowded fields (below).
+
+\subsubsection{Exposures per Visit, and Long Sequences of a Single Field}
+
+K.-T.
+Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
+This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
+Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
+things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
+The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
+To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
+Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
+{\it MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
+
+K.-T.
+has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
+One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
+The processing pipeline would need to know how to divide this sequence up into visits.
+As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
+
+\subsubsection{Alerts from Crowded Fields}
+
+Special Programs are more likely to include crowded fields than the WFD main survey area.
+Due to the increased number of sources, the number of {\tt DIASources} -- and therefore the number of Alerts -- increases as well.
+In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
+A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete.
+K.-T.
+reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed.
+This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).
+
+\subsubsection{Summary}
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey).
+For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science.
+After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018.
+(1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA?
+In the latter case, this will probably be limited by the time needed to generate a well-formed PSF.
+(2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds.
+Might this requirement be slackened to accommodate a higher cadence Special Program?
+If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping {\it vs.} delaying Alerts during observations with a fast ingest cadence.
+These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts.
+Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF).
+We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits.
+Consequential images of the same field may lead to unassociated {\tt DIASources} of the same {\tt DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately.
+\end{enumerate}
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
+
+DM will want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
+In the following three sections we summarize when and how Special Programs data might be incorporated into the pipelines and products of the Difference Imaging Analysis and Alert Stream, the Data Release Pipeline products, and/or MOPS.
+
+\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
+
+It is beneficial to transient science to include as many LSST images into the Level 1 Difference Imaging Analysis and Alert Stream as possible.
+Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST (this prospect is discussed further in Section \ref{ssec:dmplans_L3}).
+Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
+DM's ability to process the potential diversity of image types from Special Programs is explored above in Section \ref{ssec:dmplans_NSV}, and we continue the discussion of incorporating processed images into the Level 1 DIA pipeline.
+
+{\bf Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
+So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine.
+When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale).
+K.-T.
+confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible.
+This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions.
+K.-T.
+also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.
+
+\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
+
+The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries).
+Since MOPS takes {\tt DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
+As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
+There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
+
+\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
+
+Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
+We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
+When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
+For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
+In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey.
+This is mostly discussed here in context with the {\tt Source} and {\tt Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
+
+LSST DM intends to reconfigure its existing pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
+In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
+
+In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
+For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
+This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
+
+The current outstanding issue related to DM's processing of special programs data is that the work-hours needed to reconfigure and test the pipelines, run them and verify the data products for public release -- potentially on intermediate timescales that do not coincide with the prompt/yearly Level 1 and 2 timescales, e.g., monthly stacks of deep drilling fields -- have not been included in the personnel budget and thus, at the moment, constitute an expansion of scope.
+As we describe in Section \ref{ssec:docrev_dmsr} there are requirements in the DMSR that state that DM does have the responsibility to process Special Programs data.
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to investigate the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations.
+Questions to be answered include, e.g., product ownership of the reconfigured pipelines and test data products, and how do we set goals and deadlines for this?
+\end{enumerate}
+
+As described in Section \ref{ssec:docrev_dmsr}, the DMSR contains requirement DMS-REQ-0320, which states that "it shall be possible for special programs to trigger their own data processing recipes".
+Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.
+
+Also described in Section \ref{ssec:docrev_dmsr} is the DMSR requirement that ``it shall be possible for special programs to trigger their own data processing recipes" (DMS-REQ-0320).
+However, there is currently no requirement for the scheduler to initiate non-standard processing with a certain pipeline if the data was executed under a given program.
+Since the data processing is scheduler-driven, this would have to be an OSS requirement.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-2} The purpose of this ticket is to figure out whether and how auto-triggering of alternative data processing recipes for Special Programs can be supported by the DM system.
+If it can't be, a change request on DMS-REQ-0320 might be needed to down-scope.
+\end{enumerate}
+
+% % % % % % % % % % % % % % % % % % 
+\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_L3}
+
+As explained above, LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs whenever possible.
+In Section \ref{ssec:docrev_sizing} we describe how LSST has allocated 10\% of its computational and storage capacity for this processing because 10\% of the images will be generated by Special Programs.
+
+However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3).
+LSST has furthermore allocated an {\it additional} 10\% of its computing resources for user-driven processing and analysis.
+This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).
+
+In the case where a Special Program requires an intensive amount of additional computational resources (i.e., too large a fraction of the 10\% allocated), the processing will have to be done on a separate, external computational resource.
+To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
+
+In the case where user-driven processing pipelines consume a small amount of that 10\%, data export and processing on external computation resources will not be necessary.
+Users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their pipelines on LSST servers, monitor their progress, and analyze their output.
+DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
+
+However, in the case where user-driven processing pipelines consume a {\bf large} amount of that 10\%, data export and processing on external computation resources {\bf will} be necessary.
+The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.
+
+{\bf Level 3 Alert Stream?
+--} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package.
+K.-T.
+raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.
+
+{\bf Lower Latency for Data Access?
+--} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner.
+Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-1} The purpose of this ticket is to ask Gregory about whether there are aspects to the Science Platform that need to be developed to better serve the needs of the community doing Special programs processing.
+Examples include support for Level 3 alert streams and a lower latency for access to processed visit images.
+\end{enumerate}
+
+It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
+For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
+An alternative option to ``adopted" code is ``adopted" data products: situations in which Level 3 code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-2} The purpose of this ticket is to clarify whether there is a process in place, and if staffing needs have been defined, for adopting Level 3 code into DM's pipelines, and/or for adopting data products, verifying them, and hosting their public release.
+This issue is not limited to Special Programs, but is likely to be encountered in that context.
+\end{enumerate}
+
+% % % % % % % % % % % % % % % % % %
+\clearpage
+\section{DM Documentation Review}\label{sec:docrev}
+
+Here we gather every requirement or specification relevant to Special Programs in the existing DM documentation, identify areas where the intent of LSST DM is not adequately represented, and suggest clarifications, changes, or additions to these documents.
+
+% % % % % % %
+\subsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{ssec:docrev_srd}
+
+This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
+The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4).
+The latter we know to be an approximate estimate.
+
+% % % % % % %
+\subsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{ssec:docrev_lsr}
+
+The LSR is derived from the SRD, and converts the high-level specifications from the SRD into system requirements, which flow-down to and in some cases repeated in the OSS and DMSR (next sections).
+We have the version of LSE-29 last revised on August 4, 2016.
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-0} RFC (LSE-29): We propose to add a high-level requirement that DM reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, to better reflect its intent.
+This would flow-down to the DMSR.
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102).
+E.g., that data products from reconfigured pipelines will be made available on a intermediate time scale where the science goals require it.
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
+The full capabilities of DM to process non-standard visits is not yet well defined; see \ref{NSV-1}.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-2} RFC (LSE-29): We propose that LSR-REQ-0111, the requirement for DM to be capable of processing exposures that were not taken in a standard visit mode, might need to include a caveat once the DM capabilities for processing non-standard visits are truly defined (e.g., extremely short exposures).
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as "Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-3} RFC (LSE-29): We propose to add to LSR-REQ-0032, which requires that the DM system provide three classes science data products (nightly, yearly, and user-specified), the requirement to provide intermediate-cadence data products for Special Programs, whenever possible, to enable science.
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products "of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-4} RFC (LSE-29): We propose to add to LSR-REQ-0041, which requires that LSST provide software, services, and hardware resources to support Level 3 data products, the caveat that there will be technical limits on DM's ability to provide this in cases where an an intensive amount of additional computational resources is required.
+\end{enumerate}
+
+$\bullet$ LSR-REQ-0042 requires that LSST produce the data products necessary to support the four primary science missions (dark energy, solar system, transients, milky way; 2.4.5.2, Science Flowdown).
+Since the science goals of the Special Programs flow down from these four primary science goals, no change is needed here.
+
+$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
+This is discussed further in Section \ref{ssec:docrev_sizing}.
+
+$\bullet$ LSR-REQ-0075 defines that the WFD main survey science objectives will be met with 90\% of the observing time, with the remainder left to Special Programs (Section 3.3.1, Survey Time Allocation)
+
+% % % % % % %
+\subsection{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\label{ssec:docrev_oss}
+
+The OSS contains the requirements on the site, facility, and camera that are necessary to meet the requirements specified by the LSR (LSE-29) and the SRD (LPM-17).
+We have the version of LSE-30 last revised on February 10, 2017.
+Some of the aspects discussed in this section are not directly relevant to Data Management and are probably outside the scope of this study, but we keep them because they are relevant to Section \ref{ssec:data_bounds}.
+
+$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, {\tt nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies.
+\end{enumerate} % Assign to Zeljko.
+
+$\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (OSS 2.1.1.7, Visit Optimization).
+
+$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively.
+These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
+These estimates appear adequate for Special Programs in general.
+
+$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively.
+These are {\it minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.
+
+$\bullet$ Section 3.1.5 (OSS-REQ-0125 and those following) set the requirements for data products, including their variety and format (i.e., images and catalogs in Level 1 and 2), their precision (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (Alerts, Level 1, and Level 2 ).
+Would these also apply to data from Special Programs?
+For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be {\tt falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-3} RFC (LSE-30): We propose to add a statement to Section 3.1.5 of the OSS that DM will make a ``best effort" to meet the data products requirements with the Special Programs data.
+\end{enumerate}
+
+$\bullet$ Section 3.5, Photometric Calibration, puts requirements on e.g., the relative contributions to photometric errors from instrumental (OSS-REQ-0282) and atmospheric (OSS-REQ-0276) transmissions, and requires that a catalog of $10000$ reference stars of $17<r<20$ mag per field be created for use (OSS-REQ-0285).
+The latter might not apply to Special Programs fields, but this probably isn't cause for concern.
+
+$\bullet$ OSS-REQ-0319 sets the requirement that the LSST be capable of continuous operation throughout the night when all visits are separated only by the readout time, thereby enabling a deep drilling field style of observations (OSS 3.6.1.3, Continuous Exposures).
+
+$\bullet$ OSS-REQ-0291 defines the minimum exposure time as $1$ second with a stretch goal of $0.1$ seconds, with a note that the spacing between exposures should be lengthened to $15$ seconds in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times (OSS 3.6.1.4, Minimum Exposure Time).
+The implication for DM is that an exposure ingest rate will never exceed one per $15$ seconds.
+
+$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be ${\tt tFilterChange} = 120$ seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
+There is no officially required minimum time between filter changes.
+Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night (see Section \ref{ssec:data_bounds}).
+
+$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking ({\tt rotTrackTime} = 1 hour) and the half-range of the rotator motion ({\tt rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
+Questions about whether there is a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), are raised in Section \ref{ssec:data_bounds}.
+
+$\bullet$ OSS-REQ-0380 sets the rate and error limits for nonsidereal tracking, which presumably would only be relevant to a Special Program (3.6.3.7, Non-Sidereal Tracking).
+
+% % % % % % %
+\subsection{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}}\label{ssec:docrev_dmsr}
+
+The DMSR contains the top-level requirements for all of Data Management, including processing pipelines and products (revision 2017-08-11).
+
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-0} RFC: We propose that the DMSR include a high-level specification (probably in Section 1.6) that the data from Special Programs will be included in the products of the wide-fast-deep main survey only when it is (a) possible and (b) beneficial to the primary science objectives of those products.
+It should also be updated to reflect the fact that DM will reconfigure its own pipelines ``wherever possible", process the Special Programs data, verify the unique data products, and make them publicly available.
+This would be a flow-down requirement from a similar new one added to the LSR (LSE-29).
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0068 requires that each raw science image store metadata regarding the date, time, site, telescope, and camera -- missing from this might be the scheduler and program information.
+That kind of metadata will be necessary to identify images related to Special Programs (1.2.3, Raw Science Image Metadata).
+Related to this is DMS-REQ-0266, which requires the creation of an Exposure Catalog that also stores this kind of metadata independently, but also does not include scheduler and program information.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-1} RFC (LSE-61): We propose that DMS-REQ-0068 be updated to include that the scheduler and program metadata (i.e., a way to identify an image as being associated with a Special Program) with the raw science image, and/or with the Exposure Catalog (DMS-REQ-0068, -0266).
+We furthermore propose that the {\tt visitID} and the total number of exposures in the visit are also included in this updated requirement (e.g., to accommodate a deep drilling sequence of a single visit of 50 exposures).
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
+It also says that "this aspect of the processing for Special Programs data is specific to each program".
+It should be specified whether "this aspect" refers to the reduction or archiving of the processed visit images.
+If it refers to archiving, and if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
+(1.2.2, Processed Visit Images).
+This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
+We assume this applies to standard visit images that {\it can} be processed.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-2} RFC (LSE-61): We propose that DMS-REQ-0069 be clarified regarding whether processed single visits and/or difference images for Special Programs can stay archived and accessible to e.g., Level 3 pipelines, for a longer amount of time, possibly by user request.
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0274 sets the content of an Alert, and this does not currently include scheduler or program information.
+Adding this would enable users to e.g., use the LSST mini-broker to filter for only Alerts from their desired Special Program, in case they have a dedicated follow-up resource.
+Furthermore, it might facilitate follow-up more generally if users know e.g., that this Alert is from a field that is going to be observed again in $X$ minutes or $Y$ times in a night (1.3.13, Alert Content).
+However, the latter might be adequately available through the predicted visit schedule server specified by DMS-REQ-0353.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-4} RFC (LSE-61): We propose that DMS-REQ-0274, which sets the content of an Alert, be updated to include the program information and/or predicted revisit schedule (unless it is adequately available through the predicted visit schedule server specified by DMS-REQ-0353).
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
+Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.
+This is handled by \ref{reconfig-2}.
+(1.6.1 Processing of Data From Special Programs).
+
+$\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
+It should be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
+(1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-6} RFC (LSE-61): We propose that DMS-REQ-0321 and -0344, which specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey, be reworded to clarify that only Special Programs data that {\it can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products), will be incorporated into Level 1 processing and contribute to the Alert Stream.
+\end{enumerate}
+
+$\bullet$ DMS-REQ-0322 specifies that data products from Special Programs processing shall be stored in separate (but joinable) databases.
+
+$\bullet$ DMS-REQ-0312 and -0313 require that DM maintain a live L1 and the two most recent DR L1 and L2 catalogs for access by users, but there is no corresponding access requirement for the data products of Special Programs (4.1, Data Archive).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-7} RFC (LSE-61): We propose that there be a requirement for maintenance of user access to data products of Special Programs with the same timescales as the WFD main survey data; i.e., similar to DMS-REQ-0312, -0313.
+\end{enumerate}
+
+%$\bullet$ The following items are applicable to users processing data from Special Programs and/or reprocessing data from the WFD main survey.
+Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
+Section 2.9 "Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
+Sections 3.1 "Software Architecture to Enable Community Re-Use" and 3.2 "Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
+In Section 3.3 "Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
+
+% % % % % % %
+\subsection{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\label{ssec:docrev_dmad}
+
+The DMAD specifies the design and implementation of the code and algorithms that will be used in the processing of images and creation of catalogs in the Level 1 AP, Level 2 DRP, and MOPS pipelines.
+We have used the version last revised 2017-07-19.
+As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
+In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
+It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
+This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
+
+The purpose of this work on Special Programs is not to suggest any changes to the DMAD.
+Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3.
+
+\subsubsection{Extreme-Depth CoAdds} The system has been sized to hold $\sim200$ exposures in memory at once, which defined by the current maximum number of visits per field in the WFD main survey in $10$ years (from a conversation with K.-T.~Lim).
+Note that the panchromatic CoAdds would be built from the individual filter CoAdds, so the algorithm does not need to handle $\sim800$ images.
+From a computational standpoint, $200$ is the maximum number of images that can be stacked with an algorithm that requires all images to be accessible in memory at once (i.e., loading all images and calculating the median for each pixel).
+Deeper stacks might be possible with algorithms that deal with images consequentially.
+It is conceivable that a Special Program which needs to stack $>200$ images is not possible to accomplish with reconfigured pipelines, and would have to be processed with external, user-contributed resources.
+However, the exact DM capabilities in this area are not yet well known because NCSA has not yet defined the machine capabilities.
+Furthermore, the planned commissioning data will go to a $\sim20$ year depth, and so it can reasonably be expected that DM will have to be able to accommodate at least a stack that deep.
+
+\subsubsection{Deblending} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
+It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
+Level 3 deblenders for specific Special Programs fields may require development by the user community.
+
+\subsubsection{Variability Characterization} The periodic and aperiodic variability characterizations described in Section 6.21 of the DMAD are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
+From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
+It is conceivable that the goals of a particular Special Program might require different algorithms; these could be provided by DM, or written as Level 3 and either made joinable to the DM reconfigured data products or perhaps incorporated directly.
+
+\subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP {\tt Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM.
+It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.
+
+% % % % % % %
+\subsection{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\label{ssec:docrev_dpdd}
+
+The DPDD describes the data products -- the images and catalogs -- to be delivered from the Level 1 AP, Level 2 DRP, and MOPS pipelines.
+We have used the version last revised 2017-07-01.
+Section 6 describes the data products that DM will provide for Special Programs.
+It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster.
+
+The LSST Database Schema Browser\footnote{\url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}} is another way to explore the contents of the planned DM catalogs.
+We identify a couple of potential changes that might be necessary to the schema, which are related to this study on Special Programs.
+These are not high priority.
+
+$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used.
+This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
+DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
+In conversation with K.-T., we find that this isn't a problem, as it will be handled by provenance: the code configuration used and the time of processing are sufficient to identify and regenerate the template image.
+However, K.-T.
+has pointed out that the capability to regenerate the {\it exact same} template -- the pixels that were subtracted -- is not a current deliverable.
+However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.
+
+$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
+It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
+Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
+Indicates which input was used to produce a given object."} Is that all we need?
+\\
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-1} The purpose of this ticket is to clarify whether {\tt Object.prv\_inputID} in the database schema will identify whether an {\tt Object} is an externally provided coordinate.
+\end{enumerate}
+
+$\bullet$ The {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
+{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
+Are 32 floats in each of the 6 filters always going to be a large enough volume?
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-2} The purpose of this ticket is to ask if DM should consider whether the {\tt float[6x32]} for the characterization parameters for variability is of adequate size to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+This study might be similar to e.g., DMTN-049 for photometric redshifts included in the DRP {\tt Object} catalog.
+\end{enumerate}
+
+% % % % % % %
+\subsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
+
+LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
+Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}.
+Regarding the reduction and calibration of non-standard visit images, LSE-180 makes two relevant points: \\
+$\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
+$\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
+The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
+
+However, Lupton's recent work on calibrations has made much of LSE-180 obsolete, and it is not clear whether it is still needed as a separate document.
+At the time of writing, Lupton's most recent take on calibrations can be found in \url{https://github.com/lsst-dm/calibration/blob/master/calibration.pdf}.
+K.-T.
+referenced the Appendix B for the overview of calibration types, but it seems that there is not yet a final plan that can be assessed for its suitability for Special Programs.
+It is plausible that Special Programs could have their own calibration database, and obtaining additional calibration frames is not expected to cause a bottleneck (at the moment there is $\sim1.7$ hours per day for this, and exceeding this could eat into the engineering time).
+The most likely cause for trouble involving calibrations and Special Programs is the computational time needed to create the additional and/or different calibration files to be applied to non-standard Special Programs data, and/or the additional overhead required to load them in the Level 1 pipeline if the schedule interleaves SP-WFD-WFD-SP.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Cal-1} The purpose of this ticket is to define the potential additional calibration needs of Special Programs data, which might be significantly more diverse in terms of, e.g., exposure time and/or sky background levels in the case of a a twilight survey of short exposures.
+Questions include how much additional day time might be needed to obtain raw calibration frames, or the additional computational resources needed to create and/or load calibration frames or catalogs during processing.
+The goal of this would be to confirm that the DM system prepared to calibrate the diversity of data that might be generated by Special Programs.
+\end{enumerate}
+
+% % % % % % %
+\subsection{Computational Resources Sizing Documentation}\label{ssec:docrev_sizing}
+
+In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will "use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
+This is based on the estimate that Special Programs data comprise $\sim10\%$ of the total observing time.
+Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
+Based on this predicted multiple reprocessing, we might ask whether the "10\% of the computing resources" allocated for DM processing of Special Programs data is too small.
+This will depend on the exact final mix of Special Programs, which may well change from year to year, and furthermore the current error bars on the sizing are of order $\pm10\%$.
+
+The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
+LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
+DM Compute Sizing spreadsheet (LDM-138) and explanation document (LDM-140) estimates the total compute resources needed for all processes, quoting the same 10\% addition for Level 3.
+DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
+As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
+It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that {\it an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
+\end{enumerate}
+
+Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
+Have these systems been sized?"}.
+The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
+There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can ``borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-2} The purpose of this ticket is to ensure that the overall DM computational system is appropriately sized, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, {\it and} incorporated into the Level 2 data products, {\it and} have their own dedicated reconfigured pipelines and separate products, {\it and} have a higher probability for user-defined processing.
+The latter would apply to the size of the DACs.
+Although an accurate estimate of the science-driven DAC size might not be truly possible until after the call for Special Programs white papers, an initial attempt must be made because this call will need to define a computational boundary for what can be done in the DAC, and what will require the proposer to bring additional resources to the table.
+For example, if the fractional computational needs are equivalent to the fraction of total data that the program acquires, that is acceptable; but if the program requires, e.g., $>2\times$ as much computational resources, then an external contribution might be necessary.
+Defining what the call will say regarding the need for external processing is a goal of this ticket.
+\end{enumerate}
+
+% % % % % % %
+\subsection{The Limited Documentation Regarding Image Detection Efficiencies}
+
+Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
+Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
+What, exactly, are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
+Note that here we are talking about planting fakes in the single images for transient detection efficiencies with difference imaging analysis, and not planting fakes in the CoAdds for point-source limiting magnitudes).
+
+$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
+will record information on the pixel variance ...
+as well as the per-pixel masks ...
+These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
+
+$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
+detection efficiency for point sources vs.
+mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
+
+$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
+The details of what metrics will be mapped, the format and scale of the maps (e.g.
+hierarchical pixelizations vs.
+polygons), and the way the metrics will be computed are all unknown"}.
+It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
+(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
+Source detection thresholds can be estimated through the use of sky sources"}.)
+
+$\bullet$ Suchyta et al.
+(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
+
+$\bullet$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
+
+$\bullet$ However, K.-T.
+revealed that a computational budget of 40\% has been allocated for fake insertion, with the understanding that this would happen in the daytime (i.e., not within 60 seconds).
+However, there has been a failure to schedule this into the development plan and there are no milestones for this project, and no person assigned to this task.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DE-1} The purpose of this ticket is to resolve whether or not the implantation of fake sources will be done as part of the Level 1 detection efficiency characterization, or if an alternate method will be sufficient for the science needs.
+K.-T.
+says the computational system is sized to accommodate it.
+If this issue remains unclear, perhaps a study that assess the options, their computational costs, and their scientific payoff should be done so that DM can better assess the need for fake sources and allocating staffing hours as necessary.
+Note that this is not a question limited to Special Programs.
+\end{enumerate}
+
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\clearpage
+\section{Special Programs and the Science Community}\label{sec:SP_SC}
+
 So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
@@
 
-There will be some technical limitations on the mini-surveys that will be possible.
-Here's a rough draft list of such potential issues.
-It is not the job of this document to define the camera/observatory technical boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
-This list will be updated in the future (or revised to point to where these boundaries are defined).
-In Section \ref{sssec:dmplans_review_oss}, as we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document we highlight a couple more questions regarding potential boundaries on Special Programs observations.
-
-Filter Changes (and see \ref{OSS-6}): \\
-$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
-\\
-$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
-$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
-However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
-\\
-$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
-$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights.
-
-Exposure Times (and see \ref{OSS-5}): \\
-$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
-$\bullet$ The maximum exposure time is not restricted. \\
-$\bullet$ However, for exposure times there are other considerations.
-The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
-Changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
-A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Additionally, the impact on CR rejection routines is untested for long exposures.
-
-Camera Rotation (and see \ref{OSS-7}): \\
-$\bullet$ \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
-Can't remember where I heard it.}
-
-Telescope Slews: \\
-$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
-
-Calibration Images (and see \ref{OSS-2}, \ref{OSS-4}): \\
-$\bullet$ The necessary afternoon calibrations for a Special Program must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained.
-
-Operations Simulations: \\
-$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
-
-Total Time: \\
-$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey), but this could be changed if significant science is enabled.
-\\
-
 % % % % % % % % % % % % % % % % % % 
@@
 In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed.
-We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science; this Section is mainly descriptive and we do not spawn any {\bf Concerns} here.
-In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs; any {\bf Concerns} that arise are detailed in that section.
+We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science.
+In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
 Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
@@
 Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
+(Note that this nominal survey is missing the overhead for readout and shutter, 3 seconds per exposure, which means that the LSST capability is closer to 43--45 exposures per hour, minus the time for 3--4 filter changes -- not 50 exposures per hour.)
 
@@
 % % % % % % % % % % % % % % % % % % 
-\subsection{Future Input from Data Rights Holders} \label{ssec:data_comm}
+\subsection{Future Input from the Science Community} \label{ssec:data_comm}
 
@@
 (4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:data_DMconstraints}
-
-It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines, such as those with different exposure time, dither pattern, or sky background.
-Alternatively, it might make more sense that DM's processing boundaries apply not to the exposure acquisition parameters, but the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding).
-The boundaries of this parameter space might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{sssec:dmplans_review_dmad}).
-
-Regarding exposure time, images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-We consider shorter and longer exposure in turn.
-
-(1) Shorter Exposures.
-The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint.
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
-\end{center}
-\end{figure}
-
-(2) Longer Exposures.
-There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
-One possibly useful study is Chang et al.
-(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Summary}\label{ssec:data_summary}
-
-Our review of the technical boundaries and currently tabled Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, and images of crowded fields.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
-
-As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
-
-\begin{quote}
-LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
-\end{quote}
-
-In this section we review in greater detail the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
-Later on, in Section \ref{ssec:dmplans_review}, we review each piece of DM documentation in turn and identify requirements and/or specifications that do not adequately represent the intent of LSST DM, as defined in Section 6 of the DPDD, with respect to Special Programs processing.
-Where relevant we suggest clarifications, changes, or additions to these documents.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
-
-It seems obvious that LSST DM would want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
-In the following three sections we summarize when and how SP data might be incorporated into the pipelines and products of the Level 1 and Alert Stream, the Level 2 DRP, and/or MOPS.
-
-\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
-
-It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST.
-Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
-The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template.
-So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine.
-However, the suitability of shorter and longer exposures is unknown, as discussed in Section \ref{ssec:data_DMconstraints}.
-
-\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
-
-The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
-Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
-
-\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
-
-Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
-When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
-For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
-In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
-
-LSST DM intends to reconfigure its existing pipelines in order to generate imaging and catalog products for Special Programs data, whenever possible.
-In this context, possible means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
-In a possible scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build unique, separate image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
-This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Support Special Programs Processing with Level 3 Pipelines}\label{ssec:dmplans_L3}
-
-If a Special Program requires processing that cannot be accomplished with existing DM algorithms and codes, and/or requires an intensive amount of additional computational resources, it is considered Level 3 -- and in the latter case the processing might have to be undertaken on a separate computational resource.
-To support such Level 3 processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-In most cases, data export and processing on external computation resources will not be necessary.
-The majority of users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their Level 3 pipelines on LSST servers, monitor their progress, and analyze their output.
-DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
-All of this applies to both Level 3 processing for Special Programs data, and reprocessing of WFD main survey data.
-It is furthermore expected that, over time, some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team.
-For both federated and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
-
-% % % % % % % % % % % % % % % % % %
-\subsection{Documentation Review}\label{ssec:dmplans_review}
-
-Here we highlight every requirement or issue relevant to Special Programs that we could find in the existing documentation.
-Since the goal of this work is to make any needed changes to existing documentation, we have endeavored to directly tie every possible concern to the relevant sections of existing documentation (or, where possible, to requirement codes) so that we can more easily direct future changes back into the existing documentations.
-
-% % % % % % %
-\subsubsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{sssec:dmplans_review_srd}
-
-This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
-The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4).
-The latter we know to be an approximate estimate.
-
-% % % % % % %
-\subsubsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{sssec:dmplans_review_lsr}
-
-The LSR is derived from the SRD, and converts the high-level specifications from the SRD into system requirements, which flow-down to and in some cases repeated in the OSS and DMSR (next sections).
-We have the version of LSE-29 last revised on August 4, 2016.
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0} Most notably, there is no written requirement for DM to reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, as it intends to.
-Should there be, and is the LSR right right place for it?
-\end{enumerate}
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0b} Similar to \ref{LSR-0}, there is no written requirement for DM to evaluate the benefit of incorporating data from Special Programs into the products of the WFD main survey, and then to do this when it is found to enable/improve science results, as it intends to.
-Should there be, and is the LSR right right place for it?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-1} There is no requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102; LSE-29).
-Should there be?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-2} While the requirement that LSST process non-standard visit exposures (LSR-REQ-0111, LSE-29) implies that DM must be able to process the full diversity of Special Programs data, it should probably be verified that this is (a) true and (b) possible.
-For example, will DM impose limitations on the exposure times for which the Instrument Signature Removal (ISR) processing can deliver a science-grade quality image, as considered in Section \ref{ssec:data_DMconstraints}?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as "Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-3} It might be helpful to include in this section that Special Programs processing with reconfigured Level 1 and 2 codes, wherever possible, is also an effective requirement similar to LSR-REQ-0032 (LSE-29).
-(This might satisfy \ref{LSR-0}).
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products "of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-4} The requirements for LSST to provide software, services, and hardware resources to support Level 3 data products (LSR-REQ-0041, LSE-39) all have technical limits (i.e., DM's intent to create and run reconfigured pipelines for Special Programs data "unless it requires an intensive amount of additional computational resources" as discussed in Section \ref{ssec:dmplans_L3}); should these be mentioned in the LSR?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0042 requires that LSST produce the data products necessary to support the four primary science missions (dark energy, solar system, transients, milky way; 2.4.5.2, Science Flowdown).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-5} The requirement for LSST to produce the data products necessary for the four primary science missions (LSR-REQ-0042, LSE-29) could be extended with text to include Special Programs data?
-This is might also satisfy \ref{LSR-0}.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
-This is discussed further in Section \ref{sssec:dmplans_review_sizing}.
-
-$\bullet$ LSR-REQ-0075 defines that the WFD main survey science objectives will be met with 90\% of the observing time, with the remainder left to Special Programs (Section 3.3.1, Survey Time Allocation)
-
-% % % % % % %
-\subsubsection{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\label{sssec:dmplans_review_oss}
-
-The OSS contains the requirements on the site, facility, and camera that are necessary to meet the requirements specified by the LSR (LSE-29) and the SRD (LPM-17).
-We have the version of LSE-30 last revised on February 10, 2017.
-For the most part, the {\bf Concerns} raised in this section are not relevant to Data Management and are probably outside the scope of this study.
-But we keep them because they are relevant to Section \ref{ssec:data_bounds}.
-
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-1} There is a requirement on the minimum number of "science proposals" that can be optimized by the scheduler ({\tt nSciProp} = 6, OSS-REQ-0027, LSE-30), but is there also a technical constraint on the maximum?
-This might be relevant to some proposed mini-surveys with complicated observing strategies.
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (2.1.1.7, Visit Optimization).
-
-$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively.
-These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
-These estimates appear adequate for Special Programs in general.
-
-$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 per night on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-2} There is a requirement on the minimum number of calibration frames that will be supported during daytime ({\tt nCalibExpDay} = 450, OSS-REQ-0194, LSE-30); could we imagine a Special Program that would need more than that?
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0125, Section 3.1.5 deals with data products, including requirements for the variety and format of data products (i.e., images and catalogs in Level 1 and 2), the precision of the data products (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (alerts, Level 1, and the Level 2 DRP).
-(Section 3.1.5, Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-3} There are requirements on the type and quality of data products for Level 1, 2, and 3 (OSS-REQ-0125 and all those following, in LSE-30).
-Should it be specified, for Special Programs data, whether or which of these requirements DM must meet?
-For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be {\tt falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
-\end{enumerate}
-
-$\bullet$ Section 3.5, Photometric Calibration, puts requirements on e.g., the relative contributions to photometric errors from instrumental (OSS-REQ-0282) and atmospheric (OSS-REQ-0276) transmissions, and that a catalog of$10000$ reference stars of $17<r<20$ mag per field be created for use (OSS-REQ-0285).
-It seems that these might not apply to non-standard images from Special Programs (3.5, Photometric Calibration).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-4} Section 3.5 of the OSS (LSE-30) puts requirements on the photometric calibrations for WFD data, but it is unclear how and whether these requirements extend to DM providing calibration of non-standard images from Special Programs (this is related to \ref{LSR-2}).
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0319 sets the requirement that the LSST be capable of continuous operation throughout the night when all visits are separated only by the readout time, thereby enabling a deep drilling field style of observations (3.6.1.3, Continuous Exposures).
-
-$\bullet$ OSS-REQ-0291 defines the minimum exposure time as $1$ second with a stretch goal of $0.1$ seconds, with a note that the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times (3.6.1.4, Minimum Exposure Time).
-\\
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-5} There is a requirement on the minimum exposure time (OSS-REQ-0291, LSE-30), but not on the spacing between short exposures and no specified maximum exposure time.
-It would be great to get this clarified from the camera team and updated in the OSS, so we know if they will they impose boundaries on Special Programs data.
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be {\tt tFilterChange} = 120 seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
-There is no officially required minimum time between filter changes.
-Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night\footnote{From Zeljko.}.
-Additionally, changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
-(3.6.2 Filter Swaps \& Changes).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-6}   There are currently are no official requirements on the minimum time between filter changes, only that the time for a filter change will be 120 seconds (OSS-REQ-0293).
-Should the OSS include a clarification that there is no technical constraint on the frequency of filter changes?
-({\it MLG -- maybe unnecessary.}) \end{enumerate}
-
-$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking ({\tt rotTrackTime} = 1 hour) and the half-range of the rotator motion ({\tt rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-7} The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0380 sets the rate and error limits for nonsidereal tracking, which presumably would only be relevant to a Special Program (3.6.3.7, Non-Sidereal Tracking).
-
-% % % % % % %
-\subsubsection{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}}\label{sssec:dmplans_review_dmsr}
-
-The DMSR contains the top-level requirements for all of Data Management, including processing pipelines and products (revision 2017-08-11).
-
-$\bullet$ DMS-REQ-0068 requires that each raw science image store metadata regarding the date/time, site, telescope, and camera -- missing from this is the scheduler and program information?
-That kind of metadata will be necessary to identify images related to Special Programs (1.2.3, Raw Science Image Metadata).
-Related to this is DMS-REQ-0266, which requires the creation of an Exposure Catalog that also stores this kind of metadata independently, but also does not include scheduler and program information.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-1} Clarify if and how scheduler and program information (i.e., identification of an image as being associated with a Special Program) is included in the raw science image metadata and/or the Exposure Catalog (DMS-REQ-0068, -0266); if not, should it be added?
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
-It also says that "this aspect of the processing for Special Programs data is specific to each program".
-It should be specified whether "this aspect" refers to the reduction or archiving of the processed visit images; if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
-(1.2.2, Processed Visit Images).
-This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-2} Confirm whether processed single visits and/or difference images for Special Programs can stay on disk and accessible to e.g., Level 3 pipelines, for a longer amount of time by user request (DMS-REQ-0069 and -0010, LSE-61).
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0069 and DMS-REQ-0010 require that DM create processed single visit images and a difference image for each of those, but this might not be possible for Special Programs that obtain non-standard exposures (i.e., short exposures in which the PSF is not well formed, or very bright sky backgrounds).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-3} Confirm the exposure boundaries for which DM can produce processed single visits and/or difference images for Special Programs that obtain non-standard images (DMS-REQ-0069 and -0010, LSE-61).
-This is related to \ref{LSR-2} and might require a dedicated additional study with simulated data, as discussed in Section \ref{ssec:data_DMconstraints}.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0274 sets the content of an Alert, but this does not currently include scheduler or program information; should this be added?
-This would enable users to e.g., use the LSST mini-broker to filter for only Alerts from their desired Special Program, in case they have a dedicated follow-up resource.
-Furthermore, it might facilitate follow-up more generally if users know e.g., that this Alert is from a field that is going to be observed again in $X$ minutes or $Y$ times in a night (1.3.13, Alert Content).
-However, the latter might be adequately available through the predicted visit schedule server specified by DMS-REQ-0353.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-4} Could and/or should we add the program information and/or predicted revisit schedule to the Alert content (if it isn't included already)?
-(DMS-REQ-0274, -0353) \end{enumerate}
-
-$\bullet$ DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes", but this could use clarification.
-It could be misinterpreted to mean that the {\it science users} must trigger the processing, thereby classifying it as Level 3.
-More likely, it means that e.g., a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline.
-(1.6.1 Processing of Data From Special Programs)
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-5} Can we clarify the mechanism by which Special Programs trigger their own data processing recipes, as required by DMS-REQ-0320?
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing from Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
-It should probably be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
-(1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6} Clarify that only Special Programs data that {\it can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products; see \ref{DMSR-3}), will be incorporated (relevant also to \ref{LSR-0b}, \ref{LSR-2}).
-Perhaps add that in these cases they will contribute directly to the L1 Alert Stream, with a note that generating a separate Level 3 stream of alerts specific to a Special Program should also be possible (DMS-REQ-0321, -0344).
-\end{enumerate}
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6b} Related to \ref{DMSR-6} and DMS-REQ-0344, there is also a potential problem with short exposures (e.g., 5 seconds) that can be processed by  the DIA pipeline, but pose a problem for the Alert Stream because the higher data acquisition rate (e.g., $\sim$6$\times$ higher) would cause a processing back-up.
-Should this situation arise, are there options for short-term increases in parallel processing power at NCSA?
-Such options might also apply to e.g., crowded fields with $>$10k {\tt DIASources}.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0322 specifies that data products from Special Programs processing shall be stored in separate (but joinable) databases.
-
-Section 1.6 "Special Programs" might benefit from the addition of discussion and/or requirements to address the following: \\
-$\bullet$ Should we add a general specification to the DMSR (\S~1.6) to describe whether and how data from Special Programs could be included in the WFD main survey Level 1 and/or 2 data products, as also raised for the LSR in \ref{LSR-0b}, and as discussed in Section \ref{ssec:dmplans_WFD}?
-\\
-$\bullet$ Should we add a specification to the DMSR (\S~1.6) to describe DM's plans to reconfigure its own pipelines for the processing of Special Programs data, perform and verify this processing, and deliver the unique products (separate but joinable to the WFD L1 and L2 products), as also raised for the LSR in \ref{LSR-0}, and as discussed in Section \ref{ssec:dmplans_reconfig}?
-\\
-$\bullet$ Since DM plans to include Special Programs data to the WFD main survey's L1 and L2 products wherever possible, and also run dedicated processing and generate unique products for all SP data, the associated computational and storage resources for Special Programs data might need to be multiplied by a factor of two.
-This is discussed more in Section \ref{sssec:dmplans_review_sizing}, and for now this unknown seems to be well within the uncertainty of the sizing models, and so probably does not (at this time) need to be addressed in the DMSR.
-
-$\bullet$ The following items are applicable to users processing data from Special Programs and/or reprocessing data from the WFD main survey.
-Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
-Section 2.9 "Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
-Sections 3.1 "Software Architecture to Enable Community Re-Use" and 3.2 "Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
-In Section 3.3 "Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
-
-$\bullet$ DMS-REQ-0312 and -0313 require that DM maintain a live L1 and the two most recent DR L1 and L2 catalogs for access by users, but there is no corresponding access requirement for the data products of Special Programs (4.1, Data Archive).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-10} Should there be a requirement for maintenance of user access to data products of Special Programs with the same timescales as the WFD main survey data, similar to DMS-REQ-0312, -0313?
-(This is related to \ref{LSR-1}) \end{enumerate}
-
-% % % % % % %
-\subsubsection{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\label{sssec:dmplans_review_dmad}
-
-The DMAD specifies the design and implementation of the code and algorithms that will be used in the processing of images and creation of catalogs in the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-19.
-As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
-In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
-It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
-This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
-
-The purpose of this study on Special Programs is not to suggest any changes to the DMAD.
-Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3.
-{\bf Deblending --} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
-It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
-Level 3 deblenders for specific Special Programs fields may require development by the user community.
-{\bf Variability Characterization --} The periodic and aperiodic variability characterizations described in Section 6.21 are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
-From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
-It is conceivable that the goals of a particular Special Program might require different algorithms, and these would be written as Level 3 and either made joinable to the DM reconfigured data products, or perhaps federated and incorporated directly.
-{\bf Photometric Redshifts --} As described in Section 5.6.5 the Level 2 DRP {\tt Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then federated and run at scale by DM.
-It is conceivable that the photo-$z$ algorithm for a Special Programs data set product might be different from that used for the WFD main survey.
-
-% % % % % % %
-\subsubsection{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\label{sssec:dmplans_review_dpdd}
-
-The DPDD describes the data products -- the images and catalogs -- to be delivered from the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-01.
-Section 6 describes the data products that DM will provide for Special Programs.
-It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster (see Section \ref{sssec:dmplans_review_sizing}).
-
-The LSST Database Schema Browser\footnote{\url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}} is another way to explore the contents of the planned DM catalogs.
-We identify a couple of potential changes that might be necessary to the schema, which are related to this study on Special Programs.
-These are not high priority.
-
-$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used.
-This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
-DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} might need to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used).
-\end{enumerate}
-
-$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
-It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?
-\\
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an {\tt Object} is an externally provided coordinate.
-\end{enumerate}
-
-$\bullet$ As described in the DMAD and the DPDD, these are the {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization: \\
-{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{sssec:dmplans_review_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
-Are 32 floats in each of the 6 filters always going to be a large enough volume?
-({\it MLG -- I suspect yes and that if not, compression can be used, but am leaving this note here anyway}).
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} DM might want to consider whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-This might be similar to e.g., DMTN-049 for photometric redshifts included in the Level 2 DRP {\tt Object} catalog.
-\end{enumerate}
-
-% % % % % % %
-\subsubsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
-
-LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
-Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}.
-Regarding the reduction and calibration of non-standard visit imags, LSE-180 makes two relevant points: \\
-$\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
-$\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
-
-% % % % % % %
-\subsubsection{Computational Resources Sizing Documentation}\label{sssec:dmplans_review_sizing}
-
-In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will "use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
-This is based on the estimate that Special Programs data comprise $\sim10\%$ of the total observing time.
-Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
-Based on this predicted multiple reprocessing, we might ask whether the "10\% of the computing resources" allocated for DM processing of Special Programs data is too small, but it is still a bit to early to know this.
-It will depend on the exact final mix of Special Programs, which may well change from year to year, and furthermore the current error bars on the sizing are of order $\pm10\%$.
-
-The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
-LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
-DM Compute Sizing spreadsheet (LDM-138) and explanation document (LDM-140) estimates the total compute resources needed for all processes, quoting the same 10\% addition for Level 3.
-DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
-As mentioned in Section \ref{sssec:dmplans_review_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
-
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
-Have these systems been sized?"}.
-The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing "catch-up" L1, processing
-the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
-There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can "borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
-
-% % % % % % %
-\subsection{The Limited Documentation Regarding Image Detection Efficiencies}
-
-Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
-Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
-What, exactly, are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
-Note that here we are talking about planting fakes in the single images for transient detection efficiencies with difference imaging analysis, and not planting fakes in the CoAdds for point-source limiting magnitudes).
-
-$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
-will record information on the pixel variance ...
-as well as the per-pixel masks ...
-These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
-
-$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
-detection efficiency for point sources vs.
-mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
-
-$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
-The details of what metrics will be mapped, the format and scale of the maps (e.g.
-hierarchical pixelizations vs.
-polygons), and the way the metrics will be computed are all unknown"}.
-It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
-(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
-Source detection thresholds can be estimated through the use of sky sources"}.)
-
-$\bullet$ Suchyta et al.
-(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
-
-$\bullet$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
-
-$\bullet \bullet \bullet$ {\bf Summary.} {\it MLG - Perhaps I could take this on as a future study, and assess our options, their costs, and the scientific payoff.
-Since this is not a question limited to DM processing of Special Programs data, specifically, I end this side note here and no {\bf Concerns} are spawned.}
 
@@
 Step 6. Science User Level 3 Pipelines and Products. \\
-The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
+The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external process (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be adopted).
 Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
@@
 The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-1} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
-Or might there be issues due to train/test set mismatch?
-This is probably a question to be answered with testing, later, and incorporated into the DMAD?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-2} Confirm that resources for packet distribution outside of the Alert Stream will be supported (related to \ref{DMSR-4}).
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{SPCS-1} The purpose of this ticket is to resolve whether or not the internal real/bogus routine be able to run on a nightly CoAdd of deep drilling difference images.
+If there will be issues due to a mismatch between the training and testing sets, what is the cost/benefit of fixing this and enabling this capability?
+This is not urgent and not limited to Special Programs.
 \end{enumerate}
@@
 Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds (see \ref{OSS-5} and \ref{DMSR-6b}), these data could be incorporated.
+Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:dmplans_NSV}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could be incorporated.
 
@@
 If short-exposure images cannot be processed with the existing DM algorithms, would the Level 3 user-built pipeline have to start from the raw data and create new, custom algorithms for e.g., ISR?
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-3} Is it at all conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
-\end{enumerate}
+
 A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
 This would certainly be considered a Level 3 algorithm.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-4} Are any DM algorithms being developed to estimate the flux of saturated stars using e.g., the PSF wings?
-Could the brighter-fatter issue be used for some science advantage?
-\end{enumerate}
 
@@
 Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:data_DMconstraints} and \ref{sssec:dmplans_review_dmad}).
+Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:dmplans_NSV} and \ref{ssec:docrev_dmad}).
 
@@
 The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Level 1 and 2 pipelines, and not much reconfiguration would be needed.
-The biggest difference might be the federation of a user-supplied deblender algorithm optimized for very crowded fields.
+The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
 
@@
 I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
-[this isn't in the requirements right now -- right now we say that images will become available in 24hrs.
-But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have \#3, you get \#6 for free as well.]
+(This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_L3}).
 

### 2017-11-10 ms.tex
minor typos

@@
 
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to investigate the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to clarify the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations.
 Questions to be answered include, e.g., product ownership of the reconfigured pipelines and test data products, and how do we set goals and deadlines for this?
@@
 
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-0} RFC: We propose that the DMSR include a high-level specification (probably in Section 1.6) that the data from Special Programs will be included in the products of the wide-fast-deep main survey only when it is (a) possible and (b) beneficial to the primary science objectives of those products.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-0} RFC (LSE-61): We propose that the DMSR include a high-level specification (probably in Section 1.6) that the data from Special Programs will be included in the products of the wide-fast-deep main survey only when it is (a) possible and (b) beneficial to the primary science objectives of those products.
 It should also be updated to reflect the fact that DM will reconfigure its own pipelines ``wherever possible", process the Special Programs data, verify the unique data products, and make them publicly available.

### 2017-11-14 deployment/build.py
Add initial deployment files

@@ -1,0 +1,55 @@
+#!/usr/bin/env python
+
+import argparse
+import datetime
+import os
+
+import jinja2
+import pytz
+from pytz import timezone
+
+
+def parse_args():
+    parser = argparse.ArgumentParser(
+        description='Create a static website for DMTN-065.'
+    )
+    parser.add_argument(
+        'template_path',
+        help='Path of the index.html Jinja2 template.'
+    )
+    parser.add_argument(
+        'deploy_dirname',
+        help='Directory where the built index.html should be added '
+             'for deployment.'
+    )
+    return parser.parse_args()
+
+
+def main():
+    args = parse_args()
+
+    loader = jinja2.FileSystemLoader(os.path.dirname(args.template_path))
+    template_env = jinja2.Environment(
+        loader=loader,
+        autoescape=jinja2.select_autoescape(['html', 'xml']))
+    template = template_env.get_template(os.path.basename(args.template_path))
+
+    utc = pytz.utc
+    pacific_tz = timezone('US/Pacific')
+    pacific_dt = pacific_tz.normalize(datetime.datetime.utcnow().
+                                      replace(tzinfo=utc))
+
+    html = template.render(
+        branch_name=os.getenv('TRAVIS_BRANCH'),
+        pacific_timestring=pacific_dt.strftime('%Y-%m-%d %H:%M:%S %Z%z'),
+        page_title='DMTN-065',
+        page_description='Draft of DMTN-065'
+                         'TN'
+    )
+    index_html_path = os.path.join(args.deploy_dirname, 'index.html')
+    with open(index_html_path, mode='w') as f:
+        f.write(html)
+
+
+if __name__ == '__main__':
+    main()

### 2017-11-14 DMTN-065.tex
Add DMTN handle to tex file
Use more modern latex

--- 2017-11-10 ms.tex
@@
 
-\setDocRef{DMTN-nnn}
+\setDocRef{DMTN-065}
 \date{\today}
@@
 
-\setDocAbstract{{\bf This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget.
-Unresolved issues spawn {\bf Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}
+\setDocAbstract{\textbf{This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget.
+Unresolved issues spawn \textbf{Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}
 
@@
 \noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
-This study spawns {\bf Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.
-
-%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an {\bf Action} is spawned.
+This study spawns \textbf{Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.
+
+%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an \textbf{Action} is spawned.
 
@@
 Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
-%These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
-{\bf The Concerns raised by this study fall into two main categories:} \\
+%These \textbf{Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
+\textbf{The Concerns raised by this study fall into two main categories:} \\
 %1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
@@
 
-All aspects related to DM's {\it processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.
+All aspects related to DM's \textit{processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.
 
@@
 
-{\bf Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
+\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
 The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
@@
 
-{\bf Longer Exposures.} There is no maximum exposure time specified for an LSST image.
+\textbf{Longer Exposures.} There is no maximum exposure time specified for an LSST image.
 Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
@@
 They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
+Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al.
+2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or \texttt{ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
 
@@
 OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds.
-However, {\it if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated {\tt DIASources}.
-
-{\bf Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds.
+However, \textit{if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated \texttt{DIASources}.
+
+\textbf{Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds.
 This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later?
-{\bf How long does it take to save an image to disk?}
-
-{\bf Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
-(Such options might be needed anyway to process crowded fields with $>$10k {\tt DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
+\textbf{How long does it take to save an image to disk?}
+
+\textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
+(Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
 However, at NCSA there should be ways to elastically change the amount of processing power available.
@@
 
-{\bf Unassociated {\tt DIASources}:} the consistency of {\tt DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field).
-In other words, it is not guaranteed that the next string of processing would have access to the latest {\tt DIAObjects} for association.
+\textbf{Unassociated \texttt{DIASources}:} the consistency of \texttt{DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field).
+In other words, it is not guaranteed that the next string of processing would have access to the latest \texttt{DIAObjects} for association.
 It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations.
@@
 Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
-{\it MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
+\textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
 
@@
 Special Programs are more likely to include crowded fields than the WFD main survey area.
-Due to the increased number of sources, the number of {\tt DIASources} -- and therefore the number of Alerts -- increases as well.
+Due to the increased number of sources, the number of \texttt{DIASources} -- and therefore the number of Alerts -- increases as well.
 In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
@@
 Might this requirement be slackened to accommodate a higher cadence Special Program?
-If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping {\it vs.} delaying Alerts during observations with a fast ingest cadence.
+If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping \textit{vs.} delaying Alerts during observations with a fast ingest cadence.
 These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts.
@@
 We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits.
-Consequential images of the same field may lead to unassociated {\tt DIASources} of the same {\tt DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately.
+Consequential images of the same field may lead to unassociated \texttt{DIASources} of the same \texttt{DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately.
 \end{enumerate}
@@
 
-{\bf Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
+\textbf{Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
 So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine.
@@
 The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries).
-Since MOPS takes {\tt DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
+Since MOPS takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
 As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
@@
 Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
+We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP \texttt{Objects} database that is built from a set of images with nearly constant depth and cadence.
 When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
@@
 In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey.
-This is mostly discussed here in context with the {\tt Source} and {\tt Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.
+This is mostly discussed here in context with the \texttt{Source} and \texttt{Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.
 
@@
 In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
+For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
 This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
@@
 However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3).
-LSST has furthermore allocated an {\it additional} 10\% of its computing resources for user-driven processing and analysis.
+LSST has furthermore allocated an \textit{additional} 10\% of its computing resources for user-driven processing and analysis.
 This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).
@@
 
-However, in the case where user-driven processing pipelines consume a {\bf large} amount of that 10\%, data export and processing on external computation resources {\bf will} be necessary.
+However, in the case where user-driven processing pipelines consume a \textbf{large} amount of that 10\%, data export and processing on external computation resources \textbf{will} be necessary.
 The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.
 
-{\bf Level 3 Alert Stream?
+\textbf{Level 3 Alert Stream?
 --} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package.
@@
 
-{\bf Lower Latency for Data Access?
+\textbf{Lower Latency for Data Access?
 --} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner.
@@
 
-$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
+$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be \texttt{DRT1} = 1 year (1.4.2, Data Release Processing).
 \begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102).
@@
 
-$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
+$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of \texttt{minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
 The full capabilities of DM to process non-standard visits is not yet well defined; see \ref{NSV-1}.
@@
 
-$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
+$\bullet$ LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
 This is discussed further in Section \ref{ssec:docrev_sizing}.
@@
 
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, {\tt nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies.
+$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, \texttt{nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies.
 \end{enumerate} % Assign to Zeljko.
@@
 
-$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively.
+$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively.
 These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
@@
 
-$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively.
-These are {\it minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.
+$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively.
+These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.
 
@@
 Would these also apply to data from Special Programs?
-For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be {\tt falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
+For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be \texttt{falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
 \begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-3} RFC (LSE-30): We propose to add a statement to Section 3.1.5 of the OSS that DM will make a ``best effort" to meet the data products requirements with the Special Programs data.
@@
 
-$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be ${\tt tFilterChange} = 120$ seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
+$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be $\texttt{tFilterChange} = 120$ seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
 There is no officially required minimum time between filter changes.
@@
 
-$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking ({\tt rotTrackTime} = 1 hour) and the half-range of the rotator motion ({\tt rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
+$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking (\texttt{rotTrackTime} = 1 hour) and the half-range of the rotator motion (\texttt{rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
 Questions about whether there is a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), are raised in Section \ref{ssec:data_bounds}.
@@
 \begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-1} RFC (LSE-61): We propose that DMS-REQ-0068 be updated to include that the scheduler and program metadata (i.e., a way to identify an image as being associated with a Special Program) with the raw science image, and/or with the Exposure Catalog (DMS-REQ-0068, -0266).
-We furthermore propose that the {\tt visitID} and the total number of exposures in the visit are also included in this updated requirement (e.g., to accommodate a deep drilling sequence of a single visit of 50 exposures).
+We furthermore propose that the \texttt{visitID} and the total number of exposures in the visit are also included in this updated requirement (e.g., to accommodate a deep drilling sequence of a single visit of 50 exposures).
 \end{enumerate}
@@
 This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
-We assume this applies to standard visit images that {\it can} be processed.
+We assume this applies to standard visit images that \textit{can} be processed.
 \begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-2} RFC (LSE-61): We propose that DMS-REQ-0069 be clarified regarding whether processed single visits and/or difference images for Special Programs can stay archived and accessible to e.g., Level 3 pipelines, for a longer amount of time, possibly by user request.
@@
 $\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
-It should be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
+It should be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within \texttt{OTT1} = 1 minute means contributing to the L1 Alert Stream.
 (1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-6} RFC (LSE-61): We propose that DMS-REQ-0321 and -0344, which specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey, be reworded to clarify that only Special Programs data that {\it can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products), will be incorporated into Level 1 processing and contribute to the Alert Stream.
+\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-6} RFC (LSE-61): We propose that DMS-REQ-0321 and -0344, which specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey, be reworded to clarify that only Special Programs data that \textit{can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products), will be incorporated into Level 1 processing and contribute to the Alert Stream.
 \end{enumerate}
@@
 In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
-It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
+It is the DMAD's codes and algorithms that DM will be \textit{reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
 This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
@@
 
-\subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP {\tt Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM.
+\subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP \texttt{Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM.
 It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.
@@
 
-$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used.
+$\bullet$ The database schema for \texttt{DIASource} does not appear to have an element that identifies which template image that was used.
 This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
@@
 However, K.-T.
-has pointed out that the capability to regenerate the {\it exact same} template -- the pixels that were subtracted -- is not a current deliverable.
+has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable.
 However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.
 
-$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
+$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
 It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
+Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType.
 Indicates which input was used to produce a given object."} Is that all we need?
 \\
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-1} The purpose of this ticket is to clarify whether {\tt Object.prv\_inputID} in the database schema will identify whether an {\tt Object} is an externally provided coordinate.
-\end{enumerate}
-
-$\bullet$ The {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
-{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-1} The purpose of this ticket is to clarify whether \texttt{Object.prv\_inputID} in the database schema will identify whether an \texttt{Object} is an externally provided coordinate.
+\end{enumerate}
+
+$\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
+\texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+\texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
 Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
 Are 32 floats in each of the 6 filters always going to be a large enough volume?
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-2} The purpose of this ticket is to ask if DM should consider whether the {\tt float[6x32]} for the characterization parameters for variability is of adequate size to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-This study might be similar to e.g., DMTN-049 for photometric redshifts included in the DRP {\tt Object} catalog.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-2} The purpose of this ticket is to ask if DM should consider whether the \texttt{float[6x32]} for the characterization parameters for variability is of adequate size to express the additional timescales and depths covered by all possible Special Programs observing strategies.
+This study might be similar to e.g., DMTN-049 for photometric redshifts included in the DRP \texttt{Object} catalog.
 \end{enumerate}
@@
 
-LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
+LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
 Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}.
@@
 DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
-As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that {\it an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
-\end{enumerate}
-
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
+As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
+It is important to note that this 10\% is \textit{in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
+
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that \textit{an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
+\end{enumerate}
+
+Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks \textit{"whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
 Have these systems been sized?"}.
-The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
+The answer to this comes from K-T: \textit{In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
 There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can ``borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-2} The purpose of this ticket is to ensure that the overall DM computational system is appropriately sized, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, {\it and} incorporated into the Level 2 data products, {\it and} have their own dedicated reconfigured pipelines and separate products, {\it and} have a higher probability for user-defined processing.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-2} The purpose of this ticket is to ensure that the overall DM computational system is appropriately sized, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, \textit{and} incorporated into the Level 2 data products, \textit{and} have their own dedicated reconfigured pipelines and separate products, \textit{and} have a higher probability for user-defined processing.
 The latter would apply to the size of the DACs.
@@
 
-$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
+$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that \textit{"Each processed image ..
 will record information on the pixel variance ...
@@
 
-$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
+$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): \textit{"The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
 detection efficiency for point sources vs.
@@
 
-$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
+$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step \textit{"is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
 The details of what metrics will be mapped, the format and scale of the maps (e.g.
@@
 polygons), and the way the metrics will be computed are all unknown"}.
-It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
-(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
+It also states that this must be extendable to Level 3, but that \textit{"the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
+(Section 3 "Alert Production" also specifies that in LDM-151 \textit{"we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
 Source detection thresholds can be estimated through the use of sky sources"}.)
@@
 $\bullet$ Suchyta et al.
-(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
+(2016) presents \texttt{balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses \texttt{galsim} to embed shapes built of e.g., Sersic profiles).
 
@@
 In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
-Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
-
-\noindent {\bf Resources:} \\
+Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an \textit{incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
+
+\noindent \textbf{Resources:} \\
 $\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
@@
 However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
-$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
+$\bullet \bullet \bullet$ \textbf{Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
 The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures.
-Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night.
+Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 \texttt{DIASource} catalogs updated each night.
 The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
@@
 $\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}.
-For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
+For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
 \citep{SzkodyWP}.
@@
 \\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-\\
-$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
+$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+\\
+$\bullet \bullet \bullet$ \textbf{Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
 Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks.
@@
 The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
+A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
 \cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
@@
 \\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images.
+$\bullet \bullet \bullet$ \textbf{Summary.} All of these science goals appear possible with standard visit images.
 They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
@@
 $\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
+$\bullet \bullet \bullet$ \textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
 Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
@@
 $\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
+$\bullet \bullet \bullet$ \textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
 In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.
@@
 $\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
-$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
+$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
+$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
+We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
+$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
+$\bullet \bullet \bullet$ \textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
 
@@
 This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}.
-The DM team will also interact informally with the community through, e.g., a {\tt \Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
+The DM team will also interact informally with the community through, e.g., a \texttt{\Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
 
@@
 
-\subsubsection{{\tt \Large{Community}} Forum}\label{sssec:data_comm_forum}
-
-As the science community prepares their white paper proposals they have been invited to interact with Data Management through this {\tt \Large{Community}} forum on DDF.
+\subsubsection{\texttt{\Large{Community}} Forum}\label{sssec:data_comm_forum}
+
+As the science community prepares their white paper proposals they have been invited to interact with Data Management through this \texttt{\Large{Community}} forum on DDF.
 I am a watcher on this forum.
@@
 Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
 
@@
 As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would {\it all} be incorporated.
+However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.
 
@@
 Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
 
@@
 First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of {\tt DIAObject}, {\tt DIASource}, and {\tt Object} that are unique to this DDF.
+On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF.
 The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
@@
 As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would {\it all} be incorporated.
+However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would \textit{all} be incorporated.
 
@@
 Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
+Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
 
@@
 
-A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
 This would certainly be considered a Level 3 algorithm.
@@
 Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
+The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
 
@@
 It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-({\it MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
+(\textit{MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
 

### 2017-11-14 DMTN-065.tex
Cite documents in standard manner

@@
 
-The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):
+The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (\DPDD, \citeds{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):
 
@@
 The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future.
-Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.
+Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, \citeds{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.
 
@@
 \\
-$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
+$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \href{https://project.lsst.org/groups/ccb/node/938}{LCR-458}. \\
 $\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
@@
 
-As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (DPDD, LSE-163, \cite{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).
+As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (\DPDD, \citeds{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).
 
@@
 As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field.
-In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).
+In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, \citeds{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).
 
@@
 In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
+For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the \DPDD, \citedsp{LSE-163}).
 This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
@@
 % % % % % % %
-\subsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{ssec:docrev_srd}
+\subsection{Science Requirements Document, SRD, \citeds{LPM-17}}\label{ssec:docrev_srd}
 
@@
 % % % % % % %
-\subsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{ssec:docrev_lsr}
+\subsection{LSST System Requirements, LSR, \citeds{LSE-29}}\label{ssec:docrev_lsr}
 
@@
 % % % % % % %
-\subsection{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\label{ssec:docrev_oss}
+\subsection{Observatory System Specifications, OSS, \citeds{LSE-30}}\label{ssec:docrev_oss}
 
@@
 % % % % % % %
-\subsection{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}}\label{ssec:docrev_dmsr}
+\subsection{Data Management Subsystems Requirements, DMSR, \citeds{LSE-61}}\label{ssec:docrev_dmsr}
 
@@
 % % % % % % %
-\subsection{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\label{ssec:docrev_dmad}
+\subsection{Data Management Applications Design, DMAD, \citeds{LDM-151}}\label{ssec:docrev_dmad}
 
@@
 % % % % % % %
-\subsection{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\label{ssec:docrev_dpdd}
+\subsection{Data Products Definitions Document, DPDD, \citeds{LSE-163}}\label{ssec:docrev_dpdd}
 
@@
 % % % % % % %
-\subsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
+\subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}
 
@@
 So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
-\cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
-Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}).
+\citep{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
+Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}).
 We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.
@@
 \noindent \textbf{Resources:} \\
-$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
-$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
+$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \citep{2008arXiv0805.2366I} \\
+$\bullet$ The LSST Science Requirements Document (\SRD), \citeds{LPM-17} \\
 $\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
@@
 $\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
-$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
+$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
+$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \citep{2017arXiv170804058L}
 
@@
 \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
+(2008, \citep{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
 Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
@@
 $\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
-$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
+$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \citep{2017arXiv170804058L}) \\
+$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \citep{2017arXiv170804058L}) \\
+$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \citep{2017arXiv170804058L}) \\
 $\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
-Regarding SAS, \cite{BeckerWP} says that "the multi-fit algorithm ...
+Regarding SAS, \citep{BeckerWP} says that "the multi-fit algorithm ...
 naturally provides a base infrastructure for this process.
@@
 $\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
-When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}.
+When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \citep{DhitalWP}.
+\\
+$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \citep{SzkodyWP}.
 For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
@@
 \\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
 \\
@@
 $\bullet$ Transits.
-The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
-\cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
+The nominal DDF plan described in \citep{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
+A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \citep{2008arXiv0805.2366I}).
+\citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
 $\bullet$ Microlensing.
-Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
+Slower than a transit, \citep{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
 However, this will require image differencing in crowded fields.
-\cite{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
+\citep{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
 Microlensing can also be done with DDF-style observing at low Galactic latitudes.
@@
 \noindent \textbf{Supernovae} \\
-$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
+$\bullet$ The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
 The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
-\cite{KesslerWP} \\
-$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}.
+\citep{KesslerWP} \\
+$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}.
 \\
@@
 SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
+The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citep{FergusonWP}.
 \\
@@
 $\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
-\cite{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
+\citep{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
 \\
@@
 After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation.
-While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \cite{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
+While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \citep{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
 
@@
 
-A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
 This would certainly be considered a Level 3 algorithm.

### 2017-11-14 DMTN-065.tex
Replace some URLs with references and use \jira

@@
 Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or \texttt{ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
+2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} to at least simply characterize the PSF stability as a function of exposure time.
 
@@
 $\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
+$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} \\
 $\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \citep{2017arXiv170804058L}
@@
 At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call.
-The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
+The relevant Jira issue is \jira{LIT-325} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
 The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director.

### 2017-11-14 DMTN-065.tex
Use DocuShare links to White Papers
These have more longevity than the project web site URLs

@@
 $\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \citep{2017arXiv170804058L}) \\
-$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
-Regarding SAS, \citep{BeckerWP} says that "the multi-fit algorithm ...
+$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citedsp{Document-11013}.
+Regarding SAS, \citedsp{Document-11013} says that "the multi-fit algorithm ...
 naturally provides a base infrastructure for this process.
@@
 $\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
-When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \citep{DhitalWP}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \citep{SzkodyWP}.
+When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \citedsp{Publication-141}.
+\\
+$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \citedsp{Publication-145}.
 For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
-\citep{SzkodyWP}.
+\citedsp{Publication-145}.
 \\
@@
 The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
-\citep{KesslerWP} \\
+\citedsp{Publication-144} \\
 $\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}.
@@
 SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citep{FergusonWP}.
+The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citedsp{Publication-142}.
 \\
@@
 $\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
-\citep{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
+\citedsp{Publication-142} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
 \\
 $\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
-$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
+$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}. \\
 $\bullet \bullet \bullet$ \textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
@@
 We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
-$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
+$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
 $\bullet \bullet \bullet$ \textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
@@
 This Special Programs processing summary is based on Becker et al.
-(2011) white paper to find TNOs with shift-and stack (SAS) \citep{BeckerWP}.
+(2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
 See Section \ref{ssec:SPCS_SAS} for a more general processing case-study of a shift-and-stack Level 3 pipeline running regularly on a large amount of data.
@@
 Data obtained in the $g$-band filter is also acceptable.
-\citep{BeckerWP}
+\citedsp{Document-11013}
 

### 2018-06-18 DMTN-065.tex
Set repo location
major revisions; issued

@@
-\documentclass[DM,lsstdraft,toc]{lsstdoc}
+\documentclass[DM,lsstdoc,toc]{lsstdoc}
 \usepackage{graphicx}
@@
 \date{\today}
-\setDocRevision{TBD}
-\setDocStatus{draft}
+\setDocUpstreamLocation{\url{https://github.com/lsst-dm/dmtn-065}}
 
-\setDocAbstract{\textbf{This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget.
-Unresolved issues spawn \textbf{Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}
+\setDocAbstract{This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys).
+The main target audience is the LSST Data Management (DM) team, but members of the community who are preparing white papers on science-driven observing strategies may also find this document useful.
+The potential diversity of data from Special Programs is summarized, including boundaries imposed by technical limitations of the LSST hardware.
+The capability of the planned Data Management system to processes this diversity of Special Programs data is the main focus of this document.
+Case studies are provided as examples of how the LSST software and/or user-generated pipelines may be combined to process the data from Special Programs.}
 
 \setDocChangeRecord{%
-\addtohist{1}{2017-11-??}{Internal working document.}{Melissa Graham}
+\addtohist{1}{2017-11-14}{Status: internal working document.}{Melissa Graham}
+\addtohist{1}{2018-06-17}{Updated to finalize and issue.}{Melissa Graham}
 %\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
@@
 
-The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (\DPDD, \citeds{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):
+The main LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, but this is expected to be accomplished with $85$--$90$\% of the observing time available over the $10$ year survey.
+The remaining $10$--$15$\% of the time will be spent on Special Programs: alternative survey areas and/or observing strategies driven by a specific science goal.
+A call for white papers that provide scientific motivation and observing strategies for the WFD survey and Special Programs, \citeds{Document-28382}, released in June 2018, has a deadline of November 30 2018.
+It is conceivable that Special Programs might obtain imaging data that are significantly different from the WFD main survey, and/or requires special processing in order to achieve the program's science goals.
+This document provides an in-depth description of the role of the LSST Project in preparing to process the data from Special Programs (expanding on the general description in Section 6 of \citeds{LSE-163}).
+The main target audience is the LSST Data Management (DM) team, but members of the science community who are preparing white paper proposals may also find this document useful.
 
-\begin{quote}
-LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
-\end{quote}
+{\bf Special Programs Terminology -- } The WFD main survey will survey the sky with a series of {\bf visits}, and each field will be visited $>800$ times over $10$ years.
+A {\bf standard visit} is composed of $2\times15$ second exposures (commonly referred to as ``snaps") and an {\bf alternative standard visit} is composed of a single $30$ second exposure.
+A {\bf non-standard visit} is any other exposure time(s) or number of snaps.
+Special Programs are typically divided into two types: {\bf Deep Drilling}, a single pointing for which many exposures are obtained in a relatively short amount of time (e.g., $>2\times$ as many visits in six months as the WFD will obtain in 10 years); and {\bf Mini-Surveys}, which refer to either new sky areas observed with a WFD-like survey, or sky areas within the WFD but observed with a specialized strategy.
 
-\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
-This study spawns \textbf{Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.
+{\bf The LSST Project's Role in Processing Special Programs Data -- } The LSST Project will not take formal responsibility for specialized data reduction algorithms needed to process data, including images taken in non-standard modes.
+The term "specialized algorithms" refers to software that is not already within scope of the LSST Data Management (DM) science pipelines, and may include, for example: difference imaging for short exposures in which the PSF is not well-formed, shift-and-stack for faint moving objects, or any software with computational needs that significantly surpass the processing budget per image (compared to the processing of a WFD image).
+The Project will incorporate Special Programs data into the Prompt and/or Data Release processing pipelines and data products of the WFD Main Survey, such as {\tt Alerts}, {\tt CoAdds}, or {\tt Source} and {\tt Object} catalogs (with appropriate flags; LSST data products are described in \citeds{LSE-163}), whenever this (1) can be accomplished with existing software, and (2) is scientifically beneficial to that data product.
+The Project will also reconfigure its pipelines to generate separate imaging and catalog data products for Special Programs, whenever this can be accomplished with existing software.
+Finally, the Project will enable user-generated processing via the Science Platform (\citeds{LSE-319}), which will provide software tools and computational resources for (re)processing LSST data.
 
-%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an \textbf{Action} is spawned.
+{\bf Document Overview -- } The purpose of this document is to assess the potential diversity of imaging data that might be obtained by Special Programs (Section \ref{sec:data}), and to explore and clarify whether -- and to what extent -- DM's planned pipelines and user services will be able to handle this diversity of data (Section \ref{sec:dmplans}).
+As described in the call for white papers on cadence optimization (\citeds{Document-28382}), Special Programs that may require specialized or computationally-intense algorithms to meet their science goals are required to describe how these processing needs will be met.
+Sections \ref{sec:data} and \ref{sec:dmplans} have been designed to help white paper authors figure out whether their proposed Science Program(s) require processing that would be considered out of DM's scope, and in Section \ref{sec:SPCS} we have provided a series of science-driven case studies as examples.
 
-Section \ref{sec:data} compiles an overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
-We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera.
+% In Appendix \ref{sec:docrev}, we endeavor to ensure that the DM's plans for Special Programs processing are appropriately expressed in all of the major Project documents and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
+This content will only be of interest to the DM team.
 
-Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data, including the capabilities for processing non-standard visits and whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
-We also describe DM's intent for incorporating Special Programs data into the WFD main survey products, reconfiguring pipelines to create unique products, and providing support for end-user Level 3 processing.
-
-Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
-%These \textbf{Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
-\textbf{The Concerns raised by this study fall into two main categories:} \\
-%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
-%2.
-Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget.
-\\
-
-Section \ref{sec:SP_SC} gathers the existing efforts from the Science Community regarding Special Programs, summarizes the scientific needs of as many currently proposed Special Programs as we can find, and outlines the future mechanisms for LSST to gather more community input -- including some recommendations for the future call for Special Programs white paper proposals.
-
-Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys.
-These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional issues regarding the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
-
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
@@
 
-To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data.
-The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (i.e., standard visit images), then it might not be adequately handled by the DM system.
-To get an idea of the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes, in Section \ref{ssec:data_bounds}.
+To define the extent to which Data Management will be able to process data from Special Programs, a comprehensive understanding of the potential diversity of data from Special Programs -- compared to the WFD's sky survey of standard visits -- is needed.
 
-We can also get an idea of the expected diversity in exposure type and survey pattern from the current set of proposed Special Programs for LSST, which are reviewed in more detail in Section \ref{sec:SP_SC}.
-We find that most of the proposed Special Programs are likely to use standard visits, two $\sim15$ second exposures, and that the potential diversity of Special Programs data is mainly comprised of exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, long series of exposures obtained of the same field (i.e., deep drilling), and images of crowded fields.
+To build up this comprehensive understanding, in Section \ref{ssec:data_bounds} we considered any technical limitations that the facility and instrumentation will or might place on the data.
+We find that the hardware imposes few boundaries on how data can be obtained, but that a high number of filter changes and/or long slews are inefficient due to their large overheads.
+The minimum exposure time is $1$ second (stretch goal: $0.1$ seconds), but there is currently a technical boundary that limits the readout rate to $1$ every $15$ seconds.
 
-All aspects related to DM's \textit{processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.
+Next, we considered the Special Programs that have been openly discussed so far in the Science Community (Appendix \ref{sec:data_prev}).
+Based on this consideration, most of the proposed Special Programs are likely to use visits that are similar to the WFD Main Survey, but some will require exposures that are significantly shorter, or that are obtained with a bright sky background during twilight.
+The cadence and patterns may also differ from the WFD main survey, such as long series of exposures obtained of the same field (i.e., deep drilling), or a strategy optimized to find very fast-moving objects.
+In addition, images of very crowded fields in the Galactic Plane may be included (at all or more often than the WFD Main Survey).
+It does not appear that any of the previously proposed Special Programs violate the hardware-imposed boundaries discussed in Section \ref{ssec:data_bounds}.
 
-% % % % % % % % % % % % % % % % % %
-\subsection{Technical Boundaries for Special Programs Data}\label{ssec:data_bounds}
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
+\subsection{Hardware Boundaries}\label{ssec:data_bounds}
 
-We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals.
-For the purposes of this document we consider the ``observing boundaries" as those imposed by the camera, telescope, or site, and independent of any DM processing capabilities.
-It is not the job of this document to define these observing boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
+Here we consider the technical boundaries on the diversity of data products that are expected to (or may) be imposed by limitations from the camera, telescope, and/or site (boundaries imposed by DM processing capabilities are considered in Section \ref{sec:dmplans}).
 
-The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future.
-Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, \citeds{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.
+{\bf Filter Changes -- } The maximum time for filter change is $120$ seconds: $30$ seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and $90$ seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
+Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4 exposures$ in a given filter would be obtained between filter changes -- but this is not actually a technical boundary.
 
-Filter Changes: \\
-$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
-\\
-$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \href{https://project.lsst.org/groups/ccb/node/938}{LCR-458}. \\
-$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
-$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
-However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
-\\
-$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
-$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights. \\
+{\bf Filter Carousel Loads -- } As described in \citeds{Document-28382}, the filter carousel can hold five of the six LSST filters at a time, and filter loads are done in the day.
+The system is designed to support $3000$ loads and $100000$ filter changes in $15$ years, which is an average of $17$ changes per night (after accounting for filter changes during calibrations).
+Individual filters will support $30000$ changes in $15$ years.
+Based on these technical boundaries, we know that there will never be data in more than five filters in a given night.
 
-Exposure Times: \\
-$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
-$\bullet$ The maximum exposure time is not restricted. \\
-$\bullet$ However, for exposure times there are other considerations.
+{\bf Exposure Times -- } The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
+The maximum exposure time is not restricted.
+The readout time is $2$ seconds, and would be significant overhead on short exposures.
+Images with exposure times $<15$ seconds may still have to be separated by $15$ seconds for thermal tolerance; i.e., that the minimum readout rate is one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
+We therefore consider the $15$ interval between images a technical boundary on the potential diversity of data products.
+% $\bullet$ However, for exposure times there are other considerations.
 The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
@@
 
-Camera Rotation: \\
-$\bullet$ The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
-\textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
-Can't remember where I heard it.}
+{\bf Telescope Slew -- } As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no technical boundaries on the size of a single slew or the accrued slew distance.
 
-Telescope Slews: \\
-$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
-
-Total Time: \\
-$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey).
-This could be increased a bit, but otherwise imposes an upper limit on the total time request of a given Special Program.
-
-Operations Simulations: \\
-$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{Bounds-1} The purpose of this ticket is to request clarification regarding several aspects of the telescope and camera capabilities that might impose boundaries on the diversity of data from Special Programs.
-First, have the following proposed filter change tolerances been confirmed: (1) the minimum time between filter changes has no restrictions from thermal tolerances, but it is recommended to keep the filter change rate lower than once every 10 minutes in order to minimize overheads; (2) the maximum total number of filter changes is 100000 over 15 years (18/night) and the maximum number of filter swaps in and out of the carousel is 3000 in 15 years (once every two nights).
-Second, are there any restrictions on the use of the camera rotator or the total amount of slew time?
-For example, could a Special Program continuously jump between well-separated fields on the sky for an extended amount of time?
-\end{enumerate}
+{\bf Camera Rotation -- } There requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
+JIRA ticket DM-12573 is currently open and asking for clarification from the camera team on this.
+Until then, we assume there are no technical boundaries imposed by camera rotation constraints on the potential diversity of data products.
 
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
 \clearpage
-\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
+\section{Processing Special Programs Data with LSST DM Pipelines} \label{sec:dmplans}
 
-As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (\DPDD, \citeds{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).
-
-In this section we review in greater detail the planned ability of DM to process non-standard visit images (Section \ref{ssec:dmplans_NSV}), the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
+In this section we go into greater detail regarding the Projects' role in processing data from Special Programs, as introduced in Section \ref{sec:intro}.
+We discuss the ability of the planned DM pipelines to: process diverse imaging data that are unlike the WFD's standard visits in Section \ref{ssec:dmplans_NSV}; incorporate Special Programs data into the Prompt and Data Release pipelines and data products for the WFD main survey in Section \ref{ssec:dmplans_WFD}; reconfigure the pipelines and generate unique sets of data products for each Special Program in Section \ref{ssec:dmplans_reconfig}; and enable user-generated pipelines and data products in Section \ref{ssec:dmplans_user}.
 
 % % % % % % % % % % % % % % % % % %
-\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:dmplans_NSV}
+\subsection{Processing Diverse Imaging Data Unlike the WFD Main Survey}\label{ssec:dmplans_NSV}
 
-A standard visit image is considered to be two snap exposures of $\sim15$ second each, and in the WFD main survey sequential visits will be of different fields.
-As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field.
-In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, \citeds{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).
+As described in Section \ref{sec:data}, Special Programs will be able -- and likely -- to request observing modes with shorter (or longer) exposure times, long sequences of visits to the same field, and/or imaging of very crowded fields.
+We review the capability of DM's planned pipelines to process such each of these kinds of diverse data, keeping in mind that the processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters such as the number of stars with sufficient flux for photometric calibration or the source area density (crowding).
+Furthermore, that these boundaries imposed by the data quality might not be constrained until the final performance of DM's algorithms, as described in the Data Management Applications Design (\citeds{LDM-151}) document, is fully characterized.
 
@@
 
-Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-We consider shorter and longer exposure in turn.
+Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+We discuss shorter and longer exposure in turn.
 
 \textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
+The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) is under consideration (see JIRA ticket DM-12574).
 Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
 A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
+However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
+As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
+The minimum exposure time for an image to be successfully processed by the Difference Imaging Analysis (DIA) pipeline is currently under consideration (see JIRA ticket DM-12574).
 
@@
 However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+A potential maximum exposure time from a processing perspective is currently under consideration (see JIRA ticket DM-12574).
 
-In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
+% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
 One possibly useful study is Chang et al.
@@
 2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} to at least simply characterize the PSF stability as a function of exposure time.
-
-In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
-
-Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
+% In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
+% Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
 This is probably best left ot a policy issue (beyond the scope of this document).
 
-\subsubsection{Image Acquisition Rate}
+\subsubsection{Number of Exposures per Visit (Long Sequences of a Single Field)}
 
-OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds.
-However, \textit{if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated \texttt{DIASources}.
+There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
+From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.
 
-\textbf{Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds.
-This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later?
-\textbf{How long does it take to save an image to disk?}
-
-\textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
-(Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
-However, at NCSA there should be ways to elastically change the amount of processing power available.
-This is also an issue for crowded fields (below).
-
-\textbf{Unassociated \texttt{DIASources}:} the consistency of \texttt{DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field).
-In other words, it is not guaranteed that the next string of processing would have access to the latest \texttt{DIAObjects} for association.
-It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations.
-This is also an issue for crowded fields (below).
-
-\subsubsection{Exposures per Visit, and Long Sequences of a Single Field}
-
-K.-T.
+% K.-T.
 Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
@@
 
-K.-T.
+% K.-T.
 has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
@@
 
-\subsubsection{Alerts from Crowded Fields}
+\subsubsection{Images in Very Crowded Fields}
 
-Special Programs are more likely to include crowded fields than the WFD main survey area.
-Due to the increased number of sources, the number of \texttt{DIASources} -- and therefore the number of Alerts -- increases as well.
-In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
-A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete.
-K.-T.
-reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed.
-This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).
-
-\subsubsection{Summary}
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey).
-For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science.
-After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018.
-(1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA?
-In the latter case, this will probably be limited by the time needed to generate a well-formed PSF.
-(2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds.
-Might this requirement be slackened to accommodate a higher cadence Special Program?
-If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping \textit{vs.} delaying Alerts during observations with a fast ingest cadence.
-These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts.
-Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF).
-We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits.
-Consequential images of the same field may lead to unassociated \texttt{DIASources} of the same \texttt{DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately.
-\end{enumerate}
+The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
+The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields.
+These may or may not be appropriate for Special Programs data, depending on the science goals.
 
 % % % % % % % % % % % % % % % % % %
-\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
+\subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:dmplans_WFD}
 
-DM will want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
-In the following three sections we summarize when and how Special Programs data might be incorporated into the pipelines and products of the Difference Imaging Analysis and Alert Stream, the Data Release Pipeline products, and/or MOPS.
+As described in Section \ref{sec:intro}, the Project may incorporate Special Programs data into the WFD main survey's pipelines and data products whenever this is (1) possible and (2) scientifically beneficial.
+This will likely be at the discretion of the data quality assessment team during Operations.
+In the following three sections we project when and how Special Programs data might be incorporated into the pipelines and data products of the Prompt pipeline and the Alert stream (Section \ref{ssec:dmplans_prompt}) and the annual Data Release pipeline (Section \ref{ssec:dmplans_drp}).
 
-\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
+\subsubsection{The Prompt Pipeline and Alert Generation}\label{ssec:dmplans_prompt}
 
-It is beneficial to transient science to include as many LSST images into the Level 1 Difference Imaging Analysis and Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST (this prospect is discussed further in Section \ref{ssec:dmplans_L3}).
-Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
-DM's ability to process the potential diversity of image types from Special Programs is explored above in Section \ref{ssec:dmplans_NSV}, and we continue the discussion of incorporating processed images into the Level 1 DIA pipeline.
+It would be beneficial to transient science to include as many LSST images into the Difference Imaging Analysis (DIA) pipeline and the Alert Stream as possible.
+Only images that can be processed with the Prompt DIA pipeline with the $60$-second timeline can contribute to the Alert Stream.
+As discussed in Section \ref{sec:data}, this might prohibit exposures shorter than $<15$ seconds and/or visit cadences shorter than $1$ per $\sim30$ seconds.
+This might also prohibit the inclusion of very crowded fields that require more computational resources.
+A field must have an LSST template image to be processed by the DIA pipeline, which would prohibit immediate Alerts from new survey fields.
+It should be possible to load and use an alternative template image (than what would be used for that field if and when it is covered by the WFD main survey) in the Prompt pipeline for fields from a Special Program.
 
-\textbf{Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
+There may be a couple of issues encountered with Alerts from many consecutive visits of Deep Drilling fields.
+One is that, since the Alert contains the full record of all associated {\tt DIASources} from the past 12 months (\citeds{LSE-163}), for a Deep Drilling Field with significantly more visits over the year, the size of the Alert might become prohibitively large (TBD).
+Another is that the self-consistency of the \texttt{DIAObjects} catalog may suffer during consecutive visits of a single field.
+For example, the processing for image $2$ of a sequence would begin when the processing for image $1$ is only halfway complete.
+Any new {\tt DIASource} in image $1$ that cannot be associated (by coordinate) with an existing {\tt DIAObject} becomes a new {\tt DIAObject}.
+When this source is again detected in image $2$, another new {\tt DIAObject} would be created if the catalog has not yet been updated.
+The best solution might be to flag {\tt Alerts} which may suffer from incomplete {\tt DIASource}--{\tt DIAObject} associations (see JIRA ticket DM-12574).
+
+% \textbf{Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
 So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine.
@@
 
-\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
+% \textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
+(Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
+However, at NCSA there should be ways to elastically change the amount of processing power available.
+This is also an issue for crowded fields (below).
 
-The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries).
-Since MOPS takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
-There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
+% Special Programs are more likely to include crowded fields than the WFD main survey area.
+Due to the increased number of sources, the number of \texttt{DIASources} -- and therefore the number of Alerts -- increases as well.
+In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
+A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete.
+K.-T.
+reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed.
+This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).
 
-\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
+{\bf The Moving Object Processing System (MOPS) -- } Since MOPS takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
+As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:data_prev}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
+%There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
 
-Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP \texttt{Objects} database that is built from a set of images with nearly constant depth and cadence.
-When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
-For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
-In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey.
-This is mostly discussed here in context with the \texttt{Source} and \texttt{Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.
+\subsubsection{The Data Release Pipeline (DRP)}\label{ssec:dmplans_drp}
+
+This document is not the place for a full consideration of whether or not it would be ``scientifically beneficial" to include any Special Programs data in the DRP data products -- namely, the deep image CoAdds and their corresponding {\tt Source} and {\tt Object} catalogs (\citeds{LSE-163}) -- and we leave that decision for the data quality assessment team in LSST Operations.
+One example might be when Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey.
+Another may be if including some or all of the shallower Galactic Plane coverage suppresses edge effects or low-order modes in the all-sky photometric solutions.
 
 % % % % % % % % % % % % % % % % % %
-\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
+\subsection{Reconfiguring DM Pipeline Components for Special Programs}\label{ssec:dmplans_reconfig}
 
-LSST DM intends to reconfigure its existing pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
+Whether or not Special Programs images are incorporated into the WFD main survey's data products, it is anticipated that most of the Special Program's science goals will require (or benefit from) separate data products (i.e., CoAdds and/or catalogs).
+For this reason, LSST intends to reconfigure the DM's pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
 In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
@@
 For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the \DPDD, \citedsp{LSE-163}).
-This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
+This type of reconfiguration would also be possible to create as a user-generated pipeline (Section \ref{ssec:dmplans_user}), but having these products provided by the Project ensures a consistent and verified level of quality.
+%, as well as access for all users to the processed Special Programs data products, which would only increase the scientific value of Special Programs data.
+--> MLG: data rights doc will probably state that the products for SP data that is only processed by user-generated pipelines will have to be served to all users
 
-The current outstanding issue related to DM's processing of special programs data is that the work-hours needed to reconfigure and test the pipelines, run them and verify the data products for public release -- potentially on intermediate timescales that do not coincide with the prompt/yearly Level 1 and 2 timescales, e.g., monthly stacks of deep drilling fields -- have not been included in the personnel budget and thus, at the moment, constitute an expansion of scope.
-As we describe in Section \ref{ssec:docrev_dmsr} there are requirements in the DMSR that state that DM does have the responsibility to process Special Programs data.
+The above statements of intent are derived from the Data Management Subsystem Requirements document, \citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
+To ensure that the work-hours needed to reconfigure and test the pipelines, and run them and verify the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), have been included in the personnel budget, JIRA ticket DM-12575 is currently under consideration.
 
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to clarify the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations.
-Questions to be answered include, e.g., product ownership of the reconfigured pipelines and test data products, and how do we set goals and deadlines for this?
-\end{enumerate}
-
-As described in Section \ref{ssec:docrev_dmsr}, the DMSR contains requirement DMS-REQ-0320, which states that "it shall be possible for special programs to trigger their own data processing recipes".
-Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.
-
-Also described in Section \ref{ssec:docrev_dmsr} is the DMSR requirement that ``it shall be possible for special programs to trigger their own data processing recipes" (DMS-REQ-0320).
-However, there is currently no requirement for the scheduler to initiate non-standard processing with a certain pipeline if the data was executed under a given program.
-Since the data processing is scheduler-driven, this would have to be an OSS requirement.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-2} The purpose of this ticket is to figure out whether and how auto-triggering of alternative data processing recipes for Special Programs can be supported by the DM system.
-If it can't be, a change request on DMS-REQ-0320 might be needed to down-scope.
-\end{enumerate}
+DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
+A header keyword identifying an image as related to a Special Program would be sufficient to send it to a dedicated processing pipeline, and would satisfy this requirement.
+JIRA ticket DM-12576 is currently open to make sure that this happens.
 
 % % % % % % % % % % % % % % % % % %
-\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_L3}
+\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_user}
 
-As explained above, LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs whenever possible.
-In Section \ref{ssec:docrev_sizing} we describe how LSST has allocated 10\% of its computational and storage capacity for this processing because 10\% of the images will be generated by Special Programs.
+In cases where the science goals of a Special Program require specialized algorithms and cannot be achieved by reconfiguring DM's software, then user-generated pipelines will be needed.
+Towards this end, LSST DM is making all of its software open-source, and preparing the Science Platform (\citeds{LSE-319}), through which users can access the tools and computational resources to assemble data processing pipelines to achieve their science goals (whether related to Special Programs data or not).
+During Operations, there will be a method for the system to allocation processing resources in the case of over-subscription.
+JIRA ticket DM-12577 is currently open to inspire an investigation of whether additional Science Platform capabilities are needed to enable user-driven processing of Special Programs data.
 
-However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3).
-LSST has furthermore allocated an \textit{additional} 10\% of its computing resources for user-driven processing and analysis.
-This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).
-
-In the case where a Special Program requires an intensive amount of additional computational resources (i.e., too large a fraction of the 10\% allocated), the processing will have to be done on a separate, external computational resource.
+If the user-generated processing pipeline for Special Programs data requires requires significantly more computational resources than have been allocated -- where that allocation has been sized approximately, based on image processing for WFD main survey data (i.e., difference imaging, source detection, and/or stacking) -- then external computational resources may be necessary.
 To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
 
-In the case where user-driven processing pipelines consume a small amount of that 10\%, data export and processing on external computation resources will not be necessary.
-Users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their pipelines on LSST servers, monitor their progress, and analyze their output.
-DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
-
-However, in the case where user-driven processing pipelines consume a \textbf{large} amount of that 10\%, data export and processing on external computation resources \textbf{will} be necessary.
-The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.
-
-\textbf{Level 3 Alert Stream?
---} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package.
-K.-T.
-raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.
-
-\textbf{Lower Latency for Data Access?
---} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner.
-Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-1} The purpose of this ticket is to ask Gregory about whether there are aspects to the Science Platform that need to be developed to better serve the needs of the community doing Special programs processing.
-Examples include support for Level 3 alert streams and a lower latency for access to processed visit images.
-\end{enumerate}
+No user-generated pipeline may contribute Alerts to the Alert Stream, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
 
 It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
-For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
-An alternative option to ``adopted" code is ``adopted" data products: situations in which Level 3 code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-2} The purpose of this ticket is to clarify whether there is a process in place, and if staffing needs have been defined, for adopting Level 3 code into DM's pipelines, and/or for adopting data products, verifying them, and hosting their public release.
-This issue is not limited to Special Programs, but is likely to be encountered in that context.
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % %
-\clearpage
-\section{DM Documentation Review}\label{sec:docrev}
-
-Here we gather every requirement or specification relevant to Special Programs in the existing DM documentation, identify areas where the intent of LSST DM is not adequately represented, and suggest clarifications, changes, or additions to these documents.
-
-% % % % % % %
-\subsection{Science Requirements Document, SRD, \citeds{LPM-17}}\label{ssec:docrev_srd}
-
-This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
-The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4).
-The latter we know to be an approximate estimate.
-
-% % % % % % %
-\subsection{LSST System Requirements, LSR, \citeds{LSE-29}}\label{ssec:docrev_lsr}
-
-The LSR is derived from the SRD, and converts the high-level specifications from the SRD into system requirements, which flow-down to and in some cases repeated in the OSS and DMSR (next sections).
-We have the version of LSE-29 last revised on August 4, 2016.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-0} RFC (LSE-29): We propose to add a high-level requirement that DM reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, to better reflect its intent.
-This would flow-down to the DMSR.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be \texttt{DRT1} = 1 year (1.4.2, Data Release Processing).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-1} RFC (LSE-29): We propose to add a requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102).
-E.g., that data products from reconfigured pipelines will be made available on a intermediate time scale where the science goals require it.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of \texttt{minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
-The full capabilities of DM to process non-standard visits is not yet well defined; see \ref{NSV-1}.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-2} RFC (LSE-29): We propose that LSR-REQ-0111, the requirement for DM to be capable of processing exposures that were not taken in a standard visit mode, might need to include a caveat once the DM capabilities for processing non-standard visits are truly defined (e.g., extremely short exposures).
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as "Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-3} RFC (LSE-29): We propose to add to LSR-REQ-0032, which requires that the DM system provide three classes science data products (nightly, yearly, and user-specified), the requirement to provide intermediate-cadence data products for Special Programs, whenever possible, to enable science.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products "of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{LSR-4} RFC (LSE-29): We propose to add to LSR-REQ-0041, which requires that LSST provide software, services, and hardware resources to support Level 3 data products, the caveat that there will be technical limits on DM's ability to provide this in cases where an an intensive amount of additional computational resources is required.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0042 requires that LSST produce the data products necessary to support the four primary science missions (dark energy, solar system, transients, milky way; 2.4.5.2, Science Flowdown).
-Since the science goals of the Special Programs flow down from these four primary science goals, no change is needed here.
-
-$\bullet$ LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
-This is discussed further in Section \ref{ssec:docrev_sizing}.
-
-$\bullet$ LSR-REQ-0075 defines that the WFD main survey science objectives will be met with 90\% of the observing time, with the remainder left to Special Programs (Section 3.3.1, Survey Time Allocation)
-
-% % % % % % %
-\subsection{Observatory System Specifications, OSS, \citeds{LSE-30}}\label{ssec:docrev_oss}
-
-The OSS contains the requirements on the site, facility, and camera that are necessary to meet the requirements specified by the LSR (LSE-29) and the SRD (LPM-17).
-We have the version of LSE-30 last revised on February 10, 2017.
-Some of the aspects discussed in this section are not directly relevant to Data Management and are probably outside the scope of this study, but we keep them because they are relevant to Section \ref{ssec:data_bounds}.
-
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, \texttt{nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies.
-\end{enumerate} % Assign to Zeljko.
-
-$\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (OSS 2.1.1.7, Visit Optimization).
-
-$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively.
-These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
-These estimates appear adequate for Special Programs in general.
-
-$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively.
-These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.
-
-$\bullet$ Section 3.1.5 (OSS-REQ-0125 and those following) set the requirements for data products, including their variety and format (i.e., images and catalogs in Level 1 and 2), their precision (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (Alerts, Level 1, and Level 2 ).
-Would these also apply to data from Special Programs?
-For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be \texttt{falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-3} RFC (LSE-30): We propose to add a statement to Section 3.1.5 of the OSS that DM will make a ``best effort" to meet the data products requirements with the Special Programs data.
-\end{enumerate}
-
-$\bullet$ Section 3.5, Photometric Calibration, puts requirements on e.g., the relative contributions to photometric errors from instrumental (OSS-REQ-0282) and atmospheric (OSS-REQ-0276) transmissions, and requires that a catalog of $10000$ reference stars of $17<r<20$ mag per field be created for use (OSS-REQ-0285).
-The latter might not apply to Special Programs fields, but this probably isn't cause for concern.
-
-$\bullet$ OSS-REQ-0319 sets the requirement that the LSST be capable of continuous operation throughout the night when all visits are separated only by the readout time, thereby enabling a deep drilling field style of observations (OSS 3.6.1.3, Continuous Exposures).
-
-$\bullet$ OSS-REQ-0291 defines the minimum exposure time as $1$ second with a stretch goal of $0.1$ seconds, with a note that the spacing between exposures should be lengthened to $15$ seconds in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times (OSS 3.6.1.4, Minimum Exposure Time).
-The implication for DM is that an exposure ingest rate will never exceed one per $15$ seconds.
-
-$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be $\texttt{tFilterChange} = 120$ seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
-There is no officially required minimum time between filter changes.
-Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night (see Section \ref{ssec:data_bounds}).
-
-$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking (\texttt{rotTrackTime} = 1 hour) and the half-range of the rotator motion (\texttt{rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
-Questions about whether there is a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), are raised in Section \ref{ssec:data_bounds}.
-
-$\bullet$ OSS-REQ-0380 sets the rate and error limits for nonsidereal tracking, which presumably would only be relevant to a Special Program (3.6.3.7, Non-Sidereal Tracking).
-
-% % % % % % %
-\subsection{Data Management Subsystems Requirements, DMSR, \citeds{LSE-61}}\label{ssec:docrev_dmsr}
-
-The DMSR contains the top-level requirements for all of Data Management, including processing pipelines and products (revision 2017-08-11).
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-0} RFC (LSE-61): We propose that the DMSR include a high-level specification (probably in Section 1.6) that the data from Special Programs will be included in the products of the wide-fast-deep main survey only when it is (a) possible and (b) beneficial to the primary science objectives of those products.
-It should also be updated to reflect the fact that DM will reconfigure its own pipelines ``wherever possible", process the Special Programs data, verify the unique data products, and make them publicly available.
-This would be a flow-down requirement from a similar new one added to the LSR (LSE-29).
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0068 requires that each raw science image store metadata regarding the date, time, site, telescope, and camera -- missing from this might be the scheduler and program information.
-That kind of metadata will be necessary to identify images related to Special Programs (1.2.3, Raw Science Image Metadata).
-Related to this is DMS-REQ-0266, which requires the creation of an Exposure Catalog that also stores this kind of metadata independently, but also does not include scheduler and program information.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-1} RFC (LSE-61): We propose that DMS-REQ-0068 be updated to include that the scheduler and program metadata (i.e., a way to identify an image as being associated with a Special Program) with the raw science image, and/or with the Exposure Catalog (DMS-REQ-0068, -0266).
-We furthermore propose that the \texttt{visitID} and the total number of exposures in the visit are also included in this updated requirement (e.g., to accommodate a deep drilling sequence of a single visit of 50 exposures).
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
-It also says that "this aspect of the processing for Special Programs data is specific to each program".
-It should be specified whether "this aspect" refers to the reduction or archiving of the processed visit images.
-If it refers to archiving, and if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
-(1.2.2, Processed Visit Images).
-This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
-We assume this applies to standard visit images that \textit{can} be processed.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-2} RFC (LSE-61): We propose that DMS-REQ-0069 be clarified regarding whether processed single visits and/or difference images for Special Programs can stay archived and accessible to e.g., Level 3 pipelines, for a longer amount of time, possibly by user request.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0274 sets the content of an Alert, and this does not currently include scheduler or program information.
-Adding this would enable users to e.g., use the LSST mini-broker to filter for only Alerts from their desired Special Program, in case they have a dedicated follow-up resource.
-Furthermore, it might facilitate follow-up more generally if users know e.g., that this Alert is from a field that is going to be observed again in $X$ minutes or $Y$ times in a night (1.3.13, Alert Content).
-However, the latter might be adequately available through the predicted visit schedule server specified by DMS-REQ-0353.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-4} RFC (LSE-61): We propose that DMS-REQ-0274, which sets the content of an Alert, be updated to include the program information and/or predicted revisit schedule (unless it is adequately available through the predicted visit schedule server specified by DMS-REQ-0353).
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
-Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.
-This is handled by \ref{reconfig-2}.
-(1.6.1 Processing of Data From Special Programs).
-
-$\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
-It should be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within \texttt{OTT1} = 1 minute means contributing to the L1 Alert Stream.
-(1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-6} RFC (LSE-61): We propose that DMS-REQ-0321 and -0344, which specify that the Level 1 processing for data Special Programs shall be completed with the same latencies as applied to data from the WFD main survey, be reworded to clarify that only Special Programs data that \textit{can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products), will be incorporated into Level 1 processing and contribute to the Alert Stream.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0322 specifies that data products from Special Programs processing shall be stored in separate (but joinable) databases.
-
-$\bullet$ DMS-REQ-0312 and -0313 require that DM maintain a live L1 and the two most recent DR L1 and L2 catalogs for access by users, but there is no corresponding access requirement for the data products of Special Programs (4.1, Data Archive).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{DMSR-7} RFC (LSE-61): We propose that there be a requirement for maintenance of user access to data products of Special Programs with the same timescales as the WFD main survey data; i.e., similar to DMS-REQ-0312, -0313.
-\end{enumerate}
-
-%$\bullet$ The following items are applicable to users processing data from Special Programs and/or reprocessing data from the WFD main survey.
-Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
-Section 2.9 "Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
-Sections 3.1 "Software Architecture to Enable Community Re-Use" and 3.2 "Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
-In Section 3.3 "Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
-
-% % % % % % %
-\subsection{Data Management Applications Design, DMAD, \citeds{LDM-151}}\label{ssec:docrev_dmad}
-
-The DMAD specifies the design and implementation of the code and algorithms that will be used in the processing of images and creation of catalogs in the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-19.
-As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
-In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
-It is the DMAD's codes and algorithms that DM will be \textit{reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
-This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
-
-The purpose of this work on Special Programs is not to suggest any changes to the DMAD.
-Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3.
-
-\subsubsection{Extreme-Depth CoAdds} The system has been sized to hold $\sim200$ exposures in memory at once, which defined by the current maximum number of visits per field in the WFD main survey in $10$ years (from a conversation with K.-T.~Lim).
-Note that the panchromatic CoAdds would be built from the individual filter CoAdds, so the algorithm does not need to handle $\sim800$ images.
-From a computational standpoint, $200$ is the maximum number of images that can be stacked with an algorithm that requires all images to be accessible in memory at once (i.e., loading all images and calculating the median for each pixel).
-Deeper stacks might be possible with algorithms that deal with images consequentially.
-It is conceivable that a Special Program which needs to stack $>200$ images is not possible to accomplish with reconfigured pipelines, and would have to be processed with external, user-contributed resources.
-However, the exact DM capabilities in this area are not yet well known because NCSA has not yet defined the machine capabilities.
-Furthermore, the planned commissioning data will go to a $\sim20$ year depth, and so it can reasonably be expected that DM will have to be able to accommodate at least a stack that deep.
-
-\subsubsection{Deblending} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
-It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
-Level 3 deblenders for specific Special Programs fields may require development by the user community.
-
-\subsubsection{Variability Characterization} The periodic and aperiodic variability characterizations described in Section 6.21 of the DMAD are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
-From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
-It is conceivable that the goals of a particular Special Program might require different algorithms; these could be provided by DM, or written as Level 3 and either made joinable to the DM reconfigured data products or perhaps incorporated directly.
-
-\subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP \texttt{Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM.
-It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.
-
-% % % % % % %
-\subsection{Data Products Definitions Document, DPDD, \citeds{LSE-163}}\label{ssec:docrev_dpdd}
-
-The DPDD describes the data products -- the images and catalogs -- to be delivered from the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-01.
-Section 6 describes the data products that DM will provide for Special Programs.
-It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster.
-
-The LSST Database Schema Browser\footnote{\url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}} is another way to explore the contents of the planned DM catalogs.
-We identify a couple of potential changes that might be necessary to the schema, which are related to this study on Special Programs.
-These are not high priority.
-
-$\bullet$ The database schema for \texttt{DIASource} does not appear to have an element that identifies which template image that was used.
-This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
-DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
-In conversation with K.-T., we find that this isn't a problem, as it will be handled by provenance: the code configuration used and the time of processing are sufficient to identify and regenerate the template image.
-However, K.-T.
-has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable.
-However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.
-
-$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
-It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?
-\\
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-1} The purpose of this ticket is to clarify whether \texttt{Object.prv\_inputID} in the database schema will identify whether an \texttt{Object} is an externally provided coordinate.
-\end{enumerate}
-
-$\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
-\texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-\texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
-Are 32 floats in each of the 6 filters always going to be a large enough volume?
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-2} The purpose of this ticket is to ask if DM should consider whether the \texttt{float[6x32]} for the characterization parameters for variability is of adequate size to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-This study might be similar to e.g., DMTN-049 for photometric redshifts included in the DRP \texttt{Object} catalog.
-\end{enumerate}
-
-% % % % % % %
-\subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}
-
-LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
-Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}.
-Regarding the reduction and calibration of non-standard visit images, LSE-180 makes two relevant points: \\
-$\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
-$\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
-
-However, Lupton's recent work on calibrations has made much of LSE-180 obsolete, and it is not clear whether it is still needed as a separate document.
-At the time of writing, Lupton's most recent take on calibrations can be found in \url{https://github.com/lsst-dm/calibration/blob/master/calibration.pdf}.
-K.-T.
-referenced the Appendix B for the overview of calibration types, but it seems that there is not yet a final plan that can be assessed for its suitability for Special Programs.
-It is plausible that Special Programs could have their own calibration database, and obtaining additional calibration frames is not expected to cause a bottleneck (at the moment there is $\sim1.7$ hours per day for this, and exceeding this could eat into the engineering time).
-The most likely cause for trouble involving calibrations and Special Programs is the computational time needed to create the additional and/or different calibration files to be applied to non-standard Special Programs data, and/or the additional overhead required to load them in the Level 1 pipeline if the schedule interleaves SP-WFD-WFD-SP.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Cal-1} The purpose of this ticket is to define the potential additional calibration needs of Special Programs data, which might be significantly more diverse in terms of, e.g., exposure time and/or sky background levels in the case of a a twilight survey of short exposures.
-Questions include how much additional day time might be needed to obtain raw calibration frames, or the additional computational resources needed to create and/or load calibration frames or catalogs during processing.
-The goal of this would be to confirm that the DM system prepared to calibrate the diversity of data that might be generated by Special Programs.
-\end{enumerate}
-
-% % % % % % %
-\subsection{Computational Resources Sizing Documentation}\label{ssec:docrev_sizing}
-
-In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will "use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
-This is based on the estimate that Special Programs data comprise $\sim10\%$ of the total observing time.
-Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
-Based on this predicted multiple reprocessing, we might ask whether the "10\% of the computing resources" allocated for DM processing of Special Programs data is too small.
-This will depend on the exact final mix of Special Programs, which may well change from year to year, and furthermore the current error bars on the sizing are of order $\pm10\%$.
-
-The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
-LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
-DM Compute Sizing spreadsheet (LDM-138) and explanation document (LDM-140) estimates the total compute resources needed for all processes, quoting the same 10\% addition for Level 3.
-DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
-As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-It is important to note that this 10\% is \textit{in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that \textit{an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
-\end{enumerate}
-
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks \textit{"whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
-Have these systems been sized?"}.
-The answer to this comes from K-T: \textit{In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
-There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can ``borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-2} The purpose of this ticket is to ensure that the overall DM computational system is appropriately sized, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, \textit{and} incorporated into the Level 2 data products, \textit{and} have their own dedicated reconfigured pipelines and separate products, \textit{and} have a higher probability for user-defined processing.
-The latter would apply to the size of the DACs.
-Although an accurate estimate of the science-driven DAC size might not be truly possible until after the call for Special Programs white papers, an initial attempt must be made because this call will need to define a computational boundary for what can be done in the DAC, and what will require the proposer to bring additional resources to the table.
-For example, if the fractional computational needs are equivalent to the fraction of total data that the program acquires, that is acceptable; but if the program requires, e.g., $>2\times$ as much computational resources, then an external contribution might be necessary.
-Defining what the call will say regarding the need for external processing is a goal of this ticket.
-\end{enumerate}
-
-% % % % % % %
-\subsection{The Limited Documentation Regarding Image Detection Efficiencies}
-
-Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
-Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
-What, exactly, are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
-Note that here we are talking about planting fakes in the single images for transient detection efficiencies with difference imaging analysis, and not planting fakes in the CoAdds for point-source limiting magnitudes).
-
-$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that \textit{"Each processed image ..
-will record information on the pixel variance ...
-as well as the per-pixel masks ...
-These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
-
-$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): \textit{"The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
-detection efficiency for point sources vs.
-mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
-
-$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step \textit{"is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
-The details of what metrics will be mapped, the format and scale of the maps (e.g.
-hierarchical pixelizations vs.
-polygons), and the way the metrics will be computed are all unknown"}.
-It also states that this must be extendable to Level 3, but that \textit{"the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
-(Section 3 "Alert Production" also specifies that in LDM-151 \textit{"we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
-Source detection thresholds can be estimated through the use of sky sources"}.)
-
-$\bullet$ Suchyta et al.
-(2016) presents \texttt{balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses \texttt{galsim} to embed shapes built of e.g., Sersic profiles).
-
-$\bullet$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
-
-$\bullet$ However, K.-T.
-revealed that a computational budget of 40\% has been allocated for fake insertion, with the understanding that this would happen in the daytime (i.e., not within 60 seconds).
-However, there has been a failure to schedule this into the development plan and there are no milestones for this project, and no person assigned to this task.
-
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DE-1} The purpose of this ticket is to resolve whether or not the implantation of fake sources will be done as part of the Level 1 detection efficiency characterization, or if an alternate method will be sufficient for the science needs.
-K.-T.
-says the computational system is sized to accommodate it.
-If this issue remains unclear, perhaps a study that assess the options, their computational costs, and their scientific payoff should be done so that DM can better assess the need for fake sources and allocating staffing hours as necessary.
-Note that this is not a question limited to Special Programs.
-\end{enumerate}
+For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of the Prompt and DRP data products.
+An alternative option to ``adopted" code is ``adopted" data products: situations in which user-generated code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
+JIRA ticket DM-12578 is currently open to consider the staffing needs and process for adopting user-generated code or products; this issue is not necessarily limited to Special Programs.
 
@@
 \clearpage
-\section{Special Programs and the Science Community}\label{sec:SP_SC}
+\section{Special Programs Processing Case Studies}\label{sec:SPCS}
+
+For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
+This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
+Note that we are not including any analysis in these descriptions, only processing and products.
+These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
+
+Basic steps that we use to describe a processing case study: \\
+Step 1. Data Acquisition. \\
+Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
+Step 3. Delivery of LSST Processed Images. \\
+Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+Step 6. User-Generated Pipelines and Products. \\
+
+% % % % % % %
+\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
+
+This Special Programs processing summary is based on Becker et al. (2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
+
+Step 1. Data Acquisition. \\
+The observational sequence is triggered.
+In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
+This sequence is always repeated 2-3 nights later.
+This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
+Data obtained in the $g$-band filter is also acceptable.
+\citedsp{Document-11013}
+
+Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
+Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
+
+Step 3. Delivery of LSST Processed Images. \\
+The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
+
+Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+Shift-and-stack processing is beyond the scope of DM's algorithms.
+
+Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
+However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.
+
+Step 6. User-Generated Pipelines and Products. \\
+The user-generated pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external processor.
+Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
+The DRP difference imaging routine will be used with the same template tract/patch for all.
+Custom, user-generated algorithms will shift the exposures and create difference images, then DRP routines can stack and do source detection and characterization and generate an object database.
+Custom code will derive orbital parameters for the detections and add them to a {\tt SSObjects}-like database.
+
+% % % % % % %
+\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
+
+Step 1. Data Acquisition. \\
+On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
+
+Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
+Each $2\times15$ second visit is processed by the Prompt pipeline's DIA, and Alerts are released within 60 seconds.
+They are flagged to denote the image source is a DDF and that source association might be compromised.
+
+Step 3. Delivery of LSST Processed Images. \\
+The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+
+Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+The required data products for this science goal can be met by reconfiguring the DM pipelines.
+First, a template image for the field will be made using DM stacking algorithms.
+On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF.
+The LSST codes for alert packet and transport could be used to distribute the detected objects e.g., to the same brokers that receive the Alert Stream, or alternative destinations.
+However, these packets would not be distributed via the LSST {\tt Alert Stream}, and would need to be identified as, e.g., DDF Alerts.
+Note that JIRA ticket DM-12585 is currently open to investigate whether or not the internal real/bogus routine be able to run on a nightly CoAdd of deep drilling difference images.
+
+Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
+
+Step 6. User-Generated Pipelines and Products. \\
+For the science goal of searching for supernovae in nightly stacked DDF images, no separate user-generated software appears necessary.
+
+% % % % % % %
+\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
+
+Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
+The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
+
+Step 1. Data Acquisition. \\
+At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. Location and exposure times are set by the sky brightness and desired saturation limits.
+
+Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
+Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:dmplans_NSV}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.
+
+Step 3. Delivery of LSST Processed Images. \\
+Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+
+Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+This is officially not determined, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
+
+Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+These short-exposure, high sky background images would not contribute to the DRP data products created for the WFD survey.
+
+Step 6. User-Generated Pipelines and Products. \\
+If short-exposure images cannot be processed with the existing DM algorithms, a user-generated processing pipeline might be needed to reduce the raw data.
+
+Side note: A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
+This would be considered a user-generated algorithm.
+
+% % % % % % %
+\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
+
+Step 1. Data Acquisition. \\
+The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields (or shorter, for a shallower depth than the WFD main survey).
+
+Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
+Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
+Extremely crowded fields might have to be skipped if they take longer to process and violate the $60$ second latency for Alerts.
+
+Step 3. Delivery of LSST Processed Images. \\
+The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+
+Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Prompt and DRP pipelines, so it seems that not much reconfiguration would be needed.
+The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
+
+Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
+
+Step 6. User-Generated Pipelines and Products. \\
+It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
+
+%\input{spcs_sas}
+
+\clearpage
+\bibliography{local,lsst,refs,books,refs_ads}
+
+\clearpage
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+\appendix
+
+% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
+\section{Previously Proposed Special Programs}\label{sec:data_prev}
+
+In this section we compile information about the science goals and observational methods for Special Programs that have been previously proposed or discussed in the Science Community.
+We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science.
+The main resources from which we have collected information about the Community's Special Program are: \citep{2008arXiv0805.2366I}; \citep{LPM-17}; the LSST Deep Drilling Field white papers from 2011\footnote{\url{https://project.lsst.org/content/whitepapers32012}}; presentations by Niel Brandt and Stephen Ridgway at the LSST Project and Community Workshop in August 2016\footnote{\url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} and \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}}; \citep{2013arXiv1304.3455G}; and Chapter 10 of \citep{2017arXiv170804058L}.
+
+%\begin{itemize}
+%\item "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \citep{2008arXiv0805.2366I} 
+%\item The LSST Science Requirements Document (\SRD), \citeds{LPM-17} 
+%\item LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} 
+%\item "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf}
+%\item "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}
+%\item "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} 
+%\item Chapter 10 "Special Surveys" of the Observing Strategy White Paper \citep{2017arXiv170804058L}
+%\end{itemize}
 
 So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
-\citep{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
-Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}).
-We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.
-
-All other aspects of LSST Special Programs remain open to proposals, including: \\
-$\bullet$ additional deep drilling fields \\
-$\bullet$ refined observing strategies for deep drilling fields \\
-$\bullet$ optimized survey areas for the NES, South Pole, and Galactic Plane \\
-$\bullet$ refined observing strategies for the NES, South Pole, and Galactic Plane \\
-$\bullet$ additional mini-surveys \\
-$\bullet$ the total fraction of time spent on Special Programs (nominally 10\%)
-
-% % % % % % % % % % % % % % % % % %
-\subsection{Currently Proposed Special Programs} \label{ssec:data_science}
-
-In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed.
-We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science.
-In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
-Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an \textit{incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
-
-\noindent \textbf{Resources:} \\
-$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \citep{2008arXiv0805.2366I} \\
-$\bullet$ The LSST Science Requirements Document (\SRD), \citeds{LPM-17} \\
-$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
-$\bullet$ "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
-$\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} \\
-$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \citep{2017arXiv170804058L}
+There are three mini-survey areas that have been discussed extensively by the Science Community: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}).
+In Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified, along with an \textit{incomplete} list of potential mini-surveys that have been openly discussed in the Science Community.
+In Section \ref{sec:SPCS}, we create detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
 
@@
 \begin{footnotesize}
-\caption{Approved DDF and Incomplete List of Potential MS.}
+\caption{Approved DDF and Incomplete List of Potential Special Programs.}
 \label{tab:ddfms}
@@
 DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
-DDF TBD  & & TBD \\
+\hline
 North Ecliptic Spur      & & solar system objects (find and characterize) \\
@@
 
-\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \citep{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
-Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
-(Note that this nominal survey is missing the overhead for readout and shutter, 3 seconds per exposure, which means that the LSST capability is closer to 43--45 exposures per hour, minus the time for 3--4 filter changes -- not 50 exposures per hour.)
+Here we consider a variety of scientific fields in turn, the Special Programs that have been discussed in that Science Community so far, and the implications of these Programs for the diversity of data and data products.
+Generally, the types of LSST Special Programs that are open for proposals include: (i) additional deep drilling fields; (ii) refined observing strategies for deep drilling fields; (iii) optimized survey areas for the NES, South Pole, and Galactic Plane; (iv) refined observing strategies for the NES, South Pole, and Galactic Plane; and (v) additional mini-surveys (areas and observing strategies).
 
 \medskip
-\noindent \textbf{Solar System Objects (SSO)}\\
-$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
-$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \citep{2017arXiv170804058L}) \\
-$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \citep{2017arXiv170804058L}) \\
-$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \citep{2017arXiv170804058L}) \\
-$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citedsp{Document-11013}.
-Regarding SAS, \citedsp{Document-11013} says that "the multi-fit algorithm ...
-naturally provides a base infrastructure for this process.
-In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required.
-However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
-$\bullet \bullet \bullet$ \textbf{Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
-The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures.
-Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 \texttt{DIASource} catalogs updated each night.
-The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
-See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_TNO} for more detailed DM processing case studies.
+\noindent \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
+(2008, \citep{2008arXiv0805.2366I}; Section 3.1.2) describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters, repeated every two nights for four months.
+Each exposure would have a $5\sigma$ limit of $r\sim24$; the nightly stack would have a limit of $r\sim26.5$; and the final deep stack of all exposures would have a limit of $r\sim28$.
+This description does not comment on the processing mode, but, depending on the science goals the exposures could be done as either a series of 50 non-standard visits ($1\times15$ seconds) or 25 standard visits ($2\times15$ seconds).
 
-\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
-$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
-When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \citedsp{Publication-141}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \citedsp{Publication-145}.
-For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
-\citedsp{Publication-145}.
-\\
-$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog.
-\\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-\\
-$\bullet \bullet \bullet$ \textbf{Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
-Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks.
-The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars.
+\medskip
+\noindent \textbf{Solar System Objects (SSO) -- } Four of the mini-surveys in Table \ref{tab:ddfms} have science goals related to studies of SSO.
+Observations of the North Ecliptic Spur area could yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (reference: Brandt's talk).
+The Mini-Moons Mini-Survey aims to find and study the temporarily captured satellites of the Earth (Section 10.2, \citep{2017arXiv170804058L}).
+The Sweetspot Survey would use twilight fields to find NEOs in Earth-like orbits (i.e., these objects are never in opposition fields, but overhead at sunrise/sunset; Section 10.2, \citep{2017arXiv170804058L}).
+The Meter-Sized Impactors program would find and track meter-sized impactors $<2$ weeks before impact (Section 10.2, \citep{2017arXiv170804058L}).
+{\bf Summary:} most of these science goals do not seem to require non-standard visits or exposure times, with the exception of the Sweetspot survey which occurs during twilight and thus may require shorter exposures.
+The cadence and patterns of these mini-surveys may differ from the WFD main survey, especially when very fast-moving objects are sought.
+From a processing perspective, it seems that many of these science goals will be achievable by using the products of the Moving Object Processing System (MOPS), which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night.
+The exception is finding faint SSOs (e.g., Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) through shift-and-stack (SAS) processing \citedsp{Document-11013}, because SAS is not a capability being built within the DM system and cannot be done solely by reconfiguring DM pipelines.
+An example of user-generated pipeline for SAS is described in Section \ref{sec:SPCS}.
+
+\medskip
+\noindent \textbf{Stars in the Milky Way and Magellanic Clouds -- } As described in \citedsp{Publication-141}, mini-surveys of the Galactic Plane can better distinguish faints stars from faint red galaxies by including at least 3 filters of coverage (e.g., $izy$; similar to WFD), and could mitigate losses from proper motion and increase the detection rate of stellar flares by obtaining all the images in short time span (i.e., a more concentrated cadence than the WFD).
+As described in \citedsp{Publication-145}, applying the nominal DDF observing strategy over the full area of the Large and Small Magellanic Clouds can characterize stellar variability to $M_V<6.5$ on timescales from 15 seconds to 3 days.
+For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"} \citedsp{Publication-145}.
+The Twilight survey in Table \ref{tab:ddfms} proposes short exposures to enable bright stars to be put on the same photometric system as the deeper LSST WFD main survey catalog, and enable science that is based on their long monitoring baselines from historical observations.
+In Chapter 10.4 of \citep{2017arXiv170804058L}, a proposed short-exposure survey of M67 would use the camera's stretch goal of $0.1$ second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as $6$ magnitudes brighter than the saturation level"}.
+{\bf Summary:} while some of these science goals can be accomplished with standard visits, MW \& L/SMC science goals are likely to request shorter exposure times, perhaps down to $0.1$ seconds.
+These science goals are also likely to propose cadence and filter distributions that are significantly different from the WFD main survey.
+From a processing perspective, the science goals depending on shorter exposures will only be able to be met by reconfiguring the DM pipelines if the short exposures can be shown to successfully be processed (with, e.g., instrument signature removal); the science goals can likely be met with data products in the same format as the Prompt or DR Pipeline (i.e., {\tt Source} and {\tt Object} catalogs, single visits and deep CoAdds).
+Although it is not mentioned in the above paragraph, the MW \& L/SMC science community is also most likely to require special processing to extract information from saturated stars, which is outside the scope of DM.
 See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.
 
-\noindent \textbf{Exoplanets} \\
-$\bullet$ Transits.
-The nominal DDF plan described in \citep{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \citep{2008arXiv0805.2366I}).
-\citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
-$\bullet$ Microlensing.
-Slower than a transit, \citep{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
-However, this will require image differencing in crowded fields.
-\citep{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
-Microlensing can also be done with DDF-style observing at low Galactic latitudes.
-\\
-$\bullet \bullet \bullet$ \textbf{Summary.} All of these science goals appear possible with standard visit images.
-They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
+\noindent \textbf{Exoplanets -- } As described in Section 3.1.2 of \citep{2008arXiv0805.2366I}, transiting exoplanets could be detected with the nominal DDF plan, which would allow for $1\%$ variability to be detected over hour-long timescales; a DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence.
+\citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane, and how microlensing candidates can be found with $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (since microlensing events are slower; these would then require follow-up with external facilities).
+Dealing with the more crowded fields would be mitigated by the shallower images, in this case.
+One of the main points of \citep{2013arXiv1304.3455G} is that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
+\textbf{Summary.} Some of these science goals appear possible with standard visit images, and some might request shorter exposures to avoid confusion in crowded fields when the science can be done with brighter stars.
+From a processing perspective, the science goals are likely to be achievable with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
 See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.
 
-\noindent \textbf{Supernovae} \\
-$\bullet$ The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
-The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
-\citedsp{Publication-144} \\
-$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}.
-\\
-$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
+\noindent \textbf{Supernovae -- } The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
+The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$ \citedsp{Publication-144}.
+High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}.
+Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
 This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days.
@@
 The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citedsp{Publication-142}.
-\\
-$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
-$\bullet \bullet \bullet$ \textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
-Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
-In particular, the Level 2 DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
+Short-exposure observations of bright, nearby SNe may also be useful to include near-peak photometry in the LSST magnitude system, and enable full light-curve analyses.
+\textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of a target-of-opportunity short-exposure program to observe bright SNe).
+From a processing perspecitve, the science goals appear to be accessible with reconfigured DM pipelines to stack and difference the data.
+In particular, the DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
 See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
 
-\noindent \textbf{Galaxies} \\
-$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
+\noindent \textbf{Galaxies -- } The additional depth of a DDF may provide access to a larger collection of low-$\mu$ objects.
 \citedsp{Publication-142} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
-\\
-$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
-$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}. \\
-$\bullet \bullet \bullet$ \textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
-In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.
+The DDF stacks could also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects.
+Also, the DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}.
+\textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
+In addition, it seems likely that user-generated algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, and these are beyond the scope of DM.
 
-\noindent \textbf{Weak Lensing} \\
-$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
-$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
+\noindent \textbf{Weak Lensing -- } The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch).
+\textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
+%$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
+%$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
 We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
-$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
-$\bullet \bullet \bullet$ \textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
+%$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
 
-% % % % % % % % % % % % % % % % % %
-\subsection{Future Input from the Science Community} \label{ssec:data_comm}
-
-Community input on LSST Special Programs will be ingested through a formal call for white paper proposals (Section \ref{sssec:data_comm_call}).
-This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}.
-The DM team will also interact informally with the community through, e.g., a \texttt{\Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
-
-\subsubsection{Call for White Paper Proposals}\label{sssec:data_comm_call}
-
-The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
-At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call.
-The relevant Jira issue is \jira{LIT-325} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
-The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director.
-The criteria to be met by the proposals may include specifications that they: satisfy minimum technical requirements (feasibility, overheads); maximize diverse scientific objectives (serve a wide community); and generate legacy datasets and add value to the products of other facilities.
-After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation.
-While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \citep{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
-
-\textbf{Initial community suggestions regarding the call for white papers.} \\
-$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
-$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations.
-Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers.
-\\
-$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an "eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
-$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).
-
-\subsubsection{\texttt{\Large{Community}} Forum}\label{sssec:data_comm_forum}
-
-As the science community prepares their white paper proposals they have been invited to interact with Data Management through this \texttt{\Large{Community}} forum on DDF.
-I am a watcher on this forum.
-So far it has been minimally used.
-\\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
-
-\noindent The questions opened up on that forum are: \\
-(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
-(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
-"Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
-\\
-(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
-(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\clearpage
-\section{Special Programs Processing Case Studies}\label{sec:SPCS}
-
-For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
-Note that we are not including any analysis in these descriptions, only processing and products.
-These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
-
-Basic steps that we use to describe a processing case study: \\
-Step 1. Data Acquisition. \\
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Step 3. Delivery of LSST Processed Images. \\
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-Step 6. Science User Level 3 Pipelines and Products. \\
-
-% % % % % % %
-\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
-
-This Special Programs processing summary is based on Becker et al.
-(2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
-See Section \ref{ssec:SPCS_SAS} for a more general processing case-study of a shift-and-stack Level 3 pipeline running regularly on a large amount of data.
-
-Step 1. Data Acquisition. \\
-The observational sequence is triggered.
-In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
-This sequence is always repeated 2-3 nights later.
-This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
-Data obtained in the $g$-band filter is also acceptable.
-\citedsp{Document-11013}
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-Shift-and-stack processing is beyond the scope of DM's algorithms.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external process (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be adopted).
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The Level 2 difference imaging routine will be used with the same template tract/patch for all.
-Custom, user-supplied algorithms will shift the exposures and create difference images, then Level 2 routines will stack and do source detection and characterization and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to the database.
-
-% % % % % % %
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
-
-Step 1. Data Acquisition. \\
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-The required data products for this science goal can be met by reconfiguring the DM pipelines.
-First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF.
-The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{SPCS-1} The purpose of this ticket is to resolve whether or not the internal real/bogus routine be able to run on a nightly CoAdd of deep drilling difference images.
-If there will be issues due to a mismatch between the training and testing sets, what is the cost/benefit of fixing this and enabling this capability?
-This is not urgent and not limited to Special Programs.
-\end{enumerate}
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would \textit{all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, no separate Level 3, user-driven software appears necessary.
-
-% % % % % % %
-\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
-
-Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
-The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
-
-Step 1. Data Acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (i.e., 1 second).
-Location and exposure times are set by the sky brightness and desired saturation limits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:dmplans_NSV}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could be incorporated.
-
-Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-TBD, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-These data would not contribute to the Level 2 data products created for the WFD survey.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-If short-exposure images cannot be processed with the existing DM algorithms, would the Level 3 user-built pipeline have to start from the raw data and create new, custom algorithms for e.g., ISR?
-
-A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-This would certainly be considered a Level 3 algorithm.
-
-% % % % % % %
-\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
-
-Step 1. Data Acquisition. \\
-The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields, to a shallower depth than the WFD main survey.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:dmplans_NSV} and \ref{ssec:docrev_dmad}).
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Level 1 and 2 pipelines, and not much reconfiguration would be needed.
-The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-(\textit{MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
-
-% % % % % % %
-\subsection{A More General Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
-
-A Level 3 processing case study for shift-and-stack on a large number images. By Mario.
-
-\#1.
-The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec).
-[ and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].
-
-\#2.
-The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database.
-This will happen automatically for all images (perhaps within some range of exposure times?).
-
-\#3.
-The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes.
-This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it?
-or is is the same one that's used to process calibrations?
-have these systems been sized?].
-
-\#3 a).
-The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team.
-That is, we don't expect someone external to the ops team to babysit the code on a nightly basis.
-In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
-
-\#4.
-There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution.
-We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
-
-\#5.
-Once the processing finishes, the results of will be stored to a program-specific database.
-No alerts (in VOEvent sense) will be issued.
-We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store.
-[This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
-
-\#6.
-The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables).
-The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files).
-The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).
-
-\#7.
-When the images are made available to the batch system (step \#3), they also become available to *everyone*.
-I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
-(This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_L3}).
-
-\clearpage
-\bibliography{ms,lsst,refs,books,refs_ads}
+%\input{docrev}
 
 \end{document}
-
-%\begin{center}
-%\includegraphics[width=8cm]{figures/}
-%\includegraphics[width=8cm]{figures/}
-%\caption{ \label{fig:}}
-%\end{center}
-%\end{figure}

### 2018-06-18 docrev.tex
major revisions; issued

--- 2017-11-10 ms.tex
@@
-\documentclass[DM,lsstdraft,toc]{lsstdoc}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{latexsym}
-\usepackage{color}
-% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
-\usepackage{enumitem}
-
-\title[LSST Special Programs]{Data Management \\ and LSST Special Programs}
-
-\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, and E.~Bellm}
-
-\setDocRef{DMTN-nnn}
-\date{\today}
-\setDocRevision{TBD}
-\setDocStatus{draft}
-
-\setDocAbstract{{\bf This is a draft document and may yet change.} The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community, and that these plans have been appropriately sized for computational processing and are accurately represented in the work-hours budget.
-Unresolved issues spawn {\bf Actions}, which are typically documentation change requests or proposed studies to assess a potential problem.}
-
-\setDocChangeRecord{%
-\addtohist{1}{2017-11-??}{Internal working document.}{Melissa Graham}
-%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
-}
-
-\begin{document}
-
-\maketitle
-
-% CITATION EXAMPLES
-% \verb|\citellp|: \citellp{LPM-17, LSE-30} \\
-% \verb|\citell|: (SRD; \citell{LPM-17,LSE-29}) \\
-% \verb|\citep[][]|: \citep[e.g.,][are interesting]{LPM-17,LSE-29} \\
-% \verb|\cite|: \cite{LPM-17,LSE-29}
-
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\section{Introduction} \label{sec:intro}
-
-The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}; but see also the minor proposed change to this statement in \ref{Size-1}):
-
-\begin{quote}
-LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
-\end{quote}
-
-\noindent The purpose of this study is to ensure that the requirements and plans of DM are accurately expressed in all major DM documentation, and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
-This study spawns {\bf Action} items, which are typically either documentation change requests or proposals for future studies needed to resolve an issue.
-
-%In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans, in which case an {\bf Action} is spawned.
-
-Section \ref{sec:data} compiles an overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
-We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera.
-
-Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data, including the capabilities for processing non-standard visits and whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
-We also describe DM's intent for incorporating Special Programs data into the WFD main survey products, reconfiguring pipelines to create unique products, and providing support for end-user Level 3 processing.
-
-Section \ref{sec:docrev} steps through each of the relevant LSST documents and highlights specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
-%These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
-{\bf The Concerns raised by this study fall into two main categories:} \\
-%1. Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
-%2.
-Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD, and is in the staffing budget.
-\\
-
-Section \ref{sec:SP_SC} gathers the existing efforts from the Science Community regarding Special Programs, summarizes the scientific needs of as many currently proposed Special Programs as we can find, and outlines the future mechanisms for LSST to gather more community input -- including some recommendations for the future call for Special Programs white paper proposals.
-
-Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys.
-These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional issues regarding the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Potential Diversity of Special Programs Data} \label{sec:data}
-
-To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data.
-The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (i.e., standard visit images), then it might not be adequately handled by the DM system.
-To get an idea of the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes, in Section \ref{ssec:data_bounds}.
-
-We can also get an idea of the expected diversity in exposure type and survey pattern from the current set of proposed Special Programs for LSST, which are reviewed in more detail in Section \ref{sec:SP_SC}.
-We find that most of the proposed Special Programs are likely to use standard visits, two $\sim15$ second exposures, and that the potential diversity of Special Programs data is mainly comprised of exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, long series of exposures obtained of the same field (i.e., deep drilling), and images of crowded fields.
-
-All aspects related to DM's {\it processing} of Special Programs data are addressed in Section \ref{sec:dmplans}.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Technical Boundaries for Special Programs Data}\label{ssec:data_bounds}
-
-We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals.
-For the purposes of this document we consider the ``observing boundaries" as those imposed by the camera, telescope, or site, and independent of any DM processing capabilities.
-It is not the job of this document to define these observing boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
-
-The current technical limitations on what kind of observations can be obtained with the LSST telescope and camera are described below in a rough draft, to be updated in the future.
-Additionally, in Section \ref{ssec:docrev_oss}, we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document and highlight a couple more questions regarding potential boundaries on Special Programs observations.
-
-Filter Changes: \\
-$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
-\\
-$\bullet$ Note that changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
-$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
-$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
-However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
-\\
-$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
-$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights. \\
-
-Exposure Times: \\
-$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
-$\bullet$ The maximum exposure time is not restricted. \\
-$\bullet$ However, for exposure times there are other considerations.
-The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
-Changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
-A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Additionally, the impact on CR rejection routines is untested for long exposures.
-
-Camera Rotation: \\
-$\bullet$ The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
-\textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
-Can't remember where I heard it.}
-
-Telescope Slews: \\
-$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
-
-Total Time: \\
-$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey).
-This could be increased a bit, but otherwise imposes an upper limit on the total time request of a given Special Program.
-
-Operations Simulations: \\
-$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{Bounds-1} The purpose of this ticket is to request clarification regarding several aspects of the telescope and camera capabilities that might impose boundaries on the diversity of data from Special Programs.
-First, have the following proposed filter change tolerances been confirmed: (1) the minimum time between filter changes has no restrictions from thermal tolerances, but it is recommended to keep the filter change rate lower than once every 10 minutes in order to minimize overheads; (2) the maximum total number of filter changes is 100000 over 15 years (18/night) and the maximum number of filter swaps in and out of the carousel is 3000 in 15 years (once every two nights).
-Second, are there any restrictions on the use of the camera rotator or the total amount of slew time?
-For example, could a Special Program continuously jump between well-separated fields on the sky for an extended amount of time?
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
-
-As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs is that ``LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible" (DPDD, LSE-163, \cite{LSE-163}), and that an additional $\sim$10\% of the LSST computational system has been added and will be dedicated to user processing (of main survey or special programs data).
-
-In this section we review in greater detail the planned ability of DM to process non-standard visit images (Section \ref{ssec:dmplans_NSV}), the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:dmplans_NSV}
-
-A standard visit image is considered to be two snap exposures of $\sim15$ second each, and in the WFD main survey sequential visits will be of different fields.
-As we found in Section \ref{sec:data}, Special Programs are most likely to request observing modes with shorter exposure times and long sequences of visits to the same field.
-In this section we review DM's planned capability to process such non-standard visit sequences, keeping in mind that DM's processing boundaries might ultimately be defined not by exposure acquisition parameters, but by the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding), and that these boundaries might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{ssec:docrev_dmad}).
-
-\subsubsection{Exposure Times}
-
-Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-We consider shorter and longer exposure in turn.
-
-{\bf Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
-\end{center}
-\end{figure}
-
-{\bf Longer Exposures.} There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
-One possibly useful study is Chang et al.
-(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
-
-In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
-
-Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
-This is probably best left ot a policy issue (beyond the scope of this document).
-
-\subsubsection{Image Acquisition Rate}
-
-OSS-REQ-0291 requires that short exposures be spaced by 15 seconds, and so DM's ingest rate of exposures should never be more than 1 per 15 seconds, or a visit rate of 1 per 30 seconds.
-However, {\it if a higher image acquisition rate is cleared by the camera team}, then DM might encounter some issues such as the minimum time required to save an image to disk, the handling of a processing backup, and/or inconsistently associated {\tt DIASources}.
-
-{\bf Minimum Save Time:} If short exposures are taken in an unbuffered sequence, the minimum time between exposures is the shutter-crossing time (1 second) plus the readout time (2 seconds): 3 seconds.
-This is too rapid an acquisition rate for processing with Level 1, but what if the science goals accept saving them to disk and analyzing them later?
-{\bf How long does it take to save an image to disk?}
-
-{\bf Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
-(Such options might be needed anyway to process crowded fields with $>$10k {\tt DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
-However, at NCSA there should be ways to elastically change the amount of processing power available.
-This is also an issue for crowded fields (below).
-
-{\bf Unassociated {\tt DIASources}:} the consistency of {\tt DIAObjects} suffers if the same patch of sky is observed in quick succession (overlapping regions of short exposures and/or sustained 30 second visits to the same field).
-In other words, it is not guaranteed that the next string of processing would have access to the latest {\tt DIAObjects} for association.
-It is furthermore unclear whether we can flag an Alert as potentially having incomplete source-object associations.
-This is also an issue for crowded fields (below).
-
-\subsubsection{Exposures per Visit, and Long Sequences of a Single Field}
-
-K.-T.
-Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
-This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
-Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
-things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
-The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
-To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
-Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
-{\it MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
-
-K.-T.
-has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
-One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
-The processing pipeline would need to know how to divide this sequence up into visits.
-As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
-
-\subsubsection{Alerts from Crowded Fields}
-
-Special Programs are more likely to include crowded fields than the WFD main survey area.
-Due to the increased number of sources, the number of {\tt DIASources} -- and therefore the number of Alerts -- increases as well.
-In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
-A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete.
-K.-T.
-reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed.
-This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).
-
-\subsubsection{Summary}
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{NSV-1} The purpose of this ticket is to ensure that DM's capability for processing non-standard visit images, and required by LSR-REQ-0111, will meet the needs of Special Programs data (which might be considerably more diverse than data for the wide-fast-deep main survey).
-For now we have adopted a broad interpretation of ``processing" to mean both the initial ISR stage and also the Level 1 Alert Production pipeline, because allowing as many LSST images as possible to contribute to the Alert Stream is a benefit to time domain science.
-After comparing the anticipated potential diversity in Special Programs data with the currently planned DM pipelines, we have identified several open questions that need to be resolved in advance of the call for Special Programs white papers in the spring of 2018.
-(1) What is the anticipated minimum (and maximum, if applicable) exposure time for an image to be successfully processed by ISR and DIA?
-In the latter case, this will probably be limited by the time needed to generate a well-formed PSF.
-(2) Presently, OSS-REQ-0291 requires that even short exposures be spaced by 15 seconds.
-Might this requirement be slackened to accommodate a higher cadence Special Program?
-If so, and if Level 1 processing is required for the science goals, then DM will need to assess options to elastically increase NCSA computational resources and/or the policy for dropping {\it vs.} delaying Alerts during observations with a fast ingest cadence.
-These issues apply also to Special Programs in the Galactic plane, where crowded fields will increase the volume of Alerts.
-Alternatively, we could formally require that only 30 second visits will be allowed to contribute to Level 1 Alert Production (3) Long sequences of images on a single field are a likely Special Programs mode (e.g., DDF).
-We need to confirm that the processing pipeline has the ability to re-parse a scheduled $N$-exposure visit into standard $2$-exposure visits.
-Consequential images of the same field may lead to unassociated {\tt DIASources} of the same {\tt DIAObject}, and it should be confirmed that the DM pipelines can recognize this when it happens, and flag the resulting Alerts appropriately.
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
-
-DM will want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
-In the following three sections we summarize when and how Special Programs data might be incorporated into the pipelines and products of the Difference Imaging Analysis and Alert Stream, the Data Release Pipeline products, and/or MOPS.
-
-\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
-
-It is beneficial to transient science to include as many LSST images into the Level 1 Difference Imaging Analysis and Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST (this prospect is discussed further in Section \ref{ssec:dmplans_L3}).
-Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
-DM's ability to process the potential diversity of image types from Special Programs is explored above in Section \ref{ssec:dmplans_NSV}, and we continue the discussion of incorporating processed images into the Level 1 DIA pipeline.
-
-{\bf Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
-So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine.
-When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale).
-K.-T.
-confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible.
-This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions.
-K.-T.
-also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.
-
-\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
-
-The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data (aside from e.g., the North Ecliptic Spur area, which has been treated as a Mini-Survey region that will generate a larger number of SSO discoveries).
-Since MOPS takes {\tt DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
-There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
-
-\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
-
-Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
-When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
-For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
-In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs, and it will not be the default to include Special Programs data into the products for the WFD main survey.
-This is mostly discussed here in context with the {\tt Source} and {\tt Object} catalogs and the deep CoAdds, but also applies to the DIA pipeline products.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
-
-LSST DM intends to reconfigure its existing pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
-In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
-
-In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
-This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
-
-The current outstanding issue related to DM's processing of special programs data is that the work-hours needed to reconfigure and test the pipelines, run them and verify the data products for public release -- potentially on intermediate timescales that do not coincide with the prompt/yearly Level 1 and 2 timescales, e.g., monthly stacks of deep drilling fields -- have not been included in the personnel budget and thus, at the moment, constitute an expansion of scope.
-As we describe in Section \ref{ssec:docrev_dmsr} there are requirements in the DMSR that state that DM does have the responsibility to process Special Programs data.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-1} The purpose of this ticket is to clarify the DM staffing needs required to reconfigure pipelines for Special Programs, test and validate them, and then run and verify their products -- during commissioning and into operations.
-Questions to be answered include, e.g., product ownership of the reconfigured pipelines and test data products, and how do we set goals and deadlines for this?
-\end{enumerate}
-
-As described in Section \ref{ssec:docrev_dmsr}, the DMSR contains requirement DMS-REQ-0320, which states that "it shall be possible for special programs to trigger their own data processing recipes".
-Assuming that this means that a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline, it would appear that auto-triggering of alternative reduction pipelines is not currently supported by the DM system.
-
-Also described in Section \ref{ssec:docrev_dmsr} is the DMSR requirement that ``it shall be possible for special programs to trigger their own data processing recipes" (DMS-REQ-0320).
-However, there is currently no requirement for the scheduler to initiate non-standard processing with a certain pipeline if the data was executed under a given program.
-Since the data processing is scheduler-driven, this would have to be an OSS requirement.
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{reconfig-2} The purpose of this ticket is to figure out whether and how auto-triggering of alternative data processing recipes for Special Programs can be supported by the DM system.
-If it can't be, a change request on DMS-REQ-0320 might be needed to down-scope.
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_L3}
-
-As explained above, LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs whenever possible.
-In Section \ref{ssec:docrev_sizing} we describe how LSST has allocated 10\% of its computational and storage capacity for this processing because 10\% of the images will be generated by Special Programs.
-
-However, if a Special Program requires processing that cannot be accomplished by reconfiguring existing DM algorithms and codes, then the science community takes responsibility for data processing (i.e., it falls under Level 3).
-LSST has furthermore allocated an {\it additional} 10\% of its computing resources for user-driven processing and analysis.
-This 10\% is unrelated to the fraction of Special Programs data, and covers the user-driven reprocessing of any and all LSST data (WFD main survey and Special Programs).
-
-In the case where a Special Program requires an intensive amount of additional computational resources (i.e., too large a fraction of the 10\% allocated), the processing will have to be done on a separate, external computational resource.
-To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-
-In the case where user-driven processing pipelines consume a small amount of that 10\%, data export and processing on external computation resources will not be necessary.
-Users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their pipelines on LSST servers, monitor their progress, and analyze their output.
-DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
-
-However, in the case where user-driven processing pipelines consume a {\bf large} amount of that 10\%, data export and processing on external computation resources {\bf will} be necessary.
-The boundaries of what constitutes ``intensive" processing needs are discussed in Section \ref{ssec:docrev_sizing}.
-
-{\bf Level 3 Alert Stream?
---} User-driven processing cannot contribute to the Level 1 Alert Stream, but users could create an output that mimics an Alert package.
-K.-T.
-raises the issue that running servers or publishing VOEvents from DACs may be tricky due to network and security considerations, but that otherwise publishing in non-LSST Alert-like packets ought to be fine.
-
-{\bf Lower Latency for Data Access?
---} There is currently a limit of 24h before processed single visit images will be available, but Special Programs processing might need them sooner.
-Furthermore, download limits might be imposed and it isn't clear whether it will be possible for a user to pull over the full nightly amount of data if, e.g., they can fund their own server system to process it.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-1} The purpose of this ticket is to ask Gregory about whether there are aspects to the Science Platform that need to be developed to better serve the needs of the community doing Special programs processing.
-Examples include support for Level 3 alert streams and a lower latency for access to processed visit images.
-\end{enumerate}
-
-It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
-For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
-An alternative option to ``adopted" code is ``adopted" data products: situations in which Level 3 code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
-
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{L3-2} The purpose of this ticket is to clarify whether there is a process in place, and if staffing needs have been defined, for adopting Level 3 code into DM's pipelines, and/or for adopting data products, verifying them, and hosting their public release.
-This issue is not limited to Special Programs, but is likely to be encountered in that context.
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % %
-\clearpage
 \section{DM Documentation Review}\label{sec:docrev}
@@
 Here we gather every requirement or specification relevant to Special Programs in the existing DM documentation, identify areas where the intent of LSST DM is not adequately represented, and suggest clarifications, changes, or additions to these documents.
+If JIRA tickets have been spawned, they are noted by name.
+All remaining "Action Items" are associated with RFC-412, which is complete, and so should be disregarded.
 
 % % % % % % %
-\subsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{ssec:docrev_srd}
+\subsection{Science Requirements Document, SRD, \citeds{LPM-17}}\label{ssec:docrev_srd}
 
@@
 
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[resume,topsep=-10pt,after=\vspace{10pt},label= \textbf{Action \Roman*}] \item \label{OSS-1} The purpose of this ticket is to confirm that the minimum number of "science proposals" that can be optimized by the scheduler, {\tt nSciProp} = 6 from OSS-REQ-0027, is not too low, given that some proposed mini-surveys might have quite complicated observing strategies.
-\end{enumerate} % Assign to Zeljko.
+$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
+{\bf JIRA ticket DM-12579 confirmed that there is no maximum number, and so many Special Programs will be able to be included in the scheduler}.
 
@@
 In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
-It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
-This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
+It is the DMAD's codes and algorithms that DM will be \textit{reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
+This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}).
 
@@
 However, K.-T.
-has pointed out that the capability to regenerate the {\it exact same} template -- the pixels that were subtracted -- is not a current deliverable.
+has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable.
 However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.
 
-$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
+$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
 It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
+Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType.
 Indicates which input was used to produce a given object."} Is that all we need?
-\\
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-1} The purpose of this ticket is to clarify whether {\tt Object.prv\_inputID} in the database schema will identify whether an {\tt Object} is an externally provided coordinate.
-\end{enumerate}
+{\bf JIRA ticket DM-12580 clarified that  \texttt{Object.prv\_inputID} in the database schema is one possible way to identify whether an \texttt{Object} is an externally provided coordinate}.
 
-$\bullet$ The {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
-{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+$\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
+\texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
+\texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
 Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
 Are 32 floats in each of the 6 filters always going to be a large enough volume?
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DPDD-2} The purpose of this ticket is to ask if DM should consider whether the {\tt float[6x32]} for the characterization parameters for variability is of adequate size to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-This study might be similar to e.g., DMTN-049 for photometric redshifts included in the DRP {\tt Object} catalog.
-\end{enumerate}
+The Transients and Variable Stars Science Collaboration currently has a Task Force assigned to address this, and {\bf it is the topic of JIRA ticket DM-12581}.
 
 % % % % % % %
-\subsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
+\subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}
 
-LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
+LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
 Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}.
@@
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Cal-1} The purpose of this ticket is to define the potential additional calibration needs of Special Programs data, which might be significantly more diverse in terms of, e.g., exposure time and/or sky background levels in the case of a a twilight survey of short exposures.
-Questions include how much additional day time might be needed to obtain raw calibration frames, or the additional computational resources needed to create and/or load calibration frames or catalogs during processing.
-The goal of this would be to confirm that the DM system prepared to calibrate the diversity of data that might be generated by Special Programs.
-\end{enumerate}
+{\bf JIRA ticket DM-12582 is currently open, and aims to define the potential additional calibration needs of Special Programs data.}
 
@@
 DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
-As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
+As mentioned in Section \ref{ssec:docrev_lsr}, LSR-REQ-0055 requires that \texttt{userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
+It is important to note that this 10\% is \textit{in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that {\it an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
+\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-1} RFC (LSE-163): We propose to update the DPDD to better clarify that 10\% of the computational resources are allocated for processing Special Programs data because they are expected to use 10\% of the on-sky time, and that \textit{an additional 10\%} has been added to the overall computational system to accommodate user-driven processing to achieve specific science goals, which includes both reprocessing WFD main survey data and (re)processing Special Programs data.
 \end{enumerate}
 
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
+Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks \textit{"whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
 Have these systems been sized?"}.
-The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
+The answer to this comes from K-T: \textit{In the current model, the same production batch system is used for processing calibrations (CPP), processing ``catch-up" L1, processing the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
 There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can ``borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{Size-2} The purpose of this ticket is to ensure that the overall DM computational system is appropriately sized, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, {\it and} incorporated into the Level 2 data products, {\it and} have their own dedicated reconfigured pipelines and separate products, {\it and} have a higher probability for user-defined processing.
-The latter would apply to the size of the DACs.
-Although an accurate estimate of the science-driven DAC size might not be truly possible until after the call for Special Programs white papers, an initial attempt must be made because this call will need to define a computational boundary for what can be done in the DAC, and what will require the proposer to bring additional resources to the table.
-For example, if the fractional computational needs are equivalent to the fraction of total data that the program acquires, that is acceptable; but if the program requires, e.g., $>2\times$ as much computational resources, then an external contribution might be necessary.
-Defining what the call will say regarding the need for external processing is a goal of this ticket.
-\end{enumerate}
+{\bf JIRA ticket DM-12583 is open to ensure that the overall DM computational system is appropriately sized}, given it is likely that a large portion of the 10\% of data from Special Programs might be multiply-processed: incorporated into the Level 1 pipeline, \textit{and} incorporated into the Level 2 data products, \textit{and} have their own dedicated reconfigured pipelines and separate products, \textit{and} have a higher probability for user-defined processing.
 
@@
 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{DE-1} The purpose of this ticket is to resolve whether or not the implantation of fake sources will be done as part of the Level 1 detection efficiency characterization, or if an alternate method will be sufficient for the science needs.
-K.-T.
-says the computational system is sized to accommodate it.
-If this issue remains unclear, perhaps a study that assess the options, their computational costs, and their scientific payoff should be done so that DM can better assess the need for fake sources and allocating staffing hours as necessary.
-Note that this is not a question limited to Special Programs.
-\end{enumerate}
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{Special Programs and the Science Community}\label{sec:SP_SC}
-
-So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
-\cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
-Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}).
-We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.
-
-All other aspects of LSST Special Programs remain open to proposals, including: \\
-$\bullet$ additional deep drilling fields \\
-$\bullet$ refined observing strategies for deep drilling fields \\
-$\bullet$ optimized survey areas for the NES, South Pole, and Galactic Plane \\
-$\bullet$ refined observing strategies for the NES, South Pole, and Galactic Plane \\
-$\bullet$ additional mini-surveys \\
-$\bullet$ the total fraction of time spent on Special Programs (nominally 10\%)
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Currently Proposed Special Programs} \label{ssec:data_science}
-
-In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed.
-We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science.
-In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
-Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
-
-\noindent {\bf Resources:} \\
-$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
-$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
-$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
-$\bullet$ "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
-$\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
-$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
-
-\begin{table}[h]
-\begin{center}
-\begin{footnotesize}
-\caption{Approved DDF and Incomplete List of Potential MS.}
-\label{tab:ddfms}
-\begin{tabular}{lll}
-\hline \hline
-Name & Coordinates & Description  \\
-\hline
-DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
-DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
-DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
-DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
-DDF TBD  & & TBD \\
-North Ecliptic Spur      & & solar system objects (find and characterize) \\
-Galactic Plane             & & more intensive stellar surveying \\
-South Equatorial Cap  & & S/LMC and more Galactic science \\
-Twilight                        & & short exposures (0.1s) for bright stars \\
-Mini-Moons                     &  & finding mini-moons \\
-Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
-Meter-Sized Impactors     & & detection a week before impact \\
-GW Optical Counterparts & & search and recovery \\
-Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
-\hline
-\end{tabular}
-\end{footnotesize}
-\end{center}
-\end{table}
-
-\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
-Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
-(Note that this nominal survey is missing the overhead for readout and shutter, 3 seconds per exposure, which means that the LSST capability is closer to 43--45 exposures per hour, minus the time for 3--4 filter changes -- not 50 exposures per hour.)
-
-\medskip
-\noindent \textbf{Solar System Objects (SSO)}\\
-$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
-$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) \\ 
-$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
-Regarding SAS, \cite{BeckerWP} says that "the multi-fit algorithm ...
-naturally provides a base infrastructure for this process.
-In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required.
-However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
-$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
-The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures.
-Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night.
-The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
-See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_TNO} for more detailed DM processing case studies.
-
-\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
-$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
-When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}.
-For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
-\citep{SzkodyWP}.
-\\
-$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog.
-\\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-\\
-$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
-Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks.
-The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars.
-See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.
-
-\noindent \textbf{Exoplanets} \\ 
-$\bullet$ Transits.
-The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
-\cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
-$\bullet$ Microlensing.
-Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
-However, this will require image differencing in crowded fields.
-\cite{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
-Microlensing can also be done with DDF-style observing at low Galactic latitudes.
-\\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images.
-They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
-See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.
-
-\noindent \textbf{Supernovae} \\
-$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
-The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
-\cite{KesslerWP} \\
-$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}.
-\\
-$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
-This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days.
-These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss.
-SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
-\\
-$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
-Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
-In particular, the Level 2 DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
-See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
-
-\noindent \textbf{Galaxies} \\
-$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
-\cite{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
-\\
-$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
-$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
-In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.
-
-\noindent \textbf{Weak Lensing} \\
-$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
-$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Future Input from the Science Community} \label{ssec:data_comm}
-
-Community input on LSST Special Programs will be ingested through a formal call for white paper proposals (Section \ref{sssec:data_comm_call}).
-This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}.
-The DM team will also interact informally with the community through, e.g., a {\tt \Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
-
-\subsubsection{Call for White Paper Proposals}\label{sssec:data_comm_call}
-
-The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
-At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call.
-The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
-The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director.
-The criteria to be met by the proposals may include specifications that they: satisfy minimum technical requirements (feasibility, overheads); maximize diverse scientific objectives (serve a wide community); and generate legacy datasets and add value to the products of other facilities.
-After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation.
-While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \cite{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
-
-\textbf{Initial community suggestions regarding the call for white papers.} \\
-$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
-$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations.
-Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers.
-\\
-$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an "eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
-$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).
-
-\subsubsection{{\tt \Large{Community}} Forum}\label{sssec:data_comm_forum}
-
-As the science community prepares their white paper proposals they have been invited to interact with Data Management through this {\tt \Large{Community}} forum on DDF.
-I am a watcher on this forum.
-So far it has been minimally used.
-\\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
-
-\noindent The questions opened up on that forum are: \\
-(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
-(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
-"Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
-\\
-(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
-(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{Special Programs Processing Case Studies}\label{sec:SPCS}
-
-For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
-Note that we are not including any analysis in these descriptions, only processing and products.
-These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
-
-Basic steps that we use to describe a processing case study: \\
-Step 1. Data Acquisition. \\
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Step 3. Delivery of LSST Processed Images. \\
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-Step 6. Science User Level 3 Pipelines and Products. \\
-
-% % % % % % %
-\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
-
-This Special Programs processing summary is based on Becker et al.
-(2011) white paper to find TNOs with shift-and stack (SAS) \citep{BeckerWP}.
-See Section \ref{ssec:SPCS_SAS} for a more general processing case-study of a shift-and-stack Level 3 pipeline running regularly on a large amount of data.
-
-Step 1. Data Acquisition. \\ 
-The observational sequence is triggered.
-In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
-This sequence is always repeated 2-3 nights later.
-This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
-Data obtained in the $g$-band filter is also acceptable.
-\citep{BeckerWP}
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\ 
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
-Shift-and-stack processing is beyond the scope of DM's algorithms.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would {\it all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external process (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be adopted).
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The Level 2 difference imaging routine will be used with the same template tract/patch for all.
-Custom, user-supplied algorithms will shift the exposures and create difference images, then Level 2 routines will stack and do source detection and characterization and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to the database.
-
-% % % % % % %
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
-
-Step 1. Data Acquisition. \\ 
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\ 
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
-The required data products for this science goal can be met by reconfiguring the DM pipelines.
-First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of {\tt DIAObject}, {\tt DIASource}, and {\tt Object} that are unique to this DDF.
-The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*}] \item \label{SPCS-1} The purpose of this ticket is to resolve whether or not the internal real/bogus routine be able to run on a nightly CoAdd of deep drilling difference images.
-If there will be issues due to a mismatch between the training and testing sets, what is the cost/benefit of fixing this and enabling this capability?
-This is not urgent and not limited to Special Programs.
-\end{enumerate}
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would {\it all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, no separate Level 3, user-driven software appears necessary.
-
-% % % % % % %
-\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
-
-Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
-The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
-
-Step 1. Data Acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (i.e., 1 second).
-Location and exposure times are set by the sky brightness and desired saturation limits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:dmplans_NSV}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could be incorporated.
-
-Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-TBD, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-These data would not contribute to the Level 2 data products created for the WFD survey.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-If short-exposure images cannot be processed with the existing DM algorithms, would the Level 3 user-built pipeline have to start from the raw data and create new, custom algorithms for e.g., ISR?
-
-A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-This would certainly be considered a Level 3 algorithm.
-
-% % % % % % %
-\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
-
-Step 1. Data Acquisition. \\
-The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields, to a shallower depth than the WFD main survey.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:dmplans_NSV} and \ref{ssec:docrev_dmad}).
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Level 1 and 2 pipelines, and not much reconfiguration would be needed.
-The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-({\it MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
-
-% % % % % % %
-\subsection{A More General Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
-
-A Level 3 processing case study for shift-and-stack on a large number images. By Mario.
-
-\#1.
-The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec).
-[ and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].
-
-\#2.
-The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database.
-This will happen automatically for all images (perhaps within some range of exposure times?).
-
-\#3.
-The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes.
-This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it?
-or is is the same one that's used to process calibrations?
-have these systems been sized?].
-
-\#3 a).
-The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team.
-That is, we don't expect someone external to the ops team to babysit the code on a nightly basis.
-In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
-
-\#4.
-There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution.
-We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
-
-\#5.
-Once the processing finishes, the results of will be stored to a program-specific database.
-No alerts (in VOEvent sense) will be issued.
-We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store.
-[This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
-
-\#6.
-The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables).
-The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files).
-The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).
-
-\#7.
-When the images are made available to the batch system (step \#3), they also become available to *everyone*.
-I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
-(This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_L3}).
-
-\clearpage
-\bibliography{ms,lsst,refs,books,refs_ads}
-
-\end{document}
-
-%\begin{center}
-%\includegraphics[width=8cm]{figures/}
-%\includegraphics[width=8cm]{figures/}
-%\caption{ \label{fig:}}
-%\end{center}
-%\end{figure}
+{\bf JIRA ticket DM-12584 is currently open to resolve whether or not the implantation of fake sources will be done as part of the Level 1 detection efficiency characterization, or if an alternate method will be sufficient for the science needs.}

### 2018-06-18 spcs_sas.tex
major revisions; issued

--- 2017-09-01 ms.tex
@@
-\documentclass[DM,lsstdraft,toc]{lsstdoc}
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage{latexsym}
-\usepackage{color}
-% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
-\usepackage{enumitem}
-
-\title[LSST Special Programs]{Data Management \\ and LSST Special Programs}
-
-\author{M.~L.~Graham, M.~Juri\'{c}, and E.~Bellm}
-
-\setDocRef{DMTN-nnn}
-\date{\today}
-\setDocRevision{TBD}
-\setDocStatus{draft}
-
-\setDocAbstract{The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community.
-To do this, we begin with a review of the broad science goals of the previously proposed Special Programs and the potential diversity of data they might generate.
-We also consider any technical limitations on the data from the telescope, camera, and/or DM processing capabilities, and the methods by which information could be exchanged with the scientific community on these issues (i.e., the upcoming call for white paper proposals).
-We then move on to summarize the current DM plans for processing Special Programs data, and highlight specific requirements or text in the existing documentation that might need additions or clarifications to reflect these plans.
-We also develop a series of case studies for DM's processing of Special Programs data to better illuminate any complex or hidden issues; these case studies could also serve as examples for the upcoming call for community white papers.}
-
-\setDocChangeRecord{%
-\addtohist{1}{2017-04-??}{Internal working document.}{Melissa Graham}
-%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
-}
-
-\begin{document}
-
-\maketitle
-
-% CITATION EXAMPLES
-% \verb|\citellp|: \citellp{LPM-17, LSE-30} \\
-% \verb|\citell|: (SRD; \citell{LPM-17,LSE-29}) \\
-% \verb|\citep[][]|: \citep[e.g.,][are interesting]{LPM-17,LSE-29} \\
-% \verb|\cite|: \cite{LPM-17,LSE-29}
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\section{Introduction} \label{sec:intro}
-
-The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
-
-\begin{quote}
-LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
-\end{quote}
-
-\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent.
-In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans.
-
-\subsection{Study Overview}\label{ssec:intro_overview}
-
-Section \ref{sec:data} compiles a comprehensive overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters.
-We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera, the scientific needs of as many currently proposed Special Programs as we can find, and the future mechanisms for LSST to gather more community input.
-We also consider whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.
-
-Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data: reconfiguring pipelines to create unique products, incorporating data into the WFD main survey products, and enabling end-user Level 3 processing.
-Having established the {\it intent} of DM, we step through each of the relevant LSST documents and raise {\bf "Concerns"}, which highlight specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data.
-These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues.
-{\bf The Concerns raised by this study fall into two main categories:}
-\begin{enumerate}
-\item Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be.
-\item Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD.
-\end{enumerate}
-
-Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys.
-These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional {\bf Concerns} for the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Potential Diversity of Special Programs Data} \label{sec:data}
-
-To better identify what might be missing in the DM plan to accommodate Special Programs, we need a comprehensive understanding of the potential diversity in future Special Programs data.
-The concern is that if Special Programs data is significantly different from the Wide-Fast-Deep survey of tiled, nighttime, 15--30 second exposures (standard visit images), then it might not be adequately handled by the DM system.
-To get a handle on the potential diversity of Special Programs data, we first attempt to understand the technical limitations that the facility and instrumentation places on the observing modes, e.g., exposure times, filter changes (Section \ref{ssec:data_bounds}).
-Next, we explore the current set of proposed science programs for LSST (Section \ref{ssec:data_science}) and the types of observations they would want to do, and then describe when and how we can incorporate revisions to these proposals from the user community in the future (Section \ref{ssec:data_comm}).
-Finally, we explore whether DM can or should set down any technical constraints on how far an observing program can deviate from a sequence of standard visit images and still be processed by DM pipelines, be they the reconfigured and/or the standard Level 1 and 2 pipelines (Section \ref{ssec:data_DMconstraints}).
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Observing Boundaries for Special Programs}\label{ssec:data_bounds}
-
-We can constrain the potential diversity of Special Programs data if we know the observing boundaries that will be imposed on the accepted proposals.
-So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
-\cite{2008arXiv0805.2366I} describes a nominal DDF data set as e.g., a series of fifty 15 second exposures in each of the $griz$ filters obtained once every two nights for four months, which would yield a nightly stack depth of $r<26.5$ and a full stack depth of $r<28.0$ (once combined with the wide-fast-deep survey data).
-Additionally, three mini-survey areas that have already been considered in detail are likely to move forward: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \cite{2008arXiv0805.2366I}).
-We have a summary collection of past white papers and informally proposed mini-surveys in Section \ref{ssec:data_science}.
-
-All other aspects of LSST Special Programs remain open to proposals, including: \\
-$\bullet$ additional deep drilling fields \\
-$\bullet$ refined observing strategies for deep drilling fields \\
-$\bullet$ optimized survey areas for the NES, South Pole, and Galactic Plane \\
-$\bullet$ refined observing strategies for the NES, South Pole, and Galactic Plane \\
-$\bullet$ additional mini-surveys \\
-$\bullet$ the total fraction of time spent on Special Programs (nominally 10\%)
-
-There will be some technical limitations on the mini-surveys that will be possible.
-Here's a rough draft list of such potential issues.
-It is not the job of this document to define the camera/observatory technical boundaries, but we do need to know what they are in order to predict the diversity of incoming data that DM will be processing.
-This list will be updated in the future (or revised to point to where these boundaries are defined).
-In Section \ref{sssec:dmplans_review_oss}, as we review the Observatory System Specifications (OSS, LSE-30, \cite{LSE-30}) document we highlight a couple more questions regarding potential boundaries on Special Programs observations.
-
-Filter Changes (and see \ref{OSS-6}): \\
-$\bullet$ The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293).
-\\
-$\bullet$ \textit{(MLG - The next three points are from a document-in-progress by Ivezic and Ritz; update to formal citation when it's public-facing.)} \\
-$\bullet$ The minimum time between filter changes has no restrictions from e.g., thermal tolerances.
-However, based on overheads and efficiency, it is recommended to keep the filter change rate lower than once every 10 minutes.
-\\
-$\bullet$ The maximum total number of filter changes is 100,000 over 15 years, an average of 18 changes per night. \\
-$\bullet$ The maximum number of filter swaps in/out of the carousel is 3000 in 15 years, or once every two nights.
-
-Exposure Times (and see \ref{OSS-5}): \\
-$\bullet$ The minimum exposure time is 1 second, with a stretch goal of 0.1 seconds (OSS-REQ-0291). \\
-$\bullet$ The maximum exposure time is not restricted. \\
-$\bullet$ However, for exposure times there are other considerations.
-The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
-Changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
-A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Additionally, the impact on CR rejection routines is untested for long exposures.
-
-Camera Rotation (and see \ref{OSS-7}): \\
-$\bullet$ \textit{MLG - I heard there might be a lifetime maximum on the rotator, such that if a mini-survey wants to jump between patches that are far apart on the sky repeatedly, that could wear out the mechanism?
-Can't remember where I heard it.}
-
-Telescope Slews: \\
-$\bullet$ \textit{MLG - Obviously, long slews would cause additional overheads and a loss of observing efficiency, but are there other limitations?}
-
-Calibration Images (and see \ref{OSS-2}, \ref{OSS-4}): \\
-$\bullet$ The necessary afternoon calibrations for a Special Program must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained.
-
-Operations Simulations: \\
-$\bullet$ Mini-survey observing strategies will have to show that they do not negatively impact the LSST's ability to complete the wide-fast-deep survey; not necessarily in the next call for white papers, but later during the review and revise stage, once the new OpSim runs incorporating proposed mini-surveys have been generated.
-
-Total Time: \\
-$\bullet$ It has been regularly stated that 10\% of the total observing time will be used for special programs (90\% goes to the WFD main survey), but this could be changed if significant science is enabled.
-\\
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Currently Proposed Special Programs} \label{ssec:data_science}
-
-In this section we compile information about the science goals and observational methods for Special Programs that have already been proposed.
-We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science; this Section is mainly descriptive and we do not spawn any {\bf Concerns} here.
-In Section \ref{sec:SPCS}, we create much more detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs; any {\bf Concerns} that arise are detailed in that section.
-Below, we list the set of resources that we used to compile this information, and in Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified along with an {\it incomplete} list of potential mini-surveys that people are thinking about (mainly from the presentations of Brandt and Ridgway at the 2016 AHM).
-
-\noindent {\bf Resources:} \\
-$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
-$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
-$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
-$\bullet$ "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
-$\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
-$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
-$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}
-
-\begin{table}[h]
-\begin{center}
-\begin{footnotesize}
-\caption{Approved DDF and Incomplete List of Potential MS.}
-\label{tab:ddfms}
-\begin{tabular}{lll}
-\hline \hline
-Name & Coordinates & Description  \\
-\hline
-DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
-DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
-DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
-DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
-DDF TBD  & & TBD \\
-North Ecliptic Spur      & & solar system objects (find and characterize) \\
-Galactic Plane             & & more intensive stellar surveying \\
-South Equatorial Cap  & & S/LMC and more Galactic science \\
-Twilight                        & & short exposures (0.1s) for bright stars \\
-Mini-Moons                     &  & finding mini-moons \\
-Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
-Meter-Sized Impactors     & & detection a week before impact \\
-GW Optical Counterparts & & search and recovery \\
-Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
-\hline
-\end{tabular}
-\end{footnotesize}
-\end{center}
-\end{table}
-
-\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months.
-Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.
-
-\medskip
-\noindent \textbf{Solar System Objects (SSO)}\\
-$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
-$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) \\ 
-$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
-$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}.
-Regarding SAS, \cite{BeckerWP} says that "the multi-fit algorithm ...
-naturally provides a base infrastructure for this process.
-In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required.
-However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
-$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night.
-The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures.
-Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night.
-The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3.
-See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_TNO} for more detailed DM processing case studies.
-
-\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
-$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations.
-When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}.
-\\
-$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}.
-For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}.
-\citep{SzkodyWP}.
-\\
-$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog.
-\\
-$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-\\
-$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars.
-Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks.
-The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars.
-See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.
-
-\noindent \textbf{Exoplanets} \\ 
-$\bullet$ Transits.
-The nominal DDF plan described in \cite{2008arXiv0805.2366I} would allow for $1\%$ variability detection over hour-long timescales, which is suitable for detecting transits.
-A DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have ${\rm SNR}>100$ in each single exposure of the sequence (as described in Section 3.1.2 of \cite{2008arXiv0805.2366I}).
-\cite{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane.\\
-$\bullet$ Microlensing.
-Slower than a transit, \cite{2013arXiv1304.3455G} suggests that $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (i.e., the WFD nominal cadence) can find microlensing candidates, which would then need follow-up with external facilities.
-However, this will require image differencing in crowded fields.
-\cite{2013arXiv1304.3455G} suggests that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
-Microlensing can also be done with DDF-style observing at low Galactic latitudes.
-\\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images.
-They might be possible with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
-See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.
-
-\noindent \textbf{Supernovae} \\
-$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
-The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$.
-\cite{KesslerWP} \\
-$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}.
-\\
-$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
-This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days.
-These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss.
-SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}.
-\\
-$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
-$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe).
-Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data.
-In particular, the Level 2 DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
-See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
-
-\noindent \textbf{Galaxies} \\
-$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects.
-\cite{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
-\\
-$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
-$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
-In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.
-
-\noindent \textbf{Weak Lensing} \\
-$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
-$\bullet$ {\it Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
-$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
-$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Future Input from Data Rights Holders} \label{ssec:data_comm}
-
-Community input on LSST Special Programs will be ingested through a formal call for white paper proposals (Section \ref{sssec:data_comm_call}).
-This call will request that these white papers include a section that describes the processing needs that will enable their science goals; examples of processing case studies for some proposed Special Programs are given in Section \ref{sec:SPCS}.
-The DM team will also interact informally with the community through, e.g., a {\tt \Large{Community}} forum (Section \ref{sssec:data_comm_forum}) and might learn about users needs for DM processing through that means.
-
-\subsubsection{Call for White Paper Proposals}\label{sssec:data_comm_call}
-
-The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.
-At the 2017 All Hands Meeting the Science Advisory Council seemed likely to recommend these be combined into a single call.
-The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3} and the Project's initial thoughts regarding this call for white papers are described by Beth Willman at the LSST SAC Meeting in October 2016\footnote{\url{https://project.lsst.org/groups/sac/sites/lsst.org.groups.sac/files/Willman_DDF.pdf}}.
-The plan for ingesting these white paper proposals is for them to be reviewed by the Science Advisory Council based on a set of criteria set by the Project Science Team, and the SAC will make recommendations to the LSST Director.
-The criteria to be met by the proposals may include specifications that they: satisfy minimum technical requirements (feasibility, overheads); maximize diverse scientific objectives (serve a wide community); and generate legacy datasets and add value to the products of other facilities.
-After the proposals have been reviewed (and perhaps culled or combined, as recommended by the SAC) it is projected that new OpSim runs will be produced in mid-2019 for further evaluation.
-While the initial white paper proposals will be separate from the existing Observing Strategy White Paper \cite{2017arXiv170804058L}, they could be incorporated into a comprehensive analysis of the options when these simulations are ready.
-
-\textbf{Initial community suggestions regarding the call for white papers.} \\
-$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
-$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations.
-Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers.
-\\
-$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an "eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
-$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).
-
-\subsubsection{{\tt \Large{Community}} Forum}\label{sssec:data_comm_forum}
-
-As the science community prepares their white paper proposals they have been invited to interact with Data Management through this {\tt \Large{Community}} forum on DDF.
-I am a watcher on this forum.
-So far it has been minimally used.
-\\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.
-
-\noindent The questions opened up on that forum are: \\
-(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
-(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided?
-"Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely).
-\\
-(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
-(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{DM Constraints on Processing Non-Standard Visit Images}\label{ssec:data_DMconstraints}
-
-It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines, such as those with different exposure time, dither pattern, or sky background.
-Alternatively, it might make more sense that DM's processing boundaries apply not to the exposure acquisition parameters, but the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding).
-The boundaries of this parameter space might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{sssec:dmplans_review_dmad}).
-
-Regarding exposure time, images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-We consider shorter and longer exposure in turn.
-
-(1) Shorter Exposures.
-The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint.
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale).
-As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
-\end{center}
-\end{figure}
-
-(2) Longer Exposures.
-There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-With a longer exposure time the $60$ second latency requirement could still be met, but any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
-One possibly useful study is Chang et al.
-(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages {\tt PhoSim} (Peterson et al.
-2015; \url{http://adsabs.harvard.edu/abs/2015ApJS..218...14P}) or {\tt ARROYO} (\url{http://proceedings.spiedigitallibrary.org/proceeding.aspx?articleid=848256}) to at least simply characterize the PSF stability as a function of exposure time.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Summary}\label{ssec:data_summary}
-
-Our review of the technical boundaries and currently tabled Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, and images of crowded fields.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}
-
-As presented in Section \ref{sec:intro}, the intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):
-
-\begin{quote}
-LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
-\end{quote}
-
-In this section we review in greater detail the intent of DM to incorporate Special Programs data into the pipelines and products for the WFD main survey (Section \ref{ssec:dmplans_WFD}), to reconfigure the pipelines and generate unique sets of data products for each Special Program (Section \ref{ssec:dmplans_reconfig}), and to enable Level 3 processing (Section \ref{ssec:dmplans_L3}).
-Later on, in Section \ref{ssec:dmplans_review}, we review each piece of DM documentation in turn and identify requirements and/or specifications that do not adequately represent the intent of LSST DM, as defined in Section 6 of the DPDD, with respect to Special Programs processing.
-Where relevant we suggest clarifications, changes, or additions to these documents.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Include Special Programs Data in the WFD Survey Products}\label{ssec:dmplans_WFD}
-
-It seems obvious that LSST DM would want to incorporate images obtained through Special Programs into the WFD main survey's pipelines and products whenever it is beneficial for enabling science.
-In the following three sections we summarize when and how SP data might be incorporated into the pipelines and products of the Level 1 and Alert Stream, the Level 2 DRP, and/or MOPS.
-
-\subsubsection{Level 1 and the Alert Stream}\label{ssec:dmplans_WFD_L1}
-
-It is beneficial to transient science to include as many LSST images into Level 1 and the Alert Stream as possible.
-Only products of the Level 1 $60$-second processing pipeline can contribute to the Alert Stream: this means that no Level 3 difference imaging pipeline can send its detections to the Alert Stream, although a separate Level 3 alert stream should be possible if the packet/transport mechanism that LSST defines is adopted outside of LSST.
-Therefore, only images that qualify for Level 1 processing will contribute to the Alert Stream.
-The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a SP image's suitability for Level 1 is to be in a region of sky with an existing template.
-So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is obviously going to be fine.
-However, the suitability of shorter and longer exposures is unknown, as discussed in Section \ref{ssec:data_DMconstraints}.
-
-\subsubsection{MOPS}\label{ssec:dmplans_WFD_MOPS}
-
-The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data.
-Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
-
-\subsubsection{Level 2 Data Release Pipeline}\label{ssec:dmplans_WFD_L2}
-
-Unlike with transient detection and the Level 1 Alert Stream, it is not clear whether it is beneficial to the science goals that depend on the Level 2 DRP to include as many LSST images as possible.
-We propose that the simplest option is probably that LSST DM deliver a single Level 2 DRP {\tt Objects} database that is built from a set of images with nearly constant depth and cadence.
-When Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey, or improves the Level 2 DRP in some other way, it can be included in the Level 2 DRP.
-For example, photometric calibrations may require that some or all of the (shallower) Galactic Plane Special Program survey area be incorporated in order to suppress edge effects and low-order modes in the photometric solutions.
-In short, DM should probably not promise to incorporate any specific data into the Level 2 DRP images or catalogs.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Process Special Programs Data with Reconfigured DM Pipelines}\label{ssec:dmplans_reconfig}
-
-LSST DM intends to reconfigure its existing pipelines in order to generate imaging and catalog products for Special Programs data, whenever possible.
-In this context, possible means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
-In a possible scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build unique, separate image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create {\tt DIASource} and {\tt DIAObject} catalog equivalents (this example is also given in Section 6 of the DPDD, \cite{LSE-163}).
-This type of reconfiguration is also be possible for users to do as a Level 3 pipeline (see below), but having these products provided by DM ensures a consistent and verified level of quality and public access to all processed Special Programs data products, and is expected to be a large benefit to the science results.
-
-% % % % % % % % % % % % % % % % % % 
-\subsection{Intent to Support Special Programs Processing with Level 3 Pipelines}\label{ssec:dmplans_L3}
-
-If a Special Program requires processing that cannot be accomplished with existing DM algorithms and codes, and/or requires an intensive amount of additional computational resources, it is considered Level 3 -- and in the latter case the processing might have to be undertaken on a separate computational resource.
-To support such Level 3 processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-In most cases, data export and processing on external computation resources will not be necessary.
-The majority of users will access the data, build pipelines based on DM codes, and execute them using the Science Platform, through which DM provides the infrastructure for users to run their Level 3 pipelines on LSST servers, monitor their progress, and analyze their output.
-DM will also provide a prioritization system to allocate processing resources in the case of oversubscription.
-All of this applies to both Level 3 processing for Special Programs data, and reprocessing of WFD main survey data.
-It is furthermore expected that, over time, some Level 3 pipelines will become "federated", installed and operated (and change controlled) by the LSST Operations team.
-For both federated and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of Level 1 and 2.
-
-% % % % % % % % % % % % % % % % % %
-\subsection{Documentation Review}\label{ssec:dmplans_review}
-
-Here we highlight every requirement or issue relevant to Special Programs that we could find in the existing documentation.
-Since the goal of this work is to make any needed changes to existing documentation, we have endeavored to directly tie every possible concern to the relevant sections of existing documentation (or, where possible, to requirement codes) so that we can more easily direct future changes back into the existing documentations.
-
-% % % % % % %
-\subsubsection{Science Requirements Document, SRD, LPM-17, \cite{LPM-17}}\label{sssec:dmplans_review_srd}
-
-This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291).
-The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4).
-The latter we know to be an approximate estimate.
-
-% % % % % % %
-\subsubsection{LSST System Requirements, LSR, LSE-29, \cite{LSE-29}}\label{sssec:dmplans_review_lsr}
-
-The LSR is derived from the SRD, and converts the high-level specifications from the SRD into system requirements, which flow-down to and in some cases repeated in the OSS and DMSR (next sections).
-We have the version of LSE-29 last revised on August 4, 2016.
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0} Most notably, there is no written requirement for DM to reconfigure its pipelines and produce unique, separate data products for the accepted Special Programs, wherever possible, as it intends to.
-Should there be, and is the LSR right right place for it?
-\end{enumerate}
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-0b} Similar to \ref{LSR-0}, there is no written requirement for DM to evaluate the benefit of incorporating data from Special Programs into the products of the WFD main survey, and then to do this when it is found to enable/improve science results, as it intends to.
-Should there be, and is the LSR right right place for it?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0102 defines the minimum interval between data releases to be {\tt DRT1} = 1 year (1.4.2, Data Release Processing).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-1} There is no requirement on the schedule or latency for delivering Special Programs data products that is equivalent to the requirements for delivery of yearly data releases (LSR-REQ-0102; LSE-29).
-Should there be?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0111 requires that LSST be capable of obtaining \textit{and processing} exposures that were not taken in a standard visit mode, including those with minimum exposure time of {\tt minExpTime} = 1 second, with the caveat that "non-standard visit exposures may possibly be degraded in some aspects of performance (e.g., cosmic ray rejection)" (2.3.1.2, Non-Standard Visit).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-2} While the requirement that LSST process non-standard visit exposures (LSR-REQ-0111, LSE-29) implies that DM must be able to process the full diversity of Special Programs data, it should probably be verified that this is (a) true and (b) possible.
-For example, will DM impose limitations on the exposure times for which the Instrument Signature Removal (ISR) processing can deliver a science-grade quality image, as considered in Section \ref{ssec:data_DMconstraints}?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0032 requires that the LSST DM system provide three classes of science data products as "Level 1 (nightly cadence), Level 2 (data release cadence), and Level 3 (user-specified)", and subsequent sections describe the products and their timescales for delivery (2.4.5.1 Organization of Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-3} It might be helpful to include in this section that Special Programs processing with reconfigured Level 1 and 2 codes, wherever possible, is also an effective requirement similar to LSR-REQ-0032 (LSE-29).
-(This might satisfy \ref{LSR-0}).
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0041 requires that LSST support Level 3 data products "of a nature specified by users", and LSR-REQ-0106 that LSST provide software, services, and hardware resources to enable the production and storage of those products (2.4.5.1, Organization of Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-4} The requirements for LSST to provide software, services, and hardware resources to support Level 3 data products (LSR-REQ-0041, LSE-39) all have technical limits (i.e., DM's intent to create and run reconfigured pipelines for Special Programs data "unless it requires an intensive amount of additional computational resources" as discussed in Section \ref{ssec:dmplans_L3}); should these be mentioned in the LSR?
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0042 requires that LSST produce the data products necessary to support the four primary science missions (dark energy, solar system, transients, milky way; 2.4.5.2, Science Flowdown).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{LSR-5} The requirement for LSST to produce the data products necessary for the four primary science missions (LSR-REQ-0042, LSE-29) could be extended with text to include Special Programs data?
-This is might also satisfy \ref{LSR-0}.
-\end{enumerate}
-
-$\bullet$ LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products (2.5.6, Community Computing Services).
-This is discussed further in Section \ref{sssec:dmplans_review_sizing}.
-
-$\bullet$ LSR-REQ-0075 defines that the WFD main survey science objectives will be met with 90\% of the observing time, with the remainder left to Special Programs (Section 3.3.1, Survey Time Allocation)
-
-% % % % % % %
-\subsubsection{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\label{sssec:dmplans_review_oss}
-
-The OSS contains the requirements on the site, facility, and camera that are necessary to meet the requirements specified by the LSR (LSE-29) and the SRD (LPM-17).
-We have the version of LSE-30 last revised on February 10, 2017.
-For the most part, the {\bf Concerns} raised in this section are not relevant to Data Management and are probably outside the scope of this study.
-But we keep them because they are relevant to Section \ref{ssec:data_bounds}.
-
-$\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least {\tt nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (2.1.1.2, Multiple Science Programs).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-1} There is a requirement on the minimum number of "science proposals" that can be optimized by the scheduler ({\tt nSciProp} = 6, OSS-REQ-0027, LSE-30), but is there also a technical constraint on the maximum?
-This might be relevant to some proposed mini-surveys with complicated observing strategies.
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (2.1.1.7, Visit Optimization).
-
-$\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as {\tt nRawExpNightWinterAvg} = 1960 per night on average (but up to {\tt nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and {\tt nRawExpYear} = 5.5$\times10^5$ per year, respectively.
-These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
-These estimates appear adequate for Special Programs in general.
-
-$\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as {\tt nCalibExpDay} = 450 per night on average and {\tt nCalExpYear} = 1.5$\times10^5$ per year, respectively.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-2} There is a requirement on the minimum number of calibration frames that will be supported during daytime ({\tt nCalibExpDay} = 450, OSS-REQ-0194, LSE-30); could we imagine a Special Program that would need more than that?
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0125, Section 3.1.5 deals with data products, including requirements for the variety and format of data products (i.e., images and catalogs in Level 1 and 2), the precision of the data products (e.g., photometric and astrometric accuracy, completeness/spurious thresholds for DIA sources), and the timeline for the release of data products (alerts, Level 1, and the Level 2 DRP).
-(Section 3.1.5, Data Products).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-3} There are requirements on the type and quality of data products for Level 1, 2, and 3 (OSS-REQ-0125 and all those following, in LSE-30).
-Should it be specified, for Special Programs data, whether or which of these requirements DM must meet?
-For example, OSS-REQ-0157 sets the fraction of false detections in deep CoAdds caused by unremoved artifacts to be {\tt falseDeepDetect} = 0.1\%, but would this apply to a DDF stack?
-\end{enumerate}
-
-$\bullet$ Section 3.5, Photometric Calibration, puts requirements on e.g., the relative contributions to photometric errors from instrumental (OSS-REQ-0282) and atmospheric (OSS-REQ-0276) transmissions, and that a catalog of$10000$ reference stars of $17<r<20$ mag per field be created for use (OSS-REQ-0285).
-It seems that these might not apply to non-standard images from Special Programs (3.5, Photometric Calibration).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-4} Section 3.5 of the OSS (LSE-30) puts requirements on the photometric calibrations for WFD data, but it is unclear how and whether these requirements extend to DM providing calibration of non-standard images from Special Programs (this is related to \ref{LSR-2}).
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0319 sets the requirement that the LSST be capable of continuous operation throughout the night when all visits are separated only by the readout time, thereby enabling a deep drilling field style of observations (3.6.1.3, Continuous Exposures).
-
-$\bullet$ OSS-REQ-0291 defines the minimum exposure time as $1$ second with a stretch goal of $0.1$ seconds, with a note that the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times (3.6.1.4, Minimum Exposure Time).
-\\
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-5} There is a requirement on the minimum exposure time (OSS-REQ-0291, LSE-30), but not on the spacing between short exposures and no specified maximum exposure time.
-It would be great to get this clarified from the camera team and updated in the OSS, so we know if they will they impose boundaries on Special Programs data.
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0293 requires that the maximum time for an operational filter change be {\tt tFilterChange} = 120 seconds (unlikely to be much lower in practice), and OSS-REQ-0295 requires that the LSST support at least 4 changes per night (and 8 in the day).
-There is no officially required minimum time between filter changes.
-Unofficially, this is constrained by the total lifetime number of filter changes is $100,000$ over $15$ years, or an average of $18$ changes per night\footnote{From Zeljko.}.
-Additionally, changes regarding the requirements on filters in support of "deep drilling fields" are underway here: \url{https://project.lsst.org/groups/ccb/node/938}.
-(3.6.2 Filter Swaps \& Changes).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-6}   There are currently are no official requirements on the minimum time between filter changes, only that the time for a filter change will be 120 seconds (OSS-REQ-0293).
-Should the OSS include a clarification that there is no technical constraint on the frequency of filter changes?
-({\it MLG -- maybe unnecessary.}) \end{enumerate}
-
-$\bullet$ OSS-REQ-0301 and OSS-REQ-0300 set the minimum time for continuous rotation tracking ({\tt rotTrackTime} = 1 hour) and the half-range of the rotator motion ({\tt rotTrackRange} = 90 minutes) respectively, but there do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits (3.6.3.2, Field de-rotation).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{OSS-7} The OSS contains requirements for rotator capability (OSS-REQ-0301, -0300), but is there a maximum rotation speed, or a limit on the amount of rotation per night (or in a 10-year lifetime), that might constrain the distance between successive visits or the ability to jump between two widely separated fields?
-\end{enumerate}
-
-$\bullet$ OSS-REQ-0380 sets the rate and error limits for nonsidereal tracking, which presumably would only be relevant to a Special Program (3.6.3.7, Non-Sidereal Tracking).
-
-% % % % % % %
-\subsubsection{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}}\label{sssec:dmplans_review_dmsr}
-
-The DMSR contains the top-level requirements for all of Data Management, including processing pipelines and products (revision 2017-08-11).
-
-$\bullet$ DMS-REQ-0068 requires that each raw science image store metadata regarding the date/time, site, telescope, and camera -- missing from this is the scheduler and program information?
-That kind of metadata will be necessary to identify images related to Special Programs (1.2.3, Raw Science Image Metadata).
-Related to this is DMS-REQ-0266, which requires the creation of an Exposure Catalog that also stores this kind of metadata independently, but also does not include scheduler and program information.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-1} Clarify if and how scheduler and program information (i.e., identification of an image as being associated with a Special Program) is included in the raw science image metadata and/or the Exposure Catalog (DMS-REQ-0068, -0266); if not, should it be added?
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0069 requires that DM produce processed visit images (overscan trimmed, ISR, snap-combined), and that they are not archived (but can be regenerated on demand).
-It also says that "this aspect of the processing for Special Programs data is specific to each program".
-It should be specified whether "this aspect" refers to the reduction or archiving of the processed visit images; if processed visit images for Special Programs can live on disk for longer, does this affect the sizing models?
-(1.2.2, Processed Visit Images).
-This question applies also to DMS-REQ-0010, which requires that DM create one difference image for each processed visit image (1.3.3, Difference Exposures).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-2} Confirm whether processed single visits and/or difference images for Special Programs can stay on disk and accessible to e.g., Level 3 pipelines, for a longer amount of time by user request (DMS-REQ-0069 and -0010, LSE-61).
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0069 and DMS-REQ-0010 require that DM create processed single visit images and a difference image for each of those, but this might not be possible for Special Programs that obtain non-standard exposures (i.e., short exposures in which the PSF is not well formed, or very bright sky backgrounds).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-3} Confirm the exposure boundaries for which DM can produce processed single visits and/or difference images for Special Programs that obtain non-standard images (DMS-REQ-0069 and -0010, LSE-61).
-This is related to \ref{LSR-2} and might require a dedicated additional study with simulated data, as discussed in Section \ref{ssec:data_DMconstraints}.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0274 sets the content of an Alert, but this does not currently include scheduler or program information; should this be added?
-This would enable users to e.g., use the LSST mini-broker to filter for only Alerts from their desired Special Program, in case they have a dedicated follow-up resource.
-Furthermore, it might facilitate follow-up more generally if users know e.g., that this Alert is from a field that is going to be observed again in $X$ minutes or $Y$ times in a night (1.3.13, Alert Content).
-However, the latter might be adequately available through the predicted visit schedule server specified by DMS-REQ-0353.
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-4} Could and/or should we add the program information and/or predicted revisit schedule to the Alert content (if it isn't included already)?
-(DMS-REQ-0274, -0353) \end{enumerate}
-
-$\bullet$ DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes", but this could use clarification.
-It could be misinterpreted to mean that the {\it science users} must trigger the processing, thereby classifying it as Level 3.
-More likely, it means that e.g., a header keyword identifying an image as related to a Special Program is sufficient to send it to a dedicated processing pipeline.
-(1.6.1 Processing of Data From Special Programs)
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-5} Can we clarify the mechanism by which Special Programs trigger their own data processing recipes, as required by DMS-REQ-0320?
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0321 and DMS-REQ-0344 together specify that the Level 1 processing from Special Programs shall be completed with the same latencies as applied to data from the WFD main survey.
-It should probably be specified that this applies only to Special Programs data that consists of standard visit images (and not, e.g., very short or very long exposure times unless they can be shown to result in quality difference images), and that "reporting optical transients" within {\tt OTT1} = 1 minute means contributing to the L1 Alert Stream.
-(1.6.2 Level 1 Processing of Special Programs Data, 1.6.3 Constraints on Level 1 Special Program Products Generation)
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6} Clarify that only Special Programs data that {\it can} be incorporated into the Level 1 pipeline (i.e., standard visit images, or non-standard visit images that can be shown to result in quality DIA products; see \ref{DMSR-3}), will be incorporated (relevant also to \ref{LSR-0b}, \ref{LSR-2}).
-Perhaps add that in these cases they will contribute directly to the L1 Alert Stream, with a note that generating a separate Level 3 stream of alerts specific to a Special Program should also be possible (DMS-REQ-0321, -0344).
-\end{enumerate}
-
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-6b} Related to \ref{DMSR-6} and DMS-REQ-0344, there is also a potential problem with short exposures (e.g., 5 seconds) that can be processed by  the DIA pipeline, but pose a problem for the Alert Stream because the higher data acquisition rate (e.g., $\sim$6$\times$ higher) would cause a processing back-up.
-Should this situation arise, are there options for short-term increases in parallel processing power at NCSA?
-Such options might also apply to e.g., crowded fields with $>$10k {\tt DIASources}.
-\end{enumerate}
-
-$\bullet$ DMS-REQ-0322 specifies that data products from Special Programs processing shall be stored in separate (but joinable) databases.
-
-Section 1.6 "Special Programs" might benefit from the addition of discussion and/or requirements to address the following: \\
-$\bullet$ Should we add a general specification to the DMSR (\S~1.6) to describe whether and how data from Special Programs could be included in the WFD main survey Level 1 and/or 2 data products, as also raised for the LSR in \ref{LSR-0b}, and as discussed in Section \ref{ssec:dmplans_WFD}?
-\\
-$\bullet$ Should we add a specification to the DMSR (\S~1.6) to describe DM's plans to reconfigure its own pipelines for the processing of Special Programs data, perform and verify this processing, and deliver the unique products (separate but joinable to the WFD L1 and L2 products), as also raised for the LSR in \ref{LSR-0}, and as discussed in Section \ref{ssec:dmplans_reconfig}?
-\\
-$\bullet$ Since DM plans to include Special Programs data to the WFD main survey's L1 and L2 products wherever possible, and also run dedicated processing and generate unique products for all SP data, the associated computational and storage resources for Special Programs data might need to be multiplied by a factor of two.
-This is discussed more in Section \ref{sssec:dmplans_review_sizing}, and for now this unknown seems to be well within the uncertainty of the sizing models, and so probably does not (at this time) need to be addressed in the DMSR.
-
-$\bullet$ The following items are applicable to users processing data from Special Programs and/or reprocessing data from the WFD main survey.
-Since they're relevant to Special Programs, we mention them here, but they have not spawned any concerns.
-Section 2.9 "Level 3 Production" includes the requirements on access controls, external data, processing resource prioritization, consistency, provenance, and providing a software framework.
-Sections 3.1 "Software Architecture to Enable Community Re-Use" and 3.2 "Applications Software" describe aspects of the DM system that will be important to users, such as that the codes be extendable to both high-performance and desktop platforms, be able to handle simulated data and data from other instruments, and that a user interface will be provided.
-In Section 3.3 "Middleware Software", DMS-REQ-0298 requires that DM provide software to list and retrieve data, including raw data (3.3.2.2, Data Product and Raw Data Access), along with other user-access issues.
-
-$\bullet$ DMS-REQ-0312 and -0313 require that DM maintain a live L1 and the two most recent DR L1 and L2 catalogs for access by users, but there is no corresponding access requirement for the data products of Special Programs (4.1, Data Archive).
-\begin{enumerate}[topsep=-10pt,after=\vspace{10pt},label= \textbf{Concern \Roman*.},resume] \item \label{DMSR-10} Should there be a requirement for maintenance of user access to data products of Special Programs with the same timescales as the WFD main survey data, similar to DMS-REQ-0312, -0313?
-(This is related to \ref{LSR-1}) \end{enumerate}
-
-% % % % % % %
-\subsubsection{Data Management Applications Design, DMAD, LDM-151, \cite{LDM-151}}\label{sssec:dmplans_review_dmad}
-
-The DMAD specifies the design and implementation of the code and algorithms that will be used in the processing of images and creation of catalogs in the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-19.
-As the DMAD focuses on the content of the code and algorithms, it is relatively agnostic as to whether the source of the data is the WFD main survey or Special Programs -- but does (in some places) explicitly assume standard visit images of $2\times15$ seconds as the raw image input.
-In fact, the phrase "Special Programs" appears four times: three times as a source of reference catalogs or training sets, and once as a potential source of visits that have only a single $15$ second snap.
-It is the DMAD's codes and algorithms that DM will be {\it reconfiguring} to create pipelines for Special Programs -- any new algorithms that might enable science from Special Programs are considered beyond the scope of DM and will need to be developed as Level 3.
-This document will be handy to use in conjunction with Special Programs processing case studies (e.g., Section \ref{sec:SPCS}) that will be included with the next call for white paper proposals (Section \ref{ssec:data_comm}).
-
-The purpose of this study on Special Programs is not to suggest any changes to the DMAD.
-Instead, here we list a couple of examples of algorithms that sit at the threshold between DM-produced and Level 3.
-{\bf Deblending --} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
-It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
-Level 3 deblenders for specific Special Programs fields may require development by the user community.
-{\bf Variability Characterization --} The periodic and aperiodic variability characterizations described in Section 6.21 are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
-From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
-It is conceivable that the goals of a particular Special Program might require different algorithms, and these would be written as Level 3 and either made joinable to the DM reconfigured data products, or perhaps federated and incorporated directly.
-{\bf Photometric Redshifts --} As described in Section 5.6.5 the Level 2 DRP {\tt Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then federated and run at scale by DM.
-It is conceivable that the photo-$z$ algorithm for a Special Programs data set product might be different from that used for the WFD main survey.
-
-% % % % % % %
-\subsubsection{Data Products Definitions Document, DPDD, LSE-163, \cite{LSE-163}}\label{sssec:dmplans_review_dpdd}
-
-The DPDD describes the data products -- the images and catalogs -- to be delivered from the Level 1 AP, Level 2 DRP, and MOPS pipelines.
-We have used the version last revised 2017-07-01.
-Section 6 describes the data products that DM will provide for Special Programs.
-It specifies that the processing will use the same software stack as the Level 1 and 2, will make separate imaging products and separate (but joinable) catalogs, and use $\lesssim10\%$ of the computational and storage facility of the total LSST processing cluster (see Section \ref{sssec:dmplans_review_sizing}).
-
-The LSST Database Schema Browser\footnote{\url{http://lsst-web.ncsa.illinois.edu/schema/index.php?sVer=baseline}} is another way to explore the contents of the planned DM catalogs.
-We identify a couple of potential changes that might be necessary to the schema, which are related to this study on Special Programs.
-These are not high priority.
-
-$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used.
-This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
-DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} might need to have an element added that contains information about the template that was used to create the difference image, unless it can be shown that this can be handled by provenance (i.e., the code version and time of processing are sufficient to deduce which template was used).
-\end{enumerate}
-
-$\bullet$ The DMAD specifies that externally defined targets can be incorporated into the {\tt Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
-It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it "Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?
-\\
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an {\tt Object} is an externally provided coordinate.
-\end{enumerate}
-
-$\bullet$ As described in the DMAD and the DPDD, these are the {\tt Object} and {\tt DIAObject} elements that have been reserved for variability characterization: \\
-{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{sssec:dmplans_review_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
-Are 32 floats in each of the 6 filters always going to be a large enough volume?
-({\it MLG -- I suspect yes and that if not, compression can be used, but am leaving this note here anyway}).
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} DM might want to consider whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies.
-This might be similar to e.g., DMTN-049 for photometric redshifts included in the Level 2 DRP {\tt Object} catalog.
-\end{enumerate}
-
-% % % % % % %
-\subsubsection{Photometric Calibration for the LSST Survey, LSE-180, \cite{LSE-180}}
-
-LSE-180 is built on {\tt OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
-Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}.
-Regarding the reduction and calibration of non-standard visit imags, LSE-180 makes two relevant points: \\
-$\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
-$\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
-
-% % % % % % %
-\subsubsection{Computational Resources Sizing Documentation}\label{sssec:dmplans_review_sizing}
-
-In Section 6 of the DPDD, it is specified that the DM-provided processing of data from Special Programs will "use no more than $\lesssim10\%$ of computational and storage capacity of the LSST data processing cluster".
-This is based on the estimate that Special Programs data comprise $\sim10\%$ of the total observing time.
-Most Special Programs data will generate double the amount of processing and products: first, if and when they are incorporated into L1 and L2, and second, when their DM-provided reconfigured pipelines are run to generate independent and unique deep co-adds and catalogs.
-Based on this predicted multiple reprocessing, we might ask whether the "10\% of the computing resources" allocated for DM processing of Special Programs data is too small, but it is still a bit to early to know this.
-It will depend on the exact final mix of Special Programs, which may well change from year to year, and furthermore the current error bars on the sizing are of order $\pm10\%$.
-
-The Special Programs data also seems more likely than the WFD data to be processed a third time as part of science users Level 3 pipelines.
-LSST Science and Project Sizing Inputs spreadsheet (LSE-81) and explanation document (LSE-82) states that 10\% of the computing and storage resources estimated as necessary for the completion of the WFD Main Survey has been added to accommodate Level 3 data products.
-DM Compute Sizing spreadsheet (LDM-138) and explanation document (LDM-140) estimates the total compute resources needed for all processes, quoting the same 10\% addition for Level 3.
-DM Storage and I/O Sizing spreadsheet (LDM-141) and explanation document (LDM-139) estimates the total hardware needs, quoting the same 10\% addition for Level 3.
-As mentioned in Section \ref{sssec:dmplans_review_lsr}, LSR-REQ-0055 requires that {\tt userComputingFraction} = 10\% of the total LSST data processing capacity and storage space be allocated for user analysis and Level 3 pipelines and products.
-It is important to note that this 10\% is {\it in addition to} the sizing of $90\%+10\%$ for WFD and Special Programs, and the reason why Level 3 is also sized at $+10\%$ is independent of the amount of observing time spent on Special Programs.
-
-Regarding the possibly computationally intensive (e.g., shift-and-stack) but scientifically necessary processing that will need to be done for Special Programs data as Level 3, it is unclear whether the location for this has been identified; Mario asks {\it "whether this the same batch system we make available to the users for running Level 3 codes or the one that's used to process calibrations.
-Have these systems been sized?"}.
-The answer to this comes from K-T: {\it In the current model, the same production batch system is used for processing calibrations (CPP), processing "catch-up" L1, processing
-the L2 DRP, processing Special Programs as part of the L2 DRP, and processing large L3 jobs regardless of the type of data or processing.
-There is resource management and allocation applied, but there is flexibility and elasticity so that L3/DAC can "borrow" from L2 DRP as long as the integrated usage over a long time period (months, quarter, or year) does not exceed policy."}
-
-% % % % % % %
-\subsection{The Limited Documentation Regarding Image Detection Efficiencies}
-
-Characterizing the detection efficiencies will be just as important to Special Programs data as the Wide-Fast-Deep survey.
-Options such as planting fake sources may be both more manageable and more important to some Special Programs science goals.
-What, exactly, are the requirements on DM to provide detection efficiencies, and what are the current plans for providing this?
-Note that here we are talking about planting fakes in the single images for transient detection efficiencies with difference imaging analysis, and not planting fakes in the CoAdds for point-source limiting magnitudes).
-
-$\bullet$ The DPDD (LSE-163) does not have any specific data product related to detection efficiencies, but Section 3.2 "Image Characterization Data" does specify that {\it "Each processed image ..
-will record information on the pixel variance ...
-as well as the per-pixel masks ...
-These will allow the users to determine the validity and usefullness of each pixel in estimating the flux density recorded in that area of the sky"}.
-
-$\bullet$ The DMSR (LSE-61), Section 1.2.11 "Level 1 Data Quality Report Definition" (ID: DMS-REQ-0097): {\it "The DMS shall produce a Level 1 Data Quality Report that contains indicators of data quality that result from running the DMS pipelines, including at least ...
-detection efficiency for point sources vs.
-mag for each utilized filter."} However, this is a nightly data quality assessment and not a per-image product.
-
-$\bullet$ The DMAD (LDM-151), Section 5.6.3 "MakeSelectionMaps", states that this calibration step {\it "is responsible for producing multi-scale maps that describe LSST's depth and efficiency at detecting different classes of object.
-The details of what metrics will be mapped, the format and scale of the maps (e.g.
-hierarchical pixelizations vs.
-polygons), and the way the metrics will be computed are all unknown"}.
-It also states that this must be extendable to Level 3, but that {\it "the details of what DM will provide still needs to be clarified to the community"}, and notes that the reprocessing time for fake plants could be prohibitive.
-(Section 3 "Alert Production" also specifies that in LDM-151 {\it "we do not address estimation of the selection function for alert generation through the injection of simulated sources ...
-Source detection thresholds can be estimated through the use of sky sources"}.)
-
-$\bullet$ Suchyta et al.
-(2016) presents {\tt balrog}, a fake-embedding software package for detection efficiencies, and applies it in demonstration to DES data (it's not just for point sources, but uses {\tt galsim} to embed shapes built of e.g., Sersic profiles).
-
-$\bullet$ In a UW DM Brown Bag lunch meeting, it became clear that while there are many opinions on how to plant fake sources, not only are there no actual plans to do this, but there is no funding or FTQs allotted to figure out how the deliverable of detection efficiencies should be achieved.
-
-$\bullet \bullet \bullet$ {\bf Summary.} {\it MLG - Perhaps I could take this on as a future study, and assess our options, their costs, and the scientific payoff.
-Since this is not a question limited to DM processing of Special Programs data, specifically, I end this side note here and no {\bf Concerns} are spawned.}
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-\clearpage
-\section{Special Programs Processing Case Studies}\label{sec:SPCS}
-
-For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
-Note that we are not including any analysis in these descriptions, only processing and products.
-These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
-
-Basic steps that we use to describe a processing case study: \\
-Step 1. Data Acquisition. \\
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Step 3. Delivery of LSST Processed Images. \\
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-Step 6. Science User Level 3 Pipelines and Products. \\
-
-% % % % % % %
-\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
-
-This Special Programs processing summary is based on Becker et al.
-(2011) white paper to find TNOs with shift-and stack (SAS) \citep{BeckerWP}.
-See Section \ref{ssec:SPCS_SAS} for a more general processing case-study of a shift-and-stack Level 3 pipeline running regularly on a large amount of data.
-
-Step 1. Data Acquisition. \\ 
-The observational sequence is triggered.
-In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
-This sequence is always repeated 2-3 nights later.
-This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
-Data obtained in the $g$-band filter is also acceptable.
-\citep{BeckerWP}
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\ 
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
-Shift-and-stack processing is beyond the scope of DM's algorithms.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would {\it all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-The Level 3 pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform (this particular Special Program does not appear to need regular or urgent processing, and so its Level 3 code will probably not be federated).
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The Level 2 difference imaging routine will be used with the same template tract/patch for all.
-Custom, user-supplied algorithms will shift the exposures and create difference images, then Level 2 routines will stack and do source detection and characterization and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to the database.
-
-% % % % % % %
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
-
-Step 1. Data Acquisition. \\ 
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\ 
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\ 
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\ 
-The required data products for this science goal can be met by reconfiguring the DM pipelines.
-First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of {\tt DIAObject}, {\tt DIASource}, and {\tt Object} that are unique to this DDF.
-The LSST codes for alert packet and transport are used to distribute the detected objects e.g., to the same brokers that receive the Level 1 Alert Stream, or alternative destinations; however, these packets would need to be identified as having a source independent of main survey's Level 1 Alert Stream.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-1} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images?
-Or might there be issues due to train/test set mismatch?
-This is probably a question to be answered with testing, later, and incorporated into the DMAD?
-\end{enumerate}
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-2} Confirm that resources for packet distribution outside of the Alert Stream will be supported (related to \ref{DMSR-4}).
-\end{enumerate}
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\ 
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, it is unlikely that they would {\it all} be incorporated.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, no separate Level 3, user-driven software appears necessary.
-
-% % % % % % %
-\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
-
-Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
-The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
-
-Step 1. Data Acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures (i.e., 1 second).
-Location and exposure times are set by the sky brightness and desired saturation limits.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:data_DMconstraints}, \ref{LSR-2}, and \ref{DMSR-3}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds (see \ref{OSS-5} and \ref{DMSR-6b}), these data could be incorporated.
-
-Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-TBD, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-These data would not contribute to the Level 2 data products created for the WFD survey.
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-If short-exposure images cannot be processed with the existing DM algorithms, would the Level 3 user-built pipeline have to start from the raw data and create new, custom algorithms for e.g., ISR? 
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-3} Is it at all conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
-\end{enumerate}
-A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-This would certainly be considered a Level 3 algorithm.
-\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{SPCS-4} Are any DM algorithms being developed to estimate the flux of saturated stars using e.g., the PSF wings?
-Could the brighter-fatter issue be used for some science advantage?
-\end{enumerate}
-
-% % % % % % %
-\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
-
-Step 1. Data Acquisition. \\
-The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields, to a shallower depth than the WFD main survey.
-
-Step 2. Inclusion in Level 1 AP and the Alert Stream. \\
-Each $2\times15$ second visit is processed as Level 1 and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped, TBD (see Section \ref{ssec:data_DMconstraints} and \ref{sssec:dmplans_review_dmad}).
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  {\tt L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. DM-Provided Reconfigured Processing Pipelines and Products. \\
-The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Level 1 and 2 pipelines, and not much reconfiguration would be needed.
-The biggest difference might be the federation of a user-supplied deblender algorithm optimized for very crowded fields.
-
-Step 5. Inclusion in Level 2 DRP Products for WFD. \\
-It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-({\it MLG -- as I understand it, it is desirable from DM's perspective to have a single deblender algorithm used for the whole WFD main survey data products}).
-
-Step 6. Science User Level 3 Pipelines and Products. \\
-It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
-
 % % % % % % %
@@
 I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
-[this isn't in the requirements right now -- right now we say that images will become available in 24hrs.
-But I'd like to get K-T's reaction to this sort of proposal -- this kind of facility would be extremely powerful, and from a technical perspective my hunch is that once you have \#3, you get \#6 for free as well.]
-
-\clearpage
-\bibliography{ms,lsst,refs,books,refs_ads}
-
-\end{document}
-
-%\begin{center}
-%\includegraphics[width=8cm]{figures/}
-%\includegraphics[width=8cm]{figures/}
-%\caption{ \label{fig:}}
-%\end{center}
-%\end{figure}
+(This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_user}).

### 2018-06-18 DMTN-065.tex
fixed typo

@@
 Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Level 1 pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
 

### 2018-06-18 DMTN-065.tex
fixed typo

@@
 
-{\bf The LSST Project's Role in Processing Special Programs Data -- } The LSST Project will not take formal responsibility for specialized data reduction algorithms needed to process data, including images taken in non-standard modes.
+{\bf The LSST Project's Role in Processing Special Programs Data -- } The formal requirements regarding LSST's role in processing Special Programs data are in \citeds{LSE-61}, and the following statements have been derived from those requirements.
+The LSST Project will not take formal responsibility for specialized data reduction algorithms needed to process data, including images taken in non-standard modes.
 The term "specialized algorithms" refers to software that is not already within scope of the LSST Data Management (DM) science pipelines, and may include, for example: difference imaging for short exposures in which the PSF is not well-formed, shift-and-stack for faint moving objects, or any software with computational needs that significantly surpass the processing budget per image (compared to the processing of a WFD image).
@@
 {\bf Filter Changes -- } The maximum time for filter change is $120$ seconds: $30$ seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and $90$ seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
-Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4 exposures$ in a given filter would be obtained between filter changes -- but this is not actually a technical boundary.
+Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4$ exposures in a given filter would be obtained between filter changes -- but this is not actually a technical boundary.
 

### 2018-07-23 DMTN-065.tex
added LSE-61 flow-down statements
added non-sidereal tracked image processing statements

@@
 \setDocAbstract{This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys).
+The plans and description in this document flow down from the requirements in \citeds{LSE-61} regarding processing for Special Programs.
 The main target audience is the LSST Data Management (DM) team, but members of the community who are preparing white papers on science-driven observing strategies may also find this document useful.
@@
 
+{\bf Telescope Tracking --} The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
+This capability will include angular rates of up to $220$ arcseconds per second in both azimuth and elevation.
+
 {\bf Camera Rotation -- } There requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
@@
 
-\subsubsection{Exposure Times}
+\subsubsection{Exposure Times}\label{sssec:dmplans_NSV_expt}
 
@@
 These may or may not be appropriate for Special Programs data, depending on the science goals.
+
+\subsubsection{Twilight Images with a Bright Background}
+
+Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{sssec:dmplans_NSV_expt} also apply here.
+Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a User-Generated pipeline, is TBD (see also the example in Section \ref{ssec:SPCS_Twilight}).
+This may depend on the exposure time and the number of stars available in the image.
+
+\subsubsection{Images Obtained with Non-Sidereal Tracking}
+
+Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
+Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a User-Generated pipeline.
+The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.
 

### 2021-11-27 DMTN-065.tex
filter change/load as per DM-12573

@@
 {\bf Filter Changes -- } The maximum time for filter change is $120$ seconds: $30$ seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and $90$ seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
-Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4$ exposures in a given filter would be obtained between filter changes -- but this is not actually a technical boundary.
+Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4$ exposures in a given filter would be obtained between filter changes, but this is not actually a technical boundary.
+The filter change mechanism is designed to undergo a total of $100000$ changes over its lifetime, and each filter is designed to support up to $30000$ changes over its lifetime, where lifetime is $15$ years (an average of $17$ changes per night after accounting for filter changes during calibrations).
+During a given observing night, the system could support as many changes involving the 5 filters loaded in the carousel as desired (see filter change memorandum \url{ls.st/spt-494}).
 
-{\bf Filter Carousel Loads -- } As described in \citeds{Document-28382}, the filter carousel can hold five of the six LSST filters at a time, and filter loads are done in the day.
-The system is designed to support $3000$ loads and $100000$ filter changes in $15$ years, which is an average of $17$ changes per night (after accounting for filter changes during calibrations).
-Individual filters will support $30000$ changes in $15$ years.
-Based on these technical boundaries, we know that there will never be data in more than five filters in a given night.
+{\bf Filter Carousel Loads -- } The filter carousel can hold five of the six LSST filters at a time.
+The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
+Filter loads are only done in the day, and there will never be data in more than five filters in a given night.
 

### 2021-11-27 DMTN-065.tex
updated camera rotation as per DM-12573

@@
 {\bf Camera Rotation -- } There requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
-JIRA ticket DM-12573 is currently open and asking for clarification from the camera team on this.
-Until then, we assume there are no technical boundaries imposed by camera rotation constraints on the potential diversity of data products.
+Currently, there are no technical boundaries imposed by camera rotation constraints on the potential diversity of data products.
 

### 2021-11-27 DMTN-065.tex
removed references to DM-12574

@@
 
-\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds).
-The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) is under consideration (see JIRA ticket DM-12574).
+\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
+As with all requirements, this one would be verified during commissioning.
 Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
@@
 As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-The minimum exposure time for an image to be successfully processed by the Difference Imaging Analysis (DIA) pipeline is currently under consideration (see JIRA ticket DM-12574).
 
@@
 Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-A potential maximum exposure time from a processing perspective is currently under consideration (see JIRA ticket DM-12574).
 
@@
 When this source is again detected in image $2$, another new {\tt DIAObject} would be created if the catalog has not yet been updated.
-The best solution might be to flag {\tt Alerts} which may suffer from incomplete {\tt DIASource}--{\tt DIAObject} associations (see JIRA ticket DM-12574).
 

### 2021-11-27 DMTN-065.tex
removed reference to DM-12575

@@
 The above statements of intent are derived from the Data Management Subsystem Requirements document, \citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
-To ensure that the work-hours needed to reconfigure and test the pipelines, and run them and verify the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), have been included in the personnel budget, JIRA ticket DM-12575 is currently under consideration.
+Ensuring that the work-hours needed to reconfigure and test the pipelines, to run them and to verify and validate the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), is the role of Operations.
 

### 2021-11-27 DMTN-065.tex
all jira references removed now

@@
 DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
-A header keyword identifying an image as related to a Special Program would be sufficient to send it to a dedicated processing pipeline, and would satisfy this requirement.
-JIRA ticket DM-12576 is currently open to make sure that this happens.
+A header keyword identifying an image as related to a Special Program will be included (e.g., the survey name), and that will be sufficient to send it to a dedicated processing pipeline.
 
@@
 During Operations, there will be a method for the system to allocation processing resources in the case of over-subscription.
-JIRA ticket DM-12577 is currently open to inspire an investigation of whether additional Science Platform capabilities are needed to enable user-driven processing of Special Programs data.
 
@@
 An alternative option to ``adopted" code is ``adopted" data products: situations in which user-generated code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
-JIRA ticket DM-12578 is currently open to consider the staffing needs and process for adopting user-generated code or products; this issue is not necessarily limited to Special Programs.
 
@@
 However, these packets would not be distributed via the LSST {\tt Alert Stream}, and would need to be identified as, e.g., DDF Alerts.
-Note that JIRA ticket DM-12585 is currently open to investigate whether or not the internal real/bogus routine be able to run on a nightly CoAdd of deep drilling difference images.
 

### 2021-11-27 DMTN-065.tex
removed references to "MOPS" (deprecated)

@@
 \setDocChangeRecord{%
-\addtohist{1}{2017-11-14}{Status: internal working document.}{Melissa Graham}
+\addtohist{0}{2017-11-14}{Status: internal working document.}{Melissa Graham}
 \addtohist{1}{2018-06-17}{Updated to finalize and issue.}{Melissa Graham}
-%\addtohist{2}{yyyy-mm-dd}{Future changes}{Future person}
+\addtohist{2}{2021-12-01}{Updated per DM-20375.}{Melissa Graham}
 }
@@
 
-{\bf The Moving Object Processing System (MOPS) -- } Since MOPS takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can be ingested by MOPS.
-As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:data_prev}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.
+{\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
+As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:data_prev}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
 %There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
@@
 The cadence and patterns of these mini-surveys may differ from the WFD main survey, especially when very fast-moving objects are sought.
-From a processing perspective, it seems that many of these science goals will be achievable by using the products of the Moving Object Processing System (MOPS), which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night.
+From a processing perspective, it seems that many of these science goals will be achievable by using the products of Solar System Processing, which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night.
 The exception is finding faint SSOs (e.g., Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) through shift-and-stack (SAS) processing \citedsp{Document-11013}, because SAS is not a capability being built within the DM system and cannot be done solely by reconfiguring DM pipelines.

### 2021-11-27 DMTN-065.tex
added reference to RTN-008 (GW use-case)

@@
 \addtohist{1}{2018-06-17}{Updated to finalize and issue.}{Melissa Graham}
-\addtohist{2}{2021-12-01}{Updated per DM-20375.}{Melissa Graham}
+\addtohist{2}{2021-12-01}{Minor updates as per DM-20375.}{Melissa Graham}
 }
@@
 
-\textbf{Shorter Exposures.} The camera constraint on the minimum supported exposure time is currently 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
+\textbf{Shorter Exposures.}
+The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
 As with all requirements, this one would be verified during commissioning.
@@
 
+% % % % % % %
+\subsection{Gravitational Wave Event Follow-Up}\label{ssec:SPCS_GW}
+
+For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, sen \citeds{rtn-008}.
+
 \clearpage

### 2021-11-27 DMTN-065.tex
removed refs to white papers up to S3.1

@@
 \usepackage{latexsym}
-\usepackage{color}
+% \usepackage{color}
 % black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
@@
 
-\setDocAbstract{This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys).
+\setDocAbstract{
+This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys).
 The plans and description in this document flow down from the requirements in \citeds{LSE-61} regarding processing for Special Programs.
-The main target audience is the LSST Data Management (DM) team, but members of the community who are preparing white papers on science-driven observing strategies may also find this document useful.
+The main target audience is the LSST Data Management (DM) team, but members of the community who are planning to use Special Programs to reach their science goals may also find this document useful.
 The potential diversity of data from Special Programs is summarized, including boundaries imposed by technical limitations of the LSST hardware.
@@
 The remaining $10$--$15$\% of the time will be spent on Special Programs: alternative survey areas and/or observing strategies driven by a specific science goal.
-A call for white papers that provide scientific motivation and observing strategies for the WFD survey and Special Programs, \citeds{Document-28382}, released in June 2018, has a deadline of November 30 2018.
+%A call for white papers that provide scientific motivation and observing strategies for the WFD survey and Special Programs, \citeds{Document-28382}, released in June 2018, has a deadline of November 30 2018.
 It is conceivable that Special Programs might obtain imaging data that are significantly different from the WFD main survey, and/or requires special processing in order to achieve the program's science goals.
@@
 
-{\bf Document Overview -- } The purpose of this document is to assess the potential diversity of imaging data that might be obtained by Special Programs (Section \ref{sec:data}), and to explore and clarify whether -- and to what extent -- DM's planned pipelines and user services will be able to handle this diversity of data (Section \ref{sec:dmplans}).
-As described in the call for white papers on cadence optimization (\citeds{Document-28382}), Special Programs that may require specialized or computationally-intense algorithms to meet their science goals are required to describe how these processing needs will be met.
-Sections \ref{sec:data} and \ref{sec:dmplans} have been designed to help white paper authors figure out whether their proposed Science Program(s) require processing that would be considered out of DM's scope, and in Section \ref{sec:SPCS} we have provided a series of science-driven case studies as examples.
+{\bf Document Overview -- }
+The purpose of this document is to assess the potential diversity of imaging data that might be obtained by Special Programs (Section \ref{sec:data}), and to explore and clarify whether -- and to what extent -- DM's planned pipelines and user services will be able to handle this diversity of data (Section \ref{sec:dmplans}).
+Special Programs that may require specialized or computationally-intense algorithms to meet their science goals are required to describe how these processing needs will be met.
+Sections \ref{sec:data} and \ref{sec:dmplans} have been designed to help the science community figure out whether proposed Science Program(s) require processing that would be considered out of DM's scope, and in Section \ref{sec:SPCS} we have provided a series of science-driven case studies as examples.
 
@@
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
 % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %

### 2021-11-28 DMTN-065.tex
removed refs to 2018 white papers

@@
 This document provides an in-depth description of the role of the LSST Project in preparing to process the data from Special Programs (expanding on the general description in Section 6 of \citeds{LSE-163}).
-The main target audience is the LSST Data Management (DM) team, but members of the science community who are preparing white paper proposals may also find this document useful.
+The main target audience is the LSST Data Management (DM) team, but members of the community who are planning to use Special Programs to reach their science goals may also find this document useful.
 
@@
 To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-
-No user-generated pipeline may contribute Alerts to the Alert Stream, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
+No user-generated pipeline may contribute Alerts to the Alert Stream.
+%, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
 
@@
 For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-This kind of thought experiment of describing the reductions and processing could also be a required section of all future white paper proposals.
 Note that we are not including any analysis in these descriptions, only processing and products.

### 2022-02-17 DMTN-065.tex
major rewrites

@@
 
-\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, and E.~Bellm}
+\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~Guy, and the Data Management System Science Team}
 
@@
 \setDocAbstract{
-This document provides an in-depth description of the role of the LSST Project in preparing software and providing computational resources to process the data from Special Programs (deep drilling fields and/or mini-surveys).
-The plans and description in this document flow down from the requirements in \citeds{LSE-61} regarding processing for Special Programs.
-The main target audience is the LSST Data Management (DM) team, but members of the community who are planning to use Special Programs to reach their science goals may also find this document useful.
-The potential diversity of data from Special Programs is summarized, including boundaries imposed by technical limitations of the LSST hardware.
-The capability of the planned Data Management system to processes this diversity of Special Programs data is the main focus of this document.
-Case studies are provided as examples of how the LSST software and/or user-generated pipelines may be combined to process the data from Special Programs.}
+The core LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, which is expected to be accomplished with 85--90\% of the observing time.
+The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
+
+This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs (including the flow-down of related requirements, and the capabilities of the LSST Science Pipelines and the Rubin Science Platform).
+
+The specialized, science-specific aspects of Special Programs processing and analysis that are best left to the science community are also described, and illustrated with case-study examples.
+
+The main target audience of this document is Rubin Observatory staff -- the construction-era Data Management (DM) team and the operations-era Data Production (RDP) team -- but members of the science community who are planning to use Special Programs to reach their science goals may also find this document useful.
+}
 
@@
 \addtohist{1}{2018-06-17}{Updated to finalize and issue.}{Melissa Graham}
-\addtohist{2}{2021-12-01}{Minor updates as per DM-20375.}{Melissa Graham}
+\addtohist{2}{2021-12-01}{Updates per DM-20375.}{Melissa Graham}
 }
@@
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Introduction} \label{sec:intro}
+% % % % % % % % % % % % % % % % % % 
+\section{Terms and Definitions} \label{sec:terms}
 
-The main LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, but this is expected to be accomplished with $85$--$90$\% of the observing time available over the $10$ year survey.
-The remaining $10$--$15$\% of the time will be spent on Special Programs: alternative survey areas and/or observing strategies driven by a specific science goal.
-%A call for white papers that provide scientific motivation and observing strategies for the WFD survey and Special Programs, \citeds{Document-28382}, released in June 2018, has a deadline of November 30 2018.
-It is conceivable that Special Programs might obtain imaging data that are significantly different from the WFD main survey, and/or requires special processing in order to achieve the program's science goals.
-This document provides an in-depth description of the role of the LSST Project in preparing to process the data from Special Programs (expanding on the general description in Section 6 of \citeds{LSE-163}).
-The main target audience is the LSST Data Management (DM) team, but members of the community who are planning to use Special Programs to reach their science goals may also find this document useful.
+\subsection{Wide-Fast-Deep (WFD) Main Survey}\label{ssec:terms_wfd}
+The WFD Main Survey is the core science program of the LSST, designed to achieve the science goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
+The WFD will cover at least $\sim$18000 deg$^2$ and is expected to be accomplished with 85--90\% of the observing time.
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the WFD Main Survey is thought of as several contiguous areas.
+The largest is the low-dust-extinction survey area, which would also receive the most ($>$800) visits over the 10-year survey and be completed to uniform depth, likely with a rolling cadence.
+Smaller regions cover the higher-extinction high stellar density Galactic Bulge and Magellanic Clouds, the North Ecliptic Spur (NES), and the remainder of the Galactic Plane and the South Celestial Pole.
+Boundaries, re-visit cadences, and depths for all WFD areas remain to be determined \citedsp{PSTN-053}.
 
-{\bf Special Programs Terminology -- } The WFD main survey will survey the sky with a series of {\bf visits}, and each field will be visited $>800$ times over $10$ years.
-A {\bf standard visit} is composed of $2\times15$ second exposures (commonly referred to as ``snaps") and an {\bf alternative standard visit} is composed of a single $30$ second exposure.
-A {\bf non-standard visit} is any other exposure time(s) or number of snaps.
-Special Programs are typically divided into two types: {\bf Deep Drilling}, a single pointing for which many exposures are obtained in a relatively short amount of time (e.g., $>2\times$ as many visits in six months as the WFD will obtain in 10 years); and {\bf Mini-Surveys}, which refer to either new sky areas observed with a WFD-like survey, or sky areas within the WFD but observed with a specialized strategy.
+\subsubsection{Mini-Surveys}\label{sssec:terms_wfd_mini}
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``mini-survey" was being used to refer to the areas of the WFD / Main Survey outside of the large, low-dust-extinction survey area.
+This term used to refer to non-WFD regions (Special Programs).
+
+\subsection{Special Programs}\label{ssec:terms_sp}
+Special Programs are additional survey areas and/or observing strategies that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
+Special Programs will fill the remaining 10--15\% of the available survey time, and include both ``deep drilling" fields and ``micro-surveys".
+
+\subsubsection{Deep Drilling Field (DDF)}\label{sssec:terms_sp_ddf}
+A DDF is a single pointing for which many exposures are obtained during the night.
+The five DDFs (below) are expected to take $\sim$5\% of the total available time.
+Generally, the LSST observing strategy for a DDF is to obtain some or all of these exposures consecutively, and to maintain a high inter-night cadence over a short season (e.g., returning every two nights over four months, \citealt{2019ApJ...873..111I}) -- but the exact strategies and total time for the LSST DDFs remain to be determined \citedsp{PSTN-053}.
+
+\begin{itemize}
+\item Elias S1 (00:37:48, -44:00:00)
+\item XMM-LSS (02:22:50, -04:45:00)
+\item Extended Chandra Deep Field-South (03:32:30, -28:06:00)
+\item COSMOS (10:00:24, +02:10:55)
+\item Euclid Deep Field South  (04:04:58, -48:25:23)\footnote{\url{https://www.cosmos.esa.int/web/euclid/euclid-survey}}
+\end{itemize}
+
+\subsubsection{Micro-Surveys}\label{sssec:terms_sp_micro}
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``micro-survey" refers to either new sky areas observed with a WFD-like survey, or sky areas within the WFD but observed with a specialized strategy.
+Micro-surveys are expected to take up the remaining $\sim$10\% of the total available survey time over LSST's 10 years, and those which would use $>$0.3\% of the time and are currently being considered by the SCOC are listed below (see also \citeds{PSTN-053}).
+
+\begin{itemize}
+\item short-exposure twilight visits (for near-Sun and near-Earth objects; NEOs)
+\item a static short-exposure (5 sec) map of the sky in $ugrizy$ in the first year (for calibration)
+\item an extended short-exposure survey of the sky in $ugrizy$ (for transient detection and static-sky calibration)
+\item target-of-opportunity (ToO) follow-up (to identify optical counterparts to gravitational wave sources)
+\item coverage of the Roman microlensing bulge field (potentially as a DDF)
+\item deeper $g$-band imaging of 10 local volume galaxies
+\item a high-cadence survey of 2 fields in the SMC (for microlensing)
+\item annual week-long surveys of the Carina nebula and surrounding star-forming regions
+\item a limited-visit (i.e., shallow) ``northern stripe" survey to declination $<$+30 degrees
+\item a survey of the Virgo cluster to WFD depth
+\end{itemize}
+
+Note that there remains some grey area between what is a mini-survey within the WFD, and what is a WFD-like micro-survey.
+For example, the last two on the list above (the northern stripe and the Virgo cluster) could also (or might be in the future) considered WFD mini-surveys and not Special Programs.
+
+\subsection{Visit Types}\label{ssec:terms_visits}
+A visit is an observation of a single pointing at a given time, of which there are three types as listed below.
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, most WFD and DDF visits would be standard or alternative standard visits.
+However, non-standard visits with longer exposure times are being considered, especially for $u$-bands, and several of the micro-surveys are considering shorter exposure times.
+The full potential for diversity in Special Programs data is reviewed in Section~\ref{ssec:proc_datadiv}.
+
+\begin{itemize}
+\item Standard Visit -- Composed of $2\times15$ second exposures (commonly referred to as ``snaps").
+\item Alternative Standard Visit -- Composed of a single $30$ second exposure.
+\item Non-Standard Visit -- Any other exposure time(s) or number of snaps.
+\end{itemize}
+
+\clearpage
+% % % % % % % % % % % % % % % % % % 
+\section{Reduction and Processing for Special Programs Data Products}\label{sec:proc}
+
+\subsection{Requirements Related to Special Programs}
+
+{\it Use the docrev.tex file to update this.}
+
+\citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
+
+\subsection{Special Programs in the DPDD}
+
+{\it Just quote what the DPDD says. Maybe reference any relevant RFC/LCR that made those changes?}
+
+\subsection{The Potential Diversity of Special Programs Data} \label{ssec:proc_datadiv}
+
+As discussed in Section~\ref{sec:terms}, most of the Special Programs that are currently under consideration will use standard or alternative standard visits.
+However, some are likely to require non-standard visits with shorter exposures (5 sec).
+Furthermore, some are likely to acquire images with a significantly brighter sky background (e.g., twilight images) than most nighttime survey images.
+
+The cadence and patterns of Special Programs might also differ from the WFD, such as long series of exposures obtained of the same field (e.g., DDFs), or a strategy optimized to find very fast-moving objects (e.g., NEOs).
+
+It does not appear that any of the currently-proposed Special Programs are likely to violate boundaries imposed by the Rubin Observatory hardware, but there remain a few open questions about boundaries on data processing imposed by the LSST Science Pipelines.
+
+\subsubsection{Hardware Boundaries}
+
+Appendix~\ref{sec:hardbounds} lists all of the hardware boundaries that might constrain the potential diversity of Special Programs data.
+
+The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is currently a hardware boundary that limits the readout rate to 1 every 15 seconds.
+This would affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys.
+
+Hardware imposes no other boundaries on how data can be obtained, but Special Programs that request a high number of filter changes and/or long slews could be inefficient due to large overheads.
+
+\subsubsection{Processing Boundaries}
+
+Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits can be processed and calibrated by the LSST Science Pipelines, which are designed to process standard (or alternative standard) visits.
+
+Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
+
+It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
+
+The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.
+
+\subsection{Anticipated Rubin-Processed Special Programs Data Products}
 
@@
 
-{\bf Document Overview -- }
-The purpose of this document is to assess the potential diversity of imaging data that might be obtained by Special Programs (Section \ref{sec:data}), and to explore and clarify whether -- and to what extent -- DM's planned pipelines and user services will be able to handle this diversity of data (Section \ref{sec:dmplans}).
-Special Programs that may require specialized or computationally-intense algorithms to meet their science goals are required to describe how these processing needs will be met.
-Sections \ref{sec:data} and \ref{sec:dmplans} have been designed to help the science community figure out whether proposed Science Program(s) require processing that would be considered out of DM's scope, and in Section \ref{sec:SPCS} we have provided a series of science-driven case studies as examples.
+Whether or not Special Programs images are incorporated into the WFD main survey's data products, it is anticipated that most of the Special Program's science goals will require (or benefit from) separate data products (i.e., CoAdds and/or catalogs).
+For this reason, LSST intends to reconfigure the DM's pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
+In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
 
-% In Appendix \ref{sec:docrev}, we endeavor to ensure that the DM's plans for Special Programs processing are appropriately expressed in all of the major Project documents and that the computational system and human work-hours needed to accomplish Special Programs processing have been accurately evaluated.
-This content will only be of interest to the DM team.
+In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
+For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the \DPDD, \citedsp{LSE-163}).
+This type of reconfiguration would also be possible to create as a user-generated pipeline (Section \ref{ssec:dmplans_user}), but having these products provided by the Project ensures a consistent and verified level of quality.
+%, as well as access for all users to the processed Special Programs data products, which would only increase the scientific value of Special Programs data.
+--> MLG: data rights doc will probably state that the products for SP data that is only processed by user-generated pipelines will have to be served to all users
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\clearpage
-\section{The Potential Diversity of Special Programs Data} \label{sec:data}
+The above statements of intent are derived from the Data Management Subsystem Requirements document, \citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
+Ensuring that the work-hours needed to reconfigure and test the pipelines, to run them and to verify and validate the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), is the role of Operations.
 
-To define the extent to which Data Management will be able to process data from Special Programs, a comprehensive understanding of the potential diversity of data from Special Programs -- compared to the WFD's sky survey of standard visits -- is needed.
+DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
+A header keyword identifying an image as related to a Special Program will be included (e.g., the survey name), and that will be sufficient to send it to a dedicated processing pipeline.
 
-To build up this comprehensive understanding, in Section \ref{ssec:data_bounds} we considered any technical limitations that the facility and instrumentation will or might place on the data.
-We find that the hardware imposes few boundaries on how data can be obtained, but that a high number of filter changes and/or long slews are inefficient due to their large overheads.
-The minimum exposure time is $1$ second (stretch goal: $0.1$ seconds), but there is currently a technical boundary that limits the readout rate to $1$ every $15$ seconds.
+\subsection{Anticipated User-Generated Special Programs Data Products}\label{ssec:dmplans_user}
 
-Next, we considered the Special Programs that have been openly discussed so far in the Science Community (Appendix \ref{sec:data_prev}).
-Based on this consideration, most of the proposed Special Programs are likely to use visits that are similar to the WFD Main Survey, but some will require exposures that are significantly shorter, or that are obtained with a bright sky background during twilight.
-The cadence and patterns may also differ from the WFD main survey, such as long series of exposures obtained of the same field (i.e., deep drilling), or a strategy optimized to find very fast-moving objects.
-In addition, images of very crowded fields in the Galactic Plane may be included (at all or more often than the WFD Main Survey).
-It does not appear that any of the previously proposed Special Programs violate the hardware-imposed boundaries discussed in Section \ref{ssec:data_bounds}.
+In cases where the science goals of a Special Program require specialized algorithms and cannot be achieved by reconfiguring DM's software, then user-generated pipelines will be needed.
+Towards this end, LSST DM is making all of its software open-source, and preparing the Science Platform (\citeds{LSE-319}), through which users can access the tools and computational resources to assemble data processing pipelines to achieve their science goals (whether related to Special Programs data or not).
+During Operations, there will be a method for the system to allocation processing resources in the case of over-subscription.
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\subsection{Hardware Boundaries}\label{ssec:data_bounds}
+If the user-generated processing pipeline for Special Programs data requires requires significantly more computational resources than have been allocated -- where that allocation has been sized approximately, based on image processing for WFD main survey data (i.e., difference imaging, source detection, and/or stacking) -- then external computational resources may be necessary.
+To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
+No user-generated pipeline may contribute Alerts to the Alert Stream.
+%, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
 
-Here we consider the technical boundaries on the diversity of data products that are expected to (or may) be imposed by limitations from the camera, telescope, and/or site (boundaries imposed by DM processing capabilities are considered in Section \ref{sec:dmplans}).
+It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
+For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of the Prompt and DRP data products.
+An alternative option to ``adopted" code is ``adopted" data products: situations in which user-generated code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
 
-{\bf Filter Changes -- } The maximum time for filter change is $120$ seconds: $30$ seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and $90$ seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
-Assuming that most Special Programs would be designed to keep overheads $<100\%$ and would be using standard $30$ second visits, the filter change time indicates that it is likely that at least $4$ exposures in a given filter would be obtained between filter changes, but this is not actually a technical boundary.
-The filter change mechanism is designed to undergo a total of $100000$ changes over its lifetime, and each filter is designed to support up to $30000$ changes over its lifetime, where lifetime is $15$ years (an average of $17$ changes per night after accounting for filter changes during calibrations).
-During a given observing night, the system could support as many changes involving the 5 filters loaded in the carousel as desired (see filter change memorandum \url{ls.st/spt-494}).
-
-{\bf Filter Carousel Loads -- } The filter carousel can hold five of the six LSST filters at a time.
-The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
-Filter loads are only done in the day, and there will never be data in more than five filters in a given night.
-
-{\bf Exposure Times -- } The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
-The maximum exposure time is not restricted.
-The readout time is $2$ seconds, and would be significant overhead on short exposures.
-Images with exposure times $<15$ seconds may still have to be separated by $15$ seconds for thermal tolerance; i.e., that the minimum readout rate is one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
-We therefore consider the $15$ interval between images a technical boundary on the potential diversity of data products.
-% $\bullet$ However, for exposure times there are other considerations.
-The minimum exposure time needed to create an image with a PSF that is well-formed enough for difference imaging is a separate question.
-Changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming a 1 second exposure can be reduced and calibrated, its detected point sources will span $13 < r < 21$ magnitudes, whereas a 15 second exposure saturates at $r\sim15.8$ mag.
-A 150 second image would saturate at $r\sim18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Additionally, the impact on CR rejection routines is untested for long exposures.
-
-{\bf Telescope Slew -- } As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no technical boundaries on the size of a single slew or the accrued slew distance.
-
-{\bf Telescope Tracking --} The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
-This capability will include angular rates of up to $220$ arcseconds per second in both azimuth and elevation.
-
-{\bf Camera Rotation -- } There requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
-Currently, there are no technical boundaries imposed by camera rotation constraints on the potential diversity of data products.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\clearpage
-\section{Processing Special Programs Data with LSST DM Pipelines} \label{sec:dmplans}
-
-In this section we go into greater detail regarding the Projects' role in processing data from Special Programs, as introduced in Section \ref{sec:intro}.
-We discuss the ability of the planned DM pipelines to: process diverse imaging data that are unlike the WFD's standard visits in Section \ref{ssec:dmplans_NSV}; incorporate Special Programs data into the Prompt and Data Release pipelines and data products for the WFD main survey in Section \ref{ssec:dmplans_WFD}; reconfigure the pipelines and generate unique sets of data products for each Special Program in Section \ref{ssec:dmplans_reconfig}; and enable user-generated pipelines and data products in Section \ref{ssec:dmplans_user}.
-
-% % % % % % % % % % % % % % % % % %
-\subsection{Processing Diverse Imaging Data Unlike the WFD Main Survey}\label{ssec:dmplans_NSV}
-
-As described in Section \ref{sec:data}, Special Programs will be able -- and likely -- to request observing modes with shorter (or longer) exposure times, long sequences of visits to the same field, and/or imaging of very crowded fields.
-We review the capability of DM's planned pipelines to process such each of these kinds of diverse data, keeping in mind that the processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters such as the number of stars with sufficient flux for photometric calibration or the source area density (crowding).
-Furthermore, that these boundaries imposed by the data quality might not be constrained until the final performance of DM's algorithms, as described in the Data Management Applications Design (\citeds{LDM-151}) document, is fully characterized.
-
-\subsubsection{Exposure Times}\label{sssec:dmplans_NSV_expt}
-
-Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-We discuss shorter and longer exposure in turn.
-
-\textbf{Shorter Exposures.}
-The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
-As with all requirements, this one would be verified during commissioning.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes.
-A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
-As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
-\end{center}
-\end{figure}
-
-\textbf{Longer Exposures.} There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, and cosmic-ray rejection completeness might suffer (unknown), which could impact the quality of a difference image and the detected sources.
-Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute to the Level 1 and Alert Stream.
-One possibly useful study is Chang et al.
-(2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-They present a new software package to do mitigate the effects.
-Alternatively, we may need to use software packages \texttt{PhoSim} (Peterson et al.
-2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} to at least simply characterize the PSF stability as a function of exposure time.
-% In conversation with K.-T., exposure times of 10 to 120 seconds will not be a problem for the processing pipeline to accommodate, but the ingest rate of images and the number of exposure per visit are the aspects that could cause issues (discussed below).
-% Is it conceivable that the LSST would spend time on a Special Program that will acquire short exposures if DM's algorithms cannot be shown to adequately reduce and calibrate them?
-This is probably best left ot a policy issue (beyond the scope of this document).
-
-\subsubsection{Number of Exposures per Visit (Long Sequences of a Single Field)}
-
-There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
-From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.
-
-% K.-T.
-Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
-This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
-Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
-things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
-The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
-To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
-Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
-\textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
-
-% K.-T.
-has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
-One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
-The processing pipeline would need to know how to divide this sequence up into visits.
-As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
-
-\subsubsection{Images in Very Crowded Fields}
-
-The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
-The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields.
-These may or may not be appropriate for Special Programs data, depending on the science goals.
-
-\subsubsection{Twilight Images with a Bright Background}
-
-Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{sssec:dmplans_NSV_expt} also apply here.
-Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a User-Generated pipeline, is TBD (see also the example in Section \ref{ssec:SPCS_Twilight}).
-This may depend on the exposure time and the number of stars available in the image.
-
-\subsubsection{Images Obtained with Non-Sidereal Tracking}
-
-Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
-Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a User-Generated pipeline.
-The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.
-
-% % % % % % % % % % % % % % % % % %
 \subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:dmplans_WFD}
 
-As described in Section \ref{sec:intro}, the Project may incorporate Special Programs data into the WFD main survey's pipelines and data products whenever this is (1) possible and (2) scientifically beneficial.
+The Project may incorporate Special Programs data into the WFD main survey's pipelines and data products whenever this is (1) possible and (2) scientifically beneficial.
 This will likely be at the discretion of the data quality assessment team during Operations.
@@
 Only images that can be processed with the Prompt DIA pipeline with the $60$-second timeline can contribute to the Alert Stream.
-As discussed in Section \ref{sec:data}, this might prohibit exposures shorter than $<15$ seconds and/or visit cadences shorter than $1$ per $\sim30$ seconds.
+As discussed in Section \ref{ssec:proc_datadiv}, this might prohibit exposures shorter than $<15$ seconds and/or visit cadences shorter than $1$ per $\sim30$ seconds.
 This might also prohibit the inclusion of very crowded fields that require more computational resources.
@@
 {\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
-As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:data_prev}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
+As discussed under "Solar System Objects (SSO)" in Appendix \ref{ssec:proc_datadiv_prev}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
 %There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
@@
 
+\clearpage
 % % % % % % % % % % % % % % % % % % 
-\subsection{Reconfiguring DM Pipeline Components for Special Programs}\label{ssec:dmplans_reconfig}
+\section{Enabling the Discovery and Analysis of Special Programs Data Products}\label{sec:analysis}
 
-Whether or not Special Programs images are incorporated into the WFD main survey's data products, it is anticipated that most of the Special Program's science goals will require (or benefit from) separate data products (i.e., CoAdds and/or catalogs).
-For this reason, LSST intends to reconfigure the DM's pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
-In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
+\subsection{Anticipated Rubin Science Platform Capabilities for Special Programs }
 
-In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the \DPDD, \citedsp{LSE-163}).
-This type of reconfiguration would also be possible to create as a user-generated pipeline (Section \ref{ssec:dmplans_user}), but having these products provided by the Project ensures a consistent and verified level of quality.
-%, as well as access for all users to the processed Special Programs data products, which would only increase the scientific value of Special Programs data.
---> MLG: data rights doc will probably state that the products for SP data that is only processed by user-generated pipelines will have to be served to all users
+\begin{itemize}
+\item discoverability of SP data when browsing an all-sky map
+\end{itemize}
 
-The above statements of intent are derived from the Data Management Subsystem Requirements document, \citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
-Ensuring that the work-hours needed to reconfigure and test the pipelines, to run them and to verify and validate the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), is the role of Operations.
-
-DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
-A header keyword identifying an image as related to a Special Program will be included (e.g., the survey name), and that will be sufficient to send it to a dedicated processing pipeline.
-
+\clearpage
 % % % % % % % % % % % % % % % % % %
-\subsection{Support for User-Driven Processing of Special Programs Data}\label{ssec:dmplans_user}
-
-In cases where the science goals of a Special Program require specialized algorithms and cannot be achieved by reconfiguring DM's software, then user-generated pipelines will be needed.
-Towards this end, LSST DM is making all of its software open-source, and preparing the Science Platform (\citeds{LSE-319}), through which users can access the tools and computational resources to assemble data processing pipelines to achieve their science goals (whether related to Special Programs data or not).
-During Operations, there will be a method for the system to allocation processing resources in the case of over-subscription.
-
-If the user-generated processing pipeline for Special Programs data requires requires significantly more computational resources than have been allocated -- where that allocation has been sized approximately, based on image processing for WFD main survey data (i.e., difference imaging, source detection, and/or stacking) -- then external computational resources may be necessary.
-To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-No user-generated pipeline may contribute Alerts to the Alert Stream.
-%, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
-
-It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
-For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of the Prompt and DRP data products.
-An alternative option to ``adopted" code is ``adopted" data products: situations in which user-generated code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
-
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\clearpage
-\section{Special Programs Processing Case Studies}\label{sec:SPCS}
+\section{Special Programs Processing Examples}\label{sec:SPCS}
 
@@
 
-% % % % % % %
 \subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
@@
 
-% % % % % % %
+%\input{spcs_sas}
+
 \subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
@@
 
-% % % % % % %
 \subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
@@
 Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{ssec:dmplans_NSV}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.
+Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.
 
@@
 
-% % % % % % %
 \subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
@@
 
-%\input{spcs_sas}
-
-% % % % % % %
 \subsection{Gravitational Wave Event Follow-Up}\label{ssec:SPCS_GW}
 
-For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, sen \citeds{rtn-008}.
+For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, see \citeds{rtn-008}.
 
 \clearpage
+% % % % % % % % % % % % % % % % % %
 \bibliography{local,lsst,refs,books,refs_ads}
@@
 \clearpage
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
+% % % % % % % % % % % % % % % % % %
 \appendix
 
-% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % %
-\section{Previously Proposed Special Programs}\label{sec:data_prev}
+% % % % % % % % % % % % % % % % % %
+\section{Potential Hardware Boundaries on Data Diversity}\label{sec:hardbounds}
+
+The potential boundaries on the diversity of data products that could be imposed by limitations from the Rubin Observatory hardware -- camera, telescope, and/or site -- are considered.
+
+\subsection{Filter Changes}
+The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
+Assuming that most Special Programs would be designed to keep overheads $<$100\% and would be using standard 30 second visits, the filter change time indicates that it is likely that at least 4 exposures in a given filter would be obtained between filter changes, but this is not actually a hardware boundary.
+The filter change mechanism is designed to undergo a total of 100000 changes over its lifetime, and each filter is designed to support up to 30000 changes over its lifetime, where lifetime is 15 years.
+That is an average of $\sim$27 changes per day, some of which would occur in they day during calibrations (estimate, $\sim$10) and the rest at night.
+As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could support as many changes involving the 5 filters loaded in the carousel as desired, without any practical limitation"}.
+
+\subsection{Filter Carousel Loads}
+The filter carousel can hold five of the six LSST filters at a time.
+The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
+Filter loads are only done in the day, and there will never be data in more than five filters in a given night.
+
+\subsection{Exposure Times}
+The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
+The maximum exposure time is not restricted.
+
+\subsection{Readout Time}
+The readout time is $2$ seconds, and would be significant overhead on short exposures.
+
+\subsection{Inter-Image Time}
+Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
+The $15$ interval between images is a potential hardware boundary on the potential diversity of data products.
+
+\subsection{Telescope Slew}
+As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no hardware boundaries on the size of a single slew or the accrued slew distance.
+
+\subsection{Telescope Tracking}
+The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
+This capability will include angular rates of up to 220 arcseconds per second in both azimuth and elevation.
+
+\subsection{Camera Rotation}
+The requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
+Currently, there are no hardware boundaries imposed by camera rotation constraints on the potential diversity of data products.
+
+% % % % % % % % % % % % % % % % % %
+\section{Potential Processing Boundaries on Data Diversity}\label{sec:procbounds}
+
+The capability of the LSST Science Pipelines to process diverse data is explored below.
+
+Note that processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters (e.g., the number of stars with sufficient flux for photometric calibration).
+Furthermore, the processing boundaries might not be fully constrained until the final performance of the LSST Science Pipelines, as described in the Data Management Applications Design, \citeds{LDM-151}) document, is fully characterized.
+
+{\bf Summary of the conclusions below:}\\
+Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
+It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
+The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.
+
+\subsection{Exposure Times}\label{ssec:procbounds_expt}
+
+Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+
+\subsubsection{Short Exposures (Non-Standard Visits of $<$30 sec)}
+The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
+
+However, for exposure times there are other considerations, as changing the exposure time also affects the photometric and astrometric calibrations.
+Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r$$\approx$ 13--21 magnitudes.
+A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but this still leaves stars between 15.8--21.0 magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
+
+In order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
+As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
+
+\begin{figure}
+\begin{center}
+\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
+\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
+At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
+We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
+\label{fig:expt}}
+\end{center}
+\end{figure}
+
+% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute Alert Production.
+% One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
+% They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
+% They also present a new software package to do mitigate the effects.
+% Software packages \texttt{PhoSim} (Peterson et al.
+2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} could be used to characterize the PSF stability as a function of exposure time.
+
+\subsubsection{Long Exposures (Non-Standard Visits of $>$30 sec)}
+
+There is no maximum exposure time specified for an LSST image.
+Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
+However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
+Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures (unknown), which could impact the quality of a difference image and the detected sources.
+Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+
+\subsection{Twilight Images with a Bright Background}
+
+Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also apply here.
+Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a user generated pipeline, remains to be determined (see also the example in Section \ref{ssec:SPCS_Twilight}).
+This may depend on the exposure time and the number of stars available in the image.
+
+\subsection{Images Obtained with Non-Sidereal Tracking}
+
+Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
+Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a user generated pipeline.
+The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.
+
+%%% MLG removed the following in Feb 2022, it's no longer a concern.
+% \subsection{Number of Exposures per Visit (Long Sequences of a Single Field)}
+
+% There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
+% From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.
+
+% K.-T.
+Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
+This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
+Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
+things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
+The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
+To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
+Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
+\textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
+
+% K.-T.
+has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
+One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
+The processing pipeline would need to know how to divide this sequence up into visits.
+As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
+
+%%% MLG removed the following in Feb 2022, it's no longer a concern for Special Programs (crowded fields are in the WFD).
+% \subsection{Images in Very Crowded Fields}
+
+% The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
+% The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.
+
+\clearpage
+% % % % % % % % % % % % % % % % % %
+\section{Previously Proposed Special Programs}\label{ssec:proc_datadiv_prev}
+
+{\bf This section has not been updated since 2018.}
 
@@
 The main resources from which we have collected information about the Community's Special Program are: \citep{2008arXiv0805.2366I}; \citep{LPM-17}; the LSST Deep Drilling Field white papers from 2011\footnote{\url{https://project.lsst.org/content/whitepapers32012}}; presentations by Niel Brandt and Stephen Ridgway at the LSST Project and Community Workshop in August 2016\footnote{\url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} and \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}}; \citep{2013arXiv1304.3455G}; and Chapter 10 of \citep{2017arXiv170804058L}.
-
-%\begin{itemize}
-%\item "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \citep{2008arXiv0805.2366I} 
-%\item The LSST Science Requirements Document (\SRD), \citeds{LPM-17} 
-%\item LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} 
-%\item "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf}
-%\item "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}
-%\item "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \citep{2013arXiv1304.3455G} 
-%\item Chapter 10 "Special Surveys" of the Observing Strategy White Paper \citep{2017arXiv170804058L}
-%\end{itemize}
 
@@
 
-%\input{docrev}
-
 \end{document}

### 2022-02-20 DMTN-065.tex
doc rev req

@@
 The core LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, which is expected to be accomplished with 85--90\% of the observing time.
-The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
+The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which build on, or are beyond, the core science pillars of the LSST.
 
-This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs (including the flow-down of related requirements, and the capabilities of the LSST Science Pipelines and the Rubin Science Platform).
+\medskip
+This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs.
+The latter includes a discussion about the relevant system requirements, the needed capabilities of the LSST Science Pipelines and the Rubin Science Platform, and the cases in which Special Programs data products would be user-generated.
 
+\medskip
 The specialized, science-specific aspects of Special Programs processing and analysis that are best left to the science community are also described, and illustrated with case-study examples.
 
+\medskip
 The main target audience of this document is Rubin Observatory staff -- the construction-era Data Management (DM) team and the operations-era Data Production (RDP) team -- but members of the science community who are planning to use Special Programs to reach their science goals may also find this document useful.
@@
 Note that there remains some grey area between what is a mini-survey within the WFD, and what is a WFD-like micro-survey.
-For example, the last two on the list above (the northern stripe and the Virgo cluster) could also (or might be in the future) considered WFD mini-surveys and not Special Programs.
+For example, the last two on the list above (the northern stripe and the Virgo cluster) could also be (or might be in the future) called WFD mini-surveys, and not Special Programs micro-surveys.
 
@@
 
-\subsection{Requirements Related to Special Programs}
+\subsection{Rubin Requirements Related to Special Programs}
+% \lsrreq, \ossreq, \dmreq
+% \reqparam
 
-{\it Use the docrev.tex file to update this.}
+A detailed list of the requirements related to Special Programs is provided in Appendix~\ref{sec:docrev}.
+The most important requirements are summarized below.
 
-\citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
+\subsubsection{Data Products}
 
-\subsection{Special Programs in the DPDD}
+Rubin Observatory and the LSST system (the observatory and the data management systems) are required to process Special Programs data to produce unique and separate data products ``whenever possible" (LSR-REQ-0121).
 
-{\it Just quote what the DPDD says. Maybe reference any relevant RFC/LCR that made those changes?}
+The term ``whenever possible" includes cases where the original or reconfigured versions of the LSST Science Pipelines can be run, and excludes cases where the development of new algorithms or the allocation of significant additional computational resources are required (LSR-REQ-0121).
+
+The statement ``to produce unique and separate data products" typically refers to producing the same kinds of data products as will be generated by the Prompt and Data Release pipelines (processed visit images, coadded images, difference images, and catalogs of sources and objects for those images).
+
+It is a requirement that the cumulative size of the Special Programs data products generated by Rubin Observatory be no more than $\sim$10\% the size of the Data Release data products (i.e., proportional to the fraction of survey time spent; LSR-REQ-0121).
+
+It is a requirement that these Special Programs data products be distinct, and joinable with (in other words, they can be federated or cross-matched with) the Prompt and Data Release data products (DMS-REQ-0322).
+
+The derivation of value-added data products, such as HiPS or MOC maps, for Special Programs remains an open question (DMS-REQ-0379, 0383).
+
+\subsubsection{Metadata}
+
+In order to support Special Programs processing, the LSST system is required to store metadata that includes program information for every raw image, such as identifiers for images obtained as part of the Main Survey or a Special Program (DMS-REQ-0068).
+
+It is required that this metadata be sufficient for Special Programs to trigger their own real-time data processing recipes ``whenever possible" (DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).
+
+\subsubsection{Processing}
+
+It is a requirement that Special Programs processing with the Prompt pipeline (or a reconfigured version of it) is subject to the same timescales and latency constraints of 24 hours for the release of Prompt data products and 1 minute for the transmission of Alert packets (DMS-REQ-0344).
+
+It is also a requirement that Special Programs processing be done on timescales intermediate to the Prompt and Data Release processing, ``whenever possible" and whenever necessary to enable the intended science goals of the Special Program (LSR-REQ-0032).
+
+It is a requirement that the LSST system be able to process non-standard visits with short exposure times as low as 1 second, with a discussion note that such short exposures might have degraded image quality (LSR-REQ-0111).
+
+Processing for Special Programs by Rubin Observatory is expected to use no more than $\sim$10\% of computational and storage capacity of the Rubin data processing cluster (i.e., proportional to the fraction of survey time spent and the size of the Rubin-processed Special Programs data products; Section 6 of the DPDD).
+
+\subsubsection{User Processing}
+
+In cases where the science goals of a Special Program require that new algorithms or software be developed, User-Generated pipelines and data products will be needed.
+
+The 10\% of the total data processing capacity that Rubin Observatory is required to reserve for all User-Generated processing includes that applied by users to Special Programs data -- in other words, there is no additional capacity {\it for users} that will be reserved only for Special Programs data (LSR-REQ-0041).
 
@@
 {\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
-As discussed under "Solar System Objects (SSO)" in Appendix \ref{ssec:proc_datadiv_prev}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
+As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:prevpropsp}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
 %There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
@@
 \item discoverability of SP data when browsing an all-sky map
+\item to query data by tag of WFD, WFD mini-survey, or Special Program (e.g., DDF field, micro-survey identifier)
 \end{itemize}
@@
 
+% % % % % % % % % % % % % % % % % %
 \clearpage
-% % % % % % % % % % % % % % % % % %
 \bibliography{local,lsst,refs,books,refs_ads}
 
+% % % % % % % % % % % % % % % % % %
 \clearpage
-% % % % % % % % % % % % % % % % % %
 \appendix
 
-% % % % % % % % % % % % % % % % % %
-\section{Potential Hardware Boundaries on Data Diversity}\label{sec:hardbounds}
+\input{ap_docrev.tex}
 
-The potential boundaries on the diversity of data products that could be imposed by limitations from the Rubin Observatory hardware -- camera, telescope, and/or site -- are considered.
+\input{ap_hardbounds.tex}
 
-\subsection{Filter Changes}
-The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
-Assuming that most Special Programs would be designed to keep overheads $<$100\% and would be using standard 30 second visits, the filter change time indicates that it is likely that at least 4 exposures in a given filter would be obtained between filter changes, but this is not actually a hardware boundary.
-The filter change mechanism is designed to undergo a total of 100000 changes over its lifetime, and each filter is designed to support up to 30000 changes over its lifetime, where lifetime is 15 years.
-That is an average of $\sim$27 changes per day, some of which would occur in they day during calibrations (estimate, $\sim$10) and the rest at night.
-As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could support as many changes involving the 5 filters loaded in the carousel as desired, without any practical limitation"}.
+\input{ap_procbounds.tex}
 
-\subsection{Filter Carousel Loads}
-The filter carousel can hold five of the six LSST filters at a time.
-The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
-Filter loads are only done in the day, and there will never be data in more than five filters in a given night.
-
-\subsection{Exposure Times}
-The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
-The maximum exposure time is not restricted.
-
-\subsection{Readout Time}
-The readout time is $2$ seconds, and would be significant overhead on short exposures.
-
-\subsection{Inter-Image Time}
-Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
-The $15$ interval between images is a potential hardware boundary on the potential diversity of data products.
-
-\subsection{Telescope Slew}
-As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no hardware boundaries on the size of a single slew or the accrued slew distance.
-
-\subsection{Telescope Tracking}
-The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
-This capability will include angular rates of up to 220 arcseconds per second in both azimuth and elevation.
-
-\subsection{Camera Rotation}
-The requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
-Currently, there are no hardware boundaries imposed by camera rotation constraints on the potential diversity of data products.
-
-% % % % % % % % % % % % % % % % % %
-\section{Potential Processing Boundaries on Data Diversity}\label{sec:procbounds}
-
-The capability of the LSST Science Pipelines to process diverse data is explored below.
-
-Note that processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters (e.g., the number of stars with sufficient flux for photometric calibration).
-Furthermore, the processing boundaries might not be fully constrained until the final performance of the LSST Science Pipelines, as described in the Data Management Applications Design, \citeds{LDM-151}) document, is fully characterized.
-
-{\bf Summary of the conclusions below:}\\
-Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
-It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
-The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.
-
-\subsection{Exposure Times}\label{ssec:procbounds_expt}
-
-Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
-
-\subsubsection{Short Exposures (Non-Standard Visits of $<$30 sec)}
-The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
-
-However, for exposure times there are other considerations, as changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r$$\approx$ 13--21 magnitudes.
-A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but this still leaves stars between 15.8--21.0 magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
-
-In order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
-As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
-
-\begin{figure}
-\begin{center}
-\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
-\end{center}
-\end{figure}
-
-% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute Alert Production.
-% One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-% They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-% They also present a new software package to do mitigate the effects.
-% Software packages \texttt{PhoSim} (Peterson et al.
-2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} could be used to characterize the PSF stability as a function of exposure time.
-
-\subsubsection{Long Exposures (Non-Standard Visits of $>$30 sec)}
-
-There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures (unknown), which could impact the quality of a difference image and the detected sources.
-Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
-
-\subsection{Twilight Images with a Bright Background}
-
-Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also apply here.
-Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a user generated pipeline, remains to be determined (see also the example in Section \ref{ssec:SPCS_Twilight}).
-This may depend on the exposure time and the number of stars available in the image.
-
-\subsection{Images Obtained with Non-Sidereal Tracking}
-
-Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
-Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a user generated pipeline.
-The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.
-
-%%% MLG removed the following in Feb 2022, it's no longer a concern.
-% \subsection{Number of Exposures per Visit (Long Sequences of a Single Field)}
-
-% There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
-% From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.
-
-% K.-T.
-Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
-This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
-Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
-things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
-The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
-To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
-Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
-\textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
-
-% K.-T.
-has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
-One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
-The processing pipeline would need to know how to divide this sequence up into visits.
-As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
-
-%%% MLG removed the following in Feb 2022, it's no longer a concern for Special Programs (crowded fields are in the WFD).
-% \subsection{Images in Very Crowded Fields}
-
-% The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
-% The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.
-
-\clearpage
-% % % % % % % % % % % % % % % % % %
-\section{Previously Proposed Special Programs}\label{ssec:proc_datadiv_prev}
-
-{\bf This section has not been updated since 2018.}
-
-In this section we compile information about the science goals and observational methods for Special Programs that have been previously proposed or discussed in the Science Community.
-We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science.
-The main resources from which we have collected information about the Community's Special Program are: \citep{2008arXiv0805.2366I}; \citep{LPM-17}; the LSST Deep Drilling Field white papers from 2011\footnote{\url{https://project.lsst.org/content/whitepapers32012}}; presentations by Niel Brandt and Stephen Ridgway at the LSST Project and Community Workshop in August 2016\footnote{\url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} and \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}}; \citep{2013arXiv1304.3455G}; and Chapter 10 of \citep{2017arXiv170804058L}.
-
-So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}.
-There are three mini-survey areas that have been discussed extensively by the Science Community: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}).
-In Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified, along with an \textit{incomplete} list of potential mini-surveys that have been openly discussed in the Science Community.
-In Section \ref{sec:SPCS}, we create detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.
-
-\begin{table}[h]
-\begin{center}
-\begin{footnotesize}
-\caption{Approved DDF and Incomplete List of Potential Special Programs.}
-\label{tab:ddfms}
-\begin{tabular}{lll}
-\hline \hline
-Name & Coordinates & Description  \\
-\hline
-DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
-DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
-DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
-DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
-\hline
-North Ecliptic Spur      & & solar system objects (find and characterize) \\
-Galactic Plane             & & more intensive stellar surveying \\
-South Equatorial Cap  & & S/LMC and more Galactic science \\
-Twilight                        & & short exposures (0.1s) for bright stars \\
-Mini-Moons                     &  & finding mini-moons \\
-Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
-Meter-Sized Impactors     & & detection a week before impact \\
-GW Optical Counterparts & & search and recovery \\
-Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
-\hline
-\end{tabular}
-\end{footnotesize}
-\end{center}
-\end{table}
-
-Here we consider a variety of scientific fields in turn, the Special Programs that have been discussed in that Science Community so far, and the implications of these Programs for the diversity of data and data products.
-Generally, the types of LSST Special Programs that are open for proposals include: (i) additional deep drilling fields; (ii) refined observing strategies for deep drilling fields; (iii) optimized survey areas for the NES, South Pole, and Galactic Plane; (iv) refined observing strategies for the NES, South Pole, and Galactic Plane; and (v) additional mini-surveys (areas and observing strategies).
-
-\medskip
-\noindent \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al.
-(2008, \citep{2008arXiv0805.2366I}; Section 3.1.2) describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters, repeated every two nights for four months.
-Each exposure would have a $5\sigma$ limit of $r\sim24$; the nightly stack would have a limit of $r\sim26.5$; and the final deep stack of all exposures would have a limit of $r\sim28$.
-This description does not comment on the processing mode, but, depending on the science goals the exposures could be done as either a series of 50 non-standard visits ($1\times15$ seconds) or 25 standard visits ($2\times15$ seconds).
-
-\medskip
-\noindent \textbf{Solar System Objects (SSO) -- } Four of the mini-surveys in Table \ref{tab:ddfms} have science goals related to studies of SSO.
-Observations of the North Ecliptic Spur area could yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (reference: Brandt's talk).
-The Mini-Moons Mini-Survey aims to find and study the temporarily captured satellites of the Earth (Section 10.2, \citep{2017arXiv170804058L}).
-The Sweetspot Survey would use twilight fields to find NEOs in Earth-like orbits (i.e., these objects are never in opposition fields, but overhead at sunrise/sunset; Section 10.2, \citep{2017arXiv170804058L}).
-The Meter-Sized Impactors program would find and track meter-sized impactors $<2$ weeks before impact (Section 10.2, \citep{2017arXiv170804058L}).
-{\bf Summary:} most of these science goals do not seem to require non-standard visits or exposure times, with the exception of the Sweetspot survey which occurs during twilight and thus may require shorter exposures.
-The cadence and patterns of these mini-surveys may differ from the WFD main survey, especially when very fast-moving objects are sought.
-From a processing perspective, it seems that many of these science goals will be achievable by using the products of Solar System Processing, which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night.
-The exception is finding faint SSOs (e.g., Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) through shift-and-stack (SAS) processing \citedsp{Document-11013}, because SAS is not a capability being built within the DM system and cannot be done solely by reconfiguring DM pipelines.
-An example of user-generated pipeline for SAS is described in Section \ref{sec:SPCS}.
-
-\medskip
-\noindent \textbf{Stars in the Milky Way and Magellanic Clouds -- } As described in \citedsp{Publication-141}, mini-surveys of the Galactic Plane can better distinguish faints stars from faint red galaxies by including at least 3 filters of coverage (e.g., $izy$; similar to WFD), and could mitigate losses from proper motion and increase the detection rate of stellar flares by obtaining all the images in short time span (i.e., a more concentrated cadence than the WFD).
-As described in \citedsp{Publication-145}, applying the nominal DDF observing strategy over the full area of the Large and Small Magellanic Clouds can characterize stellar variability to $M_V<6.5$ on timescales from 15 seconds to 3 days.
-For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"} \citedsp{Publication-145}.
-The Twilight survey in Table \ref{tab:ddfms} proposes short exposures to enable bright stars to be put on the same photometric system as the deeper LSST WFD main survey catalog, and enable science that is based on their long monitoring baselines from historical observations.
-In Chapter 10.4 of \citep{2017arXiv170804058L}, a proposed short-exposure survey of M67 would use the camera's stretch goal of $0.1$ second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as $6$ magnitudes brighter than the saturation level"}.
-{\bf Summary:} while some of these science goals can be accomplished with standard visits, MW \& L/SMC science goals are likely to request shorter exposure times, perhaps down to $0.1$ seconds.
-These science goals are also likely to propose cadence and filter distributions that are significantly different from the WFD main survey.
-From a processing perspective, the science goals depending on shorter exposures will only be able to be met by reconfiguring the DM pipelines if the short exposures can be shown to successfully be processed (with, e.g., instrument signature removal); the science goals can likely be met with data products in the same format as the Prompt or DR Pipeline (i.e., {\tt Source} and {\tt Object} catalogs, single visits and deep CoAdds).
-Although it is not mentioned in the above paragraph, the MW \& L/SMC science community is also most likely to require special processing to extract information from saturated stars, which is outside the scope of DM.
-See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.
-
-\noindent \textbf{Exoplanets -- } As described in Section 3.1.2 of \citep{2008arXiv0805.2366I}, transiting exoplanets could be detected with the nominal DDF plan, which would allow for $1\%$ variability to be detected over hour-long timescales; a DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence.
-\citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane, and how microlensing candidates can be found with $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (since microlensing events are slower; these would then require follow-up with external facilities).
-Dealing with the more crowded fields would be mitigated by the shallower images, in this case.
-One of the main points of \citep{2013arXiv1304.3455G} is that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped.
-\textbf{Summary.} Some of these science goals appear possible with standard visit images, and some might request shorter exposures to avoid confusion in crowded fields when the science can be done with brighter stars.
-From a processing perspective, the science goals are likely to be achievable with reconfigured DM pipelines, but this depends heavily on performance in crowded fields.
-See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.
-
-\noindent \textbf{Supernovae -- } The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses.
-The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$ \citedsp{Publication-144}.
-High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}.
-Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template.
-This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days.
-These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss.
-SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science.
-The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citedsp{Publication-142}.
-Short-exposure observations of bright, nearby SNe may also be useful to include near-peak photometry in the LSST magnitude system, and enable full light-curve analyses.
-\textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of a target-of-opportunity short-exposure program to observe bright SNe).
-From a processing perspecitve, the science goals appear to be accessible with reconfigured DM pipelines to stack and difference the data.
-In particular, the DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images.
-See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.
-
-\noindent \textbf{Galaxies -- } The additional depth of a DDF may provide access to a larger collection of low-$\mu$ objects.
-\citedsp{Publication-142} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies".
-The DDF stacks could also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects.
-Also, the DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}.
-\textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources.
-In addition, it seems likely that user-generated algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, and these are beyond the scope of DM.
-
-\noindent \textbf{Weak Lensing -- } The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch).
-\textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
-%$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
-%$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline.
-We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
-%$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
+\input{ap_prevpropsp.tex}
 

### 2022-02-20 ap_docrev.tex
doc rev req

\section{Rubin Documentation Review}\label{sec:docrev}

Below are described mentions of Special Programs in Rubin documentations, with a focus on requirements and specifications.

Updates to Rubin documents related to Special Programs -- motivated by past versions of this DMTN -- were made via LCR-1309 and LCR-2265.

% LCR 1309: https://project.lsst.org/groups/ccb/node/2383
% LCR 2265: https://project.lsst.org/groups/ccb/node/4036

\subsection{Science Requirements Document (SRD)}

Version 5.2.4 (revision 2018-01-30), \citedsp{LPM-17}.

Section 3.4 ``The Full Survey Specifications" states the SRD's assumption that 90\% of the total available survey time would be spent on the main survey, and that the remaining 10\% would be spent {\it ``to obtain improved coverage of parameter space ... [or to] observe special regions"}.

\subsection{LSST System Requirements (LSR)}

Version 7.1 (revision 2020-03-05), \citeds{LSE-29}.

Note that Version 5 (2018-06-26) was an update for LCR-1309, which added requirements, specifications, and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.

In Section 1.5.1.3, ``Processing Data from Special Programs", LSR-REQ-0122\lsrreq{0122} is a requirement that the LSST system {\it ``shall deliver unique and separate data products for visits from Special Programs"} whenever possible, and that they {\it ``shall be delivered on timescales intermediate"} to the Prompt and Data Release timescales {\it ``when this enables the intended science of the Special Program"}.
The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required"}.

In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\req{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second (\reqparam{minExpTime}).
The discussion notes that {\it ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g. cosmic ray rejection on visits consisting of a single exposure), and might be incompatible with difference imaging and alert production (e.g., short exposures in which the PSF is
not fully formed)"}.

The above requirement indicates that the LSST system shall be able to process non-standard visits from Special Programs, but that the image quality might be degraded.
Improvements that require algorithms or processing outside of what the LSST Science Pipelines can provide would be left to the science community and require User-Generated pipelines and data products.

The requirement in Section 1.5.1.3 is echoed in Section 2.6.1.1, ``Organization of Data Products", in which LSR-REQ-0032\lsrreq{0032} is a requirement that the data processing system provide the means for three 'classes' of data products on different timescales (Prompt, Data Release, and User-Generated), and also to provide a means for processing Special Programs data because the {\it ``science goals of Special Programs may require that their processed data products be made available in an additional fourth class, and possibly with intermediate timescales"}.

In Section 2.6.1.1.3, ``Level 3 Data Products", LSR-REQ-0041\lsrreq{0041} specifies that the LSST system {\it ``shall support"} User-Generated data products.
The discussion clarifies that ``{\it there will be technical limits on DM's ability to meet this requirement, such as cases where an intensive amount of additional computational resources is required, because only ~10\% of the total computational system is allocated for user processing"}.
This level of support applies also to user processing of Special Programs data.
See also the reference to LSR-REQ-0055 below.

Section 2.6.1.1.4, ``Data Products for Special Programs", LSR-REQ-0121\lsrreq{0121} specifies that the LSST system {\it ``shall produce unique and separate Data Products as the result of processing data from Special Programs whenever possible, on a timescale that enables the intended science goals of the Special Program.
The cumulative size of the online Special Programs data products shall be no more than ~10\% of the size of the DRP data products from the most recent data
release"}.
The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required.
The cumulative size of the Special Programs data products is capped at ~10\% of the most recent DR because this matches the expected fractional survey area of Special Programs compared to the main survey"}.

In Section 2.7.1.6, ``Community Computing Services", LSR-REQ-0055\lsrreq{0055} requires that the LSST system {\it ``shall provide and maintain an amount of computing capacity equivalent to at least"} 10\% (\reqparam{userComputingFraction}) {\it ``of the total LSST data processing capacity (computing and storage) for the purpose of scientific analysis of LSST data and the production of"} User-Generated data products. 
The discussion clarifies that the scope of this service remains to be determined.
This level of computational resources includes user processing of Special Programs data.

In Section 3.1.3.1, ``Survey Time Allocation", LSR-REQ-0075\lsrreq{0075} requires that the {\it ``survey performance requirements shall be met utilizing approximately 90\% of the historically available observing time, leaving the remaining time available for yet to be defined special programs"}.

In Section 1.3.1.3, LSR-REQ-0124\lsrreq{0124}, the discussion specifies that an image quality parameter related to ellipticity applies only to main survey data.

\subsection{Observatory System Specifications (OSS)}

Version 19.1 (revision 2021-07-30), \citeds{LSE-30}.

Note that Version 13 (2018-06-26) was an update for LCR-1309, which added requirements, specifications, and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.

In Section 2.2.3.1, ``Standard Operating States", OSS-REQ-0044\ossreq{0044} specifies that {\it ``the LSST observatory system shall be designed and constructed to support ... manual observing - used for specific non-scheduler driven observing to support system verification and testing or specialized science programs"}.
Although most Special Programs will be executed via the survey scheduler as part of {\it ``fully automated observing"}, manual observing might be necessary for, e.g., target-of-opportunity Special Programs.

Section 3.1.5.1.2, ``Data Products Handling for Special Programs", OSS-REQ-0392\ossreq{0392} is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121), and specifies that {\it ``the handling of data products from Special Programs shall be compliant with the approach defined in LSE-163"}.

In Section 3.2.5.3, OSS-REQ-0403\ossreq{0403} is a flow-down of LSR-REQ-0124\lsrreq{0124} related to the ellipticity correlation function distribution.

In Section 3.6.1.3, ``Continuous Exposures", OSS-REQ-0319\ossreq{0319} requires that {\it ``The Observatory shall be capable of continuous operation throughout a night with the interval between successive visits equal to the FPA readout time"}.
The discussion clarifies that {\it ``this mode of observing is needed to support observations when the telescope is not being re-pointed"}, such as deep drilling fields or other Special Programs.

In Section 3.6.1.4, ``Minimum Exposure Time", OSS-REQ-0291\ossreq{0291} specifies that {\it ``the camera shall be able to obtain a single exposure with an effective minimum exposure time of no more than"} 1 second (\reqparam{minExpTime}) {\it ``with a goal of an effective minimum exposure time of"} 0.1 seconds (\reqparam{minExpTimeGoal}). 
The discussion clarifies that {\it ``if the exposure is shortened from the 15 second nominal, the spacing between successive exposures should be extended to maintain the average readout rate consistent with a 15 second exposure"}, which may increase the overheads of Special Programs using short exposure times.
The discussion also clarifies that {\it ``if the exposure is lengthened from the 15 second nominal, the thermal stability may be affected, which may affect photometric accuracy"}.

In Section 3.6.1.5, ``Publish Visit Type", OSS-REQ-0384\ossreq{0384} specifies that {\it ``the OCS [Observatory Control System] shall configure the [Data Management System] DMS (in particular Prompt Processing) with the type of visits to be processed: Standard, Alternate, or a specific type of Non-Standard"}.
The discussion clarifies that this allows the Prompt processing pipeline to be reconfigured on-the-fly in order to incorporate non-standard visits from, e.g., Special Programs.
The time required for reconfiguration might introduce some latency or cause some images to not be processed by the Prompt pipeline.

In Section 3.6.2.1.2, ``Maximum time for operational filter change", OSS-REQ-0293\ossreq{0293} specifies that {\it ``the camera system shall provide the capability of changing the operational filter with any other internal filter in a time less than"} 120 seconds (\reqparam{tFilterChange}).
This would impose a large overhead on, e.g., a Special Program that changes filters often without slewing.
See also OSS-REQ-0295\ossreq{0295}, Appendix~\ref{sec:hardbounds} of this document, and/or the filter change memorandum (\url{ls.st/spt-494}), for more information about the total lifetime number of filter changes.

In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour (\reqparam{rotTrackTime}).
The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field}.
There do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits.

In Section 3.6.3.10, ``Non-Sidereal Tracking", OSS-REQ-0380\ossreq{0380} specifies that {\it ``the LSST system shall be capable of tracking in an arbitrary direction on the sky along a parametric RA(t) and DEC(t) trajectory, at angular rates of up to"} 220 arcseconds per second (\reqparam{nonsiderealAngularRateEl} and \reqparam{nonsiderealAngularRateAZ}) {\it ``with a tracking error not to exceed"} 0.5 arcseconds per minute (\reqparam{nonsiderealTrackingError}).
The discussion notes that {\it ``this is standard capability for modern telescopes"}, but might be relevant to some Special Programs.

%%% MLG: this does not appear in OSS V13
% $\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs). {\bf JIRA ticket DM-12579 confirmed that there is no maximum number, and so many Special Programs will be able to be included in the scheduler}.

%%% MLG: this does not appear in OSS V13
% $\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (OSS 2.1.1.7, Visit Optimization).

%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
% $\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively. These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin. These estimates appear adequate for Special Programs in general.

%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
% $\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively. These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.

\subsection{Data Management Subsystems Requirements (DMSR)}

Version 9 (revision 2021-02-12), \citeds{LSE-61}.

Note that Version 8.3 (2020-05-04) was an update for LCR-2265, which updated requirements, specifications, and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.

In Section 1.2.3, ``Raw Science Image Metadata", DMS-REQ-0068\dmreq{0068} specifies that {\it ``for each raw science image, the DMS shall store image metadata"} including {\it ``Program metadata (identifier for main survey, deep drilling, etc.)"}.
The discussion clarifies that {\it ``the program metadata should be sufficient to associate an image with a specific Special Program so that DMS-REQ-0320 and DMS-REQ-0397 can be satisfied"}.

In Section 1.3.13, ``Alert Content", the discussion for DMS-REQ-0274\dmreq{0274} explains that the {\it ``program and/or scheduler metadata"} included in an alert packet {\it ``should be sufficient to identify whether the image is associated with a Special Program (such as an in-progress Deep Drilling Field)"}.

In Section 1.4.18.1, ``Produce All-Sky HiPS Map", the discussion for DMS-REQ-0379\dmreq{0379} raises the point that generating separate HiPS maps for Special Programs (e.g., DDFs) remains an open question.

In Section 1.4.18.5, ``Produce MOC Maps", DMS-REQ-0383\dmreq{0383} specifies that Data Release processing {\it ``shall include the production of Multi-Order Coverage maps for the survey data"}, and that {\it ``additional MOCs SHOULD be produced to represent special-programs datasets"}.
It is noted that a separate technical note would be created to define these MOCs.

The bulk of the DMS's requirements related to Special Programs are in Section 1.6 of the DMSR.

In Section 1.6.1, ``Processing of Data From Special Programs", DMS-REQ-0320\dmreq{0320} specifies that {\it ``it shall be possible for special programs to trigger their own data processing recipes, during the night instead of the nightly Alert Processing (but the recipes may still issue Alerts), or on alternative timescales"}.
The discussion clarifies that the {\it ``LSST will provide these recipes ... when
possible, which includes cases where DM can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required. An example of an alternative timescale is a nightly trigger to coadd all the deep-drilling field images. Decisions about which recipes are applied to which Special Programs will be made by the Operations team, after consideration of the scientific goals, computational resources, and data rights policy"}.
This requirement is derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).

In Section 1.6.2, ``Prompt/DR Processing of Data from Special Programs", DMS-REQ-0397\dmreq{0397} specifies that {\it ``it shall be possible for special programs data to be processed with the prompt and/or annual-release pipelines alongside data from the main survey"}.
The discussion further clarifies that {\it ``the data from Special Programs should only be included ... when it is (a) possible ... to do so without additional effort and (b) beneficial to the LSST's main science objectives. Decisions about which data are included ... will be made by the Operations team"}.
This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).

In Section 1.6.3, ``Level 1 Processing of Special Programs Data", DMS-REQ-0321\dmreq{0321} specifies that {\it ``all [Prompt] processing from special programs shall be completed before data arrives from the following night's observations"}.
This is essentially adding a quantifier to DMS-REQ-0397, to specify that {\it ``when it is (a) possible ... to do so"} means when it is possible to complete the processing before the next night's observations.
This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).

In Section 1.6.4, ``Constraints on Level 1 Special Program Products Generation", DMS-REQ-0344\dmreq{0344} specifies that {\it ``the publishing of [Prompt] data products from Special Programs shall be subject to the same performance requirements of"} 24 hours (\reqparam{L1PublicT}) for the release of Prompt data products and 1 minute (\reqparam{OTT1}) for the transmission of Alert packets.
This is essentially a more detailed version of DMS-REQ-0321 which includes the Alert production timescale.
This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirents from the LSR (0122, 0075, and 0121).

In Section 1.6.5, ``Special Programs Database", DMS-REQ-0322\dmreq{0322} specifies that {\it ``data products for special programs shall be stored in databases that are distinct from those used to store standard [Prompt] and [Data Release] data products"} and that {\it ``it shall be possible for these databases to be federated ... to allow cross-queries and joins"}.
This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).

In Section 4.1.16, ``Level 2 and Reprocessed Level 1 Catalog Access", DMS-REQ-0313\dmreq{0313} specifies that {\it ``the DMS shall maintain ... versions
of the most recent catalogs generated from Special Programs data"}.
As with all LSST data, {\it ``there is no requirement for older data releases to be queryable"}.

\subsection{Data Management Applications Design (DMAD)}

Version 4.3 (revision 2020-11-10), \citeds{LDM-151}.

The DMAD is not a requirements document.
Instead, it describes the scientific design of the LSST Science Pipelines: the algorithms and software that will be implemented to meet the requirements for processing the LSST data.

Special Programs are only mentioned a few times, either as a potential source of single-snap visits or as a potential source of reference images or catalogs (e.g., training sets).

As described above (e.g., LSR-REQ-0122), the LSST system shall deliver unique and separate data products for visits from Special Programs whenever this (1) enables the intended science of the Special Program and (2) can be accomplished using the original or reconfigured versions of the LSST Science Pipelines.
For cases in which the development of new algorithms or the allocation of significant additional computational resources are required to produce Special Programs data products, User-Generated pipelines and processing will be necessary.

The DMAD can be used as the reference document to decide whether a given Special Program will require User-Generated pipelines and processing.

%%% MLG: the following are all from <2018 and were not updated in 2022

% \subsubsection{Extreme-Depth CoAdds} The system has been sized to hold $\sim200$ exposures in memory at once, which defined by the current maximum number of visits per field in the WFD main survey in $10$ years (from a conversation with K.-T.~Lim). Note that the panchromatic CoAdds would be built from the individual filter CoAdds, so the algorithm does not need to handle $\sim800$ images. From a computational standpoint, $200$ is the maximum number of images that can be stacked with an algorithm that requires all images to be accessible in memory at once (i.e., loading all images and calculating the median for each pixel). Deeper stacks might be possible with algorithms that deal with images consequentially. It is conceivable that a Special Program which needs to stack $>200$ images is not possible to accomplish with reconfigured pipelines, and would have to be processed with external, user-contributed resources. However, the exact DM capabilities in this area are not yet well known because NCSA has not yet defined the machine capabilities. Furthermore, the planned commissioning data will go to a $\sim20$ year depth, and so it can reasonably be expected that DM will have to be able to accommodate at least a stack that deep.

% \subsubsection{Deblending} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey. It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal. Level 3 deblenders for specific Special Programs fields may require development by the user community.

% \subsubsection{Variability Characterization} The periodic and aperiodic variability characterizations described in Section 6.21 of the DMAD are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types. From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals. It is conceivable that the goals of a particular Special Program might require different algorithms; these could be provided by DM, or written as Level 3 and either made joinable to the DM reconfigured data products or perhaps incorporated directly.

% \subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP \texttt{Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM. It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.

\subsection{Data Products Definitions Document (DPDD)}

Version 3.6 (revision 2021-12-17), \citeds{LSE-163}.

Section 6 describes the data products for Special Programs.
The DPDD is not a requirements document; Section 6 summarizes the requirements presented above and does not introduce any new constraints or new information about Special Programs.

%%% MLG commented out the paragraphs below in 2022, they were out of date.

% $\bullet$ The database schema for \texttt{DIASource} does not appear to have an element that identifies which template image that was used. This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data. DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61). In conversation with K.-T., we find that this isn't a problem, as it will be handled by provenance: the code configuration used and the time of processing are sufficient to identify and regenerate the template image. However, K.-T. has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable. However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.

% $\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs. It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this. Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType. Indicates which input was used to produce a given object."} Is that all we need? {\bf JIRA ticket DM-12580 clarified that  \texttt{Object.prv\_inputID} in the database schema is one possible way to identify whether an \texttt{Object} is an externally provided coordinate}.

% $\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
% \texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
% \texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
% Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey. Are 32 floats in each of the 6 filters always going to be a large enough volume? The Transients and Variable Stars Science Collaboration currently has a Task Force assigned to address this, and {\bf it is the topic of JIRA ticket DM-12581}.

%%% MLG commented out this whole section in 2022, since LSE-180 is from 2013
%%%  LSE-180 is deprecated?

% \subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}

% LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey. Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}. Regarding the reduction and calibration of non-standard visit images, LSE-180 makes two relevant points: \\
% $\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
% $\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD. The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.

% However, Lupton's recent work on calibrations has made much of LSE-180 obsolete, and it is not clear whether it is still needed as a separate document. At the time of writing, Lupton's most recent take on calibrations can be found in \url{https://github.com/lsst-dm/calibration/blob/master/calibration.pdf}. K.-T. referenced the Appendix B for the overview of calibration types, but it seems that there is not yet a final plan that can be assessed for its suitability for Special Programs. It is plausible that Special Programs could have their own calibration database, and obtaining additional calibration frames is not expected to cause a bottleneck (at the moment there is $\sim1.7$ hours per day for this, and exceeding this could eat into the engineering time). The most likely cause for trouble involving calibrations and Special Programs is the computational time needed to create the additional and/or different calibration files to be applied to non-standard Special Programs data, and/or the additional overhead required to load them in the Level 1 pipeline if the schedule interleaves SP-WFD-WFD-SP.

% {\bf JIRA ticket DM-12582 is currently open, and aims to define the potential additional calibration needs of Special Programs data.}

### 2022-02-20 ap_hardbounds.tex
doc rev req

\section{Potential Hardware Boundaries on Data Diversity}\label{sec:hardbounds}

The potential boundaries on the diversity of data products that could be imposed by limitations from the Rubin Observatory hardware -- camera, telescope, and/or site -- are considered.

\subsection{Filter Changes}
The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
Assuming that most Special Programs would be designed to keep overheads $<$100\% and would be using standard 30 second visits, the filter change time indicates that it is likely that at least 4 exposures in a given filter would be obtained between filter changes, but this is not actually a hardware boundary. 
The filter change mechanism is designed to undergo a total of 100000 changes over its lifetime, and each filter is designed to support up to 30000 changes over its lifetime, where lifetime is 15 years.
That is an average of $\sim$27 changes per day, some of which would occur in they day during calibrations (estimate, $\sim$10) and the rest at night.
As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could support as many changes involving the 5 filters loaded in the carousel as desired, without any practical limitation"}.

\subsection{Filter Carousel Loads}
The filter carousel can hold five of the six LSST filters at a time.
The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}).
Filter loads are only done in the day, and there will never be data in more than five filters in a given night.

\subsection{Exposure Times}
The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
The maximum exposure time is not restricted.

\subsection{Readout Time}
The readout time is $2$ seconds, and would be significant overhead on short exposures.

\subsection{Inter-Image Time}
Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
The $15$ interval between images is a potential hardware boundary on the potential diversity of data products.

\subsection{Telescope Slew}
As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no hardware boundaries on the size of a single slew or the accrued slew distance.

\subsection{Telescope Tracking}
The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
This capability will include angular rates of up to 220 arcseconds per second in both azimuth and elevation.

\subsection{Camera Rotation}
The requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
Currently, there are no hardware boundaries imposed by camera rotation constraints on the potential diversity of data products.

### 2022-02-20 ap_procbounds.tex
doc rev req

\section{Potential Processing Boundaries on Data Diversity}\label{sec:procbounds}

The capability of the LSST Science Pipelines to process diverse data is explored below.

Note that processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters (e.g., the number of stars with sufficient flux for photometric calibration).
Furthermore, the processing boundaries might not be fully constrained until the final performance of the LSST Science Pipelines, as described in the Data Management Applications Design, \citeds{LDM-151}) document, is fully characterized.

{\bf Summary of the conclusions below:}\\
Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.

\subsection{Exposure Times}\label{ssec:procbounds_expt}

Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.

\subsubsection{Short Exposures (Non-Standard Visits of $<$30 sec)}
The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).

However, for exposure times there are other considerations, as changing the exposure time also affects the photometric and astrometric calibrations.
Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r$$\approx$ 13--21 magnitudes.
A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but this still leaves stars between 15.8--21.0 magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).

In order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").

\begin{figure}
\begin{center}
\includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin. At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center. We can see that for exposure times greater than 2 seconds, the centroid converges near its true value. \label{fig:expt}}
\end{center}
\end{figure}

% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute Alert Production.
% One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
% They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
% They also present a new software package to do mitigate the effects.
% Software packages \texttt{PhoSim} (Peterson et al. 2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} could be used to characterize the PSF stability as a function of exposure time.

\subsubsection{Long Exposures (Non-Standard Visits of $>$30 sec)}

There is no maximum exposure time specified for an LSST image.
Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures (unknown), which could impact the quality of a difference image and the detected sources.
Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).

\subsection{Twilight Images with a Bright Background}

Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also apply here.
Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a user generated pipeline, remains to be determined (see also the example in Section \ref{ssec:SPCS_Twilight}).
This may depend on the exposure time and the number of stars available in the image.

\subsection{Images Obtained with Non-Sidereal Tracking}

Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a user generated pipeline.
The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.

%%% MLG removed the following in Feb 2022, it's no longer a concern.
% \subsection{Number of Exposures per Visit (Long Sequences of a Single Field)}

% There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
% From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.

% K.-T. Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code. This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".  Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T. things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs. The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2. To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}). Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit. \textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}

% K.-T. has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler. One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence. The processing pipeline would need to know how to divide this sequence up into visits. As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.

%%% MLG removed the following in Feb 2022, it's no longer a concern for Special Programs (crowded fields are in the WFD).
% \subsection{Images in Very Crowded Fields}

% The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
% The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.

### 2022-02-20 ap_prevpropsp.tex
renamed ap prev proc sp

\section{Previously Proposed Special Programs}\label{sec:prevpropsp}

{\bf This section has not been updated since 2018.}

In this section we compile information about the science goals and observational methods for Special Programs that have been previously proposed or discussed in the Science Community. We use these to infer the potential deviations from standard visit images, and to get a basic idea of the DM processing needs that would be required to enable the science. The main resources from which we have collected information about the Community's Special Program are: \citep{2008arXiv0805.2366I}; \citep{LPM-17}; the LSST Deep Drilling Field white papers from 2011\footnote{\url{https://project.lsst.org/content/whitepapers32012}}; presentations by Niel Brandt and Stephen Ridgway at the LSST Project and Community Workshop in August 2016\footnote{\url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} and \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf}}; \citep{2013arXiv1304.3455G}; and Chapter 10 of \citep{2017arXiv170804058L}.

So far, only one aspect of the LSST Special Programs are set: the locations of the four chosen deep drilling fields\footnote{\url{https://www.lsst.org/scientists/survey-design/ddf}}. There are three mini-survey areas that have been discussed extensively by the Science Community: the North Ecliptic Spur (NES), the South Celestial Pole, and the Galactic Plane (see Figure 8 of \citep{2008arXiv0805.2366I}). In Table \ref{tab:ddfms} we list the four extragalactic deep drilling fields have already been specified, along with an \textit{incomplete} list of potential mini-surveys that have been openly discussed in the Science Community. In Section \ref{sec:SPCS}, we create detailed DM Processing Case Studies for several of these Special Programs in order to identify any potential issues with reconfiguring the DM pipelines to create specific data products for these programs.

\begin{table}[h]
\begin{center}
\begin{footnotesize}
\caption{Approved DDF and Incomplete List of Potential Special Programs.}
\label{tab:ddfms}
\begin{tabular}{lll}
\hline \hline
Name & Coordinates & Description  \\
\hline
DDF Elias S1    & 00:37:48, -44:00:00  & approved, cadence TBD \\
DDF XMM-LSS & 02:22:50, -04:45:00  & approved, cadence TBD  \\
DDF Extended Chandra Deep Field-South & 03:32:30, -28:06:00  & approved, cadence TBD  \\
DDF COSMOS  & 10:00:24, +02:10:55 & approved, cadence TBD  \\
\hline
North Ecliptic Spur      & & solar system objects (find and characterize) \\
Galactic Plane             & & more intensive stellar surveying \\
South Equatorial Cap  & & S/LMC and more Galactic science \\
Twilight                        & & short exposures (0.1s) for bright stars \\
Mini-Moons                     &  & finding mini-moons \\
Sweetspot                       & & 60 deg from Sun for NEOs on Earth-like orbits \\
Meter-Sized Impactors     & & detection a week before impact \\
GW Optical Counterparts & & search and recovery \\
Old Open Cluster M67      & dec +12 & compact survey above Galactic plane  \\
\hline
\end{tabular}
\end{footnotesize}
\end{center}
\end{table}

Here we consider a variety of scientific fields in turn, the Special Programs that have been discussed in that Science Community so far, and the implications of these Programs for the diversity of data and data products. Generally, the types of LSST Special Programs that are open for proposals include: (i) additional deep drilling fields; (ii) refined observing strategies for deep drilling fields; (iii) optimized survey areas for the NES, South Pole, and Galactic Plane; (iv) refined observing strategies for the NES, South Pole, and Galactic Plane; and (v) additional mini-surveys (areas and observing strategies).

\medskip
\noindent \textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al. (2008, \citep{2008arXiv0805.2366I}; Section 3.1.2) describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters, repeated every two nights for four months. Each exposure would have a $5\sigma$ limit of $r\sim24$; the nightly stack would have a limit of $r\sim26.5$; and the final deep stack of all exposures would have a limit of $r\sim28$. This description does not comment on the processing mode, but, depending on the science goals the exposures could be done as either a series of 50 non-standard visits ($1\times15$ seconds) or 25 standard visits ($2\times15$ seconds).

\medskip
\noindent \textbf{Solar System Objects (SSO) -- } Four of the mini-surveys in Table \ref{tab:ddfms} have science goals related to studies of SSO. Observations of the North Ecliptic Spur area could yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (reference: Brandt's talk). The Mini-Moons Mini-Survey aims to find and study the temporarily captured satellites of the Earth (Section 10.2, \citep{2017arXiv170804058L}). The Sweetspot Survey would use twilight fields to find NEOs in Earth-like orbits (i.e., these objects are never in opposition fields, but overhead at sunrise/sunset; Section 10.2, \citep{2017arXiv170804058L}). The Meter-Sized Impactors program would find and track meter-sized impactors $<2$ weeks before impact (Section 10.2, \citep{2017arXiv170804058L}). {\bf Summary:} most of these science goals do not seem to require non-standard visits or exposure times, with the exception of the Sweetspot survey which occurs during twilight and thus may require shorter exposures. The cadence and patterns of these mini-surveys may differ from the WFD main survey, especially when very fast-moving objects are sought. From a processing perspective, it seems that many of these science goals will be achievable by using the products of Solar System Processing, which runs on the Prompt Pipeline's \texttt{DIASource} catalogs after they are updated each night. The exception is finding faint SSOs (e.g., Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) through shift-and-stack (SAS) processing \citedsp{Document-11013}, because SAS is not a capability being built within the DM system and cannot be done solely by reconfiguring DM pipelines. An example of user-generated pipeline for SAS is described in Section \ref{sec:SPCS}.

\medskip
\noindent \textbf{Stars in the Milky Way and Magellanic Clouds -- } As described in \citedsp{Publication-141}, mini-surveys of the Galactic Plane can better distinguish faints stars from faint red galaxies by including at least 3 filters of coverage (e.g., $izy$; similar to WFD), and could mitigate losses from proper motion and increase the detection rate of stellar flares by obtaining all the images in short time span (i.e., a more concentrated cadence than the WFD).  As described in \citedsp{Publication-145}, applying the nominal DDF observing strategy over the full area of the Large and Small Magellanic Clouds can characterize stellar variability to $M_V<6.5$ on timescales from 15 seconds to 3 days. For this, special co-adds may be required, e.g., \textit{"to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"} \citedsp{Publication-145}. The Twilight survey in Table \ref{tab:ddfms} proposes short exposures to enable bright stars to be put on the same photometric system as the deeper LSST WFD main survey catalog, and enable science that is based on their long monitoring baselines from historical observations. In Chapter 10.4 of \citep{2017arXiv170804058L}, a proposed short-exposure survey of M67 would use the camera's stretch goal of $0.1$ second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as $6$ magnitudes brighter than the saturation level"}. {\bf Summary:} while some of these science goals can be accomplished with standard visits, MW \& L/SMC science goals are likely to request shorter exposure times, perhaps down to $0.1$ seconds. These science goals are also likely to propose cadence and filter distributions that are significantly different from the WFD main survey. From a processing perspective, the science goals depending on shorter exposures will only be able to be met by reconfiguring the DM pipelines if the short exposures can be shown to successfully be processed (with, e.g., instrument signature removal); the science goals can likely be met with data products in the same format as the Prompt or DR Pipeline (i.e., {\tt Source} and {\tt Object} catalogs, single visits and deep CoAdds). Although it is not mentioned in the above paragraph, the MW \& L/SMC science community is also most likely to require special processing to extract information from saturated stars, which is outside the scope of DM. See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.

\noindent \textbf{Exoplanets -- } As described in Section 3.1.2 of \citep{2008arXiv0805.2366I}, transiting exoplanets could be detected with the nominal DDF plan, which would allow for $1\%$ variability to be detected over hour-long timescales; a DDF field at Galactic latitude $30$ degrees would yield $10^6$ stars at $r<21$ that would have $\mathrm{SNR}>100$ in each single exposure of the sequence. \citep{2013arXiv1304.3455G} describes how transits can be extract from a wider-area survey of the Galactic Plane, and how microlensing candidates can be found with $\sim22$ mag imaging over the Galactic Plane region every 3-4 days (since microlensing events are slower; these would then require follow-up with external facilities). Dealing with the more crowded fields would be mitigated by the shallower images, in this case. One of the main points of \citep{2013arXiv1304.3455G} is that the Galactic Plane can yield a lot of science despite the fact that its eventual deep co-adds would be uselessly confusion limited, and therefore should not be skipped. \textbf{Summary.} Some of these science goals appear possible with standard visit images, and some might request shorter exposures to avoid confusion in crowded fields when the science can be done with brighter stars. From a processing perspective, the science goals are likely to be achievable with reconfigured DM pipelines, but this depends heavily on performance in crowded fields. See Section \ref{ssec:SPCS_GPVSEx} for a more detailed DM processing case study for Galactic Plane regions.

\noindent \textbf{Supernovae -- } The nominal DDF plan described in \citep{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses. The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$ \citedsp{Publication-144}. High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \citep{2014ApJ...794...23D}. Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days. These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss. SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \citedsp{Publication-142}. Short-exposure observations of bright, nearby SNe may also be useful to include near-peak photometry in the LSST magnitude system, and enable full light-curve analyses. \textbf{Summary.} All of these science goals appear possible with standard visit images (with the exception of a target-of-opportunity short-exposure program to observe bright SNe). From a processing perspecitve, the science goals appear to be accessible with reconfigured DM pipelines to stack and difference the data. In particular, the DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images. See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.

\noindent \textbf{Galaxies -- } The additional depth of a DDF may provide access to a larger collection of low-$\mu$ objects. \citedsp{Publication-142} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies". The DDF stacks could also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. Also, the DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citedsp{Publication-142,Publication-143}. \textbf{Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources. In addition, it seems likely that user-generated algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, and these are beyond the scope of DM.

\noindent \textbf{Weak Lensing -- } The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch). \textbf{Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.
%$\bullet$ \textit{Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} (\texttt{\Large{Community}} forum) \\
%$\bullet$ \textit{Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} (\texttt{\Large{Community}} forum)  \\
%$\bullet$ \textit{MLG side note -- Photo-$z$ are very important to weak lensing \citedsp{Document-10963} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\

### 2022-02-20 DMTN-065.tex
spcs

@@
 This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs.
-The latter includes a discussion about the relevant system requirements, the needed capabilities of the LSST Science Pipelines and the Rubin Science Platform, and the cases in which Special Programs data products would be user-generated.
+The latter includes a discussion about the relevant system requirements, the needed capabilities of the LSST Science Pipelines and the Rubin Science Platform, and the cases in which Special Programs data products would be User-Generated.
 
@@
 
-\subsection{Rubin Requirements Related to Special Programs}
+\subsection{Requirements Related to Special Programs}\label{ssec:proc_reqs}
 % \lsrreq, \ossreq, \dmreq
@@
 Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
+As mentioned above in Section~\ref{ssec:proc_reqs}, the LSST system is required to be able to process exposure times as low as 1 second, and it is known that such short exposures might have degraded image quality.
 
@@
 
-\subsection{Anticipated Rubin-Processed Special Programs Data Products}
+\subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:proc_wfd}
 
-{\bf The LSST Project's Role in Processing Special Programs Data -- } The formal requirements regarding LSST's role in processing Special Programs data are in \citeds{LSE-61}, and the following statements have been derived from those requirements.
-The LSST Project will not take formal responsibility for specialized data reduction algorithms needed to process data, including images taken in non-standard modes.
-The term "specialized algorithms" refers to software that is not already within scope of the LSST Data Management (DM) science pipelines, and may include, for example: difference imaging for short exposures in which the PSF is not well-formed, shift-and-stack for faint moving objects, or any software with computational needs that significantly surpass the processing budget per image (compared to the processing of a WFD image).
-The Project will incorporate Special Programs data into the Prompt and/or Data Release processing pipelines and data products of the WFD Main Survey, such as {\tt Alerts}, {\tt CoAdds}, or {\tt Source} and {\tt Object} catalogs (with appropriate flags; LSST data products are described in \citeds{LSE-163}), whenever this (1) can be accomplished with existing software, and (2) is scientifically beneficial to that data product.
-The Project will also reconfigure its pipelines to generate separate imaging and catalog data products for Special Programs, whenever this can be accomplished with existing software.
-Finally, the Project will enable user-generated processing via the Science Platform (\citeds{LSE-319}), which will provide software tools and computational resources for (re)processing LSST data.
+Instances where the Special Programs data might be included in the Prompt and/or Data Release data products alongside data from the WFD Main Survey -- i.e., the data products described by \citeds{lse-163} -- are preliminarily described.
 
-Whether or not Special Programs images are incorporated into the WFD main survey's data products, it is anticipated that most of the Special Program's science goals will require (or benefit from) separate data products (i.e., CoAdds and/or catalogs).
-For this reason, LSST intends to reconfigure the DM's pipelines in order to generate unique and separate -- but joinable -- imaging and catalog products for Special Programs data, whenever possible.
-In this context, ``possible" means that no new algorithms need to be written and that an intensive amount of additional computational resources is not required for the processing.
+Generally, Rubin Observatory might incorporate Special Programs data into the data products for the WFD Main Survey whenever this is (1) possible and (2) scientifically beneficial.
+Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.
 
-In a ``possible" scenario, DM would assemble a pipeline from existing DM codes in order to process data associated with a given Special Program and build image and catalog products that meet the science needs of that particular program.
-For example, for a DDF SN survey (see Section \ref{ssec:SPCS_SNDDF}), existing DM codes would be used to: (1) make a deep template image from a certain time window, (2) process standard single visit images, (3) create a nightly CoAdd, (4) run difference imaging analysis, (5) run source detection on the difference images, and (6) create \texttt{DIASource} and \texttt{DIAObject} catalog equivalents (this example is also given in Section 6 of the \DPDD, \citedsp{LSE-163}).
-This type of reconfiguration would also be possible to create as a user-generated pipeline (Section \ref{ssec:dmplans_user}), but having these products provided by the Project ensures a consistent and verified level of quality.
-%, as well as access for all users to the processed Special Programs data products, which would only increase the scientific value of Special Programs data.
---> MLG: data rights doc will probably state that the products for SP data that is only processed by user-generated pipelines will have to be served to all users
+\subsubsection{Prompt Processing}
 
-The above statements of intent are derived from the Data Management Subsystem Requirements document, \citeds{LSE-61}, which contains several requirements related to the processing of data from Special Programs (DMS-REQ-0069, 0320, 0321, 0322, and 0344).
-Ensuring that the work-hours needed to reconfigure and test the pipelines, to run them and to verify and validate the data products for public release (which may potentially be needed on intermediate timescales that do not coincide with the Prompt/Yearly timescales, e.g., monthly stacks of deep drilling fields), is the role of Operations.
+It would be both possible and scientifically beneficial to include all standard and alternative standard visits from Special Programs in Prompt processing and Alert Production, alongside the standard visits from the WFD Main Survey.
 
-DMS-REQ-0320 states that "it shall be possible for special programs to trigger their own data processing recipes".
-A header keyword identifying an image as related to a Special Program will be included (e.g., the survey name), and that will be sufficient to send it to a dedicated processing pipeline.
+Non-standard visits that can be processed with the Prompt pipeline in accordance with the relevant requirements might also be included.
 
-\subsection{Anticipated User-Generated Special Programs Data Products}\label{ssec:dmplans_user}
+Very short and very long exposure times might be excluded if they would need specialized algorithms or templates\footnote{If a Special Program's science goals do require specialized templates and Prompt processing, the DMS will have the capability to load and use an alternative template for some regions, based on the image metadata.
+However, there would not be enough memory to hold alternative templates for the whole sky region.}.
+Visits in sky regions that Rubin had not previously observed would not be able to contribute to Alert Production until a template could be generated.
 
-In cases where the science goals of a Special Program require specialized algorithms and cannot be achieved by reconfiguring DM's software, then user-generated pipelines will be needed.
-Towards this end, LSST DM is making all of its software open-source, and preparing the Science Platform (\citeds{LSE-319}), through which users can access the tools and computational resources to assemble data processing pipelines to achieve their science goals (whether related to Special Programs data or not).
-During Operations, there will be a method for the system to allocation processing resources in the case of over-subscription.
+The data products from Special Programs images would thus contribute to all of the Prompt data products described in Section 3 of the DPDD \citedsp{lse-163}. 
+Images and catalog rows would be tagged with a program identifier indicating their Special Program of origin, so that the associated data products can be discovered and extracted by users.
 
-If the user-generated processing pipeline for Special Programs data requires requires significantly more computational resources than have been allocated -- where that allocation has been sized approximately, based on image processing for WFD main survey data (i.e., difference imaging, source detection, and/or stacking) -- then external computational resources may be necessary.
-To support such external processing DM intends to make the data and its code base accessible to and exportable by users in the science community.
-No user-generated pipeline may contribute Alerts to the Alert Stream.
-%, although a separate stream should be possible if the packet and transport formats are adopted (see also Section \ref{ssec:dmplans_user}).
+There are two potential issues which might complicate Prompt processing for Special Programs which obtain a sequence of images without slewing, such as the Deep Drilling Fields.
+(1) As alert packets contain the full records of all associated {\tt DIASources} from the past 12 months, alerts for {\tt DIAObjects} in the DDFs might become very large.
+(2) The association of {\it new} {\tt DIASources} into {\tt DIAObjects} will be somewhat compromised for a DDF sequence.
+The processing for the second image begins when the processing for the first image is only halfway done -- when the {\tt DIAObject} catalog has not yet updated with the new {\tt DIASources} detected in the first image.
+Thus, the {\tt DIASource} from images one and two for a new transient would not be associated with a single {\tt DIAObject}, but instead would each instantiate a new {\tt DIAObject}.
 
-It is furthermore expected that, over time, some user-designed pipelines might become ``adopted", installed and operated (and change controlled) by the LSST Operations team.
-For both adopted and user-run code, whether they are for Special Programs or WFD survey data, the LSST DM team will encourage and facilitate data product databases that are built with the same schema as -- and can easily be joined with -- the tables of the Prompt and DRP data products.
-An alternative option to ``adopted" code is ``adopted" data products: situations in which user-generated code is run externally, a data catalog is returned to LSST to be ingested, verified, and made public.
+As a side note, no User-Generated pipeline may contribute Alerts to the LSST Alert Stream, and since the latency on image availability is ``within 24 hours" (\reqparam{L1PublicT}) no User-Generated pipeline would be able to obtain and process Special Programs data on a timescale similar to the Alert Stream (60 seconds, \reqparam{OTT1}).
 
-\subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:dmplans_WFD}
+\subsubsection{Solar System Processing}
 
-The Project may incorporate Special Programs data into the WFD main survey's pipelines and data products whenever this is (1) possible and (2) scientifically beneficial.
-This will likely be at the discretion of the data quality assessment team during Operations.
-In the following three sections we project when and how Special Programs data might be incorporated into the pipelines and data products of the Prompt pipeline and the Alert stream (Section \ref{ssec:dmplans_prompt}) and the annual Data Release pipeline (Section \ref{ssec:dmplans_drp}).
+Since Solar System Processing takes \texttt{DIASources} as input, so any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
 
-\subsubsection{The Prompt Pipeline and Alert Generation}\label{ssec:dmplans_prompt}
+\subsubsection{Data Release Processing}
 
-It would be beneficial to transient science to include as many LSST images into the Difference Imaging Analysis (DIA) pipeline and the Alert Stream as possible.
-Only images that can be processed with the Prompt DIA pipeline with the $60$-second timeline can contribute to the Alert Stream.
-As discussed in Section \ref{ssec:proc_datadiv}, this might prohibit exposures shorter than $<15$ seconds and/or visit cadences shorter than $1$ per $\sim30$ seconds.
-This might also prohibit the inclusion of very crowded fields that require more computational resources.
-A field must have an LSST template image to be processed by the DIA pipeline, which would prohibit immediate Alerts from new survey fields.
-It should be possible to load and use an alternative template image (than what would be used for that field if and when it is covered by the WFD main survey) in the Prompt pipeline for fields from a Special Program.
+It would be both possible and scientifically beneficial to include all standard and alternative standard visits from Special Programs in {\it some aspects} of the Data Release processing -- such as the repository of processed visit images and the {\tt Source} and {\tt Forced Source} catalogs -- alongside the standard visits from the WFD Main Survey.
+Images and catalog rows would be tagged with a program identifier indicating their Special Program of origin, so that the associated data products can be discovered and extracted by users.
 
-There may be a couple of issues encountered with Alerts from many consecutive visits of Deep Drilling fields.
-One is that, since the Alert contains the full record of all associated {\tt DIASources} from the past 12 months (\citeds{LSE-163}), for a Deep Drilling Field with significantly more visits over the year, the size of the Alert might become prohibitively large (TBD).
-Another is that the self-consistency of the \texttt{DIAObjects} catalog may suffer during consecutive visits of a single field.
-For example, the processing for image $2$ of a sequence would begin when the processing for image $1$ is only halfway complete.
-Any new {\tt DIASource} in image $1$ that cannot be associated (by coordinate) with an existing {\tt DIAObject} becomes a new {\tt DIAObject}.
-When this source is again detected in image $2$, another new {\tt DIAObject} would be created if the catalog has not yet been updated.
+However, some of the core LSST science goals require a WFD Main Survey data products of {\it uniform depth}.
+Whether and how to include any Special Programs data in the deep image coadds and the corresponding {\tt Object} catalog is left to the discretion of the appropriate Rubin teams in Operations.
 
-% \textbf{Templates:} The template images that are used in the Level 1 difference imaging pipeline will be built from the Level 2 DRP, and so the first factor affecting a Special Programs image's suitability for Level 1 is to be in a region of sky with an existing template.
-So long as there is a template, when the exposure time is equivalent to a WFD visit image, $\sim 30$ seconds, treating the image as Level 1 is going to be fine.
-When processing Special Programs data with the Level 1 pipeline, certain science cases might call for the capability to load and use a certain template that is different from the WFD template (i.e., built over a different timescale).
-K.-T.
-confirms that there is not enough memory allotted to store more than one template over the whole sky, but for sub-regions, storing and using an alternative template should be possible.
-This is not an issue limited to Special Programs, since during commissioning it is conceivable that multiple template versions.
-K.-T.
-also confirms that the capability for the processing pipeline to choose a given template based on the programID in the raw image metadata will exist.
+For example, perhaps Special Programs images will only be included when they bring additional area to the same depth as the rest of the WFD Main Survey, or when they suppress edge effects or low-order modes in the all-sky photometric solutions.
 
-% \textbf{Processing backup:} Are there options for short-term increases in parallel processing power at NCSA?
-(Such options might be needed anyway to process crowded fields with $>$10k \texttt{DIASources}.) The bandwidth needed to load templates at a faster rate is also a concern, since templates have twice as many pixels, if the data acquistion rate is surveying at twice the WFD main survey rate, that's a $4\times$ additional bandwidth load.
-However, at NCSA there should be ways to elastically change the amount of processing power available.
-This is also an issue for crowded fields (below).
+\subsection{Anticipated Unique and Separate Special Programs Data Products}\label{ssec:proc_spdp}
 
-% Special Programs are more likely to include crowded fields than the WFD main survey area.
-Due to the increased number of sources, the number of \texttt{DIASources} -- and therefore the number of Alerts -- increases as well.
-In turn, this increases the processing time and in some cases, may exceed the 60 second limit for Alert Production.
-A policy is needed on whether Alerts from crowded fields should be allowed a delay, or allowed to be incomplete.
-K.-T.
-reports that the control system can easily kill processes that are running over time and move forward with existing outputs -- but that perhaps it will be just as easy to let it keep running and elastically grab additional NCSA resources as needed.
-This appears possible because the batch system is larger than the Level 1 allocation, although we might not know that this is possible until Level 1 integration happens at NCSA (see Document LDM-230, the operations concepts).
+The anticipated divisions between what will be Rubin- versus User-Generated for the variety of anticipated ``unique and separate" Special Programs data products, which are designed to meet the specific science goals of the Special Programs, are preliminarily described.
 
-{\bf Solar System Processing -- } Since Solar System Processing takes \texttt{DIASources} as input, any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
-As discussed under "Solar System Objects (SSO)" in Appendix \ref{sec:prevpropsp}, most of the Special Programs data associated with SS science will obtain standard visit images anyway.
-%There was some concern that a large number of small-separation sources might overwhelm the processing system (i.e., from a deep drilling field with many exposures in a sequence), but upon further consideration this worry was rejected.
+For Special Programs which will require User-Generated pipelines and data products, note that users will have access to all of the LSST Science Pipelines and its data processing infrastructure, as well as dedicated computational resources next-to-the-data (via the Rubin Science Platform; \citeds{LSE-319}).
+However, computationally intense processing might require external resources.
 
-\subsubsection{The Data Release Pipeline (DRP)}\label{ssec:dmplans_drp}
+Furthermore, it is expected that some User-Generated pipelines and data products might be ``adopted" or ``federated" into the LSST Science Pipelines and the Prompt and Data Release data products. 
+Details regarding this process are to be provided elsewhere.
 
-This document is not the place for a full consideration of whether or not it would be ``scientifically beneficial" to include any Special Programs data in the DRP data products -- namely, the deep image CoAdds and their corresponding {\tt Source} and {\tt Object} catalogs (\citeds{LSE-163}) -- and we leave that decision for the data quality assessment team in LSST Operations.
-One example might be when Special Programs data brings additional area up to the same level of depth and cadence as the rest of the WFD main survey.
-Another may be if including some or all of the shallower Galactic Plane coverage suppresses edge effects or low-order modes in the all-sky photometric solutions.
+Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.
+
+\subsubsection{Deep Drilling Fields}
+
+{\bf Rubin-generated data products} might include:\\
+ - nightly-coadded images (24 h)\\
+ - nightly-coadded difference images (24 h)\\
+ - deeply-coadded images (all images to date; yearly)\\
+ - templates for the nightly-coadded difference images (yearly)\\
+ - Source- and Object-like catalogs for the night and deep coadds (yearly)\\
+ - DIASource- and DIAObject-like catalogs for the nightly-coadds (24 h)\\
+
+{\bf User-Generated data products} might include, for example, DDF images coadded on other timescales, or using algorithms outside of the LSST Science Pipelines.
+
+\subsubsection{Short-Exposure Surveys}
+
+\subsubsection{GW Target-of-Opportunity}
+
+\subsubsection{Deeper $g$-band for 10 Local Volume Galaxies}
+
+\subsubsection{Micro-Surveys}
+
+For example, the additional sky areas covered with a cadence and depth that is different from the WFD Main Survey, such as the proposed Special Programs for the Roman bulge field; deeper $g$-band imaging of 10 local volume galaxies; the SMC; the Carina nebula; the northern stripe; and the Virgo Cluster.
 
@@
 \end{itemize}
-
-\clearpage
-% % % % % % % % % % % % % % % % % %
-\section{Special Programs Processing Examples}\label{sec:SPCS}
-
-For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-Note that we are not including any analysis in these descriptions, only processing and products.
-These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
-
-Basic steps that we use to describe a processing case study: \\
-Step 1. Data Acquisition. \\
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Step 3. Delivery of LSST Processed Images. \\
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-Step 6. User-Generated Pipelines and Products. \\
-
-\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
-
-This Special Programs processing summary is based on Becker et al. (2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
-
-Step 1. Data Acquisition. \\
-The observational sequence is triggered.
-In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
-This sequence is always repeated 2-3 nights later.
-This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later.
-Data obtained in the $g$-band filter is also acceptable.
-\citedsp{Document-11013}
-
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
-
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-Shift-and-stack processing is beyond the scope of DM's algorithms.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.
-
-Step 6. User-Generated Pipelines and Products. \\
-The user-generated pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external processor.
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The DRP difference imaging routine will be used with the same template tract/patch for all.
-Custom, user-generated algorithms will shift the exposures and create difference images, then DRP routines can stack and do source detection and characterization and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to a {\tt SSObjects}-like database.
-
-%\input{spcs_sas}
-
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
-
-Step 1. Data Acquisition. \\
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
-
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed by the Prompt pipeline's DIA, and Alerts are released within 60 seconds.
-They are flagged to denote the image source is a DDF and that source association might be compromised.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-The required data products for this science goal can be met by reconfiguring the DM pipelines.
-First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF.
-The LSST codes for alert packet and transport could be used to distribute the detected objects e.g., to the same brokers that receive the Alert Stream, or alternative destinations.
-However, these packets would not be distributed via the LSST {\tt Alert Stream}, and would need to be identified as, e.g., DDF Alerts.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-
-Step 6. User-Generated Pipelines and Products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, no separate user-generated software appears necessary.
-
-\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
-
-Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
-The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
-
-Step 1. Data Acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. Location and exposure times are set by the sky brightness and desired saturation limits.
-
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.
-
-Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-This is officially not determined, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-These short-exposure, high sky background images would not contribute to the DRP data products created for the WFD survey.
-
-Step 6. User-Generated Pipelines and Products. \\
-If short-exposure images cannot be processed with the existing DM algorithms, a user-generated processing pipeline might be needed to reduce the raw data.
-
-Side note: A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-This would be considered a user-generated algorithm.
-
-\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
-
-Step 1. Data Acquisition. \\
-The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields (or shorter, for a shallower depth than the WFD main survey).
-
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped if they take longer to process and violate the $60$ second latency for Alerts.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Prompt and DRP pipelines, so it seems that not much reconfiguration would be needed.
-The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-
-Step 6. User-Generated Pipelines and Products. \\
-It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
-
-\subsection{Gravitational Wave Event Follow-Up}\label{ssec:SPCS_GW}
-
-For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, see \citeds{rtn-008}.
 
@@
 
+\input{ap_spcs.tex}
+
 \input{ap_prevpropsp.tex}

### 2022-02-20 ap_spcs.tex
spcs

\section{Special Programs Processing Examples}\label{sec:spcs}

{\bf This section has not been updated since 2018.}

For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
Note that we are not including any analysis in these descriptions, only processing and products. These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.

Basic steps that we use to describe a processing case study: \\
Step 1. Data Acquisition. \\
Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
Step 3. Delivery of LSST Processed Images. \\
Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
Step 6. User-Generated Pipelines and Products. \\

\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}

This Special Programs processing summary is based on Becker et al. (2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.

Step 1. Data Acquisition. \\
The observational sequence is triggered. In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures. This sequence is always repeated 2-3 nights later. This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later. Data obtained in the $g$-band filter is also acceptable. \citedsp{Document-11013}

Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.

Step 3. Delivery of LSST Processed Images. \\
The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.

Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
Shift-and-stack processing is beyond the scope of DM's algorithms.

Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial. However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.

Step 6. User-Generated Pipelines and Products. \\
The User-Generated pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external processor. Pipeline inputs will be the 336 processed exposures per field per re-visit sequence. The DRP difference imaging routine will be used with the same template tract/patch for all. Custom, User-Generated algorithms will shift the exposures and create difference images, then DRP routines can stack and do source detection and characterization and generate an object database. Custom code will derive orbital parameters for the detections and add them to a {\tt SSObjects}-like database.

%\input{spcs_sas}

\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}

Step 1. Data Acquisition. \\
On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.

Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
Each $2\times15$ second visit is processed by the Prompt pipeline's DIA, and Alerts are released within 60 seconds. They are flagged to denote the image source is a DDF and that source association might be compromised.

Step 3. Delivery of LSST Processed Images. \\
The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).

Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
The required data products for this science goal can be met by reconfiguring the DM pipelines. First, a template image for the field will be made using DM stacking algorithms. On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF. The LSST codes for alert packet and transport could be used to distribute the detected objects e.g., to the same brokers that receive the Alert Stream, or alternative destinations. However, these packets would not be distributed via the LSST {\tt Alert Stream}, and would need to be identified as, e.g., DDF Alerts.

Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.

Step 6. User-Generated Pipelines and Products. \\
For the science goal of searching for supernovae in nightly stacked DDF images, no separate User-Generated software appears necessary.

\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}

Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects. The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.

Step 1. Data Acquisition. \\
At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. Location and exposure times are set by the sky brightness and desired saturation limits.

Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.

Step 3. Delivery of LSST Processed Images. \\
Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).

Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
This is officially not determined, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.

Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
These short-exposure, high sky background images would not contribute to the DRP data products created for the WFD survey.

Step 6. User-Generated Pipelines and Products. \\
If short-exposure images cannot be processed with the existing DM algorithms, a User-Generated processing pipeline might be needed to reduce the raw data.

Side note: A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}. This would be considered a User-Generated algorithm.

\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}

Step 1. Data Acquisition. \\
The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields (or shorter, for a shallower depth than the WFD main survey).

Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds. Extremely crowded fields might have to be skipped if they take longer to process and violate the $60$ second latency for Alerts.

Step 3. Delivery of LSST Processed Images. \\
The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).

Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Prompt and DRP pipelines, so it seems that not much reconfiguration would be needed. The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.

Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.

Step 6. User-Generated Pipelines and Products. \\
It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.

\subsection{Gravitational Wave Event Follow-Up}\label{ssec:SPCS_GW}

For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, see \citeds{rtn-008}.

% \subsection{A More General Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}

% A Level 3 processing case study for shift-and-stack on a large number images. By Mario.

% \#1. The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec). [ and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].

% \#2. The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database. This will happen automatically for all images (perhaps within some range of exposure times?).

% \#3. The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes. This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it? or is is the same one that's used to process calibrations? have these systems been sized?].

% \#3 a). The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team. That is, we don't expect someone external to the ops team to babysit the code on a nightly basis. In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.

% \#4. There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution. We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.

% \#5. Once the processing finishes, the results of will be stored to a program-specific database. No alerts (in VOEvent sense) will be issued. We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store. [This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].

% \#6. The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables). The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files). The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).

% \#7. When the images are made available to the batch system (step \#3), they also become available to *everyone*. I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database. (This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_user}).

### 2022-02-20 DMTN-065.tex
fix lse

@@
 
-For Special Programs which will require User-Generated pipelines and data products, note that users will have access to all of the LSST Science Pipelines and its data processing infrastructure, as well as dedicated computational resources next-to-the-data (via the Rubin Science Platform; \citeds{LSE-319}).
-However, computationally intense processing might require external resources.
+For Special Programs which will require User-Generated pipelines and data products, note that users will have access to all of the LSST Science Pipelines and its data processing infrastructure, as well as dedicated computational resources next-to-the-data (via the Rubin Science Platform; \citeds{lse-319}).
+Details of the planned ``User Batch" facility for data processing are described in \citeds{dmtn-202}.
+However, very computationally intense processing might require external resources.
 

### 2022-02-20 DMTN-065.tex
fix lse

@@
 Generally, Rubin Observatory might incorporate Special Programs data into the data products for the WFD Main Survey whenever this is (1) possible and (2) scientifically beneficial.
-Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.
+{\bf Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
 
@@
 
-The anticipated divisions between what will be Rubin- versus User-Generated for the variety of anticipated ``unique and separate" Special Programs data products, which are designed to meet the specific science goals of the Special Programs, are preliminarily described.
+As mentioned in Section~\ref{ssec:proc_reqs}, Rubin Observatory and the LSST system are required to process Special Programs data to produce unique and separate data products whenever the original or reconfigured versions of the LSST Science Pipelines can be used.
+
+Cases where the development of new algorithms, or the allocation of significant additional computational resources, are required in order to produce unique and separate data products will require User-Generated processing.
+
+The anticipated divisions between what will be Rubin- versus User-Generated for the variety of anticipated ``unique and separate" Special Programs data products, which are designed to meet the specific science goals of the Special Programs, are {\it preliminarily} described.
+{\bf Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
 
@@
 
+As the DDFs will likely be observed with standard or alternative standard visits, it is likely that Rubin Observatory will be able to reconfigure existing pipelines to process the DDF data and produce ``unique and separate" DDF data products.
+
 {\bf Rubin-generated data products} might include:\\
@@
 \subsubsection{Short-Exposure Surveys}
+
+Short-exposure images obtained during twilight, which will have a very bright sky background unlike other LSST images, might require User-Generated algorithms to subtract the high sky background. 
+Short-exposure images obtained during the night would have fewer stars for astrometric and photometric calibration, and might require User-Generated processing pipelines.
 

### 2022-02-20 DMTN-065.tex
done w S 2.4

@@
 Generally, Rubin Observatory might incorporate Special Programs data into the data products for the WFD Main Survey whenever this is (1) possible and (2) scientifically beneficial.
-{\bf Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
+{\bf Decisions about when and whether to include Special Programs data in the WFD Main Survey's data products are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
 
-\subsubsection{Prompt Processing}
+\subsubsection{Prompt}
 
@@
 
-\subsubsection{Data Release Processing}
+\subsubsection{Data Release}
 
@@
 
-\subsection{Anticipated Unique and Separate Special Programs Data Products}\label{ssec:proc_spdp}
+\subsection{Anticipated Rubin- and User-Generated Special Programs Data Products}\label{ssec:proc_spdp}
 
@@
 The anticipated divisions between what will be Rubin- versus User-Generated for the variety of anticipated ``unique and separate" Special Programs data products, which are designed to meet the specific science goals of the Special Programs, are {\it preliminarily} described.
-{\bf Decisions about when and whether to do this are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
+{\bf Decisions about when and whether Rubin Observatory will generate Special Programs data products are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
 
@@
 
-{\bf Rubin-generated data products} might include:\\
+Rubin-generated data products might include:\\
  - nightly-coadded images (24 h)\\
@@
 
-{\bf User-Generated data products} might include, for example, DDF images coadded on other timescales, or using algorithms outside of the LSST Science Pipelines.
+User-Generated data products might include, for example, DDF images coadded on other timescales, or using algorithms outside of the LSST Science Pipelines.
 
@@
 
+If the Rubin Operations team decides that the LSST Science Pipelines can be reconfigured and used for short exposures, then the Rubin-generated data products would likely be similar to the Data Release data products (Section 4, \citeds{lse-163}).
+For example, a separate repository of processed visit images and coadded images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and {\tt Object} catalogs.
+
 \subsubsection{GW Target-of-Opportunity}
 
-\subsubsection{Deeper $g$-band for 10 Local Volume Galaxies}
+As the science goals of any target-of-opportunity program are the immediate processing and discovery of new transient phenomena, it is very likely that all ToO programs would be designed to use standard or alternative standard visits and Prompt processing data products such as Alerts.
+Prompt processing would proceed automatically for all ToO for which an LSST template image exists.
+
+Options for Rubin Observatory to assist with or expedite the processing of ToO programs, especially during the first year of Operations when the template coverage will be low, are discussed in more detail in \citeds{rtn-008}.
+
+Other science-driven data products from ToO programs, such as custom deep stacks and difference images to find $<$5$\sigma$ transients in the fields on intermediate timescales in cases where the optical counterpart was not detected, would require User-Generated processing.
 
@@
 
-For example, the additional sky areas covered with a cadence and depth that is different from the WFD Main Survey, such as the proposed Special Programs for the Roman bulge field; deeper $g$-band imaging of 10 local volume galaxies; the SMC; the Carina nebula; the northern stripe; and the Virgo Cluster.
+Potential micro-surveys include additional sky areas, in some cases covered to a depth that is different from the WFD Main Survey, such as the proposed micro-surveys for the Roman bulge field; deeper $g$-band imaging of 10 local volume galaxies; the SMC; the Carina nebula; the northern stripe; and the Virgo Cluster.
+
+In all of these cases, if the Rubin Operations team decides that the LSST Science Pipelines can be reconfigured and used for the visits associated with these micro-surveys -- which is likely if they're using standard or alternative standard visits --  then the Rubin-generated data products would be similar to the Data Release data products (Section 4, \citeds{lse-163}).
+For example, separate repositories of processed visit images and coadded images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and {\tt Object} catalogs would be created for all of these micro-surveys.
+
+In cases where data for the micro-survey is obtained throughout the year, then the separate data products would also be produced on a yearly basis.
+In cases where the data might be all obtained within a week (e.g., Carina nebular survey), such data products would likely be generated on a shorter (intermediate) timescales.
+
+Micro-surveys with time-domain science goals that aren't met by the Prompt pipelines -- for example if they require difference imaging with coadded images on an intermediate timescale (e.g., a weekly stack) --  might require User-Generated processing.
 

### 2022-02-21 DMTN-065.tex
fix LPG initials
fix hardbound camera quote

@@
 
-\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~Guy, and the Data Management System Science Team}
+\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~P.~Guy, and the Data Management System Science Team}
 
@@
 
-The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is currently a hardware boundary that limits the readout rate to 1 every 15 seconds.
-This would affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys.
+The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is a {\it potential} hardware boundary that limits the readout rate to 1 every 15 seconds.
+This {\it might} affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys -- but current testing of the camera system indicates that imposing waits between short exposures will not be necessary.
 

### 2022-02-21 DMTN-065.tex
fix reqparam refs

@@
 
-As a side note, no User-Generated pipeline may contribute Alerts to the LSST Alert Stream, and since the latency on image availability is ``within 24 hours" (\reqparam{L1PublicT}) no User-Generated pipeline would be able to obtain and process Special Programs data on a timescale similar to the Alert Stream (60 seconds, \reqparam{OTT1}).
+As a side note, no User-Generated pipeline may contribute Alerts to the LSST Alert Stream, and since the latency on image availability is ``within 24 hours" ({\tt L1PublicT}\reqparam{L1PublicT}) no User-Generated pipeline would be able to obtain and process Special Programs data on a timescale similar to the Alert Stream (60 seconds, {\tt OTT1}\reqparam{OTT1}).
 

### 2022-02-21 ap_docrev.tex
renamed ap prev proc sp
spcs
fix lse
fix reqparam refs

@@
-\section{Rubin Documentation Review}\label{sec:docrev}
+\section{Documentation Review for Requirements Related to Special Programs}\label{sec:docrev}
 
@@
 
-In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\req{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second (\reqparam{minExpTime}).
+In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\req{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second ({\tt minExpTime}\reqparam{minExpTime}).
 The discussion notes that {\it ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g.
@@
 
-In Section 2.7.1.6, ``Community Computing Services", LSR-REQ-0055\lsrreq{0055} requires that the LSST system {\it ``shall provide and maintain an amount of computing capacity equivalent to at least"} 10\% (\reqparam{userComputingFraction}) {\it ``of the total LSST data processing capacity (computing and storage) for the purpose of scientific analysis of LSST data and the production of"} User-Generated data products.
+In Section 2.7.1.6, ``Community Computing Services", LSR-REQ-0055\lsrreq{0055} requires that the LSST system {\it ``shall provide and maintain an amount of computing capacity equivalent to at least"} 10\% ({\tt userComputingFraction}\reqparam{userComputingFraction}) {\it ``of the total LSST data processing capacity (computing and storage) for the purpose of scientific analysis of LSST data and the production of"} User-Generated data products.
 The discussion clarifies that the scope of this service remains to be determined.
@@
 
-In Section 3.6.1.4, ``Minimum Exposure Time", OSS-REQ-0291\ossreq{0291} specifies that {\it ``the camera shall be able to obtain a single exposure with an effective minimum exposure time of no more than"} 1 second (\reqparam{minExpTime}) {\it ``with a goal of an effective minimum exposure time of"} 0.1 seconds (\reqparam{minExpTimeGoal}).
+In Section 3.6.1.4, ``Minimum Exposure Time", OSS-REQ-0291\ossreq{0291} specifies that {\it ``the camera shall be able to obtain a single exposure with an effective minimum exposure time of no more than"} 1 second ({\tt minExpTime}\reqparam{minExpTime}) {\it ``with a goal of an effective minimum exposure time of"} 0.1 seconds ({\tt minExpTimeGoal}\reqparam{minExpTimeGoal}).
 The discussion clarifies that {\it ``if the exposure is shortened from the 15 second nominal, the spacing between successive exposures should be extended to maintain the average readout rate consistent with a 15 second exposure"}, which may increase the overheads of Special Programs using short exposure times.
@@
 
-In Section 3.6.2.1.2, ``Maximum time for operational filter change", OSS-REQ-0293\ossreq{0293} specifies that {\it ``the camera system shall provide the capability of changing the operational filter with any other internal filter in a time less than"} 120 seconds (\reqparam{tFilterChange}).
+In Section 3.6.2.1.2, ``Maximum time for operational filter change", OSS-REQ-0293\ossreq{0293} specifies that {\it ``the camera system shall provide the capability of changing the operational filter with any other internal filter in a time less than"} 120 seconds ({\tt tFilterChange}\reqparam{tFilterChange}).
 This would impose a large overhead on, e.g., a Special Program that changes filters often without slewing.
@@
 
-In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour (\reqparam{rotTrackTime}).
+In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
 The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field}.
@@
 
-In Section 3.6.3.10, ``Non-Sidereal Tracking", OSS-REQ-0380\ossreq{0380} specifies that {\it ``the LSST system shall be capable of tracking in an arbitrary direction on the sky along a parametric RA(t) and DEC(t) trajectory, at angular rates of up to"} 220 arcseconds per second (\reqparam{nonsiderealAngularRateEl} and \reqparam{nonsiderealAngularRateAZ}) {\it ``with a tracking error not to exceed"} 0.5 arcseconds per minute (\reqparam{nonsiderealTrackingError}).
+In Section 3.6.3.10, ``Non-Sidereal Tracking", OSS-REQ-0380\ossreq{0380} specifies that {\it ``the LSST system shall be capable of tracking in an arbitrary direction on the sky along a parametric RA(t) and DEC(t) trajectory, at angular rates of up to"} 220 arcseconds per second ({\tt nonsiderealAngularRateEl}\reqparam{nonsiderealAngularRateEl} and {\tt nonsiderealAngularRateAZ}\reqparam{nonsiderealAngularRateAZ}) {\it ``with a tracking error not to exceed"} 0.5 arcseconds per minute ({\tt nonsiderealTrackingError}\reqparam{nonsiderealTrackingError}).
 The discussion notes that {\it ``this is standard capability for modern telescopes"}, but might be relevant to some Special Programs.
@@
 
-In Section 1.6.4, ``Constraints on Level 1 Special Program Products Generation", DMS-REQ-0344\dmreq{0344} specifies that {\it ``the publishing of [Prompt] data products from Special Programs shall be subject to the same performance requirements of"} 24 hours (\reqparam{L1PublicT}) for the release of Prompt data products and 1 minute (\reqparam{OTT1}) for the transmission of Alert packets.
+In Section 1.6.4, ``Constraints on Level 1 Special Program Products Generation", DMS-REQ-0344\dmreq{0344} specifies that {\it ``the publishing of [Prompt] data products from Special Programs shall be subject to the same performance requirements of"} 24 hours ({\tt L1PublicT}\reqparam{L1PublicT}) for the release of Prompt data products and 1 minute ({\tt OTT1}\reqparam{OTT1}) for the transmission of Alert packets.
 This is essentially a more detailed version of DMS-REQ-0321 which includes the Alert production timescale.
@@
 
-Version 3.6 (revision 2021-12-17), \citeds{LSE-163}.
+Version 3.6 (revision 2021-12-17), \citeds{lse-163}.
 

### 2022-02-21 ap_docrev.tex
update hardbounds pointing

@@
 
+In Section 3.6.3.1, ``Absolute Pointing", OSS-REQ-0298\ossreq{0298} specifies that {\it ``the LSST shall point to a defined set of sky coordinates with an RMS accuracy
+of" 2 arcseconds ({\tt absPointErr}\reqparam{absPointErr}).
+
 In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
@@
 There do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits.
+
+In Section 3.6.3.5, ``Offset Pointing", OSS-REQ-0302\ossreq{0302} specifies that {\it ``the LSST shall be capable of offset pointing within a single field-of-view with a precision of no more than" 0.2 arcseconds ({\tt offsetPointingErr}\reqparam{offsetPointingErr}).
+
+In Section 3.6.3.6, ``Open Loop Tracking", OSS-REQ-0303\ossreq{0303} specifies that {\it ``The LSST shall be capable of open loop tracking without the assistance of real
+time optical feedback to an accuracy of"} 1.0 arcseconds ({\tt openTrackErr}\reqparam{openTrackErr}) {\it ``over any 10 minute duration during normal night time operations"}.
+Note that the open loop tracking requirement is \emph{without guiding}.
+
+In Section 3.6.3.8, ``Guiding", OSS-REQ-0305\ossreq{0305} specifies that {\it ``the LSST system shall improve the tracking of the optical system to the sky by means of active guiding over the whole accessible sky"}.
 

### 2022-02-21 ap_hardbounds.tex
update hardbounds pointing

@@
 Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
-The $15$ interval between images is a potential hardware boundary on the potential diversity of data products.
+This potential $15$ interval between images is also a potential hardware boundary on the potential diversity of data products.
+Current (2022) testing of the camera system indicates that imposing waits between short exposures will not be necessary
 
@@
 
-\subsection{Telescope Tracking}
+\subsection{Pointing and Sidereal Tracking}
+The specifications for the telescope's pointing and tracking in \citeds{LSE-30} indicate that $<$0.2 arcsecond precision in field pointing (OSS-REQ-0302) and $<$1 arcsecond in open-loop tracking (OSS-REQ-0303) would not be possible, but guiding would improve the latter (OSS-REQ-0305).
+Furthermore, obtaining the \emph{exact same} alignment of the pixel grid in RA-Dec {\it ``would put demands on the camera rotator that were not planned"}\footnote{As per C.
+Claver's comments in ticket DM-12573.}.
+
+\subsection{Non-Sidereal Tracking}
 The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.

### 2022-02-21 ap_docrev.tex
fix typo

@@
 In Section 3.6.3.1, ``Absolute Pointing", OSS-REQ-0298\ossreq{0298} specifies that {\it ``the LSST shall point to a defined set of sky coordinates with an RMS accuracy
-of" 2 arcseconds ({\tt absPointErr}\reqparam{absPointErr}).
+of"} 2 arcseconds ({\tt absPointErr}\reqparam{absPointErr}).
 
 In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
-The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field}.
+The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field"}.
 There do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits.

### 2022-02-21 ap_docrev.tex
fix typo

@@
 
-In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\req{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second ({\tt minExpTime}\reqparam{minExpTime}).
+In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\lsrreq{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second ({\tt minExpTime}\reqparam{minExpTime}).
 The discussion notes that {\it ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g.
@@
 
-In Section 3.6.3.5, ``Offset Pointing", OSS-REQ-0302\ossreq{0302} specifies that {\it ``the LSST shall be capable of offset pointing within a single field-of-view with a precision of no more than" 0.2 arcseconds ({\tt offsetPointingErr}\reqparam{offsetPointingErr}).
+In Section 3.6.3.5, ``Offset Pointing", OSS-REQ-0302\ossreq{0302} specifies that {\it ``the LSST shall be capable of offset pointing within a single field-of-view with a precision of no more than"} 0.2 arcseconds ({\tt offsetPointingErr}\reqparam{offsetPointingErr}).
 
-In Section 3.6.3.6, ``Open Loop Tracking", OSS-REQ-0303\ossreq{0303} specifies that {\it ``The LSST shall be capable of open loop tracking without the assistance of real
-time optical feedback to an accuracy of"} 1.0 arcseconds ({\tt openTrackErr}\reqparam{openTrackErr}) {\it ``over any 10 minute duration during normal night time operations"}.
+In Section 3.6.3.6, ``Open Loop Tracking", OSS-REQ-0303\ossreq{0303} specifies that {\it ``The LSST shall be capable of open loop tracking without the assistance of real time optical feedback to an accuracy of"} 1.0 arcseconds ({\tt openTrackErr}\reqparam{openTrackErr}) {\it ``over any 10 minute duration during normal night time operations"}.
 Note that the open loop tracking requirement is \emph{without guiding}.

### 2022-02-21 DMTN-065.tex
update hardbounds pointing
fix WFD definition

@@
 The WFD Main Survey is the core science program of the LSST, designed to achieve the science goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
-The WFD will cover at least $\sim$18000 deg$^2$ and is expected to be accomplished with 85--90\% of the observing time.
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the WFD Main Survey is thought of as several contiguous areas.
-The largest is the low-dust-extinction survey area, which would also receive the most ($>$800) visits over the 10-year survey and be completed to uniform depth, likely with a rolling cadence.
-Smaller regions cover the higher-extinction high stellar density Galactic Bulge and Magellanic Clouds, the North Ecliptic Spur (NES), and the remainder of the Galactic Plane and the South Celestial Pole.
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the WFD Main Survey is thought of as several contiguous areas that cover at least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, and is expected to be accomplished with $\sim$85\% of the observing time.
+These areas include the low-dust-extinction survey area (the largest area), the higher-extinction high stellar density Galactic Bulge, and the Magellanic Clouds.
+
+\subsection{Mini-Surveys}\label{sssec:terms_wfd_mini}
+As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``mini-survey" refers to additional areas that are adjacent to the WFD but are covered with a few hundred (100-300) visits per pointing.
+They are expected to be accomplished with $\sim$10\% of the total survey time.
+Mini-surveys include the North Ecliptic Spur (NES), the remainder of the Galactic Plane, and the South Celestial Pole.
 Boundaries, re-visit cadences, and depths for all WFD areas remain to be determined \citedsp{PSTN-053}.
 
-\subsubsection{Mini-Surveys}\label{sssec:terms_wfd_mini}
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``mini-survey" was being used to refer to the areas of the WFD / Main Survey outside of the large, low-dust-extinction survey area.
-This term used to refer to non-WFD regions (Special Programs).
-
 \subsection{Special Programs}\label{ssec:terms_sp}
-Special Programs are additional survey areas and/or observing strategies that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
-Special Programs will fill the remaining 10--15\% of the available survey time, and include both ``deep drilling" fields and ``micro-surveys".
+Special Programs are survey areas and/or observing strategies, aside from the WFD and mini-surveys, that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
+Special Programs will fill the remaining $\sim$5\% of the available survey time, and include both ``deep drilling" fields and ``micro-surveys".
 
@@
 This {\it might} affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys -- but current testing of the camera system indicates that imposing waits between short exposures will not be necessary.
+
+Special Programs which required the \emph{exact same} field pointing and rotation for \emph{every exposure} (to sub-arcsecond levels) might run into hardware boundaries on pointing and tracking, but none of the currently-proposed programs seem likely to be affected by this potential boundary.
 

### 2022-02-22 DMTN-065.tex
change read interval text

@@
 
-The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is a {\it potential} hardware boundary that limits the readout rate to 1 every 15 seconds.
-This {\it might} affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys -- but current testing of the camera system indicates that imposing waits between short exposures will not be necessary.
+The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is a potential hardware boundary that limits the readout rate to 1 every 15 seconds.
+This would affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys.
 

### 2022-02-22 ap_hardbounds.tex
change read interval text

@@
 Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
+
+As discussed in Jira ticket DM-12573, the main issue is thermal and is related to the shutter, both the motors and the brakes; an elevated Camera skin temperature would affect image quality.
+As of 2022, early tests suggest that a sustained (30 minutes) sequence that increases the heat load by large factors would not work, but further functional testing of the system once the Camera was fully assembled are needed for full characterization of the issue.
+
 This potential $15$ interval between images is also a potential hardware boundary on the potential diversity of data products.
-Current (2022) testing of the camera system indicates that imposing waits between short exposures will not be necessary
 

### 2022-02-24 DMTN-065.tex
updates post-GPDF disc

@@
 
-\subsubsection{Data Products}
+\subsubsection{Data Products}\label{sssec:proc_reqs_dp}
 
@@
 
-\subsubsection{Metadata}
+\subsubsection{Metadata}\label{sssec:proc_reqs_meta}
 
@@
 
-\subsubsection{Processing}
+\subsubsection{Processing}\label{sssec:proc_reqs_proc}
 
-It is a requirement that Special Programs processing with the Prompt pipeline (or a reconfigured version of it) is subject to the same timescales and latency constraints of 24 hours for the release of Prompt data products and 1 minute for the transmission of Alert packets (DMS-REQ-0344).
+It is a requirement that any Special Programs processing done with the Prompt pipeline (or a reconfigured version of it) is subject to the same timescales and latency constraints of 24 hours for the release of Prompt data products and 1 minute for the transmission of Alert packets (DMS-REQ-0344).
 
@@
 
-\subsubsection{User Processing}
+\subsubsection{User Processing}\label{sssec:proc_reqs_ug}
 
@@
 
-It would be both possible and scientifically beneficial to include all standard and alternative standard visits from Special Programs in Prompt processing and Alert Production, alongside the standard visits from the WFD Main Survey.
+It is expected that all Special Programs data that {\it can} be processed by the Prompt processing and Alert Production pipelines {\it would} be, alongside the standard visits from the WFD Main Survey, as this would be scientifically beneficial to the science pillars of the LSST.
 
@@
 
-\subsection{Anticipated Rubin Science Platform Capabilities for Special Programs }
+\subsection{``Discoverability Considerations" (Visit Labels)}
+
+\begin{itemize}
+\item needs for labelling
+\item survey characterizability, selection biases
+\item provenance and origin
+\item not just for separate data products but all regions
+\item what if someone puts a micro survey in NES? labeling useful
+\end{itemize}
+
+\subsection{Anticipated Rubin Science Platform Capabilities for Special Programs}
 
@@
 \end{itemize}
+
+\subsection{Computational Resources for User (Re-)Processing}
+
+As mentioned in Section~\ref{sssec:proc_reqs_ug}, Rubin Observatory will reserve 10\% of its total data processing capacity for users.
+This component would include {\it all} user processing and re-processing of any and all LSST data, including Special Programs.
+
+As described in Section~\ref{ssec:proc_spdp} this processing capacity will be accessed via Rubin Science Platform, with a supported software environment and infrastructure for batch processing \citedsp{dmtn-202}.
+
+Very computationally intense processing (e.g., shift-and-stack for faint moving objects) will likely require external resources\footnote{For more details about the boundary between what Rubin Observatory will provide (in terms of data products and processing resources) and what will be left to the expertise of the science community, see \url{https://www.lsst.org/about/dm/data-products}}.
 

### 2023-05-01 ap_docrev.tex
update

@@
 
-Below are described mentions of Special Programs in Rubin documentations, with a focus on requirements and specifications.
+This review is limited to requirements related to Special Programs processing or data products,
+or requirements that constrain the diversity of images (e.g., exposure time limits).
 
-Updates to Rubin documents related to Special Programs -- motivated by past versions of this DMTN -- were made via LCR-1309 and LCR-2265.
+Updates to Rubin documents related to Special Programs that were motivated by past versions of this DMTN
+were made via LCR-1309 and LCR-2265.
 
@@
 
-Section 3.4 ``The Full Survey Specifications" states the SRD's assumption that 90\% of the total available survey time would be spent on the main survey, and that the remaining 10\% would be spent {\it ``to obtain improved coverage of parameter space ...
-[or to] observe special regions"}.
+Section 3.4 ``The Full Survey Specifications" states the SRD's assumption that 90\% of the total 
+available survey time would be spent on the main survey, and that the remaining 10\% would be spent 
+{\it ``to obtain improved coverage of parameter space ... [or to] observe special regions"}.
 
@@
 
-Note that Version 5 (2018-06-26) was an update for LCR-1309, which added requirements, specifications, and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.
+Note that Version 5 (2018-06-26) was an update for LCR-1309, which added requirements, specifications, 
+and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.
 
-In Section 1.5.1.3, ``Processing Data from Special Programs", LSR-REQ-0122\lsrreq{0122} is a requirement that the LSST system {\it ``shall deliver unique and separate data products for visits from Special Programs"} whenever possible, and that they {\it ``shall be delivered on timescales intermediate"} to the Prompt and Data Release timescales {\it ``when this enables the intended science of the Special Program"}.
-The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required"}.
+In Section 1.5.1.3, ``Processing Data from Special Programs", LSR-REQ-0122\lsrreq{0122} is a requirement 
+that the LSST system {\it ``shall deliver unique and separate data products for visits from Special Programs"} 
+whenever possible, and that they {\it ``shall be delivered on timescales intermediate"} to the Prompt and 
+Data Release timescales {\it ``when this enables the intended science of the Special Program"}.
+The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System 
+can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new 
+algorithms, or the allocation of significant additional computational resources, are required"}.
 
-In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\lsrreq{0111} requires that the LSST system {\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of"} 1 second ({\tt minExpTime}\reqparam{minExpTime}).
-The discussion notes that {\it ``non-standard visit exposures may possibly be degraded in some aspects of performance (e.g.
-cosmic ray rejection on visits consisting of a single exposure), and might be incompatible with difference imaging and alert production (e.g., short exposures in which the PSF is
-not fully formed)"}.
+In Section 2.4.1.1.2, ``Non-Standard Visit", LSR-REQ-0111\lsrreq{0111} requires that the LSST system 
+{\it ``be capable of obtaining and processing exposures not taken in a standard visit mode including those with a 
+minimum exposure time of"} 1 second ({\tt minExpTime}\reqparam{minExpTime}).
+The discussion notes that {\it ``non-standard visit exposures may possibly be degraded in some aspects of performance 
+(e.g. cosmic ray rejection on visits consisting of a single exposure), and might be incompatible with difference 
+imaging and alert production (e.g., short exposures in which the PSF is not fully formed)"}.
 
-The above requirement indicates that the LSST system shall be able to process non-standard visits from Special Programs, but that the image quality might be degraded.
-Improvements that require algorithms or processing outside of what the LSST Science Pipelines can provide would be left to the science community and require User-Generated pipelines and data products.
+The requirement in Section 1.5.1.3 is echoed in Section 2.6.1.1, ``Organization of Data Products", in which 
+LSR-REQ-0032\lsrreq{0032} is a requirement that the data processing system provide the means for three 'classes' of 
+data products on different timescales (Prompt, Data Release, and User-Generated), and also to provide a means for 
+processing Special Programs data because the {\it ``science goals of Special Programs may require that their processed 
+data products be made available in an additional fourth class, and possibly with intermediate timescales"}.
 
-The requirement in Section 1.5.1.3 is echoed in Section 2.6.1.1, ``Organization of Data Products", in which LSR-REQ-0032\lsrreq{0032} is a requirement that the data processing system provide the means for three 'classes' of data products on different timescales (Prompt, Data Release, and User-Generated), and also to provide a means for processing Special Programs data because the {\it ``science goals of Special Programs may require that their processed data products be made available in an additional fourth class, and possibly with intermediate timescales"}.
-
-In Section 2.6.1.1.3, ``Level 3 Data Products", LSR-REQ-0041\lsrreq{0041} specifies that the LSST system {\it ``shall support"} User-Generated data products.
-The discussion clarifies that ``{\it there will be technical limits on DM's ability to meet this requirement, such as cases where an intensive amount of additional computational resources is required, because only ~10\% of the total computational system is allocated for user processing"}.
+In Section 2.6.1.1.3, ``Level 3 Data Products", LSR-REQ-0041\lsrreq{0041} specifies that the LSST system 
+{\it ``shall support"} User-Generated data products.
+The discussion clarifies that ``{\it there will be technical limits on DM's ability to meet this requirement, such as 
+cases where an intensive amount of additional computational resources is required, because only ~10\% of the total 
+computational system is allocated for user processing"}.
 This level of support applies also to user processing of Special Programs data.
@@
 
-Section 2.6.1.1.4, ``Data Products for Special Programs", LSR-REQ-0121\lsrreq{0121} specifies that the LSST system {\it ``shall produce unique and separate Data Products as the result of processing data from Special Programs whenever possible, on a timescale that enables the intended science goals of the Special Program.
-The cumulative size of the online Special Programs data products shall be no more than ~10\% of the size of the DRP data products from the most recent data
-release"}.
-The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required.
-The cumulative size of the Special Programs data products is capped at ~10\% of the most recent DR because this matches the expected fractional survey area of Special Programs compared to the main survey"}.
+Section 2.6.1.1.4, ``Data Products for Special Programs", LSR-REQ-0121\lsrreq{0121} specifies that the LSST system 
+{\it ``shall produce unique and separate Data Products as the result of processing data from Special Programs whenever 
+possible, on a timescale that enables the intended science goals of the Special Program.
+The cumulative size of the online Special Programs data products shall be no more than ~10\% of the size of the DRP 
+data products from the most recent data release"}.
+The discussion clarifies that {\it ``the term 'whenever possible' includes cases where the Data Management System can 
+run original or reconfigured versions of existing pipelines, and excludes cases where the development of new 
+algorithms, or the allocation of significant additional computational resources, are required.
+The cumulative size of the Special Programs data products is capped at ~10\% of the most recent DR because this 
+matches the expected fractional survey area of Special Programs compared to the main survey"}.
 
-In Section 2.7.1.6, ``Community Computing Services", LSR-REQ-0055\lsrreq{0055} requires that the LSST system {\it ``shall provide and maintain an amount of computing capacity equivalent to at least"} 10\% ({\tt userComputingFraction}\reqparam{userComputingFraction}) {\it ``of the total LSST data processing capacity (computing and storage) for the purpose of scientific analysis of LSST data and the production of"} User-Generated data products.
+In Section 2.7.1.6, ``Community Computing Services", LSR-REQ-0055\lsrreq{0055} requires that the LSST system 
+{\it ``shall provide and maintain an amount of computing capacity equivalent to at least"} 10\% 
+({\tt userComputingFraction}\reqparam{userComputingFraction}) {\it ``of the total LSST data processing capacity 
+(computing and storage) for the purpose of scientific analysis of LSST data and the production of"} User-Generated 
+data products. 
 The discussion clarifies that the scope of this service remains to be determined.
@@
 
-In Section 3.1.3.1, ``Survey Time Allocation", LSR-REQ-0075\lsrreq{0075} requires that the {\it ``survey performance requirements shall be met utilizing approximately 90\% of the historically available observing time, leaving the remaining time available for yet to be defined special programs"}.
-
-In Section 1.3.1.3, LSR-REQ-0124\lsrreq{0124}, the discussion specifies that an image quality parameter related to ellipticity applies only to main survey data.
+In Section 3.1.3.1, ``Survey Time Allocation", LSR-REQ-0075\lsrreq{0075} requires that the {\it ``survey performance 
+requirements shall be met utilizing approximately 90\% of the historically available observing time, leaving the 
+remaining time available for yet to be defined special programs"}.
 
@@
 
-Version 19.1 (revision 2021-07-30), \citeds{LSE-30}.
+Version 19.3 (revision 2022-12-12), \citeds{LSE-30}.
 
@@
 
-In Section 2.2.3.1, ``Standard Operating States", OSS-REQ-0044\ossreq{0044} specifies that {\it ``the LSST observatory system shall be designed and constructed to support ...
-manual observing - used for specific non-scheduler driven observing to support system verification and testing or specialized science programs"}.
-Although most Special Programs will be executed via the survey scheduler as part of {\it ``fully automated observing"}, manual observing might be necessary for, e.g., target-of-opportunity Special Programs.
+In Section 2.2.3.1, ``Standard Operating States", OSS-REQ-0044\ossreq{0044} specifies that {\it ``the LSST 
+observatory system shall be designed and constructed to support ... manual observing - used for specific 
+non-scheduler driven observing to support system verification and testing or specialized science programs"}. 
+Although most Special Programs will be executed via the survey scheduler as part of {\it ``fully automated 
+observing"}, manual observing might be necessary for, e.g., target-of-opportunity Special Programs.
 
-Section 3.1.5.1.2, ``Data Products Handling for Special Programs", OSS-REQ-0392\ossreq{0392} is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121), and specifies that {\it ``the handling of data products from Special Programs shall be compliant with the approach defined in LSE-163"}.
+Section 3.1.5.1.2, ``Data Products Handling for Special Programs", OSS-REQ-0392\ossreq{0392} is a flow-down of 
+requirements from the LSR (0122, 0075, and 0121; LSE-29), and specifies that {\it ``the handling of data products 
+from Special Programs shall be compliant with the approach defined in LSE-163"}.
 
-In Section 3.2.5.3, OSS-REQ-0403\ossreq{0403} is a flow-down of LSR-REQ-0124\lsrreq{0124} related to the ellipticity correlation function distribution.
+In Section 3.6.1.3, ``Continuous Exposures", OSS-REQ-0319\ossreq{0319} requires that {\it ``The Observatory shall be 
+capable of continuous operation throughout a night with the interval between successive visits equal to the FPA 
+readout time"}.
+The discussion clarifies that {\it ``this mode of observing is needed to support observations when the telescope is 
+not being re-pointed. For example observing ``deep drilling" fields..."}.
 
-In Section 3.6.1.3, ``Continuous Exposures", OSS-REQ-0319\ossreq{0319} requires that {\it ``The Observatory shall be capable of continuous operation throughout a night with the interval between successive visits equal to the FPA readout time"}.
-The discussion clarifies that {\it ``this mode of observing is needed to support observations when the telescope is not being re-pointed"}, such as deep drilling fields or other Special Programs.
+In Section 3.6.1.4, ``Minimum Exposure Time", OSS-REQ-0291\ossreq{0291} specifies that {\it ``the camera shall be 
+able to obtain a single exposure with an effective minimum exposure time of no more than"} 1 second 
+({\tt minExpTime}\reqparam{minExpTime}) {\it ``with a goal of an effective minimum exposure time of"} 0.1 seconds 
+({\tt minExpTimeGoal}\reqparam{minExpTimeGoal}). 
+The discussion clarifies that {\it ``if the exposure is shortened from the 15 second nominal, the spacing between 
+successive exposures should be extended to maintain the average readout rate consistent with a 15 second exposure"}, 
+which may increase the overheads of Special Programs using short exposure times.
+The discussion also clarifies that {\it ``if the exposure is lengthened from the 15 second nominal, the thermal 
+stability may be affected, which may affect photometric accuracy"}.
 
-In Section 3.6.1.4, ``Minimum Exposure Time", OSS-REQ-0291\ossreq{0291} specifies that {\it ``the camera shall be able to obtain a single exposure with an effective minimum exposure time of no more than"} 1 second ({\tt minExpTime}\reqparam{minExpTime}) {\it ``with a goal of an effective minimum exposure time of"} 0.1 seconds ({\tt minExpTimeGoal}\reqparam{minExpTimeGoal}).
-The discussion clarifies that {\it ``if the exposure is shortened from the 15 second nominal, the spacing between successive exposures should be extended to maintain the average readout rate consistent with a 15 second exposure"}, which may increase the overheads of Special Programs using short exposure times.
-The discussion also clarifies that {\it ``if the exposure is lengthened from the 15 second nominal, the thermal stability may be affected, which may affect photometric accuracy"}.
+In Section 3.6.1.5, ``Publish Visit Type", OSS-REQ-0384\ossreq{0384} specifies that {\it ``the OCS [Observatory 
+Control System] shall configure the [Data Management System] DMS (in particular Prompt Processing) with the type 
+of visits to be processed: Standard, Alternate, or a specific type of Non-Standard"}.
+The discussion clarifies that this allows the Prompt processing pipeline to be reconfigured on-the-fly in order to 
+incorporate non-standard visits from, e.g., Special Programs.
+The time required for reconfiguration might introduce some latency or cause some images to not be processed by the 
+Prompt pipeline.
 
-In Section 3.6.1.5, ``Publish Visit Type", OSS-REQ-0384\ossreq{0384} specifies that {\it ``the OCS [Observatory Control System] shall configure the [Data Management System] DMS (in particular Prompt Processing) with the type of visits to be processed: Standard, Alternate, or a specific type of Non-Standard"}.
-The discussion clarifies that this allows the Prompt processing pipeline to be reconfigured on-the-fly in order to incorporate non-standard visits from, e.g., Special Programs.
-The time required for reconfiguration might introduce some latency or cause some images to not be processed by the Prompt pipeline.
+In Section 3.6.2.1.2, ``Maximum time for operational filter change", OSS-REQ-0293\ossreq{0293} specifies that 
+{\it ``the camera system shall provide the capability of changing the operational filter with any other internal 
+filter in a time less than"} 120 seconds ({\tt tFilterChange}\reqparam{tFilterChange}).
+This would impose a large overhead on, e.g., a Special Program that changes filters often without slewing.
+See also OSS-REQ-0295\ossreq{0295}, Appendix~\ref{sec:hardbounds} of this document, and/or the filter change 
+memorandum (\url{ls.st/spt-494}), for more information about the total lifetime number of filter changes.
 
-In Section 3.6.2.1.2, ``Maximum time for operational filter change", OSS-REQ-0293\ossreq{0293} specifies that {\it ``the camera system shall provide the capability of changing the operational filter with any other internal filter in a time less than"} 120 seconds ({\tt tFilterChange}\reqparam{tFilterChange}).
-This would impose a large overhead on, e.g., a Special Program that changes filters often without slewing.
-See also OSS-REQ-0295\ossreq{0295}, Appendix~\ref{sec:hardbounds} of this document, and/or the filter change memorandum (\url{ls.st/spt-494}), for more information about the total lifetime number of filter changes.
+In Section 3.6.3.1, ``Absolute Pointing", OSS-REQ-0298\ossreq{0298} specifies that {\it ``the LSST shall point to 
+a defined set of sky coordinates with an RMS accuracy of"} 2 arcseconds ({\tt absPointErr}\reqparam{absPointErr}).
 
-In Section 3.6.3.1, ``Absolute Pointing", OSS-REQ-0298\ossreq{0298} specifies that {\it ``the LSST shall point to a defined set of sky coordinates with an RMS accuracy
-of"} 2 arcseconds ({\tt absPointErr}\reqparam{absPointErr}).
+In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able 
+to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
+The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on 
+a single field"}.
 
-In Section 3.6.3.3, ``Rotator tracking Time, OSS-REQ-0301\ossreq{0301} specifies that {\it ``the LSST shall be able to maintain field rotation tracking over a period of at least"} 1 hour ({\tt rotTrackTime}\reqparam{rotTrackTime}).
-The discussion clarifies that this {\it ``is driven by the need to conduct extended 'deep drilling' observations on a single field"}.
-There do not seem to be any constraints on the speed of the rotator or the minimum distance between successive visits.
+In Section 3.6.3.5, ``Offset Pointing", OSS-REQ-0302\ossreq{0302} specifies that {\it ``the LSST shall be capable of 
+offset pointing within a single field-of-view with a precision of no more than"} 0.2 arcseconds 
+({\tt offsetPointingErr}\reqparam{offsetPointingErr}).
 
-In Section 3.6.3.5, ``Offset Pointing", OSS-REQ-0302\ossreq{0302} specifies that {\it ``the LSST shall be capable of offset pointing within a single field-of-view with a precision of no more than"} 0.2 arcseconds ({\tt offsetPointingErr}\reqparam{offsetPointingErr}).
-
-In Section 3.6.3.6, ``Open Loop Tracking", OSS-REQ-0303\ossreq{0303} specifies that {\it ``The LSST shall be capable of open loop tracking without the assistance of real time optical feedback to an accuracy of"} 1.0 arcseconds ({\tt openTrackErr}\reqparam{openTrackErr}) {\it ``over any 10 minute duration during normal night time operations"}.
+In Section 3.6.3.6, ``Open Loop Tracking", OSS-REQ-0303\ossreq{0303} specifies that {\it ``The LSST shall be 
+capable of open loop tracking without the assistance of real time optical feedback to an accuracy of"} 1.0 arcseconds 
+({\tt openTrackErr}\reqparam{openTrackErr}) {\it ``over any 10 minute duration during normal night time operations"}.
 Note that the open loop tracking requirement is \emph{without guiding}.
 
-In Section 3.6.3.8, ``Guiding", OSS-REQ-0305\ossreq{0305} specifies that {\it ``the LSST system shall improve the tracking of the optical system to the sky by means of active guiding over the whole accessible sky"}.
-
-In Section 3.6.3.10, ``Non-Sidereal Tracking", OSS-REQ-0380\ossreq{0380} specifies that {\it ``the LSST system shall be capable of tracking in an arbitrary direction on the sky along a parametric RA(t) and DEC(t) trajectory, at angular rates of up to"} 220 arcseconds per second ({\tt nonsiderealAngularRateEl}\reqparam{nonsiderealAngularRateEl} and {\tt nonsiderealAngularRateAZ}\reqparam{nonsiderealAngularRateAZ}) {\it ``with a tracking error not to exceed"} 0.5 arcseconds per minute ({\tt nonsiderealTrackingError}\reqparam{nonsiderealTrackingError}).
-The discussion notes that {\it ``this is standard capability for modern telescopes"}, but might be relevant to some Special Programs.
-
-%%% MLG: this does not appear in OSS V13
-% $\bullet$ OSS-REQ-0027 requires that the scheduling system be able to optimize over at least \texttt{nSciProp} = 6 "science proposals", where these "proposals" are observing targets/constraints such as the distribution of filters, the astronomical conditions, and relative priority (OSS 2.1.1.2, Multiple Science Programs).
-{\bf JIRA ticket DM-12579 confirmed that there is no maximum number, and so many Special Programs will be able to be included in the scheduler}.
-
-%%% MLG: this does not appear in OSS V13
-% $\bullet$ OSS-REQ-0381 requires that the schedule be able to handle targets of opportunity, which would be relevant for e.g., Special Programs for gravitational wave follow-up (OSS 2.1.1.7, Visit Optimization).
-
-%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
-% $\bullet$ OSS-REQ-0189 and OSS-REQ-0190 set the minimum number of raw exposures to be supported as \texttt{nRawExpNightWinterAvg} = 1960 per night on average (but up to \texttt{nRawExpNightMax} = 2800 per night if e.g., two hours of a short-exposure twilight mini-survey are included) and \texttt{nRawExpYear} = 5.5$\times10^5$ per year, respectively.
-These numbers are set by predicting the maximum number of exposures that would be acquired on the longest night of the year in WFD cadence with 2 second slews, assuming $\sim80\%$ completion, but adding a 10\% margin.
-These estimates appear adequate for Special Programs in general.
-
-%%% MLG: these appear in OSS V13, but aren't as relevant to DMTN-065 anymore
-% $\bullet$ OSS-REQ-0194 and OSS-REQ-0323 set the minimum number of calibration exposures to be supported as \texttt{nCalibExpDay} = 450 on average and \texttt{nCalExpYear} = 1.5$\times10^5$ per year, respectively.
-These are \textit{minimums}, and so if a Special Program requires additional exposures, this should be possible to accommodate.
+In Section 3.6.3.10, ``Non-Sidereal Tracking", OSS-REQ-0380\ossreq{0380} specifies that {\it ``the LSST system shall 
+be capable of tracking in an arbitrary direction on the sky along a parametric RA(t) and DEC(t) trajectory, at 
+angular rates of up to"} 220 arcseconds per second ({\tt nonsiderealAngularRateEl}\reqparam{nonsiderealAngularRateEl} 
+and {\tt nonsiderealAngularRateAZ}\reqparam{nonsiderealAngularRateAZ}) {\it ``with a tracking error not to exceed"} 
+0.5 arcseconds per minute ({\tt nonsiderealTrackingError}\reqparam{nonsiderealTrackingError}).
+The discussion notes that {\it ``this is standard capability for modern telescopes"}, but might be relevant to some 
+Special Programs.
 
@@
 
-Note that Version 8.3 (2020-05-04) was an update for LCR-2265, which updated requirements, specifications, and discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.
+Note that Version 8.3 (2020-05-04) was an update for LCR-2265, which updated requirements, specifications, and 
+discussions regarding the processing of Special Programs data based on earlier versions of this DMTN.
 
-In Section 1.2.3, ``Raw Science Image Metadata", DMS-REQ-0068\dmreq{0068} specifies that {\it ``for each raw science image, the DMS shall store image metadata"} including {\it ``Program metadata (identifier for main survey, deep drilling, etc.)"}.
-The discussion clarifies that {\it ``the program metadata should be sufficient to associate an image with a specific Special Program so that DMS-REQ-0320 and DMS-REQ-0397 can be satisfied"}.
+In Section 1.2.3, ``Raw Science Image Metadata", DMS-REQ-0068\dmreq{0068} specifies that {\it ``for each raw 
+science image, the DMS shall store image metadata"} including {\it ``Program metadata (identifier for main survey, 
+deep drilling, etc.)"}.
+The discussion clarifies that {\it ``the program metadata should be sufficient to associate an image with a specific 
+Special Program so that DMS-REQ-0320 and DMS-REQ-0397 can be satisfied"}.
 
-In Section 1.3.13, ``Alert Content", the discussion for DMS-REQ-0274\dmreq{0274} explains that the {\it ``program and/or scheduler metadata"} included in an alert packet {\it ``should be sufficient to identify whether the image is associated with a Special Program (such as an in-progress Deep Drilling Field)"}.
+In Section 1.3.13, ``Alert Content", the discussion for DMS-REQ-0274\dmreq{0274} explains that the {\it ``program 
+and/or scheduler metadata"} included in an alert packet {\it ``should be sufficient to identify whether the image is 
+associated with a Special Program (such as an in-progress Deep Drilling Field)"}.
 
-In Section 1.4.18.1, ``Produce All-Sky HiPS Map", the discussion for DMS-REQ-0379\dmreq{0379} raises the point that generating separate HiPS maps for Special Programs (e.g., DDFs) remains an open question.
+In Section 1.4.18.1, ``Produce All-Sky HiPS Map", the discussion for DMS-REQ-0379\dmreq{0379} raises the point that 
+generating separate HiPS maps for Special Programs (e.g., DDFs) remains an open question.
 
-In Section 1.4.18.5, ``Produce MOC Maps", DMS-REQ-0383\dmreq{0383} specifies that Data Release processing {\it ``shall include the production of Multi-Order Coverage maps for the survey data"}, and that {\it ``additional MOCs SHOULD be produced to represent special-programs datasets"}.
+In Section 1.4.18.5, ``Produce MOC Maps", DMS-REQ-0383\dmreq{0383} specifies that Data Release processing 
+{\it ``shall include the production of Multi-Order Coverage maps for the survey data"}, and that {\it ``additional 
+MOCs SHOULD be produced to represent special-programs datasets"}.
 It is noted that a separate technical note would be created to define these MOCs.
@@
 
-In Section 1.6.1, ``Processing of Data From Special Programs", DMS-REQ-0320\dmreq{0320} specifies that {\it ``it shall be possible for special programs to trigger their own data processing recipes, during the night instead of the nightly Alert Processing (but the recipes may still issue Alerts), or on alternative timescales"}.
-The discussion clarifies that the {\it ``LSST will provide these recipes ... when
-possible, which includes cases where DM can run original or reconfigured versions of existing pipelines, and excludes cases where the development of new algorithms, or the allocation of significant additional computational resources, are required.
-An example of an alternative timescale is a nightly trigger to coadd all the deep-drilling field images.
-Decisions about which recipes are applied to which Special Programs will be made by the Operations team, after consideration of the scientific goals, computational resources, and data rights policy"}.
-This requirement is derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).
+In Section 1.6.1, ``Processing of Data From Special Programs", DMS-REQ-0320\dmreq{0320} specifies that {\it ``it 
+shall be possible for special programs to trigger their own data processing recipes, during the night instead of 
+the nightly Alert Processing (but the recipes may still issue Alerts), or on alternative timescales"}.
+The discussion clarifies that the {\it ``LSST will provide these recipes ... when possible, which includes cases 
+where DM can run original or reconfigured versions of existing pipelines, and excludes cases where the development 
+of new algorithms, or the allocation of significant additional computational resources, are required. An example of 
+an alternative timescale is a nightly trigger to coadd all the deep-drilling field images. Decisions about which 
+recipes are applied to which Special Programs will be made by the Operations team, after consideration of the 
+scientific goals, computational resources, and data rights policy"}.
+This requirement is derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 
+0075, and 0121).
 
-In Section 1.6.2, ``Prompt/DR Processing of Data from Special Programs", DMS-REQ-0397\dmreq{0397} specifies that {\it ``it shall be possible for special programs data to be processed with the prompt and/or annual-release pipelines alongside data from the main survey"}.
-The discussion further clarifies that {\it ``the data from Special Programs should only be included ...
-when it is (a) possible ...
-to do so without additional effort and (b) beneficial to the LSST's main science objectives.
-Decisions about which data are included ...
-will be made by the Operations team"}.
-This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).
+In Section 1.6.2, ``Prompt/DR Processing of Data from Special Programs", DMS-REQ-0397\dmreq{0397} specifies that 
+{\it ``it shall be possible for special programs data to be processed with the prompt and/or annual-release pipelines 
+alongside data from the main survey"}.
+The discussion further clarifies that {\it ``the data from Special Programs should only be included ... when it is 
+(a) possible ... to do so without additional effort and (b) beneficial to the LSST's main science objectives. 
+Decisions about which data are included ... will be made by the Operations team"}.
+This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR 
+(0122, 0075, and 0121).
 
-In Section 1.6.3, ``Level 1 Processing of Special Programs Data", DMS-REQ-0321\dmreq{0321} specifies that {\it ``all [Prompt] processing from special programs shall be completed before data arrives from the following night's observations"}.
-This is essentially adding a quantifier to DMS-REQ-0397, to specify that {\it ``when it is (a) possible ...
-to do so"} means when it is possible to complete the processing before the next night's observations.
-This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).
+In Section 1.6.3, ``Level 1 Processing of Special Programs Data", DMS-REQ-0321\dmreq{0321} specifies that {\it ``all 
+[Prompt] processing from special programs shall be completed before data arrives from the following night's 
+observations"}.
+This is essentially adding a quantifier to DMS-REQ-0397, to specify that {\it ``when it is (a) possible ... to do so"} 
+means when it is possible to complete the processing before the next night's observations.
+This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR 
+(0122, 0075, and 0121).
 
-In Section 1.6.4, ``Constraints on Level 1 Special Program Products Generation", DMS-REQ-0344\dmreq{0344} specifies that {\it ``the publishing of [Prompt] data products from Special Programs shall be subject to the same performance requirements of"} 24 hours ({\tt L1PublicT}\reqparam{L1PublicT}) for the release of Prompt data products and 1 minute ({\tt OTT1}\reqparam{OTT1}) for the transmission of Alert packets.
+In Section 1.6.4, ``Constraints on Level 1 Special Program Products Generation", DMS-REQ-0344\dmreq{0344} specifies 
+that {\it ``the publishing of [Prompt] data products from Special Programs shall be subject to the same performance 
+requirements of"} 24 hours ({\tt L1PublicT}\reqparam{L1PublicT}) for the release of Prompt data products and 1 minute 
+({\tt OTT1}\reqparam{OTT1}) for the transmission of Alert packets.
 This is essentially a more detailed version of DMS-REQ-0321 which includes the Alert production timescale.
-This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirents from the LSR (0122, 0075, and 0121).
+This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirents from the LSR 
+(0122, 0075, and 0121).
 
-In Section 1.6.5, ``Special Programs Database", DMS-REQ-0322\dmreq{0322} specifies that {\it ``data products for special programs shall be stored in databases that are distinct from those used to store standard [Prompt] and [Data Release] data products"} and that {\it ``it shall be possible for these databases to be federated ...
-to allow cross-queries and joins"}.
-This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR (0122, 0075, and 0121).
+In Section 1.6.5, ``Special Programs Database", DMS-REQ-0322\dmreq{0322} specifies that {\it ``data products for 
+special programs shall be stored in databases that are distinct from those used to store standard [Prompt] and 
+[Data Release] data products"} and that {\it ``it shall be possible for these databases to be federated ... to allow 
+cross-queries and joins"}.
+This requirement is also derived from OSS-REQ-0392, which is essentially a flow-down of requirements from the LSR 
+(0122, 0075, and 0121).
 
-In Section 4.1.16, ``Level 2 and Reprocessed Level 1 Catalog Access", DMS-REQ-0313\dmreq{0313} specifies that {\it ``the DMS shall maintain ... versions
-of the most recent catalogs generated from Special Programs data"}.
+In Section 4.1.16, ``Level 2 and Reprocessed Level 1 Catalog Access", DMS-REQ-0313\dmreq{0313} specifies that 
+{\it ``the DMS shall maintain ... versions of the most recent catalogs generated from Special Programs data"}. 
 As with all LSST data, {\it ``there is no requirement for older data releases to be queryable"}.
@@
 The DMAD is not a requirements document.
-Instead, it describes the scientific design of the LSST Science Pipelines: the algorithms and software that will be implemented to meet the requirements for processing the LSST data.
+Instead, it describes the scientific design of the LSST Science Pipelines: the algorithms and software that will 
+be implemented to meet the requirements for processing the LSST data.
 
-Special Programs are only mentioned a few times, either as a potential source of single-snap visits or as a potential source of reference images or catalogs (e.g., training sets).
+Special Programs are only mentioned a few times, either as a potential source of single-snap visits or as a 
+potential source of reference images or catalogs (e.g., training sets).
 
-As described above (e.g., LSR-REQ-0122), the LSST system shall deliver unique and separate data products for visits from Special Programs whenever this (1) enables the intended science of the Special Program and (2) can be accomplished using the original or reconfigured versions of the LSST Science Pipelines.
-For cases in which the development of new algorithms or the allocation of significant additional computational resources are required to produce Special Programs data products, User-Generated pipelines and processing will be necessary.
+As described above (e.g., LSR-REQ-0122), the LSST system shall deliver unique and separate data products for visits 
+from Special Programs whenever this (1) enables the intended science of the Special Program and (2) can be 
+accomplished using the original or reconfigured versions of the LSST Science Pipelines.
+For cases in which the development of new algorithms or the allocation of significant additional computational 
+resources are required to produce Special Programs data products, User-Generated pipelines and processing will be 
+necessary.
 
 The DMAD can be used as the reference document to decide whether a given Special Program will require User-Generated pipelines and processing.
-
-%%% MLG: the following are all from <2018 and were not updated in 2022
-
-% \subsubsection{Extreme-Depth CoAdds} The system has been sized to hold $\sim200$ exposures in memory at once, which defined by the current maximum number of visits per field in the WFD main survey in $10$ years (from a conversation with K.-T.~Lim).
-Note that the panchromatic CoAdds would be built from the individual filter CoAdds, so the algorithm does not need to handle $\sim800$ images.
-From a computational standpoint, $200$ is the maximum number of images that can be stacked with an algorithm that requires all images to be accessible in memory at once (i.e., loading all images and calculating the median for each pixel).
-Deeper stacks might be possible with algorithms that deal with images consequentially.
-It is conceivable that a Special Program which needs to stack $>200$ images is not possible to accomplish with reconfigured pipelines, and would have to be processed with external, user-contributed resources.
-However, the exact DM capabilities in this area are not yet well known because NCSA has not yet defined the machine capabilities.
-Furthermore, the planned commissioning data will go to a $\sim20$ year depth, and so it can reasonably be expected that DM will have to be able to accommodate at least a stack that deep.
-
-% \subsubsection{Deblending} The deep deblender algorithm described in Section 5.3.3 will, out of necessity, be optimized for use in the bulk of the WFD main survey.
-It may or may not end up being appropriate for use in the Galactic Plane mini-survey area, depending on the science goal.
-Level 3 deblenders for specific Special Programs fields may require development by the user community.
-
-% \subsubsection{Variability Characterization} The periodic and aperiodic variability characterizations described in Section 6.21 of the DMAD are placeholders, but are representative of what is likely to be implemented: algorithms that are applicable to a broad range of variability types.
-From DM's perspective, all that is needed is sufficient information to enable relatively useful filters, from which the downstream broker/user can do additional filtering, and these parameterizations might not be sufficient for all science goals.
-It is conceivable that the goals of a particular Special Program might require different algorithms; these could be provided by DM, or written as Level 3 and either made joinable to the DM reconfigured data products or perhaps incorporated directly.
-
-% \subsubsection{Photometric Redshifts} As described in Section 5.6.5 the Level 2 DRP \texttt{Object} catalog will include a photometric redshift, but this algorithm will be produced by the science community and then adopted and run at scale by DM.
-It is conceivable that the photo-$z$ algorithm for a Special Programs data set, such as a deep drilling field, might be different from that used for the WFD main survey.
 
@@
 Section 6 describes the data products for Special Programs.
-The DPDD is not a requirements document; Section 6 summarizes the requirements presented above and does not introduce any new constraints or new information about Special Programs.
-
-%%% MLG commented out the paragraphs below in 2022, they were out of date.
-
-% $\bullet$ The database schema for \texttt{DIASource} does not appear to have an element that identifies which template image that was used.
-This will be needed for both Levels 1 and 3 differencing pipelines and products, for both Special Programs and WFD main survey data.
-DMS-REQ-0074 already does require that the identity of input exposures is stored for each difference image (LSE-61).
-In conversation with K.-T., we find that this isn't a problem, as it will be handled by provenance: the code configuration used and the time of processing are sufficient to identify and regenerate the template image.
-However, K.-T.
-has pointed out that the capability to regenerate the \textit{exact same} template -- the pixels that were subtracted -- is not a current deliverable.
-However, the stamp of the difference image will live on in the Alerts database, so we also do not foresee this as a problem.
-
-% $\bullet$ The DMAD specifies that externally defined targets can be incorporated into the \texttt{Objects} catalog (Section 3.2.5), and this may be a particular interest to Special Programs.
-It is unclear how such targets will be identified or flagged as such in the database schema, and whether we need to add an element for this.
-Currently, the \texttt{Object} database contains an element \texttt{prv\_inputId} which is an \texttt{integer}, and is described as the \textit{"Pointer to prv\_InputType.
-Indicates which input was used to produce a given object."} Is that all we need?
-{\bf JIRA ticket DM-12580 clarified that  \texttt{Object.prv\_inputID} in the database schema is one possible way to identify whether an \texttt{Object} is an externally provided coordinate}.
-
-% $\bullet$ The \texttt{Object} and \texttt{DIAObject} elements that have been reserved for variability characterization, as described in the DMAD and the DPDD, are as follows: \\
-% \texttt{Object} and \texttt{DIAObject.lcPeriodic} = \texttt{float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-% \texttt{Object} and \texttt{DIAObject.lcNonPeriodic} = \texttt{float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
-% Section 6.21 of the DMAD describes the nominal algorithms to define these parameters, but as we mentioned in Section \ref{ssec:docrev_dmad}, different kinds of variability might be measurable Special Programs cadences that are quite different from the WFD main survey.
-Are 32 floats in each of the 6 filters always going to be a large enough volume?
-The Transients and Variable Stars Science Collaboration currently has a Task Force assigned to address this, and {\bf it is the topic of JIRA ticket DM-12581}.
-
-%%% MLG commented out this whole section in 2022, since LSE-180 is from 2013
-%%%  LSE-180 is deprecated?
-
-% \subsection{Photometric Calibration for the LSST Survey, \citeds{LSE-180}}
-
-% LSE-180 is built on \texttt{OpSim} runs that do include some nominal DDF, but the photometric calibration investigated in this work does not much deal with potential issues induced by non-standard visit patterns or exposure times of Special Programs, as its scope is the WFD main survey.
-Potential issues with DM processing -- including calibrations -- of non-standard visit exposure times is raised in Section \ref{ssec:dmplans_NSV}.
-Regarding the reduction and calibration of non-standard visit images, LSE-180 makes two relevant points: \\
-% $\bullet$ In LSE-180, it is assumed that all factors affecting the system transmission are stable on 15 second timescales (page 10), but not what the upper limit of that might be. \\
-% $\bullet$ LSE-180 comments on the dither pattern for the WFD survey in that "dither patterns where the overlap is one quarter of the field of view or more produce results meeting the SRD requirements", but this is specific to photometric calibration of the WFD.
-The LSE-180 also mentions that an inappropriate dither pattern can make it hard to correct for the variation of system bandpass as a function of the focal plane position -- but so long as this is solved in the WFD, the corrections can be applied to the much smaller amount of data from the Special Programs.
-
-% However, Lupton's recent work on calibrations has made much of LSE-180 obsolete, and it is not clear whether it is still needed as a separate document.
-At the time of writing, Lupton's most recent take on calibrations can be found in \url{https://github.com/lsst-dm/calibration/blob/master/calibration.pdf}.
-K.-T.
-referenced the Appendix B for the overview of calibration types, but it seems that there is not yet a final plan that can be assessed for its suitability for Special Programs.
-It is plausible that Special Programs could have their own calibration database, and obtaining additional calibration frames is not expected to cause a bottleneck (at the moment there is $\sim1.7$ hours per day for this, and exceeding this could eat into the engineering time).
-The most likely cause for trouble involving calibrations and Special Programs is the computational time needed to create the additional and/or different calibration files to be applied to non-standard Special Programs data, and/or the additional overhead required to load them in the Level 1 pipeline if the schedule interleaves SP-WFD-WFD-SP.
-
-% {\bf JIRA ticket DM-12582 is currently open, and aims to define the potential additional calibration needs of Special Programs data.}
+The DPDD is not a requirements document; Section 6 summarizes the requirements presented above and does not introduce 
+any new constraints or new information about Special Programs.

### 2023-05-02 ap_hardbounds.tex
Update ap_hardbounds.tex

@@
 
-The potential boundaries on the diversity of data products that could be imposed by limitations from the Rubin Observatory hardware -- camera, telescope, and/or site -- are considered.
+The potential boundaries on the diversity of data products that could be imposed by 
+limitations from the Rubin Observatory hardware (camera, telescope, and/or site) 
+are considered.
 
 \subsection{Filter Changes}
-The maximum time for filter change is 120 seconds: 30 seconds for the telescope to reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
-Assuming that most Special Programs would be designed to keep overheads $<$100\% and would be using standard 30 second visits, the filter change time indicates that it is likely that at least 4 exposures in a given filter would be obtained between filter changes, but this is not actually a hardware boundary.
-The filter change mechanism is designed to undergo a total of 100000 changes over its lifetime, and each filter is designed to support up to 30000 changes over its lifetime, where lifetime is 15 years.
-That is an average of $\sim$27 changes per day, some of which would occur in they day during calibrations (estimate, $\sim$10) and the rest at night.
-As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could support as many changes involving the 5 filters loaded in the carousel as desired, without any practical limitation"}.
+
+The maximum time for filter change is 120 seconds: 30 seconds for the telescope to 
+reorient the camera to its nominal zero angle position on the rotator, and 90 seconds to 
+the camera subsystem for executing the change (OSS-REQ-0293; \citeds{LSE-30}).
+
+Assuming that most Special Programs would be designed to keep overheads $<$100\% and would 
+be using standard 30 second visits, the filter change time indicates that it is likely 
+that at least 4 exposures in a given filter would be obtained between filter changes, but 
+this is not actually a hardware boundary.
+
+The filter change mechanism is designed to undergo a total of 100000 changes over its 
+lifetime, and each filter is designed to support up to 30000 changes over its lifetime, 
+where lifetime is 15 years.
+
+That is an average of $\sim$27 changes per day, some of which would occur in they day 
+during calibrations (estimate, $\sim$10) and the rest at night.
+
+As stated in the filter change memorandum (\url{ls.st/spt-494}), {\it ``the system could 
+support as many changes involving the 5 filters loaded in the carousel as desired, without 
+any practical limitation"}.
 
 \subsection{Filter Carousel Loads}
+
 The filter carousel can hold five of the six LSST filters at a time. 
 The system is designed to support $3000$ loads in $15$ years (\url{ls.st/spt-494}). 
-Filter loads are only done in the day, and there will never be data in more than five filters in a given night.
+Filter loads are only done in the day, and there will never be data in more than five 
+filters in a given night.
 
 \subsection{Exposure Times}
-The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds (OSS-REQ-0291; \citeds{LSE-30}).
+
+The minimum exposure time is $1$ second, with a stretch goal of $0.1$ seconds 
+(OSS-REQ-0291; \citeds{LSE-30}). 
 The maximum exposure time is not restricted.
@@
 \subsection{Readout Time}
+
 The readout time is $2$ seconds, and would be significant overhead on short exposures.
@@
 \subsection{Inter-Image Time}
-Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ seconds for thermal tolerance; i.e., the minimum readout rate might be one image every $15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
 
-As discussed in Jira ticket DM-12573, the main issue is thermal and is related to the shutter, both the motors and the brakes; an elevated Camera skin temperature would affect image quality.
-As of 2022, early tests suggest that a sustained (30 minutes) sequence that increases the heat load by large factors would not work, but further functional testing of the system once the Camera was fully assembled are needed for full characterization of the issue.
+Images with exposure times $<15$ seconds {\it might} still have to be separated by $15$ 
+seconds for thermal tolerance; i.e., the minimum readout rate might be one image every 
+$15$ seconds, regardless of exposure time (OSS-REQ-0291; \citeds{LSE-30}).
 
-This potential $15$ interval between images is also a potential hardware boundary on the potential diversity of data products.
+As discussed in Jira ticket DM-12573, the main issue is thermal and is related to the 
+shutter, both the motors and the brakes; an elevated Camera skin temperature would affect 
+image quality.
+
+As of 2022, early tests suggest that a sustained (30 minutes) sequence that increases the 
+heat load by large factors would not work, but further functional testing of the system 
+once the Camera was fully assembled are needed for full characterization of the issue.
+
+This potential $15$ interval between images is also a potential hardware boundary on the 
+potential diversity of data products.
 
 \subsection{Telescope Slew}
-As described in \citeds{Document-28382}, large slews would have considerable overheads, but there are no hardware boundaries on the size of a single slew or the accrued slew distance.
+
+As described in \citeds{Document-28382}, large slews would have considerable overheads, 
+but there are no hardware boundaries on the size of a single slew or the accrued slew 
+distance.
 
 \subsection{Pointing and Sidereal Tracking}
-The specifications for the telescope's pointing and tracking in \citeds{LSE-30} indicate that $<$0.2 arcsecond precision in field pointing (OSS-REQ-0302) and $<$1 arcsecond in open-loop tracking (OSS-REQ-0303) would not be possible, but guiding would improve the latter (OSS-REQ-0305).
-Furthermore, obtaining the \emph{exact same} alignment of the pixel grid in RA-Dec {\it ``would put demands on the camera rotator that were not planned"}\footnote{As per C.
-Claver's comments in ticket DM-12573.}.
+
+The specifications for the telescope's pointing and tracking in \citeds{LSE-30} indicate 
+that $<$0.2 arcsecond precision in field pointing (OSS-REQ-0302) and $<$1 arcsecond in 
+open-loop tracking (OSS-REQ-0303) would not be possible, but guiding would improve the 
+latter (OSS-REQ-0305).
+
+Furthermore, obtaining the \emph{exact same} alignment of the pixel grid in RA-Dec 
+{\it ``would put demands on the camera rotator that were not planned"}\footnote{As per 
+C. Claver's comments in ticket DM-12573.}.
 
 \subsection{Non-Sidereal Tracking}
-The requirement that the LSST system be able to perform non-sidereal tracking is set by OSS-REQ-0380 in \citeds{LSE-30}.
-This capability will include angular rates of up to 220 arcseconds per second in both azimuth and elevation.
+
+The requirement that the LSST system be able to perform non-sidereal tracking is set by 
+OSS-REQ-0380 in \citeds{LSE-30}. 
+This capability will include angular rates of up to 220 arcseconds per second in both 
+azimuth and elevation.
 
 \subsection{Camera Rotation}
-The requirements on the rotator's capabilities do not set any limits on the per-night or total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries on the distance between successive visits or the ability to jump between two widely separated fields.
+
+The requirements on the rotator's capabilities do not set any limits on the per-night or 
+total lifetime rotation (OSS-REQ-0301, -0300; \citeds{LSE-30}) which might put boundaries 
+on the distance between successive visits or the ability to jump between two widely 
+separated fields.
+
 Currently, there are no hardware boundaries imposed by camera rotation constraints on the potential diversity of data products.

### 2023-05-02 ap_procbounds.tex
Update ap_procbounds.tex

@@
 
-Note that processing boundaries might ultimately be defined not by what is technically possible, but by the resulting image quality parameters (e.g., the number of stars with sufficient flux for photometric calibration).
-Furthermore, the processing boundaries might not be fully constrained until the final performance of the LSST Science Pipelines, as described in the Data Management Applications Design, \citeds{LDM-151}) document, is fully characterized.
+Note that processing boundaries might ultimately be defined not by what is technically 
+possible, but by the resulting image quality parameters, e.g., the number of stars with 
+sufficient flux for photometric calibration.
 
-{\bf Summary of the conclusions below:}\\
-Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
-It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
-The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.
+Furthermore, the processing boundaries might not be fully constrained until the final 
+performance of the LSST Science Pipelines, as described in the Data Management 
+Applications Design, \citeds{LDM-151}) document, is fully characterized.
+
+{\bf Summary.}\\
+
+\begin{itemize}
+\item \bf{Short or long exposures.} Very short ($<$2 sec) exposures could be difficult to 
+process due to an incompletely-formed PSF. Short (or long, $>$150 sec) exposures 
+could be difficult to calibrate due to having too few (or too few unsaturated) stars.
+\item \bf{Bright backgrounds (twilight).} It is currently unclear whether images with very 
+bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, 
+or whether user generated pipelines will be needed.
+\item \bf{Non-sidereal tracking.} The full reduction and calibration of images obtained 
+with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope 
+of the LSST Science Pipelines, and will require a user generated pipeline.
+\end{itemize}
 
@@
 
-Images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in the correction for differential chromatic refraction, in the difference imaging analysis pipeline, and/or in the photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD.
+Images which deviate significantly from the $15$ second duration for the WFD main survey 
+may encounter issues in the instrument signature removal routine, in the correction for 
+differential chromatic refraction, in the difference imaging analysis pipeline, and/or in 
+the photometric and astrometric calibrations due to a differently sampled set of standard 
+stars per CCD.
 
 \subsubsection{Short Exposures (Non-Standard Visits of $<$30 sec)}
-The LSST System Requirements document states that {\it ``The LSST shall be capable of obtaining and processing exposures not taken in a standard visit mode including those with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 seconds; LSR-REQ-0111 in \citeds{LSE-29}).
 
-However, for exposure times there are other considerations, as changing the exposure time also affects the photometric and astrometric calibrations.
-Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r$$\approx$ 13--21 magnitudes.
-A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but this still leaves stars between 15.8--21.0 magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap).
+The LSST System Requirements document states that {\it ``The LSST shall be capable of 
+obtaining and processing exposures not taken in a standard visit mode including those 
+with a minimum exposure time of} {\tt minExpTime}", which is 1 second (stretch goal 0.1 
+seconds; LSR-REQ-0111 in \citeds{LSE-29}).
 
-In order for an image to be successfully PSF-matched to the template, the PSF must be well formed (no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (be smoothly varying on some minimal scale).
-As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed").
+However, for exposure times there are other considerations, as changing the exposure time 
+also affects the photometric and astrometric calibrations.
+Assuming that 1 second exposure can be reduced and calibrated, its detected point sources 
+will span a dynamic range of $r$$\approx$ 13 to 21 magnitudes.
+A template image built on 15 second exposures will saturate at $r$$\approx$15.8 mag, but 
+this still leaves stars between 15.8 and 21.0 magnitudes to be used in the PSF-matching 
+(and all other filters have a similarly large overlap).
+
+In order for an image to be successfully PSF-matched to the template, the PSF must be 
+well formed (no speckle pattern), and have a spatial variance that the pipeline is 
+capable of modeling (be smoothly varying on some minimal scale).
+As a simple demonstration, Figure \ref{fig:expt} shows that perhaps exposure times 
+shorter than $2$ seconds do not have a well-formed PSF (using the centroid of a 2D 
+Gaussian fit as a proxy for "well-formed").
 
@@
 \includegraphics[width=14cm,trim={0cm 0cm 0cm 0cm}, clip]{figures/exptime.png}
-\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin.
-At right, blue and purple lines show the location of the centroid derived from a 2D Gaussian fit to the PSF as a function of exposure time, with the red dashed line showing the true center.
-We can see that for exposure times greater than 2 seconds, the centroid converges near its true value.
-\label{fig:expt}}
+\caption{At left, Arroyo atmosphere-only simulated PSF for LSST (with oversampled pixels) 
+with exposure times of 0.5, 2, and 15 seconds (top to bottom), courtesy of Bo Xin. At 
+right, blue and purple lines show the location of the centroid derived from a 2D Gaussian 
+fit to the PSF as a function of exposure time, with the red dashed line showing the true 
+center. We can see that for exposure times greater than 2 seconds, the centroid converges 
+near its true value. \label{fig:expt}}
 \end{center}
 \end{figure}
-
-% In conversation with DM-AP team members (Reiss, Findeisen, Connolly, Bo) there has not yet been a study of the safe range of exposure times that will be allowed to contribute Alert Production.
-% One possibly useful study is Chang et al. (2012), "Atmospheric point spread function interpolation for weak lensing in short exposure imaging data".
-% They show that a 15 second exposure contains PSF variability on short spatial scales across a 1 square degree image which, for extragalactic fields with few stars (i.e., but good for weak lensing), is hard to characterize.
-% They also present a new software package to do mitigate the effects.
-% Software packages \texttt{PhoSim} (Peterson et al.
-2015; \citep{2015ApJS..218...14P}) or \texttt{ARROYO} \citep{2004SPIE.5497..290B} could be used to characterize the PSF stability as a function of exposure time.
 
@@
 There is no maximum exposure time specified for an LSST image.
-Given that the template image will be a stack of at least a year or two of data, processing a $5$--$10$ times deeper single image through the difference imaging pipeline should be fine.
-However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and photometric calibrations.
-Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures (unknown), which could impact the quality of a difference image and the detected sources.
-Additionally, any system qualities that vary on short (but $>30$ second) timescales could inhibit photometric calibration (e.g., tracking).
+Given that the template image will be a stack of at least a year or two of data, 
+processing a $5$--$10$ times deeper single image through the difference imaging pipeline 
+should be fine.
+
+However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps 
+leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and 
+photometric calibrations.
+
+Furthermore, cosmic-ray rejection completeness might be reduced for longer exposures 
+(unknown), which could impact the quality of a difference image and the detected sources. 
+Additionally, any system qualities that vary on short (but $>30$ second) timescales could 
+inhibit photometric calibration (e.g., tracking).
 
@@
 
-Images obtained during twilight for scientific purposes are also likely to have shorter exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also apply here.
-Whether or not bright-background images can (or shall) be fully processed -- reduced, calibrated, background-subtracted, and delivered with astrometric and photometric solutions -- or whether this will require a user generated pipeline, remains to be determined (see also the example in Section \ref{ssec:SPCS_Twilight}).
+Images obtained during twilight for scientific purposes are also likely to have shorter 
+exposure times, and so the issues described in Section \ref{ssec:procbounds_expt} also 
+apply here.
+
+Whether or not bright-background images can (or shall) be fully processed -- reduced, 
+calibrated, background-subtracted, and delivered with astrometric and photometric 
+solutions -- or whether this will require a user generated pipeline, remains to be 
+determined (see also the example in Section \ref{ssec:SPCS_Twilight}). 
 This may depend on the exposure time and the number of stars available in the image.
@@
 
-Non-sidereal tracking leads to images in which stars are streaked, but the moving object appears as a point source.
-Full processing -- providing reduced, calibrated, background-subtracted images that are delivered with astrometric and photometric solutions -- of these images is beyond the scope of the DM pipelines as it would require the development of new algorithms, and will need to be done as a user generated pipeline.
-The first steps of such a pipeline, such as Instrument Signature Removal, will probably be possible to achieve by reconfiguring the relevant DM software tasks.
+Non-sidereal tracking leads to images in which stars are streaked, but the moving object 
+appears as a point source.
 
-%%% MLG removed the following in Feb 2022, it's no longer a concern.
-% \subsection{Number of Exposures per Visit (Long Sequences of a Single Field)}
-
-% There is no processing constraint on the number of consecutive exposures that could be obtained of a single field.
-% From a DM perspective, it would be best if these exposures were packaged into visits of no more than 2 exposures per visit, to minimize the need to reconfigure of the pipelines, and because the camera only ``clears" between visits.
-
-% K.-T.
-Lim has pointed out that an odd number of exposures is a non-standard visit; two snaps is hardwired into the code.
-This is baked-in to a configuration so that the pipeline can have a definition of what kind of timing delay constitutes ``late".
-Moving away from 2 exposures per visit requires a configuration change to the pipelines, which incurs an overhead (up to 1 minute) -- in fact, K.-T.
-things that between $10$ and $120$ seconds exposure times can easily be handled by the pipeline (i.e., can be run through ISR using scaled calibration frames), so long as they come in pairs.
-The real problem is knowing how long the processing should take, and not killing a process that is taking longer because there were 4 snaps in the visit instead of 2.
-To accommodate non-standard visits requires that the scheduler pass on the information of the number of snaps in the visit (\ref{DMSR-1}).
-Then the processing pipeline will know to, e.g., not attempt to difference the two snaps in the case were there is an odd number of snaps in a visit.
-\textit{MLG -- I've heard rumors of a CR regarding alternate standard visits of $1\times30$ seconds, but do not know the status or implications of this.}
-
-% K.-T.
-has also pointed out that currently, a deep drilling field would be interpreted as a single visit of 50 exposures by the scheduler.
-One implication of this is that since the camera only ``clears" prior to a new visit, it would not do this for the entire 50-exposure sequence.
-The processing pipeline would need to know how to divide this sequence up into visits.
-As there is no current requirement for DM to receive the information that the scheduler is about to do a 50 exposure visit, we need \ref{DMSR-1} to add the proposal ID and the number of exposures per visit to the meta data, and then it should be OK for DM to parse this visit information in the reduction pipeline.
-
-%%% MLG removed the following in Feb 2022, it's no longer a concern for Special Programs (crowded fields are in the WFD).
-% \subsection{Images in Very Crowded Fields}
-
-% The LSST pipelines' performance in crowded fields is documented in \citeds{DMTN-077}, which finds that, e.g., in Galactic Plane regions with a source density of $500000$ sources per square degree, the completeness drops to 50\% at $20.2$ magnitudes.
-% The slide deck at \citeds{Document-27962} also describes DM's plans for processing crowded fields. These may or may not be appropriate for Special Programs data, depending on the science goals.
+Full processing -- providing reduced, calibrated, background-subtracted images that are 
+delivered with astrometric and photometric solutions -- of these images is beyond the 
+scope of the DM pipelines as it would require the development of new algorithms, and will 
+need to be done as a user generated pipeline. 
+The first steps of such a pipeline, such as Instrument Signature Removal, will probably 
+be possible to achieve by reconfiguring the relevant DM software tasks.

### 2023-05-20 DMTN-065.tex
Update DMTN-065.tex

--- 2022-02-22 DMTN-065.tex
@@
 \usepackage{latexsym}
-% \usepackage{color}
-% black, blue, brown, cyan, darkgray, gray, green, lightgray, lime, magenta, blue, orange, pink, purple, red, teal, violet, white, yellow.
 \usepackage{enumitem}
 
-\title[LSST Special Programs]{Data Management \\ and LSST Special Programs}
+\title[LSST Special Programs]{Data Management for LSST Special Programs}
 
@@
 \setDocAbstract{
-The core LSST science goals will be met by the Wide-Fast-Deep (WFD) Main Survey, which is expected to be accomplished with 85--90\% of the observing time.
-The remaining 10--15\% of the time will be spent on Special Programs: additional survey areas and/or observing strategies that are driven by specific science goals which build on, or are beyond, the core science pillars of the LSST.
+This document provides a summary of the requirements on Rubin Data Management for processing and serving data products
+related to Special Programs: additional survey areas and/or observing strategies that are driven by specific science 
+goals which build on, or are beyond, the core science pillars of the LSST.
 
-\medskip
-This document provides a summary of the Special Programs being considered, the potential diversity of data they might produce (e.g., shorter or longer exposure times), and the role of Rubin Observatory in reducing, processing, and serving data products from Special Programs.
-The latter includes a discussion about the relevant system requirements, the needed capabilities of the LSST Science Pipelines and the Rubin Science Platform, and the cases in which Special Programs data products would be User-Generated.
+Recommendations for the specific implementations needed to fulfill these requirements and maximize science from
+Special Programs are provided on a case-by-case basis, with examples.
 
-\medskip
-The specialized, science-specific aspects of Special Programs processing and analysis that are best left to the science community are also described, and illustrated with case-study examples.
+Hardware and processing boundaries on the potential diversity of data from Special Programs 
+(e.g., longer or shorter exposure times), and the ability of the Data Management System to
+process this data, and also provided along with a discussion of when user-generated processing 
+and data products will be needed to meet some science goals.
 
-\medskip
-The main target audience of this document is Rubin Observatory staff -- the construction-era Data Management (DM) team and the operations-era Data Production (RDP) team -- but members of the science community who are planning to use Special Programs to reach their science goals may also find this document useful.
+The main target audience of this document is Rubin Observatory staff -- the construction- and operations-era Data Management
+teams -- but members of the science community who are planning to use Special Programs to reach their science goals may also 
+find this document useful.
 }
@@
 \addtohist{2}{2021-12-01}{Updates per DM-20375.}{Melissa Graham}
+\addtohist{3}{2023-05-31}{Updates per DM-32723.}{Melissa Graham}
 }
@@
 
-% % % % % % % % % % % % % % % % % % 
-\section{Terms and Definitions} \label{sec:terms}
+\input{sec_intro.tex}
+x
+\input{sec_sci.tex}
 
-\subsection{Wide-Fast-Deep (WFD) Main Survey}\label{ssec:terms_wfd}
-The WFD Main Survey is the core science program of the LSST, designed to achieve the science goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the WFD Main Survey is thought of as several contiguous areas that cover at least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, and is expected to be accomplished with $\sim$85\% of the observing time.
-These areas include the low-dust-extinction survey area (the largest area), the higher-extinction high stellar density Galactic Bulge, and the Magellanic Clouds.
+\input{sec_proc.tex}
 
-\subsection{Mini-Surveys}\label{sssec:terms_wfd_mini}
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``mini-survey" refers to additional areas that are adjacent to the WFD but are covered with a few hundred (100-300) visits per pointing.
-They are expected to be accomplished with $\sim$10\% of the total survey time.
-Mini-surveys include the North Ecliptic Spur (NES), the remainder of the Galactic Plane, and the South Celestial Pole.
-Boundaries, re-visit cadences, and depths for all WFD areas remain to be determined \citedsp{PSTN-053}.
-
-\subsection{Special Programs}\label{ssec:terms_sp}
-Special Programs are survey areas and/or observing strategies, aside from the WFD and mini-surveys, that are driven by specific science goals which either build on, or are outside of, the core science pillars of the LSST.
-Special Programs will fill the remaining $\sim$5\% of the available survey time, and include both ``deep drilling" fields and ``micro-surveys".
-
-\subsubsection{Deep Drilling Field (DDF)}\label{sssec:terms_sp_ddf}
-A DDF is a single pointing for which many exposures are obtained during the night.
-The five DDFs (below) are expected to take $\sim$5\% of the total available time.
-Generally, the LSST observing strategy for a DDF is to obtain some or all of these exposures consecutively, and to maintain a high inter-night cadence over a short season (e.g., returning every two nights over four months, \citealt{2019ApJ...873..111I}) -- but the exact strategies and total time for the LSST DDFs remain to be determined \citedsp{PSTN-053}.
-
-\begin{itemize}
-\item Elias S1 (00:37:48, -44:00:00)
-\item XMM-LSS (02:22:50, -04:45:00)
-\item Extended Chandra Deep Field-South (03:32:30, -28:06:00)
-\item COSMOS (10:00:24, +02:10:55)
-\item Euclid Deep Field South  (04:04:58, -48:25:23)\footnote{\url{https://www.cosmos.esa.int/web/euclid/euclid-survey}}
-\end{itemize}
-
-\subsubsection{Micro-Surveys}\label{sssec:terms_sp_micro}
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, the term ``micro-survey" refers to either new sky areas observed with a WFD-like survey, or sky areas within the WFD but observed with a specialized strategy.
-Micro-surveys are expected to take up the remaining $\sim$10\% of the total available survey time over LSST's 10 years, and those which would use $>$0.3\% of the time and are currently being considered by the SCOC are listed below (see also \citeds{PSTN-053}).
-
-\begin{itemize}
-\item short-exposure twilight visits (for near-Sun and near-Earth objects; NEOs)
-\item a static short-exposure (5 sec) map of the sky in $ugrizy$ in the first year (for calibration)
-\item an extended short-exposure survey of the sky in $ugrizy$ (for transient detection and static-sky calibration)
-\item target-of-opportunity (ToO) follow-up (to identify optical counterparts to gravitational wave sources)
-\item coverage of the Roman microlensing bulge field (potentially as a DDF)
-\item deeper $g$-band imaging of 10 local volume galaxies
-\item a high-cadence survey of 2 fields in the SMC (for microlensing)
-\item annual week-long surveys of the Carina nebula and surrounding star-forming regions
-\item a limited-visit (i.e., shallow) ``northern stripe" survey to declination $<$+30 degrees
-\item a survey of the Virgo cluster to WFD depth
-\end{itemize}
-
-Note that there remains some grey area between what is a mini-survey within the WFD, and what is a WFD-like micro-survey.
-For example, the last two on the list above (the northern stripe and the Virgo cluster) could also be (or might be in the future) called WFD mini-surveys, and not Special Programs micro-surveys.
-
-\subsection{Visit Types}\label{ssec:terms_visits}
-A visit is an observation of a single pointing at a given time, of which there are three types as listed below.
-As of the Phase 1 SCOC recommendations in \citeds{PSTN-053}, most WFD and DDF visits would be standard or alternative standard visits.
-However, non-standard visits with longer exposure times are being considered, especially for $u$-bands, and several of the micro-surveys are considering shorter exposure times.
-The full potential for diversity in Special Programs data is reviewed in Section~\ref{ssec:proc_datadiv}.
-
-\begin{itemize}
-\item Standard Visit -- Composed of $2\times15$ second exposures (commonly referred to as ``snaps").
-\item Alternative Standard Visit -- Composed of a single $30$ second exposure.
-\item Non-Standard Visit -- Any other exposure time(s) or number of snaps.
-\end{itemize}
-
-\clearpage
-% % % % % % % % % % % % % % % % % % 
-\section{Reduction and Processing for Special Programs Data Products}\label{sec:proc}
-
-\subsection{Requirements Related to Special Programs}\label{ssec:proc_reqs}
-% \lsrreq, \ossreq, \dmreq
-% \reqparam
-
-A detailed list of the requirements related to Special Programs is provided in Appendix~\ref{sec:docrev}.
-The most important requirements are summarized below.
-
-\subsubsection{Data Products}
-
-Rubin Observatory and the LSST system (the observatory and the data management systems) are required to process Special Programs data to produce unique and separate data products ``whenever possible" (LSR-REQ-0121).
-
-The term ``whenever possible" includes cases where the original or reconfigured versions of the LSST Science Pipelines can be run, and excludes cases where the development of new algorithms or the allocation of significant additional computational resources are required (LSR-REQ-0121).
-
-The statement ``to produce unique and separate data products" typically refers to producing the same kinds of data products as will be generated by the Prompt and Data Release pipelines (processed visit images, coadded images, difference images, and catalogs of sources and objects for those images).
-
-It is a requirement that the cumulative size of the Special Programs data products generated by Rubin Observatory be no more than $\sim$10\% the size of the Data Release data products (i.e., proportional to the fraction of survey time spent; LSR-REQ-0121).
-
-It is a requirement that these Special Programs data products be distinct, and joinable with (in other words, they can be federated or cross-matched with) the Prompt and Data Release data products (DMS-REQ-0322).
-
-The derivation of value-added data products, such as HiPS or MOC maps, for Special Programs remains an open question (DMS-REQ-0379, 0383).
-
-\subsubsection{Metadata}
-
-In order to support Special Programs processing, the LSST system is required to store metadata that includes program information for every raw image, such as identifiers for images obtained as part of the Main Survey or a Special Program (DMS-REQ-0068).
-
-It is required that this metadata be sufficient for Special Programs to trigger their own real-time data processing recipes ``whenever possible" (DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).
-
-\subsubsection{Processing}
-
-It is a requirement that Special Programs processing with the Prompt pipeline (or a reconfigured version of it) is subject to the same timescales and latency constraints of 24 hours for the release of Prompt data products and 1 minute for the transmission of Alert packets (DMS-REQ-0344).
-
-It is also a requirement that Special Programs processing be done on timescales intermediate to the Prompt and Data Release processing, ``whenever possible" and whenever necessary to enable the intended science goals of the Special Program (LSR-REQ-0032).
-
-It is a requirement that the LSST system be able to process non-standard visits with short exposure times as low as 1 second, with a discussion note that such short exposures might have degraded image quality (LSR-REQ-0111).
-
-Processing for Special Programs by Rubin Observatory is expected to use no more than $\sim$10\% of computational and storage capacity of the Rubin data processing cluster (i.e., proportional to the fraction of survey time spent and the size of the Rubin-processed Special Programs data products; Section 6 of the DPDD).
-
-\subsubsection{User Processing}
-
-In cases where the science goals of a Special Program require that new algorithms or software be developed, User-Generated pipelines and data products will be needed.
-
-The 10\% of the total data processing capacity that Rubin Observatory is required to reserve for all User-Generated processing includes that applied by users to Special Programs data -- in other words, there is no additional capacity {\it for users} that will be reserved only for Special Programs data (LSR-REQ-0041).
-
-\subsection{The Potential Diversity of Special Programs Data} \label{ssec:proc_datadiv}
-
-As discussed in Section~\ref{sec:terms}, most of the Special Programs that are currently under consideration will use standard or alternative standard visits.
-However, some are likely to require non-standard visits with shorter exposures (5 sec).
-Furthermore, some are likely to acquire images with a significantly brighter sky background (e.g., twilight images) than most nighttime survey images.
-
-The cadence and patterns of Special Programs might also differ from the WFD, such as long series of exposures obtained of the same field (e.g., DDFs), or a strategy optimized to find very fast-moving objects (e.g., NEOs).
-
-It does not appear that any of the currently-proposed Special Programs are likely to violate boundaries imposed by the Rubin Observatory hardware, but there remain a few open questions about boundaries on data processing imposed by the LSST Science Pipelines.
-
-\subsubsection{Hardware Boundaries}
-
-Appendix~\ref{sec:hardbounds} lists all of the hardware boundaries that might constrain the potential diversity of Special Programs data.
-
-The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and there is a potential hardware boundary that limits the readout rate to 1 every 15 seconds.
-This would affect the image acquisition rate and increase the overheads of the proposed short-exposure micro-surveys.
-
-Special Programs which required the \emph{exact same} field pointing and rotation for \emph{every exposure} (to sub-arcsecond levels) might run into hardware boundaries on pointing and tracking, but none of the currently-proposed programs seem likely to be affected by this potential boundary.
-
-Hardware imposes no other boundaries on how data can be obtained, but Special Programs that request a high number of filter changes and/or long slews could be inefficient due to large overheads.
-
-\subsubsection{Processing Boundaries}
-
-Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits can be processed and calibrated by the LSST Science Pipelines, which are designed to process standard (or alternative standard) visits.
-
-Very short ($<$2 sec) exposures could be difficult to reduce due to an incompletely-formed PSF, and very short or very long ($>$150 sec) exposures could be difficult to calibrate due to having too few (or too few unsaturated) stars.
-As mentioned above in Section~\ref{ssec:proc_reqs}, the LSST system is required to be able to process exposure times as low as 1 second, and it is known that such short exposures might have degraded image quality.
-
-It is currently unclear whether images with very bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, or whether user generated pipelines will be needed.
-
-The full reduction and calibration of images obtained with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope of the LSST Science Pipelines, and will require a user generated pipeline.
-
-\subsection{Including Special Programs Data in the WFD Main Survey's Data Products}\label{ssec:proc_wfd}
-
-Instances where the Special Programs data might be included in the Prompt and/or Data Release data products alongside data from the WFD Main Survey -- i.e., the data products described by \citeds{lse-163} -- are preliminarily described.
-
-Generally, Rubin Observatory might incorporate Special Programs data into the data products for the WFD Main Survey whenever this is (1) possible and (2) scientifically beneficial.
-{\bf Decisions about when and whether to include Special Programs data in the WFD Main Survey's data products are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
-
-\subsubsection{Prompt}
-
-It would be both possible and scientifically beneficial to include all standard and alternative standard visits from Special Programs in Prompt processing and Alert Production, alongside the standard visits from the WFD Main Survey.
-
-Non-standard visits that can be processed with the Prompt pipeline in accordance with the relevant requirements might also be included.
-
-Very short and very long exposure times might be excluded if they would need specialized algorithms or templates\footnote{If a Special Program's science goals do require specialized templates and Prompt processing, the DMS will have the capability to load and use an alternative template for some regions, based on the image metadata.
-However, there would not be enough memory to hold alternative templates for the whole sky region.}.
-Visits in sky regions that Rubin had not previously observed would not be able to contribute to Alert Production until a template could be generated.
-
-The data products from Special Programs images would thus contribute to all of the Prompt data products described in Section 3 of the DPDD \citedsp{lse-163}. 
-Images and catalog rows would be tagged with a program identifier indicating their Special Program of origin, so that the associated data products can be discovered and extracted by users.
-
-There are two potential issues which might complicate Prompt processing for Special Programs which obtain a sequence of images without slewing, such as the Deep Drilling Fields.
-(1) As alert packets contain the full records of all associated {\tt DIASources} from the past 12 months, alerts for {\tt DIAObjects} in the DDFs might become very large.
-(2) The association of {\it new} {\tt DIASources} into {\tt DIAObjects} will be somewhat compromised for a DDF sequence.
-The processing for the second image begins when the processing for the first image is only halfway done -- when the {\tt DIAObject} catalog has not yet updated with the new {\tt DIASources} detected in the first image.
-Thus, the {\tt DIASource} from images one and two for a new transient would not be associated with a single {\tt DIAObject}, but instead would each instantiate a new {\tt DIAObject}.
-
-As a side note, no User-Generated pipeline may contribute Alerts to the LSST Alert Stream, and since the latency on image availability is ``within 24 hours" ({\tt L1PublicT}\reqparam{L1PublicT}) no User-Generated pipeline would be able to obtain and process Special Programs data on a timescale similar to the Alert Stream (60 seconds, {\tt OTT1}\reqparam{OTT1}).
-
-\subsubsection{Solar System Processing}
-
-Since Solar System Processing takes \texttt{DIASources} as input, so any Special Programs images that can be run through the Alert Pipeline can also be incorporated into Solar System Processing.
-
-\subsubsection{Data Release}
-
-It would be both possible and scientifically beneficial to include all standard and alternative standard visits from Special Programs in {\it some aspects} of the Data Release processing -- such as the repository of processed visit images and the {\tt Source} and {\tt Forced Source} catalogs -- alongside the standard visits from the WFD Main Survey.
-Images and catalog rows would be tagged with a program identifier indicating their Special Program of origin, so that the associated data products can be discovered and extracted by users.
-
-However, some of the core LSST science goals require a WFD Main Survey data products of {\it uniform depth}.
-Whether and how to include any Special Programs data in the deep image coadds and the corresponding {\tt Object} catalog is left to the discretion of the appropriate Rubin teams in Operations.
-
-For example, perhaps Special Programs images will only be included when they bring additional area to the same depth as the rest of the WFD Main Survey, or when they suppress edge effects or low-order modes in the all-sky photometric solutions.
-
-\subsection{Anticipated Rubin- and User-Generated Special Programs Data Products}\label{ssec:proc_spdp}
-
-As mentioned in Section~\ref{ssec:proc_reqs}, Rubin Observatory and the LSST system are required to process Special Programs data to produce unique and separate data products whenever the original or reconfigured versions of the LSST Science Pipelines can be used.
-
-Cases where the development of new algorithms, or the allocation of significant additional computational resources, are required in order to produce unique and separate data products will require User-Generated processing.
-
-The anticipated divisions between what will be Rubin- versus User-Generated for the variety of anticipated ``unique and separate" Special Programs data products, which are designed to meet the specific science goals of the Special Programs, are {\it preliminarily} described.
-{\bf Decisions about when and whether Rubin Observatory will generate Special Programs data products are left to the discretion of the Rubin Operations Data Production and System Performance teams.}
-
-For Special Programs which will require User-Generated pipelines and data products, note that users will have access to all of the LSST Science Pipelines and its data processing infrastructure, as well as dedicated computational resources next-to-the-data (via the Rubin Science Platform; \citeds{lse-319}).
-Details of the planned ``User Batch" facility for data processing are described in \citeds{dmtn-202}.
-However, very computationally intense processing might require external resources.
-
-Furthermore, it is expected that some User-Generated pipelines and data products might be ``adopted" or ``federated" into the LSST Science Pipelines and the Prompt and Data Release data products. 
-Details regarding this process are to be provided elsewhere.
-
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.
-
-\subsubsection{Deep Drilling Fields}
-
-As the DDFs will likely be observed with standard or alternative standard visits, it is likely that Rubin Observatory will be able to reconfigure existing pipelines to process the DDF data and produce ``unique and separate" DDF data products.
-
-Rubin-generated data products might include:\\
- - nightly-coadded images (24 h)\\
- - nightly-coadded difference images (24 h)\\
- - deeply-coadded images (all images to date; yearly)\\
- - templates for the nightly-coadded difference images (yearly)\\
- - Source- and Object-like catalogs for the night and deep coadds (yearly)\\
- - DIASource- and DIAObject-like catalogs for the nightly-coadds (24 h)\\
-
-User-Generated data products might include, for example, DDF images coadded on other timescales, or using algorithms outside of the LSST Science Pipelines.
-
-\subsubsection{Short-Exposure Surveys}
-
-Short-exposure images obtained during twilight, which will have a very bright sky background unlike other LSST images, might require User-Generated algorithms to subtract the high sky background. 
-Short-exposure images obtained during the night would have fewer stars for astrometric and photometric calibration, and might require User-Generated processing pipelines.
-
-If the Rubin Operations team decides that the LSST Science Pipelines can be reconfigured and used for short exposures, then the Rubin-generated data products would likely be similar to the Data Release data products (Section 4, \citeds{lse-163}).
-For example, a separate repository of processed visit images and coadded images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and {\tt Object} catalogs.
-
-\subsubsection{GW Target-of-Opportunity}
-
-As the science goals of any target-of-opportunity program are the immediate processing and discovery of new transient phenomena, it is very likely that all ToO programs would be designed to use standard or alternative standard visits and Prompt processing data products such as Alerts.
-Prompt processing would proceed automatically for all ToO for which an LSST template image exists.
-
-Options for Rubin Observatory to assist with or expedite the processing of ToO programs, especially during the first year of Operations when the template coverage will be low, are discussed in more detail in \citeds{rtn-008}.
-
-Other science-driven data products from ToO programs, such as custom deep stacks and difference images to find $<$5$\sigma$ transients in the fields on intermediate timescales in cases where the optical counterpart was not detected, would require User-Generated processing.
-
-\subsubsection{Micro-Surveys}
-
-Potential micro-surveys include additional sky areas, in some cases covered to a depth that is different from the WFD Main Survey, such as the proposed micro-surveys for the Roman bulge field; deeper $g$-band imaging of 10 local volume galaxies; the SMC; the Carina nebula; the northern stripe; and the Virgo Cluster.
-
-In all of these cases, if the Rubin Operations team decides that the LSST Science Pipelines can be reconfigured and used for the visits associated with these micro-surveys -- which is likely if they're using standard or alternative standard visits --  then the Rubin-generated data products would be similar to the Data Release data products (Section 4, \citeds{lse-163}).
-For example, separate repositories of processed visit images and coadded images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and {\tt Object} catalogs would be created for all of these micro-surveys.
-
-In cases where data for the micro-survey is obtained throughout the year, then the separate data products would also be produced on a yearly basis.
-In cases where the data might be all obtained within a week (e.g., Carina nebular survey), such data products would likely be generated on a shorter (intermediate) timescales.
-
-Micro-surveys with time-domain science goals that aren't met by the Prompt pipelines -- for example if they require difference imaging with coadded images on an intermediate timescale (e.g., a weekly stack) --  might require User-Generated processing.
-
-\clearpage
-% % % % % % % % % % % % % % % % % % 
-\section{Enabling the Discovery and Analysis of Special Programs Data Products}\label{sec:analysis}
-
-\subsection{Anticipated Rubin Science Platform Capabilities for Special Programs }
-
-\begin{itemize}
-\item discoverability of SP data when browsing an all-sky map
-\item to query data by tag of WFD, WFD mini-survey, or Special Program (e.g., DDF field, micro-survey identifier)
-\end{itemize}
+\input{sec_req.tex}
 
@@
 
-\input{ap_docrev.tex}
+\input{ap_spcs.tex}
+
+\input{ap_procbounds.tex}
 
@@
 
-\input{ap_procbounds.tex}
-
-\input{ap_spcs.tex}
-
-\input{ap_prevpropsp.tex}
+\input{ap_docrev.tex}
 

### 2023-05-20 sec_intro.tex
Add files via upload

\section{Introduction} \label{sec:intro}

Special Programs is a Rubin Data Management (RDM) term used to refer to survey 
components with specific sky regions within or beyond the 
Wide-Fast-Deep (WFD) main survey, and/or distinct cadences or 
strategies from the WFD main survey.

There are several requirements on the Rubin Data Management System (DMS) 
related to Special Programs, summarized in Section~\ref{sec:req}.
This document recommends implementation aspects for these requirements in order to 
maximize science from Special Programs (Section~\ref{sec:sci}).

Special Programs are driven by specific science goals that build on or 
add to the core science pillars of the LSST.
About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
observations associated with Special Programs (and 85--90\% for WFD).

Special Programs includes non-WFD LSST components such as the Deep Drilling 
Fields (DDFs) and the mini-, micro-, and nano-surveys; some of these components
might use non-standard visits with different exposure times
(see Section~\ref{ssec:intro_terms}, below, for details).

\subsection{Additional Terms and Definitions}\label{ssec:intro_terms}

\textbf{Wide-Fast-Deep (WFD) Main Survey -- }
The core science program of the LSST, designed to achieve the science 
goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).

It is a contiguous area that covers at least $\sim$18000 deg$^2$ with 
$\gtrsim$800 visits per field, expected to be accomplished with 85--90\% 
of the observing time.
As of the Phase 2 SCOC recommendations in \citeds{PSTN-055}, the WFD area 
is considered to include the low dust extinction extragalactic regions; 
the higher extinction, high stellar density Galactic Plane and Bulge; the 
Virgo Cluster; the South Celestial Pole (SCP); and the North Ecliptic Spur 
(NES).

In the past, the Galactic, SCP, and NES regions have been referred to as 
"mini-surveys", and the scheduler software might treat them as such, but 
this does not make them "Special Programs". 
From a survey-scheduling perspective, they are now considered part of the 
WFD survey.
From a science and data-processing perspective, these components of the 
WFD do not require special processing or separate data products.

\textbf{Deep Drilling Field (DDF) -- }
A single pointing for which many (e.g., a hundred) visits are obtained 
(usually sequentially) during a single night, and repeated every few 
nights.

As of the Phase 2 SCOC recommendations in \citeds{PSTN-055}, the five 
confirmed DDFs were:

\begin{itemize}
\item Elias S1 (00:37:48, -44:00:00)
\item XMM-LSS (02:22:50, -04:45:00)
\item Extended Chandra Deep Field-South (03:32:30, -28:06:00)
\item COSMOS (10:00:24, +02:10:55)
\item Euclid Deep Field South  (04:04:58, -48:25:23)\footnote{\url{https://www.cosmos.esa.int/web/euclid/euclid-survey}}
\end{itemize}

From a science and data-processing perspective, the DDFs do require 
special processing or separate data products.

\textbf{Mini-, Micro- and Nano-Surveys -- }
Specific sky areas covered by a few hundred, a hundred, or tens of visits 
(respectively; as in Phase 2 SCOC recommendations, \citeds{PSTN-055}).
For a list of the current mini-surveys under consideration, see 
\citeds{PSTN-055}.

This document refers to them all together as mini-surveys.

These sky areas are all (so far) either within or adjacent to the WFD area.
In some cases these visits are non-standard (e.g., short exposures).

From a data-processing and science perspective, most mini-surveys are 
likely to require special processing or separate data products, and should 
be considered on a case-by-base basis.

\textbf{Visit Types -- }
A visit is an observation of a single pointing at a given time, of which 
there are three types:

\begin{itemize}
\item Standard Visit -- Composed of $2\times15$ second exposures (commonly referred to as ``snaps").
\item Alternative Standard Visit -- Composed of a single $30$ second exposure.
\item Non-Standard Visit -- Any other exposure time(s) or number of snaps.
\end{itemize}

The Phase 2 SCOC recommendations for the WFD (\citeds{PSTN-055}) are to 
use alternative standard visits in the $u$-band and standard visits in all 
other bands; variable exposure times were not recommended for the WFD.

Non-standard visits with shorter or longer exposure times are being 
considered for some mini-surveys.

### 2023-05-20 sec_proc.tex
Add files via upload

\section{Processing for Special Programs Data Products}\label{sec:proc}

Rubin Observatory and the LSST Data Management System (DMS) are required 
to process Special Programs data to produce unique and separate data 
products whenever the original or reconfigured versions of the LSST 
Science Pipelines can be used to do so (Section~\ref{sec:req}).

The LSST Science Pipelines and the DMS are designed to process standard 
(or alternative standard) visits.

It depends on the science goals related to the Special Program, but some 
Special Programs' data can and should be processed and served along with 
(or as a part of) the WFD data products (Section~\ref{ssec:proc_wfd}).
Others will require unique and separate data products 
(Section~\ref{ssec:proc_rdm}).

Cases where the Special Program obtains non-standard visits that cannot 
be processed by the DMS (Section~\ref{ssec:proc_bounds}), and/or where 
development of new algorithms or the allocation of significant additional 
computational resources are needed, will require user processing 
(Section~\ref{ssec:proc_user}).

\subsection{Boundaries on Obtaining and Processing Non-Standard Visits} \label{ssec:proc_bounds}

Most of the Special Programs that are currently under consideration will 
use standard or alternative standard visits.

However some Special Programs are likely to use non-standard visits with 
shorter exposures (5 sec), and some are likely to acquire images with a 
significantly brighter sky background (e.g., twilight images) than 
nighttime survey images.

The cadence and patterns of Special Programs might also differ from the 
WFD, such as long series of exposures obtained of the same field 
(e.g., DDFs), or a strategy optimized to find very fast-moving objects 
(e.g., NEOs).

None of the currently proposed Special Programs (\citedsp{PSTN-055}) 
seem likely to approach (or violate) any harware boundaries 
(Section~\ref{ssec:proc_bounds_hardware}).

However, some Special Programs observations seem likely to generate 
images that cannot be processed with original or reconfigured versions of 
the LSST Science Pipelines (Section~\ref{ssec:proc_bounds_processing}), 
and will require user processing.

\subsubsection{Hardware Boundaries}\label{ssec:proc_bounds_hardware}

Appendix~\ref{sec:hardbounds} lists all of the hardware boundaries that 
might constrain the potential diversity of Special Programs data.

The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and 
there is a potential hardware boundary that limits the readout rate to 1 
every 15 seconds.
This would affect the image acquisition rate and increase the overheads 
on short exposures.

Special Programs which required the \emph{exact same} field pointing and 
rotation for \emph{every exposure} (to sub-arcsecond levels) might run 
into hardware boundaries on pointing and tracking.
None of the currently-proposed programs seem likely to be affected by 
this potential boundary.

Hardware imposes no other boundaries on how data can be obtained, but 
Special Programs that request a high number of filter changes and/or long 
slews could be inefficient due to large overheads.

\subsubsection{Processing Boundaries}\label{ssec:proc_bounds_processing}

Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits can be processed and calibrated by the DMS.

Very short ($<$2 sec) exposures could be difficult to reduce due to an 
incompletely-formed PSF, and very short or very long ($>$150 sec) 
exposures could be difficult to calibrate due to having too few (or too 
few unsaturated) stars.
As mentioned above the DMS is required to be able to process exposure 
times as low as 1 second, and it is known that such short exposures might 
have degraded image quality.

It is currently unclear whether images with very bright sky backgrounds 
(twilight images) can be processed with the LSST Science Pipelines, or 
whether user generated pipelines will be needed.

The full reduction and calibration of images obtained with non-sidereal 
tracking, in which the stars are streaked, is currently beyond the scope 
of the LSST Science Pipelines, and will require a user generated pipeline.

\subsection{RDM Including Special Programs in the WFD Data Products}\label{ssec:proc_wfd}

Generally, data from Special Programs should be included in the Prompt 
and/or Data Release processing and data products alongside data from the 
WFD (i.e., included in the data products described by \citeds{lse-163}) 
whenever this is (1) possible and (2) scientifically beneficial.

Decisions about when and whether to include Special Programs data in 
the WFD's processing and data products are left to the discretion of the 
Rubin Operations Data Production and System Performance teams.

\subsubsection{WFD Prompt Processing and Alert Production}

As all visits can contribute to the time-domain astronomy science goals 
of the LSST, all Special Programs data that {\it can} be processed by the 
prompt processing and alert production pipelines {\it should} be, 
alongside visits from the WFD.

This would include all standard and alternative visits in sky regions for which a template image exists.

This might also include some non-standard visits (shorter or longer exposures) as long as they can be processed by the Prompt 
pipeline\footnote{If a Special Program's science goals do require 
specialized templates and 
Prompt processing, the DMS will have the capability to load and use an 
alternative template for some regions, based on the image metadata. 
However, there would not be enough memory to hold alternative templates 
for the whole sky region.}.

Visits with very short or very long exposure times (or very bright sky 
backgrounds) might be excluded if they would 
need specialized algorithms for difference-imaging or template-generation.

Special Programs data that is processed by the Prompt pipeline would 
contribute to the Prompt data products described in Section 3 of the 
DPDD \citedsp{lse-163}.

Difference images, {\tt DiaSources}, and alerts would have region and program labels (Section~\ref{ssec:sci_labels}), providing origin and
provenance for brokers and users.

\textbf{Two potential issues with Prompt processing for DDFs}
(or any mini-survey that obtains a sequence of images without slewing)
are:

(1) As alert packets contain the full records of all associated 
{\tt DIASources} from the past 12 months, alerts for {\tt DIAObjects} in 
the DDFs might become very large.

(2) The association of {\it new} {\tt DIASources} into {\tt DIAObjects} 
will be somewhat compromised for a DDF sequence.
The processing for the second image begins when the processing for the 
first image is only halfway done; when the {\tt DIAObject} catalog has 
not yet updated with the new {\tt DIASources} detected in the first image.
Thus, the {\tt DIASource} from images one and two for a new transient 
would not be associated with a single {\tt DIAObject}, but instead would 
each instantiate a new {\tt DIAObject}.

These two issues would not impact time-domain science.
Brokers and users would be able to use the region and program labels
in the data as context (i.e., as flags for potentially-compromised 
data products) and, for example, avoid DDF alerts should they so choose.

\subsubsection{WFD Solar System Processing}

Since Solar System Processing takes \texttt{DIASources} as input, so any 
Special Programs images that can be run through the Alert Pipeline can 
also be incorporated into Solar System Processing.

\subsubsection{WFD Data Release Processing}

\textbf{Difference images, {\tt DiaSource} and {\tt DiaObject} catalogs -- }
As all visits can contribute to the time-domain astronomy science goals 
of the LSST, all Special Programs data that {\it can} be processed with 
the difference image analysis (DIA) pipeline alongside the WFD visits 
{\it should} be.
All of this processing should use the same template image for a given
field.
Users who want, e.g., only {\tt DiaSources} a given DDF, would be 
able to write queries using the region and program labels in the catalog.

\textbf{Processed visit images, {\tt Source} and {\tt ForcedSource} catalogs -- }
For the same reasons as above, all Special Programs data that 
{\it can} be processed with the Data Release pipieline alongside the WFD 
visits {\it should} be.

\textbf{Deep coadds and the Object catalog -- }
Some of the core LSST science goals require a WFD main survey data 
products of nearly {\it uniform depth}.
Whether and how to include any Special Programs data in the deep image 
coadds and the corresponding {\tt Object} catalog is left entirely to the 
discretion of the appropriate Rubin teams in Operations.

For example, perhaps Special Programs images will only be included when 
they bring additional area to the same depth as the rest of the WFD, or when they suppress edge effects or low-order modes in the 
all-sky photometric solutions.

For stacked images and their associated catalogs, the region labels should be
included but the program labels would not be needed (or possible).

\subsection{RDM-Generated ``Unique and Separate" Data Products}\label{ssec:proc_rdm}

Rubin Data Management (RDM) is required to produce unique, separate, and 
joinable data products for Special Programs whenever this is possible with the 
original or reconfigured versions of the LSST Science Pipelines, and no new 
algorithmic development or significant computational resources are needed.

As described in Section~\ref{ssec:proc_wfd}, in some cases the science goals 
of a Special Program might be met by including the data in the WFD data 
products; especially the Prompt or DIA data products for standard visits.
Thus, most of the RDM-generated data products considered below are for 
static-sky Data Release data products, or data products for non-standard visits.

\textbf{Custom coadds -- }
As with the annual WFD deeply coadded images and associated catalogs, RMD would 
only generate one coadd (including all images to date) for a given special region.
Custom coadds (e.g., weekly, monthly) for specific science goals are considered 
user-generated data products, for both WFD and Special Programs.

Decisions about when to produce separate Special Programs data products are 
left to the discretion of RDM.

Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.

\subsubsection{Deep Drilling Fields}

As the DDFs will likely be observed with standard or alternative standard 
visits, it is likely that RDM will be able to reconfigure existing pipelines 
to process the DDF data and produce ``unique and separate" DDF data products.

For example, RDM-generated ``unique and separate" DDF data products might 
include:
\begin{itemize}
\item nightly-coadded images (24 h)
\item nightly-coadded difference images (24 h)
\item {\tt DIASource}- and {\tt DIAObject}-like catalogs for the nightly-coadds (24 h)
\item deeply-coadded images (all images to date; yearly)
\item templates for the nightly-coadded difference images (yearly)
\item {\tt Source}- and {\tt Object}-like catalogs for the nightly-coadded and deeply-coadded images (yearly)
\end{itemize}

\subsubsection{Short-Exposure Mini-Surveys}

As of the Phase 2 SCOC recommendations \citedsp{PSTN-055}, three kinds of 
short-exposure mini-surveys were under consideration: 
a static short exposure map of the sky in $ugrizy$ for calibration; 
a static to transient short exposure survey; and 
an Near-Earth Objects (NEO) twilight survey.

If either the transients or NEO short-exposure suvey is executed, and if 
RDM finds that the LSST Science Pipelines and, specifically, template images 
can be made and difference image analysis (DIA) run, then the RDM-generated 
data products may be similar to the Prompt data products (potentially 
including alerts; Section 3 of \citeds{lse-163}).

If either of the static short-exposure mini-surveys is executed, and if 
RDM finds that the LSST Science Pipelines can be reconfigured and used 
for short exposures, then the RDM-generated data products would likely be 
of a similar format to the Data Release data products 
(Section 4, \citeds{lse-163}).
For example, a separate repository of processed visit images and coadded 
images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and 
{\tt Object} catalogs.

\subsubsection{Mini-Surveys}

The cases where ``unique and separate" data products for mini-surveys are 
needed to reach the science goals of a Special Program, and are possible for 
RDM to generate, are most likely to be of a similar format to the annual 
Data Release data products: static-sky coadded images and associated 
{\tt Source}- and {\tt Object}-like cataogs for standard or alternative visits.

In cases where the observations are all obtained with a short season 
(e.g., a week spent observing a specific region), such ``annual" 
data products might be generated on an intermediate time scale, 
if resources allow.

\subsubsection{Target-of-Opportunity (TOO)}

Options for RDM to process TOO, especially during the first year of Operations 
when the template coverage will be low, are discussed in more detail in 
\citeds{rtn-008}.

\subsection{User-Generated ``Unique and Separate" Data Products}\label{ssec:proc_user}

For science goals that require data products which are not possible to create with 
the original or reconfigured versions of the 
LSST Science Pipelines, and/or for which new algorithmic development or significant 
computational resources are needed, 
user-designed and user-run processing will be needed.

As described above, custom coadds (e.g., weekly, monthly) are also left to users 
to generate, as required by their specific science goals.

Users will have access to the LSST Science Pipelines and data processing 
infrastructure, as well as dedicated computational resources next-to-the-data, 
via the Rubin Science Platform; \citeds{lse-319}.
Details of the planned ``User Batch" facility for data processing are described in 
\citeds{dmtn-202}.
Very computationally intense user processing might require external resources.

It is expected that some User-Generated pipelines and data products 
might be ``adopted" or ``federated" into the LSST Science Pipelines and the Prompt 
and Data Release data products, however, details regarding the federation of 
user-generated data products are to be provided elsewhere.

\textbf{Prompt Processing / Alert Production -- }
As the latency on processed visit image availability has an 80-hour
embargo, no user-generated pipeline will be able to process Special Programs 
data on a timescale similar to prompt processing and alert production 
(60 seconds to 24 hours).
No user-generated pipeline may contribute alerts to the LSST alert stream on 
any timescale.

\subsubsection{Deep Drilling Fields}

User-generated data products might include, for example, DDF images coadded on custom 
timescales (e.g., weekly, monthly), or coadded using algorithms outside of the LSST 
Science Pipelines.

\subsubsection{Short-Exposure Mini-Surveys}

Short-exposure images obtained during twilight, which will have a very bright sky 
background unlike other LSST images, might require User-Generated algorithms to 
subtract the high sky background.

Short-exposure images obtained during the night would have fewer stars for 
astrometric and photometric calibration, and might require user-generated 
processing pipelines.

\subsubsection{Mini-Surveys}

Mini-surveys with time-domain science goals that aren't met by the Prompt pipelines, 
e.g., those that require difference imaging with coadded images on an intermediate 
timescale (e.g., a weekly stack), might require user-generated processing.

Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.

### 2023-05-20 sec_req.tex
Add files via upload

\section{Requirements Related to Special Programs}\label{sec:req}
% \lsrreq, \ossreq, \dmreq
% \reqparam

This section proves a summary of the requirements related to Special Programs.
A more detailed list is provided in Appendix~\ref{sec:docrev}.

\subsection{Data Products}\label{ssec:req_dp}

Rubin Observatory and the LSST system (the observatory and the data 
management systems) are required to process Special Programs data to 
produce unique and separate data products ``whenever possible" 
(LSR-REQ-0121).

The term ``whenever possible" includes cases where the original or 
reconfigured versions of the LSST Science Pipelines can be run, and 
excludes cases where the development of new algorithms or the allocation 
of significant additional computational resources are required 
(LSR-REQ-0121).

The statement ``to produce unique and separate data products" typically 
refers to producing the same kinds of data products as will be generated 
by the Prompt and Data Release pipelines (processed visit images, coadded 
images, difference images, and catalogs of sources and objects for those 
images).

It is a requirement that the cumulative size of the Special Programs data 
products generated by Rubin Observatory be no more than $\sim$10\% the 
size of the Data Release data products (i.e., proportional to the fraction 
of survey time spent; LSR-REQ-0121).

It is a requirement that these Special Programs data products be distinct, 
and joinable with (in other words, they can be federated or cross-matched 
with) the Prompt and Data Release data products (DMS-REQ-0322).

The derivation of value-added data products, such as HiPS or MOC maps, for 
Special Programs remains an open question (DMS-REQ-0379, 0383).

\subsection{Metadata}\label{ssec:req_meta}

In order to support Special Programs processing, the LSST system is 
required to store metadata that includes program information for every raw 
image, such as identifiers for images obtained as part of the Main Survey 
or a Special Program (DMS-REQ-0068).

It is required that this metadata be sufficient for Special Programs to 
trigger their own real-time data processing recipes ``whenever possible" 
(DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).

\subsection{Processing}\label{ssec:req_proc}

It is a requirement that any Special Programs processing done with the 
Prompt pipeline (or a reconfigured version of it) is subject to the same 
timescales and latency constraints of 24 hours for the release of Prompt 
data products and 1 minute for the transmission of Alert packets 
(DMS-REQ-0344).

It is also a requirement that Special Programs processing be done on 
timescales intermediate to the Prompt and Data Release processing, 
``whenever possible" and whenever necessary to enable the intended science 
goals of the Special Program (LSR-REQ-0032).

It is a requirement that the LSST system be able to process non-standard 
visits with short exposure times as low as 1 second, with a discussion 
note that such short exposures might have degraded image quality 
(LSR-REQ-0111).

Processing for Special Programs by Rubin Observatory is expected to use no 
more than $\sim$10\% of computational and storage capacity of the Rubin 
data processing cluster (i.e., proportional to the fraction of survey time 
spent and the size of the Rubin-processed Special Programs data products; 
Section 6 of the DPDD).

\subsection{User Processing}\label{ssec:req_ug}

In cases where the science goals of a Special Program require that new 
algorithms or software be developed, User-Generated pipelines and data 
products will be needed.

The 10\% of the total data processing capacity that Rubin Observatory is 
required to reserve for all User-Generated processing includes that 
applied by users to Special Programs data -- in other words, there is no 
additional capacity {\it for users} that will be reserved only for Special 
Programs data (LSR-REQ-0041).

### 2023-05-20 sec_sci.tex
Add files via upload

\section{Maximizing Science from Special Programs}\label{sec:sci}

Recommended implementation aspects for the requirements related to 
Special Programs (Section~\ref{sec:req}).

\subsection{Labeling for Queryability}\label{ssec:sci_labels}

As described in Section~\ref{sec:req}, there is a requirement that metadata for 
observations associated with Special Programs is stored and is sufficient for 
triggering real-time data processing recipes.

Additionally, the Special Program science goals require that users are able to 
query and retrieve processed image and catalog data by specifiying the Special Program.

Labelling is also essential for provenance and mini-survey 
characterizability (estimate selection biases), and for Special Programs 
processing by RDM or by users.

Thus \textbf{all visits should have a region label \textit{and} a program 
label.}

\textbf{Region Labels -- } based on Right Acension and Declination.
Region labels should include, e.g., WFD low-dust, Galactic Plane, Galactic 
Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region.
Region labels should be propagated to all processed images (visit, 
difference, and deep coadds) and all cataloged sources and objects.

\textbf{Program Labels -- } based on survey cadeance or strategy.
Program labels might include, e.g., WFD, DDF, mini-survey name, or other 
options like engineering, commissioning, and director's discretionary.
Program labels should be propagated to all single-visit processed and 
difference images, and cataloged sources.

It would not make sense to propagate program labels to deep coadds or 
object tables, as these data products could be a mix of programs.

The existence and propagation of these \textit{specific} labels is not a 
requirement.
This document suggests that these labels be the adopted implementation for 
the metadata requirements (see Section~\ref{ssec:req_meta}).

\subsection{Processing and Data Products}

As described in Section~\ref{sec:req}, there is a requirement that 
Rubin Observatory produce unique, separate, and joinable data products 
whenever this is possible with the original or reconfigured versions of the 
LSST Science Pipelines (and no new algorithmic development or significant 
computational resources are needed).

From a science perspective, however, in some cases where it is \emph{possible} 
to produce such separate data products it is not \emph{necessary}.

While the sections referenced below contain examples, it is ultimately left to 
the discretion of the Rubin Operations Data Managment and System Performance teams to decide which Special Programs data products are generated by RDM.

\textbf{Possible but not necessary -- } 
For some science goals, although it would be possible to produce separate 
data products for a Special Program with the LSST Science Pipelines, this 
is not necessary because the visits can be processed with, and included in, 
the WFD data products (Section~\ref{ssec:proc_wfd}).

These cases are usually related to time domain astronomy, difference image
analysis (DIA), prompt processing, and alert production.
Users would simply incorporate the region and program labels in queries to restrict 
their analysis to only a certain Special Program (or only WFD) where the 
science goal requires it.

\textbf{Possible and necessary -- }
For some science goals, users will need distinct data products composed solely from visits associated with certain Special Programs (Section~\ref{ssec:proc_rdm}).

These cases are usually associated with image coaddition and the static-sky catalogs 
of sources detected and measured in the deeply coadded images.

\textbf{Not possible but necessary -- }
For some science goals, users might require unique and separate data products, 
new algorithms or processing routines, and/or significant additional 
computational resources for Special Programs.
In these situations, user processing and user-generated data products will be 
needed (Section~\ref{ssec:proc_user}).

\subsection{Computational Resources for User Processing}

As mentioned in Section~\ref{ssec:req_ug}, Rubin Observatory will reserve 10\% of its 
total data processing capacity for users.
This component would include {\it all} user processing and re-processing of any and 
all LSST data, including Special Programs.

As described in Section~\ref{ssec:proc_spdp} this processing capacity will be 
accessed via Rubin Science Platform, with a supported software environment and 
infrastructure for batch processing \citedsp{dmtn-202}.

Very computationally intense processing (e.g., shift-and-stack for faint moving 
objects) will likely require external resources\footnote{For more details about the boundary between what Rubin Observatory will provide (in terms of data products and processing resources) and what will be left to the expertise of the science community, see \url{https://www.lsst.org/about/dm/data-products}}.

\subsection{Rubin Science Platform Capabilities}

Users will need to be able to query for data that restricts by sky region and 
program label. 
This can be accomplished by including those labels in all image and catalog 
metadata as described in Section~\ref{ssec:sci_labels}, as the TAP service 
and butler already provide the mechanism for user-specified queries.

Users will need to be able to discover Special Programs data products when 
browsing data, such as all-sky maps.

### 2023-05-20 sec_intro.tex
Update sec_intro.tex

@@
 Special Programs includes non-WFD LSST components such as the Deep Drilling 
-Fields (DDFs) and the mini-, micro-, and nano-surveys; some of these components
+Fields (DDFs) and the mini-, micro-, and nano-surveys, some of which 
 might use non-standard visits with different exposure times
-(see Section~\ref{ssec:intro_terms}, below, for details).
+(Section~\ref{ssec:intro_terms}).
 
@@
 goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
-
 It is a contiguous area that covers at least $\sim$18000 deg$^2$ with 
@@
 of the observing time.
+
 As of the Phase 2 SCOC recommendations in \citeds{PSTN-055}, the WFD area 
 is considered to include the low dust extinction extragalactic regions; 
-the higher extinction, high stellar density Galactic Plane and Bulge; the 
+the higher extinction, high stellar density Galactic Plane (GP) and Bulge; the 
 Virgo Cluster; the South Celestial Pole (SCP); and the North Ecliptic Spur 
@@
 
-In the past, the Galactic, SCP, and NES regions have been referred to as 
-"mini-surveys", and the scheduler software might treat them as such, but 
-this does not make them "Special Programs". 
+In the past the Galactic, SCP, and NES regions have been referred to as 
+``mini-surveys" (see below) but this does not make them "Special Programs". 
 From a survey-scheduling perspective, they are now considered part of the 
-WFD survey.
+WFD survey (whether or not the scheduler software treats them as such).
+
 From a science and data-processing perspective, these components of the 
-WFD do not require special processing or separate data products.
+WFD do not require special processing or separate data products 
+(i.e., do not need to be treated by RDM as ``Special Programs").
+This is because they will use standard (or alternative) visits and can
+be incorporated into WFD Prompt processing (Section~\ref{ssec:proc_wfd}), 
+and because their distinct areas do not overlap with the other WFD components 
+and so do not require, e.g., separate deep coadds.
 
@@
 From a science and data-processing perspective, the DDFs do require 
-special processing or separate data products.
+special processing and separate data products (Section~\ref{ssec:proc_rdm}).
 
@@
 \citeds{PSTN-055}.
-
 This document refers to them all together as mini-surveys.
@@
 other bands; variable exposure times were not recommended for the WFD. 
-
 Non-standard visits with shorter or longer exposure times are being 

### 2023-05-20 sec_sci.tex
Update sec_sci.tex

@@
 
-Recommended implementation aspects for the requirements related to 
-Special Programs (Section~\ref{sec:req}).
+High-level recommendations regarding RDM implementations to meet the 
+requirements related to Special Programs summarized in Section~\ref{sec:req}.
 
@@
 
-Thus \textbf{all visits should have a region label \textit{and} a program 
-label.}
+\textbf{All visits should have a region label \textit{and} a program 
+label, and these labels should be propagated to all processed images and catalog data.}
 
@@
 
-\textbf{Program Labels -- } based on survey cadeance or strategy.
+\textbf{Program Labels -- } 
+Should be based on survey cadeance or strategy.
 Program labels might include, e.g., WFD, DDF, mini-survey name, or other 
@@
 Program labels should be propagated to all single-visit processed and 
-difference images, and cataloged sources.
+difference images, and all cataloged sources.
 
@@
 
-The existence and propagation of these \textit{specific} labels is not a 
-requirement.
+The existence and propagation of these labels is not a requirement.
 This document suggests that these labels be the adopted implementation for 
@@
 While the sections referenced below contain examples, it is ultimately left to 
-the discretion of the Rubin Operations Data Managment and System Performance teams to decide which Special Programs data products are generated by RDM.
+the discretion of the Rubin Operations Data Managment and System Performance teams 
+to decide which Special Programs data products are generated by RDM.
 
@@
 
-As described in Section~\ref{ssec:proc_spdp} this processing capacity will be 
+As described in Section~\ref{ssec:proc_user} this processing capacity will be 
 accessed via Rubin Science Platform, with a supported software environment and 

### 2023-05-20 sec_proc.tex
Update sec_proc.tex

@@
 (or as a part of) the WFD data products (Section~\ref{ssec:proc_wfd}).
-Others will require unique and separate data products 
+Others will require RDM to generate unique and separate data products 
 (Section~\ref{ssec:proc_rdm}).
@@
 development of new algorithms or the allocation of significant additional 
-computational resources are needed, will require user processing 
-(Section~\ref{ssec:proc_user}).
+computational resources are needed, will require users tp generate unique
+and separate data products (Section~\ref{ssec:proc_user}).
 
@@
 significantly brighter sky background (e.g., twilight images) than 
-nighttime survey images.
+standard (or altnerative) WFD main survey visits.
 
@@
 
-None of the currently proposed Special Programs (\citedsp{PSTN-055}) 
+None of the currently proposed Special Programs listed in the Phase 2 
+SCOC recommendations \citedsp{PSTN-055} 
 seem likely to approach (or violate) any harware boundaries 
@@
 
-Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits can be processed and calibrated by the DMS.
+Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits 
+can be processed and calibrated by the DMS.
 
@@
 
-\subsection{RDM Including Special Programs in the WFD Data Products}\label{ssec:proc_wfd}
+\subsection{Inclusion in the WFD Data Products}\label{ssec:proc_wfd}
 
@@
 
-This would include all standard and alternative visits in sky regions for which a template image exists.
-
-This might also include some non-standard visits (shorter or longer exposures) as long as they can be processed by the Prompt 
+This would include all standard and alternative visits in sky regions for which 
+a template image exists.
+This might also include some non-standard visits (shorter or longer exposures) 
+as long as they can be processed by the Prompt 
 pipeline\footnote{If a Special Program's science goals do require 
@@
 
-Difference images, {\tt DiaSources}, and alerts would have region and program labels (Section~\ref{ssec:sci_labels}), providing origin and
+Difference images, {\tt DiaSources}, and alerts would have region and 
+program labels (Section~\ref{ssec:sci_labels}), providing origin and
 provenance for brokers and users.
 
-\textbf{Two potential issues with Prompt processing for DDFs}
-(or any mini-survey that obtains a sequence of images without slewing)
-are:
+\textbf{Two potential issues with Prompt processing for DDFs}, 
+or any mini-survey that obtains a sequence of images without slewing, 
+include:
 
 (1) As alert packets contain the full records of all associated 
-{\tt DIASources} from the past 12 months, alerts for {\tt DIAObjects} in 
+{\tt DiaSources} from the past 12 months, alerts for {\tt DiaObjects} in 
 the DDFs might become very large.
 
-(2) The association of {\it new} {\tt DIASources} into {\tt DIAObjects} 
+(2) The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
 will be somewhat compromised for a DDF sequence.
 The processing for the second image begins when the processing for the 
-first image is only halfway done; when the {\tt DIAObject} catalog has 
-not yet updated with the new {\tt DIASources} detected in the first image.
-Thus, the {\tt DIASource} from images one and two for a new transient 
-would not be associated with a single {\tt DIAObject}, but instead would 
-each instantiate a new {\tt DIAObject}.
+first image is only halfway done; when the {\tt DiaObject} catalog has 
+not yet updated with the new {\tt DiaSources} detected in the first image.
+Thus, the {\tt DiaSource} from images one and two for a new transient 
+would not be associated with a single {\tt DiaObject}, but instead would 
+each instantiate a new {\tt DiaObject}.
 
@@
 
-Since Solar System Processing takes \texttt{DIASources} as input, so any 
+Since Solar System Processing takes {\tt DiaSource}s as input, so any 
 Special Programs images that can be run through the Alert Pipeline can 
@@
 
-\textbf{Deep coadds and the Object catalog -- }
+\textbf{Deep coadds and the {\tt Object} catalog -- }
 Some of the core LSST science goals require a WFD main survey data 
-products of nearly {\it uniform depth}.
+products of nearly uniform depth.
 Whether and how to include any Special Programs data in the deep image 
 coadds and the corresponding {\tt Object} catalog is left entirely to the 
-discretion of the appropriate Rubin teams in Operations.
+discretion of the RDM team in Operations.
 
 For example, perhaps Special Programs images will only be included when 
-they bring additional area to the same depth as the rest of the WFD, or when they suppress edge effects or low-order modes in the 
+they bring additional area to the same depth as the rest of the WFD, or 
+when they suppress edge effects or low-order modes in the 
 all-sky photometric solutions.
@@
 For stacked images and their associated catalogs, the region labels should be
-included but the program labels would not be needed (or possible).
+included (but the program labels would not be needed or possible).
 
@@
 
-\textbf{Custom coadds -- }
-As with the annual WFD deeply coadded images and associated catalogs, RMD would 
+\textbf{A note about intermediate-depth ``custom" coadds -- }
+As with the annual WFD deeply coadded images and associated catalogs, RDM would 
 only generate one coadd (including all images to date) for a given special region.
-Custom coadds (e.g., weekly, monthly) for specific science goals are considered 
+Custom coadds (e.g., weekly, monthly) for very specific science goals are considered 
 user-generated data products, for both WFD and Special Programs.
@@
 left to the discretion of RDM.
-
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.
+Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing 
+examples for some potential Special Programs as further illustration.
 
@@
 \item nightly-coadded difference images (24 h)
-\item {\tt DIASource}- and {\tt DIAObject}-like catalogs for the nightly-coadds (24 h)
+\item {\tt DiaSource}- and {\tt DiaObject}-like catalogs for the nightly-coadds (24 h)
 \item deeply-coadded images (all images to date; yearly)
@@
 For example, a separate repository of processed visit images and coadded 
-images, and separate (but joinable) {\tt Source}, {\tt Forced Source}, and 
+images, and separate (but joinable) {\tt Source}, {\tt ForcedSource}, and 
 {\tt Object} catalogs.
@@
 (e.g., a week spent observing a specific region), such ``annual" 
-data products might be generated on an intermediate time scale, 
+data products might be generated and released on an intermediate time scale, 
 if resources allow.
@@
 
-\textbf{Prompt Processing / Alert Production -- }
+\textbf{A note prompt processing and alert production -- }
 As the latency on processed visit image availability has an 80-hour
@@
 
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples for some potential Special Programs as further illustration.
+Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples 
+for some potential Special Programs as further illustration.

### 2023-05-20 DMTN-065.tex
Update DMTN-065.tex

@@
 
+\smallskip
 Recommendations for the specific implementations needed to fulfill these requirements and maximize science from
@@
 
+\smallskip
 Hardware and processing boundaries on the potential diversity of data from Special Programs 
 (e.g., longer or shorter exposure times), and the ability of the Data Management System to
-process this data, and also provided along with a discussion of when user-generated processing 
+process this data, are also provided along with a discussion of when user-generated processing 
 and data products will be needed to meet some science goals.
 
+\medskip
 The main target audience of this document is Rubin Observatory staff -- the construction- and operations-era Data Management
@@
 \input{sec_intro.tex}
-x
+
 \input{sec_sci.tex}

### 2023-05-20 sec_proc.tex
Update sec_proc.tex

@@
 Rubin Operations Data Production and System Performance teams.
+
+Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing 
+examples for some potential Special Programs as further illustration.
 
@@
 static-sky Data Release data products, or data products for non-standard visits.
+Decisions about when to produce separate Special Programs data products are 
+left to the discretion of RDM.
 
@@
 
-Decisions about when to produce separate Special Programs data products are 
-left to the discretion of RDM.
 Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing 
@@
 
-\subsubsection{Target-of-Opportunity (TOO)}
+\subsubsection{Target-of-Opportunity (TOO) Observations}
 
-Options for RDM to process TOO, especially during the first year of Operations 
+Options for RDM to process TOO observations, especially during the first year of Operations 
 when the template coverage will be low, are discussed in more detail in 
@@
 
-For science goals that require data products which are not possible to create with 
-the original or reconfigured versions of the 
-LSST Science Pipelines, and/or for which new algorithmic development or significant 
-computational resources are needed, 
-user-designed and user-run processing will be needed.
-
+Science goals that require data products which are not possible to create with 
+the original or reconfigured versions of the LSST Science Pipelines, 
+and/or for which new algorithmic development or significant computational resources 
+are needed, will require user processing and user-generated data products.
 As described above, custom coadds (e.g., weekly, monthly) are also left to users 
@@
 
+Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples 
+for some potential Special Programs as further illustration.
+
 \subsubsection{Deep Drilling Fields}
@@
 Short-exposure images obtained during twilight, which will have a very bright sky 
-background unlike other LSST images, might require User-Generated algorithms to 
+background unlike other LSST images, might require user-generated algorithms to 
 subtract the high sky background.
@@
 e.g., those that require difference imaging with coadded images on an intermediate 
-timescale (e.g., a weekly stack), might require user-generated processing.
-
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples 
-for some potential Special Programs as further illustration.
+timescale (e.g., a weekly stack), would require user-generated processing.

### 2023-05-20 sec_sci.tex
Update sec_sci.tex

@@
 From a science perspective, however, in some cases where it is \emph{possible} 
-to produce such separate data products it is not \emph{necessary}.
+to produce such separate data products it is not \emph{necessary}; these
+cases are discussed below.
 
-While the sections referenced below contain examples, it is ultimately left to 
-the discretion of the Rubin Operations Data Managment and System Performance teams 
-to decide which Special Programs data products are generated by RDM.
-
-\textbf{Possible but not necessary -- } 
+\textbf{Case A (possible but not necessary due to inclusion in WFD data products) -- } 
 For some science goals, although it would be possible to produce separate 
@@
 the WFD data products (Section~\ref{ssec:proc_wfd}).
-
 These cases are usually related to time domain astronomy, difference image
@@
 
-\textbf{Possible and necessary -- }
+\textbf{Case B (possible and necessary for RDM to create separate data products) -- }
 For some science goals, users will need distinct data products composed solely from visits associated with certain Special Programs (Section~\ref{ssec:proc_rdm}).
-
 These cases are usually associated with image coaddition and the static-sky catalogs 
@@
 
-\textbf{Not possible but necessary -- }
+\textbf{Case C (not possible, so user-generated data products will be needed) -- }
 For some science goals, users might require unique and separate data products, 
@@
 needed (Section~\ref{ssec:proc_user}).
+These cases are usually related to time domain science goals that utilize 
+intermediate-timescale ``custom" coadds (e.g., weekly, monthly stacks), 
+science goals that require special processing like shift-and-stack techniques, 
+or non-standard visits that are outside the boundaries of what the LSST
+Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
 

### 2023-05-20 ap_spcs.tex
Update ap_spcs.tex

@@
 
-{\bf This section has not been updated since 2018.}
-
-For further insight to the DM-related needs of potential Special Programs, we can write out all of the data acquisition and processing steps, in order, that some of the proposed Special Programs might use.
-Note that we are not including any analysis in these descriptions, only processing and products.
-These are not necessarily complete and may even be incorrect in some places, as we are not experts in the science needs of these potential Special Programs; they could use some more thought and input.
+Step-by-step examples of the processing, either by RDM or users, 
+for data from several Special Programs.
 
 Basic steps that we use to describe a processing case study: \\
-Step 1. Data Acquisition. \\
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Step 3. Delivery of LSST Processed Images. \\
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-Step 6. User-Generated Pipelines and Products. \\
+Step 1. Data acquisition. \\
+Step 2. RDM prompt processing and alert production. \\
+Step 3. RDM special processing with reconfigured pipelines. \\
+Step 4. RDM inclusion in the WFD Data Release data products. \\
+Step 5. User processing and user-generated data products. \\
 
@@
 
-This Special Programs processing summary is based on Becker et al. (2011) white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
+This Special Programs processing summary is based on Becker et al. (2011) 
+white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
 
-Step 1. Data Acquisition. \\
+Step 1. Data acquisition. \\
 The observational sequence is triggered. 
-In a single night, the 9 adjacent fields in a 3x3 grid are observed with $336$ $\times$ $15$ second $r$-band exposures.
+In a single night, the 9 adjacent fields in a 3x3 grid are observed with 
+$336$ $\times$ $15$ second $r$-band exposures. 
 This sequence is always repeated 2-3 nights later. 
@@
 Data obtained in the $g$-band filter is also acceptable.
-\citedsp{Document-11013}
 
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
+Step 2. RDM prompt processing and alert production. \\
+Each $2\times15$ second standard visit is processed by the Prompt pipeline 
+and alerts are released within 60 seconds.
+Within 24 hours, the {\tt DiaSource} and {\tt DiaObject} catalogs are updated
+to include the results of Prompt processing of these visits.
+After 80 hours, the processed visit images and difference images become available.
+All images and sources originating from this Special Program have 
+region and program labels, e.g., ``TNO Survey".
 
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104), but this is not very relevant for this program, which requires a year of dispersed observations before the processing pipelines for SAS can be run.
+Note that the results of prompt processing is not very relevant for this Special Program,
+which requires a year of dispersed observations before the processing pipelines 
+for shift-and-stack can be run, but including them in prompt processing means
+they can contribute to LSST's other time-domain and Solar System science goals.
 
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
+Step 3. RDM special processing with reconfigured pipelines. \\
+None possible. 
 Shift-and-stack processing is beyond the scope of DM's algorithms.
 
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-However, since these images are much deeper than stacks made from the WFD survey, and the strict timing of the observations might lead to their acquisition in sub-optimal conditions, it is unlikely that they would \textit{all} be incorporated.
+Step 4. RDM inclusion in the WFD Data Release data products. \\
+All standard visits are reprocessed and included in the annual Data Release.
+Data products would include processed visit images and {\tt Source} and {\tt ForcedSource}
+catalog rows, along with difference images and {\tt DiaSource} catalog rows.
+All images and sources originating from this Special Program have 
+region and program labels, e.g., ``TNO Survey".
+As with all Special Programs data, these visits might also be included in the 
+WFD deeply coadded images if RDM decides it is both possible and beneficial.
 
-Step 6. User-Generated Pipelines and Products. \\
-The User-Generated pipeline running the shift-and-stack processing will be set up and submitted for batch processing by the user through the Science Platform or on an external processor.
-Pipeline inputs will be the 336 processed exposures per field per re-visit sequence.
-The DRP difference imaging routine will be used with the same template tract/patch for all.
-Custom, User-Generated algorithms will shift the exposures and create difference images, then DRP routines can stack and do source detection and characterization and generate an object database.
-Custom code will derive orbital parameters for the detections and add them to a {\tt SSObjects}-like database.
-
-%\input{spcs_sas}
+Step 5. User processing and user-generated data products. \\
+The user-generated pipeline running the shift-and-stack processing would be set up and submitted 
+for batch processing by the user through the Science Platform or on an external processor. 
+Pipeline inputs would be the 336 processed exposures per field per re-visit sequence. 
+The LSST Science Pipelines difference imaging routine could be used with the same template tract/patch for all. 
+User-generated algorithms will shift the exposures and create difference images, and then the LSST
+Science Pipelines could be used to stack and do source detection and characterization, 
+and generate an object database. 
+Custom code will derive orbital parameters for the detections (or perhaps interface with the MPC),
+and store them in a separate {\tt SSObjects}-like database.
 
@@
 
-Step 1. Data Acquisition. \\
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 visits with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) and a small dither pattern between visits.
+Step 1. Data acquisition. \\
+On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 standard visits 
+with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) 
+and a small dither pattern between visits.
 
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed by the Prompt pipeline's DIA, and Alerts are released within 60 seconds.
-They are flagged to denote the image source is a DDF and that source association might be compromised.
+Step 2. RDM prompt processing and alert production. \\
+As in the example above, but all images and sources originating from this Special Program would have 
+region and program labels of, e.g., ``DDF Name".
 
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+Step 3. RDM special processing with reconfigured pipelines. \\
+The required data products for this science goal can be met by reconfiguring the DM pipelines. 
+First, a template image of appropriate depth for ``nightly" difference imaging 
+would be made using RDM stacking algorithms. 
+On nights when this DDF is observed, at the end of the sequence of observations, 
+reconfigured RDM algorithms would be automatically triggered to create a nightly deep stack, 
+PSF-match it with the appropriately-made template, 
+create a nightly deep difference image, run source detection on it, and update separate 
+{\tt DiaObject}- and {\tt DiaSource}-like catalogs.
+All of this could happen in, e.g., 24 hours, but note there is no requirement on this timescale.
+On an annual basis, deeply coadded images that include all of the DDF's visits to dates would
+be generated, along with a separate {\tt Object}-like catalog for the DDF.
 
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-The required data products for this science goal can be met by reconfiguring the DM pipelines.
-First, a template image for the field will be made using DM stacking algorithms.
-On nights when this DDF is observed, at the end of the sequence of observations, DM algorithms are used to create a nightly deep stack, PSF-match it with the template, create a deep difference image, run source detection on the differences, and create separate databases of \texttt{DIAObject}, \texttt{DIASource}, and \texttt{Object} that are unique to this DDF.
-The LSST codes for alert packet and transport could be used to distribute the detected objects e.g., to the same brokers that receive the Alert Stream, or alternative destinations.
-However, these packets would not be distributed via the LSST {\tt Alert Stream}, and would need to be identified as, e.g., DDF Alerts.
+Step 4. RDM inclusion in the WFD Data Release data products. \\
+As in the example above, but all images and sources originating from this Special Program would have 
+region and program labels of, e.g., ``DDF Name".
 
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-As with all Special Programs data, they might be included in the products of the WFD main survey if DM decides it is beneficial.
-
-Step 6. User-Generated Pipelines and Products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, no separate User-Generated software appears necessary.
+Step 5. User processing and user-generated data products. \\
+For the science goal of searching for supernovae in nightly stacked DDF images, 
+no separate user-generated processing or data products appear to be necessary.
 
@@
 
-Several kinds of twilight surveys with short exposures have been or might be proposed: to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the LSST photometric system and/or to observe the Sweetspot, 60 degrees from the sun, for near-Earth objects.
-The processing case study for these is currently limited by unknowns about the first step: the reduction of processed single visit images.
+Several kinds of twilight surveys with short exposures have been or might be proposed, 
+to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the 
+LSST photometric system and/or to take observations that are particularly well suited 
+(e.g., are 60 degrees from the sun) for finding Near-Earth Objects (NEOs).
+The processing case study for these is currently limited by unknowns.
 
-Step 1. Data Acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. Location and exposure times are set by the sky brightness and desired saturation limits.
+Step 1. Data acquisition. \\
+At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. 
+Location and exposure times are set by the sky brightness and desired saturation limits.
 
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) and scalable processing capabilities to incorporate a faster image-input rate than $1$ every $30$ seconds, these data could {\it potentially} be incorporated and spawn Alerts.
+Step 2. RDM prompt processing and alert production. \\
+Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) 
+these data could {\it potentially} be processed by the Prompt pipeline and lead to alerts.
 
-Step 3. Delivery of LSST Processed Images. \\
-Pending the issues mentioned above, the raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
+Step 3. RDM special processing with reconfigured pipelines. \\
+So long as the short-exposure images can be processed and have enough stars for photometric 
+and astrometric calibration, reconfigured DM pipelines will probably be sufficient for 
+creating image and catalog products from this kind of data.
 
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-This is officially not determined, but so long as the short-exposure images can be processed and have enough stars for photometric and astrometric calibration, reconfigured DM pipelines will probably be sufficient for creating image and catalog products from this kind of data.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
+Step 4. RDM inclusion in the WFD Data Release data products. \\
 These short-exposure, high sky background images would not contribute to the DRP data products created for the WFD survey.
 
-Step 6. User-Generated Pipelines and Products. \\
-If short-exposure images cannot be processed with the existing DM algorithms, a User-Generated processing pipeline might be needed to reduce the raw data.
+Step 5. User processing and user-generated data products. \\
+If short-exposure images cannot be processed with the existing DM algorithms, 
+user-generated processing would be needed to reduce the raw data.
 
-Side note: A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}.
-This would be considered a User-Generated algorithm.
-
-\subsection{The Galactic Plane Survey for Variable Stars and/or Exoplanets}\label{ssec:SPCS_GPVSEx}
-
-Step 1. Data Acquisition. \\
-The schedule incorporates fields in the Galactic Plane, and executes $2\times15$ second visits in these fields (or shorter, for a shallower depth than the WFD main survey).
-
-Step 2. Inclusion in the Prompt Pipeline and Alert Generation. \\
-Each $2\times15$ second visit is processed in the Prompt pipeline and Alerts are released within 60 seconds.
-Extremely crowded fields might have to be skipped if they take longer to process and violate the $60$ second latency for Alerts.
-
-Step 3. Delivery of LSST Processed Images. \\
-The raw, reduced, and calibrated exposures and difference images from the Prompt pipeline are made available within  \texttt{L1PublicT} (currently 24 hours; LSR-REQ-0104).
-
-Step 4. Reconfigured Processing Pipelines and Separate Data Products. \\
-The image and catalog products needed for science with the Galactic Plane are very similar to the products of the Prompt and DRP pipelines, so it seems that not much reconfiguration would be needed.
-The biggest difference might be the incorporation of a user-supplied deblender algorithm optimized for very crowded fields.
-
-Step 5. Inclusion in the DRP Data Products for the WFD Main Survey. \\
-It is quite likely that images from the Galactic Plane will be included into the products of the WFD main survey, as they could e.g., reduce edge effects and help with global photometric classification, but this will depend on deblender performance, and left to the discretion of DM.
-
-Step 6. User-Generated Pipelines and Products. \\
-It seems likely that science users will want to deploy their alternative deblending algorithms on this data set and create their own catalogs.
-
-\subsection{Gravitational Wave Event Follow-Up}\label{ssec:SPCS_GW}
-
-For a description of how target of opportunity data to search for the optical counterparts of gravitational wave events would be processed, see \citeds{rtn-008}.
-
-% \subsection{A More General Level 3 Shift-and-Stack Case Study, by Mario}\label{ssec:SPCS_SAS}
-
-% A Level 3 processing case study for shift-and-stack on a large number images. By Mario.
-
-% \#1.
-The scheduler is configured to repeatedly (e.g., 10 times) observe a field during the same night with longer exposure than usual (e.g., 120 sec).
-[ and we should take the actual numbers from the TNO-DDF whitepaper; don't have the internet right now or I would].
-
-% \#2.
-The images are processed as regular "Level 1" products within 60 seconds, and transmitted as alerts, with results stored into the regular L1 database.
-This will happen automatically for all images (perhaps within some range of exposure times?).
-
-% \#3.
-The raw images (and all necessary calibrations), calexps, and standard L1 diffims are made available within 10 minutes to the batch system for processing with special programs-specific codes.
-This is the same batch system we make available to the users for running Level 3 codes [Q for us: is it?
-or is is the same one that's used to process calibrations?
-have these systems been sized?].
-
-% \#3 a).
-The code running the shift-and-stack processing will be externally developed and delivered, but will be installed and operated (and change controlled!) by the LSST Operations team.
-That is, we don't expect someone external to the ops team to babysit the code on a nightly basis.
-In fact, it's the opposite: once the codes are delivered, any changes will go through LSST's software change control process.
-
-% \#4.
-There will be a facility to trigger program-specific processing on the batch system upon the arrival of a new image (above); this processing will then be queued up for execution.
-We assume that the policy for processing of special programs data may give it preferential treatment relative to general-purpose L3.
-
-% \#5.
-Once the processing finishes, the results of will be stored to a program-specific database.
-No alerts (in VOEvent sense) will be issued.
-We will provide a generic notification facility (perhaps something as simple as an RSS feed) that new data has been made available in a certain database/data store.
-[This is an example where I'd want to make sure somebody within DM is planning to provide such a facility.].
-
-% \#6.
-The outputs stored can be special-program specific (i.e., tables with nearly arbitrary schemas -- some columns -- like ra/dec for spatial joins -- should be present in main tables).
-The outputs can also contain images (stored in also special-program specific repository), or custom products (treated like opaque files).
-The visualizations available for these (catalogs, images, arbitrary files) through the Portal will be limited (e.g., generic table visualizations or x-y plots).
-
-% \#7.
-When the images are made available to the batch system (step \#3), they also become available to *everyone*.
-I.e., someone else could also run a custom L3 pipeline on these data, feeding their custom L3 database.
-(This isn't in the requirements right now -- we say that images will become available in 24hrs -- and is addressed in Section \ref{ssec:dmplans_user}).
+\textbf{Side note:} A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the 
+Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 
+0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform 
+photometry on stars as much as 6 magnitudes brighter than the saturation level"}. 
+This would need user-generated processing.

### 2023-05-20 sec_intro.tex
Update sec_intro.tex

@@
 
-Special Programs is a Rubin Data Management (RDM) term used to refer to survey 
+There are several requirements on the Rubin Data Management System (DMS) 
+related to Special Programs, summarized in Section~\ref{sec:req}.
+This document recommends implementation aspects for Rubin Data Management (RDM)
+with respect to these requirements in order to 
+enable science from Special Programs (Section~\ref{sec:sci}).
+
+The three main messages of this document are:
+\begin{enumerate}
+\item All visits associated with a Special Program should have a region and a program label 
+which is propagated to all processed images, catalog data, and alerts (Section~\ref{ssec:sci_labels}).
+\item All visits that can be processed by the Prompt pipelines and generate 
+alerts should be, in support of time domain and Solar System science goals (Section~\ref{ssec:proc_wfd}).
+\item ``Unique and separate" data products should be generated by RDM whenever possible 
+(e.g., coadded images and {\tt Object} catalogs that overlap, but are deeper or shallower than, the WFD) 
+\end{enumerate}
+
+The following sections provide greater detail and examples in support of these main messages.
+
+\subsection{Terms and Definitions}\label{ssec:intro_terms}
+
+\textbf{Special Programs -- } 
+A Rubin Data Management (RDM) term used to refer to survey 
 components with specific sky regions within or beyond the 
@@
 strategies from the WFD main survey.
-
-There are several requirements on the Rubin Data Management System (DMS) 
-related to Special Programs, summarized in Section~\ref{sec:req}.
-This document recommends implementation aspects for these requirements in order to 
-maximize science from Special Programs (Section~\ref{sec:sci}).
 
@@
 Fields (DDFs) and the mini-, micro-, and nano-surveys, some of which 
-might use non-standard visits with different exposure times
-(Section~\ref{ssec:intro_terms}).
-
-\subsection{Additional Terms and Definitions}\label{ssec:intro_terms}
+might use non-standard visits with different exposure times.
 
@@
 In the past the Galactic, SCP, and NES regions have been referred to as 
-``mini-surveys" (see below) but this does not make them "Special Programs". 
+``mini-surveys" (see below) but this does not make them ``Special Programs". 
 From a survey-scheduling perspective, they are now considered part of the 

### 2023-05-20 sec_intro.tex
Update sec_intro.tex

@@
 \item ``Unique and separate" data products should be generated by RDM whenever possible 
-(e.g., coadded images and {\tt Object} catalogs that overlap, but are deeper or shallower than, the WFD) 
+(e.g., coadded images and {\tt Object} catalogs that overlap, but are deeper or shallower than, the WFD; Section~\ref{ssec:proc_rdm}) 
 \end{enumerate}

### 2023-10-16 DMTN-065.tex
Update DMTN-065.tex
shorten abstract

@@
 \setDocAbstract{
+Special Programs are additional survey areas and/or observing strategies that are driven by specific science 
+goals which build on, or are beyond, the core science pillars of the LSST.
 This document provides a summary of the requirements on Rubin Data Management for processing and serving data products
-related to Special Programs: additional survey areas and/or observing strategies that are driven by specific science 
-goals which build on, or are beyond, the core science pillars of the LSST.
+related to Special Programs.
 
@@
 Special Programs are provided on a case-by-case basis, with examples.
-
-\smallskip
 Hardware and processing boundaries on the potential diversity of data from Special Programs
-(e.g., longer or shorter exposure times), and the ability of the Data Management System to
-process this data, are also provided along with a discussion of when user-generated processing 
-and data products will be needed to meet some science goals.
-
-\medskip
-The main target audience of this document is Rubin Observatory staff -- the construction- and operations-era Data Management
-teams -- but members of the science community who are planning to use Special Programs to reach their science goals may also 
-find this document useful.
+are also provided, along with a discussion of when user-generated processing 
+and data products will be needed to meet Special Programs' science goals.
 }
@@
 \addtohist{2}{2021-12-01}{Updates per DM-20375.}{Melissa Graham}
-\addtohist{3}{2023-05-31}{Updates per DM-32723.}{Melissa Graham}
+\addtohist{3}{2023-10-31}{Updates per DM-32723.}{Melissa Graham}
 }
@@
 % % % % % % % % % % % % % % % % % %
-\clearpage
 \bibliography{local,lsst,refs,books,refs_ads}
@@
 % % % % % % % % % % % % % % % % % %
-\clearpage
 \appendix

### 2023-10-16 sec_intro.tex
update terms

--- 2023-05-20 sec_intro.tex
@@
 
-Special Programs is a Rubin Data Management (RDM) term used to refer to survey 
-components with specific sky regions within or beyond the 
-Wide-Fast-Deep (WFD) main survey, and/or distinct cadences or 
-strategies from the WFD main survey.
+The three memoranda for Rubin Data Management regarding Special Programs are:
+\begin{enumerate}
+\item All visits should have both a region and a program label.
+These labels should be propagated to all derived data products
+(e.g., processed images, catalog data, and alerts; Section~\ref{ssec:sci_labels}).
+\item All visits that \emph{can} be processed by the Prompt pipelines and generate 
+alerts \emph{should} be, in support of time domain and Solar System science goals (Section~\ref{ssec:proc_wfd}).
+\item Special Processing to create ``unique and separate" data products for Special Programs
+should be done when it is possible and necessary, as defined in Section~\ref{ssec:proc_rdm}. 
+\end{enumerate}
 
-There are several requirements on the Rubin Data Management System (DMS) 
-related to Special Programs, summarized in Section~\ref{sec:req}.
-This document recommends implementation aspects for these requirements in order to 
-maximize science from Special Programs (Section~\ref{sec:sci}).
+\subsection{Terms and Definitions}\label{ssec:intro_terms}
 
-Special Programs are driven by specific science goals that build on or 
-add to the core science pillars of the LSST.
-About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
-observations associated with Special Programs (and 85--90\% for WFD).
+\subsubsection{Visit Types}
 
-Special Programs includes non-WFD LSST components such as the Deep Drilling 
-Fields (DDFs) and the mini-, micro-, and nano-surveys, some of which 
-might use non-standard visits with different exposure times
-(Section~\ref{ssec:intro_terms}).
+A visit is an observation of a single pointing at a given time, of which 
+there are three types:
 
-\subsection{Additional Terms and Definitions}\label{ssec:intro_terms}
+\begin{itemize}
+\item Standard Visit -- Composed of $2\times15$ second exposures (commonly referred to as ``snaps").
+\item Alternative Standard Visit -- Composed of a single $30$ second exposure.
+\item Non-Standard Visit -- Any other exposure time(s) or number of snaps.
+\end{itemize}
 
-\textbf{Wide-Fast-Deep (WFD) Main Survey -- }
-The core science program of the LSST, designed to achieve the science 
+Non-standard visits with shorter or longer exposure times are being 
+considered for some Special Programs.
+
+\subsubsection{Wide Fast Deep (WFD) Main Survey}
+
+The WFD program is the core science program of the LSST, designed to achieve the science 
 goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
-It is a contiguous area that covers at least $\sim$18000 deg$^2$ with 
+
+The WFD is defined from a scheduling perspective as a certain cadence and filter balance 
+(i.e., an observing strategy; a program) 
+that covers the low-dust region (extragalactic areas) and several special regions:
+the Galactic bulge and plane; the South Celestial Pole (SCP); the North Ecliptic Spur (NES);
+and the Virgo Cluster \citedsp{PSTN-055}.
+
+The WFD's regions will compose a contiguous sky footprint that covers at least $\sim$18000 deg$^2$ with 
 $\gtrsim$800 visits per field, expected to be accomplished with 85--90\% 
@@
 
-As of the Phase 2 SCOC recommendations in \citeds{PSTN-055}, the WFD area 
-is considered to include the low dust extinction extragalactic regions; 
-the higher extinction, high stellar density Galactic Plane (GP) and Bulge; the 
-Virgo Cluster; the South Celestial Pole (SCP); and the North Ecliptic Spur 
-(NES).
+The WFD is expected to use alternative standard visits in the $u$-band and 
+standard visits in all other bands \citedsp{PSTN-055}.
+Rolling cadence is expected to be implemented in the low-dust region only.
 
-In the past the Galactic, SCP, and NES regions have been referred to as 
-``mini-surveys" (see below) but this does not make them "Special Programs". 
-From a survey-scheduling perspective, they are now considered part of the 
-WFD survey (whether or not the scheduler software treats them as such).
+\subsubsection{Regular (non-special) processing}
 
-From a science and data-processing perspective, these components of the 
-WFD do not require special processing or separate data products 
-(i.e., do not need to be treated by RDM as ``Special Programs").
-This is because they will use standard (or alternative) visits and can
-be incorporated into WFD Prompt processing (Section~\ref{ssec:proc_wfd}), 
-and because their distinct areas do not overlap with the other WFD components 
-and so do not require, e.g., separate deep coadds.
+The image processing and data products described in the Data Products Definitions Document
+\citedsp{LSE-163} that are designed for, and will be applied to, the WFD program's observations.
 
-\textbf{Deep Drilling Field (DDF) -- }
+In some cases, regular processing is also appropriate for Special Programs.
+
+\subsubsection{Special Programs}
+
+This is a Rubin Data Management (RDM) term used to refer to sky regions within
+or beyond the WFD footprint which have a distinct observing strategy
+(e.g., visit type, cadence, filter distribution).
+
+Special Programs are typically driven by specific science goals that build on or 
+add to the core science pillars of the LSST.
+About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
+observations associated with Special Programs.
+Special Programs includes non-WFD LSST components such as the Deep Drilling 
+Fields (DDFs) and the mini-, micro-, and nano-surveys \citedsp{PSTN-055}.
+
+Some science goals for Special Programs can be met with regular (non-special) processing,
+but some will require Special Processing by Rubin Data Management or user-generated processing.
+
+\subsubsection{Special Processing}
+
+The data processing pipelines, built from the existing LSST Science Pipelines, which are
+applied by Rubin Data Management to observations from Special Programs.
+Special Processing creates data products that are unique and separate from those produced
+by regular processing for the WFD program.
+
+Special Processing is likely to use special inputs, configurations, 
+or timescales, as appropriate for the Special Programs' data and science goals.
+However, the development and application of specialized \emph{algorithms} or new software
+is beyond the scope of Special Processing by Rubin Data Management.
+
+\subsubsection{User-Generated Processing}
+
+Any processing of Rubin data done by users in order to reach specific science goals, including
+processing for Special Programs data, is referred to as user-generated processing.
+
+Guidelines for user-generated processing that can be run in the 
+Rubin Science Platform, and for user-generated data products that can
+be federated with the Rubin-product data products (i.e., joinable tables),
+is forthcoming.
+
+\subsubsection{Deep Drilling Field (DDF)}
+
 A single pointing for which many (e.g., a hundred) visits are obtained 
@@
 
-From a science and data-processing perspective, the DDFs do require 
-special processing and separate data products (Section~\ref{ssec:proc_rdm}).
+The DDFs will require Special Processing and unique and separate data products (Section~\ref{ssec:proc_rdm}).
 
-\textbf{Mini-, Micro- and Nano-Surveys -- }
-Specific sky areas covered by a few hundred, a hundred, or tens of visits 
-(respectively; as in Phase 2 SCOC recommendations, \citeds{PSTN-055}).
-For a list of the current mini-surveys under consideration, see 
-\citeds{PSTN-055}.
-This document refers to them all together as mini-surveys.
+\subsubsection{Mini-, Micro- and Nano-Surveys}
 
-These sky areas are all (so far) either within or adjacent to the WFD area.
-In some cases these visits are non-standard (e.g., short exposures).
+Specific sky areas covered by a few hundred, a hundred, or tens of visits (respectively).
+This document will refer to them collectively as mini-surveys.
 
-From a data-processing and science perspective, most mini-surveys are 
-likely to require special processing or separate data products, and should 
-be considered on a case-by-base basis.
+The sky areas of mini-surveys can be within, adjacent to, or detached from the WFD footprint.
+Mini-surveys can have non-standard visits.
 
-\textbf{Visit Types -- }
-A visit is an observation of a single pointing at a given time, of which 
-there are three types:
+For a list of the mini-surveys under consideration,
+see \citeds{PSTN-055}).
 
-\begin{itemize}
-\item Standard Visit -- Composed of $2\times15$ second exposures (commonly referred to as ``snaps").
-\item Alternative Standard Visit -- Composed of a single $30$ second exposure.
-\item Non-Standard Visit -- Any other exposure time(s) or number of snaps.
-\end{itemize}
-
-The Phase 2 SCOC recommendations for the WFD (\citeds{PSTN-055}) are to 
-use alternative standard visits in the $u$-band and standard visits in all 
-other bands; variable exposure times were not recommended for the WFD. 
-Non-standard visits with shorter or longer exposure times are being 
-considered for some mini-surveys.
+Most mini-surveys are likely to require Special Processing and unique and separate data products,
+and should be considered on a case-by-base basis.
+Some mini-surveys will likely require user-generated processing.

### 2023-10-16 sec_intro.tex
remove cite to lse-163, fix later

@@
 The image processing and data products described in the Data Products Definitions Document
-\citedsp{LSE-163} that are designed for, and will be applied to, the WFD program's observations.
+(LSE-163) that are designed for, and will be applied to, the WFD program's observations.
 

### 2023-10-16 sec_intro.tex
minor updates

@@
 \section{Introduction} \label{sec:intro}
+
+\subsection{Takeaways for Rubin Data Management}
 
@@
 \item All visits should have both a region and a program label.
-These labels should be propagated to all derived data products
-(e.g., processed images, catalog data, and alerts; Section~\ref{ssec:sci_labels}).
+These labels should be propagated to all derived data products, such as
+processed images, catalog data, and alerts. (Section~\ref{ssec:sci_labels})
 \item All visits that \emph{can} be processed by the Prompt pipelines and generate 
-alerts \emph{should} be, in support of time domain and Solar System science goals (Section~\ref{ssec:proc_wfd}).
+alerts \emph{should} be, in support of time domain and Solar System science goals. (Section~\ref{ssec:proc_wfd})
 \item Special Processing to create ``unique and separate" data products for Special Programs
@@
 
-\subsubsection{Wide Fast Deep (WFD) Main Survey}
+\subsubsection{Wide Fast Deep (WFD) Program}
 
-The WFD program is the core science program of the LSST, designed to achieve the science 
+The WFD is the core science program of the LSST, designed to achieve the science 
 goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
 
-The WFD is defined from a scheduling perspective as a certain cadence and filter balance 
-(i.e., an observing strategy; a program) 
-that covers the low-dust region (extragalactic areas) and several special regions:
+The WFD program is defined from a scheduling perspective as a certain observing strategy
+(i.e., a set of cadences and filter balances) applied to the low-dust region 
+(extragalactic areas) and several special regions:
 the Galactic bulge and plane; the South Celestial Pole (SCP); the North Ecliptic Spur (NES);
@@
 
-The WFD's regions will compose a contiguous sky footprint that covers at least $\sim$18000 deg$^2$ with 
-$\gtrsim$800 visits per field, expected to be accomplished with 85--90\% 
-of the observing time.
+The WFD's regions will compose a contiguous sky footprint that covers at 
+least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, 
+and is expected to be accomplished with 85--90\% of the observing time.
 
@@
 standard visits in all other bands \citedsp{PSTN-055}.
-Rolling cadence is expected to be implemented in the low-dust region only.
+Different cadences, filter balances, rotational dithers, etc., might
+be applied in the different regions composing WFD.
+For example, rolling cadence is expected to be implemented in the low-dust region only.
+WFD program data will be processed with regular (non-special) processing.
 

### 2023-10-17 sec_intro.tex
rearrange, rewrite

@@
-\section{Introduction} \label{sec:intro}
-
-\subsection{Takeaways for Rubin Data Management}
+\section{Executive Summary} \label{sec:execsum}
 
 The three memoranda for Rubin Data Management regarding Special Programs are:
+
 \begin{enumerate}
+
 \item All visits should have both a region and a program label.
@@
 processed images, catalog data, and alerts. (Section~\ref{ssec:sci_labels})
+
 \item All visits that \emph{can} be processed by the Prompt pipelines and generate 
 alerts \emph{should} be, in support of time domain and Solar System science goals. (Section~\ref{ssec:proc_wfd})
+
 \item Special Processing to create ``unique and separate" data products for Special Programs
 should be done when it is possible and necessary, as defined in Section~\ref{ssec:proc_rdm}.
+
 \end{enumerate}
 
-\subsection{Terms and Definitions}\label{ssec:intro_terms}
+\section{Introduction}\label{sec:intro}
 
-\subsubsection{Visit Types}
+The following is an introduction to the key terms related to Special Programs.
+
+\subsection{Visit Types}
 
@@
 
-\subsubsection{Wide Fast Deep (WFD) Program}
+\subsection{Wide Fast Deep (WFD) Program}
 
@@
 The WFD program is defined from a scheduling perspective as a certain observing strategy
-(i.e., a set of cadences and filter balances) applied to the low-dust region 
+(i.e., a set of cadences and filter balances) applied to a low-dust region 
 (extragalactic areas) and several special regions:
@@
 and the Virgo Cluster \citedsp{PSTN-055}.
-
-The WFD's regions will compose a contiguous sky footprint that covers at 
-least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, 
-and is expected to be accomplished with 85--90\% of the observing time.
 
@@
 For example, rolling cadence is expected to be implemented in the low-dust region only.
+
 WFD program data will be processed with regular (non-special) processing.
+This processing will produce, among other data products (LSE-163), 
+a contiguous sky footprint that covers at 
+least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, 
+and is expected to be accomplished with 85--90\% of the observing time.
+This contiguous all-sky "deep coadd" may have region-dependent processing parameters and inputs
+(e.g., different calibration parameters or deblending algorithms for crowded fields)
+and will be of variable depth, as most special regions recieve fewer visits.
 
-\subsubsection{Regular (non-special) processing}
+\subsection{Regular (non-special) processing}
 
-The image processing and data products described in the Data Products Definitions Document
+This is a term used only in this document to refer to the image processing 
+described in the Data Products Definitions Document
 (LSE-163) that are designed for, and will be applied to, the WFD program's observations.
@@
 
-\subsubsection{Special Programs}
+\subsection{Special Programs}
 
-This is a Rubin Data Management (RDM) term used to refer to sky regions within
-or beyond the WFD footprint which have a distinct observing strategy
-(e.g., visit type, cadence, filter distribution).
+This is a Rubin Data Management term used to refer to sky regions within
+or beyond the WFD footprint which have a distinct observing strategy from
+the WFD program's plans for that region (e.g., different visit type, cadence, filter distribution).
 
@@
 add to the core science pillars of the LSST.
+
 About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
@@
 
-Some science goals for Special Programs can be met with regular (non-special) processing,
+Data products from Special Programs is subject to the Rubin Data Policy \citedsp{RDO-013}
+in the same way as data products from the WFD program.
+
+Some science goals for Special Programs can be met with regular processing,
 but some will require Special Processing by Rubin Data Management or user-generated processing.
 
-\subsubsection{Special Processing}
+\subsection{Special Processing}
 
-The data processing pipelines, built from the existing LSST Science Pipelines, which are
-applied by Rubin Data Management to observations from Special Programs.
+This is a Rubin Data Management term used used to describe processing that 
+uses components of the LSST Science Pipelines and is applied by Rubin
+Data Management to images from Special Programs.
 Special Processing creates data products that are unique and separate from those produced
@@
 
-Special Processing is likely to use special inputs, configurations, 
-or timescales, as appropriate for the Special Programs' data and science goals.
+Special Processing is likely to use different inputs or configurations for the
+LSST Science Pipelines, or to run on different timescales, than regular processing - 
+as appropriate for the Special Programs' data and science goals.
 However, the development and application of specialized \emph{algorithms} or new software
-is beyond the scope of Special Processing by Rubin Data Management.
+is beyond the scope of Special Processing.
 
-\subsubsection{User-Generated Processing}
+\subsection{User-Generated Processing}
 
 Any processing of Rubin data done by users in order to reach specific science goals, including
-processing for Special Programs data, is referred to as user-generated processing.
+processing for Special Programs data, is referred to as User-Generated Processing.
 
-Guidelines for user-generated processing that can be run in the 
-Rubin Science Platform, and for user-generated data products that can
-be federated with the Rubin-product data products (i.e., joinable tables),
+User-Generated Processing for Special Programs data would be necessary in cases where
+the science goals require custom algorithms, software, or very large computational
+capacities which are beyond the scope of Special Processing or the Rubin-provided
+computational resources (Section~\ref{ssec:sci_comp}).
+
+Guidelines for User-Generated Processing, and for user-generated data products
+that can be federated with the Rubin-product data products (i.e., joinable tables),
 is forthcoming.
 
-\subsubsection{Deep Drilling Field (DDF)}
+\subsection{Deep Drilling Field (DDF)}
 
@@
 
-The DDFs will require Special Processing and unique and separate data products (Section~\ref{ssec:proc_rdm}).
+The DDFs will require Special Processing (Section~\ref{ssec:proc_rdm})
+and may also benefit from User-Generated Processing.
 
-\subsubsection{Mini-, Micro- and Nano-Surveys}
+\subsection{Mini-, Micro- and Nano-Surveys}
 
@@
 
-Most mini-surveys are likely to require Special Processing and unique and separate data products,
-and should be considered on a case-by-base basis.
-Some mini-surveys will likely require user-generated processing.
+Most mini-surveys are likely to require Special Processing (to be considered on a case-by-case basis),
+and some are also likely to require User-Generated Processing.

### 2023-10-17 sec_intro.tex
fix typos

@@
 
-This is a Rubin Data Management term used used to describe processing that 
+This is a Rubin Data Management term to describe processing that 
 uses components of the LSST Science Pipelines and is applied by Rubin
@@
 For a list of the mini-surveys under consideration,
-see \citeds{PSTN-055}).
+see \citeds{PSTN-055}.
 

### 2023-10-17 sec_sci.tex
Update sec_sci.tex
partial updates

@@
-\section{Maximizing Science from Special Programs}\label{sec:sci}
+\section{Recommended Implementations to Enable Science from Special Programs}\label{sec:sci}
 
-High-level recommendations regarding RDM implementations to meet the 
-requirements related to Special Programs summarized in Section~\ref{sec:req}.
+Recommendations regarding implementations by Rubin Data Management to meet the 
+requirements related to Special Programs summarized in Section~\ref{sec:req},
+which will enable science with data from Special Programs.
 
-\subsection{Labeling for Queryability}\label{ssec:sci_labels}
-
-As described in Section~\ref{sec:req}, there is a requirement that metadata for 
-observations associated with Special Programs is stored and is sufficient for 
-triggering real-time data processing recipes.
-
-Additionally, the Special Program science goals require that users are able to 
-query and retrieve processed image and catalog data by specifiying the Special Program.
-
-Labelling is also essential for provenance and mini-survey 
-characterizability (estimate selection biases), and for Special Programs 
-processing by RDM or by users.
+\subsection{Region and Program Labels}\label{ssec:sci_labels}
 
 \textbf{All visits should have a region label \textit{and} a program 
-label, and these labels should be propagated to all processed images and catalog data.}
+label, and these labels should be propagated to processed images and catalog data.}
 
@@
 Region labels should include, e.g., WFD low-dust, Galactic Plane, Galactic 
-Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region.
-Region labels should be propagated to all processed images (visit, 
-difference, and deep coadds) and all cataloged sources and objects.
+Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
+Region labels should be propagated to processed images (visit, 
+difference, and deep coadds), catalog sources, and catalog static-sky objects.
 
 \textbf{Program Labels -- } 
-Should be based on survey cadeance or strategy.
+Based on survey cadeance or strategy.
 Program labels might include, e.g., WFD, DDF, mini-survey name, or other 
 options like engineering, commissioning, and director's discretionary.
-Program labels should be propagated to all single-visit processed and 
-difference images, and all cataloged sources.
+Program labels should be propagated to single-visit processed and 
+difference images and catalog sources.
 
@@
 
-The existence and propagation of these labels is not a requirement.
-This document suggests that these labels be the adopted implementation for 
-the metadata requirements (see Section~\ref{ssec:req_meta}).
+The existence and propagation of these labels is not itself a requirement.
+It is recommended for implementation for three reasons.
 
-\subsection{Processing and Data Products}
+\begin{enumerate}
 
-As described in Section~\ref{sec:req}, there is a requirement that 
-Rubin Observatory produce unique, separate, and joinable data products 
-whenever this is possible with the original or reconfigured versions of the 
-LSST Science Pipelines (and no new algorithmic development or significant 
-computational resources are needed).
+\item To meet the requirement that metadata for observations associated 
+with Special Programs is stored, and is sufficient for triggering 
+real-time data processing recipes (Section~\ref{ssec:req_meta}).
 
-From a science perspective, however, in some cases where it is \emph{possible} 
-to produce such separate data products it is not \emph{necessary}; these
-cases are discussed below.
+\item To enable users to query and retrieve processed image and catalog
+data associated with a specific Special Program, and meet the science goals
+of that Special Program, when regular processing as been applied 
+(e.g., Prompt Processing, Section~\ref{ssec:sci_pproc}).
 
-\textbf{Case A (possible but not necessary due to inclusion in WFD data products) -- } 
-For some science goals, although it would be possible to produce separate 
-data products for a Special Program with the LSST Science Pipelines, this 
-is not necessary because the visits can be processed with, and included in, 
-the WFD data products (Section~\ref{ssec:proc_wfd}).
-These cases are usually related to time domain astronomy, difference image
-analysis (DIA), prompt processing, and alert production.
-Users would simply incorporate the region and program labels in queries to restrict 
-their analysis to only a certain Special Program (or only WFD) where the 
-science goal requires it.
+\item To enable provenance when Special Programs data in included in regular
+processing and the WFD program data products (e.g., if used to improve the
+all-sky coadd).
+
+\end{enumerate}
+
+\subsection{Special Processing}\label{ssec:sci_sproc}
+
+\textbf{Special Processing should be done by Rubin Data Management to 
+produce unique and separate (but joinable) data products
+for Special Programs when it is both possible and necessary.}
+
+\textbf{Possible -- } When original or reconfigured versions of the LSST
+Science Pipelines can be used, and no new algorithmic or software development,
+or significant additional computational resources, are needed.
+
+\textbf{Necessary -- } When the science goals for a Special Program cannot
+be met by only including the data in regular processing (e.g., Prompt processing).
+
+\begin{enumerate}
+
+\item To meet the requirement that Rubin Observatory produce
+unique, separate, and joinable data products whenever this is possible 
+with the original or reconfigured versions of the LSST Science Pipelines
+(Section~\ref{sec:req}).
+
+\item To enable science with Special Programs by all users, not just those
+with the time and effort to process the data, and to reduce computational
+load (and potential redundancy) in User-Generated Processing.
+
+\end{enumerate}
+
+To further illustrate this recommendation, a few examples of Special Processing
+cases are provided below.
+
+\begin{itemize}
+
+\item \textbf{Possible and necessary:}
+
+\item \textbf{Possible but not necessary:}
+a time-domain mini-survey that uses standard visits \emph{could}
+have separate difference-image analysis object and source catalogs
+generated, but this is not necessary as the science goals for the
+mini-survey can be met by processing its data with regular Prompt
+Processing (Section~\ref{ssec:sci_pproc}), and ensuring the
+data is properly labeled (Section~\ref{ssec:sci_labels}).
+
+\item \textbf{Necessary but not possible:}
+
+\end{itemize}
+
+RESTART EDITING PROCESS HERE
 
@@
 
-\subsection{Computational Resources for User Processing}
+\subsection{Prompt Processing}\label{ssec:sci_pproc}
+
+\textbf{Special Processing should produce unique and separate data products
+for Special Programs when it is both possible and necessary.}
+
+\subsection{Computational Resources for User-Generated Processing}\label{ssec:sci_comp}
 
@@
 Very computationally intense processing (e.g., shift-and-stack for faint moving 
-objects) will likely require external resources\footnote{For more details about the boundary between what Rubin Observatory will provide (in terms of data products and processing resources) and what will be left to the expertise of the science community, see \url{https://www.lsst.org/about/dm/data-products}}.
+objects) will likely require external resources\footnote{For more details about 
+the boundary between what Rubin Observatory will provide (in terms of data products 
+and processing resources) and what will be left to the expertise of the science community, 
+see \url{https://www.lsst.org/about/dm/data-products}}.
 
-\subsection{Rubin Science Platform Capabilities}
+\subsection{Rubin Science Platform Capabilities}\label{ssec:sci_rsp}
 

### 2023-10-18 sec_sci.tex
updates

@@
 
-\textbf{Necessary -- } When the science goals for a Special Program cannot
-be met by only including the data in regular processing (e.g., Prompt processing).
+\textbf{Necessary -- } When the primary science goal for a Special Program 
+cannot be met by including the data in regular processing (e.g., Prompt processing),
+or where doing so would compromise the WFD data products (e.g., introduce non-uniformity).
+
+The definitions of possible and necessary are ultimately left to the
+discretion of Rubin Data Management in Operations.
+
+There are two main reasons for this recommendation.
 
@@
 
-To further illustrate this recommendation, a few examples of Special Processing
-cases are provided below.
+Specific examples are provided only to illustrate this recommendation,
+not to place limits on what Special Processing might be done.
+Note in particular that secondary science goals may be considered
+to require User-Generated Processing.
 
@@
 \item \textbf{Possible and necessary:}
+in order to detect high-redshift (faint) galaxies in the DDFs,
+Rubin Data Management uses the LSST Science Pipelines to deeply
+coadd images for each field, and store the results of source
+detection and characterization in unique and separate tables that
+are included in the annual data release (Section~\ref{ssec:proc_rdm_ddf}).
 
@@
 
+\item \textbf{Possible but not necessary (secondary science goals):}
+a time-domain mini-survey (or DDF) has a secondary science goal of detecting
+precursor outbursts for transients, which requires coadding images
+within windows of days, weeks, and months to reach various depths.
+This set of custom coadds may be considered as "not necessary" and requiring 
+User-Generated Processing.
+
 \item \textbf{Necessary but not possible:}
+in order to find the most distant, faint Kuiper Belt Objects in the DDF,
+a specialized, computationally intensive form of "shift-and-stack" processing
+is required for detection, but such algorithms are not used by the 
+LSST Science Pipelines and so User-Generated Processing will be needed (Section~\ref{ssec:SPCS_TNO}).
+A second example here is a twilight survey that uses non-standard visits 
+which are outside the boundaries of what the LSST
+Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
 
@@
 
-RESTART EDITING PROCESS HERE
-
-\textbf{Case B (possible and necessary for RDM to create separate data products) -- }
-For some science goals, users will need distinct data products composed solely from visits associated with certain Special Programs (Section~\ref{ssec:proc_rdm}).
-These cases are usually associated with image coaddition and the static-sky catalogs 
-of sources detected and measured in the deeply coadded images.
-
-\textbf{Case C (not possible, so user-generated data products will be needed) -- }
-For some science goals, users might require unique and separate data products, 
-new algorithms or processing routines, and/or significant additional 
-computational resources for Special Programs.
-In these situations, user processing and user-generated data products will be 
-needed (Section~\ref{ssec:proc_user}).
-These cases are usually related to time domain science goals that utilize 
-intermediate-timescale ``custom" coadds (e.g., weekly, monthly stacks), 
-science goals that require special processing like shift-and-stack techniques, 
-or non-standard visits that are outside the boundaries of what the LSST
-Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
-
 \subsection{Prompt Processing}\label{ssec:sci_pproc}
 
-\textbf{Special Processing should produce unique and separate data products
-for Special Programs when it is both possible and necessary.}
+\textbf{All visits that \emph{can} be processed by the Prompt pipelines and generate 
+alerts \emph{should} be, in support of time domain and Solar System science goals.}
+
+The phrase ``can be processed" includes all standard and alternative visits in 
+sky regions for which a template image exists.
+This might also include some non-standard visits (shorter or longer exposures), 
+as long as they can be processed by the Prompt pipeline and an appropriate template image exists.
+
+Visits with very short or very long exposure times (or very bright sky 
+backgrounds) might be excluded if they would need specialized algorithms for,
+e.g., instrument signature removal, difference-imaging, template-generation 
+(Section~\ref{ssec:proc_bounds}).
+
+If a Special Program's primary science goal requires specialized templates and 
+Prompt processing, the Data Management System will have the capability to load 
+and use an alternative template for some sky regions, based on the image metadata 
+(i.e., the labels described in Section~\ref{ssec:sci_labels}). 
+However, there would not be enough memory to hold alternative templates 
+for the whole sky.
+
+Including visits from Special Programs in regular Prompt processing alongside
+vists from the WFD program is not, in general, anticipated to affect WFD science goals.
+For example, analyses for a WFD-only subset could still be done using the program and
+region labels described in Section~\ref{ssec:sci_labels}.
+However, a few concerns are raised in Section~\ref{ssec:proc_prompt}.
 
@@
 
-As mentioned in Section~\ref{ssec:req_ug}, Rubin Observatory will reserve 10\% of its 
-total data processing capacity for users.
+Rubin Observatory will reserve 10\% of its total data processing capacity for users.
+This is already a requirement, as mentioned in Section~\ref{ssec:req_ug}.
+
 This component would include {\it all} user processing and re-processing of any and 

### 2023-10-18 sec_intro.tex
Update sec_intro.tex

@@
 
-\item All visits should have both a region and a program label.
-These labels should be propagated to all derived data products, such as
-processed images, catalog data, and alerts. (Section~\ref{ssec:sci_labels})
+\item All visits should have a region label \textit{and} a program 
+label, and these labels should be propagated to processed images and catalog data.
+(Section~\ref{ssec:sci_labels})
+
+\item Special Processing should be done by Rubin Data Management to 
+produce unique and separate (but joinable) data products
+for Special Programs when it is both possible and necessary.
+(Section~\ref{ssec:sci_sproc}).
 
 \item All visits that \emph{can} be processed by the Prompt pipelines and generate 
-alerts \emph{should} be, in support of time domain and Solar System science goals. (Section~\ref{ssec:proc_wfd})
-
-\item Special Processing to create ``unique and separate" data products for Special Programs
-should be done when it is possible and necessary, as defined in Section~\ref{ssec:proc_rdm}.
+alerts \emph{should} be, in support of time domain and Solar System science goals.
+(Section~\ref{ssec:sci_pproc})
 
@@
 add to the core science pillars of the LSST.
-
-About 10--15\% of the total 10-year LSST time-span will be spent obtaining 
+They include non-WFD LSST components such as the Deep Drilling 
+Fields (DDFs) and the mini-, micro-, and nano-surveys \citedsp{PSTN-055}.
+About 10--15\% of the total 10-year LSST will be spent obtaining 
 observations associated with Special Programs.
-Special Programs includes non-WFD LSST components such as the Deep Drilling 
-Fields (DDFs) and the mini-, micro-, and nano-surveys \citedsp{PSTN-055}.
 
@@
 
-The DDFs will require Special Processing (Section~\ref{ssec:proc_rdm})
-and may also benefit from User-Generated Processing.
+Processing for the DDF images is likely to be a combination of regular (Prompt) processing,
+Special Processing, and User-Generated Processing.
 
@@
 Mini-surveys can have non-standard visits.
-
+Target-of-opportunity observations for, e.g., the discovery of optical counterparts to 
+multi-messenger astrophysical phenomena, are considered a type of mini-survey in this document.
 For a list of the mini-surveys under consideration,
@@
 
-Most mini-surveys are likely to require Special Processing (to be considered on a case-by-case basis),
-and some are also likely to require User-Generated Processing.
+Processing for the mini-surveys is likely to be a combination of regular (Prompt) processing,
+Special Processing, and User-Generated Processing.

### 2023-10-18 sec_intro.tex
add TOO acronym

@@
 Mini-surveys can have non-standard visits.
-Target-of-opportunity observations for, e.g., the discovery of optical counterparts to 
+Target-of-opportunity (TOO) observations for, e.g., the discovery of optical counterparts to 
 multi-messenger astrophysical phenomena, are considered a type of mini-survey in this document.

### 2023-10-18 sec_sci.tex
minor updates

@@
-\section{Recommended Implementations to Enable Science from Special Programs}\label{sec:sci}
+\section{Recommendations to Enable Science from Special Programs}\label{sec:sci}
 
-Recommendations regarding implementations by Rubin Data Management to meet the 
-requirements related to Special Programs summarized in Section~\ref{sec:req},
-which will enable science with data from Special Programs.
+Recommended implementations for Rubin Data Management to meet the 
+requirements related to Special Programs (Section~\ref{sec:req})
+and enable science with data from Special Programs.
 
@@
 Based on survey cadeance or strategy.
-Program labels might include, e.g., WFD, DDF, mini-survey name, or other 
+Program labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
 options like engineering, commissioning, and director's discretionary.
@@
 difference images and catalog sources.
-
 It would not make sense to propagate program labels to deep coadds or 
@@
 
-The existence and propagation of these labels is not itself a requirement.
-It is recommended for implementation for three reasons.
+There are three main reasons for this recommendation.
 
@@
 
-The definitions of possible and necessary are ultimately left to the
-discretion of Rubin Data Management in Operations.
+The definitions of possible and necessary, and the scope of Special Processing,
+are ultimately left to the discretion of Rubin Data Management in Operations.
+Note that secondary science goals may be considered as ``not necessary".
 
@@
 
-Specific examples are provided only to illustrate this recommendation,
-not to place limits on what Special Processing might be done.
-Note in particular that secondary science goals may be considered
-to require User-Generated Processing.
+Specific examples are provided only to illustrate this recommendation.
+These examples do not place limits on what Special Processing might be done.
 
@@
 within windows of days, weeks, and months to reach various depths.
-This set of custom coadds may be considered as "not necessary" and requiring 
+This set of custom coadds may be considered as ``not necessary" and requiring 
 User-Generated Processing.
@@
 in order to find the most distant, faint Kuiper Belt Objects in the DDF,
-a specialized, computationally intensive form of "shift-and-stack" processing
+a specialized, computationally intensive form of ``shift-and-stack" processing
 is required for detection, but such algorithms are not used by the 
@@
 This is already a requirement, as mentioned in Section~\ref{ssec:req_ug}.
-
 This component would include {\it all} user processing and re-processing of any and 
@@
 Users will need to be able to discover Special Programs data products when 
-browsing data, such as all-sky maps.
+browsing data.
+This could be accomplished with, e.g., a toggle to overlay region boundaries
+for Special Programs when browsing all-sky coadds built from WFD program data
+in the Portal Aspect.

### 2023-10-18 sec_proc.tex
update boundaries

@@
-\section{Processing for Special Programs Data Products}\label{sec:proc}
+\section{Processing Data from Special Programs}\label{sec:proc}
 
-Rubin Observatory and the LSST Data Management System (DMS) are required 
-to process Special Programs data to produce unique and separate data 
-products whenever the original or reconfigured versions of the LSST 
-Science Pipelines can be used to do so (Section~\ref{sec:req}).
+% Rubin Observatory and the LSST Data Management System (DMS) are required 
+% to process Special Programs data to produce unique and separate data 
+% products whenever the original or reconfigured versions of the LSST 
+% Science Pipelines can be used to do so (Section~\ref{sec:req}).
 
-The LSST Science Pipelines and the DMS are designed to process standard 
-(or alternative standard) visits.
+% The LSST Science Pipelines and the DMS are designed to process standard 
+% (or alternative standard) visits.
 
-It depends on the science goals related to the Special Program, but some 
-Special Programs' data can and should be processed and served along with 
-(or as a part of) the WFD data products (Section~\ref{ssec:proc_wfd}).
-Others will require RDM to generate unique and separate data products 
-(Section~\ref{ssec:proc_rdm}).
+% It depends on the science goals related to the Special Program, but some 
+% Special Programs' data can and should be processed and served along with 
+% (or as a part of) the WFD data products (Section~\ref{ssec:proc_wfd}).
+% Others will require RDM to generate unique and separate data products 
+% (Section~\ref{ssec:proc_rdm}).
 
-Cases where the Special Program obtains non-standard visits that cannot 
-be processed by the DMS (Section~\ref{ssec:proc_bounds}), and/or where 
-development of new algorithms or the allocation of significant additional 
-computational resources are needed, will require users tp generate unique
-and separate data products (Section~\ref{ssec:proc_user}).
+% Cases where the Special Program obtains non-standard visits that cannot 
+% be processed by the DMS (Section~\ref{ssec:proc_bounds}), and/or where 
+% development of new algorithms or the allocation of significant additional 
+% computational resources are needed, will require users to generate unique
+% and separate data products (Section~\ref{ssec:proc_user}).
 
@@
 
-Most of the Special Programs that are currently under consideration will 
-use standard or alternative standard visits.
-
-However some Special Programs are likely to use non-standard visits with 
-shorter exposures (5 sec), and some are likely to acquire images with a 
-significantly brighter sky background (e.g., twilight images) than 
-standard (or altnerative) WFD main survey visits.
-
-The cadence and patterns of Special Programs might also differ from the 
-WFD, such as long series of exposures obtained of the same field 
-(e.g., DDFs), or a strategy optimized to find very fast-moving objects 
-(e.g., NEOs).
-
-None of the currently proposed Special Programs listed in the Phase 2 
-SCOC recommendations \citedsp{PSTN-055} 
-seem likely to approach (or violate) any harware boundaries 
-(Section~\ref{ssec:proc_bounds_hardware}).
-
-However, some Special Programs observations seem likely to generate 
-images that cannot be processed with original or reconfigured versions of 
-the LSST Science Pipelines (Section~\ref{ssec:proc_bounds_processing}), 
-and will require user processing.
+Special Programs that do not use standard or alternative standard visits
+might be limited by hardware or processing boundaries.
 
@@
 
-The minimum exposure time is 1 second (stretch goal: 0.1 seconds), and 
-there is a potential hardware boundary that limits the readout rate to 1 
-every 15 seconds.
-This would affect the image acquisition rate and increase the overheads 
-on short exposures.
+In general, the currently-proposed Special Programs in \citeds{PSTN-055}
+are not anticipated to be limited by hardware boundaries.
 
-Special Programs which required the \emph{exact same} field pointing and 
+A few potential challenges to Special Programs posed by hardware
+boundaries are discussed below.
+
+\begin{itemize}
+
+\item Special Programs that use short exposures would be limited to the
+minimum exposure time is 1 second (stretch goal: 0.1 seconds).
+There is a potential hardware boundary that limits the readout rate to 1 
+every 15 seconds, which would affect the image acquisition rate and 
+increase the overheads on short exposures.
+
+\item Special Programs which require the \emph{exact same} field pointing and 
 rotation for \emph{every exposure} (to sub-arcsecond levels) might run 
 into hardware boundaries on pointing and tracking.
-None of the currently-proposed programs seem likely to be affected by 
-this potential boundary.
 
-Hardware imposes no other boundaries on how data can be obtained, but 
-Special Programs that request a high number of filter changes and/or long 
-slews could be inefficient due to large overheads.
+\item Special Programs that request a high number of filter changes and/or long 
+slews could be inefficient due to large overheads, but there are no
+boundaries on how data can be obtained.
+
+\item Special Programs that obtain twilight images will be subject to safe
+limits on sky background flux, as with any astronomical camera.
+
+\end{itemize}
 
@@
 Appendix~\ref{sec:procbounds} describes the boundaries on what types of visits 
-can be processed and calibrated by the DMS.
+can be processed and calibrated by the Data Management System and the LSST
+Science Pipelines.
 
-Very short ($<$2 sec) exposures could be difficult to reduce due to an 
-incompletely-formed PSF, and very short or very long ($>$150 sec) 
-exposures could be difficult to calibrate due to having too few (or too 
-few unsaturated) stars.
-As mentioned above the DMS is required to be able to process exposure 
-times as low as 1 second, and it is known that such short exposures might 
+Most of the currently-proposed Special Programs in \citeds{PSTN-055}
+are not anticipated to be limited by hardware boundaries.
+However, those which use non-standard visits, especially those with
+short exposure time or those obtained during twilight, might
+be affected by processing boundaries.
+
+The most likely challenges to Special Programs posed by processing
+boundaries are discussed below.
+
+\begin{itemize}
+
+\item Special Programs which use very short ($<$2 sec) exposures 
+could be difficult to reduce due to an incompletely-formed PSF 
+(Section~\ref{ssec:procbounds_expt}).
+The Data Management System is required to be able to process exposure 
+times as low as 1 second, but it is known that such short exposures might 
 have degraded image quality.
 
-It is currently unclear whether images with very bright sky backgrounds 
-(twilight images) can be processed with the LSST Science Pipelines, or 
-whether user generated pipelines will be needed.
+\item Special Programs that use very short or very long ($>$150 sec) 
+exposures could be difficult to calibrate due to having too few 
+(or too few unsaturated) stars.
 
-The full reduction and calibration of images obtained with non-sidereal 
-tracking, in which the stars are streaked, is currently beyond the scope 
-of the LSST Science Pipelines, and will require a user generated pipeline.
+\item For Special Programs which obtain images with very bright sky backgrounds
+(twilight images), it is currently unclear whether they can be processed
+with the LSST Science Pipelines; User-Generated Processing might be needed.
+
+\item The full reduction and calibration of data from any Special Programs that 
+use non-sidereal tracking, which produce images with star streaks, is
+currently beyond the scope of the LSST Science Pipelines; 
+User-Generated Processing would be needed.
+
+\end{itemize}
 

### 2023-10-18 sec_intro.tex
rm joinable, leave as detail

@@
 \item Special Processing should be done by Rubin Data Management to 
-produce unique and separate (but joinable) data products
+produce unique and separate data products
 for Special Programs when it is both possible and necessary.

### 2023-10-18 sec_sci.tex
move details to sec:proc

@@
 
-There are three main reasons for this recommendation.
+There are three main motivations for this recommendation.
 
@@
 \textbf{Special Processing should be done by Rubin Data Management to 
-produce unique and separate (but joinable) data products
+produce unique and separate data products
 for Special Programs when it is both possible and necessary.}
@@
 
-There are two main reasons for this recommendation.
+There are two main motivations for this recommendation.
 
@@
 
-Specific examples are provided only to illustrate this recommendation.
-These examples do not place limits on what Special Processing might be done.
-
-\begin{itemize}
-
-\item \textbf{Possible and necessary:}
-in order to detect high-redshift (faint) galaxies in the DDFs,
-Rubin Data Management uses the LSST Science Pipelines to deeply
-coadd images for each field, and store the results of source
-detection and characterization in unique and separate tables that
-are included in the annual data release (Section~\ref{ssec:proc_rdm_ddf}).
-
-\item \textbf{Possible but not necessary:}
-a time-domain mini-survey that uses standard visits \emph{could}
-have separate difference-image analysis object and source catalogs
-generated, but this is not necessary as the science goals for the
-mini-survey can be met by processing its data with regular Prompt
-Processing (Section~\ref{ssec:sci_pproc}), and ensuring the
-data is properly labeled (Section~\ref{ssec:sci_labels}).
-
-\item \textbf{Possible but not necessary (secondary science goals):}
-a time-domain mini-survey (or DDF) has a secondary science goal of detecting
-precursor outbursts for transients, which requires coadding images
-within windows of days, weeks, and months to reach various depths.
-This set of custom coadds may be considered as ``not necessary" and requiring 
-User-Generated Processing.
-
-\item \textbf{Necessary but not possible:}
-in order to find the most distant, faint Kuiper Belt Objects in the DDF,
-a specialized, computationally intensive form of ``shift-and-stack" processing
-is required for detection, but such algorithms are not used by the 
-LSST Science Pipelines and so User-Generated Processing will be needed (Section~\ref{ssec:SPCS_TNO}).
-A second example here is a twilight survey that uses non-standard visits 
-which are outside the boundaries of what the LSST
-Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
-
-\end{itemize}
+Further discussion on Special Processing, with examples, is provided in Section~\ref{ssec:proc_special}.
 
@@
 
-The phrase ``can be processed" includes all standard and alternative visits in 
-sky regions for which a template image exists.
-This might also include some non-standard visits (shorter or longer exposures), 
-as long as they can be processed by the Prompt pipeline and an appropriate template image exists.
+The condition ``can be processed" is ultimately left to the discretion of
+Rubin Data Management in Rubin Operations, but it is expected to include
+all standard and alternative visits in sky regions for which a template image exists.
 
-Visits with very short or very long exposure times (or very bright sky 
-backgrounds) might be excluded if they would need specialized algorithms for,
-e.g., instrument signature removal, difference-imaging, template-generation 
-(Section~\ref{ssec:proc_bounds}).
+The main motivation for this recommendation is that all time-domain and 
+moving-object science goals (two of the four science pillars for the LSST) are
+enhanced by any and all additional observations, even if they are not
+optimized for the science (as the WFD program's cadence is being optimized).
 
-If a Special Program's primary science goal requires specialized templates and 
-Prompt processing, the Data Management System will have the capability to load 
-and use an alternative template for some sky regions, based on the image metadata 
-(i.e., the labels described in Section~\ref{ssec:sci_labels}). 
-However, there would not be enough memory to hold alternative templates 
-for the whole sky.
-
-Including visits from Special Programs in regular Prompt processing alongside
-vists from the WFD program is not, in general, anticipated to affect WFD science goals.
-For example, analyses for a WFD-only subset could still be done using the program and
-region labels described in Section~\ref{ssec:sci_labels}.
-However, a few concerns are raised in Section~\ref{ssec:proc_prompt}.
+Further discussion of regular Prompt processing of Special Programs data, with
+examples and a discussion of a few challenges,
+is provided in Section~\ref{ssec:proc_reg_prompt}.
 

### 2023-10-18 sec_proc.tex
dropped in details from sec:sci

@@
 
-% Rubin Observatory and the LSST Data Management System (DMS) are required 
-% to process Special Programs data to produce unique and separate data 
-% products whenever the original or reconfigured versions of the LSST 
-% Science Pipelines can be used to do so (Section~\ref{sec:req}).
+A discussion of the anticipated details and challenges related to 
+obtaining and processing data from Special Programs.
 
-% The LSST Science Pipelines and the DMS are designed to process standard 
-% (or alternative standard) visits.
+Appendix~\ref{sec:spcs} provides detailed examples for the processing
+of data from Special Programs, including scenarios in which regular,
+Special, and User-Generate Processing are all involved.
 
-% It depends on the science goals related to the Special Program, but some 
-% Special Programs' data can and should be processed and served along with 
-% (or as a part of) the WFD data products (Section~\ref{ssec:proc_wfd}).
-% Others will require RDM to generate unique and separate data products 
-% (Section~\ref{ssec:proc_rdm}).
-
-% Cases where the Special Program obtains non-standard visits that cannot 
-% be processed by the DMS (Section~\ref{ssec:proc_bounds}), and/or where 
-% development of new algorithms or the allocation of significant additional 
-% computational resources are needed, will require users to generate unique
-% and separate data products (Section~\ref{ssec:proc_user}).
-
-\subsection{Boundaries on Obtaining and Processing Non-Standard Visits} \label{ssec:proc_bounds}
+\subsection{Boundaries on Non-Standard Visits} \label{ssec:proc_bounds}
 
 Special Programs that do not use standard or alternative standard visits
-might be limited by hardware or processing boundaries.
+might be affected by hardware or processing boundaries.
 
@@
 
-A few potential challenges to Special Programs posed by hardware
-boundaries are discussed below.
+A few potential challenges posed by hardware
+boundaries are summarized below.
 
@@
 
-The most likely challenges to Special Programs posed by processing
-boundaries are discussed below.
+The most likely challenges posed by processing
+boundaries are summarized below.
 
@@
 
-\subsection{Inclusion in the WFD Data Products}\label{ssec:proc_wfd}
+\subsection{Regular (non-special) processing}\label{ssec:proc_reg}
 
-Generally, data from Special Programs should be included in the Prompt 
-and/or Data Release processing and data products alongside data from the 
-WFD (i.e., included in the data products described by \citeds{lse-163}) 
-whenever this is (1) possible and (2) scientifically beneficial.
+Recall that the term ``regular processing" is used only within this document,
+and refers to the image processing pipelines that are designed for, 
+and will be applied to, the WFD program's observations (Section~\ref{sec:intro}).
 
-Decisions about when and whether to include Special Programs data in 
-the WFD's processing and data products are left to the discretion of the 
-Rubin Operations Data Production and System Performance teams.
+Decisions about when to apply regular processing to Special Programs data,
+or when to include it in the data products for the WFD program 
+(e.g., if it improves the all-sky coadds), 
+are ultimately left to the discretion of the Rubin Observatory's 
+Data Management and System Performance teams.
 
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing 
-examples for some potential Special Programs as further illustration.
+\subsubsection{Prompt Processing and Alert Production}\label{sssec:proc_reg_prompt}
 
-\subsubsection{WFD Prompt Processing and Alert Production}
+This might also include some non-standard visits (shorter or longer exposures), 
+as long as they can be processed by the Prompt pipeline and an appropriate template image exists.
+Visits with very short or very long exposure times (or very bright sky 
+backgrounds) might be excluded if they would need specialized algorithms for,
+e.g., instrument signature removal, difference-imaging, template-generation 
+(Section~\ref{ssec:proc_bounds}).
+
+If a Special Program's primary science goal requires specialized templates and 
+Prompt processing, the Data Management System will have the capability to load 
+and use an alternative template for some sky regions, based on the image metadata 
+(i.e., the labels described in Section~\ref{ssec:sci_labels}). 
+However, there would not be enough memory to hold alternative templates 
+for the whole sky.
+
+Including visits from Special Programs in regular Prompt processing alongside
+vists from the WFD program is not, in general, anticipated to affect WFD science goals.
+For example, analyses for a WFD-only subset could still be done using the program and
+region labels described in Section~\ref{ssec:sci_labels}.
+
+OLD
 
@@
 
-\subsubsection{WFD Solar System Processing}
+\subsubsection{Solar System Processing}\label{sssec:proc_reg_ss}
 
@@
 
-\subsubsection{WFD Data Release Processing}
+\subsubsection{Data Release Processing}\label{sssec:proc_reg_dr}
 
@@
 
-\subsection{RDM-Generated ``Unique and Separate" Data Products}\label{ssec:proc_rdm}
+\subsection{Special Processing}\label{ssec:proc_special}
 
-Rubin Data Management (RDM) is required to produce unique, separate, and 
-joinable data products for Special Programs whenever this is possible with the 
-original or reconfigured versions of the LSST Science Pipelines, and no new 
-algorithmic development or significant computational resources are needed.
+As described in Section~\ref{ssec:sci_sproc}, 
+Special Processing should be done by Rubin Data Management to 
+produce unique and separate data products
+for Special Programs when it is both possible and necessary.
 
-As described in Section~\ref{ssec:proc_wfd}, in some cases the science goals 
-of a Special Program might be met by including the data in the WFD data 
-products; especially the Prompt or DIA data products for standard visits.
-Thus, most of the RDM-generated data products considered below are for 
-static-sky Data Release data products, or data products for non-standard visits.
-Decisions about when to produce separate Special Programs data products are 
-left to the discretion of RDM.
+In short, \emph{possible} means that original or reconfigured versions of the LSST
+Science Pipelines can be used , and \emph{necessary} means the primary science goal for Special Program
+could not be met without the data products produced by the Special Processing.
+The definitions of possible and necessary, and the scope of Special Processing,
+are ultimately left to the discretion of Rubin Data Management in Operations.
+
+Specific examples are provided only to illustrate this interpretation of possible and necessary.
+These examples do not place limits on what Special Processing might be done.
+
+\begin{itemize}
+
+\item \textbf{Possible and necessary:}
+in order to detect high-redshift (faint) galaxies in the DDFs,
+Rubin Data Management uses the LSST Science Pipelines to deeply
+coadd images for each field, and store the results of source
+detection and characterization in unique and separate tables that
+are included in the annual data release (Section~\ref{ssec:proc_rdm_ddf}).
+
+\item \textbf{Possible but not necessary:}
+a time-domain mini-survey that uses standard visits \emph{could}
+have separate difference-image analysis object and source catalogs
+generated, but this is not necessary as the science goals for the
+mini-survey can be met by processing its data with regular Prompt
+Processing (Section~\ref{ssec:sci_pproc}), and ensuring the
+data is properly labeled (Section~\ref{ssec:sci_labels}).
+
+\item \textbf{Possible but not necessary (secondary science goals):}
+a time-domain mini-survey (or DDF) has a secondary science goal of detecting
+precursor outbursts for transients, which requires coadding images
+within windows of days, weeks, and months to reach various depths.
+This set of custom coadds may be considered as ``not necessary" and requiring 
+User-Generated Processing.
+
+\item \textbf{Necessary but not possible:}
+in order to find the most distant, faint Kuiper Belt Objects in the DDF,
+a specialized, computationally intensive form of ``shift-and-stack" processing
+is required for detection, but such algorithms are not used by the 
+LSST Science Pipelines and so User-Generated Processing will be needed (Section~\ref{ssec:SPCS_TNO}).
+A second example here is a twilight survey that uses non-standard visits 
+which are outside the boundaries of what the LSST
+Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
+
+\end{itemize}
+
+\textbf{Tables should be joinable -- } where feasible.
 
@@
 user-generated data products, for both WFD and Special Programs.
-
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing 
-examples for some potential Special Programs as further illustration.
 
@@
 
-\subsection{User-Generated ``Unique and Separate" Data Products}\label{ssec:proc_user}
+\subsection{User-Generated Processing}\label{ssec:proc_user}
 

### 2023-10-18 sec_proc.tex
updated thru SS

@@
 of data from Special Programs, including scenarios in which regular,
-Special, and User-Generate Processing are all involved.
+Special, and User-Generated Processing are all involved.
 
@@
 
-\item Special Programs that use short exposures would be limited to the
+\item \textbf{Short exposures --}
+Special Programs that use short exposures would be limited to the
 minimum exposure time is 1 second (stretch goal: 0.1 seconds).
@@
 
-\item Special Programs which require the \emph{exact same} field pointing and 
+\item \textbf{Repeated pointing --}
+Special Programs which require the \emph{exact same} field pointing and 
 rotation for \emph{every exposure} (to sub-arcsecond levels) might run 
@@
 
-\item Special Programs that request a high number of filter changes and/or long 
-slews could be inefficient due to large overheads, but there are no
-boundaries on how data can be obtained.
-
-\item Special Programs that obtain twilight images will be subject to safe
+\item \textbf{Twilight images --}
+Special Programs that obtain twilight images will be subject to safe
 limits on sky background flux, as with any astronomical camera.
@@
 \end{itemize}
+
+Finally, as a side note, Special Programs that request a high number of 
+filter changes and/or long slews could be inefficient due to large overheads,
+but would not be limited by hardware boundaries.
 
@@
 
-\item Special Programs which use very short ($<$2 sec) exposures 
+\item \textbf{Very short exposures --}
+Special Programs which use very short ($<$2 sec) exposures 
 could be difficult to reduce due to an incompletely-formed PSF 
@@
 
-\item Special Programs that use very short or very long ($>$150 sec) 
+\item \textbf{Very short or very long exposures --}
+Special Programs that use very short or very long ($>$150 sec) 
 exposures could be difficult to calibrate due to having too few 
@@
 
-\item For Special Programs which obtain images with very bright sky backgrounds
+\item \textbf{Twilight images --}
+For Special Programs which obtain images with very bright sky backgrounds
 (twilight images), it is currently unclear whether they can be processed
@@
 
-\item The full reduction and calibration of data from any Special Programs that 
+\item \textbf{Streaked images --}
+The full reduction and calibration of data from any Special Programs that 
 use non-sidereal tracking, which produce images with star streaks, is
@@
 
+As described in Section~\ref{ssec:sci_pproc}, 
+all visits that \emph{can} be processed by the Prompt pipelines and generate 
+alerts \emph{should} be, in support of time domain and Solar System science goals.
+
+\textbf{The meaning of ``can be processed".}
+The condition ``can be processed" is ultimately left to the discretion of
+Rubin Data Management in Rubin Operations, but is expected to include
+all standard and alternative visits in sky regions for which a template image exists.
 This might also include some non-standard visits (shorter or longer exposures), 
@@
 
+\textbf{The use of specialized (alternative) template images.}
 If a Special Program's primary science goal requires specialized templates and 
@@
 
+\textbf{WFD and Special Program data would co-reside.}
+No ``unique and separate" data products for the Special Progams would be 
+produced by regular Prompt processing.
+Special Programs data that is processed by the Prompt pipeline would 
+contribute to the Prompt data products for the WFD program as 
+described in Section 3 of the DPDD \citedsp{lse-163}. 
+These data products are the results of Difference Image Analysis (DIA),
+such as the difference images, catalogs of sources detected in difference
+images ({\tt DiaSources}) and associated static-sky {\tt DiaSources}
+into {\tt DiaObjects}, and alert packets.
+
 Including visits from Special Programs in regular Prompt processing alongside
@@
 For example, analyses for a WFD-only subset could still be done using the program and
-region labels described in Section~\ref{ssec:sci_labels}.
+region labels described in Section~\ref{ssec:sci_labels}, which would be
+propogated to difference images, difference-image catalogss, and alerts.
 
-OLD
+\subsubsubsection{Potential issues with Prompt processing for untiled sequences from Special Programs.}
 
-As all visits can contribute to the time-domain astronomy science goals 
-of the LSST, all Special Programs data that {\it can} be processed by the 
-prompt processing and alert production pipelines {\it should} be, 
-alongside visits from the WFD.
+There are two potential issues with Prompt processing for DDFs, or any mini-survey 
+that obtains a sequence of untiled images: images at the same pointing or which overlap.
+For example, a DDF which obtains a hour-long series of about a hundred images at the same coordinates,
+every few nights for a few months.
 
-This would include all standard and alternative visits in sky regions for which 
-a template image exists.
-This might also include some non-standard visits (shorter or longer exposures) 
-as long as they can be processed by the Prompt 
-pipeline\footnote{If a Special Program's science goals do require 
-specialized templates and 
-Prompt processing, the DMS will have the capability to load and use an 
-alternative template for some regions, based on the image metadata. 
-However, there would not be enough memory to hold alternative templates 
-for the whole sky region.}.
+\begin{enumerate}
 
-Visits with very short or very long exposure times (or very bright sky 
-backgrounds) might be excluded if they would 
-need specialized algorithms for difference-imaging or template-generation.
+\item {\tt DiaObject} histories may become too large for the sizing model.
+Alert packets contain the full records of all associated 
+{\tt DiaSources} from the past 12 months, but the alert
+stream bandwidth is sized for the expected histories for
+WFD program fields. 
+The Prompt pipeline resources are also sized for the
+WFD program, and it might not be possible to load up
+thousands of epochs at a time.
+The Data Management team will have to test the realized
+capabilities of Prompt processing and alert distribtuion,
+and potentially impose a mitigation strategy such as
+limiting histories to the last $N$ observations instead
+of the last 12 months in heavily-observed regions.
 
-Special Programs data that is processed by the Prompt pipeline would 
-contribute to the Prompt data products described in Section 3 of the 
-DPDD \citedsp{lse-163}.
-
-Difference images, {\tt DiaSources}, and alerts would have region and 
-program labels (Section~\ref{ssec:sci_labels}), providing origin and
-provenance for brokers and users.
-
-\textbf{Two potential issues with Prompt processing for DDFs}, 
-or any mini-survey that obtains a sequence of images without slewing, 
-include:
-
-(1) As alert packets contain the full records of all associated 
-{\tt DiaSources} from the past 12 months, alerts for {\tt DiaObjects} in 
-the DDFs might become very large.
-
-(2) The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
-will be somewhat compromised for a DDF sequence.
-The processing for the second image begins when the processing for the 
-first image is only halfway done; when the {\tt DiaObject} catalog has 
-not yet updated with the new {\tt DiaSources} detected in the first image.
-Thus, the {\tt DiaSource} from images one and two for a new transient 
+\item The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
+may be compromised.
+For consecutive images, processing for the second image begins when the processing for the 
+first image is only halfway done.
+At this point, the {\tt DiaObject} catalog has 
+not yet been updated with the new {\tt DiaSources} detected in the first image.
+Thus, the {\tt DiaSource}s from images one and two for a new transient 
 would not be associated with a single {\tt DiaObject}, but instead would 
@@
 
-These two issues would not impact time-domain science.
-Brokers and users would be able to use the region and program labels
-in the data as context (i.e., as flags for potentially-compromised 
-data products) and, for example, avoid DDF alerts should they so choose.
+\end{enumerate}
+
+These two potential issues pose challenges, but are not showstoppers in processing 
+Special Programs data with the Prompt pipelines. 
+The overall impact on time-domain science would still be positive, even 
+if mitigations are needed for these issues.
+For example, brokers and users would be able to use the region and program labels
+in the data as context (i.e., as flags) and avoid including 
+limited-history or potentially-compromised {\tt DiaSources} in their
+analyses if necessary.
 
@@
 
-Since Solar System Processing takes {\tt DiaSource}s as input, so any 
-Special Programs images that can be run through the Alert Pipeline can 
-also be incorporated into Solar System Processing.
+Since Solar System Processing takes {\tt DiaSource}s as input, any 
+Special Programs images that are processed by the Prompt pipeline
+could be incorporated into Solar System Processing as well.
 

### 2023-10-18 sec_proc.tex
rm subsubsubsection command

@@
 
-\subsubsubsection{Potential issues with Prompt processing for untiled sequences from Special Programs.}
-
+\textbf{Potential issues with Prompt processing for untiled sequences from Special Programs.}
 There are two potential issues with Prompt processing for DDFs, or any mini-survey 

### 2023-10-18 sec_sci.tex
fix ref

@@
 examples and a discussion of a few challenges,
-is provided in Section~\ref{ssec:proc_reg_prompt}.
+is provided in Section~\ref{sssec:proc_reg_prompt}.
 

### 2023-10-18 sec_proc.tex
updated through SP

@@
 \textbf{The meaning of ``can be processed".}
-The condition ``can be processed" is ultimately left to the discretion of
+This is ultimately left to the discretion of
 Rubin Data Management in Rubin Operations, but is expected to include
@@
 
-\textbf{WFD and Special Program data would co-reside.}
+\textbf{How WFD and Special Program data would co-reside.}
 No ``unique and separate" data products for the Special Progams would be 
@@
 
-\item {\tt DiaObject} histories may become too large for the sizing model.
+\item \textbf{{\tt DiaObject} histories may become too large for the sizing model.}
 Alert packets contain the full records of all associated 
@@
 
-\item The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
-may be compromised.
+\item \textbf{The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
+may be compromised.}
 For consecutive images, processing for the second image begins when the processing for the 
@@
 
-\textbf{Difference images, {\tt DiaSource} and {\tt DiaObject} catalogs -- }
-As all visits can contribute to the time-domain astronomy science goals 
-of the LSST, all Special Programs data that {\it can} be processed with 
-the difference image analysis (DIA) pipeline alongside the WFD visits 
-{\it should} be.
-All of this processing should use the same template image for a given
-field.
-Users who want, e.g., only {\tt DiaSources} a given DDF, would be 
-able to write queries using the region and program labels in the catalog.
+\textbf{Time-domain DIA data products --}
+This includes the results of the annual reprocessing of WFD program data with 
+Difference Image Analysis (DIA), and the production of Data Release
+versions of the processed images (single-visit and difference images)
+and associated catalogs ({\tt DiaSource}, {\tt DiaObject}, {\tt Source},
+{\tt ForcedSource}, {\tt DiaForcedSource}, and so on).
+These data products will be used primarily for time-domain science.
+They should include Special Programs data for the same reasons as
+provided in Section~\ref{ssec:sci_pproc}, and with the same 
+considerations as discussed in Section~\ref{sssec:proc_reg_prompt}.
+This processing should use the same template image for a given field.
 
-\textbf{Processed visit images, {\tt Source} and {\tt ForcedSource} catalogs -- }
-For the same reasons as above, all Special Programs data that 
-{\it can} be processed with the Data Release pipieline alongside the WFD 
-visits {\it should} be.
-
-\textbf{Deep coadds and the {\tt Object} catalog -- }
-Some of the core LSST science goals require a WFD main survey data 
-products of nearly uniform depth.
-Whether and how to include any Special Programs data in the deep image 
-coadds and the corresponding {\tt Object} catalog is left entirely to the 
-discretion of the RDM team in Operations.
-
-For example, perhaps Special Programs images will only be included when 
-they bring additional area to the same depth as the rest of the WFD, or 
-when they suppress edge effects or low-order modes in the 
+\textbf{Static-sky data products --}
+This includes the tessellation and coaddition of WFD program images
+and the associated multi-band {\tt Object} catalog and survey property maps.
+Whether and how to include any Special Programs data in these data products
+is left entirely to the discretion of the Rubin Data Management team in Operations.
+As an example, perhaps Special Programs images will only be included when they 
+assist with uniformity or suppress edge effects or low-order modes in the 
 all-sky photometric solutions.
-
-For stacked images and their associated catalogs, the region labels should be
-included (but the program labels would not be needed or possible).
 
@@
 could not be met without the data products produced by the Special Processing.
-The definitions of possible and necessary, and the scope of Special Processing,
+
+\textbf{Joinable tables --}
+Any tables for the unique and separate data products should be joinable to the 
+data products for the WFD program, when possible.
+
+\textbf{Survey property maps --}
+As for the tesslated all-sky coadded images made from WFD program data, survey property maps
+should be made individually as part of any Special Programs that generates
+tesselated coadded images.
+
+\textbf{Special Processing timescales --}
+The timescales for Special Processing should be adopted that best serve the
+primary science goal for the Special Program.
+For example, nighly-coadd difference-image analysis for DDFs is the most
+scientifically useful if done on a daily cadence, but the deeply coadded
+images for DDF fields could be released annually.
+
+The interpretations of possible and necessary, and the scope of Special Processing,
 are ultimately left to the discretion of Rubin Data Management in Operations.
@@
 User-Generated Processing.
+Note that for the WFD program data, such custom coadds are also considered 
+as beyond scope and in need of User-Generated Processing.
 
@@
 
-\textbf{Tables should be joinable -- } where feasible.
+Further examples of Special Processing are provided below.
 
-\textbf{A note about intermediate-depth ``custom" coadds -- }
-As with the annual WFD deeply coadded images and associated catalogs, RDM would 
-only generate one coadd (including all images to date) for a given special region.
-Custom coadds (e.g., weekly, monthly) for very specific science goals are considered 
-user-generated data products, for both WFD and Special Programs.
-
-\subsubsection{Deep Drilling Fields}
+\subsubsection{Deep Drilling Fields (DDFs)}
 
 As the DDFs will likely be observed with standard or alternative standard 
-visits, it is likely that RDM will be able to reconfigure existing pipelines 
-to process the DDF data and produce ``unique and separate" DDF data products.
+visits, Data Management will be able to reconfigure existing pipelines for
+Special Processing to produce unique and separate DDF data products.
 
-For example, RDM-generated ``unique and separate" DDF data products might 
-include:
+For example, Special Processing for the DDF data products might include:
 \begin{itemize}
@@
 
-As of the Phase 2 SCOC recommendations \citedsp{PSTN-055}, three kinds of 
-short-exposure mini-surveys were under consideration: 
-a static short exposure map of the sky in $ugrizy$ for calibration; 
-a static to transient short exposure survey; and 
-an Near-Earth Objects (NEO) twilight survey.
+As described in \citedsp{PSTN-055}, there are a few
+short-exposure mini-surveys are under consideration.
+Two examples are a short exposure map of the sky in $ugrizy$ for calibration,
+and a Near-Earth Objects (NEO) twilight survey.
 
-If either the transients or NEO short-exposure suvey is executed, and if 
-RDM finds that the LSST Science Pipelines and, specifically, template images 
-can be made and difference image analysis (DIA) run, then the RDM-generated 
-data products may be similar to the Prompt data products (potentially 
-including alerts; Section 3 of \citeds{lse-163}).
+Special Processing for short-exposure mini-surveys remains to
+be determined.
+Since it falls under the remit of Data Management to perform proper calibration,
+an evaluatation of whether short exposures for calibration are necessary
+will be done. 
+If Data Management does find that short-exposure and/or high-sky brightness images
+can be processed with reconfigured versions of the LSST Science Pipelines,
+then unique and separate data products could be generated with Special Processing.
+For the NEO twilight survey, these data products would likely be similar to the
+Prompt or DIA data products.
+For the calibration survey, these data products would like be similar to the
+annual data release tessellated coadds and associated catalog.
 
-If either of the static short-exposure mini-surveys is executed, and if 
-RDM finds that the LSST Science Pipelines can be reconfigured and used 
-for short exposures, then the RDM-generated data products would likely be 
-of a similar format to the Data Release data products 
-(Section 4, \citeds{lse-163}).
-For example, a separate repository of processed visit images and coadded 
-images, and separate (but joinable) {\tt Source}, {\tt ForcedSource}, and 
-{\tt Object} catalogs.
+\subsubsection{Standard Visit Mini-Surveys}
 
-\subsubsection{Mini-Surveys}
+Consider a Special Program in which a special region of sky is observed with standard
+visits but a a special strategy or cadence which is significantly distinct from the WFD program,
+and lasts for a limited amount of time.
+For example, a short-term survey of the Magellanic Clouds, with dual primary science
+goals in time-domain and static-sky science.
 
-The cases where ``unique and separate" data products for mini-surveys are 
-needed to reach the science goals of a Special Program, and are possible for 
-RDM to generate, are most likely to be of a similar format to the annual 
-Data Release data products: static-sky coadded images and associated 
-{\tt Source}- and {\tt Object}-like cataogs for standard or alternative visits.
-
-In cases where the observations are all obtained with a short season 
-(e.g., a week spent observing a specific region), such ``annual" 
-data products might be generated and released on an intermediate time scale, 
-if resources allow.
+In cases like this, a set of unique and separate data products with the same formats as the 
+time-domain DIA and static-sky data products described in Section~\ref{sssec:proc_reg_dr}
+should be created with Special Processing.
+They might be released with an annual data release or on an intermediate timescale, e.g.,
+within six months of the conclusion of the mini-survey observations.
 

### 2023-10-18 sec_proc.tex
updated thru UGP

@@
 
+\textbf{Computational resources -- }
 Users will have access to the LSST Science Pipelines and data processing 
@@
 
+\textbf{Adopting user code or data products --}
 It is expected that some User-Generated pipelines and data products 
 might be ``adopted" or ``federated" into the LSST Science Pipelines and the Prompt 
-and Data Release data products, however, details regarding the federation of 
-user-generated data products are to be provided elsewhere.
+and Data Release data products.
+Details regarding this are to be provided elsewhere.
 
-\textbf{A note prompt processing and alert production -- }
+\textbf{Alert production is restricted -- }
+User-Generated Processing will not be able to release alert packets in the LSST alert stream.
 As the latency on processed visit image availability has an 80-hour
@@
 (60 seconds to 24 hours).
-No user-generated pipeline may contribute alerts to the LSST alert stream on 
+Thus, no User-Generated Processing may contribute alerts to the LSST alert stream on 
 any timescale.
 
-Appendix~\ref{sec:spcs} provides more detailed, step-by-step data processing examples 
-for some potential Special Programs as further illustration.
+Further examples of User-Generated Processing are provided below.
 
@@
 
-User-generated data products might include, for example, DDF images coadded on custom 
-timescales (e.g., weekly, monthly), or coadded using algorithms outside of the LSST 
-Science Pipelines.
+User-Generated Processing and data products might include, for example, DDF images coadded 
+on custom timescales (e.g., weekly, monthly), or coadded using algorithms outside of the 
+LSST Science Pipelines.
 
@@
 Short-exposure images obtained during twilight, which will have a very bright sky 
-background unlike other LSST images, might require user-generated algorithms to 
-subtract the high sky background.
+background unlike other LSST images, might require specialized algorithms
+to subtract the high sky background which are not available in the LSST Science Pipelines,
+and might require User-Generated Processing.
 
-Short-exposure images obtained during the night would have fewer stars for 
-astrometric and photometric calibration, and might require user-generated 
-processing pipelines.
+Short-exposure images obtained during the night might have too-few stars to satisfy the
+astrometric and photometric calibration routines in the LSST Science Pipelines,
+and might require User-Generated Processing.
 
@@
 e.g., those that require difference imaging with coadded images on an intermediate 
-timescale (e.g., a weekly stack), would require user-generated processing.
+timescale (e.g., a weekly stack), would require User-Generated Processing.

### 2023-10-18 sec_proc.tex
minor updates

@@
 
-\item \textbf{{\tt DiaObject} histories may become too large for the sizing model.}
+\item \textbf{DIA Object histories may become too large for the sizing model.}
 Alert packets contain the full records of all associated 
@@
 
-\item \textbf{The association of {\it new} {\tt DiaSources} into {\tt DiaObjects} 
-may be compromised.}
+\item \textbf{The association of new DIA Sources into DIA Objects may be compromised.}
 For consecutive images, processing for the second image begins when the processing for the 
@@
 detection and characterization in unique and separate tables that
-are included in the annual data release (Section~\ref{ssec:proc_rdm_ddf}).
+are included in the annual data release.
 
@@
 mini-survey can be met by processing its data with regular Prompt
-Processing (Section~\ref{ssec:sci_pproc}), and ensuring the
-data is properly labeled (Section~\ref{ssec:sci_labels}).
+Processing, and ensuring the data is properly labeled.
 
@@
 is required for detection, but such algorithms are not used by the 
-LSST Science Pipelines and so User-Generated Processing will be needed (Section~\ref{ssec:SPCS_TNO}).
-A second example here is a twilight survey that uses non-standard visits 
-which are outside the boundaries of what the LSST
-Science Pipelines can process (Section~\ref{ssec:proc_bounds}).
+LSST Science Pipelines and so User-Generated Processing will be needed.
+A second example is a twilight survey that uses non-standard visits 
+which are outside the boundaries of what the LSST Science Pipelines can process.
 
@@
 
-Further examples of Special Processing are provided below.
+Further examples of potential Special Processing for anticipated Special Programs are provided below.
 
-\subsubsection{Deep Drilling Fields (DDFs)}
+\subsubsection{Deep Drilling Fields (DDFs)}\label{sssec:proc_special_ddf}
 
@@
 
-Further examples of User-Generated Processing are provided below.
+Further examples of potential User-Generated Processing for anticipated Special Programs are provided below.
 

### 2023-10-19 sec_proc.tex
cite sheppard et al.

@@
 (twilight images), it is currently unclear whether they can be processed
-with the LSST Science Pipelines; User-Generated Processing might be needed.
+with the LSST Science Pipelines; 
+User-Generated Processing might be needed (e.g., \citealt{2022AJ....164..168S}).
 
@@
 Special Programs images that are processed by the Prompt pipeline
-could be incorporated into Solar System Processing as well.
+could be incorporated into Solar System Processing.
 

### 2023-10-19 sec_req.tex
add bold headings

@@
 
-This section proves a summary of the requirements related to Special Programs.
-A more detailed list is provided in Appendix~\ref{sec:docrev}.
+Appendix~\ref{sec:docrev} provides a full description of the requirements 
+on the Data Management System related to Special Programs, which are summarized here.
 
@@
 
+\textbf{Produce unique and special (and joinable) data products when possible --}
 Rubin Observatory and the LSST system (the observatory and the data 
@@
 
-It is a requirement that the cumulative size of the Special Programs data 
-products generated by Rubin Observatory be no more than $\sim$10\% the 
-size of the Data Release data products (i.e., proportional to the fraction 
-of survey time spent; LSR-REQ-0121).
-
 It is a requirement that these Special Programs data products be distinct, 
@@
 
+\textbf{The data products' size should be about 10\% of the WFD programs' --}
+It is a requirement that the cumulative size of the Special Programs data 
+products generated by Rubin Observatory be no more than $\sim$10\% the 
+size of the Data Release data products (LSR-REQ-0121).
+The spirit of this requirement is that the size be proportional to the fraction 
+of survey time spent on Special Programs.
+
 The derivation of value-added data products, such as HiPS or MOC maps, for 
-Special Programs remains an open question (DMS-REQ-0379, 0383).
+Special Programs remains an open question (DMS-REQ-0379, 0383), and is not required.
+However, they are suggested to be produced with Special Processing where
+appropriate (Section~\ref{ssec:proc_special}).
 
@@
 
+\textbf{Store metadata sufficient to trigger Special Processing -- }
 In order to support Special Programs processing, the LSST system is 
@@
 or a Special Program (DMS-REQ-0068).
-
 It is required that this metadata be sufficient for Special Programs to 
@@
 
+\textbf{Latency requirements apply to alerts from Special Programs images --}
 It is a requirement that any Special Programs processing done with the 
@@
 
+\textbf{Intermediate timescales should be used to enable science --}
 It is also a requirement that Special Programs processing be done on 
@@
 
+\textbf{Exposure times of 1 second must be processable -- }
 It is a requirement that the LSST system be able to process non-standard 
@@
 
-Processing for Special Programs by Rubin Observatory is expected to use no 
-more than $\sim$10\% of computational and storage capacity of the Rubin 
-data processing cluster (i.e., proportional to the fraction of survey time 
-spent and the size of the Rubin-processed Special Programs data products; 
-Section 6 of the DPDD).
+It is not a requirement, but processing for Special Programs by Rubin Observatory is expected to use no 
+more than $\sim$10\% of computational and storage capacity of the Rubin data processing cluster 
+(i.e., proportional to the fraction of survey time spent; Section 6 of LSE-163).
 
@@
 
-In cases where the science goals of a Special Program require that new 
-algorithms or software be developed, User-Generated pipelines and data 
-products will be needed.
-
 The 10\% of the total data processing capacity that Rubin Observatory is 
-required to reserve for all User-Generated processing includes that 
-applied by users to Special Programs data -- in other words, there is no 
-additional capacity {\it for users} that will be reserved only for Special 
+required to reserve for \emph{all} User-Generated Processing includes that 
+applied by users to Special Programs data.
+There is no additional capacity {\it for users} that will be reserved only for Special 
 Programs data (LSR-REQ-0041).

### 2023-10-19 sec_proc.tex
cross-ref to reqs

@@
 The Data Management System is required to be able to process exposure 
-times as low as 1 second, but it is known that such short exposures might 
-have degraded image quality.
+times as low as 1 second (Section~\ref{ssec:req_proc}), 
+but it is known that such short exposures might have degraded image quality.
 
@@
 (Section~\ref{ssec:proc_bounds}).
+Alert latency requirements do apply to Special Programs data (Section~\ref{ssec:req_proc}),
+and if they cannot be met then Prompt processing is ``not possible".
 
@@
 
-\textbf{How WFD and Special Program data would co-reside.}
+\textbf{How WFD and Special Program data would co-exist.}
 No ``unique and separate" data products for the Special Progams would be 
@@
 
-These two potential issues pose challenges, but are not showstoppers in processing 
+These two potential issues pose challenges, but are not necessarily showstoppers in processing 
 Special Programs data with the Prompt pipelines. 
@@
 for Special Programs when it is both possible and necessary.
+This is a requirement (Section~\ref{ssec:req_dp}).
 
@@
 data products for the WFD program, when possible.
+This is a requirement (Section~\ref{ssec:req_dp}).
 
@@
 should be made individually as part of any Special Programs that generates
-tesselated coadded images.
+tesselated coadded images. This is \emph{not} a requirement (Section~\ref{ssec:req_dp}).
 

### 2023-10-19 sec_req.tex
minor edits

@@
 
-\textbf{Produce unique and special (and joinable) data products when possible --}
+\textbf{Produce unique and special (and joinable) data products when possible. --}
 Rubin Observatory and the LSST system (the observatory and the data 
@@
 (LSR-REQ-0121).
+It is a requirement that these Special Programs data products be distinct 
+and joinable with the Prompt and/or Data Release data products (DMS-REQ-0322).
 
@@
 
-It is a requirement that these Special Programs data products be distinct, 
-and joinable with (in other words, they can be federated or cross-matched 
-with) the Prompt and Data Release data products (DMS-REQ-0322).
+The term ``joinable" means the Special Programs data products can be federated 
+or cross-matched with the relevant Prompt and Data Release data products, and that
+a column of cross-matched object identifiers is provided to enable table joins.
 
-\textbf{The data products' size should be about 10\% of the WFD programs' --}
+\textbf{The size of Special Programs data products should be about 10\% of the total. --}
 It is a requirement that the cumulative size of the Special Programs data 
@@
 size of the Data Release data products (LSR-REQ-0121).
-The spirit of this requirement is that the size be proportional to the fraction 
+
+The spirit of this requirement on data volume is that the size be proportional to the fraction 
 of survey time spent on Special Programs.
@@
 
-\textbf{Store metadata sufficient to trigger Special Processing -- }
+\textbf{Program metadata that is sufficient to trigger Special Processing should be stored. -- }
 In order to support Special Programs processing, the LSST system is 
@@
 
-\textbf{Latency requirements apply to alerts from Special Programs images --}
+\textbf{Latency requirements apply to alerts from Special Programs images. --}
 It is a requirement that any Special Programs processing done with the 
@@
 
-\textbf{Intermediate timescales should be used to enable science --}
+\textbf{Intermediate timescales should be used to enable science when needed. --}
 It is also a requirement that Special Programs processing be done on 
@@
 
-\textbf{Exposure times of 1 second must be processable -- }
+\textbf{Exposure times of 1 second should be processable. -- }
 It is a requirement that the LSST system be able to process non-standard 

### 2023-10-19 sec_sci.tex
improve req links

@@
 \textbf{Special Processing should be done by Rubin Data Management to 
-produce unique and separate data products
+produce unique and separate (and joinable) data products
 for Special Programs when it is both possible and necessary.}
@@
 
-The definitions of possible and necessary, and the scope of Special Processing,
-are ultimately left to the discretion of Rubin Data Management in Operations.
-Note that secondary science goals may be considered as ``not necessary".
-
 There are two main motivations for this recommendation.
@@
 with the original or reconfigured versions of the LSST Science Pipelines
-(Section~\ref{sec:req}).
+(Section~\ref{ssec:req_dp}).
 
@@
 \end{enumerate}
+
+The definitions of possible and necessary are further illustrated with examples in Section~\ref{ssec:proc_special}.
+Note that secondary science goals may be considered as ``not necessary".
+
+The general scope of Special Processing, including situations in which cross-match and table joins 
+would be possible and scientifically relevant, 
+are ultimately left to the discretion of Rubin Data Management in Operations.
 
@@
 Rubin Observatory will reserve 10\% of its total data processing capacity for users.
-This is already a requirement, as mentioned in Section~\ref{ssec:req_ug}.
 This component would include {\it all} user processing and re-processing of any and 
 all LSST data, including Special Programs. 
+This is already a requirement, as mentioned in Section~\ref{ssec:req_proc}.
 

### 2023-10-19 sec_proc.tex
org updates

@@
-\section{Processing Data from Special Programs}\label{sec:proc}
+\section{Details and Challenges of Processing Data from Special Programs}\label{sec:proc}
 
@@
 into {\tt DiaObjects}, and alert packets.
-
 Including visits from Special Programs in regular Prompt processing alongside
@@
 There are two potential issues with Prompt processing for DDFs, or any mini-survey 
-that obtains a sequence of untiled images: images at the same pointing or which overlap.
+that obtains a sequence of untiled images (images at the same pointing or which overlap).
 For example, a DDF which obtains a hour-long series of about a hundred images at the same coordinates,
@@
 of the last 12 months in heavily-observed regions.
+Alerts from Special Programs data are subject to the same latency
+requirements as WFD program alerts (Section~\ref{ssec:req_proc}).
 
@@
 images for DDF fields could be released annually.
+
+\subsubsection{Scenarios to illustrate ``possible and necessary"}
 

### 2023-10-19 sec_proc.tex
add footnote

@@
 Special Programs that use short exposures would be limited to the
-minimum exposure time is 1 second (stretch goal: 0.1 seconds).
+minimum exposure time is 1 second (stretch goal: 0.1 seconds\footnote{A short-exposure survey 
+of the bright stars of M67, described in Chapter 10.4 of the 
+Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 
+0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform 
+photometry on stars as much as 6 magnitudes brighter than the saturation level"}. 
+This may require User-Generated Processing.}).
 There is a potential hardware boundary that limits the readout rate to 1 

### 2023-10-19 ap_spcs.tex
updates

@@
-\section{Special Programs Processing Examples}\label{sec:spcs}
+\section{Scenarios for Processing Data from Special Programs}\label{sec:spcs}
 
-Step-by-step examples of the processing, either by RDM or users, 
-for data from several Special Programs.
+\textbf{\emph{Hypothetical}} examples of Special Programs
+and the regular, Special, and User-Generated processing
+to illustrate what \emph{might be done}.
 
-Basic steps that we use to describe a processing case study: \\
+The details of the data acquisition and processing mentioned below are 
+\emph{just illustrative examples} of decisions that have yet to be made.
+
+The steps used to describe the hypothetical processing for each case scenario are: \\
 Step 1. Data acquisition. \\
-Step 2. RDM prompt processing and alert production. \\
-Step 3. RDM special processing with reconfigured pipelines. \\
-Step 4. RDM inclusion in the WFD Data Release data products. \\
-Step 5. User processing and user-generated data products. \\
+Step 2. Regular Prompt processing and Alert Production. \\
+Step 3. Special Processing with reconfigured pipelines. \\
+Step 4. Regular processing for inclusion in WFD program data products. \\
+Step 5. User-Generated Processing. \\
 
-\subsection{Searching for TNOs with Shift-and-Stack}\label{ssec:SPCS_TNO}
+\subsection{Outer solar system mini-survey}\label{ssec:SPCS_TNO}
 
-This Special Programs processing summary is based on Becker et al. (2011) 
-white paper to find TNOs with shift-and stack (SAS) \citedsp{Document-11013}.
+This hypothetical Special Programs processing summary is based on the Becker et al. (2011) 
+white paper to find outer solar system objects with shift-and stack (SAS) \citedsp{Document-11013}.
 
 Step 1. Data acquisition. \\
-The observational sequence is triggered. 
 In a single night, the 9 adjacent fields in a 3x3 grid are observed with 
-$336$ $\times$ $15$ second $r$-band exposures. 
-This sequence is always repeated 2-3 nights later. 
-This re-visit sequence is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later. 
-Data obtained in the $g$-band filter is also acceptable.
+$336$ $\times$ $15$ second $r$ or $g$-band exposures (168 standard visits). 
+These observations are repeated 2-3 nights later, and then this 2-night sequence
+is repeated 3 more times: 1.5 months, 3 months, and 13.5 months later. 
+They are not all at the same RA, Dec, but at selected ecliptic coordinates.
 
-Step 2. RDM prompt processing and alert production. \\
+Step 2. Regular Prompt processing and Alert Production. \\
 Each $2\times15$ second standard visit is processed by the Prompt pipeline 
@@
 All images and sources originating from this Special Program have 
-region and program labels, e.g., ``TNO Survey".
+region and program labels, e.g., ``SP-OSSO".
 
-Note that the results of prompt processing is not very relevant for this Special Program,
-which requires a year of dispersed observations before the processing pipelines 
-for shift-and-stack can be run, but including them in prompt processing means
+The results of Prompt processing are not very relevant for this Special Program's primary science goal,
+which requires a year of dispersed observations before the processing pipelines for shift-and-stack can be run.
+However, including these data in Prompt processing means that
 they can contribute to LSST's other time-domain and Solar System science goals.
 
-Step 3. RDM special processing with reconfigured pipelines. \\
+Step 3. Special Processing with reconfigured pipelines. \\
 None possible. 
-Shift-and-stack processing is beyond the scope of DM's algorithms.
+Shift-and-stack processing is beyond the scope of existing algorithms in the LSST Science Pipelines.
 
-Step 4. RDM inclusion in the WFD Data Release data products. \\
-All standard visits are reprocessed and included in the annual Data Release.
-Data products would include processed visit images and {\tt Source} and {\tt ForcedSource}
-catalog rows, along with difference images and {\tt DiaSource} catalog rows.
-All images and sources originating from this Special Program have 
-region and program labels, e.g., ``TNO Survey".
-As with all Special Programs data, these visits might also be included in the 
-WFD deeply coadded images if RDM decides it is both possible and beneficial.
+Step 4. Regular processing for inclusion in WFD program data products. \\
+Every year, each $2\times15$ second standard visit is reprocessed by the DIA data release pipelines
+and the results are included alongside WFD program data in the relevant DIA data products
+(e.g., processed visit images, difference images, associated source catalogs).
+In the first year after the Special Program is executed,
+Rubin Data Management finds that 10% of the standard visits from this Special Program
+had coordinates and image quality that help improve uniformity of the all-sky coadd,
+and so they are included.
+In later years, this fraction decreases (remember, this is \emph{hypothetical}).
+In all data releases, any and all processed images and catalog sources that originate in visits from this Special Program
+have the same region and program labels, e.g., ``SP-OSSO".
 
-Step 5. User processing and user-generated data products. \\
-The user-generated pipeline running the shift-and-stack processing would be set up and submitted 
-for batch processing by the user through the Science Platform or on an external processor. 
-Pipeline inputs would be the 336 processed exposures per field per re-visit sequence. 
-The LSST Science Pipelines difference imaging routine could be used with the same template tract/patch for all. 
-User-generated algorithms will shift the exposures and create difference images, and then the LSST
-Science Pipelines could be used to stack and do source detection and characterization, 
-and generate an object database. 
-Custom code will derive orbital parameters for the detections (or perhaps interface with the MPC),
-and store them in a separate {\tt SSObjects}-like database.
+Step 5. User-Generated Processing. \\
+The User-Generated Processing pipeline running the shift-and-stack processing is be set up and submitted 
+for batch processing by the user through the Science Platform or on an external system. 
+The pipeline's inputs are the processed visit images (and/or difference images) from Prompt processing.
+User-generated custom algorithms then shift-and-stack the images, and then the LSST Science Pipelines
+tasks are used to do source detection and characterization and create catalogs.
+User-generated custom code derives orbital parameters for the detections, and stores
+them in a user-generated catalog with a similar format to {\tt SSObjects}.
 
-\subsection{Searching for Supernovae in Deep Drilling Fields}\label{ssec:SPCS_SNDDF}
+\subsection{Deep Drilling Field}\label{ssec:SPCS_SNDDF}
 
 Step 1. Data acquisition. \\
-On a single deep drilling field, the scheduler obtains e.g., 5, 10, 10, 9, and 10 standard visits 
-with $2\times15$ second exposures in $grizy$ (or similar for the night's filter set) 
-and a small dither pattern between visits.
+In the COSMOS DDF, the scheduler obtains 10 standard visits in a row in each of the $griz$ filters
+with a small dither pattern between visits.
+This happens every other night for a three month season for four years.
 
-Step 2. RDM prompt processing and alert production. \\
-As in the example above, but all images and sources originating from this Special Program would have 
-region and program labels of, e.g., ``DDF Name".
+Step 2. Regular Prompt processing and Alert Production. \\
+Same as above, but the program label would be, e.g., ``SP-DDF-COSMOS".
 
-Step 3. RDM special processing with reconfigured pipelines. \\
-The required data products for this science goal can be met by reconfiguring the DM pipelines. 
-First, a template image of appropriate depth for ``nightly" difference imaging 
-would be made using RDM stacking algorithms. 
-On nights when this DDF is observed, at the end of the sequence of observations, 
-reconfigured RDM algorithms would be automatically triggered to create a nightly deep stack, 
-PSF-match it with the appropriately-made template, 
-create a nightly deep difference image, run source detection on it, and update separate 
-{\tt DiaObject}- and {\tt DiaSource}-like catalogs.
-All of this could happen in, e.g., 24 hours, but note there is no requirement on this timescale.
-On an annual basis, deeply coadded images that include all of the DDF's visits to dates would
-be generated, along with a separate {\tt Object}-like catalog for the DDF.
+Step 3. Special Processing with reconfigured pipelines. \\
+First, a template image of appropriate depth for ``nightly" difference imaging is created.
+At the end of each nightly sequence of observations, a pipeline based on recongfigured 
+components of the LSST Science Pipelines is automatically triggered.
+This pipeline creates nightly coadds in each filter and runs DIA using the template.
+Alerts are \emph{not} produced, but unique and separate catalogs with the same format
+as {\tt DiaObject} and {\tt DiaSource} are updated within 24 hours (not a requirement).
+At the end of each season, deeply coadded images that include all of the DDF's visits 
+from all years are re-generated, along with a separate {\tt Object}-like catalog.
+All images and catalogs are stored in separate butler collections and TAP tables from
+the WFD program data products.
 
-Step 4. RDM inclusion in the WFD Data Release data products. \\
-As in the example above, but all images and sources originating from this Special Program would have 
-region and program labels of, e.g., ``DDF Name".
+Step 4. Regular processing for inclusion in WFD program data products. \\
+Every year, each standard visit is reprocessed by the DIA data release pipelines
+and the results are included alongside WFD program data in the relevant DIA data products
+(e.g., processed visit images, difference images, associated source catalogs).
+Due to their small dither and lack of rotation, not even a single DDF image 
+is used to supplement the WFD program's all-sky coadd.
 
-Step 5. User processing and user-generated data products. \\
-For the science goal of searching for supernovae in nightly stacked DDF images, 
-no separate user-generated processing or data products appear to be necessary.
+Step 5. User-Generated Processing. \\
+In order to achieve a secondary science goal of finding very high-$z$ faint supernovae,
+a team of users reconfigure the LSST Science Pipelines to create weekly deep coadds
+of the COSMOS field an appropriate-depth template image, and to run DIA at the
+end of the season.
+These data products are stored in separate catalogs with the same format and schema as
+the {\tt DiaSource} and {\tt DiaObject} tables that are private to the team.
 
-\subsection{A Twilight Survey with Short Exposures}\label{ssec:SPCS_Twilight}
+\subsection{Short-exposure twilight survey}\label{ssec:SPCS_Twilight}
 
-Several kinds of twilight surveys with short exposures have been or might be proposed, 
-to put brighter stars (or transients such as supernovae) that saturate in a $15$ second image onto the 
-LSST photometric system and/or to take observations that are particularly well suited 
-(e.g., are 60 degrees from the sun) for finding Near-Earth Objects (NEOs).
-The processing case study for these is currently limited by unknowns.
+Twilight observations obtained at, e.g., 60 degrees from the Sun, are particularly
+well-suited for finding Near-Earth Objects (NEOs).
 
 Step 1. Data acquisition. \\
-At a specified time (or e.g., 6 degree twilight), the scheduler begins dither pattern of short exposures. 
-Location and exposure times are set by the sky brightness and desired saturation limits.
+At a specified time (or e.g., 6 degree twilight), the scheduler begins a dither pattern of 
+$2$-second exposures. 
+Coordinates and exposure times are set by the Sun distance, sky brightness, and desired saturation limits.
 
-Step 2. RDM prompt processing and alert production. \\
-Pending studies of short-exposure suitability for DIA (see Section \ref{sec:procbounds}) 
-these data could {\it potentially} be processed by the Prompt pipeline and lead to alerts.
+Step 2. Regular Prompt processing and Alert Production. \\
+Pending studies of DIA and Alert Production pipeline capabilities to process 
+short-exposure, high sky-background images (see Section~\ref{ssec:proc_bounds_processing}).
 
-Step 3. RDM special processing with reconfigured pipelines. \\
-So long as the short-exposure images can be processed and have enough stars for photometric 
-and astrometric calibration, reconfigured DM pipelines will probably be sufficient for 
-creating image and catalog products from this kind of data.
+Step 3. Special Processing with reconfigured pipelines. \\
+Pending studies of the LSST Science Pipelines capabilities to process 
+short-exposure, high sky-background images (see Section~\ref{ssec:proc_bounds_processing}).
 
-Step 4. RDM inclusion in the WFD Data Release data products. \\
-These short-exposure, high sky background images would not contribute to the DRP data products created for the WFD survey.
+Step 4. Regular processing for inclusion in WFD program data products. \\
+These short-exposure, high sky background images would not contribute to the data products created for the WFD program.
 
-Step 5. User processing and user-generated data products. \\
+Step 5. User-Generated Processing. \\
 If short-exposure images cannot be processed with the existing DM algorithms, 
-user-generated processing would be needed to reduce the raw data.
-
-\textbf{Side note:} A short-exposure survey of the bright stars of M67, described in Chapter 10.4 of the 
-Observing Strategy White Paper \citep{2017arXiv170804058L}, suggests using the stretch goal of 
-0.1 second exposures or, if that is not possible, \textit{"custom pixel masks to accurately perform 
-photometry on stars as much as 6 magnitudes brighter than the saturation level"}. 
-This would need user-generated processing.
+user-generated processing would be needed to reduce the raw data, and to 
+futher detect and characterize sources in the processed images.

### 2023-10-19 DMTN-065.tex
change appendix order
update abstract

@@
 Special Programs are additional survey areas and/or observing strategies that are driven by specific science 
-goals which build on, or are beyond, the core science pillars of the LSST.
-This document provides a summary of the requirements on Rubin Data Management for processing and serving data products
-related to Special Programs.
+goals which build on, or are beyond, the core science pillars of the Wide Fast Deep Main Survey.
+In order to meet the requirements and enable science related to Special Programs, this document provides 
+recommendations for Rubin Data Management regarding processing and serving data products for Special Programs.
 
 \smallskip
-Recommendations for the specific implementations needed to fulfill these requirements and maximize science from
-Special Programs are provided on a case-by-case basis, with examples.
 Hardware and processing boundaries on the potential diversity of data from Special Programs
-are also provided, along with a discussion of when user-generated processing 
-and data products will be needed to meet Special Programs' science goals.
+are discussed along with scenarios in which user-generated processing 
+and data products might be needed to meet Special Programs' science goals.
 }
@@
 \addtohist{2}{2021-12-01}{Updates per DM-20375.}{Melissa Graham}
-\addtohist{3}{2023-10-31}{Updates per DM-32723.}{Melissa Graham}
+\addtohist{3}{2023-10-19}{Updates per DM-32723.}{Melissa Graham}
 }
@@
 
+\input{ap_hardbounds.tex}
+
 \input{ap_procbounds.tex}
-
-\input{ap_hardbounds.tex}
 

### 2023-10-19 DMTN-065.tex
rm team

@@
 
-\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~P.~Guy, and the Data Management System Science Team}
+\author{M.~L.~Graham, Y.~AlSayyad, K.~Bechtol, E.~C.~Bellm, J.~F.~Bosch, J.~L.~Carlin, 
+G.~P.~Dubois-Felsmann, L.~P.~Guy, M.~Juri\'{c}, K.-T.~Lim, E.~S.~Rykoff, C.~T.~Slater}
 

### 2023-10-19 ap_spcs.tex
change appendix title
fix typo

@@
-\section{Scenarios for Processing Data from Special Programs}\label{sec:spcs}
+\section{Hypothetical Scenarios}\label{sec:spcs}
 
@@
 In the first year after the Special Program is executed,
-Rubin Data Management finds that 10% of the standard visits from this Special Program
+Rubin Data Management finds that 10\% of the standard visits from this Special Program
 had coordinates and image quality that help improve uniformity of the all-sky coadd,

### 2023-11-08 sec_intro.tex
update section titles
define Main Survey, add labels

@@
-\section{Executive Summary} \label{sec:execsum}
+\section{Memoranda} \label{sec:mem}
 
@@
 
-\item All visits should have a region label \textit{and} a program 
+\item All visits should have a region label \textit{and} an observing mode
 label, and these labels should be propagated to processed images and catalog data.
@@
 
-\section{Introduction}\label{sec:intro}
+\section{Terms and Definitions}\label{sec:intro}
 
@@
 
-\subsection{Wide Fast Deep (WFD) Program}
+\subsection{Region Label}
 
-The WFD is the core science program of the LSST, designed to achieve the science 
+Based on Right Acension and Declination.
+
+To be applied to data by Rubin Data Management, e.g., using a look-up table.
+
+Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
+Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
+
+\subsection{Observing Mode Label}
+
+Based on the scheduler mode at the time of the observation
+(i.e., survey cadence or strategy).
+
+To be applied by the Survey Scheduler team via the observations scheduler.
+
+Observing mode labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
+options like engineering, commissioning, and director's discretionary.
+
+\subsection{Main Survey and Wide-Fast-Deep (WFD)}
+
+The LSST Main Survey includes the low-dust Wide-Fast-Deep (WFD) area 
+(which includes low-dust extragalactic and Galactic regions) 
+and several special regions:
+the dusty areas of the Galactic bulge and plane; the South Celestial Pole (SCP); 
+the North Ecliptic Spur (NES); and the Virgo Cluster \citedsp{PSTN-055}.
+The WFD component is the core component of the LSST Main Survey, designed to achieve the science 
 goals defined by the Science Requirements Document (SRD; \citeds{LPM-17}).
 
-The WFD program is defined from a scheduling perspective as a certain observing strategy
-(i.e., a set of cadences and filter balances) applied to a low-dust region 
-(extragalactic areas) and several special regions:
-the Galactic bulge and plane; the South Celestial Pole (SCP); the North Ecliptic Spur (NES);
-and the Virgo Cluster \citedsp{PSTN-055}.
+The Main Survey will be executed with a set of observing modes, and 
+is expected to use alternative standard visits in the $u$-band and 
+standard visits in all other bands \citedsp{PSTN-055}.
+Different cadences, filter balances, rotational dithers, etc., 
+might be applied in the different regions composing Main Survey.
+For example, rolling cadence is expected to be implemented in the low-dust 
+WFD region only.
 
-The WFD is expected to use alternative standard visits in the $u$-band and 
-standard visits in all other bands \citedsp{PSTN-055}.
-Different cadences, filter balances, rotational dithers, etc., might
-be applied in the different regions composing WFD.
-For example, rolling cadence is expected to be implemented in the low-dust region only.
-
-WFD program data will be processed with regular (non-special) processing.
+Data from the Main Survey will be processed with regular (non-special) processing.
 This processing will produce, among other data products (LSE-163), 
 a contiguous sky footprint that covers at 
-least $\sim$18000 deg$^2$ with $\gtrsim$800 visits per field, 
+least $\sim$18000 deg$^2$ (with $\gtrsim$800 visits per field in WFD regions), 
 and is expected to be accomplished with 85--90\% of the observing time.
@@
 described in the Data Products Definitions Document
-(LSE-163) that are designed for, and will be applied to, the WFD program's observations.
+(LSE-163) that are designed for, and will be applied to, the Main Survey's observations.
 
@@
 This is a Rubin Data Management term used to refer to sky regions within
-or beyond the WFD footprint which have a distinct observing strategy from
-the WFD program's plans for that region (e.g., different visit type, cadence, filter distribution).
+or beyond the Main Survey footprint which have a distinct observing strategy from
+the Main Survey's plans for that region (e.g., different visit type, cadence, filter distribution).
 
@@
 add to the core science pillars of the LSST.
-They include non-WFD LSST components such as the Deep Drilling 
+They include LSST components such as the Deep Drilling 
 Fields (DDFs) and the mini-, micro-, and nano-surveys \citedsp{PSTN-055}.
@@
 Data products from Special Programs is subject to the Rubin Data Policy \citedsp{RDO-013}
-in the same way as data products from the WFD program.
+in the same way as data products from the Main Survey.
 
@@
 Special Processing creates data products that are unique and separate from those produced
-by regular processing for the WFD program.
+by regular processing for the Main Survey.
 
@@
 
-The sky areas of mini-surveys can be within, adjacent to, or detached from the WFD footprint.
+The sky areas of mini-surveys can be within, adjacent to, or detached from the Main Survey footprint.
 Mini-surveys can have non-standard visits.

### 2023-11-08 sec_sci.tex
add observing mode term

@@
 
-\subsection{Region and Program Labels}\label{ssec:sci_labels}
+\subsection{Region and Observing Mode Labels}\label{ssec:sci_labels}
 
-\textbf{All visits should have a region label \textit{and} a program 
+\textbf{All visits should have a region label \textit{and} an observing mode 
 label, and these labels should be propagated to processed images and catalog data.}
 
-\textbf{Region Labels -- } based on Right Acension and Declination.
-Region labels should include, e.g., WFD low-dust, Galactic Plane, Galactic 
+\textbf{Region Labels -- } 
+Based on Right Acension and Declination.
+To be applied to data by Rubin Data Management, e.g., using a look-up table.
+Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
 Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
@@
 
-\textbf{Program Labels -- } 
-Based on survey cadeance or strategy.
-Program labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
+\textbf{Observing Mode Labels -- } 
+Based on the scheduler mode at the time of the observation
+(i.e., survey cadence or strategy).
+To be applied by the Survey Scheduler team via the observations scheduler.
+Observing mode labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
 options like engineering, commissioning, and director's discretionary.
-Program labels should be propagated to single-visit processed and 
+Observing mode  labels should be propagated to single-visit processed and 
 difference images and catalog sources.
-It would not make sense to propagate program labels to deep coadds or 
-object tables, as these data products could be a mix of programs.
+It would not make sense to propagate observing mode labels to deep coadds or 
+object tables, as these data products could be a mix of modes.
 
@@
 \item To enable provenance when Special Programs data in included in regular
-processing and the WFD program data products (e.g., if used to improve the
+processing and the Main Survey's data products (e.g., if used to improve the
 all-sky coadd).
@@
 cannot be met by including the data in regular processing (e.g., Prompt processing),
-or where doing so would compromise the WFD data products (e.g., introduce non-uniformity).
+or where doing so would compromise the Main Survey data products (e.g., introduce non-uniformity).
 
@@
 enhanced by any and all additional observations, even if they are not
-optimized for the science (as the WFD program's cadence is being optimized).
+optimized for the science (as the WFD area's cadence is being optimized).
 
@@
 Users will need to be able to query for data that restricts by sky region and 
-program label. 
+observign mode label. 
 This can be accomplished by including those labels in all image and catalog 
@@
 This could be accomplished with, e.g., a toggle to overlay region boundaries
-for Special Programs when browsing all-sky coadds built from WFD program data
+for Special Programs when browsing all-sky coadds built from Main Survey data
 in the Portal Aspect.

### 2023-11-09 sec_proc.tex
fix program/WFD terms

@@
 and refers to the image processing pipelines that are designed for, 
-and will be applied to, the WFD program's observations (Section~\ref{sec:intro}).
+and will be applied to, the Main Survey's observations (Section~\ref{sec:intro}).
 
 Decisions about when to apply regular processing to Special Programs data,
-or when to include it in the data products for the WFD program 
+or when to include it in the data products for the Main Survey
 (e.g., if it improves the all-sky coadds), 
@@
 
-\textbf{How WFD and Special Program data would co-exist.}
+\textbf{How Main Survey and Special Program data would co-exist.}
 No ``unique and separate" data products for the Special Progams would be 
@@
 Special Programs data that is processed by the Prompt pipeline would 
-contribute to the Prompt data products for the WFD program as 
+contribute to the Prompt data products for the Main Survey as 
 described in Section 3 of the DPDD \citedsp{lse-163}. 
@@
 Including visits from Special Programs in regular Prompt processing alongside
-vists from the WFD program is not, in general, anticipated to affect WFD science goals.
-For example, analyses for a WFD-only subset could still be done using the program and
-region labels described in Section~\ref{ssec:sci_labels}, which would be
+vists from the Main Survey is not, in general, anticipated to affect Main Survey 
+or WFD-specific science goals.
+For example, analyses for a WFD-only subset could still be done using the observing mode
+and region labels described in Section~\ref{ssec:sci_labels}, which would be
 propogated to difference images, difference-image catalogss, and alerts.
@@
 stream bandwidth is sized for the expected histories for
-WFD program fields. 
+Main Survey fields. 
 The Prompt pipeline resources are also sized for the
-WFD program, and it might not be possible to load up
+Main Survey, and it might not be possible to load up
 thousands of epochs at a time.
@@
 Alerts from Special Programs data are subject to the same latency
-requirements as WFD program alerts (Section~\ref{ssec:req_proc}).
+requirements as Main Survey alerts (Section~\ref{ssec:req_proc}).
 
@@
 
-These two potential issues pose challenges, but are not necessarily showstoppers in processing 
-Special Programs data with the Prompt pipelines. 
+These two potential issues pose challenges, but are not necessarily showstoppers 
+in processing Special Programs data with the Prompt pipelines. 
 The overall impact on time-domain science would still be positive, even 
 if mitigations are needed for these issues.
-For example, brokers and users would be able to use the region and program labels
-in the data as context (i.e., as flags) and avoid including 
+For example, brokers and users would be able to use the region and observing mode 
+labels in the data as context (i.e., as flags) and avoid including 
 limited-history or potentially-compromised {\tt DiaSources} in their
@@
 \textbf{Time-domain DIA data products --}
-This includes the results of the annual reprocessing of WFD program data with 
+This includes the results of the annual reprocessing of Main Survey data with 
 Difference Image Analysis (DIA), and the production of Data Release
@@
 \textbf{Static-sky data products --}
-This includes the tessellation and coaddition of WFD program images
+This includes the tessellation and coaddition of Main Survey images
 and the associated multi-band {\tt Object} catalog and survey property maps.
@@
 Any tables for the unique and separate data products should be joinable to the 
-data products for the WFD program, when possible.
+data products for the Main Survey, when possible.
 This is a requirement (Section~\ref{ssec:req_dp}).
@@
 \textbf{Survey property maps --}
-As for the tesslated all-sky coadded images made from WFD program data, survey property maps
+As for the tesslated all-sky coadded images made from Main Survey data, survey property maps
 should be made individually as part of any Special Programs that generates
@@
 User-Generated Processing.
-Note that for the WFD program data, such custom coadds are also considered 
+Note that for the Main Survey data, such custom coadds are also considered 
 as beyond scope and in need of User-Generated Processing.
@@
 Consider a Special Program in which a special region of sky is observed with standard
-visits but a a special strategy or cadence which is significantly distinct from the WFD program,
-and lasts for a limited amount of time.
+visits but a a special strategy or cadence which is significantly distinct from the Main Survey
+observing modes, and lasts for a limited amount of time.
 For example, a short-term survey of the Magellanic Clouds, with dual primary science

### 2023-11-09 ap_procbounds.tex
update section title
WFD --> Main Survey

@@
-\section{Potential Processing Boundaries on Data Diversity}\label{sec:procbounds}
+\section{Potential Processing Boundaries}\label{sec:procbounds}
 
@@
 
-Images which deviate significantly from the $15$ second duration for the WFD main survey 
+Images which deviate significantly from the $15$ second duration for the Main Survey 
 may encounter issues in the instrument signature removal routine, in the correction for 
@@
 However, a $2\times150$ second exposure would saturate at $r \approx 18.3$, perhaps 
-leaving too few stars overlapping with e.g., templates or WFD images, for astrometric and 
+leaving too few stars overlapping with e.g., templates or Main Survey images, for astrometric and 
 photometric calibrations.

### 2023-11-09 ap_spcs.tex
WFD program --> Main Survey

@@
 Step 3. Special Processing with reconfigured pipelines. \\
-Step 4. Regular processing for inclusion in WFD program data products. \\
+Step 4. Regular processing for inclusion in Main Survey data products. \\
 Step 5. User-Generated Processing. \\
@@
 All images and sources originating from this Special Program have 
-region and program labels, e.g., ``SP-OSSO".
+region and observing mode labels, e.g., ``SP-OSSO".
 
@@
 
-Step 4. Regular processing for inclusion in WFD program data products. \\
+Step 4. Regular processing for inclusion in Main Survey data products. \\
 Every year, each $2\times15$ second standard visit is reprocessed by the DIA data release pipelines
-and the results are included alongside WFD program data in the relevant DIA data products
+and the results are included alongside Main Survey data in the relevant DIA data products
 (e.g., processed visit images, difference images, associated source catalogs).
@@
 In all data releases, any and all processed images and catalog sources that originate in visits from this Special Program
-have the same region and program labels, e.g., ``SP-OSSO".
+have the same region and observing mode labels, e.g., ``SP-OSSO".
 
@@
 Step 2. Regular Prompt processing and Alert Production. \\
-Same as above, but the program label would be, e.g., ``SP-DDF-COSMOS".
+Same as above, but the observing mode label might be, e.g., ``SP-DDF-COSMOS".
 
@@
 All images and catalogs are stored in separate butler collections and TAP tables from
-the WFD program data products.
+the Main Survey data products.
 
-Step 4. Regular processing for inclusion in WFD program data products. \\
+Step 4. Regular processing for inclusion in Main Survey data products. \\
 Every year, each standard visit is reprocessed by the DIA data release pipelines
-and the results are included alongside WFD program data in the relevant DIA data products
+and the results are included alongside Main Survey data in the relevant DIA data products
 (e.g., processed visit images, difference images, associated source catalogs).
 Due to their small dither and lack of rotation, not even a single DDF image 
-is used to supplement the WFD program's all-sky coadd.
+is used to supplement the Main Survey's all-sky coadd.
 
@@
 
-Step 4. Regular processing for inclusion in WFD program data products. \\
-These short-exposure, high sky background images would not contribute to the data products created for the WFD program.
+Step 4. Regular processing for inclusion in Main Survey data products. \\
+These short-exposure, high sky background images would not contribute to the data products created for the Main Survey.
 

### 2024-03-01 ap_procbounds.tex
Apply suggestions from code review
Co-authored-by: Leanne Guy <5301176+leannep@users.noreply.github.com>

@@
 \begin{itemize}
-\item \bf{Short or long exposures.} Very short ($<$2 sec) exposures could be difficult to 
+\item \textbf{Short or long exposures.} Very short ($<$2 sec) exposures could be difficult to 
 process due to an incompletely-formed PSF. Short (or long, $>$150 sec) exposures 
 could be difficult to calibrate due to having too few (or too few unsaturated) stars.
-\item \bf{Bright backgrounds (twilight).} It is currently unclear whether images with very 
+\item \textbf{Bright backgrounds (twilight).} It is currently unclear whether images with very 
 bright sky backgrounds (twilight images) can be processed with the LSST Science Pipelines, 
 or whether user generated pipelines will be needed.
-\item \bf{Non-sidereal tracking.} The full reduction and calibration of images obtained 
+\item \textbf{Non-sidereal tracking.} The full reduction and calibration of images obtained 
 with non-sidereal tracking, in which the stars are streaked, is currently beyond the scope 

### 2024-03-01 sec_intro.tex
Apply suggestions from code review
Co-authored-by: Leanne Guy <5301176+leannep@users.noreply.github.com>

@@
 
-\subsection{Main Survey and Wide-Fast-Deep (WFD)}
+\subsection{Main Survey including Wide-Fast-Deep (WFD)}
 
@@
 This is a term used only in this document to refer to the image processing 
-described in the Data Products Definitions Document
-(LSE-163) that are designed for, and will be applied to, the Main Survey's observations.
+described in the Data Management Science Pipelines Design \citeds{LDM-151} that produces the data products described in the Data Products Definitions Document
+(\citeds{LSE-163}) that are designed for, and will be applied to, the Main Survey's observations.
 
@@
 This is a Rubin Data Management term used to refer to sky regions within
-or beyond the Main Survey footprint which have a distinct observing strategy from
+or beyond the Main Survey footprint that have a distinct observing strategy from
 the Main Survey's plans for that region (e.g., different visit type, cadence, filter distribution).

### 2024-03-01 sec_proc.tex
Apply suggestions from code review
Co-authored-by: Leanne Guy <5301176+leannep@users.noreply.github.com>

@@
 Special Programs that use short exposures would be limited to the
-minimum exposure time is 1 second (stretch goal: 0.1 seconds\footnote{A short-exposure survey 
+minimum exposure time of 1 second (stretch goal: 0.1 seconds\footnote{A short-exposure survey 
 of the bright stars of M67, described in Chapter 10.4 of the 
@@
 \item \textbf{Repeated pointing --}
-Special Programs which require the \emph{exact same} field pointing and 
+Special Programs that require the \emph{exact same} field pointing and 
 rotation for \emph{every exposure} (to sub-arcsecond levels) might run 
@@
 \item \textbf{Very short exposures --}
-Special Programs which use very short ($<$2 sec) exposures 
+Special Programs that use very short ($<$2 sec) exposures 
 could be difficult to reduce due to an incompletely-formed PSF 
@@
 \item \textbf{Twilight images --}
-For Special Programs which obtain images with very bright sky backgrounds
+For Special Programs that obtain images with very bright sky backgrounds
 (twilight images), it is currently unclear whether they can be processed
@@
 are ultimately left to the discretion of the Rubin Observatory's 
-Data Management and System Performance teams.
+Data Management and System Performance departments in Rubin Operations.
 
@@
 
-Options for RDM to process TOO observations, especially during the first year of Operations 
+Options for Data Management to process TOO observations, especially during the first year of Operations 
 when the template coverage will be low, are discussed in more detail in 

### 2024-03-01 sec_sci.tex
Apply suggestions from code review
Co-authored-by: Leanne Guy <5301176+leannep@users.noreply.github.com>

@@
 cannot be met by including the data in regular processing (e.g., Prompt processing),
-or where doing so would compromise the Main Survey data products (e.g., introduce non-uniformity).
+or where doing so would compromise the Main Survey data products (e.g., introduce additional non-uniformity beyond what would be expected based on the survey strategy).
 
@@
 Users will need to be able to query for data that restricts by sky region and 
-observign mode label. 
+observing mode label. 
 This can be accomplished by including those labels in all image and catalog 

### 2024-03-01 sec_intro.tex
Update sec_intro.tex

@@
 
-Data from the Main Survey will be processed with regular (non-special) processing.
+Data from the Main Survey will be processed with standard (non-special) processing.
 This processing will produce, among other data products (LSE-163), 
@@
 and will be of variable depth, as most special regions recieve fewer visits.
-
-\subsection{Regular (non-special) processing}
-
-This is a term used only in this document to refer to the image processing 
-described in the Data Management Science Pipelines Design \citeds{LDM-151} that produces the data products described in the Data Products Definitions Document
-(\citeds{LSE-163}) that are designed for, and will be applied to, the Main Survey's observations.
-
-In some cases, regular processing is also appropriate for Special Programs.
 
@@
 
-Some science goals for Special Programs can be met with regular processing,
+Some science goals for Special Programs can be met with standard processing,
 but some will require Special Processing by Rubin Data Management or user-generated processing.
@@
 Special Processing creates data products that are unique and separate from those produced
-by regular processing for the Main Survey.
+by standard processing for the Main Survey.
 
 Special Processing is likely to use different inputs or configurations for the
-LSST Science Pipelines, or to run on different timescales, than regular processing - 
+LSST Science Pipelines, or to run on different timescales, than standard processing - 
 as appropriate for the Special Programs' data and science goals.
@@
 is beyond the scope of Special Processing.
+
+\subsection{Standard (non-special) processing}
+
+This is a term used only in this document to refer to the image processing 
+described in the Data Management Science Pipelines Design \citeds{LDM-151} that produces the data products described in the Data Products Definitions Document
+(\citeds{LSE-163}) that are designed for, and will be applied to, the Main Survey's observations.
+
+In some cases, standard processing is also appropriate for Special Programs.
 
@@
 
-Processing for the DDF images is likely to be a combination of regular (Prompt) processing,
+Processing for the DDF images is likely to be a combination of standard (Prompt) processing,
 Special Processing, and User-Generated Processing.
@@
 
-Processing for the mini-surveys is likely to be a combination of regular (Prompt) processing,
+Processing for the mini-surveys is likely to be a combination of standard (Prompt) processing,
 Special Processing, and User-Generated Processing.

### 2024-03-01 sec_intro.tex
Update sec_intro.tex

@@
 
-Data from the Main Survey will be processed with standard (non-special) processing.
+Data from the Main Survey will be processed with standard (non-special) processing (Section~\ref{ssec:intro_stdproc}).
 This processing will produce, among other data products (LSE-163), 
@@
 
-\subsection{Standard (non-special) processing}
+Examples of Special Processing include the processing
+of images with non-standard exposure times, images obtained
+during twilight, or nightly stacking and differencing
+of images in deep drilling fields.
+See \Section~\ref{ssec:proc_special} for more detailed examples of Special Processing.
+
+\subsection{Standard (non-special) processing}\label{ssec:intro_stdproc}
 

### 2024-03-01 sec_proc.tex
caps LSE

@@
 contribute to the Prompt data products for the Main Survey as 
-described in Section 3 of the DPDD \citedsp{lse-163}. 
+described in Section 3 of the DPDD \citedsp{LSE-163}. 
 These data products are the results of Difference Image Analysis (DIA),

### 2024-03-01 DMTN-065.tex

--- 2023-10-19 DMTN-065.tex
@@
 
-\author{M.~L.~Graham, M.~Juri\'{c}, K.-T.~Lim, E.~Bellm, G.~Dubois-Felsmann, L.~P.~Guy, and the Data Management System Science Team}
+\author{M.~L.~Graham, Y.~AlSayyad, K.~Bechtol, E.~C.~Bellm, J.~F.~Bosch, J.~L.~Carlin, 
+G.~P.~Dubois-Felsmann, L.~P.~Guy, M.~Juri\'{c}, K.-T.~Lim, E.~S.~Rykoff, C.~T.~Slater,
+and the Data Management System Science Team}
 

### 2024-03-01 ap_spcs.tex

@@
 \textbf{\emph{Hypothetical}} examples of Special Programs
-and the regular, Special, and User-Generated processing
+and the standard, Special, and User-Generated processing
 to illustrate what \emph{might be done}.
@@
 Step 1. Data acquisition. \\
-Step 2. Regular Prompt processing and Alert Production. \\
+Step 2. Standard Prompt processing and Alert Production. \\
 Step 3. Special Processing with reconfigured pipelines. \\
-Step 4. Regular processing for inclusion in Main Survey data products. \\
+Step 4. Standard processing for inclusion in Main Survey data products. \\
 Step 5. User-Generated Processing. \\
@@
 
-Step 2. Regular Prompt processing and Alert Production. \\
+Step 2. Standard Prompt processing and Alert Production. \\
 Each $2\times15$ second standard visit is processed by the Prompt pipeline 
@@
 
-Step 4. Regular processing for inclusion in Main Survey data products. \\
+Step 4. Standard processing for inclusion in Main Survey data products. \\
 Every year, each $2\times15$ second standard visit is reprocessed by the DIA data release pipelines
@@
 
-Step 2. Regular Prompt processing and Alert Production. \\
+Step 2. Standard Prompt processing and Alert Production. \\
 Same as above, but the observing mode label might be, e.g., ``SP-DDF-COSMOS".
@@
 
-Step 4. Regular processing for inclusion in Main Survey data products. \\
+Step 4. Standard processing for inclusion in Main Survey data products. \\
 Every year, each standard visit is reprocessed by the DIA data release pipelines
@@
 
-Step 2. Regular Prompt processing and Alert Production. \\
+Step 2. Standard Prompt processing and Alert Production. \\
 Pending studies of DIA and Alert Production pipeline capabilities to process 
@@
 
-Step 4. Regular processing for inclusion in Main Survey data products. \\
+Step 4. Standard processing for inclusion in Main Survey data products. \\
 These short-exposure, high sky background images would not contribute to the data products created for the Main Survey.

### 2024-03-01 sec_proc.tex

@@
 Appendix~\ref{sec:spcs} provides detailed examples for the processing
-of data from Special Programs, including scenarios in which regular,
+of data from Special Programs, including scenarios in which standard,
 Special, and User-Generated Processing are all involved.
@@
 
-\subsection{Regular (non-special) processing}\label{ssec:proc_reg}
+\subsection{Standard (non-special) processing}\label{ssec:proc_reg}
 
-Recall that the term ``regular processing" is used only within this document,
+Recall that the term ``standard processing" is used only within this document,
 and refers to the image processing pipelines that are designed for, 
@@
 
-Decisions about when to apply regular processing to Special Programs data,
+Decisions about when to apply standard processing to Special Programs data,
 or when to include it in the data products for the Main Survey
@@
 No ``unique and separate" data products for the Special Progams would be 
-produced by regular Prompt processing.
+produced by standard Prompt processing.
 Special Programs data that is processed by the Prompt pipeline would 
@@
 into {\tt DiaObjects}, and alert packets.
-Including visits from Special Programs in regular Prompt processing alongside
+Including visits from Special Programs in standard Prompt processing alongside
 vists from the Main Survey is not, in general, anticipated to affect Main Survey 
@@
 generated, but this is not necessary as the science goals for the
-mini-survey can be met by processing its data with regular Prompt
+mini-survey can be met by processing its data with standard Prompt
 Processing, and ensuring the data is properly labeled.

### 2024-03-01 sec_sci.tex

@@
 
-\textbf{All visits should have a region label \textit{and} an observing mode 
+\textbf{All visits should have at least one region label \textit{and} one observing mode 
 label, and these labels should be propagated to processed images and catalog data.}
@@
 difference, and deep coadds), catalog sources, and catalog static-sky objects.
+A given image might have more than one region label due to overlaping regions,
+but they could be, e.g., held in a single string as a CSV.
 
@@
 object tables, as these data products could be a mix of modes.
+It is left to the discretion of the Survey Scheduler team whether more than one
+observing mode labels might be applied at a time and, e.g., held in a single string as a CSV.
+
+A given image would probably only have one observing mode label
 
@@
 data associated with a specific Special Program, and meet the science goals
-of that Special Program, when regular processing as been applied 
+of that Special Program, when standard processing as been applied 
 (e.g., Prompt Processing, Section~\ref{ssec:sci_pproc}).
 
-\item To enable provenance when Special Programs data in included in regular
+\item To enable provenance when Special Programs data in included in standard
 processing and the Main Survey's data products (e.g., if used to improve the
@@
 \textbf{Necessary -- } When the primary science goal for a Special Program 
-cannot be met by including the data in regular processing (e.g., Prompt processing),
+cannot be met by including the data in standard processing (e.g., Prompt processing),
 or where doing so would compromise the Main Survey data products (e.g., introduce additional non-uniformity beyond what would be expected based on the survey strategy).
@@
 
-Further discussion of regular Prompt processing of Special Programs data, with
+Further discussion of standard Prompt processing of Special Programs data, with
 examples and a discussion of a few challenges,

### 2024-03-26 sec_sci.tex
update region labels

@@
 \textbf{Region Labels -- } 
-Based on Right Acension and Declination.
-To be applied to data by Rubin Data Management, e.g., using a look-up table.
+Applied to a visit based on its Right Acension and Declination.
+To be applied to data by Rubin Data Management using, e.g., a look-up table.
 Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
 Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
-Region labels should be propagated to processed images (visit, 
-difference, and deep coadds), catalog sources, and catalog static-sky objects.
-A given image might have more than one region label due to overlaping regions,
-but they could be, e.g., held in a single string as a CSV.
+A given visit is likely to have multiple region labels because, for example,
+the DDFs are within the WFD low-dust region.
+These could be stored in, e.g., a single string as a CSV.
+
+The region labels should be available to users as query constraints for
+all images and tables, via both the butler and the TAP service, but otherwise
+the details of region label implementation are left to Rubin Data Management.
+
+For example, for images, the region labels could be propagated to processed images
+(e.g., visit, difference, and deep coadds), so that users may query the ObsTAP 
+service for all images with a given region label.
+For catalogs, the region labels could be stored for individual sources and
+objects as a table column, or for TAP-accessible tables stored in a joinable
+table such as the Visit, ccdVisit, or coaddPatches tables.
+Another option may be to map regions into spatial divisions (htm or healpix),
+and enable users to query image and catalog products via their htm.
 

### 2024-03-26 sec_sci.tex
more updates to region labels

@@
 \textbf{All visits should have at least one region label \textit{and} one observing mode 
-label, and these labels should be propagated to processed images and catalog data.}
+label, and these labels should be available to users as query constraints for
+all images and tables.}
 
@@
 Applied to a visit based on its Right Acension and Declination.
+
 To be applied to data by Rubin Data Management using, e.g., a look-up table.
@@
 table such as the Visit, ccdVisit, or coaddPatches tables.
-Another option may be to map regions into spatial divisions (htm or healpix),
-and enable users to query image and catalog products via their htm.
+Another option may be to map regions into spatial divisions (HTM or HEALPix),
+and enable users to query image and catalog products this way.
+
+The spatial resolution of region labels is left to be determined as it depends
+somewhat on the implementation.
+In this document the region label is considered to be applied to a visit
+if the visit overlaps the region, as a very simple implementation.
+However, the field of view is so large that a finer resolution by ccd,
+or by a smaller-mesh HTM map if the regions are determined for sources and objects,
+would be possible (and more useful to users).
 

### 2024-03-26 sec_intro.tex
fix typo
update summary, labels text

@@
-\section{Memoranda} \label{sec:mem}
+\section{Summary} \label{sec:mem}
 
-The three memoranda for Rubin Data Management regarding Special Programs are:
+The three main takeaways for Rubin Data Management regarding Special Programs are:
 
@@
 \item All visits should have a region label \textit{and} an observing mode
-label, and these labels should be propagated to processed images and catalog data.
+label, and these labels should be available to users as query constraints for
+all images and tables.
 (Section~\ref{ssec:sci_labels})
@@
 \item Special Processing should be done by Rubin Data Management to
-produce unique and separate data products
-for Special Programs when it is both possible and necessary.
-(Section~\ref{ssec:sci_sproc}).
+produce unique and separate data products for Special Programs in
+certain cases that meet specific definitions of "possible and necessary".
+(Section~\ref{ssec:sci_sproc})
 
@@
 
-Based on Right Acension and Declination.
-
-To be applied to data by Rubin Data Management, e.g., using a look-up table.
-
-Region labels would include, e.g., WFD low-dust, Galactic Plane, Galactic 
-Bulge, SCP, NES, or Virgo; DDF fieldname; mini-survey region name.
+Applied to a visit based on its Right Acension and Declination (Section~\ref{ssec:sci_labels}).
 
@@
 
-Based on the scheduler mode at the time of the observation
-(i.e., survey cadence or strategy).
-
-To be applied by the Survey Scheduler team via the observations scheduler.
-
-Observing mode labels might include, e.g., WFD, DDF, TOO, mini-survey name, or other 
-options like engineering, commissioning, and director's discretionary.
+Applied to a visit based on the scheduler mode at the time of the observation,
+i.e., survey cadence or strategy (Section~\ref{ssec:sci_labels}).
 
@@
 of images in deep drilling fields.
-See \Section~\ref{ssec:proc_special} for more detailed examples of Special Processing.
+See Section~\ref{ssec:proc_special} for more detailed examples of Special Processing.
 

### 2024-03-26 sec_sci.tex
update obs mode text

@@
 \textbf{Observing Mode Labels -- } 
-Based on the scheduler mode at the time of the observation
+Applied to a visit based on the scheduler mode at the time of the observation
 (i.e., survey cadence or strategy).
+
 To be applied by the Survey Scheduler team via the observations scheduler.
@@
 options like engineering, commissioning, and director's discretionary.
-Observing mode labels should be propagated to single-visit processed and 
-difference images and catalog sources.
-It would not make sense to propagate observing mode labels to deep coadds or 
-object tables, as these data products could be a mix of modes.
-It is left to the discretion of the Survey Scheduler team whether more than one
+A given image would probably only have one observing mode label, but
+it is left to the discretion of the Survey Scheduler team whether more than one
 observing mode labels might be applied at a time and, e.g., held in a single string as a CSV.
 
-A given image would probably only have one observing mode label
+The observing mode labels should be available to users as query constraints for
+all single-visit processed and difference images and catalog sources,
+via both the butler and the TAP service.
+It would not make sense to apply observing mode labels to deep coadds or 
+object tables, as these data products could be a mix of modes.
 

### 2024-03-27 sec_proc.tex
update 4.2.1

@@
 into hardware boundaries on pointing and tracking.
+Atmospheric distortions also pose a major limitation on positional
+repeatability across the focal plane.
 
@@
 (Section~\ref{ssec:proc_bounds}).
-Alert latency requirements do apply to Special Programs data (Section~\ref{ssec:req_proc}),
-and if they cannot be met then Prompt processing is ``not possible".
+
+If, for example, Prompt processing takes longer on Special Programs data or short exposures demand
+additional processing because the image acquisition rate exceeds the nominal $\sim$2 per minute,
+the available compute capacity of Prompt processing might be saturated.
+This could compromise the latency of alert distribution while Special Programs are being executed,
+and could also impact alert distribution for subsequent Main Survey visits.
+It is left to the discretion of Rubin Data Management to prioritize alert distribution
+for Main Survey visits by potentially limiting Prompt processing for Special Programs in such cases.
 
@@
 (i.e., the labels described in Section~\ref{ssec:sci_labels}).
-However, there would not be enough memory to hold alternative templates 
-for the whole sky.
+Potential processes for the creation and verification of such alternative
+templates, the cost-benefit of allocating disk storage space to hold
+alternative templates, and decisions about when a science goal requires specialized
+templates, are left to the discretion of Rubin Data Management.
 

### 2024-03-27 sec_proc.tex
remove timescales from S.4.3.2

@@
 \begin{itemize}
-\item nightly-coadded images (24 h)
-\item nightly-coadded difference images (24 h)
-\item {\tt DiaSource}- and {\tt DiaObject}-like catalogs for the nightly-coadds (24 h)
-\item deeply-coadded images (all images to date; yearly)
-\item templates for the nightly-coadded difference images (yearly)
-\item {\tt Source}- and {\tt Object}-like catalogs for the nightly-coadded and deeply-coadded images (yearly)
+\item nightly-coadded images
+\item nightly-coadded difference images
+\item {\tt DiaSource}- and {\tt DiaObject}-like catalogs for the nightly-coadds
+\item deeply-coadded images
+\item templates for the nightly-coadded difference images
+\item {\tt Source}- and {\tt Object}-like catalogs for the nightly-coadded and deeply-coadded images
 \end{itemize}
+
+The release of nightly-coadded or difference images is subject to the 80-hour embargo.
+The timescales for updating catalogs based on the nightly-coadds could be daily
+or on another intermediate timescale; this is left to the discretion of Rubin Data Management.
 

### 2024-03-27 sec_proc.tex
add ref to sssec:proc_reg_prompt

@@
 For the NEO twilight survey, these data products would likely be similar to the
-Prompt or DIA data products.
+Prompt or DIA data products, but note the potential issues with alert
+latency from short exposures mentioned in Section~\ref{sssec:proc_reg_prompt}.
 For the calibration survey, these data products would like be similar to the

### 2024-03-27 sec_proc.tex
fix AP restriction text

@@
 \textbf{Alert production is restricted -- }
-User-Generated Processing will not be able to release alert packets in the LSST alert stream.
-As the latency on processed visit image availability has an 80-hour
-embargo, no user-generated pipeline will be able to process Special Programs 
-data on a timescale similar to prompt processing and alert production 
-(60 seconds to 24 hours).
-Thus, no User-Generated Processing may contribute alerts to the LSST alert stream on 
-any timescale.
+As processed visit images will embargoed for 80 hours,
+only official Rubin Prompt processing may produce alerts.
+No User-Generated Processing may contribute alerts to the LSST alert stream.
 

### 2024-03-27 sec_proc.tex
fix typo

@@
 \textbf{Alert production is restricted -- }
-As processed visit images will embargoed for 80 hours,
+As processed visit images will be embargoed for 80 hours,
 only official Rubin Prompt processing may produce alerts.

### 2024-03-27 sec_intro.tex
JB comments

@@
 The Main Survey will be executed with a set of observing modes, and 
-is expected to use alternative standard visits in the $u$-band and 
-standard visits in all other bands \citedsp{PSTN-055}.
+has been recommended by the Survey Cadence Optimization Committee (SCOC)
+to use alternative standard visits in the $u$-band;
+whether the $grizy$ bands will use alternative or standard visits
+remains to be determined \citedsp{PSTN-055}.
 Different cadences, filter balances, rotational dithers, etc., 
@@
 and is expected to be accomplished with 85--90\% of the observing time.
-This contiguous all-sky "deep coadd" may have region-dependent processing parameters and inputs
-(e.g., different calibration parameters or deblending algorithms for crowded fields)
-and will be of variable depth, as most special regions recieve fewer visits.
+This contiguous all-sky "deep coadd" might have region-dependent processing parameters and inputs
+where this is determined to be necessary by Rubin Data Management
+(e.g., different calibration parameters or deblending algorithms for crowded fields),
+and will be of variable depth as most special regions receive fewer visits.
+Some depth variation is also expected to come from, e.g., observing conditions and cadence tradeoffs.
 
@@
 
-Data products from Special Programs is subject to the Rubin Data Policy \citedsp{RDO-013}
+Data products from Special Programs are subject to the Rubin Data Policy \citedsp{RDO-013}
 in the same way as data products from the Main Survey.

### 2024-03-27 sec_proc.tex
JB comments

@@
 primary science goal for the Special Program.
-For example, nighly-coadd difference-image analysis for DDFs is the most
+For example, nightly-coadd difference-image analysis for DDFs is the most
 scientifically useful if done on a daily cadence, but the deeply coadded
@@
 time-domain DIA and static-sky data products described in Section~\ref{sssec:proc_reg_dr}
-should be created with Special Processing.
+could be created with Special Processing, if they are necessary 
+(i.e., Special Processing and data products will not be made if they will be redundant
+with the products of Standard Processing).
 They might be released with an annual data release or on an intermediate timescale, e.g.,

### 2024-03-27 sec_intro.tex
consistency

@@
 Data Management to images from Special Programs.
-Special Processing creates data products that are unique and separate from those produced
-by standard processing for the Main Survey.
+
+Special Processing will only be done in certain cases when it
+is determined, by Rubin Data Management, to be both "possible and necessary"
+under the definitions provided in Section~\ref{ssec:sci_sproc}.
 
@@
 as appropriate for the Special Programs' data and science goals.
-However, the development and application of specialized \emph{algorithms} or new software
-is beyond the scope of Special Processing.
-
+Special Processing creates data products that are unique and separate from those produced
+by standard processing for the Main Survey.
 Examples of Special Processing include the processing
@@
 This is a term used only in this document to refer to the image processing 
-described in the Data Management Science Pipelines Design \citeds{LDM-151} that produces the data products described in the Data Products Definitions Document
+described in the Data Management Science Pipelines Design \citeds{LDM-151}
+that produces the data products described in the Data Products Definitions Document
 (\citeds{LSE-163}) that are designed for, and will be applied to, the Main Survey's observations.
 
-In some cases, standard processing is also appropriate for Special Programs.
+In some cases, standard processing is also appropriate for Special Programs,
+as discussed in Sections \ref{ssec:sci_pproc} and \ref{ssec:proc_reg}.
 
@@
 
-User-Generated Processing for Special Programs data would be necessary in cases where
-the science goals require custom algorithms, software, or very large computational
-capacities which are beyond the scope of Special Processing or the Rubin-provided
+User-Generated Processing for Special Programs data will be necessary in cases that
+do not meet the conditions of "possible and necessary" as defined in
+Section~ref\{ssec:sci_sproc} and further illustrated in Section~\ref{ssec:proc_special}.
+This will include cases where the science goals require custom algorithms or software,
+or computational capacities which are beyond the Rubin-provided
 computational resources (Section~\ref{ssec:sci_comp}).

### 2024-03-27 sec_intro.tex
fix typo

@@
 do not meet the conditions of "possible and necessary" as defined in
-Section~ref\{ssec:sci_sproc} and further illustrated in Section~\ref{ssec:proc_special}.
+Section~\ref{ssec:sci_sproc} and further illustrated in Section~\ref{ssec:proc_special}.
 This will include cases where the science goals require custom algorithms or software,

### 2024-03-27 sec_intro.tex
add text for users

@@
 
-The three main takeaways for Rubin Data Management regarding Special Programs are:
+The three main takeaways for Rubin Data Management regarding Special Programs,
+and a fourth takeaway that is a message mainly for Rubin users, are:
 
@@
 (Section~\ref{ssec:sci_pproc})
+
+\item There will be cases where User-Generated Processing is required to meet
+science goals from Special Programs.
+(Section~\ref{ssec:proc_user})
 
@@
 Some science goals for Special Programs can be met with standard processing,
-but some will require Special Processing by Rubin Data Management or user-generated processing.
+but some will require Special Processing by Rubin Data Management or User-Generated processing.
 
@@
 
-Guidelines for User-Generated Processing, and for user-generated data products
-that can be federated with the Rubin-product data products (i.e., joinable tables),
-is forthcoming.
+Further details such as Rubin-provided support for User-Generated Processing is discussed
+in Section~\ref{ssec:proc_user}.
 

### 2024-03-27 sec_sci.tex
group user-related info
active section headings

@@
-\section{Recommendations to Enable Science from Special Programs}\label{sec:sci}
+\section{Recommendations for Rubin DM to Enable Science from Special Programs}\label{sec:sci}
 
@@
 
-\subsection{Region and Observing Mode Labels}\label{ssec:sci_labels}
+\subsection{Apply Region and Observing Mode Labels}\label{ssec:sci_labels}
 
@@
 
-\textbf{Region Labels -- } 
-Applied to a visit based on its Right Acension and Declination.
+\textbf{Region labels -- } 
+applied to a visit based on its Right Acension and Declination.
 
@@
 
-\textbf{Observing Mode Labels -- } 
-Applied to a visit based on the scheduler mode at the time of the observation
+\textbf{Observing mode labels -- } 
+applied to a visit based on the scheduler mode at the time of the observation
 (i.e., survey cadence or strategy).
@@
 
-\subsection{Special Processing}\label{ssec:sci_sproc}
+\subsection{Run Special Processing When Possible and Necessary}\label{ssec:sci_sproc}
 
@@
 
-\subsection{Prompt Processing}\label{ssec:sci_pproc}
+\subsection{Include Special Programs Data in Prompt Processing}\label{ssec:sci_pproc}
 
@@
 
-\subsection{Computational Resources for User-Generated Processing}\label{ssec:sci_comp}
+\subsection{Provide Computational Resources for User-Generated Processing}\label{ssec:sci_comp}
 
@@
 
-\subsection{Rubin Science Platform Capabilities}\label{ssec:sci_rsp}
+\subsubsection{Rubin Science Platform Capabilities}\label{sssec:sci_rsp}
 

### 2024-03-27 sec_sci.tex
add caveats to SProc

@@
 
-\textbf{Necessary -- } When the primary science goal for a Special Program 
+\textbf{Necessary -- } When the \emph{primary} science goal for a Special Program 
 cannot be met by including the data in standard processing (e.g., Prompt processing),
-or where doing so would compromise the Main Survey data products (e.g., introduce additional non-uniformity beyond what would be expected based on the survey strategy).
+or where doing so would compromise the Main Survey data products (e.g., introduce additional
+non-uniformity beyond what would be expected based on the survey strategy).
 
@@
 
+\textbf{Caveat I. The scope} of Special Processing is at the discretion of
+Rubin Data Management, including the timelines for data releases.
+
+\textbf{Caveat II. The definition of possible} comes from the requirement referenced
+for the first motivation in the list above (i.e., is not defined by this document).
+However, this definition failed to recognize cost-benefit tradeoffs to implementing
+and operating Special Processing, and does not account for the costs of operating,
+documenting, and maintaining additional processing campaigns and their data products.
+Furthermore, the science goals and complexity of Special Programs might evolve
+significantly over the 10 year survey.
+Even in situations that require no algorithmic or software development,
+the Rubin DM team might not have the staffing to handle Special Processing
+for all cases that are technically "possible".
+
+\textbf{Caveat III. The definition of necessary} is defined here in this document only.
+We emphasize that the term applies only to the \emph{primary} science goals, and does
+not cover secondary or "bespoke" science goals for a Special Program.
+Users should anticipate that User-Generated Processing will be needed for these cases.
+
 The definitions of possible and necessary are further illustrated with examples in Section~\ref{ssec:proc_special}.
-Note that secondary science goals may be considered as ``not necessary".
-
-The general scope of Special Processing, including situations in which cross-match and table joins 
-would be possible and scientifically relevant, 
-are ultimately left to the discretion of Rubin Data Management in Operations.
-
-Further discussion on Special Processing, with examples, is provided in Section~\ref{ssec:proc_special}.
 

### 2024-03-27 sec_sci.tex
JB comment

@@
 
-\textbf{Caveat I. The scope} of Special Processing is at the discretion of
-Rubin Data Management, including the timelines for data releases.
+The scope of Special Processing is at the discretion of
+Rubin Data Management, including the timelines for data releases,
+and there are two caveats to add to the above definitions.
 
-\textbf{Caveat II. The definition of possible} comes from the requirement referenced
+\textbf{Caveat I. The definition of possible} comes from the requirement referenced
 for the first motivation in the list above (i.e., is not defined by this document).
@@
 documenting, and maintaining additional processing campaigns and their data products.
-Furthermore, the science goals and complexity of Special Programs might evolve
-significantly over the 10 year survey.
 Even in situations that require no algorithmic or software development,
@@
 for all cases that are technically "possible".
+Current ballpark estimates are that the DM effort for Special Processing
+should be commeasurate with the 10\% level estimated for survey time,
+storage space, and computational resources allocated for Special Programs.
 
-\textbf{Caveat III. The definition of necessary} is defined here in this document only.
+\textbf{Caveat II. The definition of necessary} is defined here in this document only.
 We emphasize that the term applies only to the \emph{primary} science goals, and does
@@
 
-The definitions of possible and necessary are further illustrated with examples in Section~\ref{ssec:proc_special}.
+The definitions of possible and necessary are further illustrated with examples of when
+Special Processing vs. User-Generate Processing will be needed, in Sections
+\ref{ssec:proc_special} and \ref{ssec:proc_user}, respectively.
 

### 2024-03-27 sec_sci.tex
fix typo

@@
 Current ballpark estimates are that the DM effort for Special Processing
-should be commeasurate with the 10\% level estimated for survey time,
+should be commensurate with the 10\% level estimated for survey time,
 storage space, and computational resources allocated for Special Programs.

### 2024-03-27 sec_proc.tex
minor updates

@@
 
-\textbf{Potential issues with Prompt processing for untiled sequences from Special Programs.}
+\textbf{Potential issues with Prompt processing for untiled (deep drilling) sequences.}
 There are two potential issues with Prompt processing for DDFs, or any mini-survey 
@@
 
-In short, \emph{possible} means that original or reconfigured versions of the LSST
-Science Pipelines can be used , and \emph{necessary} means the primary science goal for Special Program
-could not be met without the data products produced by the Special Processing.
+There are several caveats and constraints on the definition of
+"possible and necessary" in Section~\ref{ssec:sci_sproc}, but
+in short \emph{possible} means that original or reconfigured versions of the LSST
+Science Pipelines can be used and \emph{necessary} means the \emph{primary} science goal
+of the Special Program could not otherwise be met.
 
@@
 should be made individually as part of any Special Programs that generates
-tesselated coadded images. This is \emph{not} a requirement (Section~\ref{ssec:req_dp}).
+tesselated coadded images.
+This is \emph{not} a requirement (Section~\ref{ssec:req_dp}).
 
@@
 
-The interpretations of possible and necessary, and the scope of Special Processing,
-are ultimately left to the discretion of Rubin Data Management in Operations.
-
-Specific examples are provided only to illustrate this interpretation of possible and necessary.
+The scope of Special Processing is at the discretion of Rubin Data Management.
+Specific examples are provided here as illustration.
 These examples do not place limits on what Special Processing might be done.
@@
 Rubin Data Management uses the LSST Science Pipelines to deeply
-coadd images for each field, and store the results of source
+coadd images for each field, and store the results of object
 detection and characterization in unique and separate tables that
@@
 mini-survey can be met by processing its data with standard Prompt
-Processing, and ensuring the data is properly labeled.
+Processing and ensuring the data is properly labeled.
 
@@
 Consider a Special Program in which a special region of sky is observed with standard
-visits but a a special strategy or cadence which is significantly distinct from the Main Survey
+visits but a pecial strategy or cadence which is significantly distinct from the Main Survey
 observing modes, and lasts for a limited amount of time.
@@
 time-domain DIA and static-sky data products described in Section~\ref{sssec:proc_reg_dr}
-could be created with Special Processing, if they are necessary 
-(i.e., Special Processing and data products will not be made if they will be redundant
-with the products of Standard Processing).
+could be created with Special Processing, if they are necessary.
+Special Processing and data products will not be made if they will be redundant
+with the products of Standard Processing.
 They might be released with an annual data release or on an intermediate timescale, e.g.,
@@
 and/or for which new algorithmic development or significant computational resources 
-are needed, will require user processing and user-generated data products.
+are needed, will require User-Generated Processing and user-generated data products.
 As described above, custom coadds (e.g., weekly, monthly) are also left to users 
@@
 and Data Release data products.
-Details regarding this are to be provided elsewhere.
+Guidelines for User-Generated Processing, and for user-generated data products
+that can be federated with the Rubin-product data products (i.e., joinable tables),
+is forthcoming.
 

### 2024-03-27 DMTN-065.tex
hotfix change record

@@
 recommendations for Rubin Data Management regarding processing and serving data products for Special Programs.
-
-\smallskip
 Hardware and processing boundaries on the potential diversity of data from Special Programs
@@
 \addtohist{2}{2021-12-01}{Updates per DM-20375.}{Melissa Graham}
-\addtohist{3}{2023-10-19}{Updates per DM-32723.}{Melissa Graham}
+\addtohist{3}{2024-03-01}{Updates per DM-32723.}{Melissa Graham}
 }

### 2024-03-27 ap_spcs.tex

@@
 At the end of each nightly sequence of observations, a pipeline based on recongfigured 
-components of the LSST Science Pipelines is automatically triggered.
+components of the LSST Science Pipelines could be automatically triggered.
 This pipeline creates nightly coadds in each filter and runs DIA using the template.
 Alerts are \emph{not} produced, but unique and separate catalogs with the same format
-as {\tt DiaObject} and {\tt DiaSource} are updated within 24 hours (not a requirement).
+as {\tt DiaObject} and {\tt DiaSource} could be updated within 24 hours (not a requirement).
+Instead of running daily, this processing might run weekly instead (TBD).
 At the end of each season, deeply coadded images that include all of the DDF's visits 

### 2024-03-27 sec_req.tex

@@
 on the Data Management System related to Special Programs, which are summarized here.
+
+\subsection{Metadata}\label{ssec:req_meta}
+
+\textbf{Program metadata that is sufficient to trigger Special Processing should be stored. -- }
+In order to support Special Programs processing, the LSST system is 
+required to store metadata that includes program information for every raw 
+image, such as identifiers for images obtained as part of the Main Survey 
+or a Special Program (DMS-REQ-0068).
+It is required that this metadata be sufficient for Special Programs to 
+trigger their own real-time data processing recipes ``whenever possible" 
+(DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).
 
@@
 
-\subsection{Metadata}\label{ssec:req_meta}
-
-\textbf{Program metadata that is sufficient to trigger Special Processing should be stored. -- }
-In order to support Special Programs processing, the LSST system is 
-required to store metadata that includes program information for every raw 
-image, such as identifiers for images obtained as part of the Main Survey 
-or a Special Program (DMS-REQ-0068).
-It is required that this metadata be sufficient for Special Programs to 
-trigger their own real-time data processing recipes ``whenever possible" 
-(DMS-REQ-0320), and be included in alert packets (DMS-REQ-0274).
-
 \subsection{Processing}\label{ssec:req_proc}
@@
 
-\subsection{User Processing}\label{ssec:req_ug}
-
 The 10\% of the total data processing capacity that Rubin Observatory is 
@@
 Programs data (LSR-REQ-0041).
+
+This is a minor detail but, there are two different "10\%" values quoted here.
+The accounting order is not in the requirements but Rubin Data Management considers
+the 10\% for User-Generated Processing as coming off of the total capacity first,
+with 90\% allocated for Rubin processing.
+Then the 10\% for Rubin Special Processing comes from that 90\%,
+and so is actually 9\% of the total capacity.

### 2024-04-05 ap_spcs.tex
change step 3 in ddf

@@
 This pipeline creates nightly coadds in each filter and runs DIA using the template.
-Alerts are \emph{not} produced, but unique and separate catalogs with the same format
-as {\tt DiaObject} and {\tt DiaSource} could be updated within 24 hours (not a requirement).
+Unique and separate catalogs with the same format
+as {\tt DiaObject} and {\tt DiaSource} could be updated within, e.g., 24 hours (not a requirement),
+and alert packets could be created and produced (with a flag set to identify them
+as longer-latency alerts from nightly stacked DDF images).
 Instead of running daily, this processing might run weekly instead (TBD).

### 2024-04-05 sec_proc.tex
fix typo

@@
 The Data Management team will have to test the realized
-capabilities of Prompt processing and alert distribtuion,
+capabilities of Prompt processing and alert distribution,
 and potentially impose a mitigation strategy such as

### 2024-04-09 sec_proc.tex
Update sec_proc.tex
Co-authored-by: Eric Bellm <ecbellm@uw.edu>

@@
 Consider a Special Program in which a special region of sky is observed with standard
-visits but a pecial strategy or cadence which is significantly distinct from the Main Survey
+visits but a special strategy or cadence which is significantly distinct from the Main Survey
 observing modes, and lasts for a limited amount of time.

### 2024-04-09 sec_intro.tex
fix user takeaway

@@
 \item There will be cases where User-Generated Processing is required to meet
-science goals from Special Programs.
-(Section~\ref{ssec:proc_user})
+science goals from Special Programs, and user processing will be supported with
+computational resources.
+(Section~\ref{ssec:sci_comp})
 
@@
 User-Generated Processing for Special Programs data will be necessary in cases that
-do not meet the conditions of "possible and necessary" as defined in
-Section~\ref{ssec:sci_sproc} and further illustrated in Section~\ref{ssec:proc_special}.
-This will include cases where the science goals require custom algorithms or software,
-or computational capacities which are beyond the Rubin-provided
-computational resources (Section~\ref{ssec:sci_comp}).
-
-Further details such as Rubin-provided support for User-Generated Processing is discussed
-in Section~\ref{ssec:proc_user}.
+\emph{do not} meet the conditions of "possible and necessary" as defined in Section~\ref{ssec:sci_sproc}.
+Computational tools, resources, and support will be provided by Rubin for User-Generated
+Processing, as described in Section~\ref{ssec:sci_comp} and Section~\ref{ssec:proc_user}.
 

### 2024-04-09 sec_req.tex

@@
 It is a requirement that any Special Programs processing done with the 
-Prompt pipeline (or a reconfigured version of it) is subject to the same 
+Prompt pipeline is subject to the same 
 timescales and latency constraints of 24 hours for the release of Prompt 
